### 2015-05-18 Draft2.tex
Initial import of Wei's Draft2

\documentclass[11pt]{article}
\usepackage{epsf}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath}
\usepackage{footnote}
\usepackage[bottom]{footmisc}
\usepackage[singlelinecheck=false]{caption}
\usepackage{caption}
\usepackage{amsthm}
\usepackage{enumitem}
\usepackage{graphics}
\usepackage{float}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{lscape}
\usepackage{array}
\textheight=8.7in
\usepackage{setspace}
\usepackage{verbatim}
\usepackage{multirow}

\def\threedigits#1{%
  \ifnum#1<100 0\fi
  \ifnum#1<10 0\fi
  \number#1}

\topmargin=0.1in \oddsidemargin=-0.1cm \evensidemargin=-0.1cm

\paperheight=11in \paperwidth=8.5in \marginparwidth=0in

\marginparsep=0in \textwidth=6.5in \headheight=0in \headsep=0in

\onehalfspacing
\def\argmax{\mathop{\rm arg\,max}}
%\usepackage{xspace,epsfig,subfig}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{claim}{Claim}
\newtheorem{proposition}{Proposition}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\newenvironment{sketch}{\noindent\emph{Proof Sketch:}}{$\quad \Box$}
%\newenvironment{proof}{\noindent\emph{Proof:}}{$\quad \Box$}

\bibpunct{(}{)}{;}{a}{,}{,}

\begin{document}
\title{Estimating the number of clusters using Cross Validation}
\author{Wei Fu \qquad Patrick O. Perry \\\\ New York University}
\date{}
\maketitle
\begin{abstract}
Many clustering methods, including K-means, require the user to specify the number of clusters as an input parameter. A variety of methods have been devised to choose the number of clusters automatically, but they often rely on strong modeling assumptions. We propose a data-driven approach to estimate the number of clusters based on a novel form of cross-validation. This differs from ordinary cross-validation, because clustering is fundamentally an unsupervised learning problem. Simulation and real data analysis results show that our proposed method outperforms existing methods, especially in high-dimensional settings with heavy-tailed data.
\end{abstract}

\section{Introduction}
As a main task of exploratory data analysis, clustering organizes unlabeled observations into groups such that observations in same group are more similar compare to those in different group. Clustering is an important topic in unsupervised learning because it can reveal the internal structure of data through grouping, segment the data through partitioning and summarize data for other purposes such as dimension reduction. It has being widely used in various fields such as psychology, biology, statistics and machine learning including pattern recognition, image segmentation etc.\\

After being proposed more than $50$ years, $K$-mean remains one of the most popular and widely used clustering algorithms \citep{jain2010data}. Like many other clustering methods, $K$-mean requires an input parameter $k$, the number of clusters, to be specified by the user. Automatically and quantitatively deciding such parameter is important and yet unsolved problem \citep{fujita2014non}. Various methods have been proposed to tackle this difficulty. One ad hoc approach is to explore the relationship between $W_k$ (within-cluster dispersion) and the number of cluster $k$ for a certain clustering method such as $K$-mean. Since $W_k$ decreases as $k$ increases, one usually find the ``elbow" of curve $W_k$ \textsl{VS} $k$ as the appropriate number of clusters. The example on the top row of Figure \ref{fig1} demonstrates such approach for data with $k=4$, where the ``elbow" point indeed reveals the true number of clusters. This is based on the idea that under partitioning data set has more impact than over partitioning data set in terms of $W_k$. However, locating the ``elbow" point is somewhat subjective and sometimes is not appropriate to select the optimal $k$. The second example on the bottom row of Figure \ref{fig1} shows a  situation where there is no clear choice of the ``elbow" point -- both $k=2$ and $k=3$ can be viewed as the ``elbow" point. What's more, the true $k=4$ can never be selected as the optimal $k$ using such approach in this case since it can hardly be viewed as the ``elbow" of the curve.
\begin{figure}
\begin{minipage}{\linewidth}
\centering
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{Data_example.pdf}
\end{minipage}
\hspace{0.05in}
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{W_k.pdf}
\end{minipage}
\end{minipage}
%\center{\caption{\label{fig1} Left plot shows the actual data, and the right plot shows the $W_k$ \textsl{VS} $k$ curve. }}
%\end{figure}

%\begin{figure}
\begin{minipage}{\linewidth}
\centering
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{Data_elbow.pdf}
\end{minipage}
\hspace{0.05in}
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{elbow.pdf}
\end{minipage}
\end{minipage}
\center{\caption{\label{fig1} Left plots show the actual data, right plots show the corresponding $W_k$ \textsl{VS} $k$ curve. }}
\end{figure}
\\

Recently, there are several new proposals to find the $k$ automatically. Gap statistics \citep{tibshirani2001estimating} estimates $k$ by comparing the change in within-cluster dispersion with that expected under an appropriate reference null distribution. Specifically, the graph of $log(W_k)$ is compared with its expectation under an appropriate null reference distribution of the data. The value of $k$ associated with the largest gap between $log(W_k)$ and the reference curve is selected as optimal $k$. \cite{sugar2003finding} proposed an approach which finds the number of clusters based on distortion, a quantity that measures the average distance, per dimension. It's backed by a rigorous theoretical justification based on information-theoretic ideas.   \cite{fraley2002model}'s Model-based method employs the EM algorithm to estimate the parameters in Gaussian mixture model, and select the best model ($k$) using BIC criterion. Stability-based criterion is also proposed to locate the best $k$ by some authors such as \cite{ben2001stability}, \cite{wang2010consistent} and \cite{fang2012selection}. \cite{chiang2010intelligent} provides a nice review of existing methods for finding the right $k$ in published literature. \\

Most existing methods are either model based method requires strong parameter assumptions or ad hoc method utilizes within-cluster dispersion. Although many view selecting the number of clusters as a model selection problem, very few approaches this problem from the prediction point of view. Select model with smallest prediction error via cross-validation is one of the simplest and most widely used model selection techniques in supervised learning. The lack of true class (label) in data set makes the adoption of cross-validation into unsupervised leaning problem difficult. Perhaps the only exception is \cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction strength. The strategy is to first cluster the test data and training data into $k$ clusters respectively. Then, for each pair of observations that are assigned to the same test cluster, algorithm determines whether they are also assigned to the same cluster based on the training centers. The intuition here is, if $k=k_0$, the true number of clusters, then the $k$ training set clusters will be similar to the $k$ test clusters, and hence will predict them well. However, a specifically defined prediction error measure is used in such procedure, which is quite different from the one commonly used in cross-validation procedure in supervised learning. Therefore the well understood properties of cross-validation procedure in supervised learning cannot be carried over to the prediction strength method, makes the interpretation of its result difficult. \\

Our proposed method is a complete data-driven approach which doesn't rely on any parametric assumptions. Trough novel form of partitioning data set, we are able to employing the cross-validation procedure in clustering exactly the same way as in supervised learning problem. Hence, it's easy for reader to understand the intuition behind our proposed method. Simulation and real data application shows the superior performance of our proposed method compare with existing methods in high-dimension settings and heavy-tailed data. Since the embedded cross-validation procedure is well understood, it also makes our method potentially easily to be extended in future study.

\section{Cross-Validation in finding the number of clusters}
Cross-Validation technique, one of the most commonly used and popular model selecting method in supervised learning, can not be used naively in unsupervised learning context. Since prediction error of new observation is calculated by its distance to the nearest cluster center, more cluster centers means much tighter fit to the feature space, and hence smaller distance (prediction error) of observation to its nearest cluster center. This holds even when the prediction is evaluated on an independent test set \citep{hastie2009elements}. Therefore, CV prefers larger $k$ if we do the CV naively.\\

We propose a new strategy to locate the optimal $k$ via Cross Validation. Recall that in supervised leaning, for any CV replicate, we have data in four parts
 $\begin{pmatrix}
  \mathbf{X}_{train} & \mathbf{Y}_{train}\\
  \mathbf{X}_{test}  & \mathbf{Y}_{test}\\
  \end{pmatrix}$.
One builds learner using $\left(\mathbf{X}_{train}, \mathbf{Y}_{train}\right)$ and apply it to $ \mathbf{X}_{test}$ in order to get $\hat{\mathbf{Y}}_{test}$. Difference between $\hat{\mathbf{Y}}_{test}$ and $\mathbf{Y}_{test}$ is measured as prediction error. In unsupervised leaning context, there is no $\mathbf{Y}$, so we only get $\begin{pmatrix}
  \mathbf{X}_{train}\\
  \mathbf{X}_{test}\\
  \end{pmatrix}$. \cite{perry2009cross} considered more general hold-outs in order for cross-validation to work in unsupervised learning. Specifically, two types of hold-outs are considered. The first one is Wold hold-out, which is ``speckled": we leave out random elements of the matrix $\mathbf{X}$ and use missing data algorithm like EM for prediction. The second one is Gabriel hold-out: we randomly partition the columns of $\mathbf{X}$ into ``predictors" and ``response" sets and then perform the CV as in supervised learning. This paper is focused on the Gabriel hold-out, we also developed the Wold hold-out version and included it in the comparison in simulation.  \\\\
We now give the details of how to implement the Gabriel cross-validation to locate the optimal cluster number $k$. The Wold cross-validation algorithm is described in Appendix $A$.

\subsection*{Gabriel CV algorithm}
The basic idea of Gabriel hold-out is to ``create" the label $(Y)$ for data without it, and then perform the cross-validation as in supervised learning. Specifically, given data set 
$\mathbf{X} \in \mathbb{R}^{n \times p}$, we partition it into four parts
\[
\begin{pmatrix}
  \mathbf{X}_{train} & \mathbf{Y}_{train}\\
  \mathbf{X}_{test}  & \mathbf{Y}_{test}\\
  \end{pmatrix} \equiv 
  \begin{pmatrix}
  \mathbf{X}^{m \times q } & \mathbf{X}^{m \times (p-q)}\\
  \mathbf{X}^{(n-m) \times q}  & \mathbf{X}^{(n-m) \times (p-q)}\\
  \end{pmatrix}
 \]
  where $ 1 < m < n$ and $1<q<p$. That is, we leave out $n-m$ rows of data as test set, and make $p-q$ features as ``response". Note that the data set looks like the one in supervised learning now.\\\\
To keep it consistent with the notation used in section $3$, we also denote above partition of data $\mathbf{X}$ as 
\[
\begin{pmatrix}
  \mathbf{X}_{train} & \mathbf{Y}_{train}\\
  \mathbf{X}_{test}  & \mathbf{Y}_{test}\\
  \end{pmatrix} \equiv 
  \begin{pmatrix}
\mathbf{X}^L_{train} & \mathbf{X}^R_{train}\\
  \mathbf{X}^L_{test}  & \mathbf{X}^R_{test}\\
  \end{pmatrix}
 \]
That is, data set $\mathbf{X} \in \mathbb{R}^{n \times p}$ been partitioned in column as $\mathbf{X}^L \in \mathbb{R}^{n \times L}$ and $\mathbf{X}^R \in \mathbb{R}^{n \times R}$, where $\mathbf{X}^L$ is treated as features ($L=q$) and $\mathbf{X}^R$ is treated as response ($R=p-q$).\\
\begin{itemize}
	\item For each $k = 1,2,...,k_{max}$, the Gabriel algorithm works as following
	\begin{enumerate}
 	\item A set of $k$ clusters (labels) $ \mathcal{G}= \{g_1,...,g_k\} $ are returned by applying K-means on $\mathbf{X}^R_{train}$ with parameter $k$, $\hat{\mu}^R \left(g_j\right)$ is the corresponding estimated cluster center for $g_j$, $1 \leq j \leq k$.  
 	\item Train a classifier $\mathcal{C}$ using $\left(\mathbf{X}^L_{train}, \mathcal{G} \right)$. We implement LDA in our algorithm.
 	\item Applying the classifier $\mathcal{C}$ on $\mathbf{X}^L_{test}$, we got a set of predicted clusters $\hat{\mathcal{G}} = \{\hat{g} \left(x^L_i\right) \in \mathcal{G}, i = m+1,...,n \}$, where $\hat{g} \left(x^L_i\right)$ is the predicted (assigned) cluster for observation $x_i$ in the testing data set   
 	\item Calculate the difference between $\hat{\mathcal{G}}$ and  $\mathbf{X}^R_{test}$ as
 	\[Error_k = \sum^n_{i = m+1} ||\hat{g} \left(x^L_i\right) - x^R_i||^2 = \sum^k_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2\]
 	\end{enumerate}
	Note that one can use classification technique other than LDA in step $2.$
\noindent
	\item The algorithm selects $k$ with minimum average $Error_k$ across all CV folds, that is $$\underset{k}{\operatorname{argmin}} \hspace{0.05 in} E\left(Error_k\right) $$
\end{itemize}
\section{Theoretical justification of Gabriel CV method}
This section gives the self-consistency proof of the proposed Gabriel method. What's more, its asymptotic property under suitable condition will be shown in this section in order to give the idea of why the Gabriel CV method works.\\

Because K-means algorithm is a key part in the proof below, it's convenient to have a review of K-means here:\\
Given a set of observations $\mathbf{X} = (x_1,...,x_n)$, and the number of clusters $K$, K-means' goal is to minimize the within cluster dispersion. That is, find the partition $\mathbb{P}_K$ of $\mathbf{X}$, where $\mathbb{P}_K(\mathbf{X})= (p_1,p_2,...,p_K)$, such that
\[	\underset{\mathbb{P}_K}{\operatorname{argmin}} \sum^K_{i=1}\sum_{x \in p_i } ||x-\hat{\mu}_i||^2\]
where $\hat{\mu}_i$ is the sample mean of $p_i$.
\subsection{Self-Consistency}
Same as in section $2$, let the data set $\mathbf{X} \in \mathbb{R}^{n \times p}$ been partitioned as 
 \[ 
 \begin{pmatrix}
\mathbf{X}^L_{train} & \mathbf{X}^R_{train}\\
  \mathbf{X}^L_{test}  & \mathbf{X}^R_{test}\\
  \end{pmatrix}
 \]
\subsection*{Assumptions}
\begin{enumerate}[label={Assumption},leftmargin=*]
\item $1.$ There are $K$ true centers $(\mu_1,...,\mu_K)$ where $\mathbf{X}$ generated from
\item $2.$ The $K$ partitioned centers $(\mu^L_1,...,\mu^L_K)$ are all distinct in $L$ dimensions: $||\mu^L_i-\mu^L_j|| \neq 0$ for any $i \neq j$; $(\mu^R_1,...,\mu^R_K)$ are all distinct in $R$ dimensions
\item $3.$ The training set contains at least one member of each cluster
\item $4.$ The test set contains at least one member of each cluster
\item $5.$ There is no noise, so that $$x_i = \mu \left(c_i\right) \in \{\mu_1,...,\mu_K\} $$ and $$\mu \left(c_i\right) = \mu_j \hspace{0.2in} \text{iff} \hspace{0.2in} c_i = j$$ where $c_i$ is the true cluster for observation $i$, which implies $$x^L_i = \mu^L\left(c_i\right) \in \{\mu^L_1,...,\mu^L_K\},\hspace{0.2in} x^R_i = \mu^R\left(c_i\right) \in \{\mu^R_1,...,\mu^R_K\}$$
\end{enumerate}

\begin{proposition}
	If assumptions $1, 2, 3, 5$ hold, then there exists a classification method works perfectly in case of ($k \leq K$): the re-substitution error is $0$. That is, for $\forall x_i \in \mathbf{X}_{train}$ prediction on $x^L_i$ always correctly predict cluster where $x^R_i$ is in.
\end{proposition}

\begin{theorem}
Given assumptions $1-5$, $ E\left(Error_k\right) > 0$ for $k<K$ and  $ E\left(Error_k\right) = 0$ for $k=K$. Hence if $k \leq K$, the Gabriel algorithm selects $k=K$ with probability $1$.
\end{theorem}
\noindent
Not that $k > K$ is not considered here because there are only $K$ unique points in $\mathbf{X}$. K-means will not work for parameter which is larger than the number of unique points in data $\mathbf{X}$.\\\\ 
\noindent
The proof of above theorem follows from following two lemmas:
\begin{lemma}
	If the assumptions of Theorem $1$ hold, then when $k<K$, the CV Error $>0$
\end{lemma}
\begin{proof}
	It is sufficient to prove that when $k<K$,  $Error_k>0$. By the definition of $Error_k$,
\begin{equation}
Error_k = \sum^k_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2
\end{equation}
	where $\hat{\mu}^R \left(g_j\right)$ is the center of cluster $g_j$ returned from K-means. By assumption $4$, test set contains at least one member of each cluster. Since there are $K$ true centers, and by assumption $5$, $x_i = \mu \left(c_i\right) \in \{\mu_1,...,\mu_K\} $, test set contains $K$ centers. It means there are $K$ distinct $x^R_i \left(=\mu^R \left(c_i\right)\right)$ in $\mathbf{X}^R_{test}$ set by assumption $2$. However, in equation $(1)$, there are only $k$ distinct $\hat{\mu}^R \left(g_j\right)$.  Because $k < K$, it means not every element $||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 = 0$ in equation $(1)$. Therefore,  $$Error_k = \sum^k_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 > 0$$
\end{proof}
\begin{lemma}
	If the assumptions of Theorem $1$ hold, when $k=K$, the CV Error $=0$
\end{lemma}
\begin{proof}
	It's sufficient to show that when $k=K$, $Error_k=0$. By assumption $3$ and $5$, training set contains $K$ centers. Which also implies $\mathbf{X}^R_{train}$ contains $K$ distinct $x^R_i$ $(\mu^R_1,...,\mu^R_K)$. When apply K-means on $\mathbf{X}^R_{train}$ with parameter $K$, the unique solution of 
	\[ \underset{\left(\hat{\mu}^R\left(g_1\right),...,\hat{\mu}^R\left(g_K\right)\right)}{\operatorname{argmin}} \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^R_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 \]
	is 
\begin{equation}	
	\{\hat{\mu}^R \left(g_j\right) = \mu^R_j, j=1,...,K\}
\end{equation}	
with 
\begin{equation}	 
	\hat{g} \left(x^R_i\right) = \sum^K_{j=1} g_j \cdot I_{c_i = j}
\end{equation}	
Where $\hat{g} \left(x^R_i\right) \in  \mathcal{G}$ is the assigned cluster of $x^R_i$ by K-means. One can easily check that such solution is unique and above objective function attains its lower bound $0$.\\\\
By Proposition $1$, $\mathcal{C}$ has re-substitution error $0$, which means its prediction error is $0$ on $\mathbf{X}^L_{test}$ as $\forall x^L_i \in \mathbf{X}^L_{test}, x^L_i \in \mathbf{X}^L_{train}$ by assumption $1,3,5$. Hence the predicted cluster $\hat{g} \left(x^L_i\right)$ for $x_i$ is the correct one $\hat{g} \left(x^R_i\right)$, which determined by $x^R_i$. By equation $(3)$, we have $x^R_i = \mu^R \left(c_i\right) = \mu^R_j$ if $\hat{g} \left(x^L_i\right) = \hat{g} \left(x^R_i\right) = g_j$. Therefore,
\[ Error_K = \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 =  \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - \mu^R_j||^2= 0\]
The last equation holds because of equation $(2)$.
\end{proof}

\begin{itemize}
	\item Proof of \textbf{Proposition} $1$.
	\begin{proof}
	To train classifier $\mathcal{C}$, we have training set $\left(\mathbf{X}^L_{train},\mathcal{G} \right) = \{\left(x^L_i,\hat{g} \left(x^R_i\right)\right), i =1,2,...,m\}$. By assumption $5$, $\forall i$ $x_i = \mu_j$, $j=1,2,...,K$ and $k \leq K$. Hence the training set has $K$ unique pairs $\{\left(\mu^L_j, \hat{g} \left(\mu^R_j\right)\right),j=1,2,...,K\}$, with $K$ unique $\mu^L_j$ by assumption $3$ and $k$ unique $\hat{g} \left(\mu^R_j\right)$. This is because each pair $(\mu^L_j,\mu^R_j)$ is unique by assumption $2$ and $\left(\mu^R_j,\hat{g} \left(\mu^R_j\right)\right)$ is unique since K-means assigns one and only one label to each $\mu^R_j$. Define the classifier $\mathcal{C}$ as following: \[ \mathcal{C}(x^L_i) = \sum^K_{j=1}\hat{g} \left(\mu^R_j\right) \cdot I_{\{c_i = j\}}  \] It's easy to see such classifier has re-substitution error $0$. 
	\end{proof}
	\item Proof of \textbf{Proposition} $1$ when the classifier is LDA.
	\begin{proof}
	Since Proposition $1$ is only used in Lemma $2$, therefore it's sufficient to prove it in case of $k = K$. Note that training set contains $K$ unique pairs $\{\left(\mu^L_j, \hat{g} \left(\mu^R_j\right)\right), j=1,2,...,K\}$, with $K$ unique $\mu^L_j$ and $K$ unique $\hat{g} \left(\mu^R_j\right)$ now, where $\mu^L_j, \hat{g} \left(\mu^R_j\right)$ has one-to-one correspondence, i.e.
\[ \left\lbrace\left(\mu^L_j,\hat{g} \left(\mu^R_j\right) \right), j=1,2,...,K\right\rbrace =  \left\lbrace\left(\mu^L_j, g_j \right), j=1,2,...,K\right\rbrace \]
By equation $(3)$.\\\\
Let the prior probability of class $g_j$ be $\pi_j, \hspace{0.05in} j=1,...,K$. By assumption $3$, $\pi_j > 0$. The class-conditional density of $X$ in class $\mathcal{G} = g_j$ is $$f_j\left(x_i\right) = \mathbb{P}\left( X=x_i | \mathcal{G} = g_j \right) $$ where $x_i \in \{\mu^L_1,...,\mu^L_K\}$ and $f_j\left(x_i\right)$ has multivariate Gaussian distribution
\[ f_j\left(x_i\right) = \frac{1}{(2\pi)^{p/2}|\Sigma_j|^{1/2}}e^{-\frac{1}{2}(x_i-\mu_j)^T\Sigma^{-1}_j(x_i-\mu_j)	}\]
LDA assumes a commom covariance matrix $\Sigma_j = \Sigma \hspace{0.1in} \forall j$.
The posterior probability is 
\[	\mathbb{P}\left( \mathcal{G} = g_j | X=x_i  \right)  = \frac{f_j\left(x_i\right) \pi_j}{\sum^K_{l=1}f_l\left(x_i\right) \pi_l}	\]
In practice, $\pi_j, \mu_j$ and $\Sigma$ are estimated from data by
 \begin{itemize}
 	\item  $\hat{\pi}_j = \frac{N_j}{N}$, where $N_j$ is the number of class-$g_j$ observation in training set and $N$ is the total observation in training set
 	\item  $\hat{\mu}_j = \sum_{c_i=j} x_i/N_j$ 
 	\item  $\hat{\Sigma} = \frac{1}{N-K} \sum^K_{j=1}\sum_{c_i=j} (x_i - \hat{\mu}_j )(x_i - \hat{\mu}_j )^T$
 \end{itemize}
In our case, $\hat{\mu}_j = \mu^L_j$ and $\hat{\Sigma}$ is zero matrix by construction. But in real case, the estimate of within-group covariance $\hat{\Sigma}$ is a positive-definite matrix. We therefore assume $\hat{\Sigma}$ to be any positive-definite matrix. Also, we assume the observations in training set are evenly distributed across all class, i.e. $\pi_j$ is a positive constant for all $j = 1,...,K$ \\\\   
LDA predicts class for each $x_i = \mu^L(c_i)\in \{\mu^L_1,...,\mu^L_K\}$ by maximizing the posterior probability
\begin{align}
     \hat{\mathcal{G}}(x_i) & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} \mathbb{P}\left( \mathcal{G} = g_j | X=x_i  \right)  \nonumber   \\ 
     & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} f_j\left(x_i\right) \pi_j \nonumber \nonumber  \\
     & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} f_j\left(x_i\right) \nonumber  \nonumber  \\    
     & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in}  \frac{1}{(2\pi)^{p/2}|\hat{\Sigma}|^{1/2}}e^{-\frac{1}{2}(x_i-\mu^L_j)^T\hat{\Sigma}^{-1}(x_i-\mu^L_j)	} \nonumber  \\  
     &=  \underset{g_j}{\operatorname{argmin}}\hspace{0.05 in} (x_i-\mu^L_j)^T\hat{\Sigma}^{-1}(x_i-\mu^L_j) \nonumber   \\  
     &= \sum^K_{j=1} g_j \cdot I_{c_i = j}
\end{align}
The last equation holds because $\hat{\Sigma}$ is assumed to be positive-definite, so is its inverse  $\hat{\Sigma}^{-1}$. Hence $(x_i-\mu^L_j)^T\hat{\Sigma}^{-1}(x_i-\mu^L_j) >0$ for any non-zero vector $(x_i-\mu^L_j)$. The lower bound $0$ is attained only when $(x_i-\mu^L_j)$ is a zero vector, which is only possible when $x_i=\mu^L(c_i) = \mu^L_j$, i.e. $c_i = j$ by assumption $2$ and $5$. \\\\
Because the RHS of equation $(4)$ coincide with RHS of equation $(3)$, the proof is complete.
\end{proof}
\end{itemize}
\subsection{Well-separated clusters with noise}
Same with previous, the data set $\mathbf{X} \in \mathbb{R}^{n \times p}$ been partitioned as $\mathbf{X}^L \in \mathbb{R}^{n \times L}$ and $\mathbf{X}^R \in \mathbb{R}^{n \times R}$, with following set of assumptions:
\begin{enumerate}[label={Assumption},leftmargin=*]
	\item $1.$ There are $K$ true centers $(\mu_1,...,\mu_K)$ where $\mathbf{X}$ generated from
	\item $2.$ The $K$ partitioned centers $(\mu^L_1,...,\mu^L_K)$ are all distinct in $L$ dimensions: $||\mu^L_i-\mu^L_j|| \neq 0$ for any $i \neq j$; $(\mu^R_1,...,\mu^R_K)$ are all distinct in $R$ dimensions
	\item $3.$ There is noise, i.e. $$x_i = \mu \left(c_i\right) + \varepsilon_i, \hspace{0.2 in} \mu \left(c_i\right) \in \{\mu_1,...,\mu_K\}$$ where vector $\varepsilon_i$ is i.i.d with zero mean and covariance matrix $\Sigma$, which is independent from any $\mu_j \in (\mu_1,...,\mu_K)$, $c_i$ is the true cluster for observation $i$, which implies $$x^L_i = \mu^L \left(c_i\right) + \varepsilon^L_i,\hspace{0.2in} x^R_i = \mu^R \left(c_i\right) + \varepsilon^R_i$$ 
with $$\mu^L\left(c_i\right) \in \{\mu^L_1,...,\mu^L_K\},\hspace{0.2in} \mu^R\left(c_i\right) \in \{\mu^R_1,...,\mu^R_K\} $$\
where $\varepsilon^L_i$ and $\varepsilon^R_i$ are independent and both have zero mean and covariance matrix $\Sigma^L$ and $\Sigma^R$ respectively, so $\Sigma = 
\begin{pmatrix}
  \Sigma^L &0\\
  0  &\Sigma^R\\
  \end{pmatrix}$
. Also $\varepsilon^R_i$ is bounded $||\varepsilon^R_i|| \leq 1$.
	\item $4.$ Clusters are well separated, i.e. \[ ||x_i-\mu \left(c_i\right)|| < ||x_i-\mu_{j}|| \hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i \] and \[ ||x^L_i-\mu^L \left(c_i\right)|| < ||x^L_i-\mu^L_{j}||-\delta,  \hspace{0.2 in}||x^R_i-\mu^R \left(c_i\right)|| < ||x^R_i-\mu^R_{j}||-\delta \hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i\] where $\delta > 0$ is a small constant 
	\item $5.$ The training set contains $n^{train}_j>0$ observations from each cluster $j$ and the test set contains $n^{test}_j > 0$ observations from each cluster $j$, $j=1,2...,K$
	\item $6.$ The prior for $x_i$ come from cluster $\mu_j$ is $\pi_j > 0$. So when the total number of observation in training set $m$ goes infinity, $\lim_{m\to \infty} n^{train}_j = m \cdot \pi_j $. Also $m=n\cdot r$, $n-m = n\cdot (1-r)$, where $0<r<1$ is a positive constant.
\end{enumerate}

\subsubsection{Consistent assumption for K-means}
\cite{pollard1981strong} found conditions that ensure the almost sure convergence of the set of cluster centers of K-means as the sample size increase. This section mainly introduce the theorem from \cite{pollard1981strong} and show the assumptions needed for our own theorem.\\\\  
Applying K-means clustering on i.i.d observations $x_1,x_2,...x_m$ from population $P$ with parameter $k$ is to choose cluster centers $a_1,a_2,...,a_k$ to minimize
\[	W_m = \frac{1}{m}\sum^m_{l=1}min_{1\leq j \leq k}||x_l-a_j||^2	\]
It's convenient to treat $W_m$ as a function of the set of cluster centers $A$ and of the  empirical distribution $P_m$ obtained from the sample by placing mass $m^{-1}$ at each of $x_1,x_2,...x_m$. Then the problem is to minimize 
\[ W(A,P_m):=\int min_{a\in A}||x-a||^2P_m(dx)\]
Over all possible choices of the set $A$ containing $k$ points. \\\\
For each probability measure $Q$ and each set $A$, defined
\[	\Phi(A,Q):= \int min_{a\in A}||x-a||^2 Q(dx)	\] and \[ \mathfrak{m}_k(Q):=inf\{\Phi(A,Q):A \hspace{0.05in} \text{contains} \hspace{0.05in} k  \hspace{0.05in} \text{or fewer points}\} \]
The theorem proved by \cite{pollard1981strong} says that
\begin{theorem}
	Suppose that $\int ||x||^2P(dx) < \infty$ and that for each $j=1,2,...,k$ there is a unique set $\bar{A}(j)$ for which $\Phi(\bar{A}(j),P)=\mathfrak{m}_j(P)$. Then $A_m \to \bar{A}(k)$ a.s and $\Phi(A_m,P_m) \to \mathfrak{m}_k(P)$ a.s 
\end{theorem}
\noindent
Where $A_m = A_m(k)$ is the set of optimal sample cluster centers satisfy $\Phi(A_m,P_m) = \mathfrak{m}_k(P_m)$ and $\bar{A}=\bar{A}(k)$ is the set of optimal population cluster center satisfy  $\Phi(\bar{A},P) = \mathfrak{m}_k(P)$. Such theorem gives the condition for the sample estimator of cluster centers be consistent estimator of optimal cluster centers.\\

In our setting, observations $x^R_1, x^R_2,...,x^R_m$ can be viewed as i.i.d observations from same distribution $\mathcal{P}$, which is a mixed distribution of $K$ distinct distributions with prior $\pi_1,\pi_2,...,\pi_K$.
\begin{theorem}
Suppose assumption $3$ holds and for each $j=1,2,...,k_{max}$ there is a unique set $\bar{A}(j)$ for which $\Phi(\bar{A}(j),\mathcal{P})=\mathfrak{m}_j(\mathcal{P})$, also $ \bar{A}(K) = \{\mu^R_1,\mu^R_2,...,\mu^R_K\}$. Then for any $k \leq k_{max}$, $A_m(k) \to \bar{A}(k)$ a.s.
\end{theorem}
\begin{proof}
By assumption $3$, $||\varepsilon^R_i||\leq 1$, which implies the first condition in theorem $2$ holds, i.e.
\[\int ||x||^2P(dx) < \infty\] 
Since the second condition of theorem $2$ is given as assumption in theorem $3$, we have almost sure convergence of cluster centers $A_m(k)$ for any $k \leq k_{max}$ by theorem $2$. 
\end{proof}
\noindent
In the rest of this section, the assumptions (conditions) given in theorem $3$ is called consistent assumptions. 
\subsubsection{Asymptotic property of Gabriel method}
\begin{definition}
Define event\\ $\mathcal{D} = \{\forall x_i \in \mathbf{X}, ||x^L_i-\hat{\mu}^L \left(c_i\right)|| < ||x^L_i-\hat{\mu}^L_{j}||, ||x^R_i-\hat{\mu}^R \left(c_i\right)|| < ||x^R_i-\hat{\mu}^R_{j}|| \hspace{0.2 in} \forall \hspace{0.01 in}j\neq c_i\}$, where $\hat{\mu}^R_{j}$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^R_{train}$and $\hat{\mu}^L_{j}$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^L_{train}$
\end{definition}

\begin{proposition}
Under assumptions $1-6$, $P(\mathcal{D}) \to 1$ with $m \to \infty$.
\end{proposition}
\begin{proof}
	By assumption $4$, \[  ||x^L_i-\mu^L \left(c_i\right)|| < ||x^L_i-\mu^L_{j}||-\delta \hspace{0.2in} \forall \hspace{0.01 in}j\neq c_i \] 
	Hence, 
	\begin{align} \nonumber
		||x^L_i-\hat{\mu}^L \left(c_i\right)|| & \leq \nonumber ||x^L_i-\mu^L \left(c_i\right)||+ ||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) ||\\ \nonumber
		&< ||x^L_i-\mu^L_{j}||-\delta + ||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) ||  \\ \nonumber
		& \leq ||x^L_i-\hat{\mu}^L_{j}||+||\mu^L_{j}-\hat{\mu}^L_{j}||-\delta + ||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) ||  \\ \nonumber
		& =  ||x^L_i-\hat{\mu}^L_{j}||-\left\lbrace \delta-||\mu^L_{j}-\hat{\mu}^L_{j}|| -||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) || \right\rbrace \\ 
m \to \infty		&<||x^L_i-\hat{\mu}^L_{j}||
	\end{align} 
The step $(5)$ is because for any given $\varepsilon > 0$, by Markov inequality
	\begin{align} \nonumber
	  P\left(||\mu^L_{j}-\hat{\mu}^L_{j}||^2 \geq \varepsilon \right) &\leq \frac{E||\mu^L_{j}-\hat{\mu}^L_{j}||^2}{\varepsilon}  \nonumber \\ \nonumber
	    &=  \frac{E\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)^T\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)}{\varepsilon}  \\ \nonumber
	    &=  \frac{E[ tr\left\lbrace\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)^T\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)\right\rbrace]}{\varepsilon}  \\ \nonumber
	   &=  \frac{E[ tr\left\lbrace\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)^T\right\rbrace]}{\varepsilon}  \\ \nonumber
	  &=  \frac{ tr\left\lbrace E\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)^T\right\rbrace}{\varepsilon}  \\ \nonumber
	  &= \frac{ tr\{\Sigma^L \}}{n^{train}_j \cdot \varepsilon}   \\ \nonumber
	% \lim_{m\to \infty} n^{train}_j = m \cdot \pi_j \to \infty \hspace{1in} &\to 0  \\ \nonumber
	\end{align} 
So
\begin{align} \nonumber
  max_{j \in \{1,2,...,K\}} \hspace{0.05in} P\left(||\mu^L_{j}-\hat{\mu}^L_{j}||^2 \geq \varepsilon \right) &\leq max_{j \in \{1,2,...,K\}} \hspace{0.05in}\frac{E||\mu^L_{j}-\hat{\mu}^L_{j}||^2}{\varepsilon}  \nonumber \\ \nonumber
  &=max_{j \in \{1,2,...,K\}} \hspace{0.05in}  \frac{ tr\{\Sigma^L \}}{n^{train}_j \cdot \varepsilon}   \\ \nonumber
m \to \infty \hspace{0.3in}  &= \frac{ tr\{\Sigma^L \}}{m\hspace{0.02in}\pi^*\hspace{0.02in} \varepsilon}  \hspace{0.5in} \text{where}\hspace{0.1in}\pi^* = min_{j \in \{1,2,...,K\}}\pi_j\\ \nonumber
&\to 0
\end{align}

Therefore $\forall j, \hspace{0.05in} ||\mu^L_{j}-\hat{\mu}^L_{j}|| \to 0$ uniformly as $m \to \infty$. Hence, 
\begin{equation}
\lim_{m\to \infty} \left\lbrace \delta-||\mu^L_{j}-\hat{\mu}^L_{j}|| -||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) || \right\rbrace >0
\end{equation}

Same argument yields 
\[||x^R_i-\hat{\mu}^R \left(c_i\right)|| < ||x^R_i-\hat{\mu}^R_{j}||\]
Because $(6)$ holds uniformly $\forall j\in\{1,2,...,K\}$, we have 
\[ \lim_{m \to \infty}P(\mathcal{D})  =1\]
	\end{proof}
\begin{proposition}
Given assumption $1-6$ and the consistent assumptions of Theorem $3$, when $m \to \infty$, $P\left[A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}\right]=1$, where $\hat{\mu}^R_j$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^R_{train}$
\end{proposition}
\begin{proof}
By theorem $3$, we know $A_m(K) \to \bar{A}(K)$ a.s. and $\bar{A}(K) = \{\mu^R_1,\mu^R_2,...,\mu^R_K\}$. Denote $A_m(K) = \{a_1, a_2,..., a_K\}$, by almost sure convergence of $A_m(K)$ to $\bar{A}(K)$, we have\\\\
For all $\varepsilon > 0$, there exists $N_{\varepsilon}$ such that if $m > N_{\varepsilon}$, then
\[ P\left(||a_j-\mu^R_j||< \varepsilon\hspace{0.1in}\forall j\right)=1\]
 Now, take $\varepsilon = \frac{\delta}{2}$, then exists $N_{\delta}$ such that for all $m > N_{\delta}$
\[P\left(||a_j-\mu^R_j||<\frac{\delta}{2} \hspace{0.1in}\forall j\right)=1\]
By assumption $4$, 
\[||x^R_i-\mu^R \left(c_i\right)|| < ||x^R_i-\mu^R_{j}||-\delta \hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i\]
Let $m > N_{\delta}$, for all $x^R_i \in \mathbf{X}^R_{train}$, with probability $1$

\begin{align}
  ||x^R_i - a(c_i)||&\leq ||x^R_i-\mu^R(c_i)|| + ||\mu^R(c_i)-a(c_i)||\nonumber \\ \nonumber
   &< ||x^R_i-\mu^R_{j}||-\delta + ||\mu^R(c_i)-a(c_i)|| \hspace{0.2 in} \forall \hspace{0.01 in}j\neq c_i  \\ \nonumber
    &\leq ||x^R_i-a_{j}|| + ||a_j - \mu^R_{j}||-\delta + ||\mu^R(c_i)-a(c_i)||  \\ \nonumber
 ||\mu^R(c_i)-a(c_i)||< \frac{\delta}{2} \hspace{0.05in}\& \hspace{0.05in}||a_j - \mu^R_{j}||< \frac{\delta}{2} \hspace{0.2in} &<||x^R_i-a_{j}||\hspace{0.2 in}  \forall \hspace{0.01 in}j\neq c_i 
\end{align}
Hence, $x^R_i$ is assigned to cluster center $a(c_i)$ by K-means. Note also that observations from same cluster are assigned to the same center. Since $a_j$ is the sample mean of all the observations assigned to it, we have
\[a_j = \hat{\mu}^R_j \hspace{0.1in}\forall j\]
 where $\hat{\mu}^R_j$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^R_{train}$. The proof is complete.
\end{proof}
\begin{proposition}
Conditioning on $\mathcal{D}$ and $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$, there exists a classification method works perfectly in case of $k = K$: the prediction error is $0$. That is, $\forall x_i \in \mathbf{X}_{test}$ prediction on $x^L_i$ always correctly predict cluster where $x^R_i$ is in. 
\end{proposition}

\begin{proof}
Given $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$,  we know from the proof of proposition $3$ that when applying K-means on $\mathbf{X}^R_{train}$ with parameter $K$, the solution of
\[ \underset{\left(\hat{\mu}^R\left(g_1\right),...,\hat{\mu}^R\left(g_K\right)\right)}{\operatorname{argmin}} \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^R_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 \]
is 
\begin{equation}
\{\hat{\mu}^R\left(g_j\right) = \hat{\mu}^R_{j}, j=1,...,K\}
\end{equation}
with 
\begin{equation}
\hat{g} \left(x^R_i\right) = \sum^K_{j=1} g_j \cdot I_{c_i = j}
\end{equation}
where $\hat{g} \left(x^R_i\right)$ is the assigned cluster by K-means for observation $x^R_i \in \mathbf{X}^R_{train}$  , $\hat{\mu}^R\left(g_j\right)$ is the sample mean (centroid) of K-means returned cluster $g_j$, $\hat{\mu}^R_{j}$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^R_{train}$, $c_i$ is the true cluster for observation $i$.\\\\
The training set of $\mathcal{C}$ consists of data in the form of 
\[
\left(\mathbf{X}^L_{train},\mathcal{G} \right) = \left\lbrace \left(x^L_i,\hat{g} \left(x^R_i\right)\right)\right\rbrace, \hspace{0.2in} i =1,2,...,m
 \]
From $(8)$, we learned that observations from the same cluster have been clustered together in the training set. 
And if $g_j = \hat{g} \left(x^R_i\right)$ for observation $i$
\[x^R_i = \mu^R \left(c_i\right) + \varepsilon^R_i = \mu^R_{j} + \varepsilon^R_i\] 
Which implies  
\[x^L_i = \mu^L \left(c_i\right) + \varepsilon^L_i = \mu^L_{j} + \varepsilon^L_i\] 
Hence training set of $\mathcal{C}$ consists of data in the form of 
\[ \left(\mathbf{X}^L_{train},\mathcal{G} \right) = \left\lbrace\left(\mu^L_{j} + \varepsilon^L_i, g_j \right)\right\rbrace\hspace{0.2in} j =1,2,...,K, \hspace{0.1in} i=1,...,m\]
So Let $\hat{\mu}^L_{j} = \frac{\sum x^L_i \cdot I_{c_i=j} }{n^{train}_j}$, $x^L_i  \in \mathbf{X}^L_{train}$ denotes the sample mean of each cluster in $\mathbf{X}^L_{train}$. Define following classifier
	 \[ \mathcal{C}(x^L_i) = \sum^K_{j=1} g_j \cdot I_{h(i) = j} \]
	 where\[h(i) = \underset{j}{\operatorname{argmin}} \hspace{0.05 in}||x^L_i - \hat{\mu}^L_j|| \]
	Note here $x^L_i  \in \mathbf{X}^L_{test}$.\\

Then, above classifier has prediction error $0$. To prove it, we need to show above classifier $\mathcal{C}(x^L_i)$ always gives the same cluster as the one in equation $(8)$, i.e. 
\begin{equation}
\mathcal{C}(x^L_i) = \sum^K_{j=1} g_j \cdot I_{c_i = j}
\end{equation}
Given $\mathcal{D}$, we have
\[||x^L_i-\hat{\mu}^L \left(c_i\right)|| < ||x^L_i-\hat{\mu}^L_{j}||\hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i\]
Which means
\[h(i) = \underset{j}{\operatorname{argmin}} \hspace{0.05 in}||x^L_i - \hat{\mu}^L_j|| = c_i \]
$(9)$ follows directly.
\end{proof}
\noindent
If we further assume equal prior are observed in training set, then we can show that the LDA classifier has prediction error $0$, that is
\begin{proposition}
Conditioning on $\mathcal{D}$ and $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$ and $n^{train}_j=\frac{m}{K}$ $\forall j$ in training set, the classifier $\mathcal{C}$ trained using LDA has prediction error $0$ in case of $k = K$.
\end{proposition}
\begin{proof}
Same as in the proof of proposition $4$, given $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$, the training set of $\mathcal{C}$ consists of data in the form of 
\[ \left(\mathbf{X}^L_{train},\mathcal{G} \right) = \left\lbrace\left(\mu^L_{j} + \varepsilon^L_i, g_j \right)\right\rbrace\hspace{0.2in} j =1,2,...,K, \hspace{0.1in} i=1,...,m\]
The class-conditional density of $X^L$ in class $\mathcal{G} = g_j$ is $$f_j\left(x^L_i\right) = \mathbb{P}\left( X^L=x^L_i | \mathcal{G} = g_j \right) $$ where $x^L_i=\mu^L(c_i)+\varepsilon^L_i$ and $f_j\left(x^L_i\right)$ has multivariate Gaussian distribution
\[ f_j\left(x^L_i\right) = \frac{1}{(2\pi)^{L/2}|\Sigma^L_j|^{1/2}}e^{-\frac{1}{2}(x^L_i-\mu^L_j)^T(\Sigma^L_j)^{-1}(x^L_i-\mu^L_j)	}\]
LDA assumes a commom covariance matrix $\Sigma^L_j = \Sigma^L \hspace{0.1in} \forall j$.
The posterior probability is 
\[	\mathbb{P}\left( \mathcal{G} = g_j | X^L=x^L_i  \right)  = \frac{f_j\left(x^L_i\right) \pi_j}{\sum^K_{l=1}f_l\left(x^L_i\right) \pi_l}	\]
where $ \pi_j$ is the prior for class $g_j$. In practice, $\pi_j, \mu^L_j$ and $\Sigma^L$ are estimated from data by
 \begin{itemize}
 	\item  $\hat{\pi}_j = \frac{n^{train}_j}{m}$, where $n^{train}_j$ is the number of class-$g_j$ observation in training set and $m$ is the total observations in training set
 	\item  $\hat{\mu}^L_{j} = \frac{\sum x^L_i \cdot I_{c_i=j} }{n^{train}_j}$, $x^L_i  \in \mathbf{X}^L_{train}$
 	\item  $\hat{\Sigma}^L = \frac{1}{m-K} \sum^K_{j=1}\sum_{c_i=j} (x^L_i - \hat{\mu}^L_j )(x^L_i - \hat{\mu}^L_j )^T$
 \end{itemize}
Because $\forall j$ $n^{train}_j=\frac{m}{K}$ , $\hat{\pi}_j = \frac{1}{K}$ $\forall j$. LDA predicts class for each $x^L_i  \in \mathbf{X}^L_{test}$ by maximizing the posterior probability
\begin{align}
     \hat{\mathcal{G}}(x_i) & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} \mathbb{P}\left( \mathcal{G} = g_j | X^L=x^L_i  \right)  \nonumber   \\ 
     & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} f_j\left(x^L_i\right) \hat{\pi}_j \nonumber \nonumber  \\
   \hat{\pi}_j = \frac{1}{K} \hspace{0.1in}\forall j  \hspace{0.2in} & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} f_j\left(x^L_i\right) \nonumber  \nonumber  \\    
     & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in}  \frac{1}{(2\pi)^{L/2}|\hat{\Sigma}^L|^{1/2}}e^{-\frac{1}{2}(x^L_i-\hat{\mu}^L_j)^T(\hat{\Sigma}^{L})^{-1}(x^L_i-\hat{\mu}^L_j)	} \nonumber  \\  
     &=  \underset{g_j}{\operatorname{argmin}}\hspace{0.05 in} (x^L_i-\hat{\mu}^L_j)^T(\hat{\Sigma}^{L})^{-1}(x^L_i-\hat{\mu}^L_j) \nonumber   \\   \nonumber 
     &= \sum^K_{j=1} g_j \cdot I_{c_i=j}
\end{align}
The last equation is because given $\mathcal{D}$, we have
\[||x^L_i-\hat{\mu}^L \left(c_i\right)|| < ||x^L_i-\hat{\mu}^L_{j}||\hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i\]
Since it coincides with equation $9$, the proof is complete. 
\end{proof}
\vspace{0.3in}
\noindent
By definition,
\begin{equation}
Error_k = \sum^k_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 = \sum^{n-m}_{i=1} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2, \hspace{0.1in}with\hspace{0.1in} g_j = \hat{g}(x^L_i)
\end{equation}
\noindent
For any parameter $k\leq k_{max}$, $x_i \in \mathbf{X}_{test}$ and $||\hat{\mu}^R \left(g_j\right) - x^R_i||^2$ in $(10)$ are i.i.d with finite expectation
\[
E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 < \infty
\]
Since $||\hat{\mu}^R \left(g_j\right) - x^R_i||$ is bounded 
\begin{align} \nonumber 
||\hat{\mu}^R \left(g_j\right) - x^R_i|| & = ||\hat{\mu}^R \left(g_j\right) - \mu^R(c_i)- \varepsilon^R_i|| \\ \nonumber 
&\leq ||\hat{\mu}^R \left(g_j\right) - \mu^R(c_i)|| + ||\varepsilon^R_i||\\ \nonumber 
&\leq \max_{i\neq j}||\mu^R_i - \mu^R_j|| + 2||\varepsilon^R_i||\\ \nonumber 
\end{align}
Given $||\varepsilon^R_i||\leq 1$.
Therefore, by strong law of large number (SLLN), when $(n-m) \to \infty$
\begin{equation}
\frac{1}{n-m} Error_k = \frac{1}{n-m}\sum^{n-m}_{i=1}||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 \hspace{0.1in}  \to \hspace{0.1in} E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 \hspace{0.1in}a.s. \hspace{0.1in} where \hspace{0.1in} g_j = \hat{g}(x^L_i)
\end{equation}

\begin{lemma}
	If assumptions $1-6$ and the consistent assumptions of Theorem $3$ hold, when $k=K$, $\frac{1}{n-m} Error_K \to tr\{\Sigma^R\}$ a.s as $n \to \infty$
\end{lemma}
\begin{proof}
Since $m= n \cdot r$ and $(n-m) = n \cdot (1-r)$, $n \to \infty$ implies $(n-m) \to \infty$ and $m \to \infty$, where event $\mathcal{D}$ and $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$ happen with probability $1$ by Proposition $2$ and Proposition $3$. By Proposition $4$, $\mathcal{C}$ has prediction error $0$ on $\mathbf{X}^L_{test}$. Hence the predicted cluster $\hat{g} \left(x^L_i\right)$ for $x^L_i$ is the correct one $\hat{g} \left(x^R_i\right)$, which determined by $x^R_i$. By $(8)$, if $g_j = \hat{g} \left(x^R_i\right) = \hat{g} \left(x^L_i\right)$
\[x^R_i = \mu^R \left(c_i\right) + \varepsilon^R_i = \mu^R_{j} + \varepsilon^R_i\] 
\noindent
Therefore,
\begin{align} \nonumber
    Error_K &= \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2   \\ \nonumber
     & = \sum^K_{j=1} \sum_{g_j} ||\hat{\mu}^R \left(g_j\right) -\mu^R_{j} - \varepsilon^R_i||^2 \\ \nonumber 
by \hspace{0.1in} (7) \hspace{0.2in}   & = \sum^K_{j=1} \sum_{g_j} ||\hat{\mu}^R_{j} -\mu^R_{j} - \varepsilon^R_i||^2 \\ 
    % &=  \sum^{n-m}_{i=1}||\hat{\mu}^R_{j} -\mu^R_{j} - \varepsilon^R_i||^2
    &= \sum^K_{j=1} \sum_{g_j}  ||\hat{\mu}^R_{j} -\mu^R_{j}||^2 + \sum^{n-m}_{i=1}||\varepsilon^R_i||^2
\end{align}
$(12)$ is true since $\varepsilon^R_i$ is independent from $\mu_j$ and $\varepsilon^L_i$ by assumption $3$. So because $\hat{\mu}^R \left(g_j\right)$ is a function of $x^L_i = \mu^L_i + \varepsilon^L_i$, $\varepsilon^R_i$ is also independent from $\hat{\mu}^R \left(g_j\right)$. \\

Hence
\begin{align} \nonumber
  E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2  & = \sum^K_{j=1}E\left(||\hat{\mu}^R \left(g_j\right) - x^R_i||^2  \hspace{0.1 in}| \hspace{0.1 in} x^R_i = \mu^R_{j} + \varepsilon^R_i\right) P\left(x^R_i = \mu^R_{j} + \varepsilon^R_i\right)\\ \nonumber
 % & = \sum^K_{j=1}E||\hat{\mu}^R_{j} -\mu^R_{j} - \varepsilon^R_i||^2\cdot\pi_j\\ \nonumber
   & = \sum^K_{j=1}E\left(||\hat{\mu}^R_{j} -\mu^R_{j}||^2 +||\varepsilon^R_i||^2\right)\cdot\pi_j \\ \nonumber
   &= \sum^K_{j=1}E||\hat{\mu}^R_{j} -\mu^R_{j}||^2\cdot\pi_j + \sum^K_{j=1}E||\varepsilon^R_i||^2\cdot\pi_j\\ \nonumber
    &= \sum^K_{j=1} \frac{tr\{\Sigma^R\}}{n^{train}_j}\cdot\pi_j + E||\varepsilon^R_i||^2\\ \nonumber
  m \to \infty \hspace{0.2in}  &= \sum^K_{j=1} \frac{tr\{\Sigma^R\}}{m\cdot\pi_j }\cdot\pi_j + E||\varepsilon^R_i||^2\\ \nonumber
      &= E||\varepsilon^R_i||^2\\ \nonumber
       &=tr\{\Sigma^R\} \\ \nonumber
\end{align} 
By $(11)$, $\frac{1}{n-m} Error_K \to tr\{\Sigma^R\}$ a.s as $n \to \infty$.
\end{proof}
\begin{lemma}
	If assumptions $1-6$ and Assumption I.(defined below) hold, as well as the consistent assumptions of Theorem $3$ and assumptions of Theorem $4$ (defined below) hold,  then if $k \neq K$, $\frac{1}{n-m} Error_K \to \Theta$ a.s as $n \to \infty$, and $\Theta > tr\{\Sigma^R\} $
\end{lemma}
\begin{proof}
The proof of almost surely convergent part follows from $(11)$ directly, so here we only need to show 
\[\Theta = E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 > tr\{\Sigma^R\}, \hspace{0.2in} where \hspace{0.1in}g_j=\hat{g}(x^L_i) \]
Because 
\begin{align} \nonumber
  E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2  &=  E||\hat{\mu}^R \left(\hat{g}(x^L_i)\right) - x^R_i||^2 \\ \nonumber
  & = \sum^K_{j=1}E\left(||\hat{\mu}^R \left(\hat{g}(x^L_i)\right) - x^R_i||^2  \hspace{0.1 in}| \hspace{0.1 in} x^R_i = \mu^R_{j} + \varepsilon^R_i\right) P\left(x^R_i = \mu^R_{j} + \varepsilon^R_i\right)\\ \nonumber
  &= \sum^K_{j=1}E\left(||\hat{\mu}^R \left(\hat{g}(x^L_i)\right) - \mu^R_{j} - \varepsilon^R_i||^2\hspace{0.1 in}| \hspace{0.1 in} x^R_i = \mu^R_{j} + \varepsilon^R_i \right)\pi_j\\ \nonumber
  &=  \sum^K_{j=1}E\left(||\hat{\mu}^R \left(\hat{g}(x^L_i)\right) - \mu^R_{j}||^2\hspace{0.1 in}| \hspace{0.1 in} x^R_i = \mu^R_{j} + \varepsilon^R_i \right)\pi_j + E||\varepsilon^R_i||^2\\ \nonumber
\end{align}
Note that unlike in $k = K$ case, here $\hat{\mu}^R \left(\hat{g}(x^L_i)\right) \neq \hat{\mu}^R_j$ could be true for $x^R_i = \mu^R_j+\varepsilon^R_i$.\\

Since $E||\varepsilon^R_i||^2 = tr\{\Sigma^R\}$, our next goal is to show as $n \to \infty$
\begin{equation}
\sum^K_{j=1}E\left(||\hat{\mu}^R \left(\hat{g}(x^L_i)\right) - \mu^R_{j}||^2\hspace{0.1 in}| \hspace{0.1 in} x^R_i = \mu^R_{j} + \varepsilon^R_i \right)\pi_j > 0
\end{equation}
Note that $\hat{\mu}^R \left(\hat{g}(x^L_i)\right) \in A_m(k)$. The necessary but not sufficient condition to have  
\[\sum^K_{j=1}E\left(||\hat{\mu}^R \left(\hat{g}(x^L_i)\right) - \mu^R_{j}||^2\hspace{0.1 in}| \hspace{0.1 in} x^R_i = \mu^R_{j} + \varepsilon^R_i \right)\pi_j = 0\]
is $\{\mu^R_1,\mu^R_2,...,\mu^R_K\} \in A_m(k)$. The strategy here is to show that at least one $\mu^R_j \notin A_m(k)$ when $k\neq K$ and $m \to \infty$, where $\mu^R_j \in \{\mu^R_1,\mu^R_2,...,\mu^R_K\}$ or equivalently, at least one $\mu^R_j \notin \bar{A}(k)$ when applying K-means on population $\mathcal{P}$ with parameter $k$. Following proof is based on the later one. \\

In case of $k < K$, the proof is trivial. This is because there are only $k$ distinct cluster centers returned from K-means, while there are $K$ distinct $\mu^R_j$. $k < K$ means at least one $\mu^R_j \notin \bar{A}(k)$. \\\\
In case of $k > K$, at least one cluster will be partitioned into at least two parts. This is because there are $K$ clusters in population $\mathcal{P}$ and if none of them be partitioned, K-means will give $K$ cluster centers instead of $k$ -- contradiction.\\

WLOG, we assume the partitioned cluster is $\mu^R_1$. In training set, there are $n^{train}_1$ observations in cluster $\mu^R_1$. Say the partition results one part with $n_1$ observations and its sample mean is $\hat{\mu}_1$, also we have \\

\textbf{Assumption I.} The partition results $n_1$ is a fixed proportion from $n^{train}_1$ when apply K-means on population $\mathcal{P}$ with parameter $k$, i.e $\lim_{m \to \infty} \frac{n_1}{n^{train}_1} = a_1$, where $0< a_1< 1$ is a positive constant. Also, the sample mean of the $n_1$ observation is different from the sample mean of the rest $n^{train}_1-n_1$ points.\\

With Assumption I., theorem $4$ below shows that $\mu^R_1 \notin \bar{A}(k)$. Therefore $(13)$ holds, which means 
\[\Theta = E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 > tr\{\Sigma^R\} \]
The proof is complete.
\end{proof}
\begin{theorem} 
Given $||\varepsilon^R_i||\leq1$, if for any $\mu^R_j$ in  $\{\mu^R_1,...,\mu^R_K\}$, WLOG, let $\mu^R_j = \mu^R_1$, $ ||\mu^R_1 - \mu|| > 2$, where $\mu$ is any weighted mean of $(\mu^R_2,...,\mu^R_{K})$ by weight vector $(n_2,...,n_{K})$, where $n_j,j=2,...,K$ is natural number and $\sum^{K}_{i=2}n_i >0$, i.e. $ \mu = \frac{n_2\mu_2+...+n_{K}\mu_{K}}{n_2+n_2...+n_{K}}$. Then $\mu^R_1 \notin \bar{A}(k)$ if $k \neq K$.
\end{theorem}
\begin{proof}
Let $\hat{\mu'}_1$ denotes the specific cluster center we are looking which is returned from K-means with parameter $k$ on population $\mathcal{P}$. By the condition given in theorem $4$, if such cluster consists of observations which are not from $\mu^R_1$, then such cluster center $\hat{\mu'}_1$ is at least $1$ unit away from $\mu^R_1$ since
\[\hat{\mu'}_1 = \frac{n_2\mu^R_2+...+n_{K}\mu^R_{K}}{n_2+n_2...+n_{K}}+\frac{\sum \varepsilon^R_i}{n_2+n_2...+n_{K}}\]
and 
\begin{align}
	||\hat{\mu'}_1-\mu^R_1|| &= ||\frac{n_2\mu^R_2+...+n_{K}\mu^R_{K}}{n_2+n_2...+n_{K}}+\frac{\sum \varepsilon^R_i}{n_2+n_2...+n_{K}}-\mu^R_1||  \nonumber \\ \nonumber
	&\geq ||\frac{n_2\mu^R_2+...+n_{K}\mu^R_{K}}{n_2+n_2...+n_{K}}-\mu^R_1||-||\frac{\sum \varepsilon^R_i}{n_2+n_2...+n_{K}}||   \\ \nonumber
	&> 2-||\frac{\sum \varepsilon^R_i}{n_2+n_2...+n_{K}}||   \\ \nonumber
	&> 1 
\end{align}
since $||\varepsilon^R_i||\leq1$. Hence, the only hope for $\hat{\mu'}_1 = \mu^R_1$ (therefore $\mu^R_1 \in \bar{A}(k)$) comes from cluster contains observations from $\mu^R_1$. WLOG, we assume those observations from $\mu^R_1$ are the $n_1$ points. 
\begin{itemize}
	\item When the $n_1$ observations form a single cluster\\
	In this case, $\hat{\mu'}_1 = \hat{\mu}_1$. We want to show 
	\begin{equation}
		 \lim_{m\to\infty}||\hat{\mu}_1-\mu^R_1|| > 0 
	\end{equation}
	It indeed is the case since when $m \to \infty$, by law of large number the sample mean converges to the population mean, so with probability $1$ 
	\begin{align}
		||\bar{\mu}_1-\mu^R_1|| &= ||\frac{n_1\hat{\mu}_1+(n^{train}_1-n_1)\hat{\mu}_{1'}}{n^{train}_1} -\mu^R_1||   \nonumber \\\nonumber		
		&= ||\frac{n_1}{n^{train}_1}\hat{\mu}_1 + \frac{n^{train}_1-n_1}{n^{train}_1}\hat{\mu}_{1'} - \mu^R_1||  \\\nonumber
		&= ||a_1\hat{\mu}_1 + (1-a_1)\hat{\mu}_{1'} - \mu^R_1 ||\\\nonumber
		&= 0\\\nonumber
    \end{align}	 
Since $\hat{\mu}_1 \neq \hat{\mu}_{1'}$, and $0<a_1<1$ by Assumption I, \[\lim_{m\to\infty}||\hat{\mu}_1-\mu^R_1|| > 0\]Where $\hat{\mu}_{1'}$ is the sample mean of the $n^{train}_1-n_1$ points of cluster $\mu^R_1$.\\
	\item When the $n_1$ observations be grouped together with $n_2$ observations from cluster $\mu^R_2$. \\\\
	Notice that it's a special case $\left(\left(n_2,...,n_{K}\right) = (n_2,0,...,0)\right)$ of the $n_1$ observations be grouped together with $\sum^{K}_{i=2}n_i$ observations from cluster $(\mu^R_2,...,\mu^R_{K})$, i.e. $\mu = \mu^R_2$. The result apply to the general cases.\\\\
	The cluster consists of $n_1$ observations from center $\mu^R_1$ with sample mean $\hat{\mu}_1$, and $n_2$ observations from center $\mu^R_2$ with sample mean $\hat{\mu}_2$. The cluster center (estimator of $\mu_1$) is then 
	\[\hat{\mu'}_1 = \frac{n_1\hat{\mu}_1+n_2\hat{\mu}_2}{n_1+n_2} \]
From above, we have $||\hat{\mu}_2-\mu^R_1|| > 1$ and $||\hat{\mu}_1-\mu^R_1|| > 0$. Since the $n_1$ observations are cluster together with $n_2$ observations from center $\mu^R_2$, and $\hat{\mu}_2$ is outside of cluster $\mu_1$ as $||\hat{\mu}_2-\mu^R_1|| > 1$ and $||\varepsilon^R_i||<1$, it's reasonable to assume in terms of $\mu^R_1$, $\hat{\mu}_1$ is biased towards $\hat{\mu}_2$. In other words, the inner product of two vector is nonnegative, i.e.
\[ <\hat{\mu}_1-\mu^R_1,\hat{\mu}_2-\mu^R_1>\hspace{0.2in} \geq 0 \] 
Therefore, when $m\to \infty$
\begin{align}
	||\hat{\mu'}_1 - \mu^R_1||^2&= || \frac{n_1\hat{\mu}_1+n_2\hat{\mu}_2}{n_1+n_2} - \mu^R_1||^2  \nonumber \\ \nonumber
	&=||\frac{n_1}{n_1+n_2}(\hat{\mu}_1-\mu^R_1)+\frac{n_2}{n_1+n_2}(\hat{\mu}_2-\mu^R_1)||^2  \\ \nonumber
	&= ||\frac{n_1}{n_1+n_2}(\hat{\mu}_1-\mu^R_1)||^2+ ||\frac{n_2}{n_1+n_2}(\hat{\mu}_2-\mu^R_1)||^2+\frac{2n_1n_2}{(n_1+n_2)^2} <\hat{\mu}_1-\mu^R_1,\hat{\mu}_2-\mu^R_1>\\ \nonumber
	&>0  \\ \nonumber
\end{align}
which means \hspace{1in} $\lim_{m\to\infty}||\hat{\mu'}_1-\mu^R_1|| > 0$
\end{itemize}
Since none of above situation returns cluster center $\hat{\mu'}_1 = \mu^R_1$, we claim $\mu^R_1 \notin \bar{A}(k)$
\end{proof}
\section{Simulation}
In this section, simulation is performed to evaluate the performance of our proposed methods in locating the ``correct" number of clusters. We compare with a basket of existing methods including Gap statistics \citep{tibshirani2001estimating}, Gaussian mixture model-based clustering \citep{fraley2002model}, CH-index \citep{calinski1974dendrite}, Hartigan statistics \citep{hartigan1975clustering}, Jump method \citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster}, Bootstrap stability \citep{fang2012selection} in following simulation settings. We select $L=R$ and $5-$fold cross-validation in row ($m=\frac{4}{5}n$) as default parameter setting for our proposed Gabriel method. Note that set $L=R$ corresponding to $2-$fold cross-validation in column.
	\begin{enumerate}
		\item[1.] Single cluster in $10$ dimensions --- $200$ observations, each observation uniformly distributed over $[0,1]$ in each dimension. 
		\item[2.] Two clusters in $4$ dimensions --- $50$ i.i.d standard normal observations are generated, i.e. generated from multivariate normal $\mathcal{N}\left(\boldsymbol\mu,\mathbf{I}\right)$ for each cluster, with cluster center $\boldsymbol\mu = (1,0,0,1)$ and $(1,3.5,3.5,1)$ respectively  
		\item[3.] Four clusters in $100$ dimensions ---  Each cluster has $100$ or $150$ i.i.d standard normal observations, with cluster centers randomly generated from multivariate normal distribution $\mathcal{N}\left(\mathbf{0},0.65^2\mathbf{I}\right)$
		\item[4.] Ten clusters in $100$ dimensions ---  Each cluster has $50$ or $100$ i.i.d standard normal observations, with cluster centers randomly generated from multivariate normal distribution $\mathcal{N}\left(\mathbf{0},0.72^2\mathbf{I}\right)$
		\item[5.] Four log-normal clusters in $16$ dimensions --- For each cluster, $30$ or $60$ i.i.d centered log-normal observations are generated from $ln\mathcal{N}\left(0,0.5^2 \right)$. The cluster centers are randomly generated from multivariate normal distribution $\mathcal{N}\left(\mathbf{0},1.2^2\mathbf{I}\right)$
		\item[6.] Three exponential clusters in $20$ dimensions with different variance --- $40$ observations in each cluster are generated from centered exponential distribution ($exp(\lambda)-1/\lambda$ in each dimension), with $\boldsymbol\mu$ randomly generated from  $\mathcal{N}\left(\mathbf{0},19\mathbf{I}\right)$ and $\lambda = 1, 1/2, 1/5 $ respectively.
	\end{enumerate}
Note that in setting $2-6$, all clusters are well-separated, i.e. no overlapping. In fact, any simulated clusters with minimum distance less than $1$ unit was discarded, so there is clear definition of true number of clusters. The parameters in setting $2-6$ are chosen such that about half of the random realization were discarded. The idea is borrowed from \cite{tibshirani2001estimating}.\\\\
Besides the number of clusters selected by each algorithm is returned, we also give the prediction error (PE) for each algorithm in each setting. The PE is defined by 
\[ PE(k) = \frac{||\hat{\mathbf{X}}_k- \boldsymbol{\mu} ||_F}{PE_{oracle}} \]
where \[PE_{oracle} = \min_k ||\hat{\mathbf{X}}_k- \boldsymbol{\mu} ||_F\]
Matrix $\boldsymbol{\mu}$ consists of true centers of observations in data matrix $\mathbf{X}$, while $\hat{\mathbf{X}}_k$ is obtained by replacing each observation in $\mathbf{X}$ by its closest cluster center result from applying K-means on $\mathbf{X}$ with parameter $k$. Here $k$ is the optimal number of clusters selected by each algorithm on data $\mathbf{X}$. This is only possible on simulated data because otherwise one would never know the true center matrix $\boldsymbol{\mu}$.\\\\
The oracle selected number for each data $\mathbf{X}$ is also given on the table, which is defined by 
\[ K_{oracle} =  \underset{k}{\operatorname{argmin}}\hspace{0.05in} ||\hat{\mathbf{X}}_k- \boldsymbol{\mu} ||_F \]
\begin{table}[H]
\centering
\caption{ \label{table1}Simulation results} 
\scalebox{0.97}{
\begin{tabular}{ lccccccccccccc}
\hline
 Number of clusters &&& $1$&$2$&$3$&$4$&$5$&$6$&$7$&$8$&$9$&$10$&PE \\ \hline
\textit{Setting $1$} &&& &&&&&&&&& \\
Oracle&&& $\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1$\\ 
Gap&&& $\mathbf{92}$&$0$&$1$&$1$&$1$&$2$&$1$&$0$&$1$&$1$&$8.0\pm28.3$\\ 
Gaussian-Mix&&& $\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
CH&&& $-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$\\ 
Hartigan&&& $-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$\\ 
Jump&&& $\mathbf{0}$&$0$&$0$&$0$&$0$&$0$&$1$&$8$&$34$&$57$&$102.3\pm 58.7$\\ 
Prediction strength&&&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
Stability&&& $-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$\\ 
Gabriel CV&&& $\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
Wold CV&&& $\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
\hline
\textit{Setting $2$} &&& &&&&&&&&& \\
Oracle&&& $0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1$\\ 
Gap&&& $0$&$\mathbf{98}$&$1$&$1$&$0$&$0$&$0$&$0$&$0$&$0$&$1.2\pm 1.3$\\ 
Gaussian-Mix&&& $0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
CH&&& $0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\
Hartigan&&& $0$&$\mathbf{0}$&$6$&$16$&$21$&$14$&$12$&$10$&$8$&$13$&$28.1\pm 50.8$\\
Jump&&& $0$&$\mathbf{70}$&$0$&$0$&$0$&$0$&$0$&$2$&$7$&$21$&$15.5\pm 55.4$\\ 
Prediction strength&&&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
Stability&&& $0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\
Gabriel CV&&& $0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
Wold CV&&& $0$&$\mathbf{66}$&$3$&$14$&$11$&$4$&$1$&$0$&$1$&$0$&$6.0\pm 10.6$\\ 
\hline
\textit{Setting $3$} &&& &&&&&&&&& \\
Oracle&&& $0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$1$\\ 
Gap&&& $0$&$0$&$0$&$\mathbf{69}$&$28$&$3$&$0$&$0$&$0$&$0$&$1.3\pm 0.5$\\ 
Gaussian-Mix&&& $85$&$2$&$3$&$\mathbf{3}$&$1$&$0$&$4$&$2$&$0$&$0$&$35.9\pm 13.7$\\ 
CH&&& $0$&$36$&$28$&$\mathbf{36}$&$0$&$0$&$0$&$0$&$0$&$0$&$12.1\pm 9.9$\\ 
Hartigan&&& $0$&$0$&$13$&$\mathbf{75}$&$11$&$0$&$1$&$0$&$0$&$0$&$2.5\pm 3.5$\\
Jump&&& $0$&$0$&$0$&$\mathbf{0}$&$0$&$0$&$0$&$0$&$14$&$86$&$5.6\pm 0.4$\\ 
Prediction strength&&&$56$&$0$&$0$&$\mathbf{44}$&$0$&$0$&$0$&$0$&$0$&$0$&$23.5\pm 20.6$\\ 
Stability&&& $0$&$0$&$0$&$\mathbf{33}$&$59$&$8$&$0$&$0$&$0$&$0$&$1.7\pm 0.5$\\
Gabriel CV&&& $0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
Wold CV&&& $0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
\hline
 Number of clusters &&& $\leq 6$&$7$&$8$&$9$&$10$&$11$&$12$&$13$&$14$&$15$ \\ \hline
\textit{Setting $4$} &&& &&&&&&&&& \\
Oracle&&& $0$&$0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$1$\\ 
Gap&&& $0$&$0$&$0$&$0$&$\mathbf{12}$&$39$&$23$&$16$&$7$&$3$&$1.5\pm 0.4$\\ 
Gaussian-Mix&&& $71$&$6$&$4$&$2$&$\mathbf{4}$&$4$&$0$&$2$&$3$&$4$&$22.2\pm 13.5$\\ 
CH&&& $84$&$6$&$6$&$4$&$\mathbf{0}$&$0$&$0$&$0$&$0$&$0$&$20.8\pm 8.9$\\ 
Hartigan&&& $13$&$4$&$13$&$21$&$\mathbf{17}$&$12$&$5$&$9$&$1$&$5$&$4.4\pm 4.8$\\
Jump&&& $0$&$0$&$0$&$20$&$\mathbf{80}$&$0$&$0$&$0$&$0$&$0$&$1.4\pm 0.9$\\ 
Prediction strength&&&$100$&$0$&$0$&$0$&$\mathbf{0}$&$0$&$0$&$0$&$0$&$0$&$35.2\pm 4.4$\\ 
Stability&&& $0$&$0$&$0$&$0$&$\mathbf{0}$&$3$&$24$&$29$&$28$&$16$&$2.0\pm 0.3$\\
Gabriel CV&&&  $0$&$0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\  
Wold CV&&&  $0$&$0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
\hline
\end{tabular}
}\\
\hspace{-2.0in} \footnotesize {In setting $1$, ``-'' means the method can not be used with parameter $k=1$}.
\end{table}
\begin{table}[H]
\centering
\begin{tabular}{ lccccccccccccc}
\hline
 Number of clusters &&& $1$&$2$&$3$&$4$&$5$&$6$&$7$&$8$&$9$&$10$&PE \\ \hline
\textit{Setting $5$} &&& &&&&&&&&& \\
Oracle&&& $0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$1$\\ 
Gap&&& $0$&$0$&$0$&$\mathbf{66}$&$25$&$7$&$1$&$1$&$0$&$0$&$1.8\pm 1.3$\\ 
Gaussian-Mix&&& $0$&$0$&$0$&$\mathbf{56}$&$36$&$6$&$2$&$0$&$0$&$0$&$2.0\pm 1.3$\\ 
CH&&& $0$&$5$&$25$&$\mathbf{53}$&$12$&$2$&$1$&$0$&$0$&$0$&$10.3\pm 17.0$\\ 
Hartigan&&& $0$&$0$&$24$&$\mathbf{61}$&$5$&$6$&$3$&$1$&$0$&$0$&$7.3\pm 11.5$\\
Jump&&& $0$&$0$&$0$&$\mathbf{73}$&$0$&$0$&$1$&$0$&$5$&$21$&$3.1\pm 3.5$\\ 
Prediction strength&&&$74$&$8$&$5$&$\mathbf{13}$&$0$&$0$&$0$&$0$&$0$&$0$&$114.5\pm 67.8$\\ 
Stability&&& $0$&$4$&$5$&$\mathbf{23}$&$36$&$23$&$7$&$2$&$0$&$0$&$5.7\pm 10.2$\\
Gabriel CV&&& $0$&$0$&$0$&$\mathbf{91}$&$8$&$1$&$0$&$0$&$0$&$0$&$1.1\pm 0.5$\\ 
Wold CV&&& $0$&$0$&$0$&$\mathbf{99}$&$1$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0.1$\\ 
\hline
\textit{Setting $6$} &&& &&&&&&&&& \\
Oracle&&& $0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1$\\ 
Gap&&& $0$&$0$&$\mathbf{10}$&$8$&$13$&$11$&$12$&$11$&$17$&$18$&$9.8\pm 5.9$\\ 
Gaussian-Mix&&& $0$&$0$&$\mathbf{88}$&$12$&$0$&$0$&$0$&$0$&$0$&$0$&$1.4\pm 1.3$\\ 
CH&&& $0$&$17$&$\mathbf{74}$&$8$&$1$&$0$&$0$&$0$&$0$&$0$&$4.0\pm 6.3$\\ 
Hartigan&&& $0$&$0$&$\mathbf{87}$&$10$&$3$&$0$&$0$&$0$&$0$&$0$&$1.6\pm 1.6$\\
Jump&&& $0$&$0$&$\mathbf{0}$&$0$&$0$&$1$&$2$&$13$&$32$&$52$&$13.7\pm 5.0$\\ 
Prediction strength&&&$19$&$2$&$\mathbf{78}$&$1$&$0$&$0$&$0$&$0$&$0$&$0$&$10.3\pm 20.5$\\ 
Stability&&& $0$&$0$&$\mathbf{24}$&$39$&$29$&$7$&$1$&$0$&$0$&$0$&$4.5\pm 3.0$\\
Gabriel CV&&& $0$&$0$&$\mathbf{97}$&$3$&$0$&$0$&$0$&$0$&$0$&$0$&$1.1\pm 0.7$\\ 
Wold CV&&& $0$&$0$&$\mathbf{89}$&$9$&$1$&$0$&$1$&$0$&$0$&$0$&$1.6\pm 1.9$\\ 
\hline
\end{tabular}\\
\hspace{-1.8in} \footnotesize {Last column gives the mean and standard deviation of PE for each algorithm}.
\end{table}
\noindent
All methods are used with their default parameter setting except for Gap statistics. We selected the $k$  corresponds to the global maximum of gap statistics, rather than the default method which selects the smallest $k$ such that the gap statistics is not more than 1 standard error away from the first local maximum. This is because among all the possible options, the global maximum criteria gives the best result in our simulation settings.\\

The ``--'' sign in setting $1$ is because for CH-index, Hartigen and Bootstrap stability, the minimum $k$ they can pick is $2$. So they are not included for comparison since the true $k=1$. Setting $2-4$ have Gaussian distribution, where the Gaussian mixture model-based clustering suppose to perform the best. It is only the case for setting $2$. In setting $3$ and $4$, it has a very poor performance even if the underlying distribution is indeed Gaussian mixture model. It highlights the difficulties that high dimension data could cause. Other methods (Gap statistics, CH-index, Hartigan, Jump method, Prediction strength and Boostrap stability) also have problems in these two high-dimension settings as we could not find any one with reasonable performance in terms of finding the correct $k$ in both cases. In terms of PE, Boostrap stability and Gap statistics have reasonable performance, but come with the price of picking larger $k$. Our proposed Gabriel method (also Wold CV method) clearly stands out from the basket of existing methods using either measure under the high-dimension setting $3$ and $4$.\\

Setting $5$ and $6$ corresponding to the situation where data are heavy-tailed instead of normal distributed, coupled with unequal number of observations or unequal variance for different clusters. From Table \ref{table1}, we can see that our proposed Gabriel method is robust and has superior performance compare with other methods. Because one can hardly tell the underlying distribution of data in practice, the resilience to non-Gaussian data gives the Gabriel method a clear edge. 
\section{Real data application}
We also applied our proposed method to three real world data sets obtained from the University of California Irvine machine learning repository. The first and third data sets are selected because there are clear number of clusters in those two data sets. The second data set is used as a benchmark data set since it was widely used in literature.\\

The first one is congress voting data which consists of voting records of $98$th United States Congress, $2$nd session. This data set includes votes for each of the U.S. House of Representatives Congressmen on the $16$ key votes identified by the \textit{CQA} (Congressional Quarterly Almanac). For each vote, each Congressman either vote positively ``yea" (voted for/paired for/announced for),  negatively ``nay'' (voted against/paired against/announced against) or position unknown ``?". We took out those records contain ``?" before comparing each algorithm. It results in $232$ remaining records, with $124$ democrat and $108$ republican. \\

The second data set is the well-known Wisconsin breast cancer data set \citep{mangasarian1990pattern}. After excluding the records with missing data, this data set consists records of $683$ patients, each with measurements of nine attributes of their biopsy specimens. It is known that there exist at least two groups of patients: $444$ patients with benign specimens and $239$ patients with malignant specimens.\\

The third data set is the Sonar data returned from two targets -- a metal cylinder and a rock with similar shape, which is first studied by \cite{gorman1988analysis} using a neural network. Both targets were impinged by pulse which was a wide-band linear FM chirp $(ka = 55.6)$. Returns were collected at a range of $10$ meters and obtained from the cylinder at aspect angles spanning $90^{\circ}$ and from the rock at aspect angles spanning $180^{\circ}$. The data set contains $208$ returns ($111$ cylinder returns and $97$ rock returns), with each composed of $60$ spectral samples, normalized to take on values between $0$ and $1$. So the data has $60$ features with clearly $2$ clusters.

\begin{table}[H]
\begin{center}
\caption{\label{table2} Number of clusters selected by each algorithm}
\begin{tabular}{lccccc}
    \hline                  
 & Congress Voting && Breast Cancer && Sonar \\ \hline                    
CH-index & $2$ && $2$  && $2$    \\
Hartigan & $3$ && $3$  &&  $3$   \\
Jump & $9$ && $9$ && $10$  \\   
Prediction strength & $2$ &&  $2$ && $1$    \\
Bootstrap stability & $2$ &&  $2$ &&  $10$   \\ 
Gap & $10$ && $9$ && $10$  \\   
Gaussian-Mix & $7$ && $5$ && $1$  \\   
Gabriel & $2$  && $3$  && $2$  \\    
Wold & $2$ && $3$ && $10$ \\  \hline 
\end{tabular}
\end{center}
\hspace{0.5in} \footnotesize {All the algorithms executed with their default parameter settings with $k$ ranges from $1$ to $10$}
\end{table} 
\noindent
Since most congressmen vote base on their parties' interest, $2$ parties (Democratic and Republican) represent two clusters in this data set. So the optimal number should be two. Close inspection shows \textit{kmeans} with $k=2$ separates the two parties very well with the lowest miss-classification error $10.43\%$ ($k=3$ has $14.78\%$). Note that CH-index and Bootstrap stability also return $k=2$, but $2$ is the lower bound those methods can select for $k$. So it's not clear they actually choose $k=2$ or they hit the lower bound (they would pick $k=1$ if allowed). For the breast cancer data, it's know to have at least $2$ cluster based on whether it's benign specimens or not. But it doesn't mean the optimal $k$ should be $2$. \cite{fujita2014non} noticed that the malign group is quite heterogeneous and can be further clustered into at least two subgroups. Hence, the result $k=3$ given by our proposed Gabriel method (as well as Wold method) make sense. For the Sonar data set which is relatively high-dimension, only our proposed Gabriel method and CH-index selects $k=2$. Majority of the methods collapse to either select the maximum or select the minimum number, underline how difficult it was to pick the right $k$ when dimension increases. 
\section{Conclusion}
In this paper, we proposed a novel approach to estimate the number of clusters. The intuition behind our proposed methods is to transfer the unsupervised learning problem into supervised learning problem via novel form of cross validation. We proved the self-consistency for our proposed Gabriel CV method as well as its asymptotic property with noise, and showed the robustness of our method by simulation. Our method has very good performance in our limited simulation settings, and clearly the superior one when data is high dimension or is heavy-tailed.
\include{...}

\section*{\textbf{APPENDIX}}
\appendix
\section{Wold CV estimation} \label{app:foobar}
\begin{itemize}
	\item For each $k = 1,2,...,k_{max}$
	\begin{enumerate}
		\item Randomly draw some entries in $\mathbf{X}$ missing, keep those hold-out values in vector $V_{true}$
		\item Impute the missing values with column mean or $0$, denote the imputed data as $\mathbf{X}_{new}$
		\item Apply the iterative procedure below until converge or stopping criteria reached
		\begin{itemize}
			\item Apply $K$-mean on data set $\mathbf{X}_{new}$ with parameter $k$
			\item Substitute each observation in $\mathbf{X}_{new}$ by its nearest center, get new data $\mathbf{X}^c_{new}$ ($\mathbf{X}_{new}$ keep the same)
			\item Replace (impute) those imputed values in $\mathbf{X}_{new}$ with the corresponding entries in $\mathbf{X}^c_{new}$
			\item Calculate the difference between the old and newly imputed values, check whether or not they coincide (converge) 
		\end{itemize}
		\item Obtain the last imputed entry values of converged $\mathbf{X}_{new}$, denote it by $V_{converge}$
		\item Calculate the prediction error  $Error_k = ||V_{true} - V_{converge}||^2$ 
	\end{enumerate}
	\item For each CV folder, repeat above procedure and obtain the $Error_k$ for each $k$
	\item Average $Error_k$ across all folders for each $k$, and then select the $k$ corresponding to the minimum average $Error_k$ 
\end{itemize}

\bibliography{myreferences}
\bibliographystyle{apalike}
\end{document}

### 2015-06-06 code/elbow plot.R
Add Wei's code

@@ -1,0 +1,29 @@
+
+library("clusterCrit")
+library(MASS)
+
+pdf("Data_elbow.pdf")
+D1 <- mvrnorm(100,c(5,5),3*diag(2))
+plot(D1[,1],D1[,2],ylim=c(0,15),xlim=c(0,15),col="red",xlab="X1",ylab="X2")
+
+D2 <- mvrnorm(20,c(5,12),diag(2))
+points(D2[,1],D2[,2],ylim=c(0,15),xlim=c(0,15),col="blue")
+
+D3 <- mvrnorm(10,c(9,9),0.5*diag(2))
+points(D3[,1],D3[,2],ylim=c(0,15),xlim=c(0,15),col="green")
+
+D4 <- mvrnorm(100,c(12,12),diag(2))
+points(D4[,1],D4[,2],ylim=c(0,15),xlim=c(0,15))
+dev.off()
+
+
+Data <- rbind(D1,D2,D3,D4)
+
+Disp <- rep(NA,10)
+for(w in 1:10){
+ 	cl <- kmeans(Data,w,nstart = 100)
+      Disp[w] <- intCriteria(Data,cl$cluster,"trace_w")[[1]]
+}
+pdf("elbow.pdf")
+plot(1:10,Disp,type="b",ylab = "Within-cluster dispersion Wk",xlab="Number of clusters k")
+dev.off()

### 2015-06-06 code/elbow plot_new.R

@@ -1,0 +1,61 @@
+
+library("clusterCrit")
+library(MASS)
+
+##---------------------Top plot of figure 1-----------------------------------
+set.seed(3)
+
+pdf("Data_example.pdf")
+D1 <- mvrnorm(20,c(1.7,3),0.15*diag(2))
+plot(D1[,1],D1[,2],ylim=c(0,4.5),xlim=c(0,4.5),col="red",xlab="X1",ylab="X2")
+
+D2 <- mvrnorm(20,c(1,1),0.15*diag(2))
+points(D2[,1],D2[,2],ylim=c(0,5),xlim=c(0,5),col="blue")
+
+D3 <- mvrnorm(20,c(3,1.5),0.15*diag(2))
+points(D3[,1],D3[,2],ylim=c(0,5),xlim=c(0,5),col="green")
+
+D4 <- mvrnorm(20,c(3.7,3.7),0.1*diag(2))
+points(D4[,1],D4[,2],ylim=c(0,5),xlim=c(0,5))
+dev.off()
+
+
+Data <- rbind(D1,D2,D3,D4)
+
+Disp <- rep(NA,10)
+for(w in 1:10){
+ 	cl <- kmeans(Data,w,nstart = 100)
+      Disp[w] <- intCriteria(Data,cl$cluster,"trace_w")[[1]]
+}
+pdf("W_k.pdf")
+plot(1:10,Disp,type="b",ylab = "Within-cluster dispersion Wk",xlab="Number of clusters k")
+dev.off()
+
+##---------------------Bottom plot of figure 1-----------------------------------
+set.seed(1)
+
+pdf("Data_elbow.pdf")
+D1 <- mvrnorm(100,c(5,5),3*diag(2))
+plot(D1[,1],D1[,2],ylim=c(0,15),xlim=c(0,15),col="red",xlab="X1",ylab="X2")
+
+D2 <- mvrnorm(20,c(5,12),diag(2))
+points(D2[,1],D2[,2],ylim=c(0,15),xlim=c(0,15),col="blue")
+
+D3 <- mvrnorm(10,c(9,9),0.5*diag(2))
+points(D3[,1],D3[,2],ylim=c(0,15),xlim=c(0,15),col="green")
+
+D4 <- mvrnorm(100,c(12,12),diag(2))
+points(D4[,1],D4[,2],ylim=c(0,15),xlim=c(0,15))
+dev.off()
+
+
+Data <- rbind(D1,D2,D3,D4)
[…]

### 2015-06-06 cvclust.tex
Move figures to 'figure' directory
Rename 'myreferences.bib' to 'references.bib'
Add counterexample

--- 2015-05-18 Draft2.tex
@@
 
+\newcommand{\E}{\operatorname{E}}
+
 \begin{document}
@@
 \begin{minipage}{0.45\linewidth}
-\includegraphics[width=\linewidth]{Data_example.pdf}
+\includegraphics[width=\linewidth]{figure/Data_example.pdf}
 \end{minipage}
@@
 \begin{minipage}{0.45\linewidth}
-\includegraphics[width=\linewidth]{W_k.pdf}
+\includegraphics[width=\linewidth]{figure/W_k.pdf}
 \end{minipage}
@@
 \begin{minipage}{0.45\linewidth}
-\includegraphics[width=\linewidth]{Data_elbow.pdf}
+\includegraphics[width=\linewidth]{figure/Data_elbow.pdf}
 \end{minipage}
@@
 \begin{minipage}{0.45\linewidth}
-\includegraphics[width=\linewidth]{elbow.pdf}
+\includegraphics[width=\linewidth]{figure/elbow.pdf}
 \end{minipage}
@@
 
-\bibliography{myreferences}
+\section{Counterexample with correlated variables}
+
+Suppose that $(X, Y)$ are jointly mean-zero multivariate normal with unit
+marginal variances and correlation $\rho$.  In this case, the data are drawn
+from a single cluster; the true number of clusters is~$1$.  We will show that
+Gabriel cross-validation is inconsistent if $\rho > 0.5$, in the sense that
+$k = 2$ will be preferred to $k = 1$ almost surely for large enough $n$.
+
+Given a sample $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to
+$\{ Y_i \}_{i=1}^{n}$.  If $k = 1$ and $n$ is large enough, then the centroid
+will be close to $\E(Y) = 0$, in the sense that the centroid converges almost
+surely to this value as $n$ tends to infinity \citep{pollard1981strong}.
+Similarly, if $k = 2$ and $n$ is large enough, then the centroids will tend to
+$\E(Y \mid Y > 0) = \sqrt{2/\pi}$ and $\E(Y \mid Y < 0) = -\sqrt{2/\pi}$.  In
+$k = 2$ case, then if $n$ is large enough, the cluster assignment will be
+determined according to whether $Y > 0$.
+
+In the $k = 1$ case, if $n$ is large enough, the average Gabriel
+cross-validation error converges to $\E(Y^2) = 1$. In the $k = 2$ case, if $n$
+is large enough, then the linear decision rule for the $X$ variables will be
+determined according to whether $X > 0$.  In this case, the error converges to
+$\E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]$, where $a =
+\sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows that $Y
+\mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$.   Thus, in
+the $k = 2$ case, the Gabriel cross-validation error converges to $1 + a^2 (1
+- 2 \rho)$.  In particular, if $\rho > 0.5$, then the asymptotic
+cross-validation error for $k = 2$ will be smaller than for $k = 1$.
+
+We confirm this with a simulation.  [TODO: add simulation]
+
+\bibliography{references}
 \bibliographystyle{apalike}

### 2015-06-06 cvclust.tex
Add moment lemma

@@
 
+\section{Technical Lemmas}
+
+\begin{lemma}\label{lem:truncated-normal-moments}
+
+If $Z$ is a standard normal random variable, then
+\[
+  \E(Z \mid a < Z < b)
+    = - \frac{\varphi(b) - \varphi(a)}
+             {\Phi(b) - \Phi(a)}
+\]
+and
+\[
+  \E\{(Z - \delta)^2 \mid a < Z < b\}
+    = \delta^2 + 1
+    - \frac{  (b - 2 \delta) \varphi(b)
+            - (a - 2 \delta) \varphi(a)}
+           {\Phi(b) - \Phi(a)}
+\]
+for all constants $a$, $b$, and $\delta$, where
+$\phi(z)$ and $\Phi(z)$ are the standard normal probability density and
+cumulative distribution functions.  These expressions are valid
+for $a = -\infty$ or $b = \infty$ by taking limits.
+
+\end{lemma}
+\begin{proof}
+We will derive the expression for the second moment.
+Integrate by parts to get
+\begin{align*}
+  \E[ (Z - \delta)^2 1\{Z < b\}]
+    &= \int_{-\infty}^b (z - \delta)^2 \varphi(z) \, dz \\
+    &= (\delta^2 + 1) \Phi(b) - (b - 2 \delta) \varphi(b).
+\end{align*}
+Now,
+\[
+  \E\{(Z - \delta)^2 \mid a < Z < b\}
+    =
+    \frac{  \E[ (Z - \delta)^2 1\{Z < b\}]
+          - \E[ (Z - \delta)^2 1\{Z < a\}]}
+         { \Phi(b) - \Phi(a) }.
+\]
+\end{proof}
+
+Lemma~\ref{lem:truncated-normal-moments} has some important special cases:
+\begin{align*}
+  \E\{Z \mid Z > 0\} &= 2 \varphi(0) = \sqrt{2 / \pi}, \\
+  \E\{(Z - \delta)^2 \mid Z > 0 \}
+    &= \delta^2 + 1 - 4 \delta \varphi(0), \\
+  \E\{(Z - \delta)^2 \mid Z < 0 \}
+    &= \delta^2 + 1 + 4 \delta \varphi(0).
+\end{align*}
+
 \section{Counterexample with correlated variables}
@@
 Similarly, if $k = 2$ and $n$ is large enough, then the centroids will tend to
-$\E(Y \mid Y > 0) = \sqrt{2/\pi}$ and $\E(Y \mid Y < 0) = -\sqrt{2/\pi}$.  In
-$k = 2$ case, then if $n$ is large enough, the cluster assignment will be
-determined according to whether $Y > 0$.
+$\E(Y \mid Y > 0) = \sqrt{2/\pi}$ and $\E(Y \mid Y < 0) = -\sqrt{2/\pi}$.  We
+have used Lemma~\ref{lem:truncated-normal-moments} to compute the
+expectations.  In $k = 2$ case, then if $n$ is large enough, the cluster
+assignment will be determined according to whether $Y > 0$.
 
@@
 determined according to whether $X > 0$.  In this case, the error converges to
-$\E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]$, where $a =
-\sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows that $Y
+$\E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]$, where
+$a = \sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows that $Y
 \mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$.   Thus, in

### 2015-06-06 code/null-1d.R
Start null 1d sims

@@ -1,0 +1,40 @@
+# code/null-1d.R
+# experiments for the null case (1 cluster) in one dimesion
+
+
+cluster <- function(x, centers, seed = 0, nstart = 100, ...)
+{
+    if (!is.na(seed)) {
+        if ((exists0 <- exists(".Random.seed", globalenv()))) {
+            seed0 <- .Random.seed
+        }
+        set.seed(seed)
+    }
+
+    cl <- kmeans(x, centers, nstart=nstart, ...)
+
+    if (!is.na(seed)) {
+        if (exists0) {
+            .Random.seed <- seed0
+        } else {
+            rm(".Random.seed", globalenv())
+        }
+    }
+
+    cl
+}
+
+
+
+set.seed(2651513) # from random.org
+
+n <- 1000000
+
+mu <- 1.0
+mean <- c(-1, 1) * mu
+
+eps <- rnorm(n)
+g <- sample.int(2, n, TRUE)
+x <- mean[g] + eps
+
+

### 2015-06-06 cvclust.tex
Add additional step to proof
change 'correlated variables' to 'correlated noise'

@@
 for all constants $a$, $b$, and $\delta$, where
-$\phi(z)$ and $\Phi(z)$ are the standard normal probability density and
+$\varphi(z)$ and $\Phi(z)$ are the standard normal probability density and
 cumulative distribution functions.  These expressions are valid
@@
 
-\section{Counterexample with correlated variables}
+\section{Counterexample with correlated noise}
 
@@
 $a = \sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows that $Y
-\mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$.   Thus, in
-the $k = 2$ case, the Gabriel cross-validation error converges to $1 + a^2 (1
-- 2 \rho)$.  In particular, if $\rho > 0.5$, then the asymptotic
+\mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$, so that
+$\E[(Y - a)^2 \mid X] = (\rho X - a)^2 + (1 - \rho^2)$.  Applying
+Lemma~\ref{lem:truncated-normal-moments}, in the $k = 2$ case we get that
+the Gabriel cross-validation error converges to $1 + a^2 (1 - 2 \rho)$.
+In particular, if $\rho > 0.5$, then the asymptotic
 cross-validation error for $k = 2$ will be smaller than for $k = 1$.

### 2015-06-07 code/null-1d.R
Add null-1d sim

@@ -1,8 +1,15 @@
 # code/null-1d.R
 # experiments for the null case (1 cluster) in one dimesion
 
+library("grid")
+library("MASS") # lda
+library("RColorBrewer") # brewer.pal
 
-cluster <- function(x, centers, seed = 0, nstart = 100, ...)
+palette(brewer.pal(6, "Set1"))
+
+
+# default seed chosen by random.org
+cluster <- function(x, centers, seed = 2651513, nstart = 100, ...)
 {
     if (!is.na(seed)) {
         if ((exists0 <- exists(".Random.seed", globalenv()))) {
@@ -26,15 +33,137 @@ cluster <- function(x, centers, seed = 0, nstart = 100, ...)
 
 
 
-set.seed(2651513) # from random.org
-
-n <- 1000000
-
-mu <- 1.0
-mean <- c(-1, 1) * mu
-
-eps <- rnorm(n)
-g <- sample.int(2, n, TRUE)
-x <- mean[g] + eps
 
 
+set.seed(0)
+
+ntrain <- 1000
+ntest <- ntrain
+n <- ntest + ntrain
+train <- seq_len(ntrain)
+test <- ntrain + seq_len(ntest)
+
+predictor <- 1
+response <- 2
+
+z <- matrix(rnorm(n * 2), n, 2)
+
+cverr <- function(correlation, centers)
+{
+    r <- correlation
+    coef <- cbind(rbind(sqrt((1 + r) / 2),  sqrt((1 - r) / 2)),
+                  rbind(sqrt((1 + r) / 2), -sqrt((1 - r) / 2)))
+    x <- z %*% coef
+
+    if (centers == 1) {
+        fit <- colMeans(x[train,response,drop=FALSE])
+        err <- scale(x[test,response], center=fit, scale=FALSE)
+    } else {
+        cl <- cluster(x[train,response], centers=centers)
+        cluster <- factor(cl$cluster, levels=seq_len(centers))
+        fit <- lda(x[train,predictor,drop=FALSE], cluster)
+        pred <- predict(fit, x[test,predictor,drop=FALSE])$class
+        err <- x[test,response] - cl$centers[pred,]
+    }
+
+    mse <- mean(err^2)
+    mse
+}
+
+
+correlation <- seq(-1, 1, length.out=21)
+mse1 <- sapply(correlation, function(r) cverr(r, 1))
+mse2 <- sapply(correlation, function(r) cverr(r, 2))
[…]

### 2015-06-07 code/null-1d.R
Switch from grid to base graphics

@@ -1,7 +1,6 @@
 # code/null-1d.R
 # experiments for the null case (1 cluster) in one dimesion
 
-library("grid")
 library("MASS") # lda
 library("RColorBrewer") # brewer.pal
 
@@ -37,7 +36,7 @@ cluster <- function(x, centers, seed = 2651513, nstart = 100, ...)
 
 set.seed(0)
 
-ntrain <- 1000
+ntrain <- 100000
 ntest <- ntrain
 n <- ntest + ntrain
 train <- seq_len(ntrain)
@@ -84,86 +83,37 @@ data <- data.frame(correlation = rep(correlation, 6),
                    centers = factor(rep(c(1, 2, 3, 4, 8, 15),
                                     each=length(correlation))))
 
-h.in <- 4
-w.in <- 4
-vplay <- grid.layout(1, 2,
-                     widths=unit(w.in * c(1, 0.5), "in"),
-                     heights=unit(h.in, "in"))
 
 
+mar <- c(3.25,3.25,0.75, 4.50 + 0.75)
+mai <- mar * par("csi")
+w <- h <- 3
+
+pdf("fig.pdf", width=sum(w, mai[c(2,4)]), height=sum(h, mai[c(1,3)]))
+
+par(las=1, mai=mai, mgp=c(2.25,0.75,0), ps=11, tcl=-0.4)
+plot(data$correlation, data$mse, type="n",
+     xlab="Correlation", ylab="CV Error")
+axis(3, label=FALSE)
+axis(4, label=FALSE)
+
+for (i in seq_len(nlevels(data$centers))) {
+    k <- levels(data$centers)[i]
+    dk <- subset(data, centers == k)
+    points(dk$correlation, dk$mse, pch=i, col=i)
+}
 
 
-gp <- gpar(fontsize=11)
+centers <- levels(data$centers)
+ncenters <- nlevels(data$centers)
 
-gplot <- local({
-    g <- gTree(x = NULL, y = NULL,
-        gp = gp,
-        childrenvp = vpTree(
-            plotViewport(c(3.25, 3.25, 0.75, 0.75), name="plotRegion"),
-            vpList(dataViewport(data$correlation, data$mse,
-                                width=unit(3, "in"),
-                                height=unit(3, "in"),
-                                name="dataRegion"))),
-        children = gList(
-            xaxisGrob(vp="plotRegion::dataRegion"),
-            xaxisGrob(vp="plotRegion::dataRegion", main=FALSE, label=FALSE),
[…]

### 2015-06-07 cvclust.tex
fix for rho < 0

@@
 cross-validation error converges to $\E(Y^2) = 1$. In the $k = 2$ case, if $n$
-is large enough, then the linear decision rule for the $X$ variables will be
-determined according to whether $X > 0$.  In this case, the error converges to
-$\E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]$, where
-$a = \sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows that $Y
-\mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$, so that
-$\E[(Y - a)^2 \mid X] = (\rho X - a)^2 + (1 - \rho^2)$.  Applying
-Lemma~\ref{lem:truncated-normal-moments}, in the $k = 2$ case we get that
-the Gabriel cross-validation error converges to $1 + a^2 (1 - 2 \rho)$.
-In particular, if $\rho > 0.5$, then the asymptotic
-cross-validation error for $k = 2$ will be smaller than for $k = 1$.
+is large enough and $\rho > 0$, then the linear decision rule for the $X$
+variables will be determined according to whether $X > 0$; if $\rho < 0$ then
+the decision is according to whether $X < 0$  In the $\rho > 0$ case, the
+error converges to $\E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]$,
+where $a = \sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows
+that $Y \mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$, so
+that $\E[(Y - a)^2 \mid X] = (\rho X - a)^2 + (1 - \rho^2)$.  Applying
+Lemma~\ref{lem:truncated-normal-moments}, in the $k = 2$ case we get that the
+Gabriel cross-validation error converges to $1 + a^2 (1 - 2 \rho)$.  In the
+$\rho < 0$ case, it converges to $1 + a^2 (1 + 2 \rho)$.  In
+particular, if $|\rho| > 0.5$, then the asymptotic cross-validation error for $k
+= 2$ will be smaller than for $k = 1$.
 

### 2015-06-08 code/classify.R
Reorganize code, add multiple replicates

@@ -1,0 +1,24 @@
+# code/classify.R
+#
+# Depends:
+#   library("MASS")
+
+
+classify_lda <- function(x, grouping, prior=c("proportions", "equal"))
+{
+    g <- as.factor(grouping)
+    prior <- match.arg(prior)
+
+    n <- length(grouping)
+    ng <- nlevels(g)
+
+    if (prior == "equal") {
+        prior <- rep(1/ng, ng)
+    } else {
+        counts <- as.vector(table(g))
+        prior <- counts / n
+    }
+
+    MASS::lda(x, g, prior)
+}
+

### 2015-06-08 code/cluster.R

@@ -1,0 +1,27 @@
+# code/cluster.R
+
+
+# default seed chosen by random.org
+cluster_kmeans <- function(x, centers, seed = 2651513, nstart = 100, ...)
+{
+    if (!is.na(seed)) {
+        if ((exists0 <- exists(".Random.seed", globalenv()))) {
+            seed0 <- .Random.seed
+        }
+        set.seed(seed)
+    }
+
+    cl <- stats::kmeans(x, centers, nstart=nstart, ...)
+
+    if (!is.na(seed)) {
+        if (exists0) {
+            .Random.seed <- seed0
+        } else {
+            rm(".Random.seed", globalenv())
+        }
+    }
+
+    cl
+}
+
+

### 2015-06-08 code/null-1d.R

@@ -1,119 +1,0 @@
-# code/null-1d.R
-# experiments for the null case (1 cluster) in one dimesion
-
-library("MASS") # lda
-library("RColorBrewer") # brewer.pal
-
-palette(brewer.pal(6, "Set1"))
-
-
-# default seed chosen by random.org
-cluster <- function(x, centers, seed = 2651513, nstart = 100, ...)
-{
-    if (!is.na(seed)) {
-        if ((exists0 <- exists(".Random.seed", globalenv()))) {
-            seed0 <- .Random.seed
-        }
-        set.seed(seed)
-    }
-
-    cl <- kmeans(x, centers, nstart=nstart, ...)
-
-    if (!is.na(seed)) {
-        if (exists0) {
-            .Random.seed <- seed0
-        } else {
-            rm(".Random.seed", globalenv())
-        }
-    }
-
-    cl
-}
-
-
-
-
-
-set.seed(0)
-
-ntrain <- 100000
-ntest <- ntrain
-n <- ntest + ntrain
-train <- seq_len(ntrain)
-test <- ntrain + seq_len(ntest)
-
-predictor <- 1
-response <- 2
-
-z <- matrix(rnorm(n * 2), n, 2)
-
-cverr <- function(correlation, centers)
-{
-    r <- correlation
-    coef <- cbind(rbind(sqrt((1 + r) / 2),  sqrt((1 - r) / 2)),
-                  rbind(sqrt((1 + r) / 2), -sqrt((1 - r) / 2)))
-    x <- z %*% coef
-
-    if (centers == 1) {
-        fit <- colMeans(x[train,response,drop=FALSE])
-        err <- scale(x[test,response], center=fit, scale=FALSE)
-    } else {
-        cl <- cluster(x[train,response], centers=centers)
-        cluster <- factor(cl$cluster, levels=seq_len(centers))
-        fit <- lda(x[train,predictor,drop=FALSE], cluster)
-        pred <- predict(fit, x[test,predictor,drop=FALSE])$class
-        err <- x[test,response] - cl$centers[pred,]
-    }
-
-    mse <- mean(err^2)
-    mse
-}
-
-
-correlation <- seq(-1, 1, length.out=21)
-mse1 <- sapply(correlation, function(r) cverr(r, 1))
-mse2 <- sapply(correlation, function(r) cverr(r, 2))
[…]

### 2015-06-08 figure/null-2d-corr.R

@@ -1,0 +1,49 @@
+# figure/null-2d-corr.R
+#
+
+
+library("RColorBrewer") # brewer.pal
+palette(brewer.pal(6, "Set1"))
+
+data <- read.csv("../sim/null-2d-corr.csv")
+data$replicate <- as.factor(data$replicate)
+data$centers <- as.factor(data$centers)
+
+mar <- c(3.25,3.25,0.75, 4.50 + 0.75)
+mai <- mar * par("csi")
+w <- h <- 3
+
+for (prior in levels(data$prior)) {
+    filename <- paste0("null-2d-corr-", prior, ".pdf")
+    pdf(filename, width=sum(w, mai[c(2,4)]), height=sum(h, mai[c(1,3)]))
+
+    d <- data[data$prior == prior,]
+
+    par(las=1, mai=mai, mgp=c(2.25,0.75,0), ps=11, tcl=-0.4)
+    plot(d$correlation, d$mse, type="n",
+         xlab="Correlation", ylab="CV Error")
+    axis(3, label=FALSE)
+    axis(4, label=FALSE)
+
+    for (i in seq_len(nlevels(data$centers))) {
+        k <- levels(data$centers)[i]
+        dk <- data[data$centers == k & data$prior == prior,]
+        points(dk$correlation, dk$mse, pch=i, col=i)
+    }
+
+
+    centers <- levels(data$centers)
+    ncenters <- nlevels(data$centers)
+
+    usr <- par("usr")
+    cxy <- par("cxy")
+    legend.x <- usr[2] + 3 * cxy[1]
+    legend.y <- usr[3]
+
+    legend(legend.x, legend.y, xjust=0, yjust=0, xpd=TRUE, bty="n",
+           title="Clusters", title.adj=0,
+           legend=centers, pch=1:ncenters, col=1:ncenters)
+
+    dev.off()
+}
+

### 2015-06-08 sim/null-2d-corr.R

@@ -1,0 +1,68 @@
+# code/null-2d-corr.R
+
+library("MASS")
+source("../code/classify.R")
+source("../code/cluster.R")
+
+
+set.seed(7627579) # from random.org
+
+ntrain <- 10000
+ntest <- ntrain
+n <- ntest + ntrain
+train <- seq_len(ntrain)
+test <- ntrain + seq_len(ntest)
+
+predictor <- 1
+response <- 2
+
+nrep <- 10
+
+sim <- NULL
+
+for (r in seq_len(nrep)) {
+    z <- matrix(rnorm(n * 2), n, 2)
+
+    cverr <- function(correlation, centers, prior)
+    {
+        r <- correlation
+        coef <- cbind(rbind(sqrt((1 + r) / 2),  sqrt((1 - r) / 2)),
+                      rbind(sqrt((1 + r) / 2), -sqrt((1 - r) / 2)))
+        x <- z %*% coef
+
+        if (centers == 1) {
+            fit <- colMeans(x[train,response,drop=FALSE])
+            err <- scale(x[test,response], center=fit, scale=FALSE)
+        } else {
+            cl <- cluster_kmeans(x[train,response], centers=centers)
+            cluster <- factor(cl$cluster, levels=seq_len(centers))
+            fit <- classify_lda(x[train,predictor,drop=FALSE], cluster, prior)
+            pred <- predict(fit, x[test,predictor,drop=FALSE])$class
+            err <- x[test,response] - cl$centers[pred,]
+        }
+
+        mse <- mean(err^2)
+        mse
+    }
+
+
+    correlation <- seq(-1, 1, length.out=21)
+    for (prior in c("equal", "proportions")) {
+        for (centers in c(1, 2, 3, 4, 5)) {
+            mse <- sapply(correlation, function(r) cverr(r, centers, prior))
+
+            sim <- rbind(sim,
+                         data.frame(replicate = r,
+                                    prior = prior,
+                                    correlation = correlation,
+                                    mse = mse,
+                                    centers = centers))
+        }
+    }
+}
+
+sim$replicate <- as.factor(sim$replicate)
+sim$centers <- as.factor(sim$centers)
+
+write.csv(sim, "null-2d-corr.csv", row.names=FALSE)
+

### 2015-06-08 cvclust.tex
Describe simulation, add figure

@@
 
-We confirm this with a simulation.  [TODO: add simulation]
+We confirm this with a simulation.  We perform $10$ replicates.  In each
+replicate, we generate $20000$ observations from a mean-zero bivariate normal
+distribution with unit marginal variances and correlation $\rho$.  We perform
+a single $2 \times 2$ fold of Gabriel cross-validation and report the
+cross-validation mean squared error for the number of clusters $k$ ranging
+from $1$ to $5$.  Figure~\ref{fig:null-2d-corr-equal} shows the
+cross-validation errors for all $10$ replicates.  The simulation demonstrates
+that in the Gabriel cross-validation criterion chooses the correct answer $k =
+1$ whenever $\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever
+$|\rho| > 0.5$.
+
+\begin{figure}
+\centering
+\includegraphics[width=25pc]{figure/null-2d-corr-equal.pdf}
+\caption{Cross-validation error on $10$ replicates, with the number of
+clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
+multivariate normal distribution with correlation $\rho$.  The Gabriel
+cross-validation criterion chooses the correct answer $k = 1$ whenever
+$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
+\label{fig:null-2d-corr-equal}
+\end{figure}
 

### 2015-06-08 cvclust.tex
Add missing period

@@
 variables will be determined according to whether $X > 0$; if $\rho < 0$ then
-the decision is according to whether $X < 0$  In the $\rho > 0$ case, the
+the decision is according to whether $X < 0$.  In the $\rho > 0$ case, the
 error converges to $\E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]$,

### 2015-06-09 cvclust.tex
changed 'integrate by parts' to 'integrate'
Revise Gabriel description

@@
 
+\newcommand{\CV}{\operatorname{CV}}
 \newcommand{\E}{\operatorname{E}}
+\newcommand{\dataX}{\mathfrak{X}}
+\newcommand{\Xtrain}{X_{\text{train}}}
+\newcommand{\Ytrain}{Y_{\text{train}}}
+\newcommand{\Xtest}{X_{\text{test}}}
+\newcommand{\Ytest}{Y_{\text{test}}}
+\newcommand{\hGX}{\hat G^{X}}
+\newcommand{\hGY}{\hat G^{Y}}
+\newcommand{\bpi}{\bar \pi}
+\newcommand{\bmuX}{\bar \mu^{X}}
+\newcommand{\bmuY}{\bar \mu^{Y}}
 
@@
 
-\section{Cross-Validation in finding the number of clusters}
-Cross-Validation technique, one of the most commonly used and popular model selecting method in supervised learning, can not be used naively in unsupervised learning context.
-Since prediction error of new observation is calculated by its distance to the nearest cluster center, more cluster centers means much tighter fit to the feature space, and hence smaller distance (prediction error) of observation to its nearest cluster center.
-This holds even when the prediction is evaluated on an independent test set \citep{hastie2009elements}.
-Therefore, CV prefers larger $k$ if we do the CV naively.\\
-
-We propose a new strategy to locate the optimal $k$ via Cross Validation. Recall that in supervised leaning, for any CV replicate, we have data in four parts
- $\begin{pmatrix}
-  \mathbf{X}_{train} & \mathbf{Y}_{train}\\
-  \mathbf{X}_{test}  & \mathbf{Y}_{test}\\
-  \end{pmatrix}$.
-One builds learner using $\left(\mathbf{X}_{train}, \mathbf{Y}_{train}\right)$ and apply it to $ \mathbf{X}_{test}$ in order to get $\hat{\mathbf{Y}}_{test}$.
-Difference between $\hat{\mathbf{Y}}_{test}$ and $\mathbf{Y}_{test}$ is measured as prediction error.
-In unsupervised leaning context, there is no $\mathbf{Y}$, so we only get $\begin{pmatrix}
-  \mathbf{X}_{train}\\
-  \mathbf{X}_{test}\\
-  \end{pmatrix}$.
-\cite{perry2009cross} considered more general hold-outs in order for cross-validation to work in unsupervised learning.
-Specifically, two types of hold-outs are considered.
-The first one is Wold hold-out, which is ``speckled": we leave out random elements of the matrix $\mathbf{X}$ and use missing data algorithm like EM for prediction.
-The second one is Gabriel hold-out: we randomly partition the columns of $\mathbf{X}$ into ``predictors" and ``response" sets and then perform the CV as in supervised learning.
-This paper is focused on the Gabriel hold-out, we also developed the Wold hold-out version and included it in the comparison in simulation.
-\\\\
-We now give the details of how to implement the Gabriel cross-validation to locate the optimal cluster number $k$. The Wold cross-validation algorithm is described in Appendix $A$.
+\section{Cross-validation for selecting the number of clusters}
+
+Cross-validation is commonly used for model selection in supervised learning
+problems.  In these settings, the data comes in the form of $N$
+predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
+\mathbb{R}^{p}$ and $Y_i \in \mathbb{R}^{q}$.  The data can be represented as
+a matrix with $N$ rows and $p + q$ columns.  We partition the data into $K$
+hold-out ``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
+``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
+matrix to get $\dataX$, a matrix with the $r$th test subset as its trailing
+rows.  We partition $\dataX$ as
+\[
+  \dataX =
+  \begin{bmatrix}
+    \Xtrain & \Ytrain \\
+    \Xtest  & \Ytest
+  \end{bmatrix}.
+\]
+We use the training rows $[ \Xtrain\ \Ytrain ]$ to fit a regression model
+$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
+test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
+or some variant thereof.  We choose the model with the smallest
+cross-validation error, averaged over all $K$ folds.
+
+In unsupervised learning problems like factor analysis and clustering, the
+features of the observations are not naturally partitioned into ``predictors''
+and ``responses'', so we cannot directly apply the cross-validation procedure
+described above.  For factor analysis, there are at least two versions of
+cross-validation.  \citet{wold78cross} proposed a ``speckled'' holdout, where
+in each fold we leave out a subset of the elements of the data matrix.  Wold's
+procedure works well empirically, but does not have any theoretical support,
+and it requires a factor analysis procedure that can handle missing data.
+\citet{owen2009bi} proposed a scheme called ``bi-cross-validation'' wherein
+each fold designates a subset of the data matrix columns to be response and a
+subset of the rows to be test data.  This generalized a procedure due to
+\citet{gabriel2002biblot}, who proposed holding out a single column and a
+single row at each fold.  Owen and Perry proved that this procedure is
+self-consistent, in the sense that it performs the correct model selection in
+the absence of noise, and \citet{perry2009cross} provided more theoretical
+support.
+
+In this report, we extend the Wold and Gabriel methods to the clustering
+problem, specifically to choose an appropriate number of clusters for a
+dataset.  We prove that the Gabriel method is self-consistent, and we analyze
+some of its properties in the presence of noise.  We compare these methods to
+state-of-the-art algorithms, and show that both are competitive.
+
+%% [POP] This paragraph doesn't fit into the section.  Maybe put it in the introduction?
+%% 
+%% Cross-Validation technique, one of the most commonly used and popular model
+%% selecting method in supervised learning, can not be used naively in
+%% unsupervised learning context. Since prediction error of new observation is
+%% calculated by its distance to the nearest cluster center, more cluster centers
+%% means much tighter fit to the feature space, and hence smaller distance
+%% (prediction error) of observation to its nearest cluster center. This holds
+%% even when the prediction is evaluated on an independent test set
+%% \citep{hastie2009elements}. Therefore, CV prefers larger $k$ if we do the CV
+%% naively.
+
+We now give the details of how to implement
+the Gabriel cross-validation to locate the optimal cluster number $k$. The
+Wold cross-validation algorithm is described in Appendix $A$.
 
 \subsection*{Gabriel CV algorithm}
-The basic idea of Gabriel hold-out is to ``create" the label $(Y)$ for data without it, and then perform the cross-validation as in supervised learning. Specifically, given data set 
-$\mathbf{X} \in \mathbb{R}^{n \times p}$, we partition it into four parts
+
+We are given a data matrix with $N$ rows and $P$ columns.  In each fold of
+cross-validation, we permute the rows and columns of the data matrix and then
+partition the rows and columns as $N = n + m$ and $P = p + q$ for 
+non-negative integers $n$, $m$, $p$, and $q$.  We treat the first $p$
+columns as ``predictors'' and the last $q$ columns as ``responses'';
+similarly, we treat the first $n$ rows as ``training'' and the last $m$ rows
+as ``test''.  In block form, the permuted data matrix is
 \[
-\begin{pmatrix}
-  \mathbf{X}_{train} & \mathbf{Y}_{train}\\
-  \mathbf{X}_{test}  & \mathbf{Y}_{test}\\
-  \end{pmatrix} \equiv 
-  \begin{pmatrix}
-  \mathbf{X}^{m \times q } & \mathbf{X}^{m \times (p-q)}\\
-  \mathbf{X}^{(n-m) \times q}  & \mathbf{X}^{(n-m) \times (p-q)}\\
-  \end{pmatrix}
+  \dataX
+  =
+  \begin{bmatrix}
+    \Xtrain & \Ytrain \\
+    \Xtest  & \Ytest
+  \end{bmatrix},
 \]
-  where $ 1 < m < n$ and $1<q<p$.
-That is, we leave out $n-m$ rows of data as test set, and make $p-q$ features as ``response".
-Note that the data set looks like the one in supervised learning now.\\\\
-To keep it consistent with the notation used in section $3$, we also denote above partition of data $\mathbf{X}$ as 
+where
+$\Xtrain \in \mathbb{R}^{n \times p}$,
+$\Ytrain \in \mathbb{R}^{n \times q}$,
+$\Xtest \in \mathbb{R}^{m \times p}$,
+and
+$\Ytest \in \mathbb{R}^{m \times q}$.
+
+Given such a partition of $\dataX$, we perform four steps for each value of
+$k$, the number of clusters:
+\begin{enumerate}
+  \item \label{step:gabriel-cluster}
+    \textbf{Cluster:}
+    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding the
+    assignment rule $\hGY : \mathbb{R}^q \to \{ 1, \dotsc, k \}$ and the
+    clusters means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
+    be the assigned cluster for row $i$.
+
+  \item \label{step:gabriel-classify}
+    \textbf{Classify:}
+    Take $X_{1}, \dotsc, X_n$, the rows of $\Xtrain$ to be predictors,
+    and take $\hGY_1, \dotsc, \hGY_n$ to be corresponding class labels.  Use
+    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to train a classifier
+    $\hGX : \mathbb{R}^p \to \{ 1, \dotsc, k \}$.
+
+  \item \label{step:gabriel-predict}
+    \textbf{Predict:}
+    Apply the classifier to $X_{n+1}, \dotsc, X_{n+m}$, the rows of
+    $\Xtest$, yielding predicted classes $\hGX_i = \hGX(X_i)$ for
+    $i = n+1, \dotsc, n+m$.  For each value of $i$ in this range, compute
+    predicted response $\hat Y_i = \bmuY(\hGX_i)$, where
+    $\bmuY(g) = \bmuY_g$.
+
+  \item \label{step:gabriel-evaluate}
+    \textbf{Evaluate:}
+    Compute the cross-validation error
     \[
-\begin{pmatrix}
-  \mathbf{X}_{train} & \mathbf{Y}_{train}\\
-  \mathbf{X}_{test}  & \mathbf{Y}_{test}\\
-  \end{pmatrix} \equiv 
-  \begin{pmatrix}
-\mathbf{X}^L_{train} & \mathbf{X}^R_{train}\\
-  \mathbf{X}^L_{test}  & \mathbf{X}^R_{test}\\
-  \end{pmatrix}
+      \CV(k) = \frac{1}{m} \sum_{i=n+1}^{n+m} \|Y_i - \hat Y_i\|^2,
     \]
-That is, data set $\mathbf{X} \in \mathbb{R}^{n \times p}$ been partitioned in column as $\mathbf{X}^L \in \mathbb{R}^{n \times L}$ and $\mathbf{X}^R \in \mathbb{R}^{n \times R}$, where $\mathbf{X}^L$ is treated as features ($L=q$) and $\mathbf{X}^R$ is treated as response ($R=p-q$).\\
-\begin{itemize}
-	\item For each $k = 1,2,...,k_{max}$, the Gabriel algorithm works as following
-	\begin{enumerate}
- 	\item A set of $k$ clusters (labels) $ \mathcal{G}= \{g_1,...,g_k\} $ are returned by applying K-means on $\mathbf{X}^R_{train}$ with parameter $k$, $\hat{\mu}^R \left(g_j\right)$ is the corresponding estimated cluster center for $g_j$, $1 \leq j \leq k$.
- 	\item Train a classifier $\mathcal{C}$ using $\left(\mathbf{X}^L_{train}, \mathcal{G} \right)$. We implement LDA in our algorithm.
- 	\item Applying the classifier $\mathcal{C}$ on $\mathbf{X}^L_{test}$, we got a set of predicted clusters $\hat{\mathcal{G}} = \{\hat{g} \left(x^L_i\right) \in \mathcal{G}, i = m+1,...,n \}$, where $\hat{g} \left(x^L_i\right)$ is the predicted (assigned) cluster for observation $x_i$ in the testing data set
- 	\item Calculate the difference between $\hat{\mathcal{G}}$ and  $\mathbf{X}^R_{test}$ as
- 	\[Error_k = \sum^n_{i = m+1} ||\hat{g} \left(x^L_i\right) - x^R_i||^2 = \sum^k_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2\]
+    where $Y_{n+1}, \dotsc, Y_{n+m}$ are the rows of $\Ytest$.
 \end{enumerate}
-	Note that one can use classification technique other than LDA in step $2.$
 \noindent
-	\item The algorithm selects $k$ with minimum average $Error_k$ across all CV folds, that is $$\underset{k}{\operatorname{argmin}} \hspace{0.05 in} E\left(Error_k\right) $$
-\end{itemize}
+In principle, we could use any clustering and classification methods in
+steps~\ref{step:gabriel-cluster} and~\ref{step:gabriel-classify}.  In this
+report, we use $k$-means as the clustering algorithm and linear discriminant
+analysis (LDA) as the classification procedure.
+
+To choose the folds, we randomly partition the rows and columns into $K$ and
+$L$ subsets, respectively.  Each fold is indexed by a pair $(r,s)$ of
+integers, with $r \in \{1, \dotsc, K\}$ and $s \in \{1, \dotsc, L\}$.  Fold
+$(r,s)$ treats the $r$th row subset as ``test'', and the $s$th column subset
+as ``response''.  Following \citet{owen2009bi}, we take $K = L = 2$.  For the
+number of clusters, we select the value of $k$ that minimizes the average of
+$\CV(k)$ over all $K \times L$ folds.
+
 \section{Theoretical justification of Gabriel CV method}
@@
 We will derive the expression for the second moment.
-Integrate by parts to get
+Integrate to get
 \begin{align*}

### 2015-06-10 cvclust.tex
Typo correction

@@
 Our proposed method is a complete data-driven approach which doesn't rely on any parametric assumptions.
-Trough novel form of partitioning data set, we are able to employing the cross-validation procedure in clustering exactly the same way as in supervised learning problem.
+Through novel form of partitioning data set, we are able to employing the cross-validation procedure in clustering exactly the same way as in supervised learning problem.
 Hence, it's easy for reader to understand the intuition behind our proposed method.

### 2015-06-10 cvclust.tex
clean up elbow plots

@@
 
-After being proposed more than $50$ years, $K$-mean remains one of the most popular and widely used clustering algorithms \citep{jain2010data}.
-Like many other clustering methods, $K$-mean requires an input parameter $k$, the number of clusters, to be specified by the user.
-Automatically and quantitatively deciding such parameter is important and yet unsolved problem \citep{fujita2014non}.
-Various methods have been proposed to tackle this difficulty.
-One ad hoc approach is to explore the relationship between $W_k$ (within-cluster dispersion) and the number of cluster $k$ for a certain clustering method such as $K$-mean.
-Since $W_k$ decreases as $k$ increases, one usually find the ``elbow" of curve $W_k$ \textsl{VS} $k$ as the appropriate number of clusters.
-The example on the top row of Figure \ref{fig1} demonstrates such approach for data with $k=4$, where the ``elbow" point indeed reveals the true number of clusters.
-This is based on the idea that under partitioning data set has more impact than over partitioning data set in terms of $W_k$.
-However, locating the ``elbow" point is somewhat subjective and sometimes is not appropriate to select the optimal $k$.
-The second example on the bottom row of Figure \ref{fig1} shows a  situation where there is no clear choice of the ``elbow" point -- both $k=2$ and $k=3$ can be viewed as the ``elbow" point.
+After being proposed more than $50$ years, $K$-mean remains one of the most
+popular and widely used clustering algorithms \citep{jain2010data}. Like many
+other clustering methods, $K$-mean requires an input parameter $k$, the number
+of clusters, to be specified by the user. Automatically and quantitatively
+deciding such parameter is important and yet unsolved problem
+\citep{fujita2014non}. Various methods have been proposed to tackle this
+difficulty. One ad hoc approach is to explore the relationship between $W_k$
+(within-cluster dispersion) and the number of cluster $k$ for a certain
+clustering method such as $K$-mean. Since $W_k$ decreases as $k$ increases,
+one usually find the ``elbow" of curve $W_k$ \textsl{VS} $k$ as the
+appropriate number of clusters. The example on the top row of Figure
+\ref{fig:elbow} demonstrates such approach for data with $k=4$, where the
+``elbow" point indeed reveals the true number of clusters. This is based on
+the idea that under partitioning data set has more impact than over
+partitioning data set in terms of $W_k$. However, locating the ``elbow" point
+is somewhat subjective and sometimes is not appropriate to select the optimal
+$k$.
+The second example on the bottom row of Figure \ref{fig:elbow} shows a  situation where there is no clear choice of the ``elbow" point -- both $k=2$ and $k=3$ can be viewed as the ``elbow" point.
 What's more, the true $k=4$ can never be selected as the optimal $k$ using such approach in this case since it can hardly be viewed as the ``elbow" of the curve.
+
 \begin{figure}
+\centering
 \begin{minipage}{\linewidth}
-\centering
   \begin{minipage}{0.45\linewidth}
-\includegraphics[width=\linewidth]{figure/Data_example.pdf}
+    \includegraphics[width=\linewidth]{figure/elbow-correct-data.pdf}
   \end{minipage}
@@
   \begin{minipage}{0.45\linewidth}
-\includegraphics[width=\linewidth]{figure/W_k.pdf}
+    \includegraphics[width=\linewidth]{figure/elbow-correct-withinss.pdf}
   \end{minipage}
 \end{minipage}
-%\center{\caption{\label{fig1} Left plot shows the actual data, and the right plot shows the $W_k$ \textsl{VS} $k$ curve. }}
-%\end{figure}
-
-%\begin{figure}
 \begin{minipage}{\linewidth}
-\centering
   \begin{minipage}{0.45\linewidth}
-\includegraphics[width=\linewidth]{figure/Data_elbow.pdf}
+    \includegraphics[width=\linewidth]{figure/elbow-incorrect-data.pdf}
   \end{minipage}
@@
   \begin{minipage}{0.45\linewidth}
-\includegraphics[width=\linewidth]{figure/elbow.pdf}
+    \includegraphics[width=\linewidth]{figure/elbow-incorrect-withinss.pdf}
   \end{minipage}
 \end{minipage}
-\center{\caption{\label{fig1} Left plots show the actual data, right plots show the corresponding $W_k$ \textsl{VS} $k$ curve. }}
+\caption{Left panels show the $(X,Y)$ data points; right panels
+  show the corresponding values of the within-cluster sum of squares $W_k$
+plotted against the number of clusters, $k$.}
+\label{fig:elbow}
 \end{figure}
-\\
 

### 2015-06-10 figure/elbow.R

@@ -1,0 +1,110 @@
+# figure/elbow
+
+library("MASS")         # mvrnorm
+library("RColorBrewer") # brewer.pal
+
+source("../code/cluster.R") # cluster_kmeans
+
+
+palette(brewer.pal(6, "Set1"))
+
+##---------------------Top plot of figure 1-----------------------------------
+
+set.seed(3) # seed chosen to make well-separated clusters
+
+x1 <- mvrnorm(20, c(1.7, 3.0), 0.15 * diag(2))
+x2 <- mvrnorm(20, c(1.0, 1.0), 0.15 * diag(2))
+x3 <- mvrnorm(20, c(3.0, 1.5), 0.15 * diag(2))
+x4 <- mvrnorm(20, c(3.7, 3.7), 0.10 * diag(2))
+
+data <- rbind(data.frame(x=x1[,1], y=x1[,2], group="1"),
+              data.frame(x=x2[,1], y=x2[,2], group="2"),
+              data.frame(x=x3[,1], y=x3[,2], group="3"),
+              data.frame(x=x4[,1], y=x4[,2], group="4"))
+
+
+
+mar <- c(3.25, 3.25, 1.75, 1.75)
+mai <- mar * par("csi")
+w <- h <- 3
+mypar <- list(las=1, mai=mai, mgp=c(2.25, 0.75, 0), ps=11, tcl=-0.4)
+
+width <- w + sum(mai[c(1,3)])
+height <- h + sum(mai[c(2,4)])
+
+
+
+pdf("elbow-correct-data.pdf", width=width, height=height)
+par(mypar)
+
+plot(data$x, data$y, type="n", xlab="X", ylab="Y")
+points(data$x, data$y, pch=as.integer(data$group),
+       col=as.integer(data$group))
+axis(3, labels=FALSE)
+axis(4, labels=FALSE)
+dev.off()
+
+
+centers <- 1:10
+tot.withinss <- rep(NA, length(centers))
+for (i in seq_along(centers)) {
+ 	cl <- cluster_kmeans(cbind(data$x, data$y), centers[i])
+    tot.withinss[i] <- cl$tot.withinss
+}
+
+
+pdf("elbow-correct-withinss.pdf", width=width, height=height)
+par(mypar)
+
+plot(centers, tot.withinss, type="b",
+     xlab = "Clusters",
+     ylab = "Within SS")
+axis(3, labels=FALSE)
+axis(4, labels=FALSE)
+
+dev.off()
+
+
[…]

### 2015-06-10 wei/elbow plot.R

@@ -1,29 +1,0 @@
-
-library("clusterCrit")
-library(MASS)
-
-pdf("Data_elbow.pdf")
-D1 <- mvrnorm(100,c(5,5),3*diag(2))
-plot(D1[,1],D1[,2],ylim=c(0,15),xlim=c(0,15),col="red",xlab="X1",ylab="X2")
-
-D2 <- mvrnorm(20,c(5,12),diag(2))
-points(D2[,1],D2[,2],ylim=c(0,15),xlim=c(0,15),col="blue")
-
-D3 <- mvrnorm(10,c(9,9),0.5*diag(2))
-points(D3[,1],D3[,2],ylim=c(0,15),xlim=c(0,15),col="green")
-
-D4 <- mvrnorm(100,c(12,12),diag(2))
-points(D4[,1],D4[,2],ylim=c(0,15),xlim=c(0,15))
-dev.off()
-
-
-Data <- rbind(D1,D2,D3,D4)
-
-Disp <- rep(NA,10)
-for(w in 1:10){
- 	cl <- kmeans(Data,w,nstart = 100)
-      Disp[w] <- intCriteria(Data,cl$cluster,"trace_w")[[1]]
-}
-pdf("elbow.pdf")
-plot(1:10,Disp,type="b",ylab = "Within-cluster dispersion Wk",xlab="Number of clusters k")
-dev.off()

### 2015-06-10 wei/elbow plot_new.R

@@ -1,61 +1,0 @@
-
-library("clusterCrit")
-library(MASS)
-
-##---------------------Top plot of figure 1-----------------------------------
-set.seed(3)
-
-pdf("Data_example.pdf")
-D1 <- mvrnorm(20,c(1.7,3),0.15*diag(2))
-plot(D1[,1],D1[,2],ylim=c(0,4.5),xlim=c(0,4.5),col="red",xlab="X1",ylab="X2")
-
-D2 <- mvrnorm(20,c(1,1),0.15*diag(2))
-points(D2[,1],D2[,2],ylim=c(0,5),xlim=c(0,5),col="blue")
-
-D3 <- mvrnorm(20,c(3,1.5),0.15*diag(2))
-points(D3[,1],D3[,2],ylim=c(0,5),xlim=c(0,5),col="green")
-
-D4 <- mvrnorm(20,c(3.7,3.7),0.1*diag(2))
-points(D4[,1],D4[,2],ylim=c(0,5),xlim=c(0,5))
-dev.off()
-
-
-Data <- rbind(D1,D2,D3,D4)
-
-Disp <- rep(NA,10)
-for(w in 1:10){
- 	cl <- kmeans(Data,w,nstart = 100)
-      Disp[w] <- intCriteria(Data,cl$cluster,"trace_w")[[1]]
-}
-pdf("W_k.pdf")
-plot(1:10,Disp,type="b",ylab = "Within-cluster dispersion Wk",xlab="Number of clusters k")
-dev.off()
-
-##---------------------Bottom plot of figure 1-----------------------------------
-set.seed(1)
-
-pdf("Data_elbow.pdf")
-D1 <- mvrnorm(100,c(5,5),3*diag(2))
-plot(D1[,1],D1[,2],ylim=c(0,15),xlim=c(0,15),col="red",xlab="X1",ylab="X2")
-
-D2 <- mvrnorm(20,c(5,12),diag(2))
-points(D2[,1],D2[,2],ylim=c(0,15),xlim=c(0,15),col="blue")
-
-D3 <- mvrnorm(10,c(9,9),0.5*diag(2))
-points(D3[,1],D3[,2],ylim=c(0,15),xlim=c(0,15),col="green")
-
-D4 <- mvrnorm(100,c(12,12),diag(2))
-points(D4[,1],D4[,2],ylim=c(0,15),xlim=c(0,15))
-dev.off()
-
-
-Data <- rbind(D1,D2,D3,D4)
-
[…]

### 2015-06-10 figure/elbow.R
dos2unix 'elbow.R'

@@ -1,6 +1,6 @@
-# figure/elbow
+# figure/elbow.R
 
-library("MASS")         # mvrnorm
+library("MASS") 	# mvrnorm
 library("RColorBrewer") # brewer.pal
 
 source("../code/cluster.R") # cluster_kmeans
@@ -18,9 +18,9 @@ x3 <- mvrnorm(20, c(3.0, 1.5), 0.15 * diag(2))
 x4 <- mvrnorm(20, c(3.7, 3.7), 0.10 * diag(2))
 
 data <- rbind(data.frame(x=x1[,1], y=x1[,2], group="1"),
-              data.frame(x=x2[,1], y=x2[,2], group="2"),
-              data.frame(x=x3[,1], y=x3[,2], group="3"),
-              data.frame(x=x4[,1], y=x4[,2], group="4"))
+	      data.frame(x=x2[,1], y=x2[,2], group="2"),
+	      data.frame(x=x3[,1], y=x3[,2], group="3"),
+	      data.frame(x=x4[,1], y=x4[,2], group="4"))
 
 
 
@@ -48,7 +48,7 @@ dev.off()
 centers <- 1:10
 tot.withinss <- rep(NA, length(centers))
 for (i in seq_along(centers)) {
- 	cl <- cluster_kmeans(cbind(data$x, data$y), centers[i])
+	cl <- cluster_kmeans(cbind(data$x, data$y), centers[i])
     tot.withinss[i] <- cl$tot.withinss
 }
 
@@ -69,15 +69,15 @@ dev.off()
 
 set.seed(1) # seed chosen to make well-separated clusters
 
-x1 <- 0.3 * mvrnorm(100, c( 5,  5), 3.0 * diag(2))
-x2 <- 0.3 * mvrnorm( 20, c( 5, 12),       diag(2))
-x3 <- 0.3 * mvrnorm( 10, c( 9,  9), 0.5 * diag(2))
-x4 <- 0.3 * mvrnorm(100, c(12, 12),       diag(2))
+x1 <- 0.3 * mvrnorm(100, c( 5,	5), 3.0 * diag(2))
+x2 <- 0.3 * mvrnorm( 20, c( 5, 12),	  diag(2))
+x3 <- 0.3 * mvrnorm( 10, c( 9,	9), 0.5 * diag(2))
+x4 <- 0.3 * mvrnorm(100, c(12, 12),	  diag(2))
 
 data <- rbind(data.frame(x=x1[,1], y=x1[,2], group="1"),
[…]

### 2015-06-10 cvclust.tex
reorg project structure

@@
   \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{figure/elbow-correct-data.pdf}
+    \includegraphics[width=\linewidth]{demo/elbow/correct-data.pdf}
   \end{minipage}
@@
   \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{figure/elbow-correct-withinss.pdf}
+    \includegraphics[width=\linewidth]{demo/elbow/correct-withinss.pdf}
   \end{minipage}
@@
   \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{figure/elbow-incorrect-data.pdf}
+    \includegraphics[width=\linewidth]{demo/elbow/incorrect-data.pdf}
   \end{minipage}
@@
   \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{figure/elbow-incorrect-withinss.pdf}
+    \includegraphics[width=\linewidth]{demo/elbow/incorrect-withinss.pdf}
   \end{minipage}
@@
 cross-validation mean squared error for the number of clusters $k$ ranging
-from $1$ to $5$.  Figure~\ref{fig:null-2d-corr-equal} shows the
+from $1$ to $5$.  Figure~\ref{fig:nullcorr-equal} shows the
 cross-validation errors for all $10$ replicates.  The simulation demonstrates
@@
 \centering
-\includegraphics[width=25pc]{figure/null-2d-corr-equal.pdf}
+\includegraphics[width=25pc]{demo/nullcorr/equal.pdf}
 \caption{Cross-validation error on $10$ replicates, with the number of
@@
 $\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
-\label{fig:null-2d-corr-equal}
+\label{fig:nullcorr-equal}
 \end{figure}

### 2015-06-10 demo/elbow/01_make_plots.R

@@ -1,12 +1,11 @@
-# figure/elbow.R
+#!/usr/bin/Rscript --vanilla
 
 library("MASS") 	# mvrnorm
 library("RColorBrewer") # brewer.pal
 
-source("../code/cluster.R") # cluster_kmeans
+source("../../code/cluster.R") # cluster_kmeans
 
 
-palette(brewer.pal(6, "Set1"))
 
 ##---------------------Top plot of figure 1-----------------------------------
 
@@ -25,7 +24,8 @@ data <- rbind(data.frame(x=x1[,1], y=x1[,2], group="1"),
 
 
 mar <- c(3.25, 3.25, 1.75, 1.75)
-mai <- mar * par("csi")
+csi <- 0.2 # default value of par("csi")
+mai <- mar * csi
 w <- h <- 3
 mypar <- list(las=1, mai=mai, mgp=c(2.25, 0.75, 0), ps=11, tcl=-0.4)
 
@@ -34,7 +34,8 @@ height <- h + sum(mai[c(2,4)])
 
 
 
-pdf("elbow-correct-data.pdf", width=width, height=height)
+pdf("correct-data.pdf", width=width, height=height)
+palette(brewer.pal(6, "Set1"))
 par(mypar)
 
 plot(data$x, data$y, type="n", xlab="X", ylab="Y")
@@ -53,7 +54,7 @@ for (i in seq_along(centers)) {
 }
 
 
-pdf("elbow-correct-withinss.pdf", width=width, height=height)
+pdf("correct-withinss.pdf", width=width, height=height)
 par(mypar)
 
 plot(centers, tot.withinss, type="b",
@@ -79,7 +80,7 @@ data <- rbind(data.frame(x=x1[,1], y=x1[,2], group="1"),
 	      data.frame(x=x3[,1], y=x3[,2], group="3"),
 	      data.frame(x=x4[,1], y=x4[,2], group="4"))
 
-pdf("elbow-incorrect-data.pdf", width=width, height=height)
+pdf("incorrect-data.pdf", width=width, height=height)
 par(mypar)
 
 plot(data$x, data$y, type="n", xlab="X", ylab="Y")
@@ -99,7 +100,7 @@ for (i in seq_along(centers)) {
 }
 
 
-pdf("elbow-incorrect-withinss.pdf", width=width, height=height)
+pdf("incorrect-withinss.pdf", width=width, height=height)
 par(mypar)
 
 plot(centers, tot.withinss, type="b",

### 2015-06-10 demo/nullcorr/01_run_simulation.R

@@ -1,8 +1,9 @@
-# code/null-2d-corr.R
+#!/usr/bin/Rscript --vanilla
+
 
 library("MASS")
-source("../code/classify.R")
-source("../code/cluster.R")
+source("../../code/classify.R")
+source("../../code/cluster.R")
 
 
 set.seed(7627579) # from random.org
@@ -64,5 +65,5 @@ for (r in seq_len(nrep)) {
 sim$replicate <- as.factor(sim$replicate)
 sim$centers <- as.factor(sim$centers)
 
-write.csv(sim, "null-2d-corr.csv", row.names=FALSE)
+write.csv(sim, "nullcorr.csv", row.names=FALSE)
 

### 2015-06-10 demo/nullcorr/02_plot_results.R

@@ -1,21 +1,21 @@
-# figure/null-2d-corr.R
-#
+#!/usr/bin/Rscript --vanilla
 
 
 library("RColorBrewer") # brewer.pal
-palette(brewer.pal(6, "Set1"))
 
-data <- read.csv("../sim/null-2d-corr.csv")
+data <- read.csv("nullcorr.csv")
 data$replicate <- as.factor(data$replicate)
 data$centers <- as.factor(data$centers)
 
 mar <- c(3.25,3.25,0.75, 4.50 + 0.75)
-mai <- mar * par("csi")
+csi <- 0.2 # default value of par("csi")
+mai <- mar * csi
 w <- h <- 3
 
 for (prior in levels(data$prior)) {
-    filename <- paste0("null-2d-corr-", prior, ".pdf")
+    filename <- paste0(prior, ".pdf")
     pdf(filename, width=sum(w, mai[c(2,4)]), height=sum(h, mai[c(1,3)]))
+    palette(brewer.pal(6, "Set1"))
 
     d <- data[data$prior == prior,]
 

### 2015-06-11 demo/bench/01_generate_data.R
split out replicate generation code

@@ -1,0 +1,29 @@
+#!/usr/bin/Rscript --vanilla
+
+library("MASS")
+source("settings.R") # settings
+
+nrep <- 100
+
+for (s in names(settings)) {
+    if (!dir.exists(s)) {
+        dir.create(s)
+    }
+    filename <- file.path(s, "replicates.rds")
+    if (!file.exists(filename)) {
+
+        cat("generating replicates for '", s, "'\n", sep="")
+
+        set.seed(settings[[s]]$seed)
+        replicates <- vector("list", nrep)
+
+        pb <- txtProgressBar(0, nrep, style=3)
+        for (r in seq_len(nrep)) {
+            replicates[[r]] <- settings[[s]]$simulate()
+            setTxtProgressBar(pb, r)
+        }
+        close(pb)
+
+        saveRDS(replicates, filename)
+    }
+}

### 2015-06-11 demo/bench/settings.R

@@ -1,0 +1,307 @@
+# demo/bench/settings.R
+#
+# Depends:
+#   library("MASS") # mvrnorm
+#
+
+
+setting1 <- function()
+{
+    nobs <- 200
+    x <- matrix(NA, nobs, 10)
+    for(s in 1:10){
+        x[,s] <- runif(nobs, 0, 1)
+    }
+
+    ngroup <- 1
+    group <- as.factor(rep(1, nobs))
+    centers <- matrix(rep(0.5, 10), byrow=TRUE, nrow=1)
+
+    mean <- centers[group,]
+
+    list(x=x, mean=mean, ngroup=ngroup, group=group, centers=centers)
+}
+
+
+location <- function(M.data,M.center)
+{
+   if(ncol(M.data)!=ncol(M.center)){return("dimention don't match")}
+   center<-rep(NA,nrow(M.data))
+   K <- nrow(M.center)
+   for (i in 1:length(center)){
+     dist <- rep(NA,K)
+     for(j in 1:K){
+	 dist[j]<- sqrt(sum((M.center[j,]-M.data[i,])^2))
+     }
+     center[i]<- which(dist == min(dist))
+   }
+   return(center)
+}
+
+
+Data2center <- function(number = 50)
+{
+	u1 = c(1,0.0,0.0,1)
+	u2 = c(1,3.5,3.5,1)
+    Centers <- matrix(c(1,1,0,3.5,0,3.5,1,1),2,4)
+
+	condition <- TRUE
+	while(condition){
+	      obs1 <- mvrnorm(n = number, u1, diag(4))
+		obs2 <- mvrnorm(n = number, u2, diag(4))
+
+		if(sum(location(obs1,Centers)==1)==number
+           & sum(location(obs2,Centers)==2)==number){
+			condition <- FALSE
+		}
+	}
+
+	Data <- rbind(obs1,obs2)
+	Mu <- rbind(matrix(u1,nrow=number,ncol=4,byrow=T),
+                matrix(u2,nrow=number,ncol=4,byrow=T))
+	ID <- sample(1:nrow(Data),nrow(Data),replace = FALSE)
[…]

### 2015-06-11 code/cluster.R
fix issues with deleting .Random.seed in cluster_kmeans

@@ -5,7 +5,7 @@
 cluster_kmeans <- function(x, centers, seed = 2651513, nstart = 100, ...)
 {
     if (!is.na(seed)) {
-        if ((exists0 <- exists(".Random.seed", globalenv()))) {
+        if ((exists0 <- exists(".Random.seed", envir=globalenv()))) {
             seed0 <- .Random.seed
         }
         set.seed(seed)
@@ -15,13 +15,11 @@ cluster_kmeans <- function(x, centers, seed = 2651513, nstart = 100, ...)
 
     if (!is.na(seed)) {
         if (exists0) {
-            .Random.seed <- seed0
-        } else {
-            rm(".Random.seed", globalenv())
+            assign(".Random.seed", seed0, envir=globalenv())
+        } else if (exists(".Random.seed", envir=globalenv())) {
+            rm(".Random.seed", envir=globalenv())
         }
     }
 
     cl
 }
-
-

### 2015-06-11 demo/bench/02_fit_kmeans.R
add code to fit kmeans to bench data

@@ -1,0 +1,32 @@
+#!/usr/bin/Rscript --vanilla
+
+source("../../code/cluster.R") # cluster_kmeans
+
+kmax <- 10
+
+
+for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
+    f.replicates <- file.path(s, "replicates.rds")
+    f.kmeans     <- file.path(s, "kmeans.rds")
+
+    if (file.exists(f.replicates) && !file.exists(f.kmeans)) {
+        cat("fitting k-means for '", s, "'\n", sep="")
+
+        replicates <- readRDS(f.replicates)
+        nrep <- length(replicates)
+        kmeans <- vector("list", nrep)
+
+        pb <- txtProgressBar(0, nrep, style=3)
+        for (r in seq_len(nrep)) {
+            kmeans[[r]] <- vector("list", kmax)
+            for (k in seq_len(kmax)) {
+                kmeans[[r]][[k]] <- cluster_kmeans(replicates[[r]]$x, k)
+            }
+            setTxtProgressBar(pb, r)
+        }
+        close(pb)
+
+        saveRDS(kmeans, f.kmeans)
+    }
+}
+

### 2015-06-11 code/cluster.R
make cluster_kmeans use iter.max=100 as default

@@ -2,7 +2,8 @@
 
 
 # default seed chosen by random.org
-cluster_kmeans <- function(x, centers, seed = 2651513, nstart = 100, ...)
+cluster_kmeans <- function(x, centers, seed = 2651513, iter.max = 100,
+                           nstart = 100, ...)
 {
     if (!is.na(seed)) {
         if ((exists0 <- exists(".Random.seed", envir=globalenv()))) {
@@ -11,7 +12,7 @@ cluster_kmeans <- function(x, centers, seed = 2651513, nstart = 100, ...)
         set.seed(seed)
     }
 
-    cl <- stats::kmeans(x, centers, nstart=nstart, ...)
+    cl <- stats::kmeans(x, centers, iter.max=iter.max, nstart=nstart, ...)
 
     if (!is.na(seed)) {
         if (exists0) {

### 2015-06-11 lib/NbClust/R/NbClust.R
patch local NbClustpackage to use cluster_kmeans

@@ -1,3 +1,15 @@
+# lib/NbClust.R
+#
+# modified version from NbClust R package with 'kmeans' replaced
+# by 'cluster_kmeans'
+#
+# Original Authors: Malika Charrad, Nadia Ghazzali, Veronique Boiteau, and Azam Niknafs
+#
+# License GPL-2.
+#
+# Depends:
+#   source("../code/cluster.R")
+
 NbClust <-function(data = NULL, diss=NULL, distance ="euclidean", min.nc=2, max.nc=15, method =NULL, index = "all", alphaBeale = 0.1)
 {
     
@@ -1215,7 +1227,7 @@ Indice.Gap <- function (x, clall, reference.distribution = "unif", B = 10,
                   pp2 <- 1:ClassNr
                 else if (method == "k-means") 
                 { set.seed(1)
-                  pp2 <- kmeans(Xnew, ClassNr, 100)$cluster
+                  pp2 <- cluster_kmeans(Xnew, ClassNr)$cluster
                 }
                 else if (method == "single" || method == "complete" || 
                   method == "average" || method == "ward.D2" || 
@@ -1247,7 +1259,7 @@ Indice.Gap <- function (x, clall, reference.distribution = "unif", B = 10,
                 else if (method == "k-means")
                 {
                   set.seed(1)
-                  pp2 <- kmeans(Xnew, ClassNr, 100)$cluster
+                  pp2 <- cluster_kmeans(Xnew, ClassNr)$cluster
                 }
                 else if (method == "single" || method == "complete" || 
                   method == "average" || method == "ward.D2" || 
@@ -1414,22 +1426,22 @@ Indice.Gap <- function (x, clall, reference.distribution = "unif", B = 10,
   	if (method == 8) 
     {
       set.seed(1)
-		  cl2 <- kmeans(jeu,nc+1)$cluster
+		  cl2 <- cluster_kmeans(jeu,nc+1)$cluster
       set.seed(1)
-		  clmax <- kmeans(jeu,max_nc)$cluster
+		  clmax <- cluster_kmeans(jeu,max_nc)$cluster
       if (nc > 2)
 		  {
         set.seed(1)
-		    cl1 <- kmeans(jeu,nc)$cluster
+		    cl1 <- cluster_kmeans(jeu,nc)$cluster
 		    clall <- cbind(cl1, cl2)
         set.seed(1)
[…]

### 2015-06-11 lib/NbClust/R/cluster.R

@@ -1,0 +1,1 @@
+../../../code/cluster.R

### 2015-06-11 lib/fpc/R/addclustermethods.R
patch fpc to use cluster_kmeans

@@ -84,7 +84,7 @@ dudahart2 <- function(x,clustering,alpha=0.001){
 
 
 kmeansruns <- function(data,krange=2:10,criterion="ch",
-                       iter.max=100,runs=100,
+                       runs=100,
                        scaledata=FALSE,alpha=0.001,
                        critout=FALSE,plot=FALSE,...){
   data <- as.matrix(data)
@@ -101,7 +101,7 @@ kmeansruns <- function(data,krange=2:10,criterion="ch",
         options(show.error.messages = FALSE)
         repeat{
 #          cat(k," ",i,"before \n")
-          kmm <- try(kmeans(data,k,iter.max=iter.max,...))
+          kmm <- try(cluster_kmeans(data,k,...))
 #          str(kmm)
           if (class(kmm) != "try-error") break
 #         cat(k," ",i,"\n")

### 2015-06-11 lib/fpc/R/classify.R

@@ -1,0 +1,1 @@
+../../../code/classify.R

### 2015-06-11 code/classify.R
whitespace

@@ -2,6 +2,7 @@
 #
 # Depends:
 #   library("MASS")
+#
 
 
 classify_lda <- function(x, grouping, prior=c("proportions", "equal"))
@@ -21,4 +22,3 @@ classify_lda <- function(x, grouping, prior=c("proportions", "equal"))
 
     MASS::lda(x, g, prior)
 }
-

### 2015-06-11 code/gabriel.R
add cluster selection code

@@ -1,0 +1,87 @@
+# code/gabriel.R
+#
+# Depends:
+#   library("bcv")
+#   library("MASS")
+#   source("classify.R")
+#   source("cluster.R")
+#
+
+
+cv.kmeans.gabriel <- function(x, krow = 5, kcol = 2, maxcenters = 10,
+                              classify.method = c("lda-equal", "lda-proportions", "svm"))
+{
+    classify.method <- match.arg(classify.method)
+    x <- as.matrix(x)
+    n <- nrow(x)
+    p <- ncol(x)
+    if (n < 2) 
+        stop("x should have at least two rows")
+    if (p < 2) 
+        stop("x should have at least two columns")
+    if ((krow > n) || (krow <= 1)) 
+        stop("krow outside allowable range")
+    if ((kcol > p) || (kcol <= 1)) 
+        stop("kcol outside allowable range")
+    if (maxcenters <= 0) 
+        stop("maxcenters should be positive")
+
+    krow.o <- krow
+    krow <- bcv:::round.fold(n, krow)
+    kcol.o <- kcol
+    kcol <- bcv:::round.fold(p, kcol)
+    if (krow != krow.o) 
+        warning("krow has been set to ", krow)
+    if (kcol != kcol.o) 
+        warning("kcol has been set to ", kcol)
+
+    s.r <- bcv:::choose.sets(n, krow)
+    s.c <- bcv:::choose.sets(p, kcol)
+    n0 <- n - max(table(s.r))
+    p0 <- p - max(table(s.c))
+    maxcenters.o <- maxcenters
+    maxcenters <- min(n0, round(maxcenters.o))
+    if (!missing(maxcenters) && maxcenters != maxcenters.o) 
+        warning("maxcenters has been set to ", maxcenters)
+
+    if (classify.method == "lda-equal") {
+        classify <- function(x, grouping)
+            classify_lda(x, grouping, prior="equal")
+    } else if (classify.method == "lda-proportions") {
+        classify <- function(x, grouping)
+            classify_lda(x, grouping, prior="proportions")
+    } else if (classify.method == "svm") {
+        stop("not yet implemented")
+    }
+
+    msep <- matrix(NA, krow * kcol, maxcenters)
+
+    for (k in seq_len(krow)) {
+        for (l in seq_len(kcol)) {
+            test <- s.r == k
+            train <- !test
+            response <- s.c == l
+            predictor <- !response
+
+            for (centers in seq_len(maxcenters)) {
+                if (centers == 1) {
[…]

### 2015-06-11 demo/bench/03_select_nclusters.R

@@ -1,0 +1,53 @@
+#!/usr/bin/Rscript --vanilla
+
+library("bcv")
+library("cluster")
+library("devtools")
+library("e1071")
+library("mclust")
+library("MASS")
+load_all("../../lib/fpc")
+load_all("../../lib/NbClust")
+source("../../code/classify.R")
+source("../../code/cluster.R")
+source("../../code/gabriel.R")
+source("../../code/jump.R")
+source("../../code/wold.R")
+source("methods.R")
+
+
+kmax <- 10
+
+for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
+    f.replicates <- file.path(s, "replicates.rds")
+    if (!file.exists(f.replicates))
+        next
+
+    replicates <- readRDS(f.replicates)
+    nrep <- length(replicates)
+
+    if (!dir.exists(file.path(s, "method")))
+        dir.create(file.path(s, "method"))
+
+    for (m in names(methods)) {
+        f.method <- file.path(s, "method", paste0(m, ".rds"))
+
+        if (!file.exists(f.method)) {
+            cat("applying method '", m, "' to '", s, "'\n", sep="")
+
+            set.seed(5453546) # from random.org
+            nclusters <- integer(nrep)
+
+            pb <- txtProgressBar(0, nrep, style=3)
+            for (r in seq_along(replicates)) {
+                x <- replicates[[r]]$x
+                nclusters[r] <- methods[[m]](x, kmax)
+                setTxtProgressBar(pb, r)
+            }
+            close(pb)
+
+            nclusters <- as.integer(nclusters)
+            saveRDS(nclusters, f.method)
+        }
+    }
+}

### 2015-06-11 demo/bench/methods.R

@@ -1,0 +1,61 @@
+# demo/bench/methods.R
+#
+# Depends:
+#   library("bcv")
+#   library("devtools")
+#   library("cluster")
+#   library("e1071")
+#   library("mclust")
+#   library("MASS")
+#   load_all("../../lib/fpc")
+#   load_all("../../lib/NbClust")
+#   source("../../lib/NbClust.R")
+#   source("../../code/classify.R")
+#   source("../../code/cluster.R")
+#   source("../../code/gabriel.R")
+#   source("../../code/jump.R")
+#   source("../../code/wold.R")
+#   
+
+methods <- list(
+    "gabriel-lda-equal" = function(x, maxcenters) {
+        cv <- cv.kmeans.gabriel(x, 2, 2, maxcenters,
+                                classify.method="lda-equal")
+        cv$centers
+    },
+    "wold" = function(x, maxcenters) {
+        Wold_holdout(x, 5, max.k=maxcenters, Errortol=0.01)
+    },
+    "gap" = function(x, maxcenters) {
+        Gap <- cluster::clusGap(x, FUN = cluster_kmeans, K.max = maxcenters)
+        which.max(Gap[[1]][,3])
+    },
+    "BIC" = function(x, maxcenters) {
+        mcluster <- mclust::Mclust(x, G = 1:maxcenters)
+        mcluster$G
+    },
+    "CH" = function(x, maxcenters) {
+        Ch <- NbClust::NbClust(x, min.nc = 2, max.nc = maxcenters,
+                               method = "kmeans", index = "ch")
+        Ch$Best.nc[[1]]
+    },
+    "Hartigan" = function(x, maxcenters) {
+        Hartigan <- NbClust::NbClust(x, min.nc = 2, max.nc = maxcenters,
+                                     method = "kmeans", index = "hartigan")
+        Hartigan$Best.nc[[1]]
+    },
+    "Jump" = function(x, maxcenters) {
+        Jump <- jump(x, plotjumps=FALSE, trace=FALSE)
+        Jump$maxjump
+    },
+    "PS" = function(x, maxcenters) {
+        PS <- fpc::prediction.strength(x, Gmin=2, Gmax=maxcenters,
+                                       clustermethod=kmeansCBI)
+        PS$optimalk
+    },
+    "SB" = function(x, maxcenters) {
+        SB <- fpc::nselectboot(x, clustermethod=kmeansCBI,
+                               classification="centroid",
+                               krange=2:maxcenters)
+        SB$kopt
+    })

### 2015-06-11 demo/bench/04_tabulate_results.R
start 'tabulate_results' step

@@ -1,0 +1,15 @@
+#!/usr/bin/Rscript --vanilla
+
+
+for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
+    replicates <- readRDS(file.path(s, "replicates.rds"))
+    kmeans <- readRDS(file.path(s, "kmeans.rds"))
+
+    nclusters <- list()
+    for (filename in list.files(file.path(s, "method"), "[.]rds$")) {
+        m <- substr(filename, 1, nchar(filename) - 4)
+        nclusters[[m]] <- readRDS(file.path(s, "method", filename))
+    }
+}
+    
+

### 2015-06-11 demo/bench/methods.R
rename method 'SB' to 'Stab'

@@ -53,7 +53,7 @@ methods <- list(
                                        clustermethod=kmeansCBI)
         PS$optimalk
     },
-    "SB" = function(x, maxcenters) {
+    "Stab" = function(x, maxcenters) {
         SB <- fpc::nselectboot(x, clustermethod=kmeansCBI,
                                classification="centroid",
                                krange=2:maxcenters)

### 2015-06-11 demo/bench/04_tabulate_results.R
Finish bench/04_tabulate_results

@@ -1,5 +1,76 @@
 #!/usr/bin/Rscript --vanilla
 
+setting.names <- list(
+    "setting1" = "Setting 1",
+    "setting2" = "Setting 2",
+    "setting3" = "Setting 3",
+    "setting4" = "Setting 4",
+    "setting5" = "Setting 5",
+    "setting6" = "Setting 6")
+
+nclust <- list(
+    "setting1" = 1L,
+    "setting2" = 2L,
+    "setting3" = 3L,
+    "setting4" = 4L,
+    "setting5" = 10L,
+    "setting6" = 4L)
+
+method.names <- list(
+    "oracle" = "Oracle",
+    "gap" = "Gap",
+    "BIC" = "Gaussian-Mix",
+    "CH" = "CH",
+    "Hartigan" = "Hartigan",
+    "Jump" = "Jump",
+    "PS" = "Prediction Strength",
+    "Stab" = "Stability",
+    "gabriel-lda-equal" = "Gabriel CV",
+    "wold" = "Wold CV")
+
+exclude <- function(setting, method)
+{
+    if (method %in% c("CH", "Hartigan", "Stab")) {
+        if (setting %in% c("setting1")) {
+            return(TRUE)
+        }
+    }
+    FALSE
+}
+
+print_table <- function(setting, summary)
+{
+    cat("\\textit{", setting.names[[setting]], "} & \\\\\n", sep="")
+
+    kmax <- length(summary[["oracle"]]$counts)
+
+    for (m in names(method.names)) {
+        s <- summary[[m]]
+        if (is.null(s))
+            next
+        cat(method.names[[m]])
+        for (k in seq_len(kmax)) {
+            cat(" & ")
+            if (exclude(setting, m)) {
+                cat("\\textendash")
+            } else if (k == nclust[[setting]]) {
+                cat("\\textbf{", s$counts[k], "}", sep="")
+            } else {
+                cat(s$counts[k])
+            }
+        }
+        if (m == "oracle") {
+            cat(" & 1")
+        } else if (exclude(setting, m)) {
+            cat(" & \\textendash")
+        } else {
+            cat(sprintf(" & %.1f $\\pm$ %.1f",
+                        s$prederr_rel$mean,
+                        s$prederr_rel$sd))
+        }
+        cat(" \\\\\n")
+    }
+}
 
 for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
[…]

### 2015-06-11 demo/bench/methods.R
add namespace qualifier to fpc::kmeansCBI

@@ -50,11 +50,11 @@ methods <- list(
     },
     "PS" = function(x, maxcenters) {
         PS <- fpc::prediction.strength(x, Gmin=2, Gmax=maxcenters,
-                                       clustermethod=kmeansCBI)
+                                       clustermethod=fpc::kmeansCBI)
         PS$optimalk
     },
     "Stab" = function(x, maxcenters) {
-        SB <- fpc::nselectboot(x, clustermethod=kmeansCBI,
+        SB <- fpc::nselectboot(x, clustermethod=fpc::kmeansCBI,
                                classification="centroid",
                                krange=2:maxcenters)
         SB$kopt

### 2015-06-11 demo/bench/03_select_nclusters.R
add parallel execution to selection_nclusters

@@ -6,8 +6,9 @@ library("devtools")
 library("e1071")
 library("mclust")
 library("MASS")
-load_all("../../lib/fpc")
-load_all("../../lib/NbClust")
+library("parallel")
+load_all("../../lib/fpc", export_all=FALSE)
+load_all("../../lib/NbClust", export_all=FALSE)
 source("../../code/classify.R")
 source("../../code/cluster.R")
 source("../../code/gabriel.R")
@@ -15,6 +16,8 @@ source("../../code/jump.R")
 source("../../code/wold.R")
 source("methods.R")
 
+
+options(cl.cores=2)
 
 kmax <- 10
 
@@ -36,15 +39,25 @@ for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
             cat("applying method '", m, "' to '", s, "'\n", sep="")
 
             set.seed(5453546) # from random.org
+            method <- methods[[m]]
             nclusters <- integer(nrep)
 
-            pb <- txtProgressBar(0, nrep, style=3)
-            for (r in seq_along(replicates)) {
-                x <- replicates[[r]]$x
-                nclusters[r] <- methods[[m]](x, kmax)
-                setTxtProgressBar(pb, r)
-            }
-            close(pb)
+            cl <- makeCluster(getOption("cl.cores", 2))
+            clusterExport(cl=cl, varlist=c("replicates", "method", "kmax",
+                                           "classify_lda",
+                                           "cluster_kmeans",
+                                           "compute.jump",
+                                           "cv.kmeans.gabriel",
+                                           "Impute",
+                                           "jump",
+                                           "kmeans.rndstart",
+                                           "mclustBIC",
+                                           "WoldIter",
+                                           "Wold_holdout"))
+            nclusters <- parSapply(cl, seq_len(nrep), function(r) {
+                    x <- replicates[[r]]$x
+                    method(x, kmax)
+                })
 
             nclusters <- as.integer(nclusters)
             saveRDS(nclusters, f.method)

### 2015-06-11 demo/bench/settings.R
fix setting definitions to max the article

@@ -73,49 +73,6 @@ setting2 <- function()
 }
 
 
-
-Data3center <- function(number = 40)
-{
-     Centers <- mvrnorm(3, rep(0,20), 19*diag(20))
-     u1 <- Centers[1,]
-     u2 <- Centers[2,]
-     u3 <- Centers[3,]
-
-     condition <- TRUE
-     while(condition){
-	      obs1 <- (matrix(u1,nrow=number,ncol=20,byrow=T)
-                   + matrix(rexp(number*20, rate = 1),ncol=20)-1)
-		U1 <- matrix(u1,nrow=number,ncol=20,byrow=T)
-		obs2 <- (matrix(u2,nrow=number,ncol=20,byrow=T)
-                 + matrix(rexp(number*20, rate = 1/2),ncol=20)-2)
-		U2 <- matrix(u2,nrow=number,ncol=20,byrow=T)
-		obs3 <- (matrix(u3,nrow=number,ncol=20,byrow=T)
-                 + matrix(rexp(number*20, rate = 1/5),ncol=20)-5)
-		U3 <-  matrix(u3,nrow=number,ncol=20,byrow=T)
-
-		if(sum(location(obs1,Centers)==1)==number
-           & sum(location(obs2,Centers)==2)==number
-           & sum(location(obs3,Centers)==3)==number){
-			condition <- FALSE
-		}
-	}
-	Data <- rbind(obs1,obs2,obs3)
-	Mu <- rbind(U1,U2,U3)
-	ID <- sample(1:nrow(Data),nrow(Data),replace = FALSE)
-	Data <- Data[ID,]
-
-	Mean <- Mu[ID,]
-	list(x = Data, mean = Mean)
-}
-
-
-setting3 <- function()
-{
-    Data3center(40)
-}
-
-
-
 Data4center <- function(di = 10, dd = 1.9){ #Another pair is di=4 and dd=5
 	condition <- TRUE
 	while(condition){
@@ -168,7 +125,7 @@ Data4center <- function(di = 10, dd = 1.9){ #Another pair is di=4 and dd=5
 }
 
 
-setting4 <- function()
+setting3 <- function()
 {
     Data4center(di = 100, dd = 0.65)
 }
@@ -239,12 +196,10 @@ Data10center <- function(di = 100, dd = 0.72){
 }
 
 
-setting5 <- function()
+setting4 <- function()
 {
[…]

### 2015-06-11 demo/bench/02_fit_kmeans.R
raise kmax from 10 to 15

@@ -2,7 +2,7 @@
 
 source("../../code/cluster.R") # cluster_kmeans
 
-kmax <- 10
+kmax <- 15
 
 
 for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {

### 2015-06-12 demo/overlap/01_run_sim.R
Directory for overlap case

@@ -1,0 +1,67 @@
+library(MASS)	
+
+LDA.pick <- function(U_x, U_y){
+
+	result <- rep(NA,10)
+	set.seed(1)
+
+	for(i in 1:10){
+		Data <- matrix(NA,20000,2)
+		ID <- sample(c(1,2),20000,replace = TRUE)
+		Data[which(ID==1),] <- mvrnorm(sum(ID==1),c(U_x, U_y), diag(2))
+		Data[which(ID==2),] <- mvrnorm(sum(ID==2),c(-U_x, -U_y), diag(2))
+
+		Xtest <- Data[10001:20000,1]
+		Ytest <- Data[10001:20000,2]
+		Xtrain <- Data[1:10000,1]
+		Ytrain <- Data[1:10000,2]
+		##----k=1-------------
+		Cond <- TRUE
+		while(Cond){
+			Kmean <- try(kmeans(Ytrain, 1,nstart = 100),silent=T)
+			if( class(Kmean)!="try-error" ){Cond <- FALSE}
+		}
+		Predict.test <- Kmean$center[rep(1,10000),]
+		CV1 <- sum((Predict.test-Ytest)^2)/10000
+		##----k=2-------------
+		Cond <- TRUE
+		while(Cond){
+			Kmean2 <- try(kmeans(Ytrain, 2,nstart = 100),silent=T)
+			if( class(Kmean)!="try-error" ){Cond <- FALSE}
+		}
+		Center2 = Kmean2$cluster
+		LDA2 <- lda(as.matrix(Xtrain),factor(Center2),prior=c(0.5,0.5))
+		Predict.test2 <- Kmean2$center[predict(LDA2,as.matrix(Xtest))$class,]
+		CV2 <- sum((Predict.test2-Ytest)^2)/10000
+		##------------------------------
+		RT <- c(CV1,CV2)
+		result[i]<-	which(RT == min(RT))
+	}
+	return(result)
+}
+
+##-----------------------------------------------------------
+##------------------------------------------------------------
+Raw <- matrix(NA,31*31*10,5)
+Raw <- as.data.frame(Raw)
+names(Raw) <-c("replicate", "U_x", "U_y", "k", "Group")
+Raw$replicate <- rep(1:10,31*31)
+Raw$Group <- rep(1:(31*31),each=10)
+
+Seq_x <- seq(0,3,0.1)
+Seq_y <- seq(0,3,0.1)
+
+G = 0
+for(X in Seq_x){
[…]

### 2015-06-12 demo/bench/04_tabulate_results.R
update correct #clusters for settings 3-6

@@ -11,10 +11,10 @@ setting.names <- list(
 nclust <- list(
     "setting1" = 1L,
     "setting2" = 2L,
-    "setting3" = 3L,
-    "setting4" = 4L,
-    "setting5" = 10L,
-    "setting6" = 4L)
+    "setting3" = 4L,
+    "setting4" = 10L,
+    "setting5" = 4L,
+    "setting6" = 3L)
 
 method.names <- list(
     "oracle" = "Oracle",

### 2015-06-15 demo/bench/03_select_nclusters.R
set kmax to 15, add call to stopCluster

@@ -18,8 +18,9 @@ source("methods.R")
 
 
 options(cl.cores=2)
+cl <- makeCluster(getOption("cl.cores", 2))
 
-kmax <- 10
+kmax <- 15
 
 for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
     f.replicates <- file.path(s, "replicates.rds")
@@ -42,7 +43,6 @@ for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
             method <- methods[[m]]
             nclusters <- integer(nrep)
 
-            cl <- makeCluster(getOption("cl.cores", 2))
             clusterExport(cl=cl, varlist=c("replicates", "method", "kmax",
                                            "classify_lda",
                                            "cluster_kmeans",
@@ -64,3 +64,6 @@ for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
         }
     }
 }
+
+stopCluster(cl)
+

### 2015-06-15 demo/bench/methods.R
specify itmax for 'mclust' method to avoid infinite loop

@@ -31,7 +31,8 @@ methods <- list(
         which.max(Gap[[1]][,3])
     },
     "BIC" = function(x, maxcenters) {
-        mcluster <- mclust::Mclust(x, G = 1:maxcenters)
+        mcluster <- mclust::Mclust(x, G = 1:maxcenters,
+                                   control=mclust::emControl(itmax=c(10000L, 10000L)))
         mcluster$G
     },
     "CH" = function(x, maxcenters) {

### 2015-06-15 demo/bench/03_select_nclusters.R
detect cores automatically

@@ -17,7 +17,7 @@ source("../../code/wold.R")
 source("methods.R")
 
 
-options(cl.cores=2)
+options(cl.cores=max(1, detectCores() - 1))
 cl <- makeCluster(getOption("cl.cores", 2))
 
 kmax <- 15

### 2015-06-15 demo/bench/03_select_nclusters.R
handle errors in select_nclusters

@@ -56,7 +56,8 @@ for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
                                            "Wold_holdout"))
             nclusters <- parSapply(cl, seq_len(nrep), function(r) {
                     x <- replicates[[r]]$x
-                    method(x, kmax)
+                    tryCatch(method(x, kmax),
+                             error=function(e) NA_integer_)
                 })
 
             nclusters <- as.integer(nclusters)

### 2015-06-15 demo/bench/04_tabulate_results.R
handle missing values in '04_tabulate_results'

@@ -59,6 +59,12 @@ print_table <- function(setting, summary)
                 cat(s$counts[k])
             }
         }
+        if (exclude(setting, m)) {
+            cat("& \\textendash")
+        } else {
+            cat(" & ", s$count_na, sep="")
+        }
+
         if (m == "oracle") {
             cat(" & 1")
         } else if (exclude(setting, m)) {
@@ -111,15 +117,26 @@ for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
     {
         x <- numeric(nrep)
         for (r in seq_len(nrep)) {
-            x[r] <- prederr[[r]][nc[r]]
+            if (is.na(nc[r])) {
+                x[r] <- NA
+            } else {
+                x[r] <- prederr[[r]][nc[r]]
+            }
         }
-        list(mean = mean(x), sd = sd(x), se_mean = sd(x) / sqrt(nrep))
+        n <- sum(!is.na(x))
+        list(mean = mean(x, na.rm=TRUE),
+             sd = sd(x, na.rm=TRUE),
+             se_mean = sd(x, na.rm=TRUE) / sqrt(n))
     }
 
     summary <- list()
     for (m in names(nclusters)) {
+        count_na <- sum(is.na(nclusters[[m]]))
+        counts <- tabulate(nclusters[[m]], kmax)
         summary[[m]] <- list(
-            counts = tabulate(nclusters[[m]], kmax),
+            count_na = count_na,
+            counts = counts,
+            proportions = counts / (length(nclusters[[m]]) - count_na),
             prederr = summarize(prederr, nclusters[[m]]),
             prederr_rel = summarize(prederr_rel, nclusters[[m]]))
     }

### 2015-06-15 demo/bench/methods.R
fix handling of maxcenters for jump method

@@ -46,7 +46,7 @@ methods <- list(
         Hartigan$Best.nc[[1]]
     },
     "Jump" = function(x, maxcenters) {
-        Jump <- jump(x, plotjumps=FALSE, trace=FALSE)
+        Jump <- jump(x, maxcenters, plotjumps=FALSE, trace=FALSE)
         Jump$maxjump
     },
     "PS" = function(x, maxcenters) {

### 2015-06-15 demo/bench/04_compute_prederr.R
insert 'compute_prederr' step

@@ -1,0 +1,25 @@
+#!/usr/bin/Rscript --vanilla
+
+for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
+    if (file.exists(file.path(s, "prederr.rds")))
+        next
+
+    replicates <- readRDS(file.path(s, "replicates.rds"))
+    kmeans <- readRDS(file.path(s, "kmeans.rds"))
+
+    nrep <- length(replicates)
+    prederr <- vector("list", nrep)
+
+    for (r in seq_len(nrep)) {
+        truth <- replicates[[r]]$mean
+        kmax <- length(kmeans[[r]])
+        prederr[[r]] <- rep(NA, kmax)
+        for(k in seq_len(kmax)) {
+            km <- kmeans[[r]][[k]]
+            pred <- km$centers[km$cluster,]
+            prederr[[r]][[k]] <- mean((truth - pred)^2)
+        }
+    }
+
+    saveRDS(prederr, file.path(s, "prederr.rds"))
+}

### 2015-06-15 demo/bench/05_tabulate_results.R

@@ -80,27 +80,16 @@ print_table <- function(setting, summary)
 
 for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
     replicates <- readRDS(file.path(s, "replicates.rds"))
+    nrep <- length(replicates)
+
     kmeans <- readRDS(file.path(s, "kmeans.rds"))
+    kmax <- length(kmeans[[1]])
+    prederr <- readRDS(file.path(s, "prederr.rds"))
 
     nclusters <- list()
     for (filename in list.files(file.path(s, "method"), "[.]rds$")) {
         m <- substr(filename, 1, nchar(filename) - 4)
         nclusters[[m]] <- readRDS(file.path(s, "method", filename))
-    }
-
-    # compute prediction error
-    nrep <- length(replicates)
-    prederr <- vector("list", nrep)
-
-    for (r in seq_len(nrep)) {
-        truth <- replicates[[r]]$mean
-        kmax <- length(kmeans[[r]])
-        prederr[[r]] <- rep(NA, kmax)
-        for(k in seq_len(kmax)) {
-            km <- kmeans[[r]][[k]]
-            pred <- km$centers[km$cluster,]
-            prederr[[r]][[k]] <- mean((truth - pred)^2)
-        }
     }
 
     # compute oracle, relative prediction error

### 2015-06-16 lib/NbClust/R/cluster.R
change cluster.R from symlink to regular file

@@ -1,1 +1,26 @@
-../../../code/cluster.R
+# code/cluster.R
+
+
+# default seed chosen by random.org
+cluster_kmeans <- function(x, centers, seed = 2651513, iter.max = 100,
+                           nstart = 100, ...)
+{
+    if (!is.na(seed)) {
+        if ((exists0 <- exists(".Random.seed", envir=globalenv()))) {
+            seed0 <- .Random.seed
+        }
+        set.seed(seed)
+    }
+
+    cl <- stats::kmeans(x, centers, iter.max=iter.max, nstart=nstart, ...)
+
+    if (!is.na(seed)) {
+        if (exists0) {
+            assign(".Random.seed", seed0, envir=globalenv())
+        } else if (exists(".Random.seed", envir=globalenv())) {
+            rm(".Random.seed", envir=globalenv())
+        }
+    }
+
+    cl
+}

### 2015-06-16 lib/fpc/R/classify.R
remove classify.R from fpc; copy cluster.R

@@ -1,1 +1,0 @@
-../../../code/classify.R

### 2015-06-16 cvclust.tex
add simulation results to main doc (needs reformatting)
start generalization of Pollard's thm

@@
 \newcommand{\bmuY}{\bar \mu^{Y}}
+\newcommand{\sE}{\mathcal{E}}
+\newcommand{\sX}{\mathcal{X}}
 
@@
 \section{Theoretical justification of Gabriel CV method}
+
 This section gives the self-consistency proof of the proposed Gabriel method.
@@
 \end{itemize}
+
+\subsection{Single Cluster With Noise}
+
+Here, we show that Gabriel Cross-Validation is consistent when the data come
+from a single cluster; the algorithm picks $k = 1$ with probability tending to
+$1$ as the number of rows in the data matrix $\dataX$ increase.
+
+Our proof
+relies on results from \citet{pollard1981strong}, who analyzed $k$-means and
+showed that the within-cluster sum of squares converges to a population
+quantity.  Specifically, for any finite set $A \subset \mathbb{R}^d$ and
+probability measure $P$ supported on $\mathbb{R}^d$, define the population
+within-cluster sum of squares as
+\[
+  W(A, P) = \int \min_{a \in A} \|x - a\|^2 P(dx).
+\]
+Set
+\[
+  m_k(P) = \inf\{ W(A, P) : \text{$A$ contains $k$ or fewer points} \}.
+\]
+Suppose that $X_1, \dotsc, X_n$ are independent draws from $P$, and
+if $P_n$ is the empirical measure formed by placing mass $1/n$ on each $X_i$.
+Let $A_n$ be a set minimizing $W(A_n, P_n)$.  Pollard gave conditions for
+$A_n$ to converge to the minimizer of $W(\cdot, P)$.  His main theorem assumes that
+the minimizer of $W(\cdot, P)$ is unique, which is too strong for our purposes
+(this fails, for example, with a bivariate normal distribution when $k \geq 2$).
+It is straightforward to generalize his result to cases with multiple
+maximizers.  To do so, we will need the following three lemmas.  The proofs of
+these lemmas are given in Sections~3 and~4 of Pollard's article.
+
+\begin{lemma}\label{lem:pollard-compact}\label{lem:pollard-first}
+Suppose that $\int \|x\|^2 P(dx) < \infty$ and that
+$m_j(P) > m_{j+1}(P)$ for $j = 1, \dotsc, k-1$.  If $A_n$ minimizes
+$W(\cdot, P_n)$, then, there exists a finite
+$M$ such that as if $n$ is large enough, then, almost surely, $A_n$ is in
+the class of sets $\sE_k = \{ A \subset B(5M) : \text{$A$ contains $k$ or fewer
+points} \}$, where $B(5M)$ denotes the Euclidean ball of radius $5M$ centered
+at the origin.
+\end{lemma}
+\begin{proof}
+Follows from Pollard's argument in Section 3 of his paper.
+\end{proof}
+
+\begin{lemma}\label{lem:polalrd-uniform}
+If $\int \|x\|^2 P(x) < \infty$, then
+\(
+  \sup_{A \in \sE_k} |W(A, P) - W(A,P_n)| \to 0\text{ a.s.}
+\)
+\end{lemma}
+\begin{proof}
+Given at the beginning of Pollard's Section~4.
+\end{proof}
+
+\begin{lemma}\label{lem:pollard-continuous}\label{lem:pollard-last}
+If $\int \|x\|^2 P(x) < \infty$, then
+the map $A \mapsto W(A, P)$ is continuous on $\sE_k$ with respect to the
+topology induced by the Hausdoff metric $H(\cdot, \cdot)$, defined for compact
+subsets $A, B$ of $\mathbb{R}^d$ by $H(A,B) < \delta$ if and only if every
+point of $A$ is within Euclidean distance $\delta$ of at least one point of
+$B$, and vice versa.
+\end{lemma}
+\begin{proof}
+Given at the end of Pollard's Section~4.
+\end{proof}
+
+Now, define $\mathcal{\bar A} = \mathcal{\bar A}(k)$ to be the class of sets
+$\bar A$ with $k$ or fewer points such that $W(\bar A,P) = m_k(P)$.
+Lemmas~\ref{lem:pollard-compact} and~\ref{lem:pollard-continuous} together
+with the extreme value theorem imply that this class is nonempty.
+
+To prove
+this, it suffices to analyze a single fold of the cross-validation procedure.
+We will assume the following setup:
+
+Observations $\{ (X_i, Y_i) \}_{i=1}^{n+N}$ are independent and identically
+distributed.  The distribution for each $(X_i, Y_i)$ pair is such that $X_i$
+and $Y_i$ have marginal distribution $P^X$ and $P^Y$, and the vectors
+$X_i$ and $Y_i$ are independent.
+
 \subsection{Well-separated clusters with noise}
-Same with previous, the data set $\mathbf{X} \in \mathbb{R}^{n \times p}$ been partitioned as $\mathbf{X}^L \in \mathbb{R}^{n \times L}$ and $\mathbf{X}^R \in \mathbb{R}^{n \times R}$, with following set of assumptions:
+
+Same with previous, the data set $\mathbf{X} \in \mathbb{R}^{n \times p}$ been
+partitioned as $\mathbf{X}^L \in \mathbb{R}^{n \times L}$ and $\mathbf{X}^R
+\in \mathbb{R}^{n \times R}$, with following set of assumptions:
+
 \begin{enumerate}[label={Assumption},leftmargin=*]
@@
 \[ K_{oracle} =  \underset{k}{\operatorname{argmin}}\hspace{0.05in} ||\hat{\mathbf{X}}_k- \boldsymbol{\mu} ||_F \]
+
+\begin{table}
+\centering
+\caption{ \label{table1}Simulation results} 
+{\scriptsize
+\begin{tabular}{lcccccccccccccccccc}
+\hline
+Number of Clusters & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA & PE \\ \hline
+\input{demo/bench/setting1/results.tex}
+\hline
+\input{demo/bench/setting2/results.tex}
+\hline
+\input{demo/bench/setting3/results.tex}
+\hline
+\input{demo/bench/setting4/results.tex}
+\hline
+\input{demo/bench/setting5/results.tex}
+\hline
+\input{demo/bench/setting6/results.tex}
+\hline
+\end{tabular}
+}\\
+\hspace{-2.0in} \footnotesize {In Setting~1, ``\textendash'' means the method can not be used with parameter $k=1$}.
+\end{table}
+
 \begin{table}[H]
 \centering
-\caption{ \label{table1}Simulation results} 
+\caption{ \label{table1}Old Simulation results}
 \scalebox{0.97}{

### 2015-06-17 demo/bench/methods.R
Revised setting 2 in simulation and multiple CV results with Gabriel

@@ -23,6 +23,21 @@ methods <- list(
                                 classify.method="lda-equal")
         cv$centers
     },
+    "gabriel-lda-equal-5X2" = function(x, maxcenters) {
+        cv <- cv.kmeans.gabriel(x, 5, 2, maxcenters,
+                              classify.method="lda-equal")
+        cv$centers
+    },
+    "gabriel-lda-proportion" = function(x, maxcenters) {
+        cv <- cv.kmeans.gabriel(x, 2, 2, maxcenters,
+                                classify.method="lda-proportions")
+        cv$centers
+    },
+    "gabriel-lda-proportion-5X2" = function(x, maxcenters) {
+        cv <- cv.kmeans.gabriel(x, 5, 2, maxcenters,
+                              classify.method="lda-proportions")
+        cv$centers
+    },
     "wold" = function(x, maxcenters) {
         Wold_holdout(x, 5, max.k=maxcenters, Errortol=0.01)
     },

### 2015-06-17 demo/bench/settings.R

@@ -41,17 +41,17 @@ location <- function(M.data,M.center)
 
 Data2center <- function(number = 50)
 {
-	u1 = c(1,0.0,0.0,1)
+	u1 = c(-1,0.0,0.0,-1)##changed
 	u2 = c(1,3.5,3.5,1)
       Centers <- matrix(c(1,1,0,3.5,0,3.5,1,1),2,4)
-      #COV <- matrix(c(1,0.2,0.2^2,0.2^3,0.2,1,0.2,0.2^2,0.2^2,0.2,1,0.2,0.2^3,0.2^2,0.2,1),4,4)
+      COV <- matrix(c(1,-0.2,(-0.2)^2,(-0.2)^3,-0.2,1,-0.2,(-0.2)^2,(-0.2)^2,-0.2,1,-0.2,(-0.2)^3,(-0.2)^2,-0.2,1),4,4)
 
 	condition <- TRUE
 	while(condition){
-	      obs1 <- mvrnorm(n = number, u1, diag(4)) #replace by #(1)# 
-		obs2 <- mvrnorm(n = number, u2, diag(4)) #replace by #(2)# 
-		#(1)# obs1 <- mvrnorm(n = number, u1, COV)
-		#(2)# obs2 <- mvrnorm(n = number, u2, COV)
+	      #obs1 <- mvrnorm(n = number, u1, diag(4)) #replace by #(1)# 
+		#obs2 <- mvrnorm(n = number, u2, diag(4)) #replace by #(2)# 
+		obs1 <- mvrnorm(n = number, u1, 0.5*COV)
+		obs2 <- mvrnorm(n = number, u2, 1.5*COV)
 
 		if(sum(location(obs1,Centers)==1)==number
            & sum(location(obs2,Centers)==2)==number){

### 2015-06-17 cvclust.tex
add proof of continuity

@@
 \newcommand{\bmuY}{\bar \mu^{Y}}
+\newcommand{\sA}{\mathcal{A}}
 \newcommand{\sE}{\mathcal{E}}
@@
 
-Our proof
-relies on results from \citet{pollard1981strong}, who analyzed $k$-means and
-showed that the within-cluster sum of squares converges to a population
-quantity.  Specifically, for any finite set $A \subset \mathbb{R}^d$ and
-probability measure $P$ supported on $\mathbb{R}^d$, define the population
-within-cluster sum of squares as
+Our proof relies on results from \citet{pollard1981strong}, who analyzed
+$k$-means and showed that the within-cluster sum of squares converges to a
+population quantity.  Specifically, for any finite set $A \subset
+\mathbb{R}^d$ and probability measure $P$ supported on $\mathbb{R}^d$, define
+the population within-cluster sum of squares as
 \[
@@
 \]
-Set
+Define
+\(
+  \sA_k
+    =
+    \{ A \subset \mathbb{R}^d : \text{$A$ contains $k$ or fewer points} \}
+\)
+and set
+\(
+  m_k(P) = \inf_{A \in \sA_k}  W(A, P). % : \text{$A$ contains $k$ or fewer points} \}.
+\)
+Suppose that $X_1, \dotsc, X_n$ are independent draws from $P$, and if $P_n$
+is the empirical measure formed by placing mass $1/n$ on each $X_i$.  Let
+$A_n$ be a set minimizing $W(A_n, P_n)$.  Pollard gave conditions for $A_n$ to
+converge to the minimizer of $W(\cdot, P)$.  His main theorem assumes that the
+minimizer of $W(\cdot, P)$ is unique, which is too strong for our purposes
+(this fails, for example, with a bivariate normal distribution with identity
+covariance whenever $k \geq 2$).  It is straightforward to generalize his
+result to cases with multiple maximizers.  To do so, we will need the
+following three lemmas.  The proofs of these lemmas are modifications of
+arguments given in Sections~3 and~4 of Pollard's article.
+
+\begin{lemma}\label{lem:pollard-continuous}\label{lem:pollard-last}
+If $\int \|x\|^2 P(x) < \infty$, then the map $A \mapsto W(A, P)$ is
+continuous on $\sA_k$ with respect to the topology induced by the Hausdoff
+metric $H(\cdot, \cdot)$, defined for compact subsets $A, B$ of $\mathbb{R}^d$
+by $H(A,B) < \delta$ if and only if every point of $A$ is within Euclidean
+distance $\delta$ of at least one point of $B$, and vice versa.
+\end{lemma}
+\begin{proof}
+This is a slight extension of the argument given at the end of Pollard's
+Section~4.  If $A, B \in \sA_k$ and $H(A, B) < \delta$, then for each $b \in
+B$ there exists an $a(b) \in A$ such that $\|b - a(b)\| < \delta$.  Then
+\begin{align*}
+  W(A, P) - W(B, P)
+    &= \int \min_{a \in A} \|x - a\|^2 - \min_{b \in B} \|x - b\|^2 P(dx) \\
+    &\leq \int \max_{b \in B} \{ \|x - a(b)\|^2 - \|x - b\|^2 \} P(dx) \\
+    &\leq \int \sum_{b \in B} \{ (\|x - b\| + \delta)^2 - \|x - b\|^2 \} P(dx) \\
+    &\leq \int \sum_{b \in B}  ( 2 \delta \|x - b\| + \delta^2) P(dx) \\
+    &\leq 2 \delta \{M^{1/2} + k \max_{b \in B} \|b\|\} + k \delta^2,
+\end{align*}
+where $M = \int \|x\|^2 P(dx)$.  Thus,
 \[
-  m_k(P) = \inf\{ W(A, P) : \text{$A$ contains $k$ or fewer points} \}.
+  |W(A, P) - W(B, P)|
+    \leq 2 \delta \{M^{1/2} + k \max_{a \in A} \|a\|\} + 3 k \delta^2.
 \]
-Suppose that $X_1, \dotsc, X_n$ are independent draws from $P$, and
-if $P_n$ is the empirical measure formed by placing mass $1/n$ on each $X_i$.
-Let $A_n$ be a set minimizing $W(A_n, P_n)$.  Pollard gave conditions for
-$A_n$ to converge to the minimizer of $W(\cdot, P)$.  His main theorem assumes that
-the minimizer of $W(\cdot, P)$ is unique, which is too strong for our purposes
-(this fails, for example, with a bivariate normal distribution when $k \geq 2$).
-It is straightforward to generalize his result to cases with multiple
-maximizers.  To do so, we will need the following three lemmas.  The proofs of
-these lemmas are given in Sections~3 and~4 of Pollard's article.
-
-\begin{lemma}\label{lem:pollard-compact}\label{lem:pollard-first}
-Suppose that $\int \|x\|^2 P(dx) < \infty$ and that
-$m_j(P) > m_{j+1}(P)$ for $j = 1, \dotsc, k-1$.  If $A_n$ minimizes
-$W(\cdot, P_n)$, then, there exists a finite
-$M$ such that as if $n$ is large enough, then, almost surely, $A_n$ is in
-the class of sets $\sE_k = \{ A \subset B(5M) : \text{$A$ contains $k$ or fewer
-points} \}$, where $B(5M)$ denotes the Euclidean ball of radius $5M$ centered
-at the origin.
+Hence if $A \in \sA_k$ is any fixed set, then for every $\varepsilon > 0$, we
+can find a $\delta$ small enough so that if $B \in \sA_k$ and $H(A, B) <
+\delta$, then $|W(A,P) - W(B,P)| < \epsilon$.  Thus, $W(\cdot, P)$ is
+continuous at $A$.
+\end{proof}
+
+\begin{lemma}
+\label{lem:pollard-compact}
+\label{lem:pollard-first}
+Suppose that $\int \|x\|^2 P(dx) < \infty$.  If $A_n$ minimizes $W(\cdot,
+P_n)$, then, there exists a finite $M$ such that as if $n$ is large enough,
+then, almost surely, $A_n$ is in the class of sets
+\(
+  \sE_k = \{ A \subset B(5M) : \text{$A$ contains $k$ or fewer points} \},
+\)
+where $B(5M)$ denotes the Euclidean ball of radius $5M$ centered at the
+origin.
 \end{lemma}
-\begin{proof}
-Follows from Pollard's argument in Section 3 of his paper.
+
+\begin{proof}
+If $P$ has finite support, then the result follows immediately.  For the
+remainder of the proof, then, we will assume that $P$ has infinite support.
+
+In Section~3 of his paper, Pollard proves the result, assuming that there
+is a unique set of $j$ points minimizing $W(\cdot, P)$ for each value of
+$j = 1, \dotsc, k$.  His proof does not use this assumption directly, only
+its implication that for each $j$ in the range $2, \dotsc, k$, there exists
+a set with $j$ points satisfying $W(A,P) < m_{j-1}(P)$.  If we can establish the
+implication, then the rest of his proof will apply.
+
+We will prove the implication by construction.  Take $j$ to be in the range
+$2, \dotsc, k$.  For any $\varepsilon > 0$, there exists a set
+$B_{\varepsilon}$ containing at most $j-1$ points such that
+$W(B_{\varepsilon}, P) < m_{j-1}(P) + \varepsilon$.  We will add a single
+point to $B_{\varepsilon}$ to get a set $A_{\varepsilon}$ with $j$ or fewer
+points satisfying $W(A, P) < m_{j-1}(P)$.
+
+If we can show that $m_{j+1}(P) > m_{j}(P)$ for $j = 1, \dotsc, k-1$, then
+we will have established the conditions required for Pollard's proof of the
+analogous result in Section 3 of his paper.
+
 \end{proof}
@@
 Given at the beginning of Pollard's Section~4.
-\end{proof}
-
-\begin{lemma}\label{lem:pollard-continuous}\label{lem:pollard-last}
-If $\int \|x\|^2 P(x) < \infty$, then
-the map $A \mapsto W(A, P)$ is continuous on $\sE_k$ with respect to the
-topology induced by the Hausdoff metric $H(\cdot, \cdot)$, defined for compact
-subsets $A, B$ of $\mathbb{R}^d$ by $H(A,B) < \delta$ if and only if every
-point of $A$ is within Euclidean distance $\delta$ of at least one point of
-$B$, and vice versa.
-\end{lemma}
-\begin{proof}
-Given at the end of Pollard's Section~4.
 \end{proof}

### 2015-06-18 cvclust.tex
more progress on proof

@@
 \newcommand{\sA}{\mathcal{A}}
+\newcommand{\sC}{\mathcal{C}}
 \newcommand{\sE}{\mathcal{E}}
+\newcommand{\sN}{\mathcal{N}}
 \newcommand{\sX}{\mathcal{X}}
@@
 continuous at $A$.
+\end{proof}
+
+\begin{lemma}
+
+If $\int \|x\|^2 P(dx) < \infty$ and the support of $P$ contains at least $k$
+points, then $m_1(P) > m_2(P) > \dotsb > m_k(P)$.  Further, there exists a
+finite $M$ such that $W(\cdot, P)$ attains its infimum over $\sA_k$ and each
+minimizer $\bar A$ is contained in the set
+\(
+  \sE_k = \{ A \in \sA_k : A \subset B(5M) \},
+\)
+where $B(5M)$ denotes the closed ball of radius~$5M$.
+\end{lemma}
+\begin{proof}
+This argument is adapted from the proof in Section~3 of Pollard's paper.  We
+first find an $M$ large enough to ensure that every minimizer of $W(\cdot, P)$
+must have at least one point in $B(M)$.  Next, we impose additional
+constraints on $M$ to ensure that all minimizers of $W(\cdot, P)$ are
+contained in $B(5M)$.
+
+Pick~$r$ so that the ball~$K$ of radius~$r$ centered at the origin has
+positive $P$~measure.  Take $M$ to be large enough such that
+\(
+  (M - r)^2 P(K) > \int \|x\|^2 P(dx).
+\)
+Let $\sC_1$ denote the class of sets $A$ with at most $k$ points and at least
+one point in in $B(M)$.  Let $A_0$ be the singleton set containing the origin.
+If no point of $A$ is in $B(M)$, then
+\(
+  W(A, P) \geq (M - r)^2 P(K) > W(A_0, P).
+\)
+Put concisely, if $A \in \sA_k \setminus \sC_1$, then
+\(
+  W(A, P) > W(A_0, P) \geq m_k(P).
+\)
+
+Next, we proceed by induction on $k$.  If $k = 1$, then there is a unique
+minimizer of $W(\cdot, P)$ over $\sA_k$, the set containing the mean $\int x
+P(dx)$.  For $k > 1$, take as given that $m_{k-1}(P) = W(\bar A_{k-1}, P)$ for
+some set $\bar A_{k-1} \in \sE_{k-1}$.  We will show that there is an $M$
+large enough so that $W(\cdot, P)$ attains its infimum over $\sA_k$ and all
+minimizers lie in $\sE_k$.
+
+We first show that if $M$ is large enough, then there exists a set $A \in \sE_k$
+and a positive $\varepsilon$ with $\varepsilon + W(A,P) < m_{k-1}(P)$.
+Since the support of $P$ contains least $k$ points, there exists an $x_0$ not
+in $\bar A_{k-1}$ that lies in the support of $P$.  Ensure that $M$ is large
+enough so that $x_0 \in B(5M)$.  Take $A \in \sE_k$ to be the set of $k$
+points gotten by adjoining $x_0$ to $\bar A_{k-1}$.  Let
+\(
+  \delta = \min_{a \in \bar A_{k-1}} \| x_0 - a \|
+\)
+and set $R = \{ x \in \mathbb{R}^d : \|x - x_0\| < \delta/3 \}$; since $x_0$
+is in the support of $P$, the neighborhood $R$ has positive $P$~measure.
+Further,
+\[
+  \int_{R} \min_{a \in \bar A_{k-1}} \|x - a\|^2 P(dx)
+  \geq
+  \int_{R} \min_{a \in \bar A_{k-1}} (\|x_0 - a\| - \delta/3)^2 P(dx)
+  = (4/9) \delta^2 P(R),
+\]
+and
+\[
+  \int_{R} \min_{a \in A} \|x - a\|^2 P(dx)
+  =
+  \int_{R} \|x - x_0\|^2 P(dx)
+  \leq
+  (1/9) \delta^2 P(R).
+\]
+Thus, if
+\(
+  \varepsilon < (3/9) \delta^2 P(R)
+\)
+then
+\(
+  m_{k-1}(P) - W(A, P) = 
+  W(\bar A_{k-1}, P) - W(A, P) > \varepsilon.
+\)
+
 \end{proof}

### 2015-06-18 cvclust.tex
finish proof of kmeans existence

@@
 
-\begin{lemma}
+\begin{lemma}\label{lem:kmeans-existence}
 
@@
 \)
-Let $\sC_1$ denote the class of sets $A$ with at most $k$ points and at least
+Let $\sC_k$ denote the class of sets $A$ with at most $k$ points and at least
 one point in in $B(M)$.  Let $A_0$ be the singleton set containing the origin.
@@
 \)
-Put concisely, if $A \in \sA_k \setminus \sC_1$, then
+Put concisely, if $A \in \sA_k \setminus \sC_k$, then
 \(
@@
 
-We first show that if $M$ is large enough, then there exists a set $A \in \sE_k$
-and a positive $\varepsilon$ with $\varepsilon + W(A,P) < m_{k-1}(P)$.
-Since the support of $P$ contains least $k$ points, there exists an $x_0$ not
+We first show that if $M$ is large enough, then there exists a set $B \in \sE_k$
+and a positive $\varepsilon$ with $W(B,P) < m_{k-1}(P) - \varepsilon$.  This
+will establish that $m_{k}(P) < m_{k-1}(P)$, and the set $B$ will have further
+utility later in the proof.
+Take $\bar A_{k-1} \in \sE_{k-1}$ to be a set with at most $k-1$ points satisfying
+$m_{k-1}(P) = W(\bar A_{k-1},P)$.
+Since the support of $P$ contains least $k$ points, there exists a point~$b_1$ not
 in $\bar A_{k-1}$ that lies in the support of $P$.  Ensure that $M$ is large
-enough so that $x_0 \in B(5M)$.  Take $A \in \sE_k$ to be the set of $k$
-points gotten by adjoining $x_0$ to $\bar A_{k-1}$.  Let
+enough so that $b_1 \in B(5M)$.  Take $B \in \sE_k$ to be the set of $k$
+points gotten by adjoining $b_1$ to $\bar A_{k-1}$.  Let
 \(
-  \delta = \min_{a \in \bar A_{k-1}} \| x_0 - a \|
+  \delta = \min_{a \in \bar A_{k-1}} \| b_1 - a \|
 \)
-and set $R = \{ x \in \mathbb{R}^d : \|x - x_0\| < \delta/3 \}$; since $x_0$
+and set $R = \{ x \in \mathbb{R}^d : \|x - b_1\| < \delta/3 \}$; since $b_1$
 is in the support of $P$, the neighborhood $R$ has positive $P$~measure.
@@
   \geq
-  \int_{R} \min_{a \in \bar A_{k-1}} (\|x_0 - a\| - \delta/3)^2 P(dx)
+  \int_{R} \min_{a \in \bar A_{k-1}} (\|b_1 - a\| - \delta/3)^2 P(dx)
   = (4/9) \delta^2 P(R),
@@
 \[
-  \int_{R} \min_{a \in A} \|x - a\|^2 P(dx)
+  \int_{R} \min_{b \in B} \|x - b\|^2 P(dx)
   =
-  \int_{R} \|x - x_0\|^2 P(dx)
+  \int_{R} \|x - b_1\|^2 P(dx)
   \leq
@@
 \(
-  m_{k-1}(P) - W(A, P) = 
-  W(\bar A_{k-1}, P) - W(A, P) > \varepsilon.
+  W(B,P) \leq W(\bar A_{k-1},P) - (3/9) \delta^2 P(R)
+         < m_{k-1}(P) - \varepsilon.
 \)
+
+Now, ensure that $M$ large enough so that
+\[
+  4 \int_{\|x\| \geq 2M} \|x\|^2 P(dx) < \varepsilon.
+\]
+We will now show that if $A \in \sC_k \setminus \sE_k$, then
+$W(A,P) > W(B,P) \geq m_{k}(P)$, which ensures that $W(\cdot, P)$ does not
+attain its infimum over $\sA_k$ outside of $\sE_k$.  To see that this holds,
+suppose that $A$ is any set $\sC_k \setminus \sE_k$.  This set has at least
+one point, $a_1$, in $B(M)$, and at least one point outside $B(5M)$.  Consider
+thet set $A^\ast$ formed by removing all points of $A$ outside $B(5M)$.  Any
+point $x$ closer to a cluster center outisde $B(5M)$ than to $a_1$ must
+satisfy $\|x\| \geq 2M$.  Thus,
+\begin{align*}
+  W(A^\ast, P)
+    &\leq W(A,P) + \int_{\|x\| \geq 2M} \|x - a_1\|^2 P(dx) \\
+    &\leq W(A,P) + \int_{\|x\| \geq 2M} (\|x\| + \|a_1\|)^2 P(dx) \\
+    &\leq W(A,P) + 4 \int_{\|x\| \geq 2M} \|x\|^2 P(dx) \\
+    &< W(A,P) + \varepsilon.
+\end{align*}
+Since $A^\ast$ has $k-1$ or fewer points, it must satisfy
+$W(A^\ast, P) \geq m_{k-1}(P)$, so that
+$W(A, P) > m_{k-1}(P) - \varepsilon > W(B, P)$.
+
+We have now established that $m_k(P) = \inf \{ W(A,P) : A \in \sE_k \}$,
+and that $W(\cdot, P)$ does not obtain its infimum over $\sA_k$ outside of
+$\sE_k$.  The final step of the proof is to show $W(\cdot, P)$ obtains its
+infimum over $\sE_k$.  This follows from the extreme value theorem, since
+$\sE_k$ is compact and, by Lemma~\ref{lem:pollard-continuous}, the map $A
+\mapsto W(A,P)$ is continuous over $\sA_k$.
 

### 2015-06-18 cvclust.tex
finish k-means generalized consistency

@@
 \newcommand{\sA}{\mathcal{A}}
+\newcommand{\sbA}{\mathcal{\bar A}}
 \newcommand{\sC}{\mathcal{C}}
@@
 
-\begin{lemma}\label{lem:pollard-continuous}\label{lem:pollard-last}
+\begin{lemma}[Continuity]\label{lem:pollard-continuous}\label{lem:pollard-last}
 If $\int \|x\|^2 P(x) < \infty$, then the map $A \mapsto W(A, P)$ is
@@
 
-\begin{lemma}\label{lem:kmeans-existence}
-
+\begin{lemma}[Existence]\label{lem:kmeans-existence}
 If $\int \|x\|^2 P(dx) < \infty$ and the support of $P$ contains at least $k$
-points, then $m_1(P) > m_2(P) > \dotsb > m_k(P)$.  Further, there exists a
-finite $M$ such that $W(\cdot, P)$ attains its infimum over $\sA_k$ and each
-minimizer $\bar A$ is contained in the set
-\(
-  \sE_k = \{ A \in \sA_k : A \subset B(5M) \},
-\)
-where $B(5M)$ denotes the closed ball of radius~$5M$.
+points, then
+\begin{enumerate}
+  \item $m_1(P) > m_2(P) > \dotsb > m_k(P)$;
+  \item $W(\cdot, P)$ achieves its infimum over $\sA_j$ for $j = 1, \dotsc,
+    k$;
+  \item the class of sets $\sbA_j = \{ A \in \sA_j : W(A,P) = m_{j}(P) \}$
+    is compact (with respect to the Hausdorff metric on $\sA_j$) for $j = 1, \dotsc, k$.
+\end{enumerate}
 \end{lemma}
@@
 Let $\sC_k$ denote the class of sets $A$ with at most $k$ points and at least
-one point in in $B(M)$.  Let $A_0$ be the singleton set containing the origin.
+one point in in $B(M)$, the closed ball with radius $M$ centered at the origin.  Let
+$A_0$ be the singleton set containing the origin.
 If no point of $A$ is in $B(M)$, then
@@
 large enough so that $W(\cdot, P)$ attains its infimum over $\sA_k$ and all
-minimizers lie in $\sE_k$.
+minimizers lie in $\sE_k$.  Since $A \mapsto W(\cdot, P)$ is continuous and
+$\sE_k$ is compact, the class $\sbA_k$, the preimage of $\{ m_k(P) \}$ under this
+map, is compact.
 
@@
 \mapsto W(A,P)$ is continuous over $\sA_k$.
-
 \end{proof}
@@
 \label{lem:pollard-compact}
-\label{lem:pollard-first}
-Suppose that $\int \|x\|^2 P(dx) < \infty$.  If $A_n$ minimizes $W(\cdot,
-P_n)$, then, there exists a finite $M$ such that as if $n$ is large enough,
+Suppose that $\int \|x\|^2 P(dx) < \infty$, that the support of $P$ has at
+least $k$ points, and that $k > 1$.  Let $A_n$ minimize $W(\cdot, P_n)$ over
+$\sA_k$, and that $B_n$ minimize $W(\cdot P_n)$ over $\sA_{k-1}$.  If
+$W(B_n, P) \to m_{k-1}(P)$ almost surely as $n \to \infty$, then
+there exists a finite $M$ such that as if $n$ is large enough,
 then, almost surely, $A_n$ is in the class of sets
@@
 \begin{proof}
-If $P$ has finite support, then the result follows immediately.  For the
-remainder of the proof, then, we will assume that $P$ has infinite support.
-
-In Section~3 of his paper, Pollard proves the result, assuming that there
+Pollard proves this result, in Section~3 of his paper,  assuming that there
 is a unique set of $j$ points minimizing $W(\cdot, P)$ for each value of
-$j = 1, \dotsc, k$.  His proof does not use this assumption directly, only
-its implication that for each $j$ in the range $2, \dotsc, k$, there exists
-a set with $j$ points satisfying $W(A,P) < m_{j-1}(P)$.  If we can establish the
-implication, then the rest of his proof will apply.
-
-We will prove the implication by construction.  Take $j$ to be in the range
-$2, \dotsc, k$.  For any $\varepsilon > 0$, there exists a set
-$B_{\varepsilon}$ containing at most $j-1$ points such that
-$W(B_{\varepsilon}, P) < m_{j-1}(P) + \varepsilon$.  We will add a single
-point to $B_{\varepsilon}$ to get a set $A_{\varepsilon}$ with $j$ or fewer
-points satisfying $W(A, P) < m_{j-1}(P)$.
-
-If we can show that $m_{j+1}(P) > m_{j}(P)$ for $j = 1, \dotsc, k-1$, then
-we will have established the conditions required for Pollard's proof of the
-analogous result in Section 3 of his paper.
-
-\end{proof}
-
-\begin{lemma}\label{lem:polalrd-uniform}
+$j = 1, \dotsc, k$.  His proof, which is very similar to the proof of
+Lemma~\ref{lem:kmeans-existence}, does not use this assumption directly, only
+its implication that $m_{k}(P) < m_{k-1}(P)$.  Since we have established his
+result in Lemma~\ref{lem:kmeans-existence}, the remainder of his proof applies
+without further modification.
+\end{proof}
+
+\begin{lemma}[Uniform SLLN]\label{lem:pollard-uniform}
 If $\int \|x\|^2 P(x) < \infty$, then
@@
 \begin{proof}
-Given at the beginning of Pollard's Section~4.
+Pollard proves this result at the beginning of his Section~4.
+\end{proof}
+
+\begin{theorem}[$k$-Means Consistency]
+Suppose that $\int \|x\|^2 P(dx) < \infty$ and that $A_n$ minimizes
+$W(\cdot, P_n)$ over $\sA_k$.  Then for all $n$ there exists an
+$\bar A_n \in \sbA_k$ such that $H(A_n, \bar A_n) \to 0\text{ a.s.}$,
+and $W(A_n, P_n) \to m_k(P)\text{ a.s.}$
+\end{theorem}
+\begin{proof}
+The proof proceeds inductively over $k$.  For $k=1$, the result follows from
+the strong law of large numbers.  For $k > 1$, assume that the result holds
+for smaller values of $k$.  If the support of $P$ has fewer than $k$ points,
+then the result is trivially true.  Otherwise, the induction hypothesis
+implies that there exists a $B_n \in \sA_{k-1}$ with $W(B_n, P_n) \to
+m_{k-1}(P)\text{ a.s.}$  Hence, Lemma~\ref{lem:pollard-compact} ensures that
+there is a finite $M$ such that $A_n \in \sE_k$ almost surely for large
+enough~$n$.  Take $M$ to be large enough so that $\sE_k$ also contains $\sbA_k$.
+
+The set $\sE_k$ is compact and the map $f: \sE_k \to \mathbb{R}$ with $f(A) =
+W(A, P)$ is continuous.  Thus, for every $\varepsilon > 0$ there exists an
+$\eta > 0$ depending on $\varepsilon$ such that if $A \in \sE_k$ and $H(A,
+\bar A) \geq \varepsilon$ for every $\bar A \in \sbA_k$, then $W(A,P) \geq
+m_{k}(P) + \eta$.  Let $\bar A_0$ be any member of $\sbA_k$.  By construction,
+$W(A_n, P_n) \leq W(\bar A, P_n)$.  Lemma~\ref{lem:pollard-uniform} ensures
+that $W(\bar A_0, P_n) \to m_k(P)\text{ a.s.}$  Thus, for $n$ large enough,
+$W(A_n, P_n) < m_{k}(P) + \eta$, which implies that there exists an $\bar A_n
+\in \sbA_k$ with $H(A_n, \bar A_n) < \varepsilon$.
+
 \end{proof}

### 2015-06-19 cvclust.tex
reorg k-means proof

@@
 $A_n$ be a set minimizing $W(A_n, P_n)$.  Pollard gave conditions for $A_n$ to
-converge to the minimizer of $W(\cdot, P)$.  His main theorem assumes that the
+converge to a minimizer of $W(\cdot, P)$.  His main theorem assumes that the
 minimizer of $W(\cdot, P)$ is unique, which is too strong for our purposes
@@
 covariance whenever $k \geq 2$).  It is straightforward to generalize his
-result to cases with multiple maximizers.  To do so, we will need the
-following three lemmas.  The proofs of these lemmas are modifications of
-arguments given in Sections~3 and~4 of Pollard's article.
-
-\begin{lemma}[Continuity]\label{lem:pollard-continuous}\label{lem:pollard-last}
+result to cases with multiple maximizers.
+
+To state the consistency theorem define
+\[
+  \sbA_k = \{ \bar A \subseteq \mathbb{R}^d
+            : W(\bar A, P) = m_k(P)
+              \text{ and $\bar A$ contains $k$ or fewer points} \}.
+\]
+We will later show that $\sbA_k$ is nonempty and compact whenever
+$\int \|x\|^2 P(dx) < \infty$.
+Let $H(\cdot, \cdot)$ denote the Hausdorff metric, defined for compact subsets
+$A, B$ of $\mathbb{R}^d$ by $H(A,B) < \delta$ if and only if every point of
+$A$ is within Euclidean distance $\delta$ of at least one point of $B$, and
+vice versa.  We can now state the generalized version of Pollard's $k$-means
+consistency theorem.
+
+\begin{theorem}[$k$-Means Consistency]\label{lem:kmeans-consistency}
+Suppose that $\int \|x\|^2 P(dx) < \infty$ and that $A_n$ minimizes
+$W(\cdot, P_n)$ over $\sA_k$.  Then for all $n$ there exists an
+$\bar A_n \in \sbA_k$ such that $H(A_n, \bar A_n) \to 0\text{ a.s.}$,
+and $W(A_n, P_n) \to m_k(P)\text{ a.s.}$
+\end{theorem}
+
+The proof of this theorem relies on a series of lemmas, which are extensions
+of results given in Pollard's original article.  First, we show that the map
+$A \mapsto W(A, P)$ is continuous.  Next, we show that $\sbA_k$ is nonempty
+and compact.  Then, it follows that for $n$ large enough, all minimizers of
+$W(\cdot, A_n)$ are contained in a compact set $\sE_k$.  The final ingredient
+of the proof is Pollard's uniform strong law of large numbers, which says that
+$\sup_{A \in \sE_k} |W(A,P) - W(A,P_n)| \to 0\text{ a.s.}$
+
+\begin{lemma}[Continuity]\label{lem:kmeans-continuous}
 If $\int \|x\|^2 P(x) < \infty$, then the map $A \mapsto W(A, P)$ is
 continuous on $\sA_k$ with respect to the topology induced by the Hausdoff
-metric $H(\cdot, \cdot)$, defined for compact subsets $A, B$ of $\mathbb{R}^d$
-by $H(A,B) < \delta$ if and only if every point of $A$ is within Euclidean
-distance $\delta$ of at least one point of $B$, and vice versa.
+metric.
 \end{lemma}
@@
 infimum over $\sE_k$.  This follows from the extreme value theorem, since
-$\sE_k$ is compact and, by Lemma~\ref{lem:pollard-continuous}, the map $A
+$\sE_k$ is compact and, by Lemma~\ref{lem:kmeans-continuous}, the map $A
 \mapsto W(A,P)$ is continuous over $\sA_k$.
@@
 
-\begin{lemma}[Uniform SLLN]\label{lem:pollard-uniform}
+\begin{lemma}[Uniform SLLN]\label{lem:kmeans-uniform-slln}
 If $\int \|x\|^2 P(x) < \infty$, then
@@
 
-\begin{theorem}[$k$-Means Consistency]
-Suppose that $\int \|x\|^2 P(dx) < \infty$ and that $A_n$ minimizes
-$W(\cdot, P_n)$ over $\sA_k$.  Then for all $n$ there exists an
-$\bar A_n \in \sbA_k$ such that $H(A_n, \bar A_n) \to 0\text{ a.s.}$,
-and $W(A_n, P_n) \to m_k(P)\text{ a.s.}$
-\end{theorem}
-\begin{proof}
+With Lemmas~\ref{lem:kmeans-continuous}--\ref{lem:kmeans-uniform-slln}, we can
+proceed to the proof of Theorem~\ref{lem:kmeans-consistency}.
+
+\begin{proof}[Proof of Theorem~\ref{lem:kmeans-consistency}]
+
 The proof proceeds inductively over $k$.  For $k=1$, the result follows from
@@
 there is a finite $M$ such that $A_n \in \sE_k$ almost surely for large
-enough~$n$.  Take $M$ to be large enough so that $\sE_k$ also contains $\sbA_k$.
+enough~$n$.  Take $M$ to be large enough so that $\sE_k$ also contains
+$\sbA_k$.
 
@@
 m_{k}(P) + \eta$.  Let $\bar A_0$ be any member of $\sbA_k$.  By construction,
-$W(A_n, P_n) \leq W(\bar A, P_n)$.  Lemma~\ref{lem:pollard-uniform} ensures
+$W(A_n, P_n) \leq W(\bar A, P_n)$.  Lemma~\ref{lem:kmeans-uniform-slln} ensures
 that $W(\bar A_0, P_n) \to m_k(P)\text{ a.s.}$  Thus, for $n$ large enough,
@@
 \in \sbA_k$ with $H(A_n, \bar A_n) < \varepsilon$.
-
 \end{proof}
@@
 $\bar A$ with $k$ or fewer points such that $W(\bar A,P) = m_k(P)$.
-Lemmas~\ref{lem:pollard-compact} and~\ref{lem:pollard-continuous} together
+Lemmas~\ref{lem:pollard-compact} and~\ref{lem:kmeans-continuous} together
 with the extreme value theorem imply that this class is nonempty.

### 2015-06-19 cvclust.tex
add kmeans separation lemma

@@
 $X_i$ and $Y_i$ are independent.
+
+\begin{lemma}
+Set $\mu = \int x P(dx)$.
+If $\int \|x\|^2 P(dx) < \infty$, then there exists an $r > 0$ such that
+for every $A \in \sbA_k$, there is at most one center $a \in A$ with
+$\|a - \mu\| \leq r$.
+\end{lemma}
+\begin{proof}
+If the support of $P$ has fewer than $k$ points, then the result is trivial.
+Otherwise, Lemma~\ref{lem:kmeans-existence} implies that
+$m_k(P) < m_{k-1}(P)$.  Define
+\[
+  d =
+    \inf_{A \in \sA_k}
+      \min \{ \|a - a'\| : a,a' \in A, a \neq a' \}.
+\]
+If $d = 0$, then there exists a sequence $A_n$ in $\sbA_k$ with
+\[
+  \min_{\substack{a,a' \in A_n \\ a \neq a'}} \|a - a'\| \to 0.
+\]
+Since $\sbA_k$ is compact, there exists a subsequence $n(m)$ and a set $A_0
+\in \sbA_k$ with $A_{n(m)} \to A_0$.
+Take $B_n \in \sA_{k-1}$ to be the set formed by removing one of the centers
+$a'$ minimizing $\min_{\substack{a \in A_n \\ a \neq a'}} \|a - a'\|$.
+By construction, $H(A_n, B_n) \to 0$, so that $B_{n(m)} \to A_0$.
+Since the map $A \mapsto W(A, P)$ is continuous on $\sA_k$, we must have
+$W(B_{n(m)}, P) \to W(A_0, P) = m_k(P)$.  However, this is impossible since
+$W(B_n,P) \geq m_{k-1}(P) > m_k(P)$.  It must be the case, then, that $d > 0$.
+
+Set $r < d/2$, and let $A \in \sbA_k$ be arbitrary.  Take $a_1$ to be the
+cluster center in $A$ closest to $\mu$ (breaking ties arbitrarily).  If
+$\|a_1 - \mu\| > r$, then $\|a - \mu\| > r$ for all centers $a \in A$.
+Otherwise, if $\|a_1 - \mu\| \leq r$ and $a \in A$ is any other cluster center, then
+\(
+  2r < \|a - a_1\| \leq \|a - \mu\| + \|a_1 - \mu\|
+\)
+so that
+\(
+  \|a - \mu\| > r.
+\)
+\end{proof}
 

### 2015-06-19 demo/bench/05_tabulate_results.R
add messages to step 05

@@ -82,6 +82,7 @@ print_table <- function(setting, summary)
 }
 
 for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
+    cat("tabulating results for '", s, "'\n", sep="")
     prederr <- readRDS(file.path(s, "prederr.rds"))
     nrep <- length(prederr)
     kmax <- length(prederr[[1]])

### 2015-06-19 demo/bench/04_compute_prederr.R
add status method to step 04

@@ -4,6 +4,8 @@ for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
     if (file.exists(file.path(s, "prederr.rds")))
         next
 
+    cat("computing prediction error for '", s, "'\n", sep="")
+
     replicates <- readRDS(file.path(s, "replicates.rds"))
     kmeans <- readRDS(file.path(s, "kmeans.rds"))
 

### 2015-06-19 demo/bench/01_generate_data.R
switch to L'Ecuyer parallel RNG

@@ -1,7 +1,15 @@
 #!/usr/bin/Rscript --vanilla
 
 library("MASS")
+library("parallel")
 source("settings.R") # settings
+
+
+# Use the L'Ecuyer RNG so that we can split the seeds, allowing
+# each replicate to have a separate stream of random numbers.
+rng_kind <- "L'Ecuyer-CMRG"
+rng_normal_kind <- "Inversion"
+
 
 nrep <- 100
 
@@ -14,12 +22,19 @@ for (s in names(settings)) {
 
         cat("generating replicates for '", s, "'\n", sep="")
 
-        set.seed(settings[[s]]$seed)
+        set.seed(settings[[s]]$seed, rng_kind, rng_normal_kind)
         replicates <- vector("list", nrep)
 
         pb <- txtProgressBar(0, nrep, style=3)
         for (r in seq_len(nrep)) {
             replicates[[r]] <- settings[[s]]$simulate()
+
+            # split the RNG and save one stream with the replicate
+            s0 <- .Random.seed
+            replicates[[r]][["rng"]] <- list(seed = nextRNGSubStream(s0),
+                                             kind = rng_kind,
+                                             normal_kind = rng_normal_kind)
+            .Random.seed <- nextRNGStream(s0)
             setTxtProgressBar(pb, r)
         }
         close(pb)

### 2015-06-19 demo/bench/03_select_nclusters.R

@@ -39,7 +39,6 @@ for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
         if (!file.exists(f.method)) {
             cat("applying method '", m, "' to '", s, "'\n", sep="")
 
-            set.seed(5453546) # from random.org
             method <- methods[[m]]
             nclusters <- integer(nrep)
 
@@ -54,9 +53,10 @@ for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
                                            "mclustBIC",
                                            "WoldIter",
                                            "Wold_holdout"))
-            nclusters <- parSapply(cl, seq_len(nrep), function(r) {
-                    x <- replicates[[r]]$x
-                    tryCatch(method(x, kmax),
+            nclusters <- parSapply(cl, replicates, function(r) {
+                    rng <- r$rng
+                    set.seed(rng$seed, rng$kind, rng$normal_kind)
+                    tryCatch(method(r$x, kmax),
                              error=function(e) NA_integer_)
                 })
 

### 2015-06-19 cvclust.tex
minor cleanups to tex file

@@
 \newcommand{\E}{\operatorname{E}}
+\newcommand{\R}{\mathbb{R}}
 \newcommand{\dataX}{\mathfrak{X}}
@@
 predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
-\mathbb{R}^{p}$ and $Y_i \in \mathbb{R}^{q}$.  The data can be represented as
-a matrix with $N$ rows and $p + q$ columns.  We partition the data into $K$
-hold-out ``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
+\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
+$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
+``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
 ``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
@@
 where
-$\Xtrain \in \mathbb{R}^{n \times p}$,
-$\Ytrain \in \mathbb{R}^{n \times q}$,
-$\Xtest \in \mathbb{R}^{m \times p}$,
+$\Xtrain \in \R^{n \times p}$,
+$\Ytrain \in \R^{n \times q}$,
+$\Xtest \in  \R^{m \times p}$,
 and
-$\Ytest \in \mathbb{R}^{m \times q}$.
+$\Ytest \in  \R^{m \times q}$.
 
@@
     Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding the
-    assignment rule $\hGY : \mathbb{R}^q \to \{ 1, \dotsc, k \}$ and the
+    assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ and the
     clusters means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
@@
     the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to train a classifier
-    $\hGX : \mathbb{R}^p \to \{ 1, \dotsc, k \}$.
+    $\hGX : \R^p \to \{ 1, \dotsc, k \}$.
 
@@
 \subsection{Self-Consistency}
-Same as in section $2$, let the data set $\mathbf{X} \in \mathbb{R}^{n \times p}$ been partitioned as 
+Same as in section $2$, let the data set $\mathbf{X} \in \R^{n \times p}$ been partitioned as
  \[ 
@@
 
-\subsection{Single Cluster With Noise}
-
-Here, we show that Gabriel Cross-Validation is consistent when the data come
-from a single cluster; the algorithm picks $k = 1$ with probability tending to
-$1$ as the number of rows in the data matrix $\dataX$ increase.
-
-Our proof relies on results from \citet{pollard1981strong}, who analyzed
-$k$-means and showed that the within-cluster sum of squares converges to a
-population quantity.  Specifically, for any finite set $A \subset
-\mathbb{R}^d$ and probability measure $P$ supported on $\mathbb{R}^d$, define
-the population within-cluster sum of squares as
+\clearpage
+
+\section{Asymptotic Analysis of $k$-Means}
+
+To analyze the asymptotic properties of of Gabriel cross validation, we first
+need to understand the properties of the $k$-means procedure.  We will use
+results from \citet{pollard1981strong}, who analyzed $k$-means and showed that
+the within-cluster sum of squares converges to a population quantity.  To
+introduce Pollard's result, for any finite set $A \subseteq \R^d$ and
+probability measure $P$ supported on $\R^d$, define the population
+within-cluster sum of squares as
 \[
@@
     =
-    \{ A \subset \mathbb{R}^d : \text{$A$ contains $k$ or fewer points} \}
+    \{ A \subseteq \R^d : \text{$A$ contains $k$ or fewer points} \}
 \)
@@
 \(
-  m_k(P) = \inf_{A \in \sA_k}  W(A, P). % : \text{$A$ contains $k$ or fewer points} \}.
+  m_k(P) = \inf_{A \in \sA_k}  W(A, P).
 \)
-Suppose that $X_1, \dotsc, X_n$ are independent draws from $P$, and if $P_n$
+Suppose that $X_1, \dotsc, X_n$ are independent draws from $P$, and that $P_n$
 is the empirical measure formed by placing mass $1/n$ on each $X_i$.  Let
@@
 
-To state the consistency theorem define
-\[
-  \sbA_k = \{ \bar A \subseteq \mathbb{R}^d
+To state the generalized version of Pollard's consistency theorem, define
+\[
+  \sbA_k = \{ \bar A \subseteq \R^d
             : W(\bar A, P) = m_k(P)
@@
 \]
-We will later show that $\sbA_k$ is nonempty and compact whenever
-$\int \|x\|^2 P(dx) < \infty$.
-Let $H(\cdot, \cdot)$ denote the Hausdorff metric, defined for compact subsets
-$A, B$ of $\mathbb{R}^d$ by $H(A,B) < \delta$ if and only if every point of
-$A$ is within Euclidean distance $\delta$ of at least one point of $B$, and
-vice versa.  We can now state the generalized version of Pollard's $k$-means
-consistency theorem.
+We will later show that $\sbA_k$ is nonempty and compact whenever $\int
+\|x\|^2 P(dx) < \infty$.  Let $H(\cdot, \cdot)$ denote the Hausdorff metric,
+defined for compact subsets $A, B$ of $\R^d$ by $H(A,B) < \delta$ if
+and only if every point of $A$ is within Euclidean distance $\delta$ of at
+least one point of $B$, and vice versa.  We can now state the generalized
+$k$-means consistency theorem.
 
 \begin{theorem}[$k$-Means Consistency]\label{lem:kmeans-consistency}
-Suppose that $\int \|x\|^2 P(dx) < \infty$ and that $A_n$ minimizes
-$W(\cdot, P_n)$ over $\sA_k$.  Then for all $n$ there exists an
-$\bar A_n \in \sbA_k$ such that $H(A_n, \bar A_n) \to 0\text{ a.s.}$,
-and $W(A_n, P_n) \to m_k(P)\text{ a.s.}$
-\end{theorem}
+Suppose that $\int \|x\|^2 P(dx) < \infty$ and that $A_n$ minimizes $W(\cdot,
+P_n)$ over $\sA_k$.  Then for all $n$ there exists an $\bar A_n \in \sbA_k$
+such that $H(A_n, \bar A_n) \to 0\text{ a.s.}$, and $W(A_n, P_n) \to
+m_k(P)\text{ a.s.}$ \end{theorem}
 
@@
   \item the class of sets $\sbA_j = \{ A \in \sA_j : W(A,P) = m_{j}(P) \}$
-    is compact (with respect to the Hausdorff metric on $\sA_j$) for $j = 1, \dotsc, k$.
+    is compact (with respect to the Hausdorff metric on $\sA_j$) for $j = 1,
+    \dotsc, k$.
 \end{enumerate}
@@
 Let $\sC_k$ denote the class of sets $A$ with at most $k$ points and at least
-one point in in $B(M)$, the closed ball with radius $M$ centered at the origin.  Let
-$A_0$ be the singleton set containing the origin.
-If no point of $A$ is in $B(M)$, then
+one point in in $B(M)$, the closed ball with radius $M$ centered at the
+origin.  Let $A_0$ be the singleton set containing the origin.  If no point of
+$A$ is in $B(M)$, then
 \(
@@
 minimizers lie in $\sE_k$.  Since $A \mapsto W(\cdot, P)$ is continuous and
-$\sE_k$ is compact, the class $\sbA_k$, the preimage of $\{ m_k(P) \}$ under this
-map, is compact.
-
-We first show that if $M$ is large enough, then there exists a set $B \in \sE_k$
-and a positive $\varepsilon$ with $W(B,P) < m_{k-1}(P) - \varepsilon$.  This
-will establish that $m_{k}(P) < m_{k-1}(P)$, and the set $B$ will have further
-utility later in the proof.
-Take $\bar A_{k-1} \in \sE_{k-1}$ to be a set with at most $k-1$ points satisfying
-$m_{k-1}(P) = W(\bar A_{k-1},P)$.
-Since the support of $P$ contains least $k$ points, there exists a point~$b_1$ not
-in $\bar A_{k-1}$ that lies in the support of $P$.  Ensure that $M$ is large
-enough so that $b_1 \in B(5M)$.  Take $B \in \sE_k$ to be the set of $k$
+$\sE_k$ is compact, the class $\sbA_k$, the preimage of $\{ m_k(P) \}$ under
+this map, is compact.
+
+We first show that if $M$ is large enough, then there exists a set $B \in
+\sE_k$ and a positive $\varepsilon$ with $W(B,P) < m_{k-1}(P) - \varepsilon$.
+This will establish that $m_{k}(P) < m_{k-1}(P)$, and the set $B$ will have
+further utility later in the proof.  Take $\bar A_{k-1} \in \sE_{k-1}$ to be a
+set with at most $k-1$ points satisfying $m_{k-1}(P) = W(\bar A_{k-1},P)$.
+Since the support of $P$ contains least $k$ points, there exists a point~$b_1$
+not in $\bar A_{k-1}$ that lies in the support of $P$.  Ensure that $M$ is
+large enough so that $b_1 \in B(5M)$.  Take $B \in \sE_k$ to be the set of $k$
 points gotten by adjoining $b_1$ to $\bar A_{k-1}$.  Let
@@
 \)
-and set $R = \{ x \in \mathbb{R}^d : \|x - b_1\| < \delta/3 \}$; since $b_1$
+and set $R = \{ x \in \R^d : \|x - b_1\| < \delta/3 \}$; since $b_1$
 is in the support of $P$, the neighborhood $R$ has positive $P$~measure.
@@
       \int_{R} \min_{a \in \bar A_{k-1}} (\|b_1 - a\| - \delta/3)^2 P(dx)
-  = (4/9) \delta^2 P(R),
+    =
+      (4/9) \delta^2 P(R),
 \]
@@
 \(
-  W(B,P) \leq W(\bar A_{k-1},P) - (3/9) \delta^2 P(R)
-         < m_{k-1}(P) - \varepsilon.
+  W(B,P)
+    \leq
+      W(\bar A_{k-1},P) - (3/9) \delta^2 P(R)
+    <
+      m_{k-1}(P) - \varepsilon.
 \)
@@
 \]
-We will now show that if $A \in \sC_k \setminus \sE_k$, then
-$W(A,P) > W(B,P) \geq m_{k}(P)$, which ensures that $W(\cdot, P)$ does not
-attain its infimum over $\sA_k$ outside of $\sE_k$.  To see that this holds,
-suppose that $A$ is any set $\sC_k \setminus \sE_k$.  This set has at least
-one point, $a_1$, in $B(M)$, and at least one point outside $B(5M)$.  Consider
-thet set $A^\ast$ formed by removing all points of $A$ outside $B(5M)$.  Any
-point $x$ closer to a cluster center outisde $B(5M)$ than to $a_1$ must
-satisfy $\|x\| \geq 2M$.  Thus,
+We will now show that if $A \in \sC_k \setminus \sE_k$, then $W(A,P) > W(B,P)
+\geq m_{k}(P)$, which ensures that $W(\cdot, P)$ does not attain its infimum
+over $\sA_k$ outside of $\sE_k$.  To see that this holds, suppose that $A$ is
+any set $\sC_k \setminus \sE_k$.  This set has at least one point, $a_1$, in
+$B(M)$, and at least one point outside $B(5M)$.  Consider thet set $A^\ast$
+formed by removing all points of $A$ outside $B(5M)$.  Any point $x$ closer to
+a cluster center outisde $B(5M)$ than to $a_1$ must satisfy $\|x\| \geq 2M$.
+Thus,
 \begin{align*}
@@
 \end{align*}
-Since $A^\ast$ has $k-1$ or fewer points, it must satisfy
-$W(A^\ast, P) \geq m_{k-1}(P)$, so that
-$W(A, P) > m_{k-1}(P) - \varepsilon > W(B, P)$.
-
-We have now established that $m_k(P) = \inf \{ W(A,P) : A \in \sE_k \}$,
-and that $W(\cdot, P)$ does not obtain its infimum over $\sA_k$ outside of
+Since $A^\ast$ has $k-1$ or fewer points, it must satisfy $W(A^\ast, P) \geq
+m_{k-1}(P)$, so that $W(A, P) > m_{k-1}(P) - \varepsilon > W(B, P)$.
+
+We have now established that $m_k(P) = \inf \{ W(A,P) : A \in \sE_k \}$, and
+that $W(\cdot, P)$ does not obtain its infimum over $\sA_k$ outside of
 $\sE_k$.  The final step of the proof is to show $W(\cdot, P)$ obtains its
@@
 least $k$ points, and that $k > 1$.  Let $A_n$ minimize $W(\cdot, P_n)$ over
-$\sA_k$, and that $B_n$ minimize $W(\cdot P_n)$ over $\sA_{k-1}$.  If
-$W(B_n, P) \to m_{k-1}(P)$ almost surely as $n \to \infty$, then
-there exists a finite $M$ such that as if $n$ is large enough,
-then, almost surely, $A_n$ is in the class of sets
+$\sA_k$, and that $B_n$ minimize $W(\cdot P_n)$ over $\sA_{k-1}$.  If $W(B_n,
+P) \to m_{k-1}(P)$ almost surely as $n \to \infty$, then there exists a finite
+$M$ such that as if $n$ is large enough, then, almost surely, $A_n$ is in the
+class of sets
 \(
-  \sE_k = \{ A \subset B(5M) : \text{$A$ contains $k$ or fewer points} \},
+  \sE_k = \{ A \subseteq B(5M) : \text{$A$ contains $k$ or fewer points} \},
 \)
@@
 \begin{proof}
-Pollard proves this result, in Section~3 of his paper,  assuming that there
-is a unique set of $j$ points minimizing $W(\cdot, P)$ for each value of
-$j = 1, \dotsc, k$.  His proof, which is very similar to the proof of
+Pollard proves this result, in Section~3 of his paper,  assuming that there is
+a unique set of $j$ points minimizing $W(\cdot, P)$ for each value of $j = 1,
+\dotsc, k$.  His proof, which is very similar to the proof of
 Lemma~\ref{lem:kmeans-existence}, does not use this assumption directly, only
@@
 \begin{proof}[Proof of Theorem~\ref{lem:kmeans-consistency}]
-
 The proof proceeds inductively over $k$.  For $k=1$, the result follows from
@@
 
-The set $\sE_k$ is compact and the map $f: \sE_k \to \mathbb{R}$ with $f(A) =
+The set $\sE_k$ is compact and the map $f: \sE_k \to \R$ with $f(A) =
 W(A, P)$ is continuous.  Thus, for every $\varepsilon > 0$ there exists an
@@
 m_{k}(P) + \eta$.  Let $\bar A_0$ be any member of $\sbA_k$.  By construction,
-$W(A_n, P_n) \leq W(\bar A, P_n)$.  Lemma~\ref{lem:kmeans-uniform-slln} ensures
-that $W(\bar A_0, P_n) \to m_k(P)\text{ a.s.}$  Thus, for $n$ large enough,
-$W(A_n, P_n) < m_{k}(P) + \eta$, which implies that there exists an $\bar A_n
-\in \sbA_k$ with $H(A_n, \bar A_n) < \varepsilon$.
-\end{proof}
-
-Now, define $\mathcal{\bar A} = \mathcal{\bar A}(k)$ to be the class of sets
-$\bar A$ with $k$ or fewer points such that $W(\bar A,P) = m_k(P)$.
-Lemmas~\ref{lem:pollard-compact} and~\ref{lem:kmeans-continuous} together
-with the extreme value theorem imply that this class is nonempty.
-
-To prove
-this, it suffices to analyze a single fold of the cross-validation procedure.
-We will assume the following setup:
-
-Observations $\{ (X_i, Y_i) \}_{i=1}^{n+N}$ are independent and identically
-distributed.  The distribution for each $(X_i, Y_i)$ pair is such that $X_i$
-and $Y_i$ have marginal distribution $P^X$ and $P^Y$, and the vectors
-$X_i$ and $Y_i$ are independent.
+$W(A_n, P_n) \leq W(\bar A_0, P_n)$.  Lemma~\ref{lem:kmeans-uniform-slln}
+ensures that $W(\bar A_0, P_n) \to m_k(P)\text{ a.s.}$  Thus, for $n$ large
+enough, $W(A_n, P_n) < m_{k}(P) + \eta$, which implies that there exists an
+$\bar A_n \in \sbA_k$ with $H(A_n, \bar A_n) < \varepsilon$.
+\end{proof}
+
+The following lemma is not required for the $k$-means consistency theorem, but
+it will prove necessary for analyzing the asymptotic behavior of Gabriel
+cross-validation.
 
 \begin{lemma}
-Set $\mu = \int x P(dx)$.
-If $\int \|x\|^2 P(dx) < \infty$, then there exists an $r > 0$ such that
-for every $A \in \sbA_k$, there is at most one center $a \in A$ with
-$\|a - \mu\| \leq r$.
+Set $\mu = \int x P(dx)$.  If $\int \|x\|^2 P(dx) < \infty$, then there exists
+an $r > 0$ such that for every $A \in \sbA_k$, there is at most one center $a
+\in A$ with $\|a - \mu\| \leq r$.
 \end{lemma}
@@
 If the support of $P$ has fewer than $k$ points, then the result is trivial.
-Otherwise, Lemma~\ref{lem:kmeans-existence} implies that
-$m_k(P) < m_{k-1}(P)$.  Define
+Otherwise, Lemma~\ref{lem:kmeans-existence} implies that $m_k(P) <
+m_{k-1}(P)$.  Define
 \[
@@
 Since $\sbA_k$ is compact, there exists a subsequence $n(m)$ and a set $A_0
-\in \sbA_k$ with $A_{n(m)} \to A_0$.
-Take $B_n \in \sA_{k-1}$ to be the set formed by removing one of the centers
-$a'$ minimizing $\min_{\substack{a \in A_n \\ a \neq a'}} \|a - a'\|$.
-By construction, $H(A_n, B_n) \to 0$, so that $B_{n(m)} \to A_0$.
-Since the map $A \mapsto W(A, P)$ is continuous on $\sA_k$, we must have
-$W(B_{n(m)}, P) \to W(A_0, P) = m_k(P)$.  However, this is impossible since
-$W(B_n,P) \geq m_{k-1}(P) > m_k(P)$.  It must be the case, then, that $d > 0$.
+\in \sbA_k$ with $A_{n(m)} \to A_0$.  Take $B_n \in \sA_{k-1}$ to be the set
+formed by removing one of the centers $a'$ minimizing $\min_{\substack{a \in
+A_n \\ a \neq a'}} \|a - a'\|$.  By construction, $H(A_n, B_n) \to 0$, so that
+$B_{n(m)} \to A_0$.  Since the map $A \mapsto W(A, P)$ is continuous on
+$\sA_k$, we must have $W(B_{n(m)}, P) \to W(A_0, P) = m_k(P)$.  However, this
+is impossible since $W(B_n,P) \geq m_{k-1}(P) > m_k(P)$.  It must be the case,
+then, that $d > 0$.
 
 Set $r < d/2$, and let $A \in \sbA_k$ be arbitrary.  Take $a_1$ to be the
-cluster center in $A$ closest to $\mu$ (breaking ties arbitrarily).  If
-$\|a_1 - \mu\| > r$, then $\|a - \mu\| > r$ for all centers $a \in A$.
-Otherwise, if $\|a_1 - \mu\| \leq r$ and $a \in A$ is any other cluster center, then
+cluster center in $A$ closest to $\mu$ (breaking ties arbitrarily).  If $\|a_1
+- \mu\| > r$, then $\|a - \mu\| > r$ for all centers $a \in A$.  Otherwise, if
+$\|a_1 - \mu\| \leq r$ and $a \in A$ is any other cluster center, then
 \(
@@
 
+\section{Asymptotic Analysis of Gabriel Cross-Validation}
+
+Here, we show that Gabriel Cross-Validation is consistent when the data come
+from a single cluster; the algorithm picks $k = 1$ with probability tending to
+$1$ as the number of rows in the data matrix $\dataX$ increase.
+
+To prove consistency of Gabriel cross-validation, it suffices to analyze a
+single fold of the cross-validation procedure.  We will assume the following
+setup:
+
+Observations $\{ (X_i, Y_i) \}_{i=1}^{n+N}$ are independent and identically
+distributed.  The distribution for each $(X_i, Y_i)$ pair is such that $X_i$
+and $Y_i$ have marginal distribution $P^X$ and $P^Y$, and the vectors $X_i$
+and $Y_i$ are independent.
+
+\clearpage
+
 \subsection{Well-separated clusters with noise}
 
-Same with previous, the data set $\mathbf{X} \in \mathbb{R}^{n \times p}$ been
-partitioned as $\mathbf{X}^L \in \mathbb{R}^{n \times L}$ and $\mathbf{X}^R
-\in \mathbb{R}^{n \times R}$, with following set of assumptions:
+Same with previous, the data set $\mathbf{X} \in \R^{n \times p}$ been
+partitioned as $\mathbf{X}^L \in \R^{n \times L}$ and $\mathbf{X}^R
+\in \R^{n \times R}$, with following set of assumptions:
 
@@
 \]
-for all constants $a$, $b$, and $\delta$, where
-$\varphi(z)$ and $\Phi(z)$ are the standard normal probability density and
-cumulative distribution functions.  These expressions are valid
-for $a = -\infty$ or $b = \infty$ by taking limits.
+for all constants $a$, $b$, and $\delta$, where $\varphi(z)$ and $\Phi(z)$ are
+the standard normal probability density and cumulative distribution functions.
+These expressions are valid for $a = -\infty$ or $b = \infty$ by taking
+limits.
 
@@
 \begin{proof}
-We will derive the expression for the second moment.
-Integrate to get
+We will derive the expression for the second moment.  Integrate to get
 \begin{align*}
@@
 from a single cluster; the true number of clusters is~$1$.  We will show that
-Gabriel cross-validation is inconsistent if $\rho > 0.5$, in the sense that
-$k = 2$ will be preferred to $k = 1$ almost surely for large enough $n$.
+Gabriel cross-validation is inconsistent if $\rho > 0.5$, in the sense that $k
+= 2$ will be preferred to $k = 1$ almost surely for large enough $n$.
 
@@
 Gabriel cross-validation error converges to $1 + a^2 (1 - 2 \rho)$.  In the
-$\rho < 0$ case, it converges to $1 + a^2 (1 + 2 \rho)$.  In
-particular, if $|\rho| > 0.5$, then the asymptotic cross-validation error for $k
-= 2$ will be smaller than for $k = 1$.
+$\rho < 0$ case, it converges to $1 + a^2 (1 + 2 \rho)$.  In particular, if
+$|\rho| > 0.5$, then the asymptotic cross-validation error for $k = 2$ will be
+smaller than for $k = 1$.
 
@@
 cross-validation mean squared error for the number of clusters $k$ ranging
-from $1$ to $5$.  Figure~\ref{fig:nullcorr-equal} shows the
-cross-validation errors for all $10$ replicates.  The simulation demonstrates
-that in the Gabriel cross-validation criterion chooses the correct answer $k =
-1$ whenever $\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever
-$|\rho| > 0.5$.
+from $1$ to $5$.  Figure~\ref{fig:nullcorr-equal} shows the cross-validation
+errors for all $10$ replicates.  The simulation demonstrates that in the
+Gabriel cross-validation criterion chooses the correct answer $k = 1$ whenever
+$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.
 

### 2015-06-19 cvclust.tex
typo

@@
     assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ and the
-    clusters means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
+    cluster means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
     be the assigned cluster for row $i$.

### 2015-06-19 cvclust.tex
typos

@@
 any set $\sC_k \setminus \sE_k$.  This set has at least one point, $a_1$, in
-$B(M)$, and at least one point outside $B(5M)$.  Consider thet set $A^\ast$
+$B(M)$, and at least one point outside $B(5M)$.  Consider the set $A^\ast$
 formed by removing all points of $A$ outside $B(5M)$.  Any point $x$ closer to
-a cluster center outisde $B(5M)$ than to $a_1$ must satisfy $\|x\| \geq 2M$.
+a cluster center outside $B(5M)$ than to $a_1$ must satisfy $\|x\| \geq 2M$.
 Thus,

### 2015-06-20 demo/bench/03_select_nclusters.R
add Gabriel classify-nearest method

@@ -6,6 +6,7 @@ library("devtools")
 library("e1071")
 library("mclust")
 library("MASS")
+library("nnet")
 library("parallel")
 load_all("../../lib/fpc", export_all=FALSE)
 load_all("../../lib/NbClust", export_all=FALSE)
@@ -45,12 +46,14 @@ for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
 
             clusterExport(cl=cl, varlist=c("replicates", "method", "kmax",
                                            "classify_lda",
+                                           "classify_nearest",
                                            "cluster_kmeans",
                                            "compute.jump",
                                            "cv.kmeans.gabriel",
                                            "Impute",
                                            "jump",
                                            "kmeans.rndstart",
+                                           "predict.classify_nearest",
                                            "mclustBIC",
                                            "WoldIter",
                                            "Wold_holdout"))

### 2015-06-20 demo/bench/methods.R

@@ -7,6 +7,7 @@
 #   library("e1071")
 #   library("mclust")
 #   library("MASS")
+#   library("nnet")
 #   load_all("../../lib/fpc")
 #   load_all("../../lib/NbClust")
 #   source("../../lib/NbClust.R")
@@ -18,6 +19,11 @@
 #   
 
 methods <- list(
+    "gabriel-nearest" = function(x, maxcenters) {
+        cv <- cv.kmeans.gabriel(x, 2, 2, maxcenters,
+                                classify.method="nearest")
+        cv$centers
+    },
     "gabriel-lda-equal" = function(x, maxcenters) {
         cv <- cv.kmeans.gabriel(x, 2, 2, maxcenters,
                                 classify.method="lda-equal")

### 2015-06-21 demo/bench/02_fit_kmeans.R
fit kmeans in parallel

@@ -1,9 +1,14 @@
 #!/usr/bin/Rscript --vanilla
 
 source("../../code/cluster.R") # cluster_kmeans
+library("parallel")
 
 kmax <- 15
 
+options(cl.cores=max(1, detectCores() - 1))
+cl <- makeCluster(getOption("cl.cores", 2))
+
+clusterExport(cl=cl, varlist=c("cluster_kmeans", "kmax"))
 
 for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
     f.replicates <- file.path(s, "replicates.rds")
@@ -13,20 +18,16 @@ for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
         cat("fitting k-means for '", s, "'\n", sep="")
 
         replicates <- readRDS(f.replicates)
-        nrep <- length(replicates)
-        kmeans <- vector("list", nrep)
-
-        pb <- txtProgressBar(0, nrep, style=3)
-        for (r in seq_len(nrep)) {
-            kmeans[[r]] <- vector("list", kmax)
-            for (k in seq_len(kmax)) {
-                kmeans[[r]][[k]] <- cluster_kmeans(replicates[[r]]$x, k)
-            }
-            setTxtProgressBar(pb, r)
-        }
-        close(pb)
+        kmeans <- parLapply(cl, replicates, function(r) {
+                km <- vector("list", kmax)
+                for (k in seq_len(kmax)) {
+                    km[[k]] <- cluster_kmeans(r$x, k)
+                }
+                km
+            })
 
         saveRDS(kmeans, f.kmeans)
     }
 }
 
+stopCluster(cl)

### 2015-06-21 demo/bench/01_generate_data.R
simplify seed distribution

@@ -5,8 +5,8 @@ library("parallel")
 source("settings.R") # settings
 
 
-# Use the L'Ecuyer RNG so that we can split the seeds, allowing
-# each replicate to have a separate stream of random numbers.
+# Use the L'Ecuyer RNG so that we can have multiple streams of
+# random numbers
 rng_kind <- "L'Ecuyer-CMRG"
 rng_normal_kind <- "Inversion"
 
@@ -22,19 +22,27 @@ for (s in names(settings)) {
 
         cat("generating replicates for '", s, "'\n", sep="")
 
+        # generate the list of seeds
         set.seed(settings[[s]]$seed, rng_kind, rng_normal_kind)
+        seeds <- vector("list", nrep)
+        seeds[[1L]] <- .Random.seed
+        for (r in seq_len(nrep - 1L))
+            seeds[[r+1L]] <- nextRNGStream(seeds[[r]])
+
+        # generate the replicates
         replicates <- vector("list", nrep)
 
         pb <- txtProgressBar(0, nrep, style=3)
         for (r in seq_len(nrep)) {
+            # set the seed for this replicate
+            .Random.seed <<- seeds[[r]]
             replicates[[r]] <- settings[[s]]$simulate()
 
-            # split the RNG and save one stream with the replicate
-            s0 <- .Random.seed
-            replicates[[r]][["rng"]] <- list(seed = nextRNGSubStream(s0),
+            # save the state of the RNG
+            replicates[[r]][["rng"]] <- list(seed = .Random.seed,
                                              kind = rng_kind,
                                              normal_kind = rng_normal_kind)
-            .Random.seed <- nextRNGStream(s0)
+
             setTxtProgressBar(pb, r)
         }
         close(pb)

### 2015-06-21 cvclust.tex

@@
 
+% operators
 \newcommand{\CV}{\operatorname{CV}}
 \newcommand{\E}{\operatorname{E}}
+\DeclareMathOperator*{\cov}{cov}
+\DeclareMathOperator*{\tr}{tr}
+\DeclareMathOperator*{\var}{var}
+
+% convergence
+\newcommand{\toas}{\overset{\mathit{a.s.}}{\to}}
+
+% sets
 \newcommand{\R}{\mathbb{R}}
-\newcommand{\dataX}{\mathfrak{X}}
-\newcommand{\Xtrain}{X_{\text{train}}}
-\newcommand{\Ytrain}{Y_{\text{train}}}
-\newcommand{\Xtest}{X_{\text{test}}}
-\newcommand{\Ytest}{Y_{\text{test}}}
-\newcommand{\hGX}{\hat G^{X}}
-\newcommand{\hGY}{\hat G^{Y}}
-\newcommand{\bpi}{\bar \pi}
-\newcommand{\bmuX}{\bar \mu^{X}}
-\newcommand{\bmuY}{\bar \mu^{Y}}
 \newcommand{\sA}{\mathcal{A}}
@@
 \newcommand{\sX}{\mathcal{X}}
+
+% scalars
+\newcommand{\bpi}{\bar \pi}
+
+% vectors
+\newcommand{\muX}{\mu^{X}}
+\newcommand{\muY}{\mu^{Y}}
+\newcommand{\bmuX}{\bar \mu^{X}}
+\newcommand{\bmuY}{\bar \mu^{Y}}
+
+% matrices
+\newcommand{\dataX}{\mathfrak{X}}
+\newcommand{\SigmaY}{\Sigma^Y}
+\newcommand{\Xtrain}{X_{\text{train}}}
+\newcommand{\Ytrain}{Y_{\text{train}}}
+\newcommand{\Xtest}{X_{\text{test}}}
+\newcommand{\Ytest}{Y_{\text{test}}}
+
+% class labels
+\newcommand{\hGX}{\hat G^{X}}
+\newcommand{\hGY}{\hat G^{Y}}
 
@@
 
-\subsection*{Gabriel CV algorithm}
+\subsection{Gabriel CV algorithm}
+\label{sec:gabriel-cv-algorithm}
 
@@
 
-\begin{theorem}[$k$-Means Consistency]\label{lem:kmeans-consistency}
+\begin{theorem}[$k$-Means Consistency]\label{thm:kmeans-consistency}
 Suppose that $\int \|x\|^2 P(dx) < \infty$ and that $A_n$ minimizes $W(\cdot,
@@
 \end{lemma}
-
 \begin{proof}
@@
 With Lemmas~\ref{lem:kmeans-continuous}--\ref{lem:kmeans-uniform-slln}, we can
-proceed to the proof of Theorem~\ref{lem:kmeans-consistency}.
-
-\begin{proof}[Proof of Theorem~\ref{lem:kmeans-consistency}]
+proceed to the proof of Theorem~\ref{thm:kmeans-consistency}.
+
+\begin{proof}[Proof of Theorem~\ref{thm:kmeans-consistency}]
 The proof proceeds inductively over $k$.  For $k=1$, the result follows from
@@
 
-\begin{lemma}
+\begin{lemma}\label{lem:kmeans-separation}
 Set $\mu = \int x P(dx)$.  If $\int \|x\|^2 P(dx) < \infty$, then there exists
@@
 
-Observations $\{ (X_i, Y_i) \}_{i=1}^{n+N}$ are independent and identically
-distributed.  The distribution for each $(X_i, Y_i)$ pair is such that $X_i$
-and $Y_i$ have marginal distribution $P^X$ and $P^Y$, and the vectors $X_i$
-and $Y_i$ are independent.
+\begin{theorem}
+For each $n$, let $N(n)$ be a nonnegative integer with $N(n) \to \infty$ as $n
+\to \infty$, and let $\{ (X_{n,i}, Y_{n,i}) \}_{i=1}^{n+N(n)}$ be independent
+identically distributed random variables.  For any fixed $k$, let $\CV(k)$
+be the cross-validation error from a single Gabriel CV fold, computed as
+described in Sec.~\ref{sec:gabriel-cv-algorithm}.  Suppose that
+\begin{enumerate}[label=(\roman*)]
+\item the distribution for each $(X, Y)$ pair is such that $X$ and $Y$ have
+  marginal distributions $P^X$ and $P^Y$, and the vectors $X$ and $Y$ are
+  independent of each other;
+
+\item the distribution $P^Y$ satisfies $\int \|y\|^2 P^Y(dy) < \infty$;
+
+\item the support of $P^Y$ contains at least two points.
+\end{enumerate}
+Then for any fixed $k > 1$, almost surely eventually  $\CV(k) > \CV(1)$.
+\end{theorem}
+\begin{proof}
+Set $\muY = \E(Y)$ and $\SigmaY = \cov(Y)$.  The strong law of large
+numbers implies that
+\(
+  \CV(1) \toas \int \|y - \mu^Y\|^2 P(dy) = \tr(\SigmaY).
+\)
+If $k > 1$, then Theorem~\ref{thm:kmeans-consistency} and
+Lemma~\ref{lem:kmeans-separation} imply that there exists an $r > 0$ such that
+almost surely eventually at most one cluster center in
+$\{ \bmuY_1, \dotsc, \bmuY_k \}$ is within distance $r$ of $\muY$.  Without
+loss of generality suppose that $\bmuY_1$ is the closest cluster center to
+$\muY$.
+\end{proof}
 

### 2015-06-22 demo/bench/05_tabulate_results.R
add gabriel nearest results; use consistent names for gabriel methods

@@ -25,10 +25,12 @@ method.names <- list(
     "Jump" = "Jump",
     "PS" = "Prediction Strength",
     "Stab" = "Stability",
-    "gabriel-lda-equal" = "Gabriel CV",
-    "gabriel-lda-equal-5X2" = "Gabriel-5X2 CV",
-    "gabriel-lda-proportion" = "Gabriel-proportion CV",
-    "gabriel-lda-proportion-5X2" = "Gabriel-proportion-5X2 CV",
+    "gabriel-nearest-2x2" = "Gabriel CV (Nearest; $2 \\times 2$)",
+    "gabriel-nearest-5x2" = "Gabriel CV (Nearest; $5 \\times 2$)",
+    "gabriel-lda-equal-2x2" = "Gabriel CV (LDA Equal; $2 \\times 2$)",
+    "gabriel-lda-equal-5x2" = "Gabriel CV (LDA Equal; $5 \\times 2$)",
+    "gabriel-lda-proportion-2x2" = "Gabriel (CV LDA Prop.; $2 \\times 2$)",
+    "gabriel-lda-proportion-5x2" = "Gabriel (CV LDA Prop.; $5 \\times 2$)",
     "wold" = "Wold CV")
 
 exclude <- function(setting, method)

### 2015-06-22 demo/bench/methods.R

@@ -19,27 +19,32 @@
 #   
 
 methods <- list(
-    "gabriel-nearest" = function(x, maxcenters) {
+    "gabriel-nearest-2x2" = function(x, maxcenters) {
         cv <- cv.kmeans.gabriel(x, 2, 2, maxcenters,
                                 classify.method="nearest")
         cv$centers
     },
-    "gabriel-lda-equal" = function(x, maxcenters) {
+    "gabriel-nearest-5x2" = function(x, maxcenters) {
+        cv <- cv.kmeans.gabriel(x, 5, 2, maxcenters,
+                                classify.method="nearest")
+        cv$centers
+    },
+    "gabriel-lda-equal-2x2" = function(x, maxcenters) {
         cv <- cv.kmeans.gabriel(x, 2, 2, maxcenters,
                                 classify.method="lda-equal")
         cv$centers
     },
-    "gabriel-lda-equal-5X2" = function(x, maxcenters) {
+    "gabriel-lda-equal-5x2" = function(x, maxcenters) {
         cv <- cv.kmeans.gabriel(x, 5, 2, maxcenters,
                               classify.method="lda-equal")
         cv$centers
     },
-    "gabriel-lda-proportion" = function(x, maxcenters) {
+    "gabriel-lda-proportion-2x2" = function(x, maxcenters) {
         cv <- cv.kmeans.gabriel(x, 2, 2, maxcenters,
                                 classify.method="lda-proportions")
         cv$centers
     },
-    "gabriel-lda-proportion-5X2" = function(x, maxcenters) {
+    "gabriel-lda-proportion-5x2" = function(x, maxcenters) {
         cv <- cv.kmeans.gabriel(x, 5, 2, maxcenters,
                               classify.method="lda-proportions")
         cv$centers

### 2015-06-22 cvclust.tex
switch to gabriel nearest 5x2 in document

@@
 steps~\ref{step:gabriel-cluster} and~\ref{step:gabriel-classify}.  In this
-report, we use $k$-means as the clustering algorithm and linear discriminant
-analysis (LDA) as the classification procedure.
+report, we use $k$-means as the clustering algorithm.  For the classification
+step, we compute the mean value of $X$ for each class; we assign an
+observation to class $g$ if that class has the closest mean (randomly breaking
+ties between classes).  The classification step is equivalent to linear
+discriminant analysis with equal class priors and identity noise covariance
+matrix.
 

### 2015-06-22 demo/bench/05_tabulate_results.R

@@ -25,12 +25,12 @@ method.names <- list(
     "Jump" = "Jump",
     "PS" = "Prediction Strength",
     "Stab" = "Stability",
-    "gabriel-nearest-2x2" = "Gabriel CV (Nearest; $2 \\times 2$)",
-    "gabriel-nearest-5x2" = "Gabriel CV (Nearest; $5 \\times 2$)",
-    "gabriel-lda-equal-2x2" = "Gabriel CV (LDA Equal; $2 \\times 2$)",
-    "gabriel-lda-equal-5x2" = "Gabriel CV (LDA Equal; $5 \\times 2$)",
-    "gabriel-lda-proportion-2x2" = "Gabriel (CV LDA Prop.; $2 \\times 2$)",
-    "gabriel-lda-proportion-5x2" = "Gabriel (CV LDA Prop.; $5 \\times 2$)",
+#    "gabriel-nearest-2x2" = "Gabriel CV ($2 \\times 2$)",
+    "gabriel-nearest-5x2" = "Gabriel CV ($5 \\times 2$)",
+#    "gabriel-lda-equal-2x2" = "Gabriel CV (LDA Equal; $2 \\times 2$)",
+#    "gabriel-lda-equal-5x2" = "Gabriel CV (LDA Equal; $5 \\times 2$)",
+#    "gabriel-lda-proportion-2x2" = "Gabriel (CV LDA Prop.; $2 \\times 2$)",
+#    "gabriel-lda-proportion-5x2" = "Gabriel (CV LDA Prop.; $5 \\times 2$)",
     "wold" = "Wold CV")
 
 exclude <- function(setting, method)

### 2015-06-22 cvclust.tex
start revising self-consistency section

@@
 % operators
+\DeclareMathOperator*{\argmin}{arg\,min}
 \newcommand{\CV}{\operatorname{CV}}
@@
 
-\section{Theoretical justification of Gabriel CV method}
+\section{Self-Consistency of Gabriel CV method}
 
 This section gives the self-consistency proof of the proposed Gabriel method.
-What's more, its asymptotic property under suitable condition will be shown in this section in order to give the idea of why the Gabriel CV method works.\\
-
-Because K-means algorithm is a key part in the proof below, it's convenient to have a review of K-means here:\\
-Given a set of observations $\mathbf{X} = (x_1,...,x_n)$, and the number of clusters $K$, K-means' goal is to minimize the within cluster dispersion.
-That is, find the partition $\mathbb{P}_K$ of $\mathbf{X}$, where $\mathbb{P}_K(\mathbf{X})= (p_1,p_2,...,p_K)$, such that
-\[	\underset{\mathbb{P}_K}{\operatorname{argmin}} \sum^K_{i=1}\sum_{x \in p_i } ||x-\hat{\mu}_i||^2\]
-where $\hat{\mu}_i$ is the sample mean of $p_i$.
-\subsection{Self-Consistency}
-Same as in section $2$, let the data set $\mathbf{X} \in \R^{n \times p}$ been partitioned as
- \[ 
- \begin{pmatrix}
-\mathbf{X}^L_{train} & \mathbf{X}^R_{train}\\
-  \mathbf{X}^L_{test}  & \mathbf{X}^R_{test}\\
-  \end{pmatrix}
- \]
-\subsection*{Assumptions}
-\begin{enumerate}[label={Assumption},leftmargin=*]
-\item $1.$ There are $K$ true centers $(\mu_1,...,\mu_K)$ where $\mathbf{X}$ generated from
-\item $2.$ The $K$ partitioned centers $(\mu^L_1,...,\mu^L_K)$ are all distinct in $L$ dimensions: $||\mu^L_i-\mu^L_j|| \neq 0$ for any $i \neq j$; $(\mu^R_1,...,\mu^R_K)$ are all distinct in $R$ dimensions
-\item $3.$ The training set contains at least one member of each cluster
-\item $4.$ The test set contains at least one member of each cluster
-\item $5.$ There is no noise, so that $$x_i = \mu \left(c_i\right) \in \{\mu_1,...,\mu_K\} $$ and $$\mu \left(c_i\right) = \mu_j \hspace{0.2in} \text{iff} \hspace{0.2in} c_i = j$$ where $c_i$ is the true cluster for observation $i$, which implies $$x^L_i = \mu^L\left(c_i\right) \in \{\mu^L_1,...,\mu^L_K\},\hspace{0.2in} x^R_i = \mu^R\left(c_i\right) \in \{\mu^R_1,...,\mu^R_K\}$$
+Specifically, we will show that under appropriate conditions, in the absence
+of noise, the Gabriel cross-validation procedure finds the optimal number of
+clusters.
+
+Because $k$-means algorithm is essential to the method, we review the
+procedure here.  Given a set of observations $\{ x_1, \dotsc ,x_n \}$, and a
+specified the number of clusters $k$, the goal of the $k$-means procedure is
+to find a set of $k$ or cluster centers $A = \{ a_1, \dotsc, a_k \} $
+minimizing the within cluster dispersion
+\[
+  W(A) = \sum_{i=1}^{n} \min_{a \in A} \|x_i - a\|^2.
+\]
+This implicitly defines a cluster assignment rule
+\[
+  g(x) = \argmin_{g \in \{1, \dotsc, k\}} \|x - a_g\|^2,
+\]
+with ties broken arbitrarily.  We will assume that the $k$-means procedure
+finds an optimal solution, $A$, but we will not assume that this solution is
+unique.
+
+It will suffice to analyze a single fold of the cross-validation procedure.
+As in in section~\ref{sec:gabriel-cv-algorithm} we assume that the $P$
+variables of the data set have been partitioned into $p$ predictor variables
+represented in vector~$X$ and $q$ response variables represented in
+vector~$Y$.  The $N$ observations have been divided into two sets: $n$ train
+observations and $m$ test observations.
+
+We make the following assumptions:
+\begin{enumerate}[label={\arabic*.},leftmargin=*]
+
+  \item There are $k$ true centers $\mu(1), \dotsc,\mu(k)$, with the $g$th
+    cluster center partitioned as $\mu(g) = \bigl(\muX(g), \muY(g)\bigr)$ with
+    $\muX(g) \in \R^p$ and $\muY(g) \in \R^q$ for $g = 1, \dotsc, k$.
+
+  \item The vectors $\muX(1), \dotsc,\muX(k)$ are all distinct; likewise,
+    $\muY(1), \dotsc,\muY(k)$ are all distinct.
+
+  \item The training set contains at least one member of each cluster.
+
+  \item The test set contains at least one member of each cluster.
+
+  \item Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, k \}$.
+    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
+    $i = 1, \dotsc, n+m$.
+
 \end{enumerate}
@@
 \begin{proposition}
-	If assumptions $1, 2, 3, 5$ hold, then there exists a classification method works perfectly in case of ($k \leq K$): the re-substitution error is $0$.
-That is, for $\forall x_i \in \mathbf{X}_{train}$ prediction on $x^L_i$ always correctly predict cluster where $x^R_i$ is in.
+If assumptions $1, 2, 3, 5$ hold, then there exists a classification method
+works perfectly in case of ($k \leq K$): the re-substitution error is $0$.
+That is, for $\forall x_i \in \mathbf{X}_{train}$ prediction on $x^L_i$ always
+correctly predict cluster where $x^R_i$ is in.
 \end{proposition}
+\begin{proof}
+To train classifier $\mathcal{C}$, we have training set $\left(\mathbf{X}^L_{train},\mathcal{G} \right) = \{\left(x^L_i,\hat{g} \left(x^R_i\right)\right), i =1,2,...,m\}$.
+By assumption $5$, $\forall i$ $x_i = \mu_j$, $j=1,2,...,K$ and $k \leq K$.
+Hence the training set has $K$ unique pairs $\{\left(\mu^L_j, \hat{g} \left(\mu^R_j\right)\right),j=1,2,...,K\}$, with $K$ unique $\mu^L_j$ by assumption $3$ and $k$ unique $\hat{g} \left(\mu^R_j\right)$.
+This is because each pair $(\mu^L_j,\mu^R_j)$ is unique by assumption $2$ and $\left(\mu^R_j,\hat{g} \left(\mu^R_j\right)\right)$ is unique since K-means assigns one and only one label to each $\mu^R_j$.
+Define the classifier $\mathcal{C}$ as following: \[ \mathcal{C}(x^L_i) = \sum^K_{j=1}\hat{g} \left(\mu^R_j\right) \cdot I_{\{c_i = j\}}  \] It's easy to see such classifier has re-substitution error $0$.
+\end{proof}
 
@@
 
-\begin{itemize}
-	\item Proof of \textbf{Proposition} $1$.
-	\begin{proof}
-	To train classifier $\mathcal{C}$, we have training set $\left(\mathbf{X}^L_{train},\mathcal{G} \right) = \{\left(x^L_i,\hat{g} \left(x^R_i\right)\right), i =1,2,...,m\}$.
-By assumption $5$, $\forall i$ $x_i = \mu_j$, $j=1,2,...,K$ and $k \leq K$.
-Hence the training set has $K$ unique pairs $\{\left(\mu^L_j, \hat{g} \left(\mu^R_j\right)\right),j=1,2,...,K\}$, with $K$ unique $\mu^L_j$ by assumption $3$ and $k$ unique $\hat{g} \left(\mu^R_j\right)$.
-This is because each pair $(\mu^L_j,\mu^R_j)$ is unique by assumption $2$ and $\left(\mu^R_j,\hat{g} \left(\mu^R_j\right)\right)$ is unique since K-means assigns one and only one label to each $\mu^R_j$.
-Define the classifier $\mathcal{C}$ as following: \[ \mathcal{C}(x^L_i) = \sum^K_{j=1}\hat{g} \left(\mu^R_j\right) \cdot I_{\{c_i = j\}}  \] It's easy to see such classifier has re-substitution error $0$.
-	\end{proof}
-	\item Proof of \textbf{Proposition} $1$ when the classifier is LDA.
-	\begin{proof}
-	Since Proposition $1$ is only used in Lemma $2$, therefore it's sufficient to prove it in case of $k = K$.
-Note that training set contains $K$ unique pairs $\{\left(\mu^L_j, \hat{g} \left(\mu^R_j\right)\right), j=1,2,...,K\}$, with $K$ unique $\mu^L_j$ and $K$ unique $\hat{g} \left(\mu^R_j\right)$ now, where $\mu^L_j, \hat{g} \left(\mu^R_j\right)$ has one-to-one correspondence, i.e.
-\[ \left\lbrace\left(\mu^L_j,\hat{g} \left(\mu^R_j\right) \right), j=1,2,...,K\right\rbrace =  \left\lbrace\left(\mu^L_j, g_j \right), j=1,2,...,K\right\rbrace \]
-By equation $(3)$.\\\\
-Let the prior probability of class $g_j$ be $\pi_j, \hspace{0.05in} j=1,...,K$.
-By assumption $3$, $\pi_j > 0$.
-The class-conditional density of $X$ in class $\mathcal{G} = g_j$ is $$f_j\left(x_i\right) = \mathbb{P}\left( X=x_i | \mathcal{G} = g_j \right) $$ where $x_i \in \{\mu^L_1,...,\mu^L_K\}$ and $f_j\left(x_i\right)$ has multivariate Gaussian distribution
-\[ f_j\left(x_i\right) = \frac{1}{(2\pi)^{p/2}|\Sigma_j|^{1/2}}e^{-\frac{1}{2}(x_i-\mu_j)^T\Sigma^{-1}_j(x_i-\mu_j)	}\]
-LDA assumes a commom covariance matrix $\Sigma_j = \Sigma \hspace{0.1in} \forall j$.
-The posterior probability is 
-\[	\mathbb{P}\left( \mathcal{G} = g_j | X=x_i  \right)  = \frac{f_j\left(x_i\right) \pi_j}{\sum^K_{l=1}f_l\left(x_i\right) \pi_l}	\]
-In practice, $\pi_j, \mu_j$ and $\Sigma$ are estimated from data by
- \begin{itemize}
- 	\item  $\hat{\pi}_j = \frac{N_j}{N}$, where $N_j$ is the number of class-$g_j$ observation in training set and $N$ is the total observation in training set
- 	\item  $\hat{\mu}_j = \sum_{c_i=j} x_i/N_j$ 
- 	\item  $\hat{\Sigma} = \frac{1}{N-K} \sum^K_{j=1}\sum_{c_i=j} (x_i - \hat{\mu}_j )(x_i - \hat{\mu}_j )^T$
- \end{itemize}
-In our case, $\hat{\mu}_j = \mu^L_j$ and $\hat{\Sigma}$ is zero matrix by construction.
-But in real case, the estimate of within-group covariance $\hat{\Sigma}$ is a positive-definite matrix.
-We therefore assume $\hat{\Sigma}$ to be any positive-definite matrix.
-Also, we assume the observations in training set are evenly distributed across all class, i.e.
-$\pi_j$ is a positive constant for all $j = 1,...,K$ \\\\
-LDA predicts class for each $x_i = \mu^L(c_i)\in \{\mu^L_1,...,\mu^L_K\}$ by maximizing the posterior probability
-\begin{align}
-     \hat{\mathcal{G}}(x_i) & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} \mathbb{P}\left( \mathcal{G} = g_j | X=x_i  \right)  \nonumber   \\ 
-     & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} f_j\left(x_i\right) \pi_j \nonumber \nonumber  \\
-     & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} f_j\left(x_i\right) \nonumber  \nonumber  \\    
-     & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in}  \frac{1}{(2\pi)^{p/2}|\hat{\Sigma}|^{1/2}}e^{-\frac{1}{2}(x_i-\mu^L_j)^T\hat{\Sigma}^{-1}(x_i-\mu^L_j)	} \nonumber  \\  
-     &=  \underset{g_j}{\operatorname{argmin}}\hspace{0.05 in} (x_i-\mu^L_j)^T\hat{\Sigma}^{-1}(x_i-\mu^L_j) \nonumber   \\  
-     &= \sum^K_{j=1} g_j \cdot I_{c_i = j}
-\end{align}
-The last equation holds because $\hat{\Sigma}$ is assumed to be positive-definite, so is its inverse  $\hat{\Sigma}^{-1}$.
-Hence $(x_i-\mu^L_j)^T\hat{\Sigma}^{-1}(x_i-\mu^L_j) >0$ for any non-zero vector $(x_i-\mu^L_j)$.
-The lower bound $0$ is attained only when $(x_i-\mu^L_j)$ is a zero vector, which is only possible when $x_i=\mu^L(c_i) = \mu^L_j$, i.e.
-$c_i = j$ by assumption $2$ and $5$.
-\\\\
-Because the RHS of equation $(4)$ coincide with RHS of equation $(3)$, the proof is complete.
-\end{proof}
-\end{itemize}
-
 \clearpage
@@
 \begin{theorem}
-For each $n$, let $N(n)$ be a nonnegative integer with $N(n) \to \infty$ as $n
-\to \infty$, and let $\{ (X_{n,i}, Y_{n,i}) \}_{i=1}^{n+N(n)}$ be independent
+For each $n$, let $m(n)$ be a nonnegative integer with $m(n) \to \infty$ as $n
+\to \infty$, and let $\{ (X_{n,i}, Y_{n,i}) \}_{i=1}^{n+m(n)}$ be independent
 identically distributed random variables.  For any fixed $k$, let $\CV(k)$
@@
 loss of generality suppose that $\bmuY_1$ is the closest cluster center to
-$\muY$.
+$\muY$. [TODO: finish proof]
 \end{proof}

### 2015-06-23 cvclust.tex
revise self-consistency lemma 1

@@
 $(r,s)$ treats the $r$th row subset as ``test'', and the $s$th column subset
-as ``response''.  Following \citet{owen2009bi}, we take $K = L = 2$.  For the
-number of clusters, we select the value of $k$ that minimizes the average of
-$\CV(k)$ over all $K \times L$ folds.
+as ``response''.  We typically take $K = 5$ and $L = 2$.  For the number of
+clusters, we select the value of $k$ that minimizes the average of $\CV(k)$
+over all $K \times L$ folds (choosing the smallest value of $k$ in the event
+of a tie).
+
+\clearpage
 
@@
 vector~$Y$.  The $N$ observations have been divided into two sets: $n$ train
-observations and $m$ test observations.
-
-We make the following assumptions:
-\begin{enumerate}[label={\arabic*.},leftmargin=*]
-
-  \item There are $k$ true centers $\mu(1), \dotsc,\mu(k)$, with the $g$th
+observations and $m$ test observations.  The following theorem gives
+conditions for Gabriel CV to recover the true number of clusters in the
+absence of noise.
+
+\begin{theorem}\label{thm:self-consistency}
+Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
+cross-validation.  For any $k$, let $CV(k)$ be the cross-validation error for
+this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
+Suppose that
+\begin{enumerate}[label=(\roman*)]
+
+  \item There are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with the $g$th
     cluster center partitioned as $\mu(g) = \bigl(\muX(g), \muY(g)\bigr)$ with
-    $\muX(g) \in \R^p$ and $\muY(g) \in \R^q$ for $g = 1, \dotsc, k$.
-
-  \item The vectors $\muX(1), \dotsc,\muX(k)$ are all distinct; likewise,
-    $\muY(1), \dotsc,\muY(k)$ are all distinct.
-
-  \item The training set contains at least one member of each cluster.
-
-  \item The test set contains at least one member of each cluster.
-
-  \item Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, k \}$.
+    $\muX(g) \in \R^p$ and $\muY(g) \in \R^q$ for $g = 1, \dotsc, K$.
+
+  \item The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct; likewise,
+    $\muY(1), \dotsc,\muY(K)$ are all distinct.
+
+  \item \label{asn:self-consistency-noiseless}
+    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
     There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
@@
 
+  \item The training set contains at least one member of each cluster:
+    for all $g$ in the range $1, \dotsc, K$, there exists at least one $i$ in
+    the range $1, \dotsc, n$ such that $G_i = g$.
+
+  \item \label{asn:self-consistency-test}
+    The test set contains at least one member of each cluster: for all $g$ in
+    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
+    \dotsc, n+m$ such that $G_i = g$.
+
 \end{enumerate}
+Then $CV(k) < CV(K)$ for $k < K$, and $CV(k) = CV(K)$ for $k > K$, so that
+Gabriel CV correctly chooses $k = K$.
+\end{theorem}
 
@@
 \begin{proof}
-To train classifier $\mathcal{C}$, we have training set $\left(\mathbf{X}^L_{train},\mathcal{G} \right) = \{\left(x^L_i,\hat{g} \left(x^R_i\right)\right), i =1,2,...,m\}$.
-By assumption $5$, $\forall i$ $x_i = \mu_j$, $j=1,2,...,K$ and $k \leq K$.
-Hence the training set has $K$ unique pairs $\{\left(\mu^L_j, \hat{g} \left(\mu^R_j\right)\right),j=1,2,...,K\}$, with $K$ unique $\mu^L_j$ by assumption $3$ and $k$ unique $\hat{g} \left(\mu^R_j\right)$.
-This is because each pair $(\mu^L_j,\mu^R_j)$ is unique by assumption $2$ and $\left(\mu^R_j,\hat{g} \left(\mu^R_j\right)\right)$ is unique since K-means assigns one and only one label to each $\mu^R_j$.
-Define the classifier $\mathcal{C}$ as following: \[ \mathcal{C}(x^L_i) = \sum^K_{j=1}\hat{g} \left(\mu^R_j\right) \cdot I_{\{c_i = j\}}  \] It's easy to see such classifier has re-substitution error $0$.
-\end{proof}
-
-\begin{theorem}
-Given assumptions $1-5$, $ E\left(Error_k\right) > 0$ for $k<K$ and  $ E\left(Error_k\right) = 0$ for $k=K$. Hence if $k \leq K$, the Gabriel algorithm selects $k=K$ with probability $1$.
-\end{theorem}
-\noindent
-Not that $k > K$ is not considered here because there are only $K$ unique points in $\mathbf{X}$.
-K-means will not work for parameter which is larger than the number of unique points in data $\mathbf{X}$.\\\\
-\noindent
+To train classifier $\mathcal{C}$, we have training set
+$\left(\mathbf{X}^L_{train},\mathcal{G} \right) = \{\left(x^L_i,\hat{g}
+\left(x^R_i\right)\right), i =1,2,...,m\}$. By assumption $5$, $\forall i$
+$x_i = \mu_j$, $j=1,2,...,K$ and $k \leq K$. Hence the training set has $K$
+unique pairs $\{\left(\mu^L_j, \hat{g}
+\left(\mu^R_j\right)\right),j=1,2,...,K\}$, with $K$ unique $\mu^L_j$ by
+assumption $3$ and $k$ unique $\hat{g} \left(\mu^R_j\right)$. This is because
+each pair $(\mu^L_j,\mu^R_j)$ is unique by assumption $2$ and
+$\left(\mu^R_j,\hat{g} \left(\mu^R_j\right)\right)$ is unique since K-means
+assigns one and only one label to each $\mu^R_j$. Define the classifier
+$\mathcal{C}$ as following: \[ \mathcal{C}(x^L_i) = \sum^K_{j=1}\hat{g}
+\left(\mu^R_j\right) \cdot I_{\{c_i = j\}}  \] It's easy to see such
+classifier has re-substitution error $0$.
+\end{proof}
+
 The proof of above theorem follows from following two lemmas:
 \begin{lemma}
-	If the assumptions of Theorem $1$ hold, then when $k<K$, the CV Error $>0$
+Suppose that the assumptions of Theorem $1$ are in force.  If $k < K$, then
+$\CV(k) > 0$.
 \end{lemma}
 \begin{proof}
-	It is sufficient to prove that when $k<K$,  $Error_k>0$. By the definition of $Error_k$,
-\begin{equation}
-Error_k = \sum^k_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2
-\end{equation}
-	where $\hat{\mu}^R \left(g_j\right)$ is the center of cluster $g_j$ returned from K-means.
-By assumption $4$, test set contains at least one member of each cluster.
-Since there are $K$ true centers, and by assumption $5$, $x_i = \mu \left(c_i\right) \in \{\mu_1,...,\mu_K\} $, test set contains $K$ centers.
-It means there are $K$ distinct $x^R_i \left(=\mu^R \left(c_i\right)\right)$ in $\mathbf{X}^R_{test}$ set by assumption $2$.
-However, in equation $(1)$, there are only $k$ distinct $\hat{\mu}^R \left(g_j\right)$.
-Because $k < K$, it means not every element $||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 = 0$ in equation $(1)$.
-Therefore,  $$Error_k = \sum^k_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 > 0$$
-\end{proof}
+By definition,
+\[
+  \CV(k)
+    =
+      \sum_{i=n+1}^{n+m}
+        \| Y_i - \bmuY (\hGX_i) \|^2,
+\]
+where $\bmuY(g)$ is the center of cluster $g$ returned from applying $k$-means
+to $Y_1, \dotsc, Y_n$.  Assumptions~\ref{asn:self-consistency-noiseless}
+and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
+set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all $K$ distinct values
+in $\{ \muY(1), \dotsc, \muY(K) \}$.  However, there are only $k$ distinct
+values of $\bmuY(g)$.  Thus, at least one summand
+\(
+  \| Y_i - \bmuY(\hGX_i) \|^2
+\)
+is nonzero.  Therefore,
+\(
+  \CV(k) > 0.
+\)
+\end{proof}
+
 \begin{lemma}

### 2015-06-23 cvclust.tex
finish revising self-consistency proof

@@
 
-\clearpage
-
 \section{Self-Consistency of Gabriel CV method}
@@
 \begin{theorem}\label{thm:self-consistency}
+
 Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
@@
 this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
-Suppose that
+We will assume that there are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with
+the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
+\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
 \begin{enumerate}[label=(\roman*)]
-
-  \item There are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with the $g$th
-    cluster center partitioned as $\mu(g) = \bigl(\muX(g), \muY(g)\bigr)$ with
-    $\muX(g) \in \R^p$ and $\muY(g) \in \R^q$ for $g = 1, \dotsc, K$.
-
-  \item The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct; likewise,
-    $\muY(1), \dotsc,\muY(K)$ are all distinct.
-
   \item \label{asn:self-consistency-noiseless}
@@
 
-  \item The training set contains at least one member of each cluster:
-    for all $g$ in the range $1, \dotsc, K$, there exists at least one $i$ in
-    the range $1, \dotsc, n$ such that $G_i = g$.
+  \item \label{asn:self-consistency-distinct-mux}
+    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
+
+  \item \label{asn:self-consistency-distinct-muy}
+    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
+
+  \item \label{asn:self-consistency-train}
+    The training set contains at least one member of each cluster: for all $g$
+    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
+    $1, \dotsc, n$ such that $G_i = g$.
 
@@
 
-\begin{proposition}
-If assumptions $1, 2, 3, 5$ hold, then there exists a classification method
-works perfectly in case of ($k \leq K$): the re-substitution error is $0$.
-That is, for $\forall x_i \in \mathbf{X}_{train}$ prediction on $x^L_i$ always
-correctly predict cluster where $x^R_i$ is in.
-\end{proposition}
-\begin{proof}
-To train classifier $\mathcal{C}$, we have training set
-$\left(\mathbf{X}^L_{train},\mathcal{G} \right) = \{\left(x^L_i,\hat{g}
-\left(x^R_i\right)\right), i =1,2,...,m\}$. By assumption $5$, $\forall i$
-$x_i = \mu_j$, $j=1,2,...,K$ and $k \leq K$. Hence the training set has $K$
-unique pairs $\{\left(\mu^L_j, \hat{g}
-\left(\mu^R_j\right)\right),j=1,2,...,K\}$, with $K$ unique $\mu^L_j$ by
-assumption $3$ and $k$ unique $\hat{g} \left(\mu^R_j\right)$. This is because
-each pair $(\mu^L_j,\mu^R_j)$ is unique by assumption $2$ and
-$\left(\mu^R_j,\hat{g} \left(\mu^R_j\right)\right)$ is unique since K-means
-assigns one and only one label to each $\mu^R_j$. Define the classifier
-$\mathcal{C}$ as following: \[ \mathcal{C}(x^L_i) = \sum^K_{j=1}\hat{g}
-\left(\mu^R_j\right) \cdot I_{\{c_i = j\}}  \] It's easy to see such
-classifier has re-substitution error $0$.
-\end{proof}
-
-The proof of above theorem follows from following two lemmas:
+This theorem is implied by the following two lemmas.
+
 \begin{lemma}
-Suppose that the assumptions of Theorem $1$ are in force.  If $k < K$, then
-$\CV(k) > 0$.
+Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
+force.  If $k < K$, then $\CV(k) > 0$.
 \end{lemma}
@@
 and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
-set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all $K$ distinct values
-in $\{ \muY(1), \dotsc, \muY(K) \}$.  However, there are only $k$ distinct
-values of $\bmuY(g)$.  Thus, at least one summand
+set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all distinct values in
+$\{ \muY(1), \dotsc, \muY(K) \}$.
+Assumption~\ref{asn:self-consistency-distinct-muy} implies that there are
+exactly $K$ such distinct values.  However, there are only $k$ distinct values
+of $\bmuY(g)$.  Thus, at least one summand
 \(
@@
 \begin{lemma}
-	If the assumptions of Theorem $1$ hold, when $k=K$, the CV Error $=0$
+Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
+force.  If $k \geq K$, then $\CV(k) = 0$.
 \end{lemma}
 \begin{proof}
-	It's sufficient to show that when $k=K$, $Error_k=0$.
-By assumption $3$ and $5$, training set contains $K$ centers.
-Which also implies $\mathbf{X}^R_{train}$ contains $K$ distinct $x^R_i$ $(\mu^R_1,...,\mu^R_K)$.
-When apply K-means on $\mathbf{X}^R_{train}$ with parameter $K$, the unique solution of
-	\[ \underset{\left(\hat{\mu}^R\left(g_1\right),...,\hat{\mu}^R\left(g_K\right)\right)}{\operatorname{argmin}} \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^R_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 \]
-	is 
-\begin{equation}	
-	\{\hat{\mu}^R \left(g_j\right) = \mu^R_j, j=1,...,K\}
-\end{equation}	
-with 
-\begin{equation}	 
-	\hat{g} \left(x^R_i\right) = \sum^K_{j=1} g_j \cdot I_{c_i = j}
-\end{equation}	
-Where $\hat{g} \left(x^R_i\right) \in  \mathcal{G}$ is the assigned cluster of $x^R_i$ by K-means.
-One can easily check that such solution is unique and above objective function attains its lower bound $0$.\\\\
-By Proposition $1$, $\mathcal{C}$ has re-substitution error $0$, which means its prediction error is $0$ on $\mathbf{X}^L_{test}$ as $\forall x^L_i \in \mathbf{X}^L_{test}, x^L_i \in \mathbf{X}^L_{train}$ by assumption $1,3,5$.
-Hence the predicted cluster $\hat{g} \left(x^L_i\right)$ for $x_i$ is the correct one $\hat{g} \left(x^R_i\right)$, which determined by $x^R_i$.
-By equation $(3)$, we have $x^R_i = \mu^R \left(c_i\right) = \mu^R_j$ if $\hat{g} \left(x^L_i\right) = \hat{g} \left(x^R_i\right) = g_j$.
-Therefore,
-\[ Error_K = \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 =  \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - \mu^R_j||^2= 0\]
-The last equation holds because of equation $(2)$.
-\end{proof}
-
-\clearpage
+From assumptions~\ref{asn:self-consistency-noiseless},
+\ref{asn:self-consistency-distinct-muy},
+and~\ref{asn:self-consistency-train}, we know the cluster centers
+gotten from applying $k$-means to $Y_1, \dotsc, Y_n$ must include
+$\muY(1), \dotsc, \muY(K)$.  Without loss of generality, suppose that
+$\bmuY(g) = \muY(g)$ for $g = 1, \dotsc, K$.  This implies that
+$\hGY_i = G_i$ for $i = 1, \dotsc, n$.  Thus, employing
+assumption~\ref{asn:self-consistency-noiseless} again, we get that
+$\bmuX(g) = \muX(g)$ for $g = 1, \dotsc, K$.
+
+Since assumption~\ref{asn:self-consistency-distinct-mux} ensures that
+$\muX(1), \dotsc, \muX(K)$ are all distinct, we must have that $\hGX_i = G_i$
+for all $i = 1, \dotsc, m+n$.  In particular, this implies that $\bmuY(\hGX_i)
+= Y_i$ for $i = 1, \dotsc, m+n$, so that $\CV(k) = 0$.
+\end{proof}
 

### 2015-06-23 cvclust.tex
fix paragraph spacing in introduction

@@
 Clustering is an important topic in unsupervised learning because it can reveal the internal structure of data through grouping, segment the data through partitioning and summarize data for other purposes such as dimension reduction.
-It has being widely used in various fields such as psychology, biology, statistics and machine learning including pattern recognition, image segmentation etc.\\
+It has being widely used in various fields such as psychology, biology, statistics and machine learning including pattern recognition, image segmentation etc.
 
@@
 \cite{chiang2010intelligent} provides a nice review of existing methods for finding the right $k$ in published literature.
-\\
 
@@
 Therefore the well understood properties of cross-validation procedure in supervised learning cannot be carried over to the prediction strength method, makes the interpretation of its result difficult.
-\\
 

### 2015-06-23 cvclust.tex
minor formatting edits
cut out old well-separated cluster proof; simplify simulation results table

@@
 \usepackage{multirow}
+\usepackage{booktabs}
+\usepackage{longtable}
 
@@
 \newcommand{\CV}{\operatorname{CV}}
+\newcommand{\PE}{\operatorname{PE}}
 \newcommand{\E}{\operatorname{E}}
@@
 
-\section{Asymptotic Analysis of $k$-Means}
+\section{Asymptotic Analysis of $k$-means}
 
@@
 
+\subsection{Single Cluster}
+
 Here, we show that Gabriel Cross-Validation is consistent when the data come
@@
 
+\subsection{Well-Separated Clusters with Noise}
+
+\section{Simulation}
+
+In this section, simulation is performed to evaluate the performance of our
+proposed methods in locating the ``correct" number of clusters. We compare
+with a basket of existing methods including Gap statistics
+\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
+\citep{fraley2002model}, CH-index \citep{calinski1974dendrite}, Hartigan
+statistics \citep{hartigan1975clustering}, Jump method
+\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
+Bootstrap stability \citep{fang2012selection} in following simulation
+settings. We select $L=R$ and $5-$fold cross-validation in row
+($m=\frac{4}{5}n$) as default parameter setting for our proposed Gabriel
+method. Note that set $L=R$ corresponding to $2-$fold cross-validation in
+column.
+\begin{enumerate}
+
+  \item Single cluster in $10$ dimensions --- $200$ observations, each
+    observation uniformly distributed over $[0,1]$ in each dimension.
+
+  \item Two clusters in $4$ dimensions --- $50$ i.i.d standard normal
+    observations are generated, i.e. generated from multivariate normal
+    $\mathcal{N}\left(\boldsymbol\mu,\mathbf{I}\right)$ for each cluster, with
+    cluster center $\boldsymbol\mu = (1,0,0,1)$ and $(1,3.5,3.5,1)$
+    respectively
+
+  \item Four clusters in $100$ dimensions ---  Each cluster has $100$ or $150$
+    i.i.d standard normal observations, with cluster centers randomly
+    generated from multivariate normal distribution
+    $\mathcal{N}\left(\mathbf{0},0.65^2\mathbf{I}\right)$
+
+  \item Ten clusters in $100$ dimensions ---  Each cluster has $50$ or $100$
+    i.i.d standard normal observations, with cluster centers randomly
+    generated from multivariate normal distribution
+    $\mathcal{N}\left(\mathbf{0},0.72^2\mathbf{I}\right)$
+
+  \item Four log-normal clusters in $16$ dimensions --- For each cluster, $30$
+    or $60$ i.i.d centered log-normal observations are generated from
+    $ln\mathcal{N}\left(0,0.5^2 \right)$. The cluster centers are randomly
+    generated from multivariate normal distribution
+    $\mathcal{N}\left(\mathbf{0},1.2^2\mathbf{I}\right)$
+
+  \item[6.] Three exponential clusters in $20$ dimensions with different
+    variance --- $40$ observations in each cluster are generated from centered
+    exponential distribution ($exp(\lambda)-1/\lambda$ in each dimension),
+    with $\boldsymbol\mu$ randomly generated from
+    $\mathcal{N}\left(\mathbf{0},19\mathbf{I}\right)$ and $\lambda = 1, 1/2,
+    1/5$ respectively.
+\end{enumerate}
+Note that in setting 2--6, all clusters are well-separated, i.e. no
+overlapping. In fact, any simulated clusters with minimum distance less than
+1 unit was discarded, so there is clear definition of true number of
+clusters. The parameters in setting 2--6 are chosen such that about half of
+the random realization were discarded. The idea is borrowed from
+\cite{tibshirani2001estimating}.
+
+Besides the number of clusters selected by each algorithm is returned, we also
+give the prediction error (PE) for each algorithm in each setting. The PE is
+defined by 
+\[
+  \PE(k)
+    =
+      \frac{\|\hat{\mathbf{X}}_k- \boldsymbol{\mu} \|_F}
+           {PE_{\text{oracle}}}
+\]
+where
+\[
+  \PE_{\text{oracle}} = \min_k \|\hat{\mathbf{X}}_k- \boldsymbol{\mu} \|_F
+\]
+Matrix $\boldsymbol{\mu}$ consists of true centers of observations in data
+matrix $\mathbf{X}$, while $\hat{\mathbf{X}}_k$ is obtained by replacing each
+observation in $\mathbf{X}$ by its closest cluster center result from applying
+K-means on $\mathbf{X}$ with parameter $k$. Here $k$ is the optimal number of
+clusters selected by each algorithm on data $\mathbf{X}$. This is only
+possible on simulated data because otherwise one would never know the true
+center matrix~$\boldsymbol{\mu}$.
+
+The oracle selected number for each data $\mathbf{X}$ is also given on the
+table, which is defined by 
+\[
+  K_{\text{oracle}}
+    =
+      \argmin_{k} \|\hat{\mathbf{X}}_k - \boldsymbol{\mu} \|_F
+\]
+
 \clearpage
 
-\subsection{Well-separated clusters with noise}
-
-Same with previous, the data set $\mathbf{X} \in \R^{n \times p}$ been
-partitioned as $\mathbf{X}^L \in \R^{n \times L}$ and $\mathbf{X}^R
-\in \R^{n \times R}$, with following set of assumptions:
-
-\begin{enumerate}[label={Assumption},leftmargin=*]
-	\item $1.$ There are $K$ true centers $(\mu_1,...,\mu_K)$ where $\mathbf{X}$ generated from
-	\item $2.$ The $K$ partitioned centers $(\mu^L_1,...,\mu^L_K)$ are all distinct in $L$ dimensions: $||\mu^L_i-\mu^L_j|| \neq 0$ for any $i \neq j$; $(\mu^R_1,...,\mu^R_K)$ are all distinct in $R$ dimensions
-	\item $3.$ There is noise, i.e.
-$$x_i = \mu \left(c_i\right) + \varepsilon_i, \hspace{0.2 in} \mu \left(c_i\right) \in \{\mu_1,...,\mu_K\}$$ where vector $\varepsilon_i$ is i.i.d with zero mean and covariance matrix $\Sigma$, which is independent from any $\mu_j \in (\mu_1,...,\mu_K)$, $c_i$ is the true cluster for observation $i$, which implies $$x^L_i = \mu^L \left(c_i\right) + \varepsilon^L_i,\hspace{0.2in} x^R_i = \mu^R \left(c_i\right) + \varepsilon^R_i$$
-with $$\mu^L\left(c_i\right) \in \{\mu^L_1,...,\mu^L_K\},\hspace{0.2in} \mu^R\left(c_i\right) \in \{\mu^R_1,...,\mu^R_K\} $$\
-where $\varepsilon^L_i$ and $\varepsilon^R_i$ are independent and both have zero mean and covariance matrix $\Sigma^L$ and $\Sigma^R$ respectively, so $\Sigma = 
-\begin{pmatrix}
-  \Sigma^L &0\\
-  0  &\Sigma^R\\
-  \end{pmatrix}$
-. Also $\varepsilon^R_i$ is bounded $||\varepsilon^R_i|| \leq 1$.
-	\item $4.$ Clusters are well separated, i.e.
-\[ ||x_i-\mu \left(c_i\right)|| < ||x_i-\mu_{j}|| \hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i \] and \[ ||x^L_i-\mu^L \left(c_i\right)|| < ||x^L_i-\mu^L_{j}||-\delta,  \hspace{0.2 in}||x^R_i-\mu^R \left(c_i\right)|| < ||x^R_i-\mu^R_{j}||-\delta \hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i\] where $\delta > 0$ is a small constant
-	\item $5.$ The training set contains $n^{train}_j>0$ observations from each cluster $j$ and the test set contains $n^{test}_j > 0$ observations from each cluster $j$, $j=1,2...,K$
-	\item $6.$ The prior for $x_i$ come from cluster $\mu_j$ is $\pi_j > 0$.
-So when the total number of observation in training set $m$ goes infinity, $\lim_{m\to \infty} n^{train}_j = m \cdot \pi_j $.
-Also $m=n\cdot r$, $n-m = n\cdot (1-r)$, where $0<r<1$ is a positive constant.
-\end{enumerate}
-
-\subsubsection{Consistent assumption for K-means}
-\cite{pollard1981strong} found conditions that ensure the almost sure convergence of the set of cluster centers of K-means as the sample size increase.
-This section mainly introduce the theorem from \cite{pollard1981strong} and show the assumptions needed for our own theorem.\\\\
-Applying K-means clustering on i.i.d observations $x_1,x_2,...x_m$ from population $P$ with parameter $k$ is to choose cluster centers $a_1,a_2,...,a_k$ to minimize
-\[	W_m = \frac{1}{m}\sum^m_{l=1}min_{1\leq j \leq k}||x_l-a_j||^2	\]
-It's convenient to treat $W_m$ as a function of the set of cluster centers $A$ and of the  empirical distribution $P_m$ obtained from the sample by placing mass $m^{-1}$ at each of $x_1,x_2,...x_m$.
-Then the problem is to minimize
-\[ W(A,P_m):=\int min_{a\in A}||x-a||^2P_m(dx)\]
-Over all possible choices of the set $A$ containing $k$ points. \\\\
-For each probability measure $Q$ and each set $A$, defined
-\[	\Phi(A,Q):= \int min_{a\in A}||x-a||^2 Q(dx)	\] and \[ \mathfrak{m}_k(Q):=inf\{\Phi(A,Q):A \hspace{0.05in} \text{contains} \hspace{0.05in} k  \hspace{0.05in} \text{or fewer points}\} \]
-The theorem proved by \cite{pollard1981strong} says that
-\begin{theorem}
-	Suppose that $\int ||x||^2P(dx) < \infty$ and that for each $j=1,2,...,k$ there is a unique set $\bar{A}(j)$ for which $\Phi(\bar{A}(j),P)=\mathfrak{m}_j(P)$.
-Then $A_m \to \bar{A}(k)$ a.s and $\Phi(A_m,P_m) \to \mathfrak{m}_k(P)$ a.s
-\end{theorem}
+\begin{center}
+{%\footnotesize
+\begin{longtable}{lcccccccccccccccc}
+\caption{\label{table1} Simulation Results}\\
+\toprule
+\phantom{Pred.~Strength}
+& \multicolumn{16}{c}{Estimated Number of Clusters} \\
+\cmidrule(lr){2-17}
+Method
+  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15
+  & \textsc{na} \\
+\midrule
+\endfirsthead
+\multicolumn{17}{c}%
+{\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
+\toprule
+\phantom{Pred.~Strength}
+  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
+\cmidrule(lr){2-17}
+Method
+  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15
+  & \textsc{na} \\
+\midrule
+\endhead
+%\bottomrule
+\multicolumn{17}{r}{\textit{Continued on next page}} \\
+\endfoot
+\bottomrule
+\multicolumn{17}{l}{
+{\footnotesize {In Setting~1, ``\textendash'' means the method can not be used
+with parameter $k=1$}.}}
+\endlastfoot
+\input{demo/bench/setting1/results.tex}
+\midrule
+\input{demo/bench/setting2/results.tex}
+\midrule
+\input{demo/bench/setting3/results.tex}
+\midrule
+\input{demo/bench/setting4/results.tex}
+\midrule
+\input{demo/bench/setting5/results.tex}
+\midrule
+\input{demo/bench/setting6/results.tex}
+\end{longtable}
+}
+\end{center}
+
 \noindent
-Where $A_m = A_m(k)$ is the set of optimal sample cluster centers satisfy $\Phi(A_m,P_m) = \mathfrak{m}_k(P_m)$ and $\bar{A}=\bar{A}(k)$ is the set of optimal population cluster center satisfy  $\Phi(\bar{A},P) = \mathfrak{m}_k(P)$.
-Such theorem gives the condition for the sample estimator of cluster centers be consistent estimator of optimal cluster centers.\\
-
-In our setting, observations $x^R_1, x^R_2,...,x^R_m$ can be viewed as i.i.d observations from same distribution $\mathcal{P}$, which is a mixed distribution of $K$ distinct distributions with prior $\pi_1,\pi_2,...,\pi_K$.
-\begin{theorem}
-Suppose assumption $3$ holds and for each $j=1,2,...,k_{max}$ there is a unique set $\bar{A}(j)$ for which $\Phi(\bar{A}(j),\mathcal{P})=\mathfrak{m}_j(\mathcal{P})$, also $ \bar{A}(K) = \{\mu^R_1,\mu^R_2,...,\mu^R_K\}$.
-Then for any $k \leq k_{max}$, $A_m(k) \to \bar{A}(k)$ a.s.
-\end{theorem}
-\begin{proof}
-By assumption $3$, $||\varepsilon^R_i||\leq 1$, which implies the first condition in theorem $2$ holds, i.e.
-\[\int ||x||^2P(dx) < \infty\] 
-Since the second condition of theorem $2$ is given as assumption in theorem $3$, we have almost sure convergence of cluster centers $A_m(k)$ for any $k \leq k_{max}$ by theorem $2$. 
-\end{proof}
-\noindent
-In the rest of this section, the assumptions (conditions) given in theorem $3$ is called consistent assumptions. 
-\subsubsection{Asymptotic property of Gabriel method}
-\begin{definition}
-Define event\\ $\mathcal{D} = \{\forall x_i \in \mathbf{X}, ||x^L_i-\hat{\mu}^L \left(c_i\right)|| < ||x^L_i-\hat{\mu}^L_{j}||, ||x^R_i-\hat{\mu}^R \left(c_i\right)|| < ||x^R_i-\hat{\mu}^R_{j}|| \hspace{0.2 in} \forall \hspace{0.01 in}j\neq c_i\}$, where $\hat{\mu}^R_{j}$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^R_{train}$and $\hat{\mu}^L_{j}$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^L_{train}$
-\end{definition}
-
-\begin{proposition}
-Under assumptions $1-6$, $P(\mathcal{D}) \to 1$ with $m \to \infty$.
-\end{proposition}
-\begin{proof}
-	By assumption $4$, \[  ||x^L_i-\mu^L \left(c_i\right)|| < ||x^L_i-\mu^L_{j}||-\delta \hspace{0.2in} \forall \hspace{0.01 in}j\neq c_i \] 
-	Hence, 
-	\begin{align} \nonumber
-		||x^L_i-\hat{\mu}^L \left(c_i\right)|| & \leq \nonumber ||x^L_i-\mu^L \left(c_i\right)||+ ||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) ||\\ \nonumber
-		&< ||x^L_i-\mu^L_{j}||-\delta + ||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) ||  \\ \nonumber
-		& \leq ||x^L_i-\hat{\mu}^L_{j}||+||\mu^L_{j}-\hat{\mu}^L_{j}||-\delta + ||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) ||  \\ \nonumber
-		& =  ||x^L_i-\hat{\mu}^L_{j}||-\left\lbrace \delta-||\mu^L_{j}-\hat{\mu}^L_{j}|| -||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) || \right\rbrace \\ 
-m \to \infty		&<||x^L_i-\hat{\mu}^L_{j}||
-	\end{align} 
-The step $(5)$ is because for any given $\varepsilon > 0$, by Markov inequality
-	\begin{align} \nonumber
-	  P\left(||\mu^L_{j}-\hat{\mu}^L_{j}||^2 \geq \varepsilon \right) &\leq \frac{E||\mu^L_{j}-\hat{\mu}^L_{j}||^2}{\varepsilon}  \nonumber \\ \nonumber
-	    &=  \frac{E\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)^T\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)}{\varepsilon}  \\ \nonumber
-	    &=  \frac{E[ tr\left\lbrace\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)^T\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)\right\rbrace]}{\varepsilon}  \\ \nonumber
-	   &=  \frac{E[ tr\left\lbrace\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)^T\right\rbrace]}{\varepsilon}  \\ \nonumber
-	  &=  \frac{ tr\left\lbrace E\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)^T\right\rbrace}{\varepsilon}  \\ \nonumber
-	  &= \frac{ tr\{\Sigma^L \}}{n^{train}_j \cdot \varepsilon}   \\ \nonumber
-	% \lim_{m\to \infty} n^{train}_j = m \cdot \pi_j \to \infty \hspace{1in} &\to 0  \\ \nonumber
-	\end{align} 
-So
-\begin{align} \nonumber
-  max_{j \in \{1,2,...,K\}} \hspace{0.05in} P\left(||\mu^L_{j}-\hat{\mu}^L_{j}||^2 \geq \varepsilon \right) &\leq max_{j \in \{1,2,...,K\}} \hspace{0.05in}\frac{E||\mu^L_{j}-\hat{\mu}^L_{j}||^2}{\varepsilon}  \nonumber \\ \nonumber
-  &=max_{j \in \{1,2,...,K\}} \hspace{0.05in}  \frac{ tr\{\Sigma^L \}}{n^{train}_j \cdot \varepsilon}   \\ \nonumber
-m \to \infty \hspace{0.3in}  &= \frac{ tr\{\Sigma^L \}}{m\hspace{0.02in}\pi^*\hspace{0.02in} \varepsilon}  \hspace{0.5in} \text{where}\hspace{0.1in}\pi^* = min_{j \in \{1,2,...,K\}}\pi_j\\ \nonumber
-&\to 0
-\end{align}
-
-Therefore $\forall j, \hspace{0.05in} ||\mu^L_{j}-\hat{\mu}^L_{j}|| \to 0$ uniformly as $m \to \infty$. Hence, 
-\begin{equation}
-\lim_{m\to \infty} \left\lbrace \delta-||\mu^L_{j}-\hat{\mu}^L_{j}|| -||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) || \right\rbrace >0
-\end{equation}
-
-Same argument yields 
-\[||x^R_i-\hat{\mu}^R \left(c_i\right)|| < ||x^R_i-\hat{\mu}^R_{j}||\]
-Because $(6)$ holds uniformly $\forall j\in\{1,2,...,K\}$, we have 
-\[ \lim_{m \to \infty}P(\mathcal{D})  =1\]
-	\end{proof}
-\begin{proposition}
-Given assumption $1-6$ and the consistent assumptions of Theorem $3$, when $m \to \infty$, $P\left[A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}\right]=1$, where $\hat{\mu}^R_j$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^R_{train}$
-\end{proposition}
-\begin{proof}
-By theorem $3$, we know $A_m(K) \to \bar{A}(K)$ a.s.
-and $\bar{A}(K) = \{\mu^R_1,\mu^R_2,...,\mu^R_K\}$.
-Denote $A_m(K) = \{a_1, a_2,..., a_K\}$, by almost sure convergence of $A_m(K)$ to $\bar{A}(K)$, we have\\\\
-For all $\varepsilon > 0$, there exists $N_{\varepsilon}$ such that if $m > N_{\varepsilon}$, then
-\[ P\left(||a_j-\mu^R_j||< \varepsilon\hspace{0.1in}\forall j\right)=1\]
- Now, take $\varepsilon = \frac{\delta}{2}$, then exists $N_{\delta}$ such that for all $m > N_{\delta}$
-\[P\left(||a_j-\mu^R_j||<\frac{\delta}{2} \hspace{0.1in}\forall j\right)=1\]
-By assumption $4$, 
-\[||x^R_i-\mu^R \left(c_i\right)|| < ||x^R_i-\mu^R_{j}||-\delta \hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i\]
-Let $m > N_{\delta}$, for all $x^R_i \in \mathbf{X}^R_{train}$, with probability $1$
-
-\begin{align}
-  ||x^R_i - a(c_i)||&\leq ||x^R_i-\mu^R(c_i)|| + ||\mu^R(c_i)-a(c_i)||\nonumber \\ \nonumber
-   &< ||x^R_i-\mu^R_{j}||-\delta + ||\mu^R(c_i)-a(c_i)|| \hspace{0.2 in} \forall \hspace{0.01 in}j\neq c_i  \\ \nonumber
-    &\leq ||x^R_i-a_{j}|| + ||a_j - \mu^R_{j}||-\delta + ||\mu^R(c_i)-a(c_i)||  \\ \nonumber
- ||\mu^R(c_i)-a(c_i)||< \frac{\delta}{2} \hspace{0.05in}\& \hspace{0.05in}||a_j - \mu^R_{j}||< \frac{\delta}{2} \hspace{0.2in} &<||x^R_i-a_{j}||\hspace{0.2 in}  \forall \hspace{0.01 in}j\neq c_i 
-\end{align}
-Hence, $x^R_i$ is assigned to cluster center $a(c_i)$ by K-means.
-Note also that observations from same cluster are assigned to the same center.
-Since $a_j$ is the sample mean of all the observations assigned to it, we have
-\[a_j = \hat{\mu}^R_j \hspace{0.1in}\forall j\]
- where $\hat{\mu}^R_j$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^R_{train}$. The proof is complete.
-\end{proof}
-\begin{proposition}
-Conditioning on $\mathcal{D}$ and $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$, there exists a classification method works perfectly in case of $k = K$: the prediction error is $0$.
-That is, $\forall x_i \in \mathbf{X}_{test}$ prediction on $x^L_i$ always correctly predict cluster where $x^R_i$ is in.
-\end{proposition}
-
-\begin{proof}
-Given $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$,  we know from the proof of proposition $3$ that when applying K-means on $\mathbf{X}^R_{train}$ with parameter $K$, the solution of
-\[ \underset{\left(\hat{\mu}^R\left(g_1\right),...,\hat{\mu}^R\left(g_K\right)\right)}{\operatorname{argmin}} \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^R_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 \]
-is 
-\begin{equation}
-\{\hat{\mu}^R\left(g_j\right) = \hat{\mu}^R_{j}, j=1,...,K\}
-\end{equation}
-with 
-\begin{equation}
-\hat{g} \left(x^R_i\right) = \sum^K_{j=1} g_j \cdot I_{c_i = j}
-\end{equation}
-where $\hat{g} \left(x^R_i\right)$ is the assigned cluster by K-means for observation $x^R_i \in \mathbf{X}^R_{train}$  , $\hat{\mu}^R\left(g_j\right)$ is the sample mean (centroid) of K-means returned cluster $g_j$, $\hat{\mu}^R_{j}$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^R_{train}$, $c_i$ is the true cluster for observation $i$.\\\\
-The training set of $\mathcal{C}$ consists of data in the form of 
-\[
-\left(\mathbf{X}^L_{train},\mathcal{G} \right) = \left\lbrace \left(x^L_i,\hat{g} \left(x^R_i\right)\right)\right\rbrace, \hspace{0.2in} i =1,2,...,m
- \]
-From $(8)$, we learned that observations from the same cluster have been clustered together in the training set. 
-And if $g_j = \hat{g} \left(x^R_i\right)$ for observation $i$
-\[x^R_i = \mu^R \left(c_i\right) + \varepsilon^R_i = \mu^R_{j} + \varepsilon^R_i\] 
-Which implies  
-\[x^L_i = \mu^L \left(c_i\right) + \varepsilon^L_i = \mu^L_{j} + \varepsilon^L_i\] 
-Hence training set of $\mathcal{C}$ consists of data in the form of 
-\[ \left(\mathbf{X}^L_{train},\mathcal{G} \right) = \left\lbrace\left(\mu^L_{j} + \varepsilon^L_i, g_j \right)\right\rbrace\hspace{0.2in} j =1,2,...,K, \hspace{0.1in} i=1,...,m\]
-So Let $\hat{\mu}^L_{j} = \frac{\sum x^L_i \cdot I_{c_i=j} }{n^{train}_j}$, $x^L_i  \in \mathbf{X}^L_{train}$ denotes the sample mean of each cluster in $\mathbf{X}^L_{train}$.
-Define following classifier
-	 \[ \mathcal{C}(x^L_i) = \sum^K_{j=1} g_j \cdot I_{h(i) = j} \]
-	 where\[h(i) = \underset{j}{\operatorname{argmin}} \hspace{0.05 in}||x^L_i - \hat{\mu}^L_j|| \]
-	Note here $x^L_i  \in \mathbf{X}^L_{test}$.\\
-
-Then, above classifier has prediction error $0$. To prove it, we need to show above classifier $\mathcal{C}(x^L_i)$ always gives the same cluster as the one in equation $(8)$, i.e. 
-\begin{equation}
-\mathcal{C}(x^L_i) = \sum^K_{j=1} g_j \cdot I_{c_i = j}
-\end{equation}
-Given $\mathcal{D}$, we have
-\[||x^L_i-\hat{\mu}^L \left(c_i\right)|| < ||x^L_i-\hat{\mu}^L_{j}||\hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i\]
-Which means
-\[h(i) = \underset{j}{\operatorname{argmin}} \hspace{0.05 in}||x^L_i - \hat{\mu}^L_j|| = c_i \]
-$(9)$ follows directly.
-\end{proof}
-\noindent
-If we further assume equal prior are observed in training set, then we can show that the LDA classifier has prediction error $0$, that is
-\begin{proposition}
-Conditioning on $\mathcal{D}$ and $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$ and $n^{train}_j=\frac{m}{K}$ $\forall j$ in training set, the classifier $\mathcal{C}$ trained using LDA has prediction error $0$ in case of $k = K$.
-\end{proposition}
-\begin{proof}
-Same as in the proof of proposition $4$, given $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$, the training set of $\mathcal{C}$ consists of data in the form of 
-\[ \left(\mathbf{X}^L_{train},\mathcal{G} \right) = \left\lbrace\left(\mu^L_{j} + \varepsilon^L_i, g_j \right)\right\rbrace\hspace{0.2in} j =1,2,...,K, \hspace{0.1in} i=1,...,m\]
-The class-conditional density of $X^L$ in class $\mathcal{G} = g_j$ is $$f_j\left(x^L_i\right) = \mathbb{P}\left( X^L=x^L_i | \mathcal{G} = g_j \right) $$ where $x^L_i=\mu^L(c_i)+\varepsilon^L_i$ and $f_j\left(x^L_i\right)$ has multivariate Gaussian distribution
-\[ f_j\left(x^L_i\right) = \frac{1}{(2\pi)^{L/2}|\Sigma^L_j|^{1/2}}e^{-\frac{1}{2}(x^L_i-\mu^L_j)^T(\Sigma^L_j)^{-1}(x^L_i-\mu^L_j)	}\]
-LDA assumes a commom covariance matrix $\Sigma^L_j = \Sigma^L \hspace{0.1in} \forall j$.
-The posterior probability is 
-\[	\mathbb{P}\left( \mathcal{G} = g_j | X^L=x^L_i  \right)  = \frac{f_j\left(x^L_i\right) \pi_j}{\sum^K_{l=1}f_l\left(x^L_i\right) \pi_l}	\]
-where $ \pi_j$ is the prior for class $g_j$. In practice, $\pi_j, \mu^L_j$ and $\Sigma^L$ are estimated from data by
- \begin{itemize}
- 	\item  $\hat{\pi}_j = \frac{n^{train}_j}{m}$, where $n^{train}_j$ is the number of class-$g_j$ observation in training set and $m$ is the total observations in training set
- 	\item  $\hat{\mu}^L_{j} = \frac{\sum x^L_i \cdot I_{c_i=j} }{n^{train}_j}$, $x^L_i  \in \mathbf{X}^L_{train}$
- 	\item  $\hat{\Sigma}^L = \frac{1}{m-K} \sum^K_{j=1}\sum_{c_i=j} (x^L_i - \hat{\mu}^L_j )(x^L_i - \hat{\mu}^L_j )^T$
- \end{itemize}
-Because $\forall j$ $n^{train}_j=\frac{m}{K}$ , $\hat{\pi}_j = \frac{1}{K}$ $\forall j$. LDA predicts class for each $x^L_i  \in \mathbf{X}^L_{test}$ by maximizing the posterior probability
-\begin{align}
-     \hat{\mathcal{G}}(x_i) & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} \mathbb{P}\left( \mathcal{G} = g_j | X^L=x^L_i  \right)  \nonumber   \\ 
-     & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} f_j\left(x^L_i\right) \hat{\pi}_j \nonumber \nonumber  \\
-   \hat{\pi}_j = \frac{1}{K} \hspace{0.1in}\forall j  \hspace{0.2in} & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} f_j\left(x^L_i\right) \nonumber  \nonumber  \\    
-     & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in}  \frac{1}{(2\pi)^{L/2}|\hat{\Sigma}^L|^{1/2}}e^{-\frac{1}{2}(x^L_i-\hat{\mu}^L_j)^T(\hat{\Sigma}^{L})^{-1}(x^L_i-\hat{\mu}^L_j)	} \nonumber  \\
-     &=  \underset{g_j}{\operatorname{argmin}}\hspace{0.05 in} (x^L_i-\hat{\mu}^L_j)^T(\hat{\Sigma}^{L})^{-1}(x^L_i-\hat{\mu}^L_j) \nonumber   \\   \nonumber 
-     &= \sum^K_{j=1} g_j \cdot I_{c_i=j}
-\end{align}
-The last equation is because given $\mathcal{D}$, we have
-\[||x^L_i-\hat{\mu}^L \left(c_i\right)|| < ||x^L_i-\hat{\mu}^L_{j}||\hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i\]
-Since it coincides with equation $9$, the proof is complete. 
-\end{proof}
-\vspace{0.3in}
-\noindent
-By definition,
-\begin{equation}
-Error_k = \sum^k_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 = \sum^{n-m}_{i=1} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2, \hspace{0.1in}with\hspace{0.1in} g_j = \hat{g}(x^L_i)
-\end{equation}
-\noindent
-For any parameter $k\leq k_{max}$, $x_i \in \mathbf{X}_{test}$ and $||\hat{\mu}^R \left(g_j\right) - x^R_i||^2$ in $(10)$ are i.i.d with finite expectation
-\[
-E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 < \infty
-\]
-Since $||\hat{\mu}^R \left(g_j\right) - x^R_i||$ is bounded 
-\begin{align} \nonumber 
-||\hat{\mu}^R \left(g_j\right) - x^R_i|| & = ||\hat{\mu}^R \left(g_j\right) - \mu^R(c_i)- \varepsilon^R_i|| \\ \nonumber 
-&\leq ||\hat{\mu}^R \left(g_j\right) - \mu^R(c_i)|| + ||\varepsilon^R_i||\\ \nonumber 
-&\leq \max_{i\neq j}||\mu^R_i - \mu^R_j|| + 2||\varepsilon^R_i||\\ \nonumber 
-\end{align}
-Given $||\varepsilon^R_i||\leq 1$.
-Therefore, by strong law of large number (SLLN), when $(n-m) \to \infty$
-\begin{equation}
-\frac{1}{n-m} Error_k = \frac{1}{n-m}\sum^{n-m}_{i=1}||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 \hspace{0.1in}  \to \hspace{0.1in} E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 \hspace{0.1in}a.s.
-\hspace{0.1in} where \hspace{0.1in} g_j = \hat{g}(x^L_i)
-\end{equation}
-
-\begin{lemma}
-	If assumptions $1-6$ and the consistent assumptions of Theorem $3$ hold, when $k=K$, $\frac{1}{n-m} Error_K \to tr\{\Sigma^R\}$ a.s as $n \to \infty$
-\end{lemma}
-\begin{proof}
-Since $m= n \cdot r$ and $(n-m) = n \cdot (1-r)$, $n \to \infty$ implies $(n-m) \to \infty$ and $m \to \infty$, where event $\mathcal{D}$ and $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$ happen with probability $1$ by Proposition $2$ and Proposition $3$.
-By Proposition $4$, $\mathcal{C}$ has prediction error $0$ on $\mathbf{X}^L_{test}$.
-Hence the predicted cluster $\hat{g} \left(x^L_i\right)$ for $x^L_i$ is the correct one $\hat{g} \left(x^R_i\right)$, which determined by $x^R_i$.
-By $(8)$, if $g_j = \hat{g} \left(x^R_i\right) = \hat{g} \left(x^L_i\right)$
-\[x^R_i = \mu^R \left(c_i\right) + \varepsilon^R_i = \mu^R_{j} + \varepsilon^R_i\] 
-\noindent
-Therefore,
-\begin{align} \nonumber
-    Error_K &= \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2   \\ \nonumber
-     & = \sum^K_{j=1} \sum_{g_j} ||\hat{\mu}^R \left(g_j\right) -\mu^R_{j} - \varepsilon^R_i||^2 \\ \nonumber 
-by \hspace{0.1in} (7) \hspace{0.2in}   & = \sum^K_{j=1} \sum_{g_j} ||\hat{\mu}^R_{j} -\mu^R_{j} - \varepsilon^R_i||^2 \\ 
-    % &=  \sum^{n-m}_{i=1}||\hat{\mu}^R_{j} -\mu^R_{j} - \varepsilon^R_i||^2
-    &= \sum^K_{j=1} \sum_{g_j}  ||\hat{\mu}^R_{j} -\mu^R_{j}||^2 + \sum^{n-m}_{i=1}||\varepsilon^R_i||^2
-\end{align}
-$(12)$ is true since $\varepsilon^R_i$ is independent from $\mu_j$ and $\varepsilon^L_i$ by assumption $3$.
-So because $\hat{\mu}^R \left(g_j\right)$ is a function of $x^L_i = \mu^L_i + \varepsilon^L_i$, $\varepsilon^R_i$ is also independent from $\hat{\mu}^R \left(g_j\right)$.
-\\
-
-Hence
-\begin{align} \nonumber
-  E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2  & = \sum^K_{j=1}E\left(||\hat{\mu}^R \left(g_j\right) - x^R_i||^2  \hspace{0.1 in}| \hspace{0.1 in} x^R_i = \mu^R_{j} + \varepsilon^R_i\right) P\left(x^R_i = \mu^R_{j} + \varepsilon^R_i\right)\\ \nonumber
- % & = \sum^K_{j=1}E||\hat{\mu}^R_{j} -\mu^R_{j} - \varepsilon^R_i||^2\cdot\pi_j\\ \nonumber
-   & = \sum^K_{j=1}E\left(||\hat{\mu}^R_{j} -\mu^R_{j}||^2 +||\varepsilon^R_i||^2\right)\cdot\pi_j \\ \nonumber
-   &= \sum^K_{j=1}E||\hat{\mu}^R_{j} -\mu^R_{j}||^2\cdot\pi_j + \sum^K_{j=1}E||\varepsilon^R_i||^2\cdot\pi_j\\ \nonumber
-    &= \sum^K_{j=1} \frac{tr\{\Sigma^R\}}{n^{train}_j}\cdot\pi_j + E||\varepsilon^R_i||^2\\ \nonumber
-  m \to \infty \hspace{0.2in}  &= \sum^K_{j=1} \frac{tr\{\Sigma^R\}}{m\cdot\pi_j }\cdot\pi_j + E||\varepsilon^R_i||^2\\ \nonumber
-      &= E||\varepsilon^R_i||^2\\ \nonumber
-       &=tr\{\Sigma^R\} \\ \nonumber
-\end{align} 
-By $(11)$, $\frac{1}{n-m} Error_K \to tr\{\Sigma^R\}$ a.s as $n \to \infty$.
-\end{proof}
-\begin{lemma}
-	If assumptions $1-6$ and Assumption I.(defined below) hold, as well as the consistent assumptions of Theorem $3$ and assumptions of Theorem $4$ (defined below) hold,  then if $k \neq K$, $\frac{1}{n-m} Error_K \to \Theta$ a.s as $n \to \infty$, and $\Theta > tr\{\Sigma^R\} $
-\end{lemma}
-\begin{proof}
-The proof of almost surely convergent part follows from $(11)$ directly, so here we only need to show 
-\[\Theta = E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 > tr\{\Sigma^R\}, \hspace{0.2in} where \hspace{0.1in}g_j=\hat{g}(x^L_i) \]
-Because 
-\begin{align} \nonumber
-  E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2  &=  E||\hat{\mu}^R \left(\hat{g}(x^L_i)\right) - x^R_i||^2 \\ \nonumber
-  & = \sum^K_{j=1}E\left(||\hat{\mu}^R \left(\hat{g}(x^L_i)\right) - x^R_i||^2  \hspace{0.1 in}| \hspace{0.1 in} x^R_i = \mu^R_{j} + \varepsilon^R_i\right) P\left(x^R_i = \mu^R_{j} + \varepsilon^R_i\right)\\ \nonumber
-  &= \sum^K_{j=1}E\left(||\hat{\mu}^R \left(\hat{g}(x^L_i)\right) - \mu^R_{j} - \varepsilon^R_i||^2\hspace{0.1 in}| \hspace{0.1 in} x^R_i = \mu^R_{j} + \varepsilon^R_i \right)\pi_j\\ \nonumber
-  &=  \sum^K_{j=1}E\left(||\hat{\mu}^R \left(\hat{g}(x^L_i)\right) - \mu^R_{j}||^2\hspace{0.1 in}| \hspace{0.1 in} x^R_i = \mu^R_{j} + \varepsilon^R_i \right)\pi_j + E||\varepsilon^R_i||^2\\ \nonumber
-\end{align}
-Note that unlike in $k = K$ case, here $\hat{\mu}^R \left(\hat{g}(x^L_i)\right) \neq \hat{\mu}^R_j$ could be true for $x^R_i = \mu^R_j+\varepsilon^R_i$.\\
-
-Since $E||\varepsilon^R_i||^2 = tr\{\Sigma^R\}$, our next goal is to show as $n \to \infty$
-\begin{equation}
-\sum^K_{j=1}E\left(||\hat{\mu}^R \left(\hat{g}(x^L_i)\right) - \mu^R_{j}||^2\hspace{0.1 in}| \hspace{0.1 in} x^R_i = \mu^R_{j} + \varepsilon^R_i \right)\pi_j > 0
-\end{equation}
-Note that $\hat{\mu}^R \left(\hat{g}(x^L_i)\right) \in A_m(k)$. The necessary but not sufficient condition to have  
-\[\sum^K_{j=1}E\left(||\hat{\mu}^R \left(\hat{g}(x^L_i)\right) - \mu^R_{j}||^2\hspace{0.1 in}| \hspace{0.1 in} x^R_i = \mu^R_{j} + \varepsilon^R_i \right)\pi_j = 0\]
-is $\{\mu^R_1,\mu^R_2,...,\mu^R_K\} \in A_m(k)$.
-The strategy here is to show that at least one $\mu^R_j \notin A_m(k)$ when $k\neq K$ and $m \to \infty$, where $\mu^R_j \in \{\mu^R_1,\mu^R_2,...,\mu^R_K\}$ or equivalently, at least one $\mu^R_j \notin \bar{A}(k)$ when applying K-means on population $\mathcal{P}$ with parameter $k$.
-Following proof is based on the later one.
-\\
-
-In case of $k < K$, the proof is trivial.
-This is because there are only $k$ distinct cluster centers returned from K-means, while there are $K$ distinct $\mu^R_j$.
-$k < K$ means at least one $\mu^R_j \notin \bar{A}(k)$.
-\\\\
-In case of $k > K$, at least one cluster will be partitioned into at least two parts.
-This is because there are $K$ clusters in population $\mathcal{P}$ and if none of them be partitioned, K-means will give $K$ cluster centers instead of $k$ -- contradiction.\\
-
-WLOG, we assume the partitioned cluster is $\mu^R_1$.
-In training set, there are $n^{train}_1$ observations in cluster $\mu^R_1$.
-Say the partition results one part with $n_1$ observations and its sample mean is $\hat{\mu}_1$, also we have \\
-
-\textbf{Assumption I.} The partition results $n_1$ is a fixed proportion from $n^{train}_1$ when apply K-means on population $\mathcal{P}$ with parameter $k$, i.e $\lim_{m \to \infty} \frac{n_1}{n^{train}_1} = a_1$, where $0< a_1< 1$ is a positive constant.
-Also, the sample mean of the $n_1$ observation is different from the sample mean of the rest $n^{train}_1-n_1$ points.\\
-
-With Assumption I., theorem $4$ below shows that $\mu^R_1 \notin \bar{A}(k)$. Therefore $(13)$ holds, which means 
-\[\Theta = E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 > tr\{\Sigma^R\} \]
-The proof is complete.
-\end{proof}
-\begin{theorem} 
-Given $||\varepsilon^R_i||\leq1$, if for any $\mu^R_j$ in  $\{\mu^R_1,...,\mu^R_K\}$, WLOG, let $\mu^R_j = \mu^R_1$, $ ||\mu^R_1 - \mu|| > 2$, where $\mu$ is any weighted mean of $(\mu^R_2,...,\mu^R_{K})$ by weight vector $(n_2,...,n_{K})$, where $n_j,j=2,...,K$ is natural number and $\sum^{K}_{i=2}n_i >0$, i.e.
-$ \mu = \frac{n_2\mu_2+...+n_{K}\mu_{K}}{n_2+n_2...+n_{K}}$.
-Then $\mu^R_1 \notin \bar{A}(k)$ if $k \neq K$.
-\end{theorem}
-\begin{proof}
-Let $\hat{\mu'}_1$ denotes the specific cluster center we are looking which is returned from K-means with parameter $k$ on population $\mathcal{P}$.
-By the condition given in theorem $4$, if such cluster consists of observations which are not from $\mu^R_1$, then such cluster center $\hat{\mu'}_1$ is at least $1$ unit away from $\mu^R_1$ since
-\[\hat{\mu'}_1 = \frac{n_2\mu^R_2+...+n_{K}\mu^R_{K}}{n_2+n_2...+n_{K}}+\frac{\sum \varepsilon^R_i}{n_2+n_2...+n_{K}}\]
-and 
-\begin{align}
-	||\hat{\mu'}_1-\mu^R_1|| &= ||\frac{n_2\mu^R_2+...+n_{K}\mu^R_{K}}{n_2+n_2...+n_{K}}+\frac{\sum \varepsilon^R_i}{n_2+n_2...+n_{K}}-\mu^R_1||  \nonumber \\ \nonumber
-	&\geq ||\frac{n_2\mu^R_2+...+n_{K}\mu^R_{K}}{n_2+n_2...+n_{K}}-\mu^R_1||-||\frac{\sum \varepsilon^R_i}{n_2+n_2...+n_{K}}||   \\ \nonumber
-	&> 2-||\frac{\sum \varepsilon^R_i}{n_2+n_2...+n_{K}}||   \\ \nonumber
-	&> 1 
-\end{align}
-since $||\varepsilon^R_i||\leq1$.
-Hence, the only hope for $\hat{\mu'}_1 = \mu^R_1$ (therefore $\mu^R_1 \in \bar{A}(k)$) comes from cluster contains observations from $\mu^R_1$.
-WLOG, we assume those observations from $\mu^R_1$ are the $n_1$ points.
-\begin{itemize}
-	\item When the $n_1$ observations form a single cluster\\
-	In this case, $\hat{\mu'}_1 = \hat{\mu}_1$. We want to show 
-	\begin{equation}
-		 \lim_{m\to\infty}||\hat{\mu}_1-\mu^R_1|| > 0 
-	\end{equation}
-	It indeed is the case since when $m \to \infty$, by law of large number the sample mean converges to the population mean, so with probability $1$ 
-	\begin{align}
-		||\bar{\mu}_1-\mu^R_1|| &= ||\frac{n_1\hat{\mu}_1+(n^{train}_1-n_1)\hat{\mu}_{1'}}{n^{train}_1} -\mu^R_1||   \nonumber \\\nonumber		
-		&= ||\frac{n_1}{n^{train}_1}\hat{\mu}_1 + \frac{n^{train}_1-n_1}{n^{train}_1}\hat{\mu}_{1'} - \mu^R_1||  \\\nonumber
-		&= ||a_1\hat{\mu}_1 + (1-a_1)\hat{\mu}_{1'} - \mu^R_1 ||\\\nonumber
-		&= 0\\\nonumber
-    \end{align}	 
-Since $\hat{\mu}_1 \neq \hat{\mu}_{1'}$, and $0<a_1<1$ by Assumption I, \[\lim_{m\to\infty}||\hat{\mu}_1-\mu^R_1|| > 0\]Where $\hat{\mu}_{1'}$ is the sample mean of the $n^{train}_1-n_1$ points of cluster $\mu^R_1$.\\
-	\item When the $n_1$ observations be grouped together with $n_2$ observations from cluster $\mu^R_2$. \\\\
-	Notice that it's a special case $\left(\left(n_2,...,n_{K}\right) = (n_2,0,...,0)\right)$ of the $n_1$ observations be grouped together with $\sum^{K}_{i=2}n_i$ observations from cluster $(\mu^R_2,...,\mu^R_{K})$, i.e.
-$\mu = \mu^R_2$.
-The result apply to the general cases.\\\\
-	The cluster consists of $n_1$ observations from center $\mu^R_1$ with sample mean $\hat{\mu}_1$, and $n_2$ observations from center $\mu^R_2$ with sample mean $\hat{\mu}_2$.
-The cluster center (estimator of $\mu_1$) is then
-	\[\hat{\mu'}_1 = \frac{n_1\hat{\mu}_1+n_2\hat{\mu}_2}{n_1+n_2} \]
-From above, we have $||\hat{\mu}_2-\mu^R_1|| > 1$ and $||\hat{\mu}_1-\mu^R_1|| > 0$.
-Since the $n_1$ observations are cluster together with $n_2$ observations from center $\mu^R_2$, and $\hat{\mu}_2$ is outside of cluster $\mu_1$ as $||\hat{\mu}_2-\mu^R_1|| > 1$ and $||\varepsilon^R_i||<1$, it's reasonable to assume in terms of $\mu^R_1$, $\hat{\mu}_1$ is biased towards $\hat{\mu}_2$.
-In other words, the inner product of two vector is nonnegative, i.e.
-\[ <\hat{\mu}_1-\mu^R_1,\hat{\mu}_2-\mu^R_1>\hspace{0.2in} \geq 0 \] 
-Therefore, when $m\to \infty$
-\begin{align}
-	||\hat{\mu'}_1 - \mu^R_1||^2&= || \frac{n_1\hat{\mu}_1+n_2\hat{\mu}_2}{n_1+n_2} - \mu^R_1||^2  \nonumber \\ \nonumber
-	&=||\frac{n_1}{n_1+n_2}(\hat{\mu}_1-\mu^R_1)+\frac{n_2}{n_1+n_2}(\hat{\mu}_2-\mu^R_1)||^2  \\ \nonumber
-	&= ||\frac{n_1}{n_1+n_2}(\hat{\mu}_1-\mu^R_1)||^2+ ||\frac{n_2}{n_1+n_2}(\hat{\mu}_2-\mu^R_1)||^2+\frac{2n_1n_2}{(n_1+n_2)^2} <\hat{\mu}_1-\mu^R_1,\hat{\mu}_2-\mu^R_1>\\ \nonumber
-	&>0  \\ \nonumber
-\end{align}
-which means \hspace{1in} $\lim_{m\to\infty}||\hat{\mu'}_1-\mu^R_1|| > 0$
-\end{itemize}
-Since none of above situation returns cluster center $\hat{\mu'}_1 = \mu^R_1$, we claim $\mu^R_1 \notin \bar{A}(k)$
-\end{proof}
-\section{Simulation}
-In this section, simulation is performed to evaluate the performance of our proposed methods in locating the ``correct" number of clusters.
-We compare with a basket of existing methods including Gap statistics \citep{tibshirani2001estimating}, Gaussian mixture model-based clustering \citep{fraley2002model}, CH-index \citep{calinski1974dendrite}, Hartigan statistics \citep{hartigan1975clustering}, Jump method \citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster}, Bootstrap stability \citep{fang2012selection} in following simulation settings.
-We select $L=R$ and $5-$fold cross-validation in row ($m=\frac{4}{5}n$) as default parameter setting for our proposed Gabriel method.
-Note that set $L=R$ corresponding to $2-$fold cross-validation in column.
-	\begin{enumerate}
-		\item[1.] Single cluster in $10$ dimensions --- $200$ observations, each observation uniformly distributed over $[0,1]$ in each dimension. 
-		\item[2.] Two clusters in $4$ dimensions --- $50$ i.i.d standard normal observations are generated, i.e.
-generated from multivariate normal $\mathcal{N}\left(\boldsymbol\mu,\mathbf{I}\right)$ for each cluster, with cluster center $\boldsymbol\mu = (1,0,0,1)$ and $(1,3.5,3.5,1)$ respectively
-		\item[3.] Four clusters in $100$ dimensions ---  Each cluster has $100$ or $150$ i.i.d standard normal observations, with cluster centers randomly generated from multivariate normal distribution $\mathcal{N}\left(\mathbf{0},0.65^2\mathbf{I}\right)$
-		\item[4.] Ten clusters in $100$ dimensions ---  Each cluster has $50$ or $100$ i.i.d standard normal observations, with cluster centers randomly generated from multivariate normal distribution $\mathcal{N}\left(\mathbf{0},0.72^2\mathbf{I}\right)$
-		\item[5.] Four log-normal clusters in $16$ dimensions --- For each cluster, $30$ or $60$ i.i.d centered log-normal observations are generated from $ln\mathcal{N}\left(0,0.5^2 \right)$.
-The cluster centers are randomly generated from multivariate normal distribution $\mathcal{N}\left(\mathbf{0},1.2^2\mathbf{I}\right)$
-		\item[6.] Three exponential clusters in $20$ dimensions with different variance --- $40$ observations in each cluster are generated from centered exponential distribution ($exp(\lambda)-1/\lambda$ in each dimension), with $\boldsymbol\mu$ randomly generated from  $\mathcal{N}\left(\mathbf{0},19\mathbf{I}\right)$ and $\lambda = 1, 1/2, 1/5 $ respectively.
-	\end{enumerate}
-Note that in setting $2-6$, all clusters are well-separated, i.e.
-no overlapping.
-In fact, any simulated clusters with minimum distance less than $1$ unit was discarded, so there is clear definition of true number of clusters.
-The parameters in setting $2-6$ are chosen such that about half of the random realization were discarded.
-The idea is borrowed from \cite{tibshirani2001estimating}.\\\\
-Besides the number of clusters selected by each algorithm is returned, we also give the prediction error (PE) for each algorithm in each setting. The PE is defined by 
-\[ PE(k) = \frac{||\hat{\mathbf{X}}_k- \boldsymbol{\mu} ||_F}{PE_{oracle}} \]
-where \[PE_{oracle} = \min_k ||\hat{\mathbf{X}}_k- \boldsymbol{\mu} ||_F\]
-Matrix $\boldsymbol{\mu}$ consists of true centers of observations in data matrix $\mathbf{X}$, while $\hat{\mathbf{X}}_k$ is obtained by replacing each observation in $\mathbf{X}$ by its closest cluster center result from applying K-means on $\mathbf{X}$ with parameter $k$.
-Here $k$ is the optimal number of clusters selected by each algorithm on data $\mathbf{X}$.
-This is only possible on simulated data because otherwise one would never know the true center matrix $\boldsymbol{\mu}$.\\\\
-The oracle selected number for each data $\mathbf{X}$ is also given on the table, which is defined by 
-\[ K_{oracle} =  \underset{k}{\operatorname{argmin}}\hspace{0.05in} ||\hat{\mathbf{X}}_k- \boldsymbol{\mu} ||_F \]
-
-\begin{table}
-\centering
-\caption{ \label{table1}Simulation results} 
-{\scriptsize
-\begin{tabular}{lcccccccccccccccccc}
-\hline
-Number of Clusters & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA & PE \\ \hline
-\input{demo/bench/setting1/results.tex}
-\hline
-\input{demo/bench/setting2/results.tex}
-\hline
-\input{demo/bench/setting3/results.tex}
-\hline
-\input{demo/bench/setting4/results.tex}
-\hline
-\input{demo/bench/setting5/results.tex}
-\hline
-\input{demo/bench/setting6/results.tex}
-\hline
-\end{tabular}
-}\\
-\hspace{-2.0in} \footnotesize {In Setting~1, ``\textendash'' means the method can not be used with parameter $k=1$}.
-\end{table}
-
-\begin{table}[H]
-\centering
-\caption{ \label{table1}Old Simulation results}
-\scalebox{0.97}{
-\begin{tabular}{ lccccccccccccc}
-\hline
- Number of clusters &&& $1$&$2$&$3$&$4$&$5$&$6$&$7$&$8$&$9$&$10$&PE \\ \hline
-\textit{Setting $1$} &&& &&&&&&&&& \\
-Oracle&&& $\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1$\\ 
-Gap&&& $\mathbf{92}$&$0$&$1$&$1$&$1$&$2$&$1$&$0$&$1$&$1$&$8.0\pm28.3$\\ 
-Gaussian-Mix&&& $\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
-CH&&& $-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$\\ 
-Hartigan&&& $-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$\\ 
-Jump&&& $\mathbf{0}$&$0$&$0$&$0$&$0$&$0$&$1$&$8$&$34$&$57$&$102.3\pm 58.7$\\ 
-Prediction strength&&&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
-Stability&&& $-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$&$-$\\ 
-Gabriel CV&&& $\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
-Wold CV&&& $\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
-\hline
-\textit{Setting $2$} &&& &&&&&&&&& \\
-Oracle&&& $0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1$\\ 
-Gap&&& $0$&$\mathbf{98}$&$1$&$1$&$0$&$0$&$0$&$0$&$0$&$0$&$1.2\pm 1.3$\\ 
-Gaussian-Mix&&& $0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
-CH&&& $0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\
-Hartigan&&& $0$&$\mathbf{0}$&$6$&$16$&$21$&$14$&$12$&$10$&$8$&$13$&$28.1\pm 50.8$\\
-Jump&&& $0$&$\mathbf{70}$&$0$&$0$&$0$&$0$&$0$&$2$&$7$&$21$&$15.5\pm 55.4$\\ 
-Prediction strength&&&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
-Stability&&& $0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\
-Gabriel CV&&& $0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
-Wold CV&&& $0$&$\mathbf{66}$&$3$&$14$&$11$&$4$&$1$&$0$&$1$&$0$&$6.0\pm 10.6$\\ 
-\hline
-\textit{Setting $3$} &&& &&&&&&&&& \\
-Oracle&&& $0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$1$\\ 
-Gap&&& $0$&$0$&$0$&$\mathbf{69}$&$28$&$3$&$0$&$0$&$0$&$0$&$1.3\pm 0.5$\\ 
-Gaussian-Mix&&& $85$&$2$&$3$&$\mathbf{3}$&$1$&$0$&$4$&$2$&$0$&$0$&$35.9\pm 13.7$\\ 
-CH&&& $0$&$36$&$28$&$\mathbf{36}$&$0$&$0$&$0$&$0$&$0$&$0$&$12.1\pm 9.9$\\ 
-Hartigan&&& $0$&$0$&$13$&$\mathbf{75}$&$11$&$0$&$1$&$0$&$0$&$0$&$2.5\pm 3.5$\\
-Jump&&& $0$&$0$&$0$&$\mathbf{0}$&$0$&$0$&$0$&$0$&$14$&$86$&$5.6\pm 0.4$\\ 
-Prediction strength&&&$56$&$0$&$0$&$\mathbf{44}$&$0$&$0$&$0$&$0$&$0$&$0$&$23.5\pm 20.6$\\ 
-Stability&&& $0$&$0$&$0$&$\mathbf{33}$&$59$&$8$&$0$&$0$&$0$&$0$&$1.7\pm 0.5$\\
-Gabriel CV&&& $0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
-Wold CV&&& $0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
-\hline
- Number of clusters &&& $\leq 6$&$7$&$8$&$9$&$10$&$11$&$12$&$13$&$14$&$15$ \\ \hline
-\textit{Setting $4$} &&& &&&&&&&&& \\
-Oracle&&& $0$&$0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$1$\\ 
-Gap&&& $0$&$0$&$0$&$0$&$\mathbf{12}$&$39$&$23$&$16$&$7$&$3$&$1.5\pm 0.4$\\ 
-Gaussian-Mix&&& $71$&$6$&$4$&$2$&$\mathbf{4}$&$4$&$0$&$2$&$3$&$4$&$22.2\pm 13.5$\\ 
-CH&&& $84$&$6$&$6$&$4$&$\mathbf{0}$&$0$&$0$&$0$&$0$&$0$&$20.8\pm 8.9$\\ 
-Hartigan&&& $13$&$4$&$13$&$21$&$\mathbf{17}$&$12$&$5$&$9$&$1$&$5$&$4.4\pm 4.8$\\
-Jump&&& $0$&$0$&$0$&$20$&$\mathbf{80}$&$0$&$0$&$0$&$0$&$0$&$1.4\pm 0.9$\\ 
-Prediction strength&&&$100$&$0$&$0$&$0$&$\mathbf{0}$&$0$&$0$&$0$&$0$&$0$&$35.2\pm 4.4$\\ 
-Stability&&& $0$&$0$&$0$&$0$&$\mathbf{0}$&$3$&$24$&$29$&$28$&$16$&$2.0\pm 0.3$\\
-Gabriel CV&&&  $0$&$0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\  
-Wold CV&&&  $0$&$0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0$\\ 
-\hline
-\end{tabular}
-}\\
-\hspace{-2.0in} \footnotesize {In setting $1$, ``-'' means the method can not be used with parameter $k=1$}.
-\end{table}
-\begin{table}[H]
-\centering
-\begin{tabular}{ lccccccccccccc}
-\hline
- Number of clusters &&& $1$&$2$&$3$&$4$&$5$&$6$&$7$&$8$&$9$&$10$&PE \\ \hline
-\textit{Setting $5$} &&& &&&&&&&&& \\
-Oracle&&& $0$&$0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$1$\\ 
-Gap&&& $0$&$0$&$0$&$\mathbf{66}$&$25$&$7$&$1$&$1$&$0$&$0$&$1.8\pm 1.3$\\ 
-Gaussian-Mix&&& $0$&$0$&$0$&$\mathbf{56}$&$36$&$6$&$2$&$0$&$0$&$0$&$2.0\pm 1.3$\\ 
-CH&&& $0$&$5$&$25$&$\mathbf{53}$&$12$&$2$&$1$&$0$&$0$&$0$&$10.3\pm 17.0$\\ 
-Hartigan&&& $0$&$0$&$24$&$\mathbf{61}$&$5$&$6$&$3$&$1$&$0$&$0$&$7.3\pm 11.5$\\
-Jump&&& $0$&$0$&$0$&$\mathbf{73}$&$0$&$0$&$1$&$0$&$5$&$21$&$3.1\pm 3.5$\\ 
-Prediction strength&&&$74$&$8$&$5$&$\mathbf{13}$&$0$&$0$&$0$&$0$&$0$&$0$&$114.5\pm 67.8$\\ 
-Stability&&& $0$&$4$&$5$&$\mathbf{23}$&$36$&$23$&$7$&$2$&$0$&$0$&$5.7\pm 10.2$\\
-Gabriel CV&&& $0$&$0$&$0$&$\mathbf{91}$&$8$&$1$&$0$&$0$&$0$&$0$&$1.1\pm 0.5$\\ 
-Wold CV&&& $0$&$0$&$0$&$\mathbf{99}$&$1$&$0$&$0$&$0$&$0$&$0$&$1.0\pm 0.1$\\ 
-\hline
-\textit{Setting $6$} &&& &&&&&&&&& \\
-Oracle&&& $0$&$0$&$\mathbf{100}$&$0$&$0$&$0$&$0$&$0$&$0$&$0$&$1$\\ 
-Gap&&& $0$&$0$&$\mathbf{10}$&$8$&$13$&$11$&$12$&$11$&$17$&$18$&$9.8\pm 5.9$\\ 
-Gaussian-Mix&&& $0$&$0$&$\mathbf{88}$&$12$&$0$&$0$&$0$&$0$&$0$&$0$&$1.4\pm 1.3$\\ 
-CH&&& $0$&$17$&$\mathbf{74}$&$8$&$1$&$0$&$0$&$0$&$0$&$0$&$4.0\pm 6.3$\\ 
-Hartigan&&& $0$&$0$&$\mathbf{87}$&$10$&$3$&$0$&$0$&$0$&$0$&$0$&$1.6\pm 1.6$\\
-Jump&&& $0$&$0$&$\mathbf{0}$&$0$&$0$&$1$&$2$&$13$&$32$&$52$&$13.7\pm 5.0$\\ 
-Prediction strength&&&$19$&$2$&$\mathbf{78}$&$1$&$0$&$0$&$0$&$0$&$0$&$0$&$10.3\pm 20.5$\\ 
-Stability&&& $0$&$0$&$\mathbf{24}$&$39$&$29$&$7$&$1$&$0$&$0$&$0$&$4.5\pm 3.0$\\
-Gabriel CV&&& $0$&$0$&$\mathbf{97}$&$3$&$0$&$0$&$0$&$0$&$0$&$0$&$1.1\pm 0.7$\\ 
-Wold CV&&& $0$&$0$&$\mathbf{89}$&$9$&$1$&$0$&$1$&$0$&$0$&$0$&$1.6\pm 1.9$\\ 
-\hline
-\end{tabular}\\
-\hspace{-1.8in} \footnotesize {Last column gives the mean and standard deviation of PE for each algorithm}.
-\end{table}
-\noindent
-All methods are used with their default parameter setting except for Gap statistics.
-We selected the $k$  corresponds to the global maximum of gap statistics, rather than the default method which selects the smallest $k$ such that the gap statistics is not more than 1 standard error away from the first local maximum.
-This is because among all the possible options, the global maximum criteria gives the best result in our simulation settings.\\
-
-The ``--'' sign in setting $1$ is because for CH-index, Hartigen and Bootstrap stability, the minimum $k$ they can pick is $2$.
-So they are not included for comparison since the true $k=1$.
-Setting $2-4$ have Gaussian distribution, where the Gaussian mixture model-based clustering suppose to perform the best.
-It is only the case for setting $2$.
-In setting $3$ and $4$, it has a very poor performance even if the underlying distribution is indeed Gaussian mixture model.
-It highlights the difficulties that high dimension data could cause.
-Other methods (Gap statistics, CH-index, Hartigan, Jump method, Prediction strength and Boostrap stability) also have problems in these two high-dimension settings as we could not find any one with reasonable performance in terms of finding the correct $k$ in both cases.
-In terms of PE, Boostrap stability and Gap statistics have reasonable performance, but come with the price of picking larger $k$.
-Our proposed Gabriel method (also Wold CV method) clearly stands out from the basket of existing methods using either measure under the high-dimension setting $3$ and $4$.\\
-
-Setting $5$ and $6$ corresponding to the situation where data are heavy-tailed instead of normal distributed, coupled with unequal number of observations or unequal variance for different clusters.
-From Table \ref{table1}, we can see that our proposed Gabriel method is robust and has superior performance compare with other methods.
-Because one can hardly tell the underlying distribution of data in practice, the resilience to non-Gaussian data gives the Gabriel method a clear edge.
+All methods are used with their default parameter setting except for Gap
+statistics. We selected the $k$  corresponds to the global maximum of gap
+statistics, rather than the default method which selects the smallest $k$ such
+that the gap statistics is not more than 1 standard error away from the first
+local maximum. This is because among all the possible options, the global
+maximum criteria gives the best result in our simulation settings.
+
+The ``\textendash'' sign in Setting~1 is because for CH-index, Hartigen and
+Bootstrap stability, the minimum $k$ they can pick is $2$. So they are not
+included for comparison since the true $k=1$. Setting $2-4$ have Gaussian
+distribution, where the Gaussian mixture model-based clustering suppose to
+perform the best. It is only the case for setting $2$. In setting $3$ and $4$,
+it has a very poor performance even if the underlying distribution is indeed
+Gaussian mixture model. It highlights the difficulties that high dimension
+data could cause. Other methods (Gap statistics, CH-index, Hartigan, Jump
+method, Prediction strength and Boostrap stability) also have problems in
+these two high-dimension settings as we could not find any one with reasonable
+performance in terms of finding the correct $k$ in both cases. In terms of PE,
+Boostrap stability and Gap statistics have reasonable performance, but come
+with the price of picking larger $k$. Our proposed Gabriel method (also Wold
+CV method) clearly stands out from the basket of existing methods using either
+measure under the high-dimension setting $3$ and $4$.
+
+Setting $5$ and $6$ corresponding to the situation where data are heavy-tailed
+instead of normal distributed, coupled with unequal number of observations or
+unequal variance for different clusters. From Table \ref{table1}, we can see
+that our proposed Gabriel method is robust and has superior performance
+compare with other methods. Because one can hardly tell the underlying
+distribution of data in practice, the resilience to non-Gaussian data gives
+the Gabriel method a clear edge.
+
 \section{Real data application}
-We also applied our proposed method to three real world data sets obtained from the University of California Irvine machine learning repository.
-The first and third data sets are selected because there are clear number of clusters in those two data sets.
-The second data set is used as a benchmark data set since it was widely used in literature.\\
-
-The first one is congress voting data which consists of voting records of $98$th United States Congress, $2$nd session.
-This data set includes votes for each of the U.S.
-House of Representatives Congressmen on the $16$ key votes identified by the \textit{CQA} (Congressional Quarterly Almanac).
-For each vote, each Congressman either vote positively ``yea" (voted for/paired for/announced for),  negatively ``nay'' (voted against/paired against/announced against) or position unknown ``?".
-We took out those records contain ``?" before comparing each algorithm.
-It results in $232$ remaining records, with $124$ democrat and $108$ republican.
-\\
+
+We also applied our proposed method to three real world data sets obtained
+from the University of California Irvine machine learning repository. The
+first and third data sets are selected because there are clear number of
+clusters in those two data sets. The second data set is used as a benchmark
+data set since it was widely used in literature.
+
+The first one is congress voting data which consists of voting records of
+$98$th United States Congress, $2$nd session. This data set includes votes for
+each of the U.S. House of Representatives Congressmen on the $16$ key votes
+identified by the \textit{CQA} (Congressional Quarterly Almanac). For each
+vote, each Congressman either vote positively ``yea" (voted for/paired
+for/announced for),  negatively ``nay'' (voted against/paired
+against/announced against) or position unknown ``?". We took out those records
+contain ``?" before comparing each algorithm. It results in $232$ remaining
+records, with $124$ democrat and $108$ republican.
 
@@
 
-The third data set is the Sonar data returned from two targets -- a metal cylinder and a rock with similar shape, which is first studied by \cite{gorman1988analysis} using a neural network.
-Both targets were impinged by pulse which was a wide-band linear FM chirp $(ka = 55.6)$.
-Returns were collected at a range of $10$ meters and obtained from the cylinder at aspect angles spanning $90^{\circ}$ and from the rock at aspect angles spanning $180^{\circ}$.
-The data set contains $208$ returns ($111$ cylinder returns and $97$ rock returns), with each composed of $60$ spectral samples, normalized to take on values between $0$ and $1$.
-So the data has $60$ features with clearly $2$ clusters.
+The third data set is the Sonar data returned from two targets -- a metal
+cylinder and a rock with similar shape, which is first studied by
+\cite{gorman1988analysis} using a neural network. Both targets were impinged
+by pulse which was a wide-band linear FM chirp $(ka = 55.6)$. Returns were
+collected at a range of $10$ meters and obtained from the cylinder at aspect
+angles spanning $90^{\circ}$ and from the rock at aspect angles spanning
+$180^{\circ}$. The data set contains $208$ returns ($111$ cylinder returns and
+$97$ rock returns), with each composed of $60$ spectral samples, normalized to
+take on values between $0$ and $1$. So the data has $60$ features with clearly
+$2$ clusters.
 
@@
 \end{table}
-\noindent
-Since most congressmen vote base on their parties' interest, $2$ parties (Democratic and Republican) represent two clusters in this data set.
-So the optimal number should be two.
-Close inspection shows \textit{kmeans} with $k=2$ separates the two parties very well with the lowest miss-classification error $10.43\%$ ($k=3$ has $14.78\%$).
-Note that CH-index and Bootstrap stability also return $k=2$, but $2$ is the lower bound those methods can select for $k$.
-So it's not clear they actually choose $k=2$ or they hit the lower bound (they would pick $k=1$ if allowed).
-For the breast cancer data, it's know to have at least $2$ cluster based on whether it's benign specimens or not.
-But it doesn't mean the optimal $k$ should be $2$.
-\cite{fujita2014non} noticed that the malign group is quite heterogeneous and can be further clustered into at least two subgroups.
-Hence, the result $k=3$ given by our proposed Gabriel method (as well as Wold method) make sense.
-For the Sonar data set which is relatively high-dimension, only our proposed Gabriel method and CH-index selects $k=2$.
-Majority of the methods collapse to either select the maximum or select the minimum number, underline how difficult it was to pick the right $k$ when dimension increases.
+
+Since most congressmen vote base on their parties' interest, $2$ parties
+(Democratic and Republican) represent two clusters in this data set. So the
+optimal number should be two. Close inspection shows \textit{kmeans} with
+$k=2$ separates the two parties very well with the lowest miss-classification
+error $10.43\%$ ($k=3$ has $14.78\%$). Note that CH-index and Bootstrap
+stability also return $k=2$, but $2$ is the lower bound those methods can
+select for $k$. So it's not clear they actually choose $k=2$ or they hit the
+lower bound (they would pick $k=1$ if allowed). For the breast cancer data,
+it's know to have at least $2$ cluster based on whether it's benign specimens
+or not. But it doesn't mean the optimal $k$ should be $2$.
+\cite{fujita2014non} noticed that the malign group is quite heterogeneous and
+can be further clustered into at least two subgroups. Hence, the result $k=3$
+given by our proposed Gabriel method (as well as Wold method) make sense. For
+the Sonar data set which is relatively high-dimension, only our proposed
+Gabriel method and CH-index selects $k=2$. Majority of the methods collapse to
+either select the maximum or select the minimum number, underline how
+difficult it was to pick the right $k$ when dimension increases.
+
 \section{Conclusion}
-In this paper, we proposed a novel approach to estimate the number of clusters.
-The intuition behind our proposed methods is to transfer the unsupervised learning problem into supervised learning problem via novel form of cross validation.
-We proved the self-consistency for our proposed Gabriel CV method as well as its asymptotic property with noise, and showed the robustness of our method by simulation.
-Our method has very good performance in our limited simulation settings, and clearly the superior one when data is high dimension or is heavy-tailed.
-\include{...}
+
+In this paper, we proposed a novel approach to estimate the number of
+clusters. The intuition behind our proposed methods is to transfer the
+unsupervised learning problem into supervised learning problem via novel form
+of cross validation. We proved the self-consistency for our proposed Gabriel
+CV method as well as its asymptotic property with noise, and showed the
+robustness of our method by simulation. Our method has very good performance
+in our limited simulation settings, and clearly the superior one when data is
+high dimension or is heavy-tailed.
 

### 2015-06-23 demo/bench/05_tabulate_results.R

@@ -23,10 +23,10 @@ method.names <- list(
     "CH" = "CH",
     "Hartigan" = "Hartigan",
     "Jump" = "Jump",
-    "PS" = "Prediction Strength",
+    "PS" = "Pred.~Strength",
     "Stab" = "Stability",
 #    "gabriel-nearest-2x2" = "Gabriel CV ($2 \\times 2$)",
-    "gabriel-nearest-5x2" = "Gabriel CV ($5 \\times 2$)",
+    "gabriel-nearest-5x2" = "Gabriel CV",
 #    "gabriel-lda-equal-2x2" = "Gabriel CV (LDA Equal; $2 \\times 2$)",
 #    "gabriel-lda-equal-5x2" = "Gabriel CV (LDA Equal; $5 \\times 2$)",
 #    "gabriel-lda-proportion-2x2" = "Gabriel (CV LDA Prop.; $2 \\times 2$)",
@@ -43,15 +43,15 @@ exclude <- function(setting, method)
     FALSE
 }
 
-print_table <- function(setting, summary)
+print_table <- function(setting, summary, pe=FALSE, zero=FALSE)
 {
-    cat("\\textit{", setting.names[[setting]], "} & \\\\\n", sep="")
+    cat("\\textit{", setting.names[[setting]], "} & \\\\*\n", sep="")
 
     kmax <- length(summary[["oracle"]]$counts)
 
     for (m in names(method.names)) {
         s <- summary[[m]]
-        if (is.null(s))
+        if (is.null(s) || (m == "oracle" && !pe))
             next
         cat(method.names[[m]])
         for (k in seq_len(kmax)) {
@@ -61,25 +61,36 @@ print_table <- function(setting, summary)
             } else if (k == nclust[[setting]]) {
                 cat("\\textbf{", s$counts[k], "}", sep="")
             } else {
-                cat(s$counts[k])
+                if (zero || s$counts[k] != 0) {
+                    cat(s$counts[k])
+                } else {
+                    cat("$\\cdot$")
+                }
             }
         }
         if (exclude(setting, m)) {
             cat("& \\textendash")
         } else {
-            cat(" & ", s$count_na, sep="")
+            cat(" & ")
+            if (zero || s$count_na != 0) {
+                cat(s$count_na)
+            } else {
+                cat("$\\cdot$")
+            }
         }
 
-        if (m == "oracle") {
-            cat(" & 1")
-        } else if (exclude(setting, m)) {
-            cat(" & \\textendash")
-        } else {
[…]

### 2015-06-24 demo/bench/05_tabulate_results.R
minor tweaks to simulation results table

@@ -54,10 +54,12 @@ print_table <- function(setting, summary, pe=FALSE, zero=FALSE)
         if (is.null(s) || (m == "oracle" && !pe))
             next
         cat(method.names[[m]])
+        if (exclude(setting, m))
+            cat("\\textsuperscript{$\\ast$}")
         for (k in seq_len(kmax)) {
             cat(" & ")
             if (exclude(setting, m)) {
-                cat("\\textendash")
+                cat("$\\cdot$")
             } else if (k == nclust[[setting]]) {
                 cat("\\textbf{", s$counts[k], "}", sep="")
             } else {
@@ -69,7 +71,7 @@ print_table <- function(setting, summary, pe=FALSE, zero=FALSE)
             }
         }
         if (exclude(setting, m)) {
-            cat("& \\textendash")
+            cat("& $\\cdot$")
         } else {
             cat(" & ")
             if (zero || s$count_na != 0) {
@@ -83,7 +85,7 @@ print_table <- function(setting, summary, pe=FALSE, zero=FALSE)
             if (m == "oracle") {
                 cat(" & 1")
             } else if (exclude(setting, m)) {
-                cat(" & \\textendash")
+                cat(" & $\\cdot$")
             } else {
                 cat(sprintf(" & %.1f $\\pm$ %.1f",
                             s$prederr_rel$mean,

### 2015-06-24 cvclust.tex
minor tweaks to simulation results table
minor tweaks to intro

@@
 \begin{abstract}
-Many clustering methods, including K-means, require the user to specify the number of clusters as an input parameter.
-A variety of methods have been devised to choose the number of clusters automatically, but they often rely on strong modeling assumptions.
-We propose a data-driven approach to estimate the number of clusters based on a novel form of cross-validation.
-This differs from ordinary cross-validation, because clustering is fundamentally an unsupervised learning problem.
-Simulation and real data analysis results show that our proposed method outperforms existing methods, especially in high-dimensional settings with heavy-tailed data.
+Many clustering methods, including $k$-means, require the user to specify the
+number of clusters as an input parameter. A variety of methods have been
+devised to choose the number of clusters automatically, but they often rely on
+strong modeling assumptions. We propose a data-driven approach to estimate the
+number of clusters based on a novel form of cross-validation. This differs
+from ordinary cross-validation, because clustering is fundamentally an
+unsupervised learning problem. Simulation and real data analysis results show
+that our proposed method outperforms existing methods, especially in
+high-dimensional settings with heavy-tailed data.
 \end{abstract}
@@
 \section{Introduction}
-As a main task of exploratory data analysis, clustering organizes unlabeled observations into groups such that observations in same group are more similar compare to those in different group.
-Clustering is an important topic in unsupervised learning because it can reveal the internal structure of data through grouping, segment the data through partitioning and summarize data for other purposes such as dimension reduction.
-It has being widely used in various fields such as psychology, biology, statistics and machine learning including pattern recognition, image segmentation etc.
-
-After being proposed more than $50$ years, $K$-mean remains one of the most
+
+As a main task of exploratory data analysis, clustering organizes unlabeled
+observations into groups such that observations in same group are more similar
+compare to those in different group. Clustering is an important topic in
+unsupervised learning because it can reveal the internal structure of data
+through grouping, segment the data through partitioning and summarize data for
+other purposes such as dimension reduction. It has being widely used in
+various fields such as psychology, biology, statistics and machine learning
+including pattern recognition, image segmentation etc.
+
+After being proposed more than $50$ years, $k$-means remains one of the most
 popular and widely used clustering algorithms \citep{jain2010data}. Like many
-other clustering methods, $K$-mean requires an input parameter $k$, the number
-of clusters, to be specified by the user. Automatically and quantitatively
-deciding such parameter is important and yet unsolved problem
+other clustering methods, $k$-means requires an input parameter $k$, the
+number of clusters, to be specified by the user. Automatically and
+quantitatively deciding such parameter is important and yet unsolved problem
 \citep{fujita2014non}. Various methods have been proposed to tackle this
@@
 (within-cluster dispersion) and the number of cluster $k$ for a certain
-clustering method such as $K$-mean. Since $W_k$ decreases as $k$ increases,
-one usually find the ``elbow" of curve $W_k$ \textsl{VS} $k$ as the
-appropriate number of clusters. The example on the top row of Figure
-\ref{fig:elbow} demonstrates such approach for data with $k=4$, where the
-``elbow" point indeed reveals the true number of clusters. This is based on
-the idea that under partitioning data set has more impact than over
+clustering method such as $k$-means. Since $W_k$ decreases as $k$ increases,
+one usually find the ``elbow" of curve obtain by plotting $W_k$ versus $k$ as
+the appropriate number of clusters. The example on the top row of
+Figure~\ref{fig:elbow} demonstrates such approach for data with $k=4$, where
+the ``elbow" point indeed reveals the true number of clusters. This is based
+on the idea that under partitioning data set has more impact than over
 partitioning data set in terms of $W_k$. However, locating the ``elbow" point
 is somewhat subjective and sometimes is not appropriate to select the optimal
-$k$.
-The second example on the bottom row of Figure \ref{fig:elbow} shows a  situation where there is no clear choice of the ``elbow" point -- both $k=2$ and $k=3$ can be viewed as the ``elbow" point.
-What's more, the true $k=4$ can never be selected as the optimal $k$ using such approach in this case since it can hardly be viewed as the ``elbow" of the curve.
+$k$. The second example on the bottom row of Figure \ref{fig:elbow} shows a
+situation where there is no clear choice of the ``elbow" point -- both $k=2$
+and $k=3$ can be viewed as the ``elbow" point. What's more, the true $k=4$ can
+never be selected as the optimal $k$ using such approach in this case since it
+can hardly be viewed as the ``elbow" of the curve.
 
@@
 
-Recently, there are several new proposals to find the $k$ automatically.
-Gap statistics \citep{tibshirani2001estimating} estimates $k$ by comparing the change in within-cluster dispersion with that expected under an appropriate reference null distribution.
-Specifically, the graph of $log(W_k)$ is compared with its expectation under an appropriate null reference distribution of the data.
-The value of $k$ associated with the largest gap between $log(W_k)$ and the reference curve is selected as optimal $k$.
-\cite{sugar2003finding} proposed an approach which finds the number of clusters based on distortion, a quantity that measures the average distance, per dimension.
-It's backed by a rigorous theoretical justification based on information-theoretic ideas.
-\cite{fraley2002model}'s Model-based method employs the EM algorithm to estimate the parameters in Gaussian mixture model, and select the best model ($k$) using BIC criterion.
-Stability-based criterion is also proposed to locate the best $k$ by some authors such as \cite{ben2001stability}, \cite{wang2010consistent} and \cite{fang2012selection}.
-\cite{chiang2010intelligent} provides a nice review of existing methods for finding the right $k$ in published literature.
-
-Most existing methods are either model based method requires strong parameter assumptions or ad hoc method utilizes within-cluster dispersion.
-Although many view selecting the number of clusters as a model selection problem, very few approaches this problem from the prediction point of view.
-Select model with smallest prediction error via cross-validation is one of the simplest and most widely used model selection techniques in supervised learning.
-The lack of true class (label) in data set makes the adoption of cross-validation into unsupervised leaning problem difficult.
-Perhaps the only exception is \cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction strength.
-The strategy is to first cluster the test data and training data into $k$ clusters respectively.
-Then, for each pair of observations that are assigned to the same test cluster, algorithm determines whether they are also assigned to the same cluster based on the training centers.
-The intuition here is, if $k=k_0$, the true number of clusters, then the $k$ training set clusters will be similar to the $k$ test clusters, and hence will predict them well.
-However, a specifically defined prediction error measure is used in such procedure, which is quite different from the one commonly used in cross-validation procedure in supervised learning.
-Therefore the well understood properties of cross-validation procedure in supervised learning cannot be carried over to the prediction strength method, makes the interpretation of its result difficult.
-
-Our proposed method is a complete data-driven approach which doesn't rely on any parametric assumptions.
-Through novel form of partitioning data set, we are able to employing the cross-validation procedure in clustering exactly the same way as in supervised learning problem.
-Hence, it's easy for reader to understand the intuition behind our proposed method.
-Simulation and real data application shows the superior performance of our proposed method compare with existing methods in high-dimension settings and heavy-tailed data.
-Since the embedded cross-validation procedure is well understood, it also makes our method potentially easily to be extended in future study.
+Recently, there are several new proposals to find the $k$ automatically. Gap
+statistics \citep{tibshirani2001estimating} estimates $k$ by comparing the
+change in within-cluster dispersion with that expected under an appropriate
+reference null distribution. Specifically, the graph of $\log(W_k)$ is
+compared with its expectation under an appropriate null reference distribution
+of the data. The value of $k$ associated with the largest gap between
+$\log(W_k)$ and the reference curve is selected as optimal $k$.
+\citet{sugar2003finding} proposed an approach which finds the number of
+clusters based on distortion, a quantity that measures the average distance,
+per dimension. It's backed by a rigorous theoretical justification based on
+information-theoretic ideas.   \citet{fraley2002model}'s Model-based method
+employs the EM algorithm to estimate the parameters in Gaussian mixture model,
+and select the best model ($k$) using BIC criterion. Stability-based criterion
+is also proposed to locate the best $k$ by some authors such as
+\citet{ben2001stability}, \citet{wang2010consistent} and
+\citet{fang2012selection}. \citet{chiang2010intelligent} provides a nice review
+of existing methods for finding the right $k$ in published literature.
+
+Most existing methods are either model based method requires strong parameter
+assumptions or ad hoc method utilizes within-cluster dispersion. Although many
+view selecting the number of clusters as a model selection problem, very few
+approaches this problem from the prediction point of view. Select model with
+smallest prediction error via cross-validation is one of the simplest and most
+widely used model selection techniques in supervised learning. The lack of
+true class (label) in data set makes the adoption of cross-validation into
+unsupervised leaning problem difficult. Perhaps the only exception is
+\cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction
+strength. The strategy is to first cluster the test data and training data
+into $k$ clusters respectively. Then, for each pair of observations that are
+assigned to the same test cluster, algorithm determines whether they are also
+assigned to the same cluster based on the training centers. The intuition here
+is, if $k=k_0$, the true number of clusters, then the $k$ training set
+clusters will be similar to the $k$ test clusters, and hence will predict them
+well. However, a specifically defined prediction error measure is used in such
+procedure, which is quite different from the one commonly used in
+cross-validation procedure in supervised learning. Therefore the well
+understood properties of cross-validation procedure in supervised learning
+cannot be carried over to the prediction strength method, makes the
+interpretation of its result difficult.
+
+Our proposed method is a complete data-driven approach which doesn't rely on
+any parametric assumptions. Through novel form of partitioning data set, we
+are able to employing the cross-validation procedure in clustering exactly the
+same way as in supervised learning problem. Hence, it's easy for reader to
+understand the intuition behind our proposed method. Simulation and real data
+application shows the superior performance of our proposed method compare with
+existing methods in high-dimension settings and heavy-tailed data. Since the
+embedded cross-validation procedure is well understood, it also makes our
+method potentially easily to be extended in future study.
 
@@
 \begin{center}
-{%\footnotesize
+\footnotesize
 \begin{longtable}{lcccccccccccccccc}
@@
 Method
-  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15
-  & \textsc{na} \\
+  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
 \midrule
@@
   {\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
-\toprule
+\midrule
 \phantom{Pred.~Strength}
@@
 Method
-  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15
-  & \textsc{na} \\
+  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
 \midrule
 \endhead
-%\bottomrule
 \multicolumn{17}{r}{\textit{Continued on next page}} \\
@@
 \bottomrule
-\multicolumn{17}{l}{
-{\footnotesize {In Setting~1, ``\textendash'' means the method can not be used
-with parameter $k=1$}.}}
+\multicolumn{17}{l}{\footnotesize
+  \textsuperscript{$\ast$} CH, Hartigan, and Stability are excluded from
+  Setting~1 because they cannot select $k=1$.
+}
 \endlastfoot
@@
 \end{longtable}
-}
 \end{center}

### 2015-08-15 cvclust.tex
add one more step in proof

@@
 \)
-If $k > 1$, then Theorem~\ref{thm:kmeans-consistency} and
+
+Now suppose that $k > 1$.  Let $\bmuY_1, \dotsc, \bmuY_k$ be the cluster
+centers gotten from applying $k$-means to $Y_{n,1}, \dotsc, Y_{n,n}$.  Since
+the support of $P^Y$ has at least two points,
+Lemma~\ref{lem:kmeans-existence}, implies that almost surely eventually at
+least two of the clusters will be nonempty and have distinct centers.  Without
+loss of generality, assume that $k$ is less than or equal to the cardinality
+of the support of $P^Y$.  Theorem~\ref{thm:kmeans-consistency} and
 Lemma~\ref{lem:kmeans-separation} imply that there exists an $r > 0$ such that
-almost surely eventually at most one cluster center in
-$\{ \bmuY_1, \dotsc, \bmuY_k \}$ is within distance $r$ of $\muY$.  Without
-loss of generality suppose that $\bmuY_1$ is the closest cluster center to
-$\muY$. [TODO: finish proof]
+almost surely eventually at most one cluster center is within distance $r$ of
+$\muY$.  Without loss of generality suppose that $\bmuY_1$ is the closest
+cluster center to $\muY$.
+
+Let $\bmuX_1, \dotsc, \bmuX_k$ be the corresponding sample means computed from
+$X_{n,1}, \dotsc, X_{n,n}$, with the $i$th observation assigned to cluster
+$\hGY_{n,i} = \argmin_g \| Y_{n,i} - \bmuY_{n,i} \|^2.$  Consider the
+classification rule $\hGY(X)$, which assigns $X$ to the class $g$ minimizing
+$\|X - \bmuX_g\|^2$, breaking ties uniformly at random between classes of
+equal distance.
+
+[TODO: finish]
+
 \end{proof}

### 2015-08-20 cvclust.tex
remove kmeans asymptotics

@@
 = Y_i$ for $i = 1, \dotsc, m+n$, so that $\CV(k) = 0$.
-\end{proof}
-
-\section{Asymptotic Analysis of $k$-means}
-
-To analyze the asymptotic properties of of Gabriel cross validation, we first
-need to understand the properties of the $k$-means procedure.  We will use
-results from \citet{pollard1981strong}, who analyzed $k$-means and showed that
-the within-cluster sum of squares converges to a population quantity.  To
-introduce Pollard's result, for any finite set $A \subseteq \R^d$ and
-probability measure $P$ supported on $\R^d$, define the population
-within-cluster sum of squares as
-\[
-  W(A, P) = \int \min_{a \in A} \|x - a\|^2 P(dx).
-\]
-Define
-\(
-  \sA_k
-    =
-    \{ A \subseteq \R^d : \text{$A$ contains $k$ or fewer points} \}
-\)
-and set
-\(
-  m_k(P) = \inf_{A \in \sA_k}  W(A, P).
-\)
-Suppose that $X_1, \dotsc, X_n$ are independent draws from $P$, and that $P_n$
-is the empirical measure formed by placing mass $1/n$ on each $X_i$.  Let
-$A_n$ be a set minimizing $W(A_n, P_n)$.  Pollard gave conditions for $A_n$ to
-converge to a minimizer of $W(\cdot, P)$.  His main theorem assumes that the
-minimizer of $W(\cdot, P)$ is unique, which is too strong for our purposes
-(this fails, for example, with a bivariate normal distribution with identity
-covariance whenever $k \geq 2$).  It is straightforward to generalize his
-result to cases with multiple maximizers.
-
-To state the generalized version of Pollard's consistency theorem, define
-\[
-  \sbA_k = \{ \bar A \subseteq \R^d
-            : W(\bar A, P) = m_k(P)
-              \text{ and $\bar A$ contains $k$ or fewer points} \}.
-\]
-We will later show that $\sbA_k$ is nonempty and compact whenever $\int
-\|x\|^2 P(dx) < \infty$.  Let $H(\cdot, \cdot)$ denote the Hausdorff metric,
-defined for compact subsets $A, B$ of $\R^d$ by $H(A,B) < \delta$ if
-and only if every point of $A$ is within Euclidean distance $\delta$ of at
-least one point of $B$, and vice versa.  We can now state the generalized
-$k$-means consistency theorem.
-
-\begin{theorem}[$k$-Means Consistency]\label{thm:kmeans-consistency}
-Suppose that $\int \|x\|^2 P(dx) < \infty$ and that $A_n$ minimizes $W(\cdot,
-P_n)$ over $\sA_k$.  Then for all $n$ there exists an $\bar A_n \in \sbA_k$
-such that $H(A_n, \bar A_n) \to 0\text{ a.s.}$, and $W(A_n, P_n) \to
-m_k(P)\text{ a.s.}$ \end{theorem}
-
-The proof of this theorem relies on a series of lemmas, which are extensions
-of results given in Pollard's original article.  First, we show that the map
-$A \mapsto W(A, P)$ is continuous.  Next, we show that $\sbA_k$ is nonempty
-and compact.  Then, it follows that for $n$ large enough, all minimizers of
-$W(\cdot, A_n)$ are contained in a compact set $\sE_k$.  The final ingredient
-of the proof is Pollard's uniform strong law of large numbers, which says that
-$\sup_{A \in \sE_k} |W(A,P) - W(A,P_n)| \to 0\text{ a.s.}$
-
-\begin{lemma}[Continuity]\label{lem:kmeans-continuous}
-If $\int \|x\|^2 P(x) < \infty$, then the map $A \mapsto W(A, P)$ is
-continuous on $\sA_k$ with respect to the topology induced by the Hausdoff
-metric.
-\end{lemma}
-\begin{proof}
-This is a slight extension of the argument given at the end of Pollard's
-Section~4.  If $A, B \in \sA_k$ and $H(A, B) < \delta$, then for each $b \in
-B$ there exists an $a(b) \in A$ such that $\|b - a(b)\| < \delta$.  Then
-\begin{align*}
-  W(A, P) - W(B, P)
-    &= \int \min_{a \in A} \|x - a\|^2 - \min_{b \in B} \|x - b\|^2 P(dx) \\
-    &\leq \int \max_{b \in B} \{ \|x - a(b)\|^2 - \|x - b\|^2 \} P(dx) \\
-    &\leq \int \sum_{b \in B} \{ (\|x - b\| + \delta)^2 - \|x - b\|^2 \} P(dx) \\
-    &\leq \int \sum_{b \in B}  ( 2 \delta \|x - b\| + \delta^2) P(dx) \\
-    &\leq 2 \delta \{M^{1/2} + k \max_{b \in B} \|b\|\} + k \delta^2,
-\end{align*}
-where $M = \int \|x\|^2 P(dx)$.  Thus,
-\[
-  |W(A, P) - W(B, P)|
-    \leq 2 \delta \{M^{1/2} + k \max_{a \in A} \|a\|\} + 3 k \delta^2.
-\]
-Hence if $A \in \sA_k$ is any fixed set, then for every $\varepsilon > 0$, we
-can find a $\delta$ small enough so that if $B \in \sA_k$ and $H(A, B) <
-\delta$, then $|W(A,P) - W(B,P)| < \epsilon$.  Thus, $W(\cdot, P)$ is
-continuous at $A$.
-\end{proof}
-
-\begin{lemma}[Existence]\label{lem:kmeans-existence}
-If $\int \|x\|^2 P(dx) < \infty$ and the support of $P$ contains at least $k$
-points, then
-\begin{enumerate}
-  \item $m_1(P) > m_2(P) > \dotsb > m_k(P)$;
-  \item $W(\cdot, P)$ achieves its infimum over $\sA_j$ for $j = 1, \dotsc,
-    k$;
-  \item the class of sets $\sbA_j = \{ A \in \sA_j : W(A,P) = m_{j}(P) \}$
-    is compact (with respect to the Hausdorff metric on $\sA_j$) for $j = 1,
-    \dotsc, k$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
-This argument is adapted from the proof in Section~3 of Pollard's paper.  We
-first find an $M$ large enough to ensure that every minimizer of $W(\cdot, P)$
-must have at least one point in $B(M)$.  Next, we impose additional
-constraints on $M$ to ensure that all minimizers of $W(\cdot, P)$ are
-contained in $B(5M)$.
-
-Pick~$r$ so that the ball~$K$ of radius~$r$ centered at the origin has
-positive $P$~measure.  Take $M$ to be large enough such that
-\(
-  (M - r)^2 P(K) > \int \|x\|^2 P(dx).
-\)
-Let $\sC_k$ denote the class of sets $A$ with at most $k$ points and at least
-one point in in $B(M)$, the closed ball with radius $M$ centered at the
-origin.  Let $A_0$ be the singleton set containing the origin.  If no point of
-$A$ is in $B(M)$, then
-\(
-  W(A, P) \geq (M - r)^2 P(K) > W(A_0, P).
-\)
-Put concisely, if $A \in \sA_k \setminus \sC_k$, then
-\(
-  W(A, P) > W(A_0, P) \geq m_k(P).
-\)
-
-Next, we proceed by induction on $k$.  If $k = 1$, then there is a unique
-minimizer of $W(\cdot, P)$ over $\sA_k$, the set containing the mean $\int x
-P(dx)$.  For $k > 1$, take as given that $m_{k-1}(P) = W(\bar A_{k-1}, P)$ for
-some set $\bar A_{k-1} \in \sE_{k-1}$.  We will show that there is an $M$
-large enough so that $W(\cdot, P)$ attains its infimum over $\sA_k$ and all
-minimizers lie in $\sE_k$.  Since $A \mapsto W(\cdot, P)$ is continuous and
-$\sE_k$ is compact, the class $\sbA_k$, the preimage of $\{ m_k(P) \}$ under
-this map, is compact.
-
-We first show that if $M$ is large enough, then there exists a set $B \in
-\sE_k$ and a positive $\varepsilon$ with $W(B,P) < m_{k-1}(P) - \varepsilon$.
-This will establish that $m_{k}(P) < m_{k-1}(P)$, and the set $B$ will have
-further utility later in the proof.  Take $\bar A_{k-1} \in \sE_{k-1}$ to be a
-set with at most $k-1$ points satisfying $m_{k-1}(P) = W(\bar A_{k-1},P)$.
-Since the support of $P$ contains least $k$ points, there exists a point~$b_1$
-not in $\bar A_{k-1}$ that lies in the support of $P$.  Ensure that $M$ is
-large enough so that $b_1 \in B(5M)$.  Take $B \in \sE_k$ to be the set of $k$
-points gotten by adjoining $b_1$ to $\bar A_{k-1}$.  Let
-\(
-  \delta = \min_{a \in \bar A_{k-1}} \| b_1 - a \|
-\)
-and set $R = \{ x \in \R^d : \|x - b_1\| < \delta/3 \}$; since $b_1$
-is in the support of $P$, the neighborhood $R$ has positive $P$~measure.
-Further,
-\[
-  \int_{R} \min_{a \in \bar A_{k-1}} \|x - a\|^2 P(dx)
-    \geq
-      \int_{R} \min_{a \in \bar A_{k-1}} (\|b_1 - a\| - \delta/3)^2 P(dx)
-    =
-      (4/9) \delta^2 P(R),
-\]
-and
-\[
-  \int_{R} \min_{b \in B} \|x - b\|^2 P(dx)
-    =
-      \int_{R} \|x - b_1\|^2 P(dx)
-    \leq
-      (1/9) \delta^2 P(R).
-\]
-Thus, if
-\(
-  \varepsilon < (3/9) \delta^2 P(R)
-\)
-then
-\(
-  W(B,P)
-    \leq
-      W(\bar A_{k-1},P) - (3/9) \delta^2 P(R)
-    <
-      m_{k-1}(P) - \varepsilon.
-\)
-
-Now, ensure that $M$ large enough so that
-\[
-  4 \int_{\|x\| \geq 2M} \|x\|^2 P(dx) < \varepsilon.
-\]
-We will now show that if $A \in \sC_k \setminus \sE_k$, then $W(A,P) > W(B,P)
-\geq m_{k}(P)$, which ensures that $W(\cdot, P)$ does not attain its infimum
-over $\sA_k$ outside of $\sE_k$.  To see that this holds, suppose that $A$ is
-any set $\sC_k \setminus \sE_k$.  This set has at least one point, $a_1$, in
-$B(M)$, and at least one point outside $B(5M)$.  Consider the set $A^\ast$
-formed by removing all points of $A$ outside $B(5M)$.  Any point $x$ closer to
-a cluster center outside $B(5M)$ than to $a_1$ must satisfy $\|x\| \geq 2M$.
-Thus,
-\begin{align*}
-  W(A^\ast, P)
-    &\leq W(A,P) + \int_{\|x\| \geq 2M} \|x - a_1\|^2 P(dx) \\
-    &\leq W(A,P) + \int_{\|x\| \geq 2M} (\|x\| + \|a_1\|)^2 P(dx) \\
-    &\leq W(A,P) + 4 \int_{\|x\| \geq 2M} \|x\|^2 P(dx) \\
-    &< W(A,P) + \varepsilon.
-\end{align*}
-Since $A^\ast$ has $k-1$ or fewer points, it must satisfy $W(A^\ast, P) \geq
-m_{k-1}(P)$, so that $W(A, P) > m_{k-1}(P) - \varepsilon > W(B, P)$.
-
-We have now established that $m_k(P) = \inf \{ W(A,P) : A \in \sE_k \}$, and
-that $W(\cdot, P)$ does not obtain its infimum over $\sA_k$ outside of
-$\sE_k$.  The final step of the proof is to show $W(\cdot, P)$ obtains its
-infimum over $\sE_k$.  This follows from the extreme value theorem, since
-$\sE_k$ is compact and, by Lemma~\ref{lem:kmeans-continuous}, the map $A
-\mapsto W(A,P)$ is continuous over $\sA_k$.
-\end{proof}
-
-\begin{lemma}
-\label{lem:pollard-compact}
-Suppose that $\int \|x\|^2 P(dx) < \infty$, that the support of $P$ has at
-least $k$ points, and that $k > 1$.  Let $A_n$ minimize $W(\cdot, P_n)$ over
-$\sA_k$, and that $B_n$ minimize $W(\cdot P_n)$ over $\sA_{k-1}$.  If $W(B_n,
-P) \to m_{k-1}(P)$ almost surely as $n \to \infty$, then there exists a finite
-$M$ such that as if $n$ is large enough, then, almost surely, $A_n$ is in the
-class of sets
-\(
-  \sE_k = \{ A \subseteq B(5M) : \text{$A$ contains $k$ or fewer points} \},
-\)
-where $B(5M)$ denotes the Euclidean ball of radius $5M$ centered at the
-origin.
-\end{lemma}
-\begin{proof}
-Pollard proves this result, in Section~3 of his paper,  assuming that there is
-a unique set of $j$ points minimizing $W(\cdot, P)$ for each value of $j = 1,
-\dotsc, k$.  His proof, which is very similar to the proof of
-Lemma~\ref{lem:kmeans-existence}, does not use this assumption directly, only
-its implication that $m_{k}(P) < m_{k-1}(P)$.  Since we have established his
-result in Lemma~\ref{lem:kmeans-existence}, the remainder of his proof applies
-without further modification.
-\end{proof}
-
-\begin{lemma}[Uniform SLLN]\label{lem:kmeans-uniform-slln}
-If $\int \|x\|^2 P(x) < \infty$, then
-\(
-  \sup_{A \in \sE_k} |W(A, P) - W(A,P_n)| \to 0\text{ a.s.}
-\)
-\end{lemma}
-\begin{proof}
-Pollard proves this result at the beginning of his Section~4.
-\end{proof}
-
-With Lemmas~\ref{lem:kmeans-continuous}--\ref{lem:kmeans-uniform-slln}, we can
-proceed to the proof of Theorem~\ref{thm:kmeans-consistency}.
-
-\begin{proof}[Proof of Theorem~\ref{thm:kmeans-consistency}]
-The proof proceeds inductively over $k$.  For $k=1$, the result follows from
-the strong law of large numbers.  For $k > 1$, assume that the result holds
-for smaller values of $k$.  If the support of $P$ has fewer than $k$ points,
-then the result is trivially true.  Otherwise, the induction hypothesis
-implies that there exists a $B_n \in \sA_{k-1}$ with $W(B_n, P_n) \to
-m_{k-1}(P)\text{ a.s.}$  Hence, Lemma~\ref{lem:pollard-compact} ensures that
-there is a finite $M$ such that $A_n \in \sE_k$ almost surely for large
-enough~$n$.  Take $M$ to be large enough so that $\sE_k$ also contains
-$\sbA_k$.
-
-The set $\sE_k$ is compact and the map $f: \sE_k \to \R$ with $f(A) =
-W(A, P)$ is continuous.  Thus, for every $\varepsilon > 0$ there exists an
-$\eta > 0$ depending on $\varepsilon$ such that if $A \in \sE_k$ and $H(A,
-\bar A) \geq \varepsilon$ for every $\bar A \in \sbA_k$, then $W(A,P) \geq
-m_{k}(P) + \eta$.  Let $\bar A_0$ be any member of $\sbA_k$.  By construction,
-$W(A_n, P_n) \leq W(\bar A_0, P_n)$.  Lemma~\ref{lem:kmeans-uniform-slln}
-ensures that $W(\bar A_0, P_n) \to m_k(P)\text{ a.s.}$  Thus, for $n$ large
-enough, $W(A_n, P_n) < m_{k}(P) + \eta$, which implies that there exists an
-$\bar A_n \in \sbA_k$ with $H(A_n, \bar A_n) < \varepsilon$.
 \end{proof}

### 2015-08-20 cvclust.tex
move counterexample to main text

@@
 \newcommand{\toas}{\overset{\mathit{a.s.}}{\to}}
+
+\newcommand{\OhP}{O_p}
 
@@
 
-\section{Asymptotic Analysis of Gabriel Cross-Validation}
+\section{Analysis of Gabriel Cross-Validation with Gaussian Noise}
 
@@
 
-Here, we show that Gabriel Cross-Validation is consistent when the data come
-from a single cluster; the algorithm picks $k = 1$ with probability tending to
-$1$ as the number of rows in the data matrix $\dataX$ increase.
+Now we we analyze the asymptotic performance of Gabriel Cross-Validation, in
+the case of Gaussian noise. Our main result is that with single-cluster
+Gaussian data, if the predictor and
+response columns of $\dataX$ are weakly correlated or independent, then the
+method will correctly choose to $k = 1$ cluster. We state the result in the
+case where $\dataX$ has 2 columns, but it is straightforward to generalize
+this result to higher dimensions.
 
-To prove consistency of Gabriel cross-validation, it suffices to analyze a
-single fold of the cross-validation procedure.  We will assume the following
-setup:
+\begin{proposition}
+Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is that data from a single fold
+of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
+independent draw from a mean-zero multivariate normal distribution with unit
+marginal variances and correlation $\rho$.  In this case, the data are drawn
+from a single cluster; the true number of clusters is~$1$.  If if $\rho <
+0.5$, then $k = 1$ will be preferred to $k = 2$ for large enough $n$ and $m$.
+\end{proposition}
 
-\begin{theorem}
-For each $n$, let $m(n)$ be a nonnegative integer with $m(n) \to \infty$ as $n
-\to \infty$, and let $\{ (X_{n,i}, Y_{n,i}) \}_{i=1}^{n+m(n)}$ be independent
-identically distributed random variables.  For any fixed $k$, let $\CV(k)$
-be the cross-validation error from a single Gabriel CV fold, computed as
-described in Sec.~\ref{sec:gabriel-cv-algorithm}.  Suppose that
-\begin{enumerate}[label=(\roman*)]
-\item the distribution for each $(X, Y)$ pair is such that $X$ and $Y$ have
-  marginal distributions $P^X$ and $P^Y$, and the vectors $X$ and $Y$ are
-  independent of each other;
+\begin{proof}(Outline.)
 
-\item the distribution $P^Y$ satisfies $\int \|y\|^2 P^Y(dy) < \infty$;
+Given $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to $\{ Y_i
+\}_{i=1}^{n}$. With $k = 1$, the single-cluster centroid will
+be equal to $\bar Y_n$, the sample mean of the $Y_1, \dotsc, Y_n$, approximately equal to
+$\E(Y) = 0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error
+will be
+\[
+  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
+         = 1 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
+\]
 
-\item the support of $P^Y$ contains at least two points.
-\end{enumerate}
-Then for any fixed $k > 1$, almost surely eventually  $\CV(k) > \CV(1)$.
-\end{theorem}
-\begin{proof}
-Set $\muY = \E(Y)$ and $\SigmaY = \cov(Y)$.  The strong law of large
-numbers implies that
-\(
-  \CV(1) \toas \int \|y - \mu^Y\|^2 P(dy) = \tr(\SigmaY).
-\)
+Now we will consider the $k = 2$ case.  If $n$ is large enough, then
+\citet{pollard1981strong} showed that the centroids $\bmuY_1$ and $\bmuY_2$
+will be close to $\E(Y \mid Y > 0) = \sqrt{2/\pi}$ and $\E(Y \mid Y < 0) =
+-\sqrt{2/\pi}$.  We have used Lemma~\ref{lem:truncated-normal-moments}
+(Appendix~\ref{app:technical-lemmas}) to compute the expectations.  Further,
+\citet{pollardclt} showed that the errors will be of size $\OhP(n^{-1/2})$.
 
-Now suppose that $k > 1$.  Let $\bmuY_1, \dotsc, \bmuY_k$ be the cluster
-centers gotten from applying $k$-means to $Y_{n,1}, \dotsc, Y_{n,n}$.  Since
-the support of $P^Y$ has at least two points,
-Lemma~\ref{lem:kmeans-existence}, implies that almost surely eventually at
-least two of the clusters will be nonempty and have distinct centers.  Without
-loss of generality, assume that $k$ is less than or equal to the cardinality
-of the support of $P^Y$.  Theorem~\ref{thm:kmeans-consistency} and
-Lemma~\ref{lem:kmeans-separation} imply that there exists an $r > 0$ such that
-almost surely eventually at most one cluster center is within distance $r$ of
-$\muY$.  Without loss of generality suppose that $\bmuY_1$ is the closest
-cluster center to $\muY$.
+If $\rho > 0$ and $n$ is large enough, then classification rule learned from
+$\{(X_i, \hGY_i)\}_{i=1}^{n}$ variables will be determined according to
+whether $X > 0$; if $\rho < 0$ then the decision is according to whether $X <
+0$.  More specifically, the decision boundary will be at $0 + \OhP(n^{-1/2})$.
 
-Let $\bmuX_1, \dotsc, \bmuX_k$ be the corresponding sample means computed from
-$X_{n,1}, \dotsc, X_{n,n}$, with the $i$th observation assigned to cluster
-$\hGY_{n,i} = \argmin_g \| Y_{n,i} - \bmuY_{n,i} \|^2.$  Consider the
-classification rule $\hGY(X)$, which assigns $X$ to the class $g$ minimizing
-$\|X - \bmuX_g\|^2$, breaking ties uniformly at random between classes of
-equal distance.
+In the $\rho > 0$ case, the cross-validation error will be
+\begin{align*}
+  \CV(2)
+  &=
+    \frac{1}{m}
+    \sum_{i=n+1}^{n+m}
+      \| (Y_i - \bmuY_1) 1\{\hGX_i = 1\} \|^2
+      +
+      \| (Y_i - \bmuY_2) 1\{\hGX_i = 2\} \|^2
+\\
+  &=
+    \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]
+      + \OhP(m^{-1/2}) + \OhP(n^{-1/2}),
+\end{align*}
+where $a = \sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows
+that $Y \mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$, so
+that $\E[(Y - a)^2 \mid X] = (\rho X - a)^2 + (1 - \rho^2)$.  Applying
+Lemma~\ref{lem:truncated-normal-moments}, we get that for large $m$ and $n$,
+the Gabriel cross-validation error is close to
+$1 + a^2 (1 - 2 \rho)$.
 
-[TODO: finish]
-
+In the $\rho < 0$ case, a similar calculation shows that $\CV(2)$ is close to
+$1 + a^2 (1 + 2 \rho)$.  In particular, if $|\rho| < 0.5$, then with
+probability tending to $1$ and $m$ and $n$ increase, the asymptotic
+cross-validation error for $k = 1$ will be smaller than for $k = 2$.
 \end{proof}
 
-\subsection{Well-Separated Clusters with Noise}
+We confirm this result with a simulation.  We perform $10$ replicates.  In each
+replicate, we generate $20000$ observations from a mean-zero bivariate normal
+distribution with unit marginal variances and correlation $\rho$.  We perform
+a single $2 \times 2$ fold of Gabriel cross-validation and report the
+cross-validation mean squared error for the number of clusters $k$ ranging
+from $1$ to $5$.  Figure~\ref{fig:nullcorr-equal} shows the cross-validation
+errors for all $10$ replicates.  The simulation demonstrates that in the
+Gabriel cross-validation criterion chooses the correct answer $k = 1$ whenever
+$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.
+
+\begin{figure}
+\centering
+\includegraphics[width=25pc]{demo/nullcorr/equal.pdf}
+\caption{Cross-validation error on $10$ replicates, with the number of
+clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
+multivariate normal distribution with correlation $\rho$.  The Gabriel
+cross-validation criterion chooses the correct answer $k = 1$ whenever
+$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
+\label{fig:nullcorr-equal}
+\end{figure}
+
+\subsection{Two Clusters}
 
@@
 \section{Technical Lemmas}
+\label{app:technical-lemmas}
 
@@
 
-\section{Counterexample with correlated noise}
-
-Suppose that $(X, Y)$ are jointly mean-zero multivariate normal with unit
-marginal variances and correlation $\rho$.  In this case, the data are drawn
-from a single cluster; the true number of clusters is~$1$.  We will show that
-Gabriel cross-validation is inconsistent if $\rho > 0.5$, in the sense that $k
-= 2$ will be preferred to $k = 1$ almost surely for large enough $n$.
-
-Given a sample $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to
-$\{ Y_i \}_{i=1}^{n}$.  If $k = 1$ and $n$ is large enough, then the centroid
-will be close to $\E(Y) = 0$, in the sense that the centroid converges almost
-surely to this value as $n$ tends to infinity \citep{pollard1981strong}.
-Similarly, if $k = 2$ and $n$ is large enough, then the centroids will tend to
-$\E(Y \mid Y > 0) = \sqrt{2/\pi}$ and $\E(Y \mid Y < 0) = -\sqrt{2/\pi}$.  We
-have used Lemma~\ref{lem:truncated-normal-moments} to compute the
-expectations.  In $k = 2$ case, then if $n$ is large enough, the cluster
-assignment will be determined according to whether $Y > 0$.
-
-In the $k = 1$ case, if $n$ is large enough, the average Gabriel
-cross-validation error converges to $\E(Y^2) = 1$. In the $k = 2$ case, if $n$
-is large enough and $\rho > 0$, then the linear decision rule for the $X$
-variables will be determined according to whether $X > 0$; if $\rho < 0$ then
-the decision is according to whether $X < 0$.  In the $\rho > 0$ case, the
-error converges to $\E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]$,
-where $a = \sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows
-that $Y \mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$, so
-that $\E[(Y - a)^2 \mid X] = (\rho X - a)^2 + (1 - \rho^2)$.  Applying
-Lemma~\ref{lem:truncated-normal-moments}, in the $k = 2$ case we get that the
-Gabriel cross-validation error converges to $1 + a^2 (1 - 2 \rho)$.  In the
-$\rho < 0$ case, it converges to $1 + a^2 (1 + 2 \rho)$.  In particular, if
-$|\rho| > 0.5$, then the asymptotic cross-validation error for $k = 2$ will be
-smaller than for $k = 1$.
-
-We confirm this with a simulation.  We perform $10$ replicates.  In each
-replicate, we generate $20000$ observations from a mean-zero bivariate normal
-distribution with unit marginal variances and correlation $\rho$.  We perform
-a single $2 \times 2$ fold of Gabriel cross-validation and report the
-cross-validation mean squared error for the number of clusters $k$ ranging
-from $1$ to $5$.  Figure~\ref{fig:nullcorr-equal} shows the cross-validation
-errors for all $10$ replicates.  The simulation demonstrates that in the
-Gabriel cross-validation criterion chooses the correct answer $k = 1$ whenever
-$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.
-
-\begin{figure}
-\centering
-\includegraphics[width=25pc]{demo/nullcorr/equal.pdf}
-\caption{Cross-validation error on $10$ replicates, with the number of
-clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
-multivariate normal distribution with correlation $\rho$.  The Gabriel
-cross-validation criterion chooses the correct answer $k = 1$ whenever
-$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
-\label{fig:nullcorr-equal}
-\end{figure}
-
 \bibliography{references}

### 2015-08-20 cvclust.tex
finish single-cluster; start 2 cluster

@@
 the case of Gaussian noise. Our main result is that with single-cluster
-Gaussian data, if the predictor and
-response columns of $\dataX$ are weakly correlated or independent, then the
-method will correctly choose to $k = 1$ cluster. We state the result in the
-case where $\dataX$ has 2 columns, but it is straightforward to generalize
-this result to higher dimensions.
+Gaussian data, if the predictor and response columns of $\dataX$ are weakly
+correlated or independent, then the method will correctly prefer $k = 1$ to
+$k = 2$ clusters. We state the result in the case where $\dataX$ has 2 columns,
+but it is straightforward to generalize this result to higher dimensions.
 
 \begin{proposition}
-Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is that data from a single fold
+Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
 of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
@@
 marginal variances and correlation $\rho$.  In this case, the data are drawn
-from a single cluster; the true number of clusters is~$1$.  If if $\rho <
-0.5$, then $k = 1$ will be preferred to $k = 2$ for large enough $n$ and $m$.
+from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 0.5$,
+then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
 \end{proposition}
 
-\begin{proof}(Outline.)
+\begin{proof}
 
@@
 (Appendix~\ref{app:technical-lemmas}) to compute the expectations.  Further,
-\citet{pollardclt} showed that the errors will be of size $\OhP(n^{-1/2})$.
+\citet{pollard1982central} showed that the errors will be of size $\OhP(n^{-1/2})$.
 
@@
 \subsection{Two Clusters}
+
+We will now analyze a simple two-cluster setting, and derive conditions for
+Gabriel cross-validation to correctly prefer $k=2$ clusters to $k=1$.
+
+\begin{proposition}
+Suppose that $\{(X_i,Y_i)\}_{i=1}^{n+m}$ is data from a single fold of Gabriel
+cross-validation, where each $(X,Y)$ pair in $\R^2$ is an independent draw
+from an equiprobable mixture of two multivariate normal distributions with
+identity covariance. Suppose that the first mixture component has mean $(\muX,
+\muY)$, and the second has mean $(-\muX, -\muY)$, where $\muX > 0$ and
+$\muY > 0$.  If [Wei, add condition here], then $\CV(2) < \CV(1)$ with
+probability tending to one as $m$ and $n$ increase.
+\end{proposition}
+
+\begin{proof}
+  [Wei, add proof here; you can take most of this from your document,
+  ``LDA\_prior\_equal.tex'']
+\end{proof}
+
+We confirm the proposition with a simulation.
+
+[Wei, describe simulation here, and include the figure]
 

### 2015-08-20 cvclust.tex
remove old lemma

@@
 
-The following lemma is not required for the $k$-means consistency theorem, but
-it will prove necessary for analyzing the asymptotic behavior of Gabriel
-cross-validation.
-
-\begin{lemma}\label{lem:kmeans-separation}
-Set $\mu = \int x P(dx)$.  If $\int \|x\|^2 P(dx) < \infty$, then there exists
-an $r > 0$ such that for every $A \in \sbA_k$, there is at most one center $a
-\in A$ with $\|a - \mu\| \leq r$.
-\end{lemma}
-\begin{proof}
-If the support of $P$ has fewer than $k$ points, then the result is trivial.
-Otherwise, Lemma~\ref{lem:kmeans-existence} implies that $m_k(P) <
-m_{k-1}(P)$.  Define
-\[
-  d =
-    \inf_{A \in \sA_k}
-      \min \{ \|a - a'\| : a,a' \in A, a \neq a' \}.
-\]
-If $d = 0$, then there exists a sequence $A_n$ in $\sbA_k$ with
-\[
-  \min_{\substack{a,a' \in A_n \\ a \neq a'}} \|a - a'\| \to 0.
-\]
-Since $\sbA_k$ is compact, there exists a subsequence $n(m)$ and a set $A_0
-\in \sbA_k$ with $A_{n(m)} \to A_0$.  Take $B_n \in \sA_{k-1}$ to be the set
-formed by removing one of the centers $a'$ minimizing $\min_{\substack{a \in
-A_n \\ a \neq a'}} \|a - a'\|$.  By construction, $H(A_n, B_n) \to 0$, so that
-$B_{n(m)} \to A_0$.  Since the map $A \mapsto W(A, P)$ is continuous on
-$\sA_k$, we must have $W(B_{n(m)}, P) \to W(A_0, P) = m_k(P)$.  However, this
-is impossible since $W(B_n,P) \geq m_{k-1}(P) > m_k(P)$.  It must be the case,
-then, that $d > 0$.
-
-Set $r < d/2$, and let $A \in \sbA_k$ be arbitrary.  Take $a_1$ to be the
-cluster center in $A$ closest to $\mu$ (breaking ties arbitrarily).  If $\|a_1
-- \mu\| > r$, then $\|a - \mu\| > r$ for all centers $a \in A$.  Otherwise, if
-$\|a_1 - \mu\| \leq r$ and $a \in A$ is any other cluster center, then
-\(
-  2r < \|a - a_1\| \leq \|a - \mu\| + \|a_1 - \mu\|
-\)
-so that
-\(
-  \|a - \mu\| > r.
-\)
-\end{proof}
-
 \section{Analysis of Gabriel Cross-Validation with Gaussian Noise}

### 2015-08-22 cvclust.tex
Add proof and complete .tex file

@@
 Most existing methods are either model based method requires strong parameter
-assumptions or ad hoc method utilizes within-cluster dispersion. Although many
-view selecting the number of clusters as a model selection problem, very few
+assumptions or otherwise lack of clear interpretation and theoretical justification.
+Although many view selecting the number of clusters as a model selection problem, very few
 approaches this problem from the prediction point of view. Select model with
@@
 true class (label) in data set makes the adoption of cross-validation into
-unsupervised leaning problem difficult. Perhaps the only exception is
+unsupervised leaning problem difficult. One exception is
 \cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction
@@
 cannot be carried over to the prediction strength method, makes the
-interpretation of its result difficult.
+interpretation of its result difficult. \citet{wang2010consistent} also uses 
+cross-validation to select the optimal $k$. Instead of selecting $k$ which minimizes 
+prediction error, such method picks $k$ which minimizes the specifically defined clustering instability.
 
@@
 
-\begin{figure}
+\begin{figure}[H]
 \centering
@@
 \muY)$, and the second has mean $(-\muX, -\muY)$, where $\muX > 0$ and
-$\muY > 0$.  If [Wei, add condition here], then $\CV(2) < \CV(1)$ with
+$\muY > 0$.  If $1 + 2\Phi(\mu^Y)+ \frac{2\varphi(\mu^Y)}{\mu^Y} < 4\Phi(\mu^X)$, then $\CV(2) < \CV(1)$ with
 probability tending to one as $m$ and $n$ increase.
@@
 \begin{proof}
-  [Wei, add proof here; you can take most of this from your document,
-  ``LDA\_prior\_equal.tex'']
+There are two clusters $G_1$ and $G_2$, where observations from $G_1$ are distributed as
+\[	\mathcal{N}\left( \begin{pmatrix} 
+    \mu^X \\
+    \mu^Y \\
+  \end{pmatrix}, \mathbf{I} \right)	\]
+and observations from $G_2$ are distributed as
+\[	\mathcal{N}\left( \begin{pmatrix} 
+    -\mu^X \\
+    -\mu^Y \\
+  \end{pmatrix}, \mathbf{I} \right)	\]
+where $\mu^X_1 > 0$ and $\mu^Y_1 > 0$. Let $G_i$ be the true cluster where observation $i$ is generated from, by assumption
+\[	P(G_i=G_1) = P(G_i=G_2) = 1/2	\]
+After applying $k$-means on $\{ Y_i\}_{i=1}^{n}$ with $k=2$, if $n$ is large enough, we have the estimated centroids $\bar{\mu}^Y_1$ and $\bar{\mu}^Y_2$ be close to $E(Y \mid Y>0)$ and $\E(Y \mid Y < 0)$,
+with errors will be of size $\OhP(n^{-1/2})$. Here
+\begin{align}
+E(Y \mid Y>0) &= E(Y_1 \mid Y_1 > 0) \cdot P(Y_1>0) + E(Y_2 \mid Y_2 > 0) \cdot P(Y_2>0) \nonumber \\ 
+   &= 2\varphi(\mu^Y)+2\mu^Y\Phi(\mu^Y)-\mu^Y  \\  \nonumber 
+\end{align}
+where $Y_1 \sim N(\mu^Y, 1)$ and $Y_2 \sim N(-\mu^Y, 1)$, and we used Lemma~\ref{lem:truncated-normal-moments}
+(Appendix~\ref{app:technical-lemmas}). Similarly, we have
+\begin{align}
+E(Y \mid Y<0) &= E(Y_1 \mid Y_1 < 0) \cdot P(Y_1<0) + E(Y_2 \mid Y_2 < 0) \cdot P(Y_2<0) \nonumber \\ 
+   &= -2\varphi(\mu^Y)-2\mu^Y\Phi(\mu^Y)+\mu^Y  \\  \nonumber 
+\end{align}
+ where $\varphi()$ and $\Phi()$ are the standard normal probability and cumulative distribution function respectively.
+
+Same as in single cluster case, the classification rule learned from
+$\{(X_i, \hGY_i)\}_{i=1}^{n}$ variables will be determined according to
+whether $X > 0$, with the decision boundary be at $0 + \OhP(n^{-1/2})$.
+By symmetry, the CV error for points from $G_1$ is same as the points from $G_2$. Because $P(G=G_1) = P(G=G_2) =1/2$, the CV error can be calculated solely from $G_2$, that is
+
+\begin{align*}
+  \CV(2)
+  &=
+    \frac{1}{m}
+    \sum_{i=n+1}^{n+m}
+      \| (Y_i - \bmuY_1) 1\{\hGX_i = 1\} \|^2
+      +
+      \| (Y_i - \bmuY_2) 1\{\hGX_i = 2\} \|^2, \hspace{0.2in}Y\sim N(-\mu^Y,1)
+\\
+  &=
+    \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]
+      + \OhP(m^{-1/2}) + \OhP(n^{-1/2})\\
+   \end{align*}
+   With 
+   \begin{align}
+ \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]  =& P(\hGX_i = 1)\cdot E[(Y - a)^2] + P(\hGX_i = 2)\cdot E[(Y + a)^2] \nonumber \\  \nonumber
+ =& [1-\Phi(\mu^X)][var(Y)+(-\mu^Y - a)^2]+\\ \nonumber
+  & \Phi(\mu^X)[var(Y)+(-\mu^Y + a)^2]\\  \nonumber
+  =& [1-\Phi(\mu^X)][1+(\mu^Y + a)^2]+\Phi(\mu^X)[1+(\mu^Y -a )^2]   \\  \nonumber
+  =& 1+(\mu^Y + a)^2 -4a\Phi(\mu^X)\mu^Y  \nonumber
+\end{align}
+where $a$ is given by $(1)$.
+
+When $k=1$, the result is straight forward since the estimated centroid will approximately equal to
+$0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error
+will be
+\[
+  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
+         = 1 + (\mu^Y)^2 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
+\]
+So if we have $1 + 2\Phi(\mu^Y)+ \frac{2\varphi(\mu^Y)}{\mu^Y} < 4\Phi(\mu^X)$, we have $\CV(2) < \CV(1)$
 \end{proof}
 
-We confirm the proposition with a simulation.
+We confirm this result with a simulation.  We perform $10$ replicates for each pair of $(\mu^X, \mu^Y)$, where both $\mu^X$ and $\mu^Y$ take value on grid of $[0,3]$ with step $0.1$.  In each
+replicate, we generate $20000$ observations from two multivariate normal distributions with identity covariance, where one has mean $(\muX, \muY)$ and the other one has mean $(-\muX, -\muY)$.
+We perform a single $2 \times 2$ fold of Gabriel cross-validation and report the
+times (out of $10$ replicates) when $k=2$ is selected by the algorithm in stead of $k=1$.
+Figure~\ref{fig:overlap-color_plot} shows the frequency $k=2$ is selected by the algorithm for each pair of $(\mu^X, \mu^Y)$.
+The red spot means high number (close to $10$) is selected by the algorithm, which means algorithm very likely will pick $k=2$ over $k=1$ for the corresponding $(\mu^X, \mu^Y)$.
+While blue spot means algorithm prefer $k=1$ for the corresponding value of $(\mu^X, \mu^Y)$.
+We can see the simulation result perfectly align with the theoretical curve (the black line), which separates the $k=2$ zone from the $k=1$ zone.
+It demonstrates that the Gabriel cross-validation works exactly as it suppose to under such setting.
+The position of red spots shows that when the two clusters are reasonably apart (not overlapping too heavily) in both dimensions, the Gabriel cross-validation is asymptotically consistent.
 
-[Wei, describe simulation here, and include the figure]
+\begin{figure}[H]
+\centering
+\includegraphics[width=25pc]{demo/overlap/color_plot.pdf}
+\caption{Number of times $k=2$ is selected out of $10$ replicates for each pair of $(\muX, \muY)$.
+The heat map shows the frequency $k=2$ is selected by the algorithm, with blue means low number (of $k=2$) is selected and red color means high number is selected.
+The black line is the theoretical curve, above which the algorithm suppose to pick $k=2$ and below which algorithm select $k=1$.
+}
+\label{fig:overlap-color_plot}
+\end{figure}
 
@@
 Bootstrap stability \citep{fang2012selection} in following simulation
-settings. We select $L=R$ and $5-$fold cross-validation in row
-($m=\frac{4}{5}n$) as default parameter setting for our proposed Gabriel
-method. Note that set $L=R$ corresponding to $2-$fold cross-validation in
+settings. We select $p=q$ and $5-$fold cross-validation in row
+($m=\frac{1}{4}n$) as default parameter setting for our proposed Gabriel
+method. Note that set $p=q$ corresponding to $2-$fold cross-validation in
 column.
@@
 
-  \item Two clusters in $4$ dimensions --- $50$ i.i.d standard normal
-    observations are generated, i.e. generated from multivariate normal
-    $\mathcal{N}\left(\boldsymbol\mu,\mathbf{I}\right)$ for each cluster, with
-    cluster center $\boldsymbol\mu = (1,0,0,1)$ and $(1,3.5,3.5,1)$
-    respectively
+  \item Two clusters in $4$ dimensions --- $50$ i.i.d
+    observations are generated from both multivariate normal
+    $\mathcal{N}\left(\mu_1,0.5\Sigma\right)$ and multivariate normal
+    $\mathcal{N}\left(\mu_2,1.5\Sigma\right)$, where 
+    cluster center $\mu_1 = (1,0,0,1)$ and $\mu_2 =(1,3.5,3.5,1)$
+    respectively. $\Sigma$ has $AR(1)$ structure with $\rho=-0.2$ and $\sigma=1$.
 
@@
 
-Besides the number of clusters selected by each algorithm is returned, we also
-give the prediction error (PE) for each algorithm in each setting. The PE is
-defined by 
-\[
-  \PE(k)
-    =
-      \frac{\|\hat{\mathbf{X}}_k- \boldsymbol{\mu} \|_F}
-           {PE_{\text{oracle}}}
-\]
-where
-\[
-  \PE_{\text{oracle}} = \min_k \|\hat{\mathbf{X}}_k- \boldsymbol{\mu} \|_F
-\]
-Matrix $\boldsymbol{\mu}$ consists of true centers of observations in data
-matrix $\mathbf{X}$, while $\hat{\mathbf{X}}_k$ is obtained by replacing each
-observation in $\mathbf{X}$ by its closest cluster center result from applying
-K-means on $\mathbf{X}$ with parameter $k$. Here $k$ is the optimal number of
-clusters selected by each algorithm on data $\mathbf{X}$. This is only
-possible on simulated data because otherwise one would never know the true
-center matrix~$\boldsymbol{\mu}$.
-
-The oracle selected number for each data $\mathbf{X}$ is also given on the
-table, which is defined by 
-\[
-  K_{\text{oracle}}
-    =
-      \argmin_{k} \|\hat{\mathbf{X}}_k - \boldsymbol{\mu} \|_F
-\]
-
+Table \ref{table1} shows the distribution of $k$ selected by each algorithm in 
+each simulation setting. We run $k$ from $1$ to $15$, and the best $k$ within this range
+is selected by the algorithms. [ NA in the table means...]
 \clearpage
@@
 \begin{longtable}{lcccccccccccccccc}
+\captionsetup{justification=centering}
 \caption{\label{table1} Simulation Results}\\
@@
 
-The ``\textendash'' sign in Setting~1 is because for CH-index, Hartigen and
-Bootstrap stability, the minimum $k$ they can pick is $2$. So they are not
-included for comparison since the true $k=1$. Setting $2-4$ have Gaussian
-distribution, where the Gaussian mixture model-based clustering suppose to
-perform the best. It is only the case for setting $2$. In setting $3$ and $4$,
+For CH-index, Hartigen and Bootstrap stability, the minimum $k$ they can pick is $2$. So they are not
+included for comparison since the true $k=1$. It also demonstrates the difficulty 
+that null scenario poses for some existing algorithms. For those could select $k=1$,
+most of them work well under such setting except for Jump method. Setting $2-4$ have Gaussian
+distribution, where the Gaussian mixture model-based method suppose to
+perform the best. However, it only performs reasonably well in setting $2$. In setting $3$ and $4$,
 it has a very poor performance even if the underlying distribution is indeed
-Gaussian mixture model. It highlights the difficulties that high dimension
-data could cause. Other methods (Gap statistics, CH-index, Hartigan, Jump
+Gaussian mixture model. It highlights the difficulty that high dimension
+data could cause. Other methods (CH-index, Hartigan, Jump
 method, Prediction strength and Boostrap stability) also have problems in
 these two high-dimension settings as we could not find any one with reasonable
-performance in terms of finding the correct $k$ in both cases. In terms of PE,
-Boostrap stability and Gap statistics have reasonable performance, but come
-with the price of picking larger $k$. Our proposed Gabriel method (also Wold
-CV method) clearly stands out from the basket of existing methods using either
-measure under the high-dimension setting $3$ and $4$.
+performance in terms of finding the correct $k$ in both cases. The only exception is
+Gap statistics, which works perfectly in both settings. Our proposed Gabriel method (also Wold
+CV method) clearly stands out from the basket of existing methods under the 
+high-dimension setting $3$ and $4$, as we can see they both perform perfectly under
+such settings.
 
@@
 instead of normal distributed, coupled with unequal number of observations or
-unequal variance for different clusters. From Table \ref{table1}, we can see
-that our proposed Gabriel method is robust and has superior performance
+unequal variance for different clusters. From Table \ref{table1}, we can see none of 
+the existing methods has good performance in both settings. In contrast,
+our proposed Gabriel method is robust and has superior performance
 compare with other methods. Because one can hardly tell the underlying
@@
 The first one is congress voting data which consists of voting records of
-$98$th United States Congress, $2$nd session. This data set includes votes for
-each of the U.S. House of Representatives Congressmen on the $16$ key votes
+$98$th United States Congress, $2$nd session \citep{schlimmer1987concept}. 
+This data set includes votes for each of the U.S. House of Representatives Congressmen on the $16$ key votes
 identified by the \textit{CQA} (Congressional Quarterly Almanac). For each
@@
 After excluding the records with missing data, this data set consists records of $683$ patients, each with measurements of nine attributes of their biopsy specimens.
-It is known that there exist at least two groups of patients: $444$ patients with benign specimens and $239$ patients with malignant specimens.\\
+It is known that there exist at least two groups of patients: $444$ patients with benign specimens and $239$ patients with malignant specimens.
 
@@
 \begin{center}
+\captionsetup{justification=centering}
 \caption{\label{table2} Number of clusters selected by each algorithm}
@@
 (Democratic and Republican) represent two clusters in this data set. So the
-optimal number should be two. Close inspection shows \textit{kmeans} with
+optimal number should be two. Close inspection shows $k$-means with
 $k=2$ separates the two parties very well with the lowest miss-classification
@@
 lower bound (they would pick $k=1$ if allowed). For the breast cancer data,
-it's know to have at least $2$ cluster based on whether it's benign specimens
+it's known to have at least $2$ cluster based on whether it's benign specimens
 or not. But it doesn't mean the optimal $k$ should be $2$.

### 2015-08-22 demo/bench/settings.R

@@ -41,7 +41,7 @@ location <- function(M.data,M.center)
 
 Data2center <- function(number = 50)
 {
-	u1 = c(-1,0.0,0.0,-1)##changed
+	u1 = c(1,0.0,0.0,1)
 	u2 = c(1,3.5,3.5,1)
       Centers <- matrix(c(1,1,0,3.5,0,3.5,1,1),2,4)
       COV <- matrix(c(1,-0.2,(-0.2)^2,(-0.2)^3,-0.2,1,-0.2,(-0.2)^2,(-0.2)^2,-0.2,1,-0.2,(-0.2)^3,(-0.2)^2,-0.2,1),4,4)

### 2015-08-26 demo/bench/05_tabulate_results.R
Add setting 7

@@ -6,7 +6,8 @@ setting.names <- list(
     "setting3" = "Setting 3",
     "setting4" = "Setting 4",
     "setting5" = "Setting 5",
-    "setting6" = "Setting 6")
+    "setting6" = "Setting 6",
+    "setting7" = "Setting 7")
 
 nclust <- list(
     "setting1" = 1L,
@@ -14,7 +15,8 @@ nclust <- list(
     "setting3" = 4L,
     "setting4" = 10L,
     "setting5" = 4L,
-    "setting6" = 3L)
+    "setting6" = 3L,
+    "setting7" = 1L)
 
 method.names <- list(
     "oracle" = "Oracle",

### 2015-08-26 demo/bench/settings.R

@@ -295,12 +295,27 @@ setting6 <- function()
 }
 
 
+setting7 <- function()
+{
+    p <- 10
+    n <- 200
+    mean <- numeric(p)
+    cov <- matrix(.9, p, p)
+    diag(cov) <- 1
+    mu <- matrix(0, n, p)
+    eps <- mvrnorm(n, mean, cov)
+
+    list(x = mu + eps, mean = mu)
+}
+
+
 # seeds taken from random.org, uniform { 1, ..., 10000000 }
 settings <- list("setting1" = list(simulate = setting1, seed = 4244250),
                  "setting2" = list(simulate = setting2, seed = 5513442),
                  "setting3" = list(simulate = setting3, seed = 5685887),
                  "setting4" = list(simulate = setting4, seed = 1243061),
                  "setting5" = list(simulate = setting5, seed = 7997590),
-                 "setting6" = list(simulate = setting6, seed = 5427086))
+                 "setting6" = list(simulate = setting6, seed = 5427086),
+                 "setting7" = list(simulate = setting7, seed = 1595473))
 
 

### 2015-08-27 cvclust.tex
Add setting 7
add single cluster, multiple dimension case

@@
 \DeclareMathOperator*{\cov}{cov}
+\DeclareMathOperator*{\corr}{corr}
 \DeclareMathOperator*{\tr}{tr}
 \DeclareMathOperator*{\var}{var}
+\newcommand{\T}{T}
 
@@
 
-\subsection{Single Cluster}
+\subsection{Single cluster in two dimensions}
 
@@
 
-\begin{proposition}
+\begin{proposition}\label{prop:1clust-2dim}
 Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
@@
 
-\subsection{Two Clusters}
+\subsection{Single cluster in higher dimensions}
+
+To generalize the result of the previous section to higher dimensions, 
+suppose that $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ is data from a single fold of
+Gabriel Cross-Validation, with each $(X, Y)$ pair in $\R^p \times \R^q$ a
+draw a mean-zero multivariate normal distribution. To analyze this setting, it
+suffices to consider the mean-zero case with $\cov(X) = I_p$.
+
+Suppose that $\cov(Y) = U \Lambda U^\T$ is the eigendecomposition of
+$\cov(Y)$. For simplicity, suppose that the leading eigenvalue $\lambda_1$
+has unit multiplicity, so that the first principal subspace of $Y$ is a line,
+given by the span of the vector $u_1 \in \R^q$. When we apply $k$-means to
+$Y_1, \dotsc, Y_n$, the centroids will be close to
+$\bmuY_1 =  \sqrt{2\lambda_1 /\pi} u_1$ and
+$\bmuY_2 = -\sqrt{2\lambda_1 /\pi} u_1$, with errors of size $\OhP(n^{-1/2})$.
+
+The means for the $X$ variables will be close to
+$\bmuX_1 = \E(X \mid u_1^\T Y > 0)$ and
+$\bmuX_2 = \E(X \mid u_1^\T Y < 0)$,
+with errors of size $\OhP(n^{-1/2})$.  Let
+\[
+  \begin{bmatrix}
+    I_p & \lambda_1^{1/2} \rho \\
+    \lambda_1^{1/2} \rho^T & \lambda_1
+  \end{bmatrix}
+\]
+be the covariance matrix of $(X, u_1^\T Y)$,
+where $\rho = (\rho_1, \dotsc, \rho_p)$ is such that $\rho_i = \corr(X_i,
+u_1^\T Y)$ for $i = 1, \dotsc, p$.
+. Then,
+\[
+  \bmuX_1 = \sqrt{2 \lambda_1 / \pi} \rho,
+  \qquad
+  \bmuX_2 = -\sqrt{2 \lambda_1 / \pi} \rho;
+\]
+thus, the classification rule learned from $\{ (X_i, \hGY_i) \}_{i=1}^{n}$
+will send observations to class $1$ or $2$ according to whether
+$\rho^\T X + \OhP(n^{-1/2})$ is positive or negative.
+
+We have reduced the analysis of cross-validation error to the two-dimensional
+case. In particular, following the calculations from the proof of
+Prop.~\ref{prop:1clust-2dim}, if $|\corr(\rho^\T X, u_1^\T Y)| < 0.5$, then with
+probability tending to one as $m$ and $n$ increase, $\CV(1) < \CV(2)$.
+The condition $|\corr(\rho^\T X, u_1^\T)| < 0.5$ reduces to
+\[
+  \sum_{i=1}^p \rho_i^2 < 0.25.
+\]
+A sufficient (but not necessary) condition to ensure that this holds is that
+the magnitude of the canonical correlation between the $X$ variables and
+the $Y$ variables is below $0.5 / \sqrt{p}$.
+
+\subsection{Two clusters}
 
@@
     1/5$ respectively.
+
+  \item[7.] One cluster in $10$ dimensions with correlated noise -- $200$
+    mean-zero multivariate normal observations; the covariance matrix $\Sigma$
+    has $\Sigma_{ij} = 0.9$ for $i \neq j$ and $\Sigma_{ii} = 1$ for $i = 1,
+    \dotsc, 10$.
 \end{enumerate}
@@
 \input{demo/bench/setting6/results.tex}
+\midrule
+\input{demo/bench/setting7/results.tex}
 \end{longtable}

### 2015-08-27 cvclust.tex
minor clarification

@@
 A sufficient (but not necessary) condition to ensure that this holds is that
-the magnitude of the canonical correlation between the $X$ variables and
+the magnitude of the first canonical correlation between the $X$ variables and
 the $Y$ variables is below $0.5 / \sqrt{p}$.

### 2015-08-27 cvclust.tex
fix some typos

@@
 cross-validation criterion chooses the correct answer $k = 1$ whenever
-$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
+$|\rho| < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
 \label{fig:nullcorr-equal}
@@
 given by the span of the vector $u_1 \in \R^q$. When we apply $k$-means to
-$Y_1, \dotsc, Y_n$, the centroids will be close to
+$Y_1, \dotsc, Y_n$ with $k = 2$, the centroids will be close to
 $\bmuY_1 =  \sqrt{2\lambda_1 /\pi} u_1$ and

### 2015-09-27 technometrics.tex
add technometrics journal version

--- 2015-08-27 cvclust.tex
@@
-\documentclass[11pt]{article}
-\usepackage{epsf}
+\documentclass[12pt]{article}
+\usepackage{amsmath}
+\usepackage{graphicx,psfrag,epsf}
+\usepackage{enumerate}
+\usepackage{natbib}
+\usepackage{url} % not crucial - just used below for the URL 
+%---------We added packages---------------------
+%\usepackage{epsf}
 \usepackage{hyperref}
-\usepackage{graphicx}
+%\usepackage{graphicx}
 \usepackage{subfigure}
-\usepackage[margin=2.5cm]{geometry}
-\usepackage{amsmath}
+%\usepackage[margin=2.5cm]{geometry}
+%\usepackage{amsmath}
 \usepackage{footnote}
@@
 \usepackage[singlelinecheck=false]{caption}
-\usepackage{caption}
+%\usepackage{caption}
 \usepackage{amsthm}
 \usepackage{enumitem}
-\usepackage{graphics}
+%\usepackage{graphics}
 \usepackage{float}
 \usepackage{amssymb}
-\usepackage{natbib}
+%\usepackage{natbib}
 \usepackage{lscape}
 \usepackage{array}
-\textheight=8.7in
+%\textheight=8.7in
 \usepackage{setspace}
@@
 
-\def\threedigits#1{%
-  \ifnum#1<100 0\fi
-  \ifnum#1<10 0\fi
-  \number#1}
-
-\topmargin=0.1in \oddsidemargin=-0.1cm \evensidemargin=-0.1cm
-
-\paperheight=11in \paperwidth=8.5in \marginparwidth=0in
-
-\marginparsep=0in \textwidth=6.5in \headheight=0in \headsep=0in
+%\def\threedigits#1{%
+%  \ifnum#1<100 0\fi
+%  \ifnum#1<10 0\fi
+%  \number#1}
+
+%\topmargin=0.1in \oddsidemargin=-0.1cm \evensidemargin=-0.1cm
+
+%\paperheight=11in \paperwidth=8.5in \marginparwidth=0in
+
+%\marginparsep=0in \textwidth=6.5in \headheight=0in \headsep=0in
 
@@
 \DeclareMathOperator*{\cov}{cov}
-\DeclareMathOperator*{\corr}{corr}
 \DeclareMathOperator*{\tr}{tr}
 \DeclareMathOperator*{\var}{var}
-\newcommand{\T}{T}
 
@@
 \newcommand{\hGY}{\hat G^{Y}}
+%----------------------We added packages----------------------------
+\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{0}
+
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.3in}%
+\addtolength{\topmargin}{-.8in}%
 
 \begin{document}
-\title{Estimating the number of clusters using Cross Validation}
-\author{Wei Fu \qquad Patrick O. Perry \\\\ New York University}
-\date{}
+
+%\bibliographystyle{natbib}
+
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\if0\blind
+{
+  \title{\bf Estimating the number of clusters using Cross Validation}
+  \author{Wei Fu \thanks{
+    The authors gratefully acknowledge}\hspace{.2cm}\\
+    Department of IOMS, New York University\\
+    and \\
+    Patrick O. Perry \\
+    Department of IOMS, New York University}
   \maketitle
+} \fi
+
+\if1\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf Estimating the number of clusters using Cross Validation}
+\end{center}
+  \medskip
+} \fi
+
+\bigskip
 \begin{abstract}
@@
 
+\noindent%
+{\it Keywords:}  clustering, unsupervised learning, data-driven method 
+\vfill
+\hfill {\tiny technometrics tex template (do not remove)}
+
+\newpage
+\spacingset{1.45} % DON'T change the spacing!
 \section{Introduction}
+\label{sec:intro}
 
@@
 \section{Cross-validation for selecting the number of clusters}
-
+\label{sec:meth}
 Cross-validation is commonly used for model selection in supervised learning
@@
 \label{sec:gabriel-cv-algorithm}
-
 We are given a data matrix with $N$ rows and $P$ columns.  In each fold of
@@
     be the assigned cluster for row $i$.
-
   \item \label{step:gabriel-classify}
@@
     $\hGX : \R^p \to \{ 1, \dotsc, k \}$.
-
   \item \label{step:gabriel-predict}
@@
     $\bmuY(g) = \bmuY_g$.
-
   \item \label{step:gabriel-evaluate}
@@
 of a tie).
-
+%------------------------------------------------------------------
 \section{Self-Consistency of Gabriel CV method}
@@
     $i = 1, \dotsc, n+m$.
-
   \item \label{asn:self-consistency-distinct-mux}
     The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
-
   \item \label{asn:self-consistency-distinct-muy}
     The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
-
   \item \label{asn:self-consistency-train}
@@
     $1, \dotsc, n$ such that $G_i = g$.
-
   \item \label{asn:self-consistency-test}
@@
     \dotsc, n+m$ such that $G_i = g$.
-
 \end{enumerate}
@@
 \end{proof}
-
+%-------------------------------------------------------------
 \section{Analysis of Gabriel Cross-Validation with Gaussian Noise}
 
-\subsection{Single cluster in two dimensions}
-
+\subsection{Single Cluster}
 Now we we analyze the asymptotic performance of Gabriel Cross-Validation, in
@@
 correlated or independent, then the method will correctly prefer $k = 1$ to
-$k = 2$ clusters. We state the result in the case where $\dataX$ has 2 columns,
-but it is straightforward to generalize this result to higher dimensions.
-
-\begin{proposition}\label{prop:1clust-2dim}
+$k = 2$ clusters. We first state the result in the case where $\dataX$ has two 
+columns, and later generalize this result to higher dimensions.
+
+\begin{proposition}
 Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
@@
 \begin{proof}
-
 Given $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to $\{ Y_i
@@
 \centering
-\includegraphics[width=25pc]{demo/nullcorr/equal.pdf}
+\includegraphics[width=3in]{demo/nullcorr/equal.pdf}
 \caption{Cross-validation error on $10$ replicates, with the number of
@@
 
-\subsection{Single cluster in higher dimensions}
-
-To generalize the result of the previous section to higher dimensions, 
-suppose that $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ is data from a single fold of
-Gabriel Cross-Validation, with each $(X, Y)$ pair in $\R^p \times \R^q$ a
-draw a mean-zero multivariate normal distribution. To analyze this setting, it
-suffices to consider the mean-zero case with $\cov(X) = I_p$.
-
-Suppose that $\cov(Y) = U \Lambda U^\T$ is the eigendecomposition of
-$\cov(Y)$. For simplicity, suppose that the leading eigenvalue $\lambda_1$
-has unit multiplicity, so that the first principal subspace of $Y$ is a line,
-given by the span of the vector $u_1 \in \R^q$. When we apply $k$-means to
-$Y_1, \dotsc, Y_n$, the centroids will be close to
-$\bmuY_1 =  \sqrt{2\lambda_1 /\pi} u_1$ and
-$\bmuY_2 = -\sqrt{2\lambda_1 /\pi} u_1$, with errors of size $\OhP(n^{-1/2})$.
-
-The means for the $X$ variables will be close to
-$\bmuX_1 = \E(X \mid u_1^\T Y > 0)$ and
-$\bmuX_2 = \E(X \mid u_1^\T Y < 0)$,
-with errors of size $\OhP(n^{-1/2})$.  Let
+\begin{proposition}
+Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
+of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^{p+q}$ is an
+independent draw from a mean-zero multivariate normal distribution with 
+covariance matrix $\Sigma_{XY}= \left( \begin{smallmatrix} \Sigma_{XX} & \Sigma_{XY} \\ 
+ \Sigma_{YX} & \Sigma_{YY} \end{smallmatrix}\right)$, with $\Sigma_{YY}$ has leading 
+eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. In this case, the data are drawn
+from a single cluster; the true number of clusters is~$1$.  If  $\frac{\sqrt{\lambda_1}}{2}
+ > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}}$,
+then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
+\end{proposition}
+
+\begin{proof}
+Let $X$ and $Y$ be jointly multivariate normal distributed with mean $\mathbf{0}$ and covariance matrix $\Sigma_{XY}$, i.e.
+\[	(X,Y) \sim \mathcal{N} \left( \mathbf{0}, \Sigma_{XY}\right)	\]
+where $\Sigma_{XY}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} \\  \Sigma_{YX} & \Sigma_{YY} \end{bmatrix}$.
+
+Let $\Sigma_{YY} = U \Lambda U^T$  be the eigendecomposition of $\Sigma_{YY}$, with leading eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$.
+Then the centroid of $k$-means applying on $(y_1,..,y_n)$ is on the first PC
+of $Y$,\[	E(u^T_1 Y|u^T_1 Y>0) = \bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1\] and 
+\[	E(u^T_1 Y|u^T_1 Y<0) = \bar{\mu}^Y_2 =-\sqrt{2 \lambda_1/\pi}u_1\]
+ where $u^T_1 Y \sim \mathcal{N}(0,\lambda_1)$.
+
+To compute $\bar{\mu}^X_1 = E(X|u^T_1 Y>0)$, we need to know the conditional distribution $X|u^T_1 Y$.
+Since $(X,Y)$ has multivariate normal distribution, $(X,u^T_1 Y)$ also has a multivariate normal distribution with mean $\mathbf{0}$ and covariance matrix
+$$\Sigma_{X,u^T_1 Y}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} u_1 \\  u^T_1 \Sigma_{YX} & \lambda_1 \end{bmatrix}$$
+The conditional distribution $X|u^T_1 Y$ is hence normal with mean
+ $$\mu_{X|u^T_1 Y} = \Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y $$
+Therefore, 
+\begin{align}
+\bar{\mu}^X_1 &= E(X \mid u^T_1 Y>0) \nonumber \\ \nonumber
+ 			  &= E\left(E[X \mid u^T_1 Y] \mid u^T_1 Y>0\right) \\ \nonumber
+ 			  &=  E\left(\Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y \mid u^T_1 Y>0\right)\\ \nonumber
+ 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 E(u^T_1 Y \mid u^T_1 Y>0) \\ \nonumber
+ 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 \sqrt{2 \lambda_1/\pi} \\ \nonumber
+ 			  &= \sqrt{2 / \lambda_1 \pi} \Sigma_{XY}u_1
+\end{align}
+Similar calculation yields $\bar{\mu}^X_2 = -\sqrt{2 / \lambda_1 \pi} \Sigma_{XY}u_1$.
+The decision rule to classify any observed value of $X$ to $\bar{\mu}^X_1$ is therefore
+\[	(\bar{\mu}^X_1)^T X >0	\hspace{0.2in}\text{or} \hspace{0.2in} u^T_1\Sigma_{YX}X>0\] 
+Since $u^T_1\Sigma_{YX}X$ is a linear combination of $X$, it also has normal distribution 
+\[	\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)	\]
+And $(Y,u^T_1\Sigma_{YX}X)$ also have multivariate normal distribution with mean $\mathbf{0}$ 
+and covariance matrix
 \[
 \begin{bmatrix}
-    I_p & \lambda_1^{1/2} \rho \\
-    \lambda_1^{1/2} \rho^T & \lambda_1
+\Sigma_{YY} & \Sigma_{YX}\Sigma_{XY}u_1  \\
+u^T_1\Sigma_{YX}\Sigma_{XY} &  u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1
 \end{bmatrix}
 \]
-be the covariance matrix of $(X, u_1^\T Y)$,
-where $\rho = (\rho_1, \dotsc, \rho_p)$ is such that $\rho_i = \corr(X_i,
-u_1^\T Y)$ for $i = 1, \dotsc, p$.
-. Then,
-\[
-  \bmuX_1 = \sqrt{2 \lambda_1 / \pi} \rho,
-  \qquad
-  \bmuX_2 = -\sqrt{2 \lambda_1 / \pi} \rho;
-\]
-thus, the classification rule learned from $\{ (X_i, \hGY_i) \}_{i=1}^{n}$
-will send observations to class $1$ or $2$ according to whether
-$\rho^\T X + \OhP(n^{-1/2})$ is positive or negative.
-
-We have reduced the analysis of cross-validation error to the two-dimensional
-case. In particular, following the calculations from the proof of
-Prop.~\ref{prop:1clust-2dim}, if $|\corr(\rho^\T X, u_1^\T Y)| < 0.5$, then with
-probability tending to one as $m$ and $n$ increase, $\CV(1) < \CV(2)$.
-The condition $|\corr(\rho^\T X, u_1^\T)| < 0.5$ reduces to
-\[
-  \sum_{i=1}^p \rho_i^2 < 0.25.
-\]
-A sufficient (but not necessary) condition to ensure that this holds is that
-the magnitude of the canonical correlation between the $X$ variables and
-the $Y$ variables is below $0.5 / \sqrt{p}$.
-
-\subsection{Two clusters}
+The conditional distribution of $Y|u^T_1\Sigma_{YX}X$ is also multivariate normal with mean 
+\[	
+\mu_{Y|u^T_1\Sigma_{YX}X } = \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1}u^T_1\Sigma_{YX}X	
+\]
+The $Y$ center for $u^T_1\Sigma_{YX}X>0$ is
+\begin{align}
+\hat{\mu}^Y_1 &= E(Y|u^T_1\Sigma_{YX}X>0) \nonumber \\ \nonumber
+ 		      & =  \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) \\ \nonumber
+\end{align}
+Note that $u^T_1\Sigma_{YX}X$ has normal distribution $\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)$, so
+\[
+E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) = \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}
+\]
+Therefore, we have the $Y$ center for $u^T_1\Sigma_{YX}X>0$ be
+\begin{align*}
+\hat{\mu}^Y_1 &= \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1} \hspace{0.1in} \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} \\
+&=\frac{\sqrt{2/\pi}}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} \Sigma_{YX}\Sigma_{XY}u_1
+\end{align*}
+
+Recall that $\bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1$, to judge if $CV(2) > CV(1)$, one only need to compare the distance between  $\hat{\mu}^Y_1$  and  $\bar{\mu}^Y_1$ with distance between  $\hat{\mu}^Y_1$ and grand mean $0$.
+By variance and bias decomposition of prediction MSE, when variance is the same, only bias influence the MSE.
+
+After some linear algebra manipulation, we get
+$||\hat{\mu}^Y_1 - \bar{\mu}^Y_1||^2 > ||\hat{\mu}^Y_1||^2$ or $CV(2) > CV(1)$ iff
+\[
+ \frac{\sqrt{\lambda_1}}{2} > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} 
+\]
+\end{proof}
+
+\subsection{Two Clusters}
 
@@
 \centering
-\includegraphics[width=25pc]{demo/overlap/color_plot.pdf}
+\includegraphics[width=3in]{demo/overlap/color_plot.pdf}
 \caption{Number of times $k=2$ is selected out of $10$ replicates for each pair of $(\muX, \muY)$.
@@
 \end{figure}
-
+%----------------------------------------------------------------
 \section{Simulation}
-
 In this section, simulation is performed to evaluate the performance of our
@@
 column.
+
 \begin{enumerate}
-
   \item Single cluster in $10$ dimensions --- $200$ observations, each
     observation uniformly distributed over $[0,1]$ in each dimension. 
-
   \item Two clusters in $4$ dimensions --- $50$ i.i.d
@@
     respectively. $\Sigma$ has $AR(1)$ structure with $\rho=-0.2$ and $\sigma=1$.
-
   \item Four clusters in $100$ dimensions ---  Each cluster has $100$ or $150$
@@
     $\mathcal{N}\left(\mathbf{0},0.65^2\mathbf{I}\right)$
-
   \item Ten clusters in $100$ dimensions ---  Each cluster has $50$ or $100$
@@
     $\mathcal{N}\left(\mathbf{0},0.72^2\mathbf{I}\right)$
-
   \item Four log-normal clusters in $16$ dimensions --- For each cluster, $30$
@@
     $\mathcal{N}\left(\mathbf{0},1.2^2\mathbf{I}\right)$
-
   \item[6.] Three exponential clusters in $20$ dimensions with different
@@
     1/5$ respectively.
-
-  \item[7.] One cluster in $10$ dimensions with correlated noise -- $200$
-    mean-zero multivariate normal observations; the covariance matrix $\Sigma$
-    has $\Sigma_{ij} = 0.9$ for $i \neq j$ and $\Sigma_{ii} = 1$ for $i = 1,
-    \dotsc, 10$.
 \end{enumerate}
-Note that in setting 2--6, all clusters are well-separated, i.e. no
+Note that in setting $2$--$6$, all clusters are well-separated, i.e. no
 overlapping. In fact, any simulated clusters with minimum distance less than
 1 unit was discarded, so there is clear definition of true number of
-clusters. The parameters in setting 2--6 are chosen such that about half of
+clusters. The parameters in setting $2$--$6$ are chosen such that about half of
 the random realization were discarded. The idea is borrowed from
@@
 \input{demo/bench/setting6/results.tex}
-\midrule
-\input{demo/bench/setting7/results.tex}
 \end{longtable}
@@
 the Gabriel method a clear edge. 
-
+%----------------------------------------------------------------
 \section{Real data application}
-
 We also applied our proposed method to three real world data sets obtained
@@
 \section{Conclusion}
-
+\label{sec:conc}
 In this paper, we proposed a novel approach to estimate the number of
@@
 high dimension or is heavy-tailed.
+\clearpage
 
@@
 \end{align*}
+\clearpage
 

### 2015-10-08 technometrics.tex
edit technometrics version

@@
 various fields such as psychology, biology, statistics and machine learning
-including pattern recognition, image segmentation etc.
+including pattern recognition, image segmentation etc \citep{jain1999data}.
 
@@
 
-Most existing methods are either model based method requires strong parameter
+Most existing methods are either model based method requires strong modeling
 assumptions or otherwise lack of clear interpretation and theoretical justification.
@@
 true class (label) in data set makes the adoption of cross-validation into
-unsupervised leaning problem difficult. One exception is
-\cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction
+unsupervised leaning problem difficult.
+
+One exception is \cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction
 strength. The strategy is to first cluster the test data and training data
@@
 procedure, which is quite different from the one commonly used in
-cross-validation procedure in supervised learning. Therefore the well
-understood properties of cross-validation procedure in supervised learning
-cannot be carried over to the prediction strength method, makes the
-interpretation of its result difficult. \citet{wang2010consistent} also uses 
+cross-validation procedure in supervised learning. \citet{wang2010consistent} also uses 
 cross-validation to select the optimal $k$. Instead of selecting $k$ which minimizes 
-prediction error, such method picks $k$ which minimizes the specifically defined clustering instability.
-
-Our proposed method is a complete data-driven approach which doesn't rely on
-any parametric assumptions. Through novel form of partitioning data set, we
-are able to employing the cross-validation procedure in clustering exactly the
-same way as in supervised learning problem. Hence, it's easy for reader to
+prediction error, such method picks $k$ which minimizes the specifically defined 
+clustering instability. Note that these methods minimize measures that are specifically defined,
+which makes these methods hard to interpret or compare to other methods through analogy because
+the underlying measures are unique and not well understood.
+
+The prediction error measure in our proposed method is exactly the same as in supervised learning.
+Our method is a complete data-driven approach which doesn't rely on
+any model assumption. Through novel form of partitioning data set, we effectively 
+transferred an unsupervised learning problem into a supervised leaning problem, which is one of
+the kind. By doing so, we are able to employing the cross-validation procedure in clustering 
+a similar way as in supervised learning problem, so that much of the intuition from supervised learning
+carries over. Hence, it's easy for reader to
 understand the intuition behind our proposed method. Simulation and real data
@@
 
+Proposition $1$ gives a very simple condition for the Gabriel method to correctly 
+pick $k=1$ with single cluster in $2$ dimensions. The following proposition generalizes
+such condition for data in arbitrary dimension.
+
 \begin{proposition}
@@
 \end{proof}
+
+Although the condition given by Proposition $2$ is succinct in such that it's straight forward to see how
+the structure of covariance matrix $\Sigma_{XY}$ affects the performance of Gabriel CV method, it's not simple and easy to interpret such condition as the one in Proposition $1$.
+In specific, each block of matrix $\Sigma_{XY}$ influences the condition while the combined effect
+determines whether Gabriel CV method chooses $k=1$ or $k=2$.
+
+In practice, one can check such condition using estimated sample covariance matrix for single cluster case. 
+If such condition doesn't met, one solution is rotating the matrix to eliminate the correlation between each dimension then perform Gabriel CV method on the rotated data matrix.
+Another empirically observed solution is
+to leave most columns for clustering, i.e. increase the dimension of $\Sigma_{YY}$. Such solution is especially 
+useful for data containing multiple clusters, where neither estimating sample covariance matrices nor
+rotating to eliminate correlation is feasible. Although the theory for such solution is available for special cases,
+its general version hasn't been developed.
 
@@
 Figure~\ref{fig:overlap-color_plot} shows the frequency $k=2$ is selected by the algorithm for each pair of $(\mu^X, \mu^Y)$.
-The red spot means high number (close to $10$) is selected by the algorithm, which means algorithm very likely will pick $k=2$ over $k=1$ for the corresponding $(\mu^X, \mu^Y)$.
-While blue spot means algorithm prefer $k=1$ for the corresponding value of $(\mu^X, \mu^Y)$.
+The dark spot means high number (close to $10$) is selected by the algorithm, which means algorithm very likely will pick $k=2$ over $k=1$ for the corresponding $(\mu^X, \mu^Y)$.
+While light spot means algorithm prefer $k=1$ for the corresponding value of $(\mu^X, \mu^Y)$.
 We can see the simulation result perfectly align with the theoretical curve (the black line), which separates the $k=2$ zone from the $k=1$ zone.
 It demonstrates that the Gabriel cross-validation works exactly as it suppose to under such setting.
-The position of red spots shows that when the two clusters are reasonably apart (not overlapping too heavily) in both dimensions, the Gabriel cross-validation is asymptotically consistent.
+The position of dark spots shows that when the two clusters are reasonably apart (not overlapping too heavily) in both dimensions, the Gabriel cross-validation is asymptotically consistent.
 
@@
 \caption{Number of times $k=2$ is selected out of $10$ replicates for each pair of $(\muX, \muY)$.
-The heat map shows the frequency $k=2$ is selected by the algorithm, with blue means low number (of $k=2$) is selected and red color means high number is selected.
+The heat map shows the frequency $k=2$ is selected by the algorithm, with light means low number (of $k=2$) is selected and dark color means high number is selected.
 The black line is the theoretical curve, above which the algorithm suppose to pick $k=2$ and below which algorithm select $k=1$.
@@
 each simulation setting. We run $k$ from $1$ to $15$, and the best $k$ within this range
-is selected by the algorithms. [ NA in the table means...]
+is selected by the algorithms. NA in the table means algorithm did not converge or otherwise 
+failed. 
 \clearpage
@@
 
-\section{Conclusion}
+\section{Discussion}
 \label{sec:conc}
@@
 unsupervised learning problem into supervised learning problem via novel form
-of cross validation. We proved the self-consistency for our proposed Gabriel
-CV method as well as its asymptotic property with noise, and showed the
-robustness of our method by simulation. Our method has very good performance
-in our limited simulation settings, and clearly the superior one when data is
-high dimension or is heavy-tailed.
+of cross validation. Such approach is quite different from previous methods which 
+utilize the within/between cluster dispersion or stability criterion for selecting the 
+optimal $k$. Our method utilizes the connection between different dimensions (columns)
+of data through the uniqueness of each cluster center. We proved the self-consistency 
+for our proposed Gabriel CV method as well as its asymptotic property with Gaussian noise,
+and showed the robustness of our method by simulation. Our method has very good performance
+in our limited simulation settings and real data application, and clearly the superior
+one when data is high dimension or is heavy-tailed.
+
+Besides no modeling assumption is required, our proposed method is robust against data set 
+with variance heterogeneity, unequal number of observations, non-Gaussian noise and high-dimension. 
+Such robustness is important in practice because for any given data, it's hard to tell whether
+its clusters have different number of observations, is the variance equal for each cluster,
+or what underlying noise distribution it has. The weakness of our proposed method is 
+that its theory assumes only week correlation between ``predictor'' columns and ``response'' 
+columns. In practice, many data sets don't exhibit high correlations between columns, where
+our proposed method can be safely applied. In case the high correlation does exist, some procedure
+such as rotating the data set and/or leave out most columns for clustering may be used to
+reduce its effect as previously mentioned. However, the theory for those procedure have not
+been fully developed and it could be the future research topic.
+
+Another situation where our proposed method cannot be directly used is when the clusters 
+are non-convex. In such situation, $k$-means itself doesn't work well, for example two 
+concentric circles share the same cluster center \citep{hastie2009elements}. However, it's 
+possible that our proposed Gabriel CV method can be used on the transformed data set. In the
+concentric circles case where spectral clustering is appropriate, our proposed method can be applied 
+on the eigenvector subspace of the graph Laplacian matrix inside the spectral clustering algorithm
+to find the optimal $k$. This can also be the future research topic.
+
 \clearpage

### 2015-10-09 technometrics.tex
Revised the rotation method for correcting high correlation

@@
 
-In practice, one can check such condition using estimated sample covariance matrix for single cluster case. 
-If such condition doesn't met, one solution is rotating the matrix to eliminate the correlation between each dimension then perform Gabriel CV method on the rotated data matrix.
-Another empirically observed solution is
-to leave most columns for clustering, i.e. increase the dimension of $\Sigma_{YY}$. Such solution is especially 
-useful for data containing multiple clusters, where neither estimating sample covariance matrices nor
-rotating to eliminate correlation is feasible. Although the theory for such solution is available for special cases,
+In practice, check such condition is tricky.
+One need to know the covariance matrix of each cluster to check if the condition holds.
+However, in order to estimate the covariance matrix we need to know the cluster center of each cluster, which in turn need to know the number of clusters in the data.
+A possible solution is following
+\begin{enumerate}
+	\item For parameter $k$ starts at $k=1$ do the following
+	\begin{enumerate}[label=(\roman*)]
+		\item Estimate $k$ cluster centers by applying $k$-means with parameter $k$
+		\item Estimate the covariance matrix $\hat{\Sigma}$ for each cluster 
+		\item Check the condition for each $\hat{\Sigma}$
+		\item If the the condition doesn't hold, rotate that cluster through 
+		rotation matrix $R^T$, where $\hat{\Sigma} = RLR^T$ 
+		\item Apply previously proposed Gabriel CV method on the rotated data, get estimated number $k^*$
+	\end{enumerate}
+	\item If $k^* = k$, stop the algorithm and return $k$; otherwise repeat step $1$ for parameter $k+1$
+\end{enumerate}
+from which we can see the condition checking step can not be separated from the proposed Gabriel CV method.
+Another empirically observed solution is to leave out most columns for clustering, i.e. increase the 
+dimension of $\Sigma_{YY}$. Such method is observed reducing the impact of correlation on the performance 
+of the proposed method. Although the theory for such solution is available for special cases,
 its general version hasn't been developed.

### 2015-11-13 demo/bench/05_tabulate_results.R
add rotated gabriel sim

@@ -29,6 +29,7 @@ method.names <- list(
     "Stab" = "Stability",
 #    "gabriel-nearest-2x2" = "Gabriel CV ($2 \\times 2$)",
     "gabriel-nearest-5x2" = "Gabriel CV",
+    "gabriel-rot-5x2" = "Gabriel CV (Rot)",
 #    "gabriel-lda-equal-2x2" = "Gabriel CV (LDA Equal; $2 \\times 2$)",
 #    "gabriel-lda-equal-5x2" = "Gabriel CV (LDA Equal; $5 \\times 2$)",
 #    "gabriel-lda-proportion-2x2" = "Gabriel (CV LDA Prop.; $2 \\times 2$)",

### 2015-11-13 demo/bench/methods.R

@@ -24,6 +24,23 @@ methods <- list(
                                 classify.method="nearest")
         cv$centers
     },
+    "gabriel-rot-5x2" = function(x, maxcenters) {
+        # generate a random rotation
+        dim <- ncol(x)
+        z <- matrix(rnorm(dim * dim), dim, dim)
+        qr <- qr(z)
+        q <- qr.Q(qr)
+        sign <- sample(c(-1, 1), dim, replace=TRUE)
+        rot <- q %*% diag(sign, dim)
+
+        # rotate the columns of the data matrix
+        x_rot <- x %*% rot
+
+        # apply Gabriel CV to rotated data
+        cv <- cv.kmeans.gabriel(x_rot, 2, 2, maxcenters,
+                                classify.method="nearest")
+        cv$centers
+    },
     "gabriel-nearest-5x2" = function(x, maxcenters) {
         cv <- cv.kmeans.gabriel(x, 5, 2, maxcenters,
                                 classify.method="nearest")

### 2015-12-05 New simulation/technometrics.tex
Add new directory contains new simulation

--- 2015-10-08 technometrics.tex
@@
 
+Among all the proposed methods, very few provide theoretical consistency result for choosing $k$. 
+Assume data is a mixture of $G$ $p$-dimension clusters with equal prior, identically distributed with 
+common covariance matrix and finite fourth moments in each dimension, \citet{sugar2003finding} shows
+the Jump method will pick the correct $k$ under the conditions that cluster centers are sufficiently
+separated and an appropriate transformation is used. \cite{tibshirani2005cluster} proved that the prediction 
+strength method is consistent under the setting that the $k$ clusters are uniformly distributed on $k$ unit balls
+in $p$-dimension, with pairwise distance of centers are at least four.
+
 \section{Cross-validation for selecting the number of clusters}
@@
 
-Although the condition given by Proposition $2$ is succinct in such that it's straight forward to see how
-the structure of covariance matrix $\Sigma_{XY}$ affects the performance of Gabriel CV method, it's not simple and easy to interpret such condition as the one in Proposition $1$.
-In specific, each block of matrix $\Sigma_{XY}$ influences the condition while the combined effect
-determines whether Gabriel CV method chooses $k=1$ or $k=2$.
-
-In practice, one can check such condition using estimated sample covariance matrix for single cluster case. 
-If such condition doesn't met, one solution is rotating the matrix to eliminate the correlation between each dimension then perform Gabriel CV method on the rotated data matrix.
-Another empirically observed solution is
-to leave most columns for clustering, i.e. increase the dimension of $\Sigma_{YY}$. Such solution is especially 
-useful for data containing multiple clusters, where neither estimating sample covariance matrices nor
-rotating to eliminate correlation is feasible. Although the theory for such solution is available for special cases,
-its general version hasn't been developed.
+Above equation gives the condition of when Gabriel CV method would correctly choose $k=1$ over $k=2$.
+Although the expression is succinct, it's not straight forward to see how the structure of covariance matrix $\Sigma_{XY} \in \mathbb{R}^{(p+q) \times (p+q)}$ affects the performance of Gabriel CV method.
+Here, assuming covariance matrix has compound symmetric structure where only the matrix dimension $p+q$ and $\rho$ are variables, i.e.
+$$\Sigma_{XY} = \begin{pmatrix}
+1 & \rho & \cdots & \rho \\
+\rho & 1 & \cdots & \rho \\
+\vdots & \vdots & \cdots & \rho \\
+\rho& \rho & \cdots & 1 \\
+\end{pmatrix}
+$$
+we are able to feel what above equation implies for this specific case.
+Another reason to use the compound symmetric structure covariance matrix is that it's invariant under the permutation of each column vector in the matrix.
+Given that Gabriel CV method randomly choose $p$ columns as $X$ ($q$ columns as $Y$), compound symmetric structure insures that the covariance matrix $\Sigma_{XY}$ always look the same no matter which $p$ columns selected as $X$.
+
+If we do $2$ fold cross-validation in the column, i.e. $p=q$, then the result in Proposition $2$ implies that the boundary value of $\rho$ is 
+\[
+	\rho^* = \frac{1}{p+1}
+\]
+under the compound symmetric structure given above, where $\Sigma_{XY}$ has $p+q=2p$ dimension.
+If $\rho < \rho^*$, then Gabriel CV prefers $CV(1)$ over $CV(2)$ and vice versa.
+It means the boundary value $\rho^*$ depends on the dimension of $\Sigma_{XY}$ linearly.
+
+If dimension is $p+q = 2$ with $p=q$, then above boundary condition shows the boundary value $\rho^* = 0.5$, while for dimension $p+q = 100$, the  boundary value $\rho^* = \frac{1}{51}$.
+However, such value also depends on the ratio of $p$ over $q$.
+Also in $p+q = 100$ dimension, if we pick $p=2$ and $q=98$ (leave majority of the columns for clustering), the the boundary value $\rho^* \in (\frac{1}{6},\frac{1}{7})$.
 
@@
 \end{figure}
-%----------------------------------------------------------------
+
+\subsection{Correlation adjustment for Gabriel method}
+When the correlation between dimensions are high, the proposed Gabriel CV method tends to overestimate the number $k$.
+A simply remedy for the high correlation is available if we assume common covariance structure among the $k$ clusters.
+\begin{enumerate}
+	\item Apply Gabriel CV method on the original data $\dataX$, get estimated number of cluster $\hat{k}$
+	\item Estimate the pooled covariance matrix $\hat{\Sigma}$ from the $\hat{k}$ clusters.
+	\item Let $\hat{\Sigma} = \Gamma\Lambda\Gamma'$ be the Eigendecomposition of $\hat{\Sigma}$, we rescale and rotate the original data $\dataX$ to get $ \widetilde{\dataX} = \dataX\Gamma\Lambda^{-1/2}Q$, where $Q$ is a random  orthonormal rotation matrix.
+	\item Apply Gabriel CV method again on the transformed data $ \widetilde{\dataX}$ to estimate $k$. 
+\end{enumerate}
+
 \section{Simulation}
@@
 \begin{enumerate}
-  \item Single cluster in $10$ dimensions --- $200$ observations, each
-    observation uniformly distributed over $[0,1]$ in each dimension. 
-  \item Two clusters in $4$ dimensions --- $50$ i.i.d
-    observations are generated from both multivariate normal
-    $\mathcal{N}\left(\mu_1,0.5\Sigma\right)$ and multivariate normal
-    $\mathcal{N}\left(\mu_2,1.5\Sigma\right)$, where 
-    cluster center $\mu_1 = (1,0,0,1)$ and $\mu_2 =(1,3.5,3.5,1)$
-    respectively. $\Sigma$ has $AR(1)$ structure with $\rho=-0.2$ and $\sigma=1$.
-  \item Four clusters in $100$ dimensions ---  Each cluster has $100$ or $150$
-    i.i.d standard normal observations, with cluster centers randomly
-    generated from multivariate normal distribution
-    $\mathcal{N}\left(\mathbf{0},0.65^2\mathbf{I}\right)$
-  \item Ten clusters in $100$ dimensions ---  Each cluster has $50$ or $100$
-    i.i.d standard normal observations, with cluster centers randomly
-    generated from multivariate normal distribution
-    $\mathcal{N}\left(\mathbf{0},0.72^2\mathbf{I}\right)$
-  \item Four log-normal clusters in $16$ dimensions --- For each cluster, $30$
-    or $60$ i.i.d centered log-normal observations are generated from
-    $ln\mathcal{N}\left(0,0.5^2 \right)$. The cluster centers are randomly
-    generated from multivariate normal distribution
-    $\mathcal{N}\left(\mathbf{0},1.2^2\mathbf{I}\right)$
-  \item[6.] Three exponential clusters in $20$ dimensions with different
-    variance --- $40$ observations in each cluster are generated from centered
-    exponential distribution ($exp(\lambda)-1/\lambda$ in each dimension),
-    with $\boldsymbol\mu$ randomly generated from
-    $\mathcal{N}\left(\mathbf{0},19\mathbf{I}\right)$ and $\lambda = 1, 1/2,
-    1/5$ respectively.
+
+  \item \textbf{Correlation between dimensions} -- Six clusters in $10$ dimensions. 
+  	Each cluster has $100$ or $50$
+    multivariate normal observations with common covariance matrix $\Sigma$ 
+    which has compound symmetric structure with $1$ in diagonal and $\rho$ 
+    off diagonal. For each $\rho$, $100$ data are simulated and the proportion each
+    method successfully picks $k=6$ is reported. $\rho$ takes value in $\{0,0.1,...,0.9\}$.
+
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting1/Overlay.pdf}
+	\label{fig:setting1}
+	\end{figure}
+
+  \item \textbf{Noise dimensions} --- Three clusters in $6$ dimensions. Each cluster has $100$ or $50$
+    mean zero multivariate normal observations with identity covariance matrix. 
+    We add $P$ dimensions of noise to the data, which is randomly generated from uniform$[0,1]$. The noise
+    dimension $P$ takes values in $\{0,6,...,54\}$.
+
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting2/Overlay.pdf}
+	\label{fig:setting2}
+	\end{figure}
+
+  \item \textbf{High dimension} --- Eight clusters in $p$ dimensions, $p$ takes values 
+  	in $\{10,20,...,100\}$. Each cluster has $100$ or $50$ mean zero multivariate normal 
+  	observations with identity covariance matrix.
+
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting3/Overlay.pdf}
+	\label{fig:setting3}
+	\end{figure}
+
+     \item \textbf{Variance heterogeneity} --- Three clusters in $20$ dimensions, each has
+     $60$ observations in it. Observations are generated from 
+     $\mathcal{N}\left(\mathbf{0},\sigma_1^2\mathbf{I}\right)$,
+      $\mathcal{N}\left(\mathbf{0},\sigma_2^2\mathbf{I}\right)$ and
+       $\mathcal{N}\left(\mathbf{0},\sigma_3^2\mathbf{I}\right)$ where 
+       $\sigma_1^2 : \sigma_2^2: \sigma_3^2 = 1:\frac{1+R}{2}:R$. The maximum ratio $R$
+       takes values in $\{1,5,10,...,45\}$
+
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting4/Overlay.pdf}
+	\label{fig:setting4}
+	\end{figure}
+
+	\item \textbf{Heavy tail data} --- Five clusters in $15$ dimensions, each has
+     $80$ observations in it. Observations have independent $t$ distribution in each dimension  
+     with degree of freedom $\nu$, which takes value in $\{2,3,...,11\}$
+
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting5/Overlay.pdf}
+	\label{fig:setting4}
+	\end{figure}
 \end{enumerate}
-Note that in setting $2$--$6$, all clusters are well-separated, i.e. no
-overlapping. In fact, any simulated clusters with minimum distance less than
-1 unit was discarded, so there is clear definition of true number of
-clusters. The parameters in setting $2$--$6$ are chosen such that about half of
-the random realization were discarded. The idea is borrowed from
+
+Note that in all settings above, cluster centers are randomly generated from 
+multivariate normal distribution $\mathcal{N}\left(\mathbf{0},\varsigma \mathbf{I}\right)$. 
+All clusters are well-separated, i.e. no overlapping. In fact, any simulated 
+clusters with minimum distance less than $1$ unit was discarded, so there is
+clear definition of true number of clusters. The parameters $\varsigma$ is chosen 
+such that about half of the random realization were discarded. The idea is borrowed from
 \cite{tibshirani2001estimating}.
 
-Table \ref{table1} shows the distribution of $k$ selected by each algorithm in 
-each simulation setting. We run $k$ from $1$ to $15$, and the best $k$ within this range
-is selected by the algorithms. NA in the table means algorithm did not converge or otherwise 
-failed. 
-\clearpage
-
-\begin{center}
-\footnotesize
-\begin{longtable}{lcccccccccccccccc}
-\captionsetup{justification=centering}
-\caption{\label{table1} Simulation Results}\\
-\toprule
-\phantom{Pred.~Strength}
-  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
-\cmidrule(lr){2-17}
-Method
-  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
-\midrule
-\endfirsthead
-\multicolumn{17}{c}%
-  {\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
-\midrule
-\phantom{Pred.~Strength}
-  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
-\cmidrule(lr){2-17}
-Method
-  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
-\midrule
-\endhead
-\multicolumn{17}{r}{\textit{Continued on next page}} \\
-\endfoot
-\bottomrule
-\multicolumn{17}{l}{\footnotesize
-  \textsuperscript{$\ast$} CH, Hartigan, and Stability are excluded from
-  Setting~1 because they cannot select $k=1$.
-}
-\endlastfoot
-\input{demo/bench/setting1/results.tex}
-\midrule
-\input{demo/bench/setting2/results.tex}
-\midrule
-\input{demo/bench/setting3/results.tex}
-\midrule
-\input{demo/bench/setting4/results.tex}
-\midrule
-\input{demo/bench/setting5/results.tex}
-\midrule
-\input{demo/bench/setting6/results.tex}
-\end{longtable}
-\end{center}
-
-\noindent
-All methods are used with their default parameter setting except for Gap
-statistics. We selected the $k$  corresponds to the global maximum of gap
-statistics, rather than the default method which selects the smallest $k$ such
-that the gap statistics is not more than 1 standard error away from the first
-local maximum. This is because among all the possible options, the global
-maximum criteria gives the best result in our simulation settings.
-
-For CH-index, Hartigen and Bootstrap stability, the minimum $k$ they can pick is $2$. So they are not
-included for comparison since the true $k=1$. It also demonstrates the difficulty 
-that null scenario poses for some existing algorithms. For those could select $k=1$,
-most of them work well under such setting except for Jump method. Setting $2-4$ have Gaussian
-distribution, where the Gaussian mixture model-based method suppose to
-perform the best. However, it only performs reasonably well in setting $2$. In setting $3$ and $4$,
-it has a very poor performance even if the underlying distribution is indeed
-Gaussian mixture model. It highlights the difficulty that high dimension
-data could cause. Other methods (CH-index, Hartigan, Jump
-method, Prediction strength and Boostrap stability) also have problems in
-these two high-dimension settings as we could not find any one with reasonable
-performance in terms of finding the correct $k$ in both cases. The only exception is
-Gap statistics, which works perfectly in both settings. Our proposed Gabriel method (also Wold
-CV method) clearly stands out from the basket of existing methods under the 
-high-dimension setting $3$ and $4$, as we can see they both perform perfectly under
-such settings.
-
-Setting $5$ and $6$ corresponding to the situation where data are heavy-tailed
-instead of normal distributed, coupled with unequal number of observations or
-unequal variance for different clusters. From Table \ref{table1}, we can see none of 
-the existing methods has good performance in both settings. In contrast,
-our proposed Gabriel method is robust and has superior performance
-compare with other methods. Because one can hardly tell the underlying
-distribution of data in practice, the resilience to non-Gaussian data gives
-the Gabriel method a clear edge. 
-%----------------------------------------------------------------
 \section{Real data application}

### 2015-12-05 New simulation/code/gabriel.R

@@ -1,0 +1,200 @@
+# code/gabriel.R
+#
+# Depends:
+#   library("bcv")
+#   library("MASS")
+#   library("nnet")
+#   source("classify.R")
+#   source("cluster.R")
+#
+
+
+cv.kmeans.gabriel <- function(x, krow = 2, kcol = 2, maxcenters = 10,
+                              classify.method = c("nearest", "lda-equal", "lda-proportions", "svm"))
+{
+    classify.method <- match.arg(classify.method)
+    x <- as.matrix(x)
+    n <- nrow(x)
+    p <- ncol(x)
+    if (n < 2) 
+        stop("x should have at least two rows")
+    if (p < 2) 
+        stop("x should have at least two columns")
+    if ((krow > n) || (krow <= 1)) 
+        stop("krow outside allowable range")
+    if ((kcol > p) || (kcol <= 1)) 
+        stop("kcol outside allowable range")
+    if (maxcenters <= 0) 
+        stop("maxcenters should be positive")
+
+    krow.o <- krow
+    krow <- bcv:::round.fold(n, krow)
+    kcol.o <- kcol
+    kcol <- bcv:::round.fold(p, kcol)
+    if (krow != krow.o) 
+        warning("krow has been set to ", krow)
+    if (kcol != kcol.o) 
+        warning("kcol has been set to ", kcol)
+
+    s.r <- bcv:::choose.sets(n, krow)
+    s.c <- bcv:::choose.sets(p, kcol)
+    n0 <- n - max(table(s.r))
+    p0 <- p - max(table(s.c))
+    maxcenters.o <- maxcenters
+    maxcenters <- min(n0, round(maxcenters.o))
+    if (!missing(maxcenters) && maxcenters != maxcenters.o) 
+        warning("maxcenters has been set to ", maxcenters)
+
+    if (classify.method == "nearest") {
+        classify <- classify_nearest
+    } else if (classify.method == "lda-equal") {
+        classify <- function(x, grouping)
+            classify_lda(x, grouping, prior="equal")
+    } else if (classify.method == "lda-proportions") {
+        classify <- function(x, grouping)
+            classify_lda(x, grouping, prior="proportions")
+    } else if (classify.method == "svm") {
+        stop("not yet implemented")
+    }
+
[…]

### 2015-12-05 New simulation/demo/bench/04_plot_results.R

@@ -1,0 +1,82 @@
+
+library(ggplot2)
+
+nclusters <- list("setting1" = 6L,
+		     "setting2" = 3L,
+			 "setting3" = 8L,
+			 "setting4" = 3L,
+			 "setting5" = 5L)
+
+Scale_X <-  list("setting1" = seq(0.0,0.9,0.1),
+		     "setting2" = 6*seq(0,9,1),
+			 "setting3" = seq(10,100,10),
+			 "setting4" = c(1,seq(5,45,5)),
+			 "setting5" = seq(2,11,1))
+
+Label_X <- list("setting1" = "Rho",
+		    "setting2" = "Noise Dimension",
+			"setting3" = "Data Dimension",
+			"setting4" = "Max Variance Ratio",
+			"setting5" = "Degree of freedom")
+
+
+for (s in list.dirs(full.names=FALSE, recursive=FALSE)){
+	## remove the object so the if cond checkd out for next iter
+	rm(Result_data)
+	rm(Fit_result)
+
+	Result.path <- file.path(s,"result.rds")
+
+	if(!file.exists(Result.path)){
+		Fit.result <- list()
+
+		for (m in names(methods)) {
+	 		Matrix_result <- matrix(NA,100,10)
+			for(c in 1:10){
+				f.replicates <- file.path(s, paste0("C" ,c), "method",paste0(m ,".rds"))
+				if(file.exists(f.replicates)){
+					Matrix_result[,c]<- readRDS(f.replicates)
+				}
+			}
+			Fit.result[[m]] = Matrix_result
+		}
+		saveRDS(Fit.result, Result.path)
+	}
+	####----------------------Read from RDS and make plot-----------------
+
+	Fit_result <- readRDS(Result.path)
+	
+	for(i in 1:length(Fit_result)){
+		Proportion <- as.numeric(apply(Fit_result[[i]], 2, function(x) sum(x==nclusters[[s]],na.rm=T)))
+		X_axis <- Scale_X[[s]]
+		Methods <- rep(names(Fit_result)[i],10)
+
+		if(!exists("Result_data")){
+			Result_data <- as.data.frame(cbind(X_axis,Proportion,Methods))
+		}else{
+			Result_data <- rbind(Result_data,as.data.frame(cbind(X_axis,Proportion,Methods)))
+		}	
+	}
+	Result_data[,1] <- as.numeric(as.character(Result_data[,1]))
[…]

### 2015-12-05 New simulation/demo/bench/settings.R

@@ -1,0 +1,358 @@
+# demo/bench/settings.R
+#
+# Depends:
+#   library("MASS") # mvrnorm
+#
+##-----------------------------------------------
+Cov <- function(dim = 2, rho = 0){
+	COV <- diag(dim)
+	for(i in 1:dim){
+		for(j in 1:dim){
+			if(i!= j){COV[i,j] = rho}
+		}
+	}
+	return (COV)
+}
+##-------------------------------------------------
+Data_generate <- function(Dim = 10, C.num = 4, Para = 1, Rho = 0, Obs.num = 100){ 
+	condition <- TRUE
+	N = 0
+  	while(condition){
+		N=N+1
+		del = 1
+		nn = sample(c(Obs.num/2,Obs.num),size=C.num,replace=T)
+		cl = rep(1:C.num,nn)
+		 
+		Centers <- mvrnorm(C.num, mu=rep(0,Dim),Sigma = (Para^2)*diag(Dim))
+
+		for(j in 1:nrow(Centers)){
+			if(j==1){
+				X.data = mvrnorm(nn[j], mu = Centers[j,], Sigma = Cov(Dim,Rho))
+			}else{
+				DATA = mvrnorm(nn[j], mu = Centers[j,], Sigma = Cov(Dim,Rho))
+				X.data = rbind(X.data, DATA)
+			}
+		}
+		
+		ss=dist(rbind(X.data,Centers))
+		d=matrix(0,nrow=nrow(X.data)+C.num,ncol=nrow(X.data)+C.num)
+ 		d[row(d)>col(d)]=ss
+ 		DD=d[(nrow(X.data)+1):(nrow(X.data)+C.num),1:nrow(X.data)]
+		
+		for(i in 1:ncol(DD)){ DD[cl[i],i]=DD[cl[i],i]+del}
+		ff=apply(DD,2,which.min)
+
+		if(sum(ff==cl)==nrow(X.data)){
+ 			condition <- FALSE
+ 		}
+	}#end while loop
+	ID <- sample(1:nrow(X.data),nrow(X.data),replace = FALSE)
+    	result <- list(Data = X.data[ID,], Number = N)
+    	return(result)
+}
+
+##MM <- Data_generate(dim = 10, C.num = 10, Para = 0.72, Rho = 0, Obs.num = 100)
+###-----------------------------------------------------
+NumA <- function(P = 1, dim, c.num, rho, obs.num){
+	set.seed(1)
+	NN <- rep(NA,50)
+	for(W in 1:50){
[…]

### 2015-12-05 New simulation/demo/elbow/01_make_plots.R

@@ -1,0 +1,111 @@
+#!/usr/bin/Rscript --vanilla
+
+library("MASS") 	# mvrnorm
+library("RColorBrewer") # brewer.pal
+
+source("../../code/cluster.R") # cluster_kmeans
+
+
+
+##---------------------Top plot of figure 1-----------------------------------
+
+set.seed(3) # seed chosen to make well-separated clusters
+
+x1 <- mvrnorm(20, c(1.7, 3.0), 0.15 * diag(2))
+x2 <- mvrnorm(20, c(1.0, 1.0), 0.15 * diag(2))
+x3 <- mvrnorm(20, c(3.0, 1.5), 0.15 * diag(2))
+x4 <- mvrnorm(20, c(3.7, 3.7), 0.10 * diag(2))
+
+data <- rbind(data.frame(x=x1[,1], y=x1[,2], group="1"),
+	      data.frame(x=x2[,1], y=x2[,2], group="2"),
+	      data.frame(x=x3[,1], y=x3[,2], group="3"),
+	      data.frame(x=x4[,1], y=x4[,2], group="4"))
+
+
+
+mar <- c(3.25, 3.25, 1.75, 1.75)
+csi <- 0.2 # default value of par("csi")
+mai <- mar * csi
+w <- h <- 3
+mypar <- list(las=1, mai=mai, mgp=c(2.25, 0.75, 0), ps=11, tcl=-0.4)
+
+width <- w + sum(mai[c(1,3)])
+height <- h + sum(mai[c(2,4)])
+
+
+
+pdf("correct-data.pdf", width=width, height=height)
+palette(brewer.pal(6, "Set1"))
+par(mypar)
+
+plot(data$x, data$y, type="n", xlab="X", ylab="Y")
+points(data$x, data$y, pch=as.integer(data$group),
+       col=as.integer(data$group))
+axis(3, labels=FALSE)
+axis(4, labels=FALSE)
+dev.off()
+
+
+centers <- 1:10
+tot.withinss <- rep(NA, length(centers))
+for (i in seq_along(centers)) {
+	cl <- cluster_kmeans(cbind(data$x, data$y), centers[i])
+    tot.withinss[i] <- cl$tot.withinss
+}
+
+
+pdf("correct-withinss.pdf", width=width, height=height)
+par(mypar)
+
+plot(centers, tot.withinss, type="b",
+     xlab = "Clusters",
+     ylab = "Within SS")
+axis(3, labels=FALSE)
[…]

### 2015-12-05 New simulation/demo/overlap/02_make_plot.R

@@ -1,0 +1,43 @@
+
+Raw <- read.csv("Raw_result.csv")
+## "Group" column accidently deleted from above data file
+Raw$Group <- rep(1:(31*31),each=10)
+
+Number <- rep(NA,30*30)
+X_ax <- rep(NA,30*30)
+Y_ax <- rep(NA,30*30)
+
+for(i in unique(Raw$Group) ){
+	X_ax[i] <- unique(Raw[Raw$Group==i,]$U_x)
+	Y_ax[i] <- unique(Raw[Raw$Group==i,]$U_y)
+	Number[i] <-  sum(Raw[Raw$Group==i,]$k == 2)
+}
+
+library(plotrix)
+
+pdf("color_plot.pdf")
+
+plot(X_ax,Y_ax,col=color.scale(-Number,c(2,1,0),c(0,2,4),c(3,5,6)),pch=16,cex=2,xlab="U_x", ylab="U_y")
+##-----------------------------------------------
+
+YtoX <- function(U_y){
+	Num <- 2*dnorm(U_y)+2*U_y*pnorm(U_y)-U_y
+	NM <- (Num/U_y +2)/4
+	if(NM<1){
+		return(qnorm(NM))
+	}else{
+		return(NA)
+	}
+}
+##-----------------------------------
+Y_vector <- seq(0.437,3,0.01)
+X_vector <- rep(NA,257)
+
+for(j in 1:257){
+	X_vector[j] <- YtoX(Y_vector[j])
+}
+
+
+lines(X_vector,Y_vector, lwd=3) 
+
+dev.off()

### 2015-12-16 New simulation/technometrics.tex
Add error bar, more consist theory and some editing

@@
 
-Among all the proposed methods, very few provide theoretical consistency result for choosing $k$. 
+Some proposed methods provide theoretical consistency result for choosing $k$. 
 Assume data is a mixture of $G$ $p$-dimension clusters with equal prior, identically distributed with 
@@
 the Jump method will pick the correct $k$ under the conditions that cluster centers are sufficiently
-separated and an appropriate transformation is used. \cite{tibshirani2005cluster} proved that the prediction 
-strength method is consistent under the setting that the $k$ clusters are uniformly distributed on $k$ unit balls
-in $p$-dimension, with pairwise distance of centers are at least four.
+separated and an appropriate transformation is used. Specifically, let $\Delta\sqrt{p}$ denotes the 
+minimum Euclidean distance between cluster centers, \citet{sugar2003finding} shows that the jump will
+be maximized at $k=G$ given that $\Delta > 6$ and the existence of a positive $Y$ such that
+\[	\left(\frac{p\Delta^2W}{9G}\right)^{-Y}+\left(W\left[\frac{2^{2H^*(X)}}{K^2_{max}2\pi e}\right]-(\frac{\Delta}{6})^2(1-W) \right)^{-Y} < 2	\]
+and $\left(\frac{p\Delta^2W}{9G}\right)^{-Y} <1/2$, where $H^*(X)$ is the minimum entropy of the cluster 
+membership over each dimension and $W =1- \frac{6^4V_{\mathbf{X}}}{(\Delta^2-36)^2}$ with $V_{\mathbf{X}} = var\left( 1/p ||\mathbf{X}-\mu_j||^2_{\Gamma^{-1}} \mid \mathbf{X} \hspace{0.05in}\text{in jth cluster}\right)$
+
+\cite{tibshirani2005cluster} proved that the prediction strength method is consistent under the setting 
+that the observations in $p$-dimension are uniformly distributed within one unit from their respective cluster centers, with minimum distance between the $G$ cluster centers is four unit.
+Under such well-separated setting,
+they show that 
+\[ ps(G) = 1+ \OhP(1), \hspace{0.5in} \underset{k \geq G+1}{\text{sup}} ps(k)\leq \frac{2}{3}+ \OhP(1) \]
+therefore $\hat{k}$ is consistent in estimating $G$.  
+When data does come from mixture of Gaussian distribution, the model-based method of \citet{fraley2002model} is consistent in estimating $k$ in low dimension.
 
@@
 Bootstrap stability \citep{fang2012selection} in following simulation
-settings. We select $p=q$ and $5-$fold cross-validation in row
+settings. All methods are executed with their default parameter settings. 
+We select $p=q$ and $5-$fold cross-validation in row
 ($m=\frac{1}{4}n$) as default parameter setting for our proposed Gabriel
 method. Note that set $p=q$ corresponding to $2-$fold cross-validation in
-column.
+column. Wold method (detailed in Appendix $A$) and correlation-corrected 
+Gabriel method (section $4.3$) are also included for comparison.
+
+Note that in all settings, cluster centers are randomly generated from 
+multivariate normal distribution $\mathcal{N}\left(\mathbf{0},\varsigma \mathbf{I}\right)$. 
+All clusters are well-separated, i.e. no overlapping. In fact, any simulated 
+clusters with minimum distance less than one unit was discarded, so there is
+clear definition of true number of clusters. The parameters $\varsigma$ is chosen 
+such that about half of the random realization were discarded. The idea is borrowed from
+\cite{tibshirani2001estimating}. The proportion of each method successfully picks the correct
+$k$ out of $100$ simulation trials is reported, along with its corresponding confident interval
+by Wilson's method \citep{wilson1927probable}.
 
@@
     which has compound symmetric structure with $1$ in diagonal and $\rho$ 
-    off diagonal. For each $\rho$, $100$ data are simulated and the proportion each
-    method successfully picks $k=6$ is reported. $\rho$ takes value in $\{0,0.1,...,0.9\}$.
+    off diagonal. $\rho$ takes value in $\{0,0.1,...,0.9\}$.
 
@@
 	\centering
-	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting1/Overlay.pdf}
+	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting1/Facet.pdf}
 	\label{fig:setting1}
@@
 
-  \item \textbf{Noise dimensions} --- Three clusters in $6$ dimensions. Each cluster has $100$ or $50$
+  \item \textbf{Noise dimensions} --- Three clusters in $6$ dimensions. Each cluster has $1000$ or $500$
     mean zero multivariate normal observations with identity covariance matrix. 
@@
 
-	\begin{figure}[H]
-	\centering
-	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting2/Overlay.pdf}
-	\label{fig:setting2}
-	\end{figure}
+	%\begin{figure}[H]
+	%\centering
+	%\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting2/Facet.pdf}
+	%\label{fig:setting2}
+	%\end{figure}
 
@@
 	\centering
-	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting3/Overlay.pdf}
+	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting3/Facet.pdf}
 	\label{fig:setting3}
@@
 	\centering
-	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting4/Overlay.pdf}
+	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting4/Facet.pdf}
 	\label{fig:setting4}
@@
      $80$ observations in it. Observations have independent $t$ distribution in each dimension  
-     with degree of freedom $\nu$, which takes value in $\{2,3,...,11\}$
+     with degree of freedom $\nu$, which takes value in $\{11,10,...,2\}$
 
@@
 	\centering
-	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting5/Overlay.pdf}
+	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting5/Facet.pdf}
 	\label{fig:setting4}
@@
 
-Note that in all settings above, cluster centers are randomly generated from 
-multivariate normal distribution $\mathcal{N}\left(\mathbf{0},\varsigma \mathbf{I}\right)$. 
-All clusters are well-separated, i.e. no overlapping. In fact, any simulated 
-clusters with minimum distance less than $1$ unit was discarded, so there is
-clear definition of true number of clusters. The parameters $\varsigma$ is chosen 
-such that about half of the random realization were discarded. The idea is borrowed from
-\cite{tibshirani2001estimating}.
+From the graph of setting $1$, we can see high correlation between dimensions cause problem for most 
+existing methods as well as the proposed Gabriel CV method and Wold method. Higher correlation clearly 
+degenerate most algorithms' ability to find the correct $k$ as data no longer spherical in $p$-dimension.
+And the recent proposed methods (Gabriel CV method, Gap statistic, Jump method) seem to be more sensitive to
+high correlation than those old methods such as CH-index and Hartigan statistics. The only two methods 
+work well in high correlation are the Gaussian model-based BIC method \citep{fraley2002model} and the 
+correlation-corrected Gabriel method.
+
+Plot of setting $3$ shows some methods are insensitive to higher dimension such as Jump method and Gap 
+statistics, while others deteriorate quickly with increasing dimension, most notably the Gaussian 
+model-based BIC method. Note here the data are still generated from mixture of Gaussian. In contrast, our 
+proposed Gabriel CV method as well as its correlation-corrected version actually working better in higher 
+dimension.
+
+Result of setting $4$ indicates most previously methods are sensitive to variance heterogeneity among 
+clusters, most notably the Gap statistics and model-based BIC method. The proposed Gabriel CV method
+and its correlation-corrected version consistently perform well in estimating $k$ and quite insensitive
+to variance heterogeneity. The Wold method also performs well in this setting.
+
+The last non-Gaussian setting examines methods' performance in heavy-tail data. With degree of freedom 
+decreases, the tail becomes more flat and the Gaussian assumption becomes more inappropriate. It explains
+the sharp plunge of Gaussian model-based BIC method in the plot. For most methods, their performances are 
+relatively stable until the tail gets very heavy. In case where degree of freedom is $2$, Gap statistics 
+and Jump method dived considerably compare to the Gabriel method and Wold method.
+
+In conclusion, the proposed Gabriel CV method compares well with current state-of-the-art methods and 
+is very robust against variance heterogeneity, high dimension and heavy-tail data. One weakness of the 
+Gabriel CV method is that it's sensitive to the high correlation among dimensions. In such case, its 
+correlation-corrected version will perform well under the assumption of common covariance structure.
 
@@
 in our limited simulation settings and real data application, and clearly the superior
-one when data is high dimension or is heavy-tailed.
-
-Besides no modeling assumption is required, our proposed method is robust against data set 
+one when data has heterogeneous variance or is heavy-tailed.
+
+Besides no strong modeling assumption is required, our proposed method is robust against data set 
 with variance heterogeneity, unequal number of observations, non-Gaussian noise and high-dimension. 
@@
 columns. In practice, many data sets don't exhibit high correlations between columns, where
-our proposed method can be safely applied. In case the high correlation does exist, some procedure
-such as rotating the data set and/or leave out most columns for clustering may be used to
-reduce its effect as previously mentioned. However, the theory for those procedure have not
+our proposed method can be safely applied. In case the high correlation does exist, 
+the correlation-corrected version of Gabriel CV method in section $4.3$ can be used if we can 
+assume common covariance structure. Other procedure such as leave out most columns for clustering 
+may also be used to reduce its effect. However, the theory for such procedure has not
 been fully developed and it could be the future research topic.

### 2015-12-16 New simulation/demo/bench/04_plot_results.R

@@ -1,4 +1,4 @@
-
+library("PropCIs")
 library(ggplot2)
 
 nclusters <- list("setting1" = 6L,
@@ -59,21 +59,23 @@ for (s in list.dirs(full.names=FALSE, recursive=FALSE)){
 	}
 	Result_data[,1] <- as.numeric(as.character(Result_data[,1]))
 	Result_data[,2] <- as.numeric(as.character(Result_data[,2]))
-
+	
+	Result_data$lower <- sapply(Result_data$Proportion,function(x) 100*scoreci(x,100,0.95)$conf.int[1])
+	Result_data$upper <- sapply(Result_data$Proportion,function(x) 100*scoreci(x,100,0.95)$conf.int[2])
 	###==================================================================
 	if(s == "setting5"){
 		ggplot(Result_data, aes(X_axis,Proportion, group = Methods, shape = Methods, colour = Methods))+labs(x = Label_X[[s]])+ geom_line(size = 1.3) + scale_x_reverse()
 		ggsave(file.path(s,"Overlay.pdf"),width = 11, height = 7)
 
 	
-		ggplot(Result_data, aes(X_axis,Proportion, group = Methods, colour = Methods))+ geom_line(size = 1.3)+labs(x = Label_X[[s]])+facet_wrap(~Methods,scales = "free_y",nrow=3)+scale_x_reverse()
+		ggplot(Result_data, aes(X_axis,Proportion, group = Methods, colour = Methods))+ geom_line(size = 1.3)+labs(x = Label_X[[s]])+facet_wrap(~Methods,nrow=2)+guides(colour=FALSE)+geom_errorbar(aes(ymax = upper, ymin=lower),colour = "black")+scale_x_reverse()
 		ggsave(file.path(s,"Facet.pdf"),width = 11, height = 7)
 	}else{
 		ggplot(Result_data, aes(X_axis,Proportion, group = Methods, shape = Methods, colour = Methods))+labs(x = Label_X[[s]])+ geom_line(size = 1.3)
 		ggsave(file.path(s,"Overlay.pdf"),width = 11, height = 7)
 
 	
-		ggplot(Result_data, aes(X_axis,Proportion, group = Methods, colour = Methods))+ geom_line(size = 1.3)+labs(x = Label_X[[s]])+facet_wrap(~Methods,scales = "free_y",nrow=3)
[…]

### 2015-12-16 New simulation/demo/bench/settings.R

@@ -295,11 +295,13 @@ setting1 <- function(c){
 			list(x=MM$Data, ngroup = 6)
 }
 
-## Para3_6dim = Para.hunt(Dim = 6, C.num = 3, Rho = 0, Obs.num=100, Initial = 5)
-Para3_6dim = 2.9
+###### Para3_6dim = Para.hunt(Dim = 6, C.num = 3, Rho = 0, Obs.num=100, Initial = 5)
+###### Para3_6dim = 2.9
+## Para3_6dim = Para.hunt(Dim = 6, C.num = 3, Rho = 0, Obs.num=1000, Initial = 4)
+Para3_6dim = 3.2375
 
 setting2 <- function(c){
-			MM <- Data_generate(Dim = 6, C.num = 3, Para = Para3_6dim, Rho = 0, Obs.num = 100)
+			MM <- Data_generate(Dim = 6, C.num = 3, Para = Para3_6dim, Rho = 0, Obs.num = 1000)
 			if(c==1){
 				Data.x =  MM$Data
 			}else{

### 2015-12-16 New simulation/wei/Data Example.R

@@ -1,0 +1,209 @@
+
+   library("bcv")
+   library("devtools")
+   library("cluster")
+   library("e1071")
+   library("mclust")
+   library("MASS")
+   library("nnet")
+   load_all("../lib/fpc")
+   load_all("../lib/NbClust")
+   source("../lib/NbClust.R")
+   source("../code/classify.R")
+   source("../code/cluster.R")
+   source("../code/gabriel.R")
+   source("../code/jump.R")
+   source("../code/wold.R")
+
+Vote <- read.table("Voting.txt",sep=",")
+
+Id <- vector()
+
+for(i in 1:nrow(Vote)){
+      if(sum(Vote[i,] == "?")!=0){
+		Id <- c(Id,i)
+	}
+}
+
+vote <- Vote[-Id,-1]
+
+VV <- matrix(NA,232,16)
+
+for(i in 1:232){
+	for(j in 1:16){
+		if(vote[i,j]=="n"){
+			VV[i,j]=0
+		}else if(vote[i,j]=="y"){
+			VV[i,j]=1
+		}
+	}
+}
+
+set.seed(3)
+ID <- sample(1:232, 232)
+
+Vote <- VV[ID,]
+set.seed(0)
+cv.kmeans.gabriel(Vote, 5, 2, maxcenters=10, classify.method="nearest")$centers
+
+set.seed(0)		
+gabriel_cor_correct(Vote, maxcenters=10, type = 2)
+
+set.seed(0)		
+Wold_holdout(data = Vote, CV = 5, Errortol = 0.01,  max.k =10)
+
+set.seed(0)		
+Gap <- clusGap(Vote, FUN = kmeans, K.max =10)
+which(Gap[[1]][,3] == max(Gap[[1]][,3]))
+			
+set.seed(0)	
+mcluster <- Mclust(Vote, G = 1:10)
+ mcluster$G
+						
+set.seed(0)
+Ch <- NbClust(Vote, min.nc = 2, max.nc = 10,method = "kmeans", index = "ch")
+Ch$Best.nc[1]
+	
+set.seed(0)
+Hartigan <- NbClust(Vote, min.nc = 2, max.nc = 10,method = "kmeans", index = "hartigan")
+Hartigan$Best.nc[1]
+			
+set.seed(0)
+Jump <- jump(Vote,plotjumps=F,rand=10,trace=F)
+Jump$maxjump
+	
+set.seed(0)
[…]

### 2015-12-31 New simulation/technometrics.tex
Finish revisions

@@
 true class (label) in data set makes the adoption of cross-validation into
-unsupervised leaning problem difficult.
+unsupervised learning problem difficult. A naive algorithm may works as following:
+with the data split by rows into training set and test set, cluster the training 
+data with parameter $k$; then predict each observation in test data by the closest
+cluster center returned in last step. The prediction error is calculated by the 
+difference between the predicted cluster centers and observations' true value, and algorithm 
+picks the $k$ which minimizes such error. Such algorithm will always pick the maximum number 
+of $k$ allowed because more cluster centers means more tight fit to the data space, therefore
+smaller within-cluster dispersion even if the prediction is evaluated on an independent
+test set \citep{hastie2009elements}.
 
@@
 The prediction error measure in our proposed method is exactly the same as in supervised learning.
-Our method is a complete data-driven approach which doesn't rely on
-any model assumption. Through novel form of partitioning data set, we effectively 
+Our method is a complete data-driven approach which doesn't rely on strong model assumption. Through novel form of partitioning data set, we effectively 
 transferred an unsupervised learning problem into a supervised leaning problem, which is one of
@@
 they show that 
-\[ ps(G) = 1+ \OhP(1), \hspace{0.5in} \underset{k \geq G+1}{\text{sup}} ps(k)\leq \frac{2}{3}+ \OhP(1) \]
+\[ ps(G) = 1+ o_p(1), \hspace{0.5in} \underset{k \geq G+1}{\text{sup}} ps(k)\leq \frac{2}{3}+ o_p(1) \]
 therefore $\hat{k}$ is consistent in estimating $G$.  
 When data does come from mixture of Gaussian distribution, the model-based method of \citet{fraley2002model} is consistent in estimating $k$ in low dimension.
+
+Let $\Psi$ denotes any given base-clustering algorithm (e.g $k$-means), and $\Psi(Z,k)$ the clustering obtained by applying $\Psi$ on data $Z$ with parameter $k$.
+Also let $k_0$ be the optimal number of cluster and $d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ be the measure of distance between clustering $\Psi(Z_1,k)$ and $\Psi(Z_2,k)$ where $Z_1$ and $Z_2$ are two independent samples from population.
+\citet{wang2010consistent} shows that their proposed method will consistently select $k = k_0$, given such $k_0$ exists and $d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ converges to zero at proper rate $r_k$.
+However, $k_0$ is defined as the most stable number for $\Psi$ among all the possible $k$, instead of the number of components in mixture model or well-separated compact regions commonly seen in literature.
+Therefore, their consistent result actually is convergent result, i.e.
+the selected $k$ converges to its limit $k_0$ given such limit exists.
 
@@
 
+The reason why Gabriel CV method tends to select larger $k$ when correlation is high 
+between dimensions is that it resembles the naive method mentioned in section \ref{sec:intro} 
+under such circumstance. In fact, if $X$, $Y$ are perfectly correlated, Gabriel method is 
+equivalent to the naive method.
+
 Proposition $1$ gives a very simple condition for the Gabriel method to correctly 
@@
 Also in $p+q = 100$ dimension, if we pick $p=2$ and $q=98$ (leave majority of the columns for clustering), the the boundary value $\rho^* \in (\frac{1}{6},\frac{1}{7})$.
+
+\begin{theorem}
+Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
+of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
+independent draw from a mean-zero multivariate normal distribution with unit
+marginal variances and correlation $\rho$.  In this case, the data are drawn
+from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 0.5$,
+then $\CV(1) < \CV(k)$, $k > 1$ with probability tending to one as $m$ and $n$ increase.
+\end{theorem}
+
+\begin{proof}
+Given $(X_1, Y_1), \dotsc, (X_n, Y_n)$, let $A_k = \{a_1,a_2,...,a_k\}$ denotes the set of cluster centers from applying $k$-means to $\{ Y_i\}_{i=1}^{n}$ with parameter $k$.
+Because $\{ Y_i\}_{i=1}^{n}$ is symmetric, it's easy to see that the cluster centers are symmetric around $0$, i.e.
+if $a^* \in A_k$ then $-a^* \in A_k$.
+The $k$ clusters are symmetric in the same sense.
+Also, if $k$ is odd, then only one cluster contains both negative and positive $Y$ and it's symmetric with center $0$.
+The rest have pure positive or negative intervals; if $k$ is even, then all clusters have pure positive or negative intervals.
+
+Let $a_i$ denotes a cluster center with interval $[b_i, c_i]$.
+By above argument, either $0 \leq b_i < c_i$ ($b_i < c_i \leq 0$) or $b_i = -c_i$ which corresponding to $a_i=0$ with $k$ be odd.
+By symmetry, it's sufficient to only consider case $0 \leq b_i < c_i$ (with $a_i > 0 $) and $b_i = -c_i$ (with $a_i = 0 $) with $\rho \geq 0$.
+
+From joint normality of $X$ and $Y$, it follows that $X \mid Y$ is normal with mean $\rho Y$. Therefore, the mean value $\bar{X}_i$ of $X$ in cluster with $Y$ center $a_i$ is
+
+\begin{align*}
+	\bar{X}_i &= E[X \mid b_i \leq Y \leq c_i] \\
+			&= E[ E(X \mid Y) \mid b_i \leq Y \leq c_i] \\
+			&= E[\rho Y \mid b_i \leq Y \leq c_i] \\
+			&= \rho E[ Y \mid b_i \leq Y \leq c_i] \\
+			&= \rho a_i
+\end{align*}
+
+That is, the mean values of $X$ in the $k$ clusters is $\{\rho a_1,\rho a_2,...,\rho a_k\}$.
+Because we assign each observation to the closest $\bar{X}_i$, $i=1,2,...,k$, it's easy to see the corresponding interval on $X$ for $\bar{X}_i = \rho a_i$ is $[\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}]$, where $a_{i-1}$ and $a_{i+1}$ are the two adjacent centers to $a_i$ on $Y$.
+Note that $[\frac{ \left( a_{i-1}+a_i\right)}{2}, \frac{\left( a_{i}+a_{i+1}\right)}{2}] = [b_i, c_i]$ by the algorithm definition of $k$-means.
+The $Y$ center for such interval on $X$ is
+
+\begin{align*}
+	\hat{\mu}^Y_i &= E[Y \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
+	 &= E[ E(Y \mid X) \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
+	 &= E[ \rho X \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
+	 & =  \rho E[X \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
+	  & =  \rho E[X \mid \rho b_i \leq X \leq \rho c_i] \\
+	  &\leq \rho a_i
+\end{align*}
+
+where the equality holds if $b_i = -c_i$ or $\rho = 1$ on the last step. This is because
+
+\begin{align*}
+	 E[X \mid \rho b_i \leq X \leq \rho c_i] &= E[Y \mid \rho b_i \leq Y \leq \rho c_i] \\
+	\mathtt{0 \leq b_i < c_i}, \hspace{0.05in} \mathtt{0 \leq \rho \leq 1} \hspace{0.1in} & \leq E[Y \mid b_i \leq Y \leq c_i] \\
+	&= a_i
+\end{align*}
+
+The first equation above is because the same marginal distribution of $X$ and $Y$.
+
+Note that Gabriel CV method predicts $Y$ to be $a_i$ for $X \in [\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}]$, and if $k=1$ the predicted $Y = 0$ for all $X$.
+Hence, to see if $CV(k) > CV(1)$ for  $X \in [\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}]$, one only need to see if $ a_i-\hat{\mu}^Y_i > \hat{\mu}^Y_i - 0$.
+Since $ \hat{\mu}^Y_i \leq \rho a_i$, it's clear that $CV(k) > CV(1)$ whenever $\rho < 0.5$.
+This is true to all segment $[\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] = [\rho b_i, \rho c_i]$ except the one with $b_i = -c_i$, in which $CV(k) = CV(1)$.
+Since $k>1$ and at most one segment has $CV(k) = CV(1)$ while the rest have $CV(k) > CV(1)$ if $\rho < 0.5$, it is clear that the overall prediction error $CV(k) > CV(1)$.
+By symmetry, we can see $CV(k) > CV(1)$ if $|\rho| < 0.5$.
+The proof is complete.
+\end{proof}
 
@@
 
-	%\begin{figure}[H]
-	%\centering
-	%\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting2/Facet.pdf}
-	%\label{fig:setting2}
-	%\end{figure}
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting2/Facet.pdf}
+	\label{fig:setting2}
+	\end{figure}
 

### 2016-01-28 New simulation/technometrics.tex
Update real data example--high dimension data

@@
 \section{Real data application}
-We also applied our proposed method to three real world data sets obtained
+We also applied our proposed method to three real world data sets, two were obtained
 from the University of California Irvine machine learning repository. The
@@
 
-The third data set is the Sonar data returned from two targets -- a metal
-cylinder and a rock with similar shape, which is first studied by
-\cite{gorman1988analysis} using a neural network. Both targets were impinged
-by pulse which was a wide-band linear FM chirp $(ka = 55.6)$. Returns were
-collected at a range of $10$ meters and obtained from the cylinder at aspect
-angles spanning $90^{\circ}$ and from the rock at aspect angles spanning
-$180^{\circ}$. The data set contains $208$ returns ($111$ cylinder returns and
-$97$ rock returns), with each composed of $60$ spectral samples, normalized to
-take on values between $0$ and $1$. So the data has $60$ features with clearly
-$2$ clusters.
+The third data set is gene expression data of $5$ types of brain tumours from \cite{de2008clustering}, which contains $42$ observations including $10$ medulloblastomas, $10$ malignant gliomas, $10$ atypical teratoid/rhabdoid tumours (AT/RTs), $8$ primitive neuroectodermal tumours (PNETs) and $4$ normal cerebella.
+It was originally used by \cite{pomeroy2002prediction} to show that medulloblastomas are molecularly distinct from other types of brain tumours.
+There were $7129$ genes originally, but after feature selection, $1379$ genes left.
+So the data set is $42 \times 1379$, which serves as a high dimensional example.
 
@@
     \hline                  
- & Congress Voting && Breast Cancer && Sonar \\ \hline                    
+ & Congress Voting && Breast Cancer && Brain Tumours \\ \hline                    
 CH-index & $2$ && $2$  && $2$    \\
-Hartigan & $3$ && $3$  &&  $3$   \\
-Jump & $9$ && $9$ && $10$  \\   
+Hartigan & $3$ && $3$  &&  $4$   \\
+Jump & $10$ && $9$ && $1$  \\   
 Prediction strength & $2$ &&  $2$ && $1$    \\
-Bootstrap stability & $2$ &&  $2$ &&  $10$   \\ 
-Gap & $10$ && $9$ && $10$  \\   
-Gaussian-Mix & $7$ && $5$ && $1$  \\   
-Gabriel & $2$  && $3$  && $2$  \\    
-Wold & $2$ && $3$ && $10$ \\  \hline 
+Bootstrap stability & $2$ &&  $2$ &&  $7$   \\ 
+Gap & $8$ && $10$ && $10$  \\   
+Gaussian-Mix & $2$ && $5$ && $2$  \\   
+Gabriel & $2$  && $3$  && $5$  \\    
+Wold & $2$ && $3$ && $4$ \\  \hline
+
 \end{tabular}
@@
 can be further clustered into at least two subgroups. Hence, the result $k=3$
-given by our proposed Gabriel method (as well as Wold method) make sense. For
-the Sonar data set which is relatively high-dimension, only our proposed
-Gabriel method and CH-index selects $k=2$. Majority of the methods collapse to
-either select the maximum or select the minimum number, underline how
-difficult it was to pick the right $k$ when dimension increases.
+given by our proposed Gabriel method make sense. For
+the brain tumours gene expression data set which is high-dimension, it is also clear that 
+the correct number of clusters should be $5$, each cluster corresponding to one type of 
+brain tumour. From the table we can see only our proposed
+Gabriel method correctly selects $k=5$, underline how
+difficult it was to pick the right $k$ when dimension is high.
 

### 2016-01-28 New simulation/wei/Data Example.R

@@ -1,4 +1,5 @@
 
+
    library("bcv")
    library("devtools")
    library("cluster")
@@ -15,6 +16,8 @@
    source("../code/jump.R")
    source("../code/wold.R")
 
+
+######------------Voting data-----------------------############
 Vote <- read.table("Voting.txt",sep=",")
 
 Id <- vector()
@@ -41,42 +44,42 @@ for(i in 1:232){
 
 set.seed(3)
 ID <- sample(1:232, 232)
-
 Vote <- VV[ID,]
-set.seed(0)
+
+set.seed(1)
 cv.kmeans.gabriel(Vote, 5, 2, maxcenters=10, classify.method="nearest")$centers
 
-set.seed(0)		
+set.seed(1)		
 gabriel_cor_correct(Vote, maxcenters=10, type = 2)
 
-set.seed(0)		
+set.seed(1)		
 Wold_holdout(data = Vote, CV = 5, Errortol = 0.01,  max.k =10)
 
-set.seed(0)		
+set.seed(1)		
 Gap <- clusGap(Vote, FUN = kmeans, K.max =10)
 which(Gap[[1]][,3] == max(Gap[[1]][,3]))
 			
-set.seed(0)	
+set.seed(1)	
 mcluster <- Mclust(Vote, G = 1:10)
  mcluster$G
 						
-set.seed(0)
+set.seed(1)
 Ch <- NbClust(Vote, min.nc = 2, max.nc = 10,method = "kmeans", index = "ch")
 Ch$Best.nc[1]
 	
-set.seed(0)
+set.seed(1)
 Hartigan <- NbClust(Vote, min.nc = 2, max.nc = 10,method = "kmeans", index = "hartigan")
 Hartigan$Best.nc[1]
 			
-set.seed(0)
-Jump <- jump(Vote,plotjumps=F,rand=10,trace=F)
+set.seed(1)
+Jump <- jump(Vote,10, plotjumps=FALSE, trace=FALSE)
 Jump$maxjump
 	
-set.seed(0)
+set.seed(1)
 PS <- prediction.strength(Vote, Gmin=2, Gmax=10)
  PS$optimalk
 			
-set.seed(0)
+set.seed(1)
 SB <- nselectboot(Vote,clustermethod=kmeansCBI,classification="centroid",krange=2:10)
 SB$kopt
 
@@ -105,95 +108,108 @@ set.seed(3)
[…]

### 2016-02-29 demo/rotate/01_run_sim.R
add rotate demo

@@ -1,0 +1,58 @@
+#!/usr/bin/Rscript --vanilla
+
+library("RColorBrewer")
+
+
+rot_sim <- function(corr, nrep=1000)
+{
+    # start with a covariance matrix with high correlation
+    dim <- 5
+    #cov <- matrix(c(1, corr, corr, 1), 2, 2)
+    cov <- matrix(1, dim, dim)
+    cov[1,2] <- corr
+    cov[2,1] <- corr
+
+    corr_rot <- rep(NA, nrep)
+
+    for (r in seq_len(nrep)) {
+        # generate a random rotation
+        z <- matrix(rnorm(dim * dim), dim, dim)
+        qr <- qr(z)
+        q <- qr.Q(qr)
+        sign <- sample(c(-1, 1), dim, replace=TRUE)
+        rot <- q %*% diag(sign, dim)
+
+        # compute the covariance matrix for the rotated data
+        cov_rot <- t(rot) %*% cov %*% rot
+        sd_rot <- sqrt(diag(cov_rot))
+
+        # compute the correlation for the rotated data
+        corr_rot[r] <- cov_rot[1,2] / (sd_rot[1] * sd_rot[2])
+    }
+
+    corr_rot
+}
+
+
+set.seed(0)
+
+nrep <- 1000
+corr <- seq(0, 1, len=100)
+corr_rot_mean <- rep(NA, length(corr))
+corr_rot_sd <- rep(NA, length(corr))
+
+for (i in seq_along(corr)) {
+    sim <- rot_sim(corr[i], nrep)
+    corr_rot_mean[i] <- mean(abs(sim))
+    corr_rot_sd[i] <- sd(abs(sim))
+}
+
+palette(brewer.pal(6, "Set1"))
+plot(corr, corr_rot_mean, col=2)
+segments(corr, corr_rot_mean - corr_rot_sd, corr, corr_rot_mean + corr_rot_sd,
+         col=2)
+abline(h=0.5, col=1)
+
+i <- max(which(corr_rot_mean < 0.5))
+abline(v=0.5 * (corr[i] + corr[i+1]), col=1, lty=2)
+

### 2016-02-29 draft2/technometrics.tex
add draft2

--- 2016-01-28 New simulation/technometrics.tex
@@
 Gap & $8$ && $10$ && $10$  \\   
-Gaussian-Mix & $2$ && $5$ && $2$  \\   
+Gaussian-BIC & $2$ && $5$ && $2$  \\   
 Gabriel & $2$  && $3$  && $5$  \\    
+Gabriel-corr-correct & $2$  && $2$  && $5$  \\  
 Wold & $2$ && $3$ && $4$ \\  \hline
@@
 
-Since most congressmen vote base on their parties' interest, $2$ parties
+Since most congressmen vote based on their parties' interest, two parties
 (Democratic and Republican) represent two clusters in this data set. So the
@@
 $k=2$ separates the two parties very well with the lowest miss-classification
-error $10.43\%$ ($k=3$ has $14.78\%$). Note that CH-index and Bootstrap
+error $10.43\%$ ($k=3$ has $14.78\%$). Our proposed Gabriel CV method, along with 
+several other algorithms correctly pick $k=2$ for this data set. Note that CH-index and Bootstrap
 stability also return $k=2$, but $2$ is the lower bound those methods can
 select for $k$. So it's not clear they actually choose $k=2$ or they hit the
-lower bound (they would pick $k=1$ if allowed). For the breast cancer data,
-it's known to have at least $2$ cluster based on whether it's benign specimens
-or not. But it doesn't mean the optimal $k$ should be $2$.
-\cite{fujita2014non} noticed that the malign group is quite heterogeneous and
-can be further clustered into at least two subgroups. Hence, the result $k=3$
-given by our proposed Gabriel method make sense. For
-the brain tumours gene expression data set which is high-dimension, it is also clear that 
+lower bound (they would pick $k=1$ if allowed).
+
+For the breast cancer data, it's known to have $2$ clusters based on 
+whether it's benign specimens or malignant specimens. Although the proposed Gabriel method
+picks $k=3$, its correlation-corrected version correctly picks $k=2$. It shows the necessity 
+to correct for the high correlation among data features. However, for this data $k=3$ is also
+an acceptable answer, as \cite{fujita2014non} noticed that the malign group is quite heterogeneous and
+can be further clustered into at least two subgroups.
+
+The brain tumours gene expression data is a typical high-dimension data with $p > n$. It is clear that 
 the correct number of clusters should be $5$, each cluster corresponding to one type of 
-brain tumour. From the table we can see only our proposed
-Gabriel method correctly selects $k=5$, underline how
-difficult it was to pick the right $k$ when dimension is high.
+brain tumour.
+From table \ref{table2} we can see only our proposed Gabriel method (as well as its correlation-corrected version) correctly selects $k=5$, underline how difficult it was to pick the right $k$ when dimension is high.
 

### 2016-04-08 New simulation/technometrics.tex
Brain tumor data: add preprocessing detail and discussion

@@
 It was originally used by \cite{pomeroy2002prediction} to show that medulloblastomas are molecularly distinct from other types of brain tumours.
-There were $7129$ genes originally, but after feature selection, $1379$ genes left.
+
+Data was collected with Affymetrix microarrays, which estimates the number of RNA copies found in the cell sample (frozen specimens).
+Data was preprocessed as following: as a common procedure, all genes with expression level below $10$ are set to a minimum level threshold of $10$.
+The maximum threshold is set at $16,000$.
+Values below or above these thresholds are often unreliable.
+To remove uninformative genes, for each gene $j$ (attribute), the mean $m_j$ among the samples was calculated.
+The $10\%$ largest and smallest values are discarded in order to remove extreme values.
+Based on the mean, following transformation is applied on every value $x_{ij}$ of sample $i$ and gene j
+\[	y_{ij} = \log_2 (x_{ij}/m_j)	\] 
+Genes with expression levels differing by at least $l$-flod in at least $c$ samples from their mean expression level across the sample were selected.
+The parameter $l$ and $c$ were chosen so as to produce a filtered data set with at least $10\%$ of the original number of genes \citep{de2008clustering}.
+Such preprocessing procedure results in $1379$ genes selected.
+After filtering step, data is restored in the original scale (i.e.
+$x_{ij}$).
 So the data set is $42 \times 1379$, which serves as a high dimensional example.
@@
 $k=2$ separates the two parties very well with the lowest miss-classification
-error $10.43\%$ ($k=3$ has $14.78\%$). Note that CH-index and Bootstrap
+error $10.43\%$ ($k=3$ has $14.78\%$). Our proposed Gabriel CV method, along with 
+several other algorithms correctly pick $k=2$ for this data set. Note that CH-index and Bootstrap
 stability also return $k=2$, but $2$ is the lower bound those methods can
 select for $k$. So it's not clear they actually choose $k=2$ or they hit the
-lower bound (they would pick $k=1$ if allowed). For the breast cancer data,
-it's known to have at least $2$ cluster based on whether it's benign specimens
-or not. But it doesn't mean the optimal $k$ should be $2$.
-\cite{fujita2014non} noticed that the malign group is quite heterogeneous and
-can be further clustered into at least two subgroups. Hence, the result $k=3$
-given by our proposed Gabriel method make sense. For
-the brain tumours gene expression data set which is high-dimension, it is also clear that 
-the correct number of clusters should be $5$, each cluster corresponding to one type of 
-brain tumour. From the table we can see only our proposed
-Gabriel method correctly selects $k=5$, underline how
-difficult it was to pick the right $k$ when dimension is high.
+lower bound (they would pick $k=1$ if allowed).
+
+For the breast cancer data, it's known to have $2$ clusters based on 
+whether it's benign specimens or malignant specimens. Although the proposed Gabriel method
+picks $k=3$, its correlation-corrected version correctly picks $k=2$. It shows the necessity 
+to correct for the high correlation among data features. However, for this data $k=3$ is also
+an acceptable answer, as \cite{fujita2014non} noticed that the malign group is quite heterogeneous and
+can be further clustered into at least two subgroups.
+
+The brain tumours gene expression data is a typical high-dimension data with $p > n$. It is clear that 
+the correct number of clusters should be $5$, the number of brain tumour types.
+Indeed, $k$-means results with $k=5$ show that the $4$ normal human cerebella are well separated from brain tumours as they form a single cluster.
+Medulloblastomas, malignant gliomas and AT/RT tumours also form $3$ respective clusters, which means they can be separated from each other.
+Such results are similar to those found in \cite{pomeroy2002prediction}, whose analysis
+is based on the first $3$ principle components of the original data.    
+From table \ref{table2} we can see only our proposed Gabriel method (as well as its correlation-corrected version) correctly selects $k=5$, underline how difficult it was to pick the right $k$ when dimension is high.
 

### 2016-04-18 cvclust.tex
revise introduction; start revising theory

--- 2015-08-27 cvclust.tex
@@
-\documentclass[11pt]{article}
-\usepackage{epsf}
-\usepackage{hyperref}
-\usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage[margin=2.5cm]{geometry}
+\documentclass[12pt]{article}
 \usepackage{amsmath}
-\usepackage{footnote}
-\usepackage[bottom]{footmisc}
+\usepackage{amssymb}
+\usepackage{amsthm}
+\usepackage{enumerate}
+\usepackage{enumitem}
+\usepackage{float}
+%\usepackage{footnote}
+%\usepackage[bottom]{footmisc}
 \usepackage[singlelinecheck=false]{caption}
-\usepackage{caption}
-\usepackage{amsthm}
-\usepackage{enumitem}
-\usepackage{graphics}
-\usepackage{float}
-\usepackage{amssymb}
+
+\usepackage{graphicx,psfrag,epsf}
+\usepackage[space]{grffile}
 \usepackage{natbib}
-\usepackage{lscape}
-\usepackage{array}
-\textheight=8.7in
-\usepackage{setspace}
-\usepackage{verbatim}
-\usepackage{multirow}
-\usepackage{booktabs}
-\usepackage{longtable}
-
-\def\threedigits#1{%
-  \ifnum#1<100 0\fi
-  \ifnum#1<10 0\fi
-  \number#1}
-
-\topmargin=0.1in \oddsidemargin=-0.1cm \evensidemargin=-0.1cm
-
-\paperheight=11in \paperwidth=8.5in \marginparwidth=0in
-
-\marginparsep=0in \textwidth=6.5in \headheight=0in \headsep=0in
-
-\onehalfspacing
-\def\argmax{\mathop{\rm arg\,max}}
-%\usepackage{xspace,epsfig,subfig}
+\usepackage{url} % not crucial - just used below for the URL
+
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{1}
+
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{-.3in}%
+\addtolength{\topmargin}{-.8in}%
+
 \newtheorem{theorem}{Theorem}
@@
 \newtheorem{corollary}{Corollary}
-\newenvironment{sketch}{\noindent\emph{Proof Sketch:}}{$\quad \Box$}
-%\newenvironment{proof}{\noindent\emph{Proof:}}{$\quad \Box$}
-
-\bibpunct{(}{)}{;}{a}{,}{,}
 
@@
 \begin{document}
-\title{Estimating the number of clusters using Cross Validation}
-\author{Wei Fu \qquad Patrick O. Perry \\\\ New York University}
-\date{}
+
+%\bibliographystyle{natbib}
+
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\if1\blind
+{
+  \title{\bf Estimating the number of clusters using cross-validation}
+  \author{Wei Fu and Patrick O. Perry \\
+  Stern School of Business, New York University}
   \maketitle
+} \fi
+
+\if0\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf Estimating the number of clusters using cross-validation}
+\end{center}
+  \medskip
+} \fi
+
+\bigskip
 \begin{abstract}
@@
 
+\noindent%
+{\it Keywords:} clustering, unsupervised learning, model selection
+\vfill
+
+\newpage
+\spacingset{1.45} % DON'T change the spacing!
 \section{Introduction}
-
-As a main task of exploratory data analysis, clustering organizes unlabeled
-observations into groups such that observations in same group are more similar
-compare to those in different group. Clustering is an important topic in
-unsupervised learning because it can reveal the internal structure of data
-through grouping, segment the data through partitioning and summarize data for
-other purposes such as dimension reduction. It has being widely used in
-various fields such as psychology, biology, statistics and machine learning
-including pattern recognition, image segmentation etc.
-
-After being proposed more than $50$ years, $k$-means remains one of the most
-popular and widely used clustering algorithms \citep{jain2010data}. Like many
-other clustering methods, $k$-means requires an input parameter $k$, the
-number of clusters, to be specified by the user. Automatically and
-quantitatively deciding such parameter is important and yet unsolved problem
-\citep{fujita2014non}. Various methods have been proposed to tackle this
-difficulty. One ad hoc approach is to explore the relationship between $W_k$
-(within-cluster dispersion) and the number of cluster $k$ for a certain
-clustering method such as $k$-means. Since $W_k$ decreases as $k$ increases,
-one usually find the ``elbow" of curve obtain by plotting $W_k$ versus $k$ as
-the appropriate number of clusters. The example on the top row of
-Figure~\ref{fig:elbow} demonstrates such approach for data with $k=4$, where
-the ``elbow" point indeed reveals the true number of clusters. This is based
-on the idea that under partitioning data set has more impact than over
-partitioning data set in terms of $W_k$. However, locating the ``elbow" point
-is somewhat subjective and sometimes is not appropriate to select the optimal
-$k$. The second example on the bottom row of Figure \ref{fig:elbow} shows a
-situation where there is no clear choice of the ``elbow" point -- both $k=2$
-and $k=3$ can be viewed as the ``elbow" point. What's more, the true $k=4$ can
-never be selected as the optimal $k$ using such approach in this case since it
-can hardly be viewed as the ``elbow" of the curve.
-
-\begin{figure}
-\centering
-\begin{minipage}{\linewidth}
-  \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{demo/elbow/correct-data.pdf}
-  \end{minipage}
-  \hspace{0.05in}
-  \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{demo/elbow/correct-withinss.pdf}
-  \end{minipage}
-\end{minipage}
-\begin{minipage}{\linewidth}
-  \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{demo/elbow/incorrect-data.pdf}
-  \end{minipage}
-  \hspace{0.05in}
-  \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{demo/elbow/incorrect-withinss.pdf}
-  \end{minipage}
-\end{minipage}
-\caption{Left panels show the $(X,Y)$ data points; right panels
-  show the corresponding values of the within-cluster sum of squares $W_k$
-plotted against the number of clusters, $k$.}
-\label{fig:elbow}
-\end{figure}
-
-Recently, there are several new proposals to find the $k$ automatically. Gap
-statistics \citep{tibshirani2001estimating} estimates $k$ by comparing the
-change in within-cluster dispersion with that expected under an appropriate
-reference null distribution. Specifically, the graph of $\log(W_k)$ is
-compared with its expectation under an appropriate null reference distribution
-of the data. The value of $k$ associated with the largest gap between
-$\log(W_k)$ and the reference curve is selected as optimal $k$.
-\citet{sugar2003finding} proposed an approach which finds the number of
-clusters based on distortion, a quantity that measures the average distance,
-per dimension. It's backed by a rigorous theoretical justification based on
-information-theoretic ideas.   \citet{fraley2002model}'s Model-based method
-employs the EM algorithm to estimate the parameters in Gaussian mixture model,
-and select the best model ($k$) using BIC criterion. Stability-based criterion
-is also proposed to locate the best $k$ by some authors such as
-\citet{ben2001stability}, \citet{wang2010consistent} and
-\citet{fang2012selection}. \citet{chiang2010intelligent} provides a nice review
-of existing methods for finding the right $k$ in published literature.
-
-Most existing methods are either model based method requires strong parameter
-assumptions or otherwise lack of clear interpretation and theoretical justification.
-Although many view selecting the number of clusters as a model selection problem, very few
-approaches this problem from the prediction point of view. Select model with
-smallest prediction error via cross-validation is one of the simplest and most
-widely used model selection techniques in supervised learning. The lack of
-true class (label) in data set makes the adoption of cross-validation into
-unsupervised leaning problem difficult. One exception is
-\cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction
-strength. The strategy is to first cluster the test data and training data
-into $k$ clusters respectively. Then, for each pair of observations that are
-assigned to the same test cluster, algorithm determines whether they are also
-assigned to the same cluster based on the training centers. The intuition here
-is, if $k=k_0$, the true number of clusters, then the $k$ training set
-clusters will be similar to the $k$ test clusters, and hence will predict them
-well. However, a specifically defined prediction error measure is used in such
-procedure, which is quite different from the one commonly used in
-cross-validation procedure in supervised learning. Therefore the well
-understood properties of cross-validation procedure in supervised learning
-cannot be carried over to the prediction strength method, makes the
-interpretation of its result difficult. \citet{wang2010consistent} also uses 
-cross-validation to select the optimal $k$. Instead of selecting $k$ which minimizes 
-prediction error, such method picks $k$ which minimizes the specifically defined clustering instability.
-
-Our proposed method is a complete data-driven approach which doesn't rely on
-any parametric assumptions. Through novel form of partitioning data set, we
-are able to employing the cross-validation procedure in clustering exactly the
-same way as in supervised learning problem. Hence, it's easy for reader to
-understand the intuition behind our proposed method. Simulation and real data
-application shows the superior performance of our proposed method compare with
-existing methods in high-dimension settings and heavy-tailed data. Since the
-embedded cross-validation procedure is well understood, it also makes our
-method potentially easily to be extended in future study.
+\label{sec:intro}
+
+A clustering procedure segments a collection of items into smaller groups,
+with the property that items in the same group are more similar to each other
+than items in different groups \citep{hartigan1975clustering}.  These
+procedures are used in two main applicaitons: (a) exploratory analysis, where clusters
+reveal homogeneous sub-groups within a large sample \cite{TODO}; (b) data
+reduction, where high-dimensional item attribute vectors get reduced discrete
+cluster labels \citep{jain1999data}.
+
+With many clustering methods, including the popular $k$-means clustering
+procedure, the user must specify $k$, the number of clusters
+\citep{jain2010data}.  One popular ad-hoc device for selecting the number of
+clusters is to use an analogue of the principal components scree plot: plot
+the within-cluster dispersion $W_k$, as a function of the number of clusters
+$k$, looking for an ``elbow'' in the plot.  This approach is simple and often
+performs well, but it requires subjective judgment as to where the elbow is
+located, and as we demonstrate in Appendix~\ref{sec:elbow-fail}, the approach
+can easily fail.  In this report, we propose a new method to choose the number
+of clusters automatically.
+
+The problem of choosing $k$ has been well-studied, and dozens of methods have
+been proposed~\citep{chiang2010intelligent,fujita2014non}. The main difficulty
+in choosing $k$ is that clustering is fundamentally an ``unsupervised''
+learning problem, meaning that there is no obvious way to use ``prediction
+ability'' to drive the model selection \citep{hastie2009elements}.  Most
+existing methods for choosing $k$ instead rely on on explicit or implicit
+assumptions about the data distribution, including it shape, scale, and
+correlation structure.
+
+Several authors advocate choosing $k$ by performing a sequence of hypothesis
+tests will null and alternative hypotheses of the form $H_0 : k = k_0$ and
+$H_1: k > k_0$, starting with $k_0 = 1$ and proceeding sequentially with
+higher values of $k_0$ until a test fails to reject $H_0$. The gap statistic
+method typifies this class of methods, with a test statistic that measures the
+within-cluster dispersion relative to what is expected under a reference
+distribution \citep{tibshirani2001estimating}. Similar methods include those
+developed by \citet{TODO}.
+
+Other authors have proposed choosing $k$ by using information
+criteria.  \citet{sugar2003finding} proposed an approach that
+minimizes the estimated ``distortion'', the average distance per dimension.
+\citet{fraley2002model}'s model-based method fits Gaussian mixture model
+models to the data, then selects the number of mixture components, $k$, using
+the Bayesian Information Criterion (BIC).
+
+A third set of approaches is based on the idea of ``stability'', that clusters
+are meaningful if they manifest in multiple independent samples from the same
+population. \citet{ben2001stability}, \citet{tibshirani2005cluster},
+\citet{wang2010consistent} and \citet{fang2012selection} developed methods
+based on this idea.
+
+The procedure we propose in this report is based on a form of
+cross-validation, and it is adaptive to the characteristics of the data
+distribution. The essential idea is to devise a way to measure a form 
+of internal prediction error associated each choose of $k$, and then choose
+the $k$ with the smallest associated error.
+
+In this report, we extend the Wold and Gabriel methods to the clustering
+problem, specifically to choose an appropriate number of clusters for a
+dataset.  We prove that the Gabriel method is self-consistent, and we analyze
+some of its properties in the presence of noise.  We compare these methods to
+state-of-the-art algorithms, and show that both are competitive.
+
+\textit{TODO: Add outline}
 
 \section{Cross-validation for selecting the number of clusters}
-
+\label{sec:meth}
 Cross-validation is commonly used for model selection in supervised learning
@@
 
-In this report, we extend the Wold and Gabriel methods to the clustering
-problem, specifically to choose an appropriate number of clusters for a
-dataset.  We prove that the Gabriel method is self-consistent, and we analyze
-some of its properties in the presence of noise.  We compare these methods to
-state-of-the-art algorithms, and show that both are competitive.
-
 %% [POP] This paragraph doesn't fit into the section.  Maybe put it in the introduction?
@@
 
-\section{Analysis of Gabriel Cross-Validation with Gaussian Noise}
-
-\subsection{Single cluster in two dimensions}
+\section{Analysis of Gabriel cross-validation with Gaussian noise}
+
+\subsection{Single Cluster in two dimensions}
 
@@
 the case of Gaussian noise. Our main result is that with single-cluster
-Gaussian data, if the predictor and response columns of $\dataX$ are weakly
-correlated or independent, then the method will correctly prefer $k = 1$ to
-$k = 2$ clusters. We state the result in the case where $\dataX$ has 2 columns,
-but it is straightforward to generalize this result to higher dimensions.
-
-\begin{proposition}\label{prop:1clust-2dim}
+Gaussian data in two-dimensions, if the predictor and response columns of $\dataX$ are weakly
+correlated or independent, then the method will correctly prefer $k = 1$
+clusters. In section~\ref{} we will generalize this result to 
+cases where $\dataX$ has more than two columns.
+
+\begin{theorem}
 Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
@@
 from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 0.5$,
-then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
-\end{proposition}
+then $\CV(1) < \CV(k)$, $k > 1$ with probability tending to one as $m$ and $n$ increase.
+\end{theorem}
 
 \begin{proof}
-
-Given $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to $\{ Y_i
-\}_{i=1}^{n}$. With $k = 1$, the single-cluster centroid will
-be equal to $\bar Y_n$, the sample mean of the $Y_1, \dotsc, Y_n$, approximately equal to
-$\E(Y) = 0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error
-will be
-\[
-  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
-         = 1 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
-\]
-
-Now we will consider the $k = 2$ case.  If $n$ is large enough, then
-\citet{pollard1981strong} showed that the centroids $\bmuY_1$ and $\bmuY_2$
-will be close to $\E(Y \mid Y > 0) = \sqrt{2/\pi}$ and $\E(Y \mid Y < 0) =
--\sqrt{2/\pi}$.  We have used Lemma~\ref{lem:truncated-normal-moments}
-(Appendix~\ref{app:technical-lemmas}) to compute the expectations.  Further,
-\citet{pollard1982central} showed that the errors will be of size $\OhP(n^{-1/2})$.
-
-If $\rho > 0$ and $n$ is large enough, then classification rule learned from
-$\{(X_i, \hGY_i)\}_{i=1}^{n}$ variables will be determined according to
-whether $X > 0$; if $\rho < 0$ then the decision is according to whether $X <
-0$.  More specifically, the decision boundary will be at $0 + \OhP(n^{-1/2})$.
-
-In the $\rho > 0$ case, the cross-validation error will be
+Given $(X_1, Y_1), \dotsc, (X_n, Y_n)$, let $A_k = \{a_1,a_2,...,a_k\}$
+denotes the set of cluster centers from applying $k$-means to $\{
+Y_i\}_{i=1}^{n}$ with parameter $k$. Because $\{ Y_i\}_{i=1}^{n}$ is
+symmetric, it's easy to see that the cluster centers are symmetric around $0$,
+i.e. if $a^* \in A_k$ then $-a^* \in A_k$. The $k$ clusters are symmetric in
+the same sense. Also, if $k$ is odd, then only one cluster contains both
+negative and positive $Y$ and it's symmetric with center $0$. The rest have
+pure positive or negative intervals; if $k$ is even, then all clusters have
+pure positive or negative intervals.
+
+Let $a_i$ denotes a cluster center with interval $[b_i, c_i]$. By above
+argument, either $0 \leq b_i < c_i$ ($b_i < c_i \leq 0$) or $b_i = -c_i$ which
+corresponding to $a_i=0$ with $k$ be odd. By symmetry, it's sufficient to only
+consider case $0 \leq b_i < c_i$ (with $a_i > 0 $) and $b_i = -c_i$ (with $a_i
+= 0 $) with $\rho \geq 0$.
+
+From joint normality of $X$ and $Y$, it follows that $X \mid Y$ is normal with
+mean $\rho Y$. Therefore, the mean value $\bar{X}_i$ of $X$ in cluster with
+$Y$ center $a_i$ is
+
 \begin{align*}
-  \CV(2)
-  &=
-    \frac{1}{m}
-    \sum_{i=n+1}^{n+m}
-      \| (Y_i - \bmuY_1) 1\{\hGX_i = 1\} \|^2
-      +
-      \| (Y_i - \bmuY_2) 1\{\hGX_i = 2\} \|^2
-\\
-  &=
-    \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]
-      + \OhP(m^{-1/2}) + \OhP(n^{-1/2}),
+	\bar{X}_i &= E[X \mid b_i \leq Y \leq c_i] \\
+			&= E[ E(X \mid Y) \mid b_i \leq Y \leq c_i] \\
+			&= E[\rho Y \mid b_i \leq Y \leq c_i] \\
+			&= \rho E[ Y \mid b_i \leq Y \leq c_i] \\
+			&= \rho a_i
 \end{align*}
-where $a = \sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows
-that $Y \mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$, so
-that $\E[(Y - a)^2 \mid X] = (\rho X - a)^2 + (1 - \rho^2)$.  Applying
-Lemma~\ref{lem:truncated-normal-moments}, we get that for large $m$ and $n$,
-the Gabriel cross-validation error is close to
-$1 + a^2 (1 - 2 \rho)$.
-
-In the $\rho < 0$ case, a similar calculation shows that $\CV(2)$ is close to
-$1 + a^2 (1 + 2 \rho)$.  In particular, if $|\rho| < 0.5$, then with
-probability tending to $1$ and $m$ and $n$ increase, the asymptotic
-cross-validation error for $k = 1$ will be smaller than for $k = 2$.
+
+That is, the mean values of $X$ in the $k$ clusters is $\{\rho a_1,\rho
+a_2,...,\rho a_k\}$. Because we assign each observation to the closest
+$\bar{X}_i$, $i=1,2,...,k$, it's easy to see the corresponding interval on $X$
+for $\bar{X}_i = \rho a_i$ is $[\frac{\rho \left( a_{i-1}+a_i\right)}{2},
+\frac{\rho \left( a_{i}+a_{i+1}\right)}{2}]$, where $a_{i-1}$ and $a_{i+1}$
+are the two adjacent centers to $a_i$ on $Y$. Note that $[\frac{ \left(
+a_{i-1}+a_i\right)}{2}, \frac{\left( a_{i}+a_{i+1}\right)}{2}] = [b_i, c_i]$
+by the algorithm definition of $k$-means. The $Y$ center for such interval on
+$X$ is
+
+\begin{align*}
+	\hat{\mu}^Y_i &= E[Y \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
+	 &= E[ E(Y \mid X) \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
+	 &= E[ \rho X \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
+	 & =  \rho E[X \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
+	  & =  \rho E[X \mid \rho b_i \leq X \leq \rho c_i] \\
+	  &\leq \rho a_i
+\end{align*}
+
+where the equality holds if $b_i = -c_i$ or $\rho = 1$ on the last step. This is because
+
+\begin{align*}
+	 E[X \mid \rho b_i \leq X \leq \rho c_i] &= E[Y \mid \rho b_i \leq Y \leq \rho c_i] \\
+	\mathtt{0 \leq b_i < c_i}, \hspace{0.05in} \mathtt{0 \leq \rho \leq 1} \hspace{0.1in} & \leq E[Y \mid b_i \leq Y \leq c_i] \\
+	&= a_i
+\end{align*}
+
+The first equation above is because the same marginal distribution of $X$ and $Y$.
+
+Note that Gabriel CV method predicts $Y$ to be $a_i$ for $X \in [\frac{\rho
+\left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}]$,
+and if $k=1$ the predicted $Y = 0$ for all $X$. Hence, to see if $CV(k) >
+CV(1)$ for  $X \in [\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho
+\left( a_{i}+a_{i+1}\right)}{2}]$, one only need to see if $ a_i-\hat{\mu}^Y_i
+> \hat{\mu}^Y_i - 0$. Since $ \hat{\mu}^Y_i \leq \rho a_i$, it's clear that
+$CV(k) > CV(1)$ whenever $\rho < 0.5$. This is true to all segment
+$[\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left(
+a_{i}+a_{i+1}\right)}{2}] = [\rho b_i, \rho c_i]$ except the one with $b_i =
+-c_i$, in which $CV(k) = CV(1)$. Since $k>1$ and at most one segment has
+$CV(k) = CV(1)$ while the rest have $CV(k) > CV(1)$ if $\rho < 0.5$, it is
+clear that the overall prediction error $CV(k) > CV(1)$. By symmetry, we can
+see $CV(k) > CV(1)$ if $|\rho| < 0.5$. The proof is complete.
 \end{proof}
@@
 
-\subsection{Single cluster in higher dimensions}
-
-To generalize the result of the previous section to higher dimensions, 
-suppose that $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ is data from a single fold of
-Gabriel Cross-Validation, with each $(X, Y)$ pair in $\R^p \times \R^q$ a
-draw a mean-zero multivariate normal distribution. To analyze this setting, it
-suffices to consider the mean-zero case with $\cov(X) = I_p$.
-
-Suppose that $\cov(Y) = U \Lambda U^\T$ is the eigendecomposition of
-$\cov(Y)$. For simplicity, suppose that the leading eigenvalue $\lambda_1$
-has unit multiplicity, so that the first principal subspace of $Y$ is a line,
-given by the span of the vector $u_1 \in \R^q$. When we apply $k$-means to
-$Y_1, \dotsc, Y_n$, the centroids will be close to
-$\bmuY_1 =  \sqrt{2\lambda_1 /\pi} u_1$ and
-$\bmuY_2 = -\sqrt{2\lambda_1 /\pi} u_1$, with errors of size $\OhP(n^{-1/2})$.
-
-The means for the $X$ variables will be close to
-$\bmuX_1 = \E(X \mid u_1^\T Y > 0)$ and
-$\bmuX_2 = \E(X \mid u_1^\T Y < 0)$,
-with errors of size $\OhP(n^{-1/2})$.  Let
-\[
-  \begin{bmatrix}
-    I_p & \lambda_1^{1/2} \rho \\
-    \lambda_1^{1/2} \rho^T & \lambda_1
-  \end{bmatrix}
-\]
-be the covariance matrix of $(X, u_1^\T Y)$,
-where $\rho = (\rho_1, \dotsc, \rho_p)$ is such that $\rho_i = \corr(X_i,
-u_1^\T Y)$ for $i = 1, \dotsc, p$.
-. Then,
-\[
-  \bmuX_1 = \sqrt{2 \lambda_1 / \pi} \rho,
-  \qquad
-  \bmuX_2 = -\sqrt{2 \lambda_1 / \pi} \rho;
-\]
-thus, the classification rule learned from $\{ (X_i, \hGY_i) \}_{i=1}^{n}$
-will send observations to class $1$ or $2$ according to whether
-$\rho^\T X + \OhP(n^{-1/2})$ is positive or negative.
-
-We have reduced the analysis of cross-validation error to the two-dimensional
-case. In particular, following the calculations from the proof of
-Prop.~\ref{prop:1clust-2dim}, if $|\corr(\rho^\T X, u_1^\T Y)| < 0.5$, then with
-probability tending to one as $m$ and $n$ increase, $\CV(1) < \CV(2)$.
-The condition $|\corr(\rho^\T X, u_1^\T)| < 0.5$ reduces to
-\[
-  \sum_{i=1}^p \rho_i^2 < 0.25.
-\]
-A sufficient (but not necessary) condition to ensure that this holds is that
-the magnitude of the canonical correlation between the $X$ variables and
-the $Y$ variables is below $0.5 / \sqrt{p}$.
-
-\subsection{Two clusters}
+The reason why Gabriel CV method tends to select larger $k$ when correlation is high 
+between dimensions is that it resembles the naive method mentioned in section \ref{sec:intro} 
+under such circumstance. In fact, if $X$, $Y$ are perfectly correlated, Gabriel method is 
+equivalent to the naive method.
+
+\subsection{Single Cluster in higher dimensions}
+
+\textit{todo: add and revise old proof}
+
+\subsection{Two clusters in two dimensions}
 
@@
 \centering
-\includegraphics[width=25pc]{demo/overlap/color_plot.pdf}
+\includegraphics[width=3in]{demo/overlap/color_plot.pdf}
 \caption{Number of times $k=2$ is selected out of $10$ replicates for each pair of $(\muX, \muY)$.
-The heat map shows the frequency $k=2$ is selected by the algorithm, with blue means low number (of $k=2$) is selected and red color means high number is selected.
+The heat map shows the frequency $k=2$ is selected by the algorithm, with light means low number (of $k=2$) is selected and dark color means high number is selected.
 The black line is the theoretical curve, above which the algorithm suppose to pick $k=2$ and below which algorithm select $k=1$.
@@
 
-\section{Simulation}
+\section{Adjusting for correlation}
+
+\textit{TODO: add more discussion here}
+
+When the correlation between dimensions are high, the proposed Gabriel CV
+method tends to overestimate the number $k$. A simply remedy for the high
+correlation is available if we assume common covariance structure among the
+$k$ clusters.
+
+\begin{enumerate}
+	\item Apply Gabriel CV method on the original data $\dataX$, get estimated number of cluster $\hat{k}$
+	\item Estimate the pooled covariance matrix $\hat{\Sigma}$ from the $\hat{k}$ clusters.
+	\item Let $\hat{\Sigma} = \Gamma\Lambda\Gamma'$ be the Eigendecomposition of $\hat{\Sigma}$, we rescale and rotate the original data $\dataX$ to get $ \widetilde{\dataX} = \dataX\Gamma\Lambda^{-1/2}Q$, where $Q$ is a random  orthonormal rotation matrix.
+	\item Apply Gabriel CV method again on the transformed data $ \widetilde{\dataX}$ to estimate $k$. 
+\end{enumerate}
+
+\section{Performance in simulations}
 
 In this section, simulation is performed to evaluate the performance of our
-proposed methods in locating the ``correct" number of clusters. We compare
+proposed methods in locating the ``correct'' number of clusters. We compare
 with a basket of existing methods including Gap statistics
@@
 Bootstrap stability \citep{fang2012selection} in following simulation
-settings. We select $p=q$ and $5-$fold cross-validation in row
+settings. All methods are executed with their default parameter settings. 
+We select $p=q$ and $5-$fold cross-validation in row
 ($m=\frac{1}{4}n$) as default parameter setting for our proposed Gabriel
 method. Note that set $p=q$ corresponding to $2-$fold cross-validation in
-column.
+column. Wold method (detailed in Appendix $A$) and correlation-corrected 
+Gabriel method (section $4.3$) are also included for comparison.
+
+Note that in all settings, cluster centers are randomly generated from 
+multivariate normal distribution $\mathcal{N}\left(\mathbf{0},\varsigma \mathbf{I}\right)$. 
+All clusters are well-separated, i.e. no overlapping. In fact, any simulated 
+clusters with minimum distance less than one unit was discarded, so there is
+clear definition of true number of clusters. The parameters $\varsigma$ is chosen 
+such that about half of the random realization were discarded. The idea is borrowed from
+\cite{tibshirani2001estimating}. The proportion of each method successfully picks the correct
+$k$ out of $100$ simulation trials is reported, along with its corresponding confident interval
+by Wilson's method \citep{wilson1927probable}.
+
 \begin{enumerate}
 
-  \item Single cluster in $10$ dimensions --- $200$ observations, each
-    observation uniformly distributed over $[0,1]$ in each dimension.
-
-  \item Two clusters in $4$ dimensions --- $50$ i.i.d
-    observations are generated from both multivariate normal
-    $\mathcal{N}\left(\mu_1,0.5\Sigma\right)$ and multivariate normal
-    $\mathcal{N}\left(\mu_2,1.5\Sigma\right)$, where 
-    cluster center $\mu_1 = (1,0,0,1)$ and $\mu_2 =(1,3.5,3.5,1)$
-    respectively. $\Sigma$ has $AR(1)$ structure with $\rho=-0.2$ and $\sigma=1$.
-
-  \item Four clusters in $100$ dimensions ---  Each cluster has $100$ or $150$
-    i.i.d standard normal observations, with cluster centers randomly
-    generated from multivariate normal distribution
-    $\mathcal{N}\left(\mathbf{0},0.65^2\mathbf{I}\right)$
-
-  \item Ten clusters in $100$ dimensions ---  Each cluster has $50$ or $100$
-    i.i.d standard normal observations, with cluster centers randomly
-    generated from multivariate normal distribution
-    $\mathcal{N}\left(\mathbf{0},0.72^2\mathbf{I}\right)$
-
-  \item Four log-normal clusters in $16$ dimensions --- For each cluster, $30$
-    or $60$ i.i.d centered log-normal observations are generated from
-    $ln\mathcal{N}\left(0,0.5^2 \right)$. The cluster centers are randomly
-    generated from multivariate normal distribution
-    $\mathcal{N}\left(\mathbf{0},1.2^2\mathbf{I}\right)$
-
-  \item[6.] Three exponential clusters in $20$ dimensions with different
-    variance --- $40$ observations in each cluster are generated from centered
-    exponential distribution ($exp(\lambda)-1/\lambda$ in each dimension),
-    with $\boldsymbol\mu$ randomly generated from
-    $\mathcal{N}\left(\mathbf{0},19\mathbf{I}\right)$ and $\lambda = 1, 1/2,
-    1/5$ respectively.
-
-  \item[7.] One cluster in $10$ dimensions with correlated noise -- $200$
-    mean-zero multivariate normal observations; the covariance matrix $\Sigma$
-    has $\Sigma_{ij} = 0.9$ for $i \neq j$ and $\Sigma_{ii} = 1$ for $i = 1,
-    \dotsc, 10$.
+  \item \textbf{Correlation between dimensions} -- Six clusters in $10$ dimensions. 
+  	Each cluster has $100$ or $50$
+    multivariate normal observations with common covariance matrix $\Sigma$ 
+    which has compound symmetric structure with $1$ in diagonal and $\rho$ 
+    off diagonal. $\rho$ takes value in $\{0,0.1,...,0.9\}$.
+
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting1/Facet.pdf}
+	\label{fig:setting1}
+	\end{figure}
+
+  \item \textbf{Noise dimensions} --- Three clusters in $6$ dimensions. Each cluster has $1000$ or $500$
+    mean zero multivariate normal observations with identity covariance matrix. 
+    We add $P$ dimensions of noise to the data, which is randomly generated from uniform$[0,1]$. The noise
+    dimension $P$ takes values in $\{0,6,...,54\}$.
+
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting2/Facet.pdf}
+	\label{fig:setting2}
+	\end{figure}
+
+  \item \textbf{High dimension} --- Eight clusters in $p$ dimensions, $p$ takes values 
+  	in $\{10,20,...,100\}$. Each cluster has $100$ or $50$ mean zero multivariate normal 
+  	observations with identity covariance matrix.
+
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting3/Facet.pdf}
+	\label{fig:setting3}
+	\end{figure}
+
+     \item \textbf{Variance heterogeneity} --- Three clusters in $20$ dimensions, each has
+     $60$ observations in it. Observations are generated from 
+     $\mathcal{N}\left(\mathbf{0},\sigma_1^2\mathbf{I}\right)$,
+      $\mathcal{N}\left(\mathbf{0},\sigma_2^2\mathbf{I}\right)$ and
+       $\mathcal{N}\left(\mathbf{0},\sigma_3^2\mathbf{I}\right)$ where 
+       $\sigma_1^2 : \sigma_2^2: \sigma_3^2 = 1:\frac{1+R}{2}:R$. The maximum ratio $R$
+       takes values in $\{1,5,10,...,45\}$
+
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting4/Facet.pdf}
+	\label{fig:setting4}
+	\end{figure}
+
+	\item \textbf{Heavy tail data} --- Five clusters in $15$ dimensions, each has
+     $80$ observations in it. Observations have independent $t$ distribution in each dimension  
+     with degree of freedom $\nu$, which takes value in $\{11,10,...,2\}$
+
+	\begin{figure}[H]
+	\centering
+	\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting5/Facet.pdf}
+	\label{fig:setting4}
+	\end{figure}
 \end{enumerate}
-Note that in setting 2--6, all clusters are well-separated, i.e. no
-overlapping. In fact, any simulated clusters with minimum distance less than
-1 unit was discarded, so there is clear definition of true number of
-clusters. The parameters in setting 2--6 are chosen such that about half of
-the random realization were discarded. The idea is borrowed from
-\cite{tibshirani2001estimating}.
-
-Table \ref{table1} shows the distribution of $k$ selected by each algorithm in 
-each simulation setting. We run $k$ from $1$ to $15$, and the best $k$ within this range
-is selected by the algorithms. [ NA in the table means...]
-\clearpage
-
-\begin{center}
-\footnotesize
-\begin{longtable}{lcccccccccccccccc}
-\captionsetup{justification=centering}
-\caption{\label{table1} Simulation Results}\\
-\toprule
-\phantom{Pred.~Strength}
-  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
-\cmidrule(lr){2-17}
-Method
-  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
-\midrule
-\endfirsthead
-\multicolumn{17}{c}%
-  {\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
-\midrule
-\phantom{Pred.~Strength}
-  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
-\cmidrule(lr){2-17}
-Method
-  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
-\midrule
-\endhead
-\multicolumn{17}{r}{\textit{Continued on next page}} \\
-\endfoot
-\bottomrule
-\multicolumn{17}{l}{\footnotesize
-  \textsuperscript{$\ast$} CH, Hartigan, and Stability are excluded from
-  Setting~1 because they cannot select $k=1$.
-}
-\endlastfoot
-\input{demo/bench/setting1/results.tex}
-\midrule
-\input{demo/bench/setting2/results.tex}
-\midrule
-\input{demo/bench/setting3/results.tex}
-\midrule
-\input{demo/bench/setting4/results.tex}
-\midrule
-\input{demo/bench/setting5/results.tex}
-\midrule
-\input{demo/bench/setting6/results.tex}
-\midrule
-\input{demo/bench/setting7/results.tex}
-\end{longtable}
-\end{center}
-
-\noindent
-All methods are used with their default parameter setting except for Gap
-statistics. We selected the $k$  corresponds to the global maximum of gap
-statistics, rather than the default method which selects the smallest $k$ such
-that the gap statistics is not more than 1 standard error away from the first
-local maximum. This is because among all the possible options, the global
-maximum criteria gives the best result in our simulation settings.
-
-For CH-index, Hartigen and Bootstrap stability, the minimum $k$ they can pick is $2$. So they are not
-included for comparison since the true $k=1$. It also demonstrates the difficulty 
-that null scenario poses for some existing algorithms. For those could select $k=1$,
-most of them work well under such setting except for Jump method. Setting $2-4$ have Gaussian
-distribution, where the Gaussian mixture model-based method suppose to
-perform the best. However, it only performs reasonably well in setting $2$. In setting $3$ and $4$,
-it has a very poor performance even if the underlying distribution is indeed
-Gaussian mixture model. It highlights the difficulty that high dimension
-data could cause. Other methods (CH-index, Hartigan, Jump
-method, Prediction strength and Boostrap stability) also have problems in
-these two high-dimension settings as we could not find any one with reasonable
-performance in terms of finding the correct $k$ in both cases. The only exception is
-Gap statistics, which works perfectly in both settings. Our proposed Gabriel method (also Wold
-CV method) clearly stands out from the basket of existing methods under the 
-high-dimension setting $3$ and $4$, as we can see they both perform perfectly under
-such settings.
-
-Setting $5$ and $6$ corresponding to the situation where data are heavy-tailed
-instead of normal distributed, coupled with unequal number of observations or
-unequal variance for different clusters. From Table \ref{table1}, we can see none of 
-the existing methods has good performance in both settings. In contrast,
-our proposed Gabriel method is robust and has superior performance
-compare with other methods. Because one can hardly tell the underlying
-distribution of data in practice, the resilience to non-Gaussian data gives
-the Gabriel method a clear edge.
-
-\section{Real data application}
-
-We also applied our proposed method to three real world data sets obtained
-from the University of California Irvine machine learning repository. The
-first and third data sets are selected because there are clear number of
+
+From the graph of setting $1$, we can see high correlation between dimensions cause problem for most 
+existing methods as well as the proposed Gabriel CV method and Wold method. Higher correlation clearly 
+degenerate most algorithms' ability to find the correct $k$ as data no longer spherical in $p$-dimension.
+And the recent proposed methods (Gabriel CV method, Gap statistic, Jump method) seem to be more sensitive to
+high correlation than those old methods such as CH-index and Hartigan statistics. The only two methods 
+work well in high correlation are the Gaussian model-based BIC method \citep{fraley2002model} and the 
+correlation-corrected Gabriel method.
+
+Plot of setting $3$ shows some methods are insensitive to higher dimension such as Jump method and Gap 
+statistics, while others deteriorate quickly with increasing dimension, most notably the Gaussian 
+model-based BIC method. Note here the data are still generated from mixture of Gaussian. In contrast, our 
+proposed Gabriel CV method as well as its correlation-corrected version actually working better in higher 
+dimension.
+
+Result of setting $4$ indicates most previously methods are sensitive to variance heterogeneity among 
+clusters, most notably the Gap statistics and model-based BIC method. The proposed Gabriel CV method
+and its correlation-corrected version consistently perform well in estimating $k$ and quite insensitive
+to variance heterogeneity. The Wold method also performs well in this setting.
+
+The last non-Gaussian setting examines methods' performance in heavy-tail data. With degree of freedom 
+decreases, the tail becomes more flat and the Gaussian assumption becomes more inappropriate. It explains
+the sharp plunge of Gaussian model-based BIC method in the plot. For most methods, their performances are 
+relatively stable until the tail gets very heavy. In case where degree of freedom is $2$, Gap statistics 
+and Jump method dived considerably compare to the Gabriel method and Wold method.
+
+In conclusion, the proposed Gabriel CV method compares well with current state-of-the-art methods and 
+is very robust against variance heterogeneity, high dimension and heavy-tail data. One weakness of the 
+Gabriel CV method is that it's sensitive to the high correlation among dimensions. In such case, its 
+correlation-corrected version will perform well under the assumption of common covariance structure.
+
+\section{Empirical validation}
+
+We also applied our proposed method to three real world data sets, two were
+obtained from the University of California Irvine machine learning repository.
+The first and third data sets are selected because there are clear number of
 clusters in those two data sets. The second data set is used as a benchmark
@@
 $98$th United States Congress, $2$nd session \citep{schlimmer1987concept}.
-This data set includes votes for each of the U.S. House of Representatives Congressmen on the $16$ key votes
-identified by the \textit{CQA} (Congressional Quarterly Almanac). For each
-vote, each Congressman either vote positively ``yea" (voted for/paired
-for/announced for),  negatively ``nay'' (voted against/paired
-against/announced against) or position unknown ``?". We took out those records
-contain ``?" before comparing each algorithm. It results in $232$ remaining
-records, with $124$ democrat and $108$ republican.
-
-The second data set is the well-known Wisconsin breast cancer data set \citep{mangasarian1990pattern}.
-After excluding the records with missing data, this data set consists records of $683$ patients, each with measurements of nine attributes of their biopsy specimens.
-It is known that there exist at least two groups of patients: $444$ patients with benign specimens and $239$ patients with malignant specimens.
-
-The third data set is the Sonar data returned from two targets -- a metal
-cylinder and a rock with similar shape, which is first studied by
-\cite{gorman1988analysis} using a neural network. Both targets were impinged
-by pulse which was a wide-band linear FM chirp $(ka = 55.6)$. Returns were
-collected at a range of $10$ meters and obtained from the cylinder at aspect
-angles spanning $90^{\circ}$ and from the rock at aspect angles spanning
-$180^{\circ}$. The data set contains $208$ returns ($111$ cylinder returns and
-$97$ rock returns), with each composed of $60$ spectral samples, normalized to
-take on values between $0$ and $1$. So the data has $60$ features with clearly
-$2$ clusters.
+This data set includes votes for each of the U.S. House of Representatives
+Congressmen on the $16$ key votes identified by the \textit{CQA}
+(Congressional Quarterly Almanac). For each vote, each Congressman either vote
+positively ``yea" (voted for/paired for/announced for),  negatively ``nay''
+(voted against/paired against/announced against) or position unknown ``?". We
+took out those records contain ``?" before comparing each algorithm. It
+results in $232$ remaining records, with $124$ democrat and $108$ republican.
+
+The second data set is the well-known Wisconsin breast cancer data set
+\citep{mangasarian1990pattern}. After excluding the records with missing data,
+this data set consists records of $683$ patients, each with measurements of
+nine attributes of their biopsy specimens. It is known that there exist at
+least two groups of patients: $444$ patients with benign specimens and $239$
+patients with malignant specimens.
+
+The third data set is gene expression data of $5$ types of brain tumours from
+\cite{de2008clustering}, which contains $42$ observations including $10$
+medulloblastomas, $10$ malignant gliomas, $10$ atypical teratoid/rhabdoid
+tumours (AT/RTs), $8$ primitive neuroectodermal tumours (PNETs) and $4$ normal
+cerebella. It was originally used by \cite{pomeroy2002prediction} to show that
+medulloblastomas are molecularly distinct from other types of brain tumours.
+
+Data was collected with Affymetrix microarrays, which estimates the number of
+RNA copies found in the cell sample (frozen specimens). Data was preprocessed
+as following: as a common procedure, all genes with expression level below
+$10$ are set to a minimum level threshold of $10$. The maximum threshold is
+set at $16,000$. Values below or above these thresholds are often unreliable.
+To remove uninformative genes, for each gene $j$ (attribute), the mean $m_j$
+among the samples was calculated. The $10\%$ largest and smallest values are
+discarded in order to remove extreme values. Based on the mean, following
+transformation is applied on every value $x_{ij}$ of sample $i$ and gene j \[
+y_{ij} = \log_2 (x_{ij}/m_j)	\]
+
+Genes with expression levels differing by at least $l$-flod in at least $c$
+samples from their mean expression level across the sample were selected. The
+parameter $l$ and $c$ were chosen so as to produce a filtered data set with at
+least $10\%$ of the original number of genes \citep{de2008clustering}. Such
+preprocessing procedure results in $1379$ genes selected. After filtering
+step, data is restored in the original scale (i.e. $x_{ij}$). So the data set
+is $42 \times 1379$, which serves as a high dimensional example.
 
@@
     \hline                  
- & Congress Voting && Breast Cancer && Sonar \\ \hline                    
+ & Congress Voting && Breast Cancer && Brain Tumours \\ \hline                    
 CH-index & $2$ && $2$  && $2$    \\
-Hartigan & $3$ && $3$  &&  $3$   \\
-Jump & $9$ && $9$ && $10$  \\   
+Hartigan & $3$ && $3$  &&  $4$   \\
+Jump & $10$ && $9$ && $1$  \\   
 Prediction strength & $2$ &&  $2$ && $1$    \\
-Bootstrap stability & $2$ &&  $2$ &&  $10$   \\ 
-Gap & $10$ && $9$ && $10$  \\   
-Gaussian-Mix & $7$ && $5$ && $1$  \\   
-Gabriel & $2$  && $3$  && $2$  \\    
-Wold & $2$ && $3$ && $10$ \\  \hline 
+Bootstrap stability & $2$ &&  $2$ &&  $7$   \\ 
+Gap & $8$ && $10$ && $10$  \\   
+Gaussian-BIC & $2$ && $5$ && $2$  \\   
+Gabriel & $2$  && $3$  && $5$  \\    
+Gabriel-corr-correct & $2$  && $2$  && $5$  \\  
+Wold & $2$ && $3$ && $4$ \\  \hline
+
 \end{tabular}
@@
 
-Since most congressmen vote base on their parties' interest, $2$ parties
+Since most congressmen vote based on their parties' interest, two parties
 (Democratic and Republican) represent two clusters in this data set. So the
-optimal number should be two. Close inspection shows $k$-means with
-$k=2$ separates the two parties very well with the lowest miss-classification
-error $10.43\%$ ($k=3$ has $14.78\%$). Note that CH-index and Bootstrap
-stability also return $k=2$, but $2$ is the lower bound those methods can
-select for $k$. So it's not clear they actually choose $k=2$ or they hit the
-lower bound (they would pick $k=1$ if allowed). For the breast cancer data,
-it's known to have at least $2$ cluster based on whether it's benign specimens
-or not. But it doesn't mean the optimal $k$ should be $2$.
+optimal number should be two. Close inspection shows $k$-means with $k=2$
+separates the two parties very well with the lowest miss-classification error
+$10.43\%$ ($k=3$ has $14.78\%$). Our proposed Gabriel CV method, along with
+several other algorithms correctly pick $k=2$ for this data set. Note that
+CH-index and Bootstrap stability also return $k=2$, but $2$ is the lower bound
+those methods can select for $k$. So it's not clear they actually choose $k=2$
+or they hit the lower bound (they would pick $k=1$ if allowed).
+
+For the breast cancer data, it's known to have $2$ clusters based on whether
+it's benign specimens or malignant specimens. Although the proposed Gabriel
+method picks $k=3$, its correlation-corrected version correctly picks $k=2$.
+It shows the necessity to correct for the high correlation among data
+features. However, for this data $k=3$ is also an acceptable answer, as
 \cite{fujita2014non} noticed that the malign group is quite heterogeneous and
-can be further clustered into at least two subgroups. Hence, the result $k=3$
-given by our proposed Gabriel method (as well as Wold method) make sense. For
-the Sonar data set which is relatively high-dimension, only our proposed
-Gabriel method and CH-index selects $k=2$. Majority of the methods collapse to
-either select the maximum or select the minimum number, underline how
-difficult it was to pick the right $k$ when dimension increases.
-
-\section{Conclusion}
+can be further clustered into at least two subgroups.
+
+The brain tumours gene expression data is a typical high-dimension data with
+$p > n$. It is clear that the correct number of clusters should be $5$, the
+number of brain tumour types. Indeed, $k$-means results with $k=5$ show that
+the $4$ normal human cerebella are well separated from brain tumours as they
+form a single cluster.  Medulloblastomas, malignant gliomas and AT/RT tumours
+also form $3$ respective clusters, which means they can be separated from each
+other. Such results are similar to those found in
+\cite{pomeroy2002prediction}, whose analysis is based on the first $3$
+principle components of the original data.    From table \ref{table2} we can
+see only our proposed Gabriel method (as well as its correlation-corrected
+version) correctly selects $k=5$, underline how difficult it was to pick the
+right $k$ when dimension is high.
+
+\section{Application}
+
+\textit{TODO: add data applicaiton where number of clusters is unknown}
+
+\section{Discussion}
+\label{sec:conc}
 
@@
 unsupervised learning problem into supervised learning problem via novel form
-of cross validation. We proved the self-consistency for our proposed Gabriel
-CV method as well as its asymptotic property with noise, and showed the
-robustness of our method by simulation. Our method has very good performance
-in our limited simulation settings, and clearly the superior one when data is
-high dimension or is heavy-tailed.
-
-\section*{\textbf{APPENDIX}}
+of cross validation. Such approach is quite different from previous methods which 
+utilize the within/between cluster dispersion or stability criterion for selecting the 
+optimal $k$. Our method utilizes the connection between different dimensions (columns)
+of data through the uniqueness of each cluster center. We proved the self-consistency 
+for our proposed Gabriel CV method as well as its asymptotic property with Gaussian noise,
+and showed the robustness of our method by simulation. Our method has very good performance
+in our limited simulation settings and real data application, and clearly the superior
+one when data has heterogeneous variance or is heavy-tailed.
+
+Besides no strong modeling assumption is required, our proposed method is robust against data set 
+with variance heterogeneity, unequal number of observations, non-Gaussian noise and high-dimension. 
+Such robustness is important in practice because for any given data, it's hard to tell whether
+its clusters have different number of observations, is the variance equal for each cluster,
+or what underlying noise distribution it has. The weakness of our proposed method is 
+that its theory assumes only week correlation between ``predictor'' columns and ``response'' 
+columns. In practice, many data sets don't exhibit high correlations between columns, where
+our proposed method can be safely applied. In case the high correlation does exist, 
+the correlation-corrected version of Gabriel CV method in section $4.3$ can be used if we can 
+assume common covariance structure. Other procedure such as leave out most columns for clustering 
+may also be used to reduce its effect. However, the theory for such procedure has not
+been fully developed and it could be the future research topic.
+
+Another situation where our proposed method cannot be directly used is when the clusters 
+are non-convex. In such situation, $k$-means itself doesn't work well, for example two 
+concentric circles share the same cluster center \citep{hastie2009elements}. However, it's 
+possible that our proposed Gabriel CV method can be used on the transformed data set. In the
+concentric circles case where spectral clustering is appropriate, our proposed method can be applied 
+on the eigenvector subspace of the graph Laplacian matrix inside the spectral clustering algorithm
+to find the optimal $k$. This can also be the future research topic.
+
 \appendix
+
+\section{Clustering scree plot}
+
+\label{sec:elbow-fail}
+
+The top row of Figure~\ref{fig:elbow}
+displays an example where the elbow in $W_k$ corresponds to the true number $k
+= 4$ of mixture components in the data-generating mechanism. The elbow
+approach is simple and often performs well, but it requires subjective
+judgment as to where the elbow is located, and, as the bottom row of
+Figure~\ref{fig:elbow} demonstrates, the approach can easily fail.
+
+\begin{figure}
+\centering
+\begin{minipage}{\linewidth}
+  \begin{minipage}{0.45\linewidth}
+    \includegraphics[width=\linewidth]{demo/elbow/correct-data.pdf}
+  \end{minipage}
+  \hspace{0.05in}
+  \begin{minipage}{0.45\linewidth}
+    \includegraphics[width=\linewidth]{demo/elbow/correct-withinss.pdf}
+  \end{minipage}
+\end{minipage}
+\begin{minipage}{\linewidth}
+  \begin{minipage}{0.45\linewidth}
+    \includegraphics[width=\linewidth]{demo/elbow/incorrect-data.pdf}
+  \end{minipage}
+  \hspace{0.05in}
+  \begin{minipage}{0.45\linewidth}
+    \includegraphics[width=\linewidth]{demo/elbow/incorrect-withinss.pdf}
+  \end{minipage}
+\end{minipage}
+\caption{Left panels show the $(X,Y)$ data points; right panels
+  show the corresponding values of the within-cluster sum of squares $W_k$
+plotted against the number of clusters, $k$.}
+\label{fig:elbow}
+\end{figure}
+
+\section{Review of consistency results for clustering}
+
+Some proposed methods provide theoretical consistency result for choosing $k$.
+Assume data is a mixture of $G$ $p$-dimension clusters with equal prior,
+identically distributed with common covariance matrix and finite fourth
+moments in each dimension, \citet{sugar2003finding} shows the Jump method will
+pick the correct $k$ under the conditions that cluster centers are
+sufficiently separated and an appropriate transformation is used.
+Specifically, let $\Delta\sqrt{p}$ denotes the minimum Euclidean distance
+between cluster centers, \citet{sugar2003finding} shows that the jump will be
+maximized at $k=G$ given that $\Delta > 6$ and the existence of a positive $Y$
+such that
+\[
+  \left(\frac{p\Delta^2W}{9G}\right)^{-Y}
+    + \left(
+        W \left[\frac{2^{2H^*(X)}}{K^2_{max}2\pi e}\right]
+        - (\frac{\Delta}{6})^2(1-W)
+      \right)^{-Y} < 2
+\]
+and $\left(\frac{p\Delta^2W}{9G}\right)^{-Y} <1/2$, where $H^*(X)$ is the
+minimum entropy of the cluster membership over each dimension and $W =1-
+\frac{6^4V_{\mathbf{X}}}{(\Delta^2-36)^2}$ with $V_{\mathbf{X}} = \var\left(
+  1/p ||\mathbf{X}-\mu_j||^2_{\Gamma^{-1}} \mid \mathbf{X}
+\hspace{0.05in}\text{in jth cluster}\right)$
+
+\cite{tibshirani2005cluster} proved that the prediction strength method is
+consistent under the setting that the observations in $p$-dimension are
+uniformly distributed within one unit from their respective cluster centers,
+with minimum distance between the $G$ cluster centers is four unit. Under such
+well-separated setting, they show that 
+\[ ps(G) = 1+ o_p(1), \hspace{0.5in} \underset{k \geq G+1}{\text{sup}} ps(k)\leq \frac{2}{3}+ o_p(1) \]
+therefore $\hat{k}$ is consistent in estimating $G$.  When data does come from
+mixture of Gaussian distribution, the model-based method of
+\citet{fraley2002model} is consistent in estimating $k$ in low dimension.
+
+Let $\Psi$ denotes any given base-clustering algorithm (e.g $k$-means), and
+$\Psi(Z,k)$ the clustering obtained by applying $\Psi$ on data $Z$ with
+parameter $k$. Also let $k_0$ be the optimal number of cluster and
+$d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ be the measure of distance between clustering
+$\Psi(Z_1,k)$ and $\Psi(Z_2,k)$ where $Z_1$ and $Z_2$ are two independent
+samples from population. \citet{wang2010consistent} shows that their proposed
+method will consistently select $k = k_0$, given such $k_0$ exists and
+$d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ converges to zero at proper rate $r_k$.
+However, $k_0$ is defined as the most stable number for $\Psi$ among all the
+possible $k$, instead of the number of components in mixture model or
+well-separated compact regions commonly seen in literature. Therefore, their
+consistent result actually is convergent result, i.e. the selected $k$
+converges to its limit $k_0$ given such limit exists.
+
 \section{Wold CV estimation} \label{app:foobar}
@@
 
-\section{Technical Lemmas}
-\label{app:technical-lemmas}
-
-\begin{lemma}\label{lem:truncated-normal-moments}
-
-If $Z$ is a standard normal random variable, then
-\[
-  \E(Z \mid a < Z < b)
-    = - \frac{\varphi(b) - \varphi(a)}
-             {\Phi(b) - \Phi(a)}
-\]
-and
-\[
-  \E\{(Z - \delta)^2 \mid a < Z < b\}
-    = \delta^2 + 1
-    - \frac{  (b - 2 \delta) \varphi(b)
-            - (a - 2 \delta) \varphi(a)}
-           {\Phi(b) - \Phi(a)}
-\]
-for all constants $a$, $b$, and $\delta$, where $\varphi(z)$ and $\Phi(z)$ are
-the standard normal probability density and cumulative distribution functions.
-These expressions are valid for $a = -\infty$ or $b = \infty$ by taking
-limits.
-
-\end{lemma}
-\begin{proof}
-We will derive the expression for the second moment.  Integrate to get
-\begin{align*}
-  \E[ (Z - \delta)^2 1\{Z < b\}]
-    &= \int_{-\infty}^b (z - \delta)^2 \varphi(z) \, dz \\
-    &= (\delta^2 + 1) \Phi(b) - (b - 2 \delta) \varphi(b).
-\end{align*}
-Now,
-\[
-  \E\{(Z - \delta)^2 \mid a < Z < b\}
-    =
-    \frac{  \E[ (Z - \delta)^2 1\{Z < b\}]
-          - \E[ (Z - \delta)^2 1\{Z < a\}]}
-         { \Phi(b) - \Phi(a) }.
-\]
-\end{proof}
-
-Lemma~\ref{lem:truncated-normal-moments} has some important special cases:
-\begin{align*}
-  \E\{Z \mid Z > 0\} &= 2 \varphi(0) = \sqrt{2 / \pi}, \\
-  \E\{(Z - \delta)^2 \mid Z > 0 \}
-    &= \delta^2 + 1 - 4 \delta \varphi(0), \\
-  \E\{(Z - \delta)^2 \mid Z < 0 \}
-    &= \delta^2 + 1 + 4 \delta \varphi(0).
-\end{align*}
+\bibliographystyle{apalike}
 
 \bibliography{references}
-\bibliographystyle{apalike}
 \end{document}

### 2016-04-19 New simulation/technometrics.tex
Revised Yeast data application

@@
 
+\subsection{Application on Yeast data set}
+The yeast Saccharomyces cerevisiae data set was collected by \cite{cho1998genome} to study the genome-wide characterization of mRNA transcript levels during the cell cycles of the budding yeast S.
+cerevisiae.
+To obtain data,
+cdc28-13 cells were synchronized in late G1 phase by heat treatment, and the cell cycle was reinitiated by shifting cells to cooler temperature.
+Data was collected at $17$ time points at every $10$ minutes intervals covering almost $2$ complete cell cycles.
+Oligonucleotide microarrays was used to query $6,220$ gene expression profiles at these $17$ time points.
+This data has been the subject of several studies and is generally accepted as a reference \citep{dortet2008model}.
+
+\cite{tavazoie1999systematic} applied $k$-means clustering on the preprocessed data with $k=30$ and discovered several periodic clusters.
+The preprocessing procedure includes choosing the most variable $2,945$ ORFs among the total $6,220$ by metric of variation based on the normalized dispersion across time points (s.d/mean).
+Data at time points $90$ and $100$ minutes were removed because of their less efficient labelling of mRNA during the original chip hybridization.
+It results a $2,945$ by $15$ data matrix.
+The last step of the preprocessing procedure is to variance normalize each gene across the $15$ time points, which is done by subtracting the mean and then dividing by the standard deviation.
+The result is each gene has mean zero and standard deviation equal to one across the time points.
+
+\cite{dortet2008model} performed model-based clustering on the preprocessed data of \cite{tavazoie1999systematic} and  found $27$ clusters.
+They classified the $27$ clusters into $3$ major types: cyclic gene clusters, stress gene clusters and the clusters neither cyclically nor sensitive to the induced stress.
+
+We also applied our proposed Gabriel method on the preprocessed data of \cite{tavazoie1999systematic}.
+The algorithm finds $9$ clusters.
+Since the dimensions (time points) are likely to be correlated, we also applied the correlation-corrected version of Gabriel method on the data and find $5$ clusters.
+Both numbers are much smaller compare to those of \cite{tavazoie1999systematic} and \cite{dortet2008model}.
+We will focus our analysis on the $5$ clusters case, whose result is summarized in Figure \ref{fig:5_clusters}, where each cluster is presented by the mean expression level of all the genes within that cluster over $15$ time points.
+Similar as \cite{tavazoie1999systematic}, we found the members of each cluster to be significantly enriched for genes function in similar biological process.
+Each gene is mapped with the Gene Ontology Annotation in SGD database, which categorizes the genes into $103$ groups in terms of biological process.
+The $p$-value of observing $k$ genes from a particular category group within a cluster of size $m$ is given by:
+\[	P = 1-\sum^{k-1}_{i=1} \frac{{{2945-n}\choose{m-i}}{{n}\choose{i}} }{ {{2945} \choose{m}} }\] 
+where $n$ is the total ORFs in that group. The result is shown in Table \ref{table:enrich}.
+
+\begin{table}
+\begin{center}
+\captionsetup{justification=centering}
+\caption{\label{table:enrich} Enrichment of clusters for ORFs with biological process categories}
+\scalebox{0.8}{%
+\begin{tabular}{cclcl}
+\hline
+   Cluster & Cluster size($m$) & Biological process category(total ORFs) & ORFs within category($k$) &$p$-value \\ \hline
+  $1$& $514$ &  DNA repair ($115$)  &  $54$  & $7.5\times10^{-14}$     \\                 
+ 	 &     &  cellular response to DNA damage stimulus ($141$)  &  $61$  & $1.7\times10^{-13}$     \\  
+ 	 &     &  DNA replication ($77$)  & $41$   &  $4.7\times10^{-13}$   \\  
+  	 &     &  mitotic cell cycle ($169$)  &  $65$  &  $1.8\times10^{-11}$   \\  
+ $2$ & $583$ &  transcription from RNA polymerase II promoter ($209$) & $74$ &  $3.0\times10^{-8}$    \\  
+  	 &     & chromatin organization ($147$) & $52$ 	& $4.2\times10^{-6}$    \\  
+  	 &     & mitotic cell cycle ($169$)  & $57$ 	& $7.9\times10^{-6}$  \\  
+  $3$& $682$  &  mitochondrion organization ($193$) & $119$ &  $1.0\times10^{-32}$  \\  
+  	 &     &  mitochondrial translation ($73$) & $57$ 	&  $1.0\times10^{-23}$  \\  
+  	 &     &  generation of precursor metabolites and energy ($104$) &  $45$ & $3.0\times10^{-6}$   \\  
+  $5$& $603$  & cytoplasmic translation ($107$)  &  $93$ &  $1.6\times10^{-51}$   \\  
+     &     &  ribosomal subunit biogenesis $(115)$ & $59$ 	&  $6.1\times10^{-14}$  \\  
+  	 &     &  rRNA processing ($110$) &  $51$	&  $4.3\times10^{-10}$  \\  
+  	 &     &  cellular amino acid metabolic process $121$ &  $52$	&  $8.2\times10^{-9}$  \\
+
+\hline
+\end{tabular}
+}
+\end{center}
+\hspace{0.5in} \footnotesize {}
+\end{table}
+
+It turns out that each cluster contains genes that tend to participate in the common processes.
+It is easy to see that cluster $1$ and $2$ are periodic from Figure \ref{fig:5_clusters}.
+Indeed, more than $70\%$ genes which function in mitotic cell cycle process are concentrated in these two clusters.
+Genes participated in the cell-cycle phase related processes such as DNA replication and DNA repair are highly concentrated in cluster $1$.
+Genes related to budding and cell polarity, as well as chromatin organization and transcription are commonly found in cluster $2$.
+These cyclic gene clusters are also reported in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
+In fact, the mean expression profile plot of cluster $1$ resembles those of cluster $2$, $14$ of \cite{tavazoie1999systematic} and cluster $2$, $3$ of \cite{dortet2008model}, while cluster $2$ resembles cluster $7$ of \cite{tavazoie1999systematic}.
+
+Our analysis also found genes govern mitochondrial translation and mitochondrion organization are heavily concentrated in cluster $3$, while genes related to cytoplasmic translation are highlighted in cluster $5$.
+Genes encode ribosomes are also overly presented in cluster $5$.
+Those results are not reported in previous studies, meaning our analysis leads to new discovery of gene grouping.
+
+In conclusion, functional grouping of gene occurred in the $5$ clusters found by our proposed method.
+In particular, the periodic cell cycle related genes are concentrated in cluster 1 and cluster 2, which broadly consistent with the findings in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
+The number of clusters found by our method is much smaller than the number of clusters used in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
+However, both studies found that most of their clusters have no characteristics, suggesting overestimate of the underlying diversity of biological expression classes in the data set \citep{tavazoie1999systematic}.
+Our result is more compact, while still able to identify the functional grouping of genes in clusters highlighted in previous studies.
+
+	\begin{figure}[H]
+		\centering
+	\includegraphics[width=5.5in, height=3.3in]{wei/5_clusters.pdf}
+	\caption{Yeast data set--mean expression profiles of the 5 clusters found by correlation-corrected Gabriel method}
+	\label{fig:5_clusters}
+	\end{figure}
+
 \section{Discussion}

### 2016-04-19 New simulation/wei/Yeast Data Example.R

@@ -1,0 +1,138 @@
+
+Data <- read.table("Yeast/cdc15.txt",header=T,sep="\t")
+head(Data)
+DATA = Data[,52:68]
+
+##impute NAs by the column mean
+Test <- as.matrix(DATA)
+for(i in 1:ncol(Test)){
+ 	IDs <- which(is.na(Test[,i]));
+	Impute <- median(Test[,i],na.rm=T);
+	Test[,i][IDs] <- Impute;
+}
+##impute NAs by K-NN with k =5
+##library("DMwR")
+##Test <- as.matrix(DATA)
+##Test <- knnImputation(Test,k=15)
+
+##select the top 3000/2945 most variable genes(s.d/mean)
+Rank <- rep(NA,nrow(Test))
+
+for(i in 1:nrow(Test)){
+	Rank[i] = sd(Test[i,])/mean(Test[i,])
+}
+
+Order <- order(Rank,decreasing = T)
+Id <- Order[1:2945]
+Test <- Test[Id,]
+Gene <- as.character(Data$X)[Id]
+Gene <- as.character(lapply(Gene,function(x)substr(x,1,7)))
+
+##variance normalize each gene
+Test <- Test[,-c(10,11)]
+
+for(i in 1:nrow(Test)){
+	Mean = mean(Test[i,]);
+	SD = sd(Test[i,]);
+	for(j in 1:ncol(Test)){
+		Test[i,j] = (Test[i,j]-Mean)/SD
+	}
+}
+################################################################################
+
+set.seed(1)
+K.original <- cv.kmeans.gabriel(Test, 5, 2, maxcenters=40, classify.method="nearest")$centers
+K.original
+DATA_new <- Uncorrelate2(Test,K.original)
+
+K2<- cv.kmeans.gabriel(DATA_new, 5, 2, maxcenters=40, classify.method="nearest")$centers
+K2
+
+####gabriel_cor_correct(Test, maxcenters=40, type = 2)
+#set.seed(1)		
+#Gap <- clusGap(Test, FUN = kmeans, K.max =45)
+#which(Gap[[1]][,3] == max(Gap[[1]][,3]))
+			
+#set.seed(1)	
+#mcluster <- Mclust(Test, G = 20:40)
+#mcluster$G
+###################################################################################						
+###==============================================================
+Data <- as.data.frame(Test)
+Data$Gene = Gene
+
+##set.seed(1)
[…]

### 2016-04-19 New simulation/technometrics.tex
Revised Yeast data application

@@
 We also applied our proposed Gabriel method on the preprocessed data of \cite{tavazoie1999systematic}.
-The algorithm finds $9$ clusters.
-Since the dimensions (time points) are likely to be correlated, we also applied the correlation-corrected version of Gabriel method on the data and find $5$ clusters.
-Both numbers are much smaller compare to those of \cite{tavazoie1999systematic} and \cite{dortet2008model}.
-We will focus our analysis on the $5$ clusters case, whose result is summarized in Figure \ref{fig:5_clusters}, where each cluster is presented by the mean expression level of all the genes within that cluster over $15$ time points.
+The algorithm found $5$ clusters.
+Since the dimensions (time points) are likely to be correlated, we also applied the correlation-corrected version of Gabriel method on the data which also result $5$ clusters.
+This numbers is much smaller compare to those of \cite{tavazoie1999systematic} and \cite{dortet2008model}.
+We will focus our analysis on this $5$ clusters found by the proposed method, whose result is summarized in Figure \ref{fig:5_clusters}, where each cluster is presented by the mean expression level of all the genes within that cluster over $15$ time points.
 Similar as \cite{tavazoie1999systematic}, we found the members of each cluster to be significantly enriched for genes function in similar biological process.
@@
 \[	P = 1-\sum^{k-1}_{i=1} \frac{{{2945-n}\choose{m-i}}{{n}\choose{i}} }{ {{2945} \choose{m}} }\] 
-where $n$ is the total ORFs in that group. The result is shown in Table \ref{table:enrich}.
+where $n$ is the total ORFs in that group. The enrichment of each category is shown in Table \ref{table:enrich}.
 
@@
    Cluster & Cluster size($m$) & Biological process category(total ORFs) & ORFs within category($k$) &$p$-value \\ \hline
-  $1$& $514$ &  DNA repair ($115$)  &  $54$  & $7.5\times10^{-14}$     \\                 
- 	 &     &  cellular response to DNA damage stimulus ($141$)  &  $61$  & $1.7\times10^{-13}$     \\  
- 	 &     &  DNA replication ($77$)  & $41$   &  $4.7\times10^{-13}$   \\  
-  	 &     &  mitotic cell cycle ($169$)  &  $65$  &  $1.8\times10^{-11}$   \\  
- $2$ & $583$ &  transcription from RNA polymerase II promoter ($209$) & $74$ &  $3.0\times10^{-8}$    \\  
-  	 &     & chromatin organization ($147$) & $52$ 	& $4.2\times10^{-6}$    \\  
-  	 &     & mitotic cell cycle ($169$)  & $57$ 	& $7.9\times10^{-6}$  \\  
-  $3$& $682$  &  mitochondrion organization ($193$) & $119$ &  $1.0\times10^{-32}$  \\  
-  	 &     &  mitochondrial translation ($73$) & $57$ 	&  $1.0\times10^{-23}$  \\  
-  	 &     &  generation of precursor metabolites and energy ($104$) &  $45$ & $3.0\times10^{-6}$   \\  
-  $5$& $603$  & cytoplasmic translation ($107$)  &  $93$ &  $1.6\times10^{-51}$   \\  
-     &     &  ribosomal subunit biogenesis $(115)$ & $59$ 	&  $6.1\times10^{-14}$  \\  
-  	 &     &  rRNA processing ($110$) &  $51$	&  $4.3\times10^{-10}$  \\  
-  	 &     &  cellular amino acid metabolic process $121$ &  $52$	&  $8.2\times10^{-9}$  \\
+ $1$ & $550$  &  response to oxidative stress ($55$)  &    $24$  & $1.5\times10^{-5}$ \\
+     &     &  response to chemical ($213$)    &    $64$     & $2.2\times10^{-5}$\\
+  $2$& $590$  &  mitochondrion organization ($159$) & $79$ &  $1.1\times10^{-16}$  \\  
+  	 &     &  mitochondrial translation ($51$) & $28$ 	&  $2.9\times10^{-8}$  \\  
+  	 &     &  generation of precursor metabolites and energy ($80$) &  $37$ & $7.3\times10^{-8}$   \\  
+ $3$ & $654$ &  transcription from RNA polymerase II promoter ($214$) & $75$ &  $5.5\times10^{-6}$    \\  
+  	 &     & mRNA processing ($67$) & $30$ 	& $2.7\times10^{-5}$    \\  
+  	 &     & mitotic cell cycle ($183$)  & $63$ 	& $6.2\times10^{-5}$  \\ 
+  $4$& $634$  & cytoplasmic translation ($134$)  &  $105$ &  $3.3\times10^{-47}$   \\  
+     &     &  ribosomal subunit biogenesis $(138)$ & $73$ 	&  $7.7\times10^{-17}$  \\  
+  	 &     &  rRNA processing ($131$) &  $61$	&  $5.7\times10^{-11}$  \\  
+  	 &     &  ribosome assembly ($36$) &  $21$	&  $1.5\times10^{-6}$  \\  
+$5$  &   $517$   &  chromosome segregation($106$) &  $53$      &  $6.0\times10^{-15}$ \\
+     &     &  cellular response to DNA damage stimulus ($172$)  &  $71$  & $3.6\times10^{-14}$     \\  
+  	 &     &  DNA repair ($147$)  &  $64$  & $3.7\times10^{-14}$     \\                 
+ 	 &     &  DNA replication ($78$)  & $42$   &  $1.8\times10^{-13}$   \\  
+  	 &     &  mitotic cell cycle ($183$)  &  $70$  &  $4.8\times10^{-12}$   \\
 
@@
 It turns out that each cluster contains genes that tend to participate in the common processes.
-It is easy to see that cluster $1$ and $2$ are periodic from Figure \ref{fig:5_clusters}.
+It is easy to see that cluster $5$ and $3$ are periodic from Figure \ref{fig:5_clusters}.
 Indeed, more than $70\%$ genes which function in mitotic cell cycle process are concentrated in these two clusters.
-Genes participated in the cell-cycle phase related processes such as DNA replication and DNA repair are highly concentrated in cluster $1$.
-Genes related to budding and cell polarity, as well as chromatin organization and transcription are commonly found in cluster $2$.
+Genes participated in the cell-cycle phase related processes such as DNA replication and DNA repair are highly concentrated in cluster $5$.
+Genes related to budding and cell polarity, as well as chromatin organization and transcription are commonly found in cluster $3$.
 These cyclic gene clusters are also reported in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
-In fact, the mean expression profile plot of cluster $1$ resembles those of cluster $2$, $14$ of \cite{tavazoie1999systematic} and cluster $2$, $3$ of \cite{dortet2008model}, while cluster $2$ resembles cluster $7$ of \cite{tavazoie1999systematic}.
-
-Our analysis also found genes govern mitochondrial translation and mitochondrion organization are heavily concentrated in cluster $3$, while genes related to cytoplasmic translation are highlighted in cluster $5$.
-Genes encode ribosomes are also overly presented in cluster $5$.
+In fact, the mean expression profile plot of cluster $5$ resembles those of cluster $2$, $14$ of \cite{tavazoie1999systematic} and cluster $2$, $3$ of \cite{dortet2008model}, while cluster $3$ corresponding to cluster $7$ of \cite{tavazoie1999systematic}.
+
+Cluster $1$ concentrates genes that somatize the cell stress, such as oxidative heat-induce proteins. This cluster corresponding to the cluster $5$ and $11$ in \cite{dortet2008model}.
+Our analysis also found genes govern mitochondrial translation and mitochondrion organization are heavily concentrated in cluster $2$, while genes related to cytoplasmic translation are highlighted in cluster $4$.
+Genes encode ribosomes are also overly presented in cluster $4$.
 Those results are not reported in previous studies, meaning our analysis leads to new discovery of gene grouping.
@@
 In conclusion, functional grouping of gene occurred in the $5$ clusters found by our proposed method.
-In particular, the periodic cell cycle related genes are concentrated in cluster 1 and cluster 2, which broadly consistent with the findings in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
+In particular, the periodic cell cycle related genes are concentrated in cluster $5$ and cluster $3$, which broadly consistent with the findings in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
 The number of clusters found by our method is much smaller than the number of clusters used in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
@@
 		\centering
-	\includegraphics[width=5.5in, height=3.3in]{wei/5_clusters.pdf}
+	\includegraphics[width=5.5in, height=4in]{wei/5_clusters.pdf}
 	\caption{Yeast data set--mean expression profiles of the 5 clusters found by correlation-corrected Gabriel method}

### 2016-04-19 New simulation/wei/Yeast Data Example.R

@@ -38,41 +38,35 @@ for(i in 1:nrow(Test)){
 		Test[i,j] = (Test[i,j]-Mean)/SD
 	}
 }
-################################################################################
-
+############## abandon above ########################################
+##Use processed data
+Yeast <- read.table("yeast3000.txt")
+Test <- as.matrix(Yeast[,2:16])
+#######################################
 set.seed(1)
 K.original <- cv.kmeans.gabriel(Test, 5, 2, maxcenters=40, classify.method="nearest")$centers
 K.original
-DATA_new <- Uncorrelate2(Test,K.original)
 
-K2<- cv.kmeans.gabriel(DATA_new, 5, 2, maxcenters=40, classify.method="nearest")$centers
-K2
+set.seed(0)
+Test2 <- Uncorrelate2(Test,K.original)
+cv.kmeans.gabriel(Test2, 5, 2, maxcenters=40, classify.method="nearest")$centers
 
-####gabriel_cor_correct(Test, maxcenters=40, type = 2)
-#set.seed(1)		
-#Gap <- clusGap(Test, FUN = kmeans, K.max =45)
-#which(Gap[[1]][,3] == max(Gap[[1]][,3]))
-			
-#set.seed(1)	
-#mcluster <- Mclust(Test, G = 20:40)
-#mcluster$G
 ###################################################################################						
 ###==============================================================
-Data <- as.data.frame(Test)
-Data$Gene = Gene
+Data <- Yeast
+names(Data)[1] = "Gene"
+Data$Gene = as.character(Data$Gene)
+Data$Gene <- as.character(lapply(Data$Gene,function(x)substr(x,1,7)))
 
-##set.seed(1)
-##Kfit <- kmeans(Data[,1:15],9, nstart = 100)
-set.seed(12)
-Kfit <- kmeans(DATA_new,5,nstart = 100)
+set.seed(1)
+Kfit <- kmeans(Test,5,nstart = 100)
 Data$Cluster <- Kfit$cluster
 
-Names_cluster <- c("Cluster 1; size=514","Cluster 2; size=583","Cluster 3; size=682","Cluster 4; size=563","Cluster 5; size=603")
[…]

### 2016-04-20 New simulation/technometrics.tex
Details about enrichment anlaysis to aid understanding of table 2

@@
 \cite{tavazoie1999systematic} applied $k$-means clustering on the preprocessed data with $k=30$ and discovered several periodic clusters.
-The preprocessing procedure includes choosing the most variable $2,945$ ORFs among the total $6,220$ by metric of variation based on the normalized dispersion across time points (s.d/mean).
+The preprocessing procedure includes choosing the most variable $2,945$ geness among the total $6,220$ by metric of variation based on the normalized dispersion across time points (s.d/mean).
 Data at time points $90$ and $100$ minutes were removed because of their less efficient labelling of mRNA during the original chip hybridization.
@@
 Similar as \cite{tavazoie1999systematic}, we found the members of each cluster to be significantly enriched for genes function in similar biological process.
-Each gene is mapped with the Gene Ontology Annotation in SGD database, which categorizes the genes into $103$ groups in terms of biological process.
-The $p$-value of observing $k$ genes from a particular category group within a cluster of size $m$ is given by:
+Each gene is mapped with the Gene Ontology (GO) Annotation in Saccharomyces Genome Database (SGD) database, which provides Gene Ontology (GO) annotation of the entire genome using GO slims, which are high-level subsets of GO terms that allow grouping of genes into broad categories, such as biological process, molecular function or cellular component.
+We choose to group genes in terms of biological process, which is a recognized series of events or molecular functions.
+
+To identify biologically meaningful gene clusters from spurious ones, \cite{tavazoie1999systematic} introduced a method, enrichment analysis, to assess the statistical significance of the over-represented gene class in a cluster using hypergeometric distribution.
+Specifically, the $p$-value of observing $k$ genes from a particular category group within a cluster of size $m$ is given by:
 \[	P = 1-\sum^{k-1}_{i=1} \frac{{{2945-n}\choose{m-i}}{{n}\choose{i}} }{ {{2945} \choose{m}} }\] 
-where $n$ is the total ORFs in that group. The enrichment of each category is shown in Table \ref{table:enrich}.
+where $n$ is the total genes in that group.
+It can be seen as testing the null hypothesis that genes are randomly  distributed among each cluster.
+The enrichment of each category is shown in Table \ref{table:enrich}.
 
@@
 \captionsetup{justification=centering}
-\caption{\label{table:enrich} Enrichment of clusters for ORFs with biological process categories}
+\caption{\label{table:enrich} Enrichment of clusters for genes within biological process categories}
 \scalebox{0.8}{%
@@
 \hline
-   Cluster & Cluster size($m$) & Biological process category(total ORFs) & ORFs within category($k$) &$p$-value \\ \hline
+   Cluster & Cluster size($m$) & Biological process category(total genes) & genes within category($k$) &$p$-value \\ \hline
  $1$ & $550$  &  response to oxidative stress ($55$)  &    $24$  & $1.5\times10^{-5}$ \\

### 2016-04-23 New simulation/technometrics.tex
Add confusion matrix

@@
 \usepackage{subfigure}
+\usepackage{threeparttable}
 %\usepackage[margin=2.5cm]{geometry}
@@
 These cyclic gene clusters are also reported in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
-In fact, the mean expression profile plot of cluster $5$ resembles those of cluster $2$, $14$ of \cite{tavazoie1999systematic} and cluster $2$, $3$ of \cite{dortet2008model}, while cluster $3$ corresponding to cluster $7$ of \cite{tavazoie1999systematic}.
 
@@
 Genes encode ribosomes are also overly presented in cluster $4$.
-Those results are not reported in previous studies, meaning our analysis leads to new discovery of gene grouping.
+
+Our results are well align with the results in \cite{tavazoie1999systematic}, which clustered the data with $k=30$.
+We prepared a confusion matrix (Table \ref{table:confusion}) to compare our results with those in \cite{tavazoie1999systematic}.
+In specific, \cite{tavazoie1999systematic} profiled three periodic clusters: cluster $2$, cluster $7$ and cluster $14$, which corresponding to our cluster $3$ and $5$.
+Indeed, we can easily see from Table \ref{table:confusion} that their cluster $2$ is entirely contained in our cluster $5$, their cluster $7$ is contained in our cluster $3$ and the periodic cluster $14$ is divided into our cluster $3$ and $5$.
+So their periodic clusters coincide with ours.
+
+In addition, they also profiled several non-periodic clusters.
+Among those, their most notable functional grouping cluster $1$ is well contained inside our cluster $4$, their cluster $3$ is also in our cluster $3$ and their cluster $8$ is contained in our cluster $2$
 
 In conclusion, functional grouping of gene occurred in the $5$ clusters found by our proposed method.
-In particular, the periodic cell cycle related genes are concentrated in cluster $5$ and cluster $3$, which broadly consistent with the findings in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
+In particular, the periodic cell cycle related genes are concentrated in cluster $5$ and cluster $3$, which consistent with the findings in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
 The number of clusters found by our method is much smaller than the number of clusters used in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
@@
 	\end{figure}
+
+\begin{table}[H]
+\begin{center}
+\captionsetup{justification=centering}
+\caption{\label{table:confusion} Confusion matrix comparing $5$ clusters to $30$ clusters }
+\begin{threeparttable}
+\begin{tabular}{lccccccc}
+\hline
+   &  Cluster $1$  & Cluster $2$   &  Cluster $3$  &  Cluster $4$ &  Cluster $7$  & Cluster $8$  & Cluster $14$ \\ \hline 
+ Cluster $1$  & $0$  &  $1$  & $0$  &  $0$  & $1$  & $3$  & $0$ \\
+ Cluster $2$  & $0$  &  $0$  & $0$  & $102$ & $10$ & $145$& $1$ \\
+ Cluster $3$  & $1$  &  $0$  & $91$ &  $2$  & $83$ & $0$  & $29$ \\
+ Cluster $4$  & $161$&  $0$  & $11$ &  $66$ & $7$  & $0$  & $6$ \\
+ Cluster $5$  & $2$  & $185$ & $2$  &  $0$  & $0$  & $0$  & $38$ \\ \hline      
+\end{tabular}
+ \begin{tablenotes}[para,flushleft]
+\footnotesize {Only the profiled clusters in \cite{tavazoie1999systematic} are listed here}.
+  \end{tablenotes}
+\end{threeparttable}
+\end{center}
+\end{table}
 

### 2016-04-27 cvclust.tex
revise general null proof

@@
 \newcommand{\bmuY}{\bar \mu^{Y}}
+\newcommand{\hmuY}{\hat \mu^{Y}}
 
@@
 
-\section{Analysis of Gabriel cross-validation with Gaussian noise}
-
-\subsection{Single Cluster in two dimensions}
-
-Now we we analyze the asymptotic performance of Gabriel Cross-Validation, in
-the case of Gaussian noise. Our main result is that with single-cluster
-Gaussian data in two-dimensions, if the predictor and response columns of $\dataX$ are weakly
-correlated or independent, then the method will correctly prefer $k = 1$
-clusters. In section~\ref{} we will generalize this result to 
-cases where $\dataX$ has more than two columns.
+\section{Analysis under Gaussian noise}
+
+\subsection{Single cluster in two dimensions}
+
+Theorem~\ref{thm:self-consistency} tells us that the Gabriel cross-validation
+method recovers the true number of clusters then the noise is negligible.
+While this result gives us some assurance that the procedure is well-behaved,
+we can bolster our confidence and gain insight into its workings by analyzing
+its behavior in the presence of noise.  We first study the case of a single
+cluster in two dimensions with correlated Gaussian noise.
 
@@
 marginal variances and correlation $\rho$.  In this case, the data are drawn
-from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 0.5$,
-then $\CV(1) < \CV(k)$, $k > 1$ with probability tending to one as $m$ and $n$ increase.
+from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 1/2$,
+and $k > 1$, then $\CV(1) < \CV(k)$ with probability tending to one as $m$
+and $n$ increase.
 \end{theorem}
@@
 \begin{proof}
-Given $(X_1, Y_1), \dotsc, (X_n, Y_n)$, let $A_k = \{a_1,a_2,...,a_k\}$
-denotes the set of cluster centers from applying $k$-means to $\{
-Y_i\}_{i=1}^{n}$ with parameter $k$. Because $\{ Y_i\}_{i=1}^{n}$ is
-symmetric, it's easy to see that the cluster centers are symmetric around $0$,
-i.e. if $a^* \in A_k$ then $-a^* \in A_k$. The $k$ clusters are symmetric in
-the same sense. Also, if $k$ is odd, then only one cluster contains both
-negative and positive $Y$ and it's symmetric with center $0$. The rest have
-pure positive or negative intervals; if $k$ is even, then all clusters have
-pure positive or negative intervals.
-
-Let $a_i$ denotes a cluster center with interval $[b_i, c_i]$. By above
-argument, either $0 \leq b_i < c_i$ ($b_i < c_i \leq 0$) or $b_i = -c_i$ which
-corresponding to $a_i=0$ with $k$ be odd. By symmetry, it's sufficient to only
-consider case $0 \leq b_i < c_i$ (with $a_i > 0 $) and $b_i = -c_i$ (with $a_i
-= 0 $) with $\rho \geq 0$.
-
-From joint normality of $X$ and $Y$, it follows that $X \mid Y$ is normal with
-mean $\rho Y$. Therefore, the mean value $\bar{X}_i$ of $X$ in cluster with
-$Y$ center $a_i$ is
-
+
+Throughout the proof we will assume that $\rho \geq 0$; a similar argument
+holds with minor modification when $\rho < 0$.
+
+Set $\hGY_1, \dotsc, \hGY_n$ to be the cluster labels gotten from applying
+$k$-means to $Y_1, \dotsc, Y_n$. Denote the cluster means by $\bmuY_1 \leq
+\bmuY_2 \leq \dotsb \leq \bmuY_k$.  Pollard's \citeyearpar{pollard1981strong}
+strong consistency theorem for $k$-means implies that for large $n$, the
+cluster centers are close to population clusters centers $a_1 < a_2 < \dotsb <
+a_k$. Specifically, $\bmuY_j = a_j + \OhP(n^{-1/2})$.  Since the distribution
+of $Y$ is symmetric, the population centers $a_1, a_2, \dotsc, a_k$ are
+symmetric about the origin.
+
+For $j$ in $1, \dotsc, k$, set
+\[
+  \bmuX_j = \frac{\sum_{i=1}^{n} 1\{ \hGY_i = j \} X_i}
+                 {\sum_{i=1}^{n} 1\{ \hGY_i = j \}}.
+\]
+The classification rule $\hGX$ is defined by
+\(
+  \hGX(X) = \argmin_j \| \bmuX_j - X \|.
+\)
+Denote the boundaries between the population clusters as
+$b_j = (a_j + a_{j+1})/2$ for $j = 1, \dotsc, k-1$.
+Set $b_0 = -\infty$ and $b_k = +\infty$. 
+Then, $\bmuX_j$ is within $\OhP(n^{-1/2})$ of the following expectation:
 \begin{align*}
-	\bar{X}_i &= E[X \mid b_i \leq Y \leq c_i] \\
-			&= E[ E(X \mid Y) \mid b_i \leq Y \leq c_i] \\
-			&= E[\rho Y \mid b_i \leq Y \leq c_i] \\
-			&= \rho E[ Y \mid b_i \leq Y \leq c_i] \\
-			&= \rho a_i
+  \E(X \mid b_j \leq Y \leq b_{j+1})
+    &= \E\{ E(X \mid Y) \mid b_j \leq Y \leq b_{j+1}\} \\
+    &= \E\{ \rho Y \mid b_j \leq Y \leq b_{j+1}\} \\
+    &= \rho \E\{ Y \mid b_j \leq Y \leq b_{j+1}\} \\
+    &= \rho a_j.
 \end{align*}
-
-That is, the mean values of $X$ in the $k$ clusters is $\{\rho a_1,\rho
-a_2,...,\rho a_k\}$. Because we assign each observation to the closest
-$\bar{X}_i$, $i=1,2,...,k$, it's easy to see the corresponding interval on $X$
-for $\bar{X}_i = \rho a_i$ is $[\frac{\rho \left( a_{i-1}+a_i\right)}{2},
-\frac{\rho \left( a_{i}+a_{i+1}\right)}{2}]$, where $a_{i-1}$ and $a_{i+1}$
-are the two adjacent centers to $a_i$ on $Y$. Note that $[\frac{ \left(
-a_{i-1}+a_i\right)}{2}, \frac{\left( a_{i}+a_{i+1}\right)}{2}] = [b_i, c_i]$
-by the algorithm definition of $k$-means. The $Y$ center for such interval on
-$X$ is
-
+That is, $\bmuX_j = \rho a_j + \OhP(n^{-1/2})$. For $j = 1, \dotsc, k-1$,
+the boundary between sample the classification based on $X$ to
+labels $j$ and $j+1$ is
+$(\bmuX_j + \bmuX_{j+1}) /2 = \rho b_j + \OhP(n^{-1/2})$.
+
+Set $\hGX_i = \hGX(X_i)$.
+The cross-validation error is
 \begin{align*}
-	\hat{\mu}^Y_i &= E[Y \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
-	 &= E[ E(Y \mid X) \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
-	 &= E[ \rho X \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
-	 & =  \rho E[X \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
-	  & =  \rho E[X \mid \rho b_i \leq X \leq \rho c_i] \\
-	  &\leq \rho a_i
+  \CV(k)
+  &= \frac{1}{m} \sum_{i=n+1}^{n+m} \|Y_i - \hat Y_i\|^2
+  \\
+  &= \frac{1}{m} \sum_{i=n+1}^{n+m}
+  \sum_{j=1}^{k} \|Y_i - \bmuY_j\|^2 1\{ \hGX_i = j \}
+  \\
+  &=
+  \sum_{j=1}^{k}
+    \pi_j
+    \E[ \|Y - a_j\|^2 \mid \rho b_j < X < \rho b_{j+1}]
+  + \OhP(n^{-1/2})
+  + \OhP(m^{-1/2}),
 \end{align*}
-
-where the equality holds if $b_i = -c_i$ or $\rho = 1$ on the last step. This is because
-
+where $\pi_j = \Pr(\rho b_j < X < \rho b_{j+1}).$
+
+For $j = 1, \dotsc, k$,
+set
+\(
+  \hmuY_j =
+  \E[ Y  \mid \rho b_j < X < \rho b_{j+1}].
+\)
+Note that
+\[
+  \CV(1)
+    = 
+  \sum_{j=1}^{k}
+    \pi_j
+    \E[ \|Y - 0\|^2 \mid \rho b_j < X < \rho b_{j+1}]
+  + \OhP(n^{-1/2})
+  + \OhP(m^{-1/2}).
+\]
+Thus, the difference in cross-validation errors is
 \begin{align*}
-	 E[X \mid \rho b_i \leq X \leq \rho c_i] &= E[Y \mid \rho b_i \leq Y \leq \rho c_i] \\
-	\mathtt{0 \leq b_i < c_i}, \hspace{0.05in} \mathtt{0 \leq \rho \leq 1} \hspace{0.1in} & \leq E[Y \mid b_i \leq Y \leq c_i] \\
-	&= a_i
+\CV(k) - \CV(1)
+  &=
+  \sum_{j=1}^{k}
+    \pi_j
+    a_j
+    (a_j - 2 \hmuY_j)
+  + \OhP(n^{-1/2})
+  + \OhP(m^{-1/2}).
 \end{align*}
-
-The first equation above is because the same marginal distribution of $X$ and $Y$.
-
-Note that Gabriel CV method predicts $Y$ to be $a_i$ for $X \in [\frac{\rho
-\left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}]$,
-and if $k=1$ the predicted $Y = 0$ for all $X$. Hence, to see if $CV(k) >
-CV(1)$ for  $X \in [\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho
-\left( a_{i}+a_{i+1}\right)}{2}]$, one only need to see if $ a_i-\hat{\mu}^Y_i
-> \hat{\mu}^Y_i - 0$. Since $ \hat{\mu}^Y_i \leq \rho a_i$, it's clear that
-$CV(k) > CV(1)$ whenever $\rho < 0.5$. This is true to all segment
-$[\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left(
-a_{i}+a_{i+1}\right)}{2}] = [\rho b_i, \rho c_i]$ except the one with $b_i =
--c_i$, in which $CV(k) = CV(1)$. Since $k>1$ and at most one segment has
-$CV(k) = CV(1)$ while the rest have $CV(k) > CV(1)$ if $\rho < 0.5$, it is
-clear that the overall prediction error $CV(k) > CV(1)$. By symmetry, we can
-see $CV(k) > CV(1)$ if $|\rho| < 0.5$. The proof is complete.
+For arbitrary $j$,
+\begin{align*}
+  \hmuY_j &=
+  \E[ E(Y\mid X)  \mid \rho b_j < X < \rho b_{j+1}]
+\\
+  &= \rho \E[ X  \mid \rho b_j < X < \rho b_{j+1}].
+\end{align*}
+Since $0 \leq \rho \leq 1$,
+in cases where $0 \leq b_j < b_{j+1}$, we have that
+$\hmuY_j \leq \rho a_j$; similarly, when $b_j < b_{j+1} \leq 0$, it follows
+that $\hmuY_j \geq \rho a_j$. In either of these two situations, if $\rho <
+1/2$, then
+\[
+  a_j (a_j - 2 \hmuY_j) > 0.
+\]
+The last situation to consider is when $b_j < 0 < b_{j+1}$, in which case
+$b_j = - b_{j+1}$; here, $\hmuY_j = a_j = 0$.
+Putting this all together, we have that as $n$ and $m$ tend to infinity,
+the probability that
+\(
+  \CV(k) > \CV(1)
+\)
+tends to one.
 \end{proof}

### 2016-04-28 New simulation/technometrics.tex
Add confusion plots in Data example

@@
 \usepackage{longtable}
-
+\usepackage{rotating} % for landscape table
 %\def\threedigits#1{%
@@
 This numbers is much smaller compare to those of \cite{tavazoie1999systematic} and \cite{dortet2008model}.
-We will focus our analysis on this $5$ clusters found by the proposed method, whose result is summarized in Figure \ref{fig:5_clusters}, where each cluster is presented by the mean expression level of all the genes within that cluster over $15$ time points.
+We will focus our analysis on this $5$ clusters found by the proposed method, whose result is summarized on the top panel of Figure \ref{fig:5_clusters}, where each cluster is presented by the mean expression level of all the genes within that cluster over $15$ time points.
 Similar as \cite{tavazoie1999systematic}, we found the members of each cluster to be significantly enriched for genes function in similar biological process.
@@
 We choose to group genes in terms of biological process, which is a recognized series of events or molecular functions.
+There are total $103$ categories for biological process.
 
@@
 It can be seen as testing the null hypothesis that genes are randomly  distributed among each cluster.
-The enrichment of each category is shown in Table \ref{table:enrich}.
+The categories with the most significant grouping of gene inside each of the $5$ clusters are shown in Table \ref{table:enrich}.
+
+To better compare our results with those of \cite{tavazoie1999systematic}, we also prepared confusion Table \ref{table:confusion} and Figure \ref{fig:5_clusters}.
+The number in cell $(i,j)$ of Table \ref{table:confusion} shows the number of observations that belong to both the $i$th cluster of ours and the $j$th cluster of \cite{tavazoie1999systematic}.
+The plots in Figure \ref{fig:5_clusters} have similar meaning -- the mini plot on first row and the $4$th column is the mean gene expression level of observations belong to both the first cluster of \cite{tavazoie1999systematic} and the $4$th cluster of ours.
+We only plot the mean expression level if the number of observations in that cell is greater than $20$.
+Although \cite{tavazoie1999systematic} found $30$ clusters, not all of them were discussed in \cite{tavazoie1999systematic}.
+We only compare our results with those clusters that were profiled in \cite{tavazoie1999systematic}.
+The complete confusion matrix is in Appendix \ref{app:confusion}.
+The original result of \cite{tavazoie1999systematic} is obtained from site \url{http://arep.med.harvard.edu/network_discovery/}.
 
@@
 
-It turns out that each cluster contains genes that tend to participate in the common processes.
-It is easy to see that cluster $5$ and $3$ are periodic from Figure \ref{fig:5_clusters}.
+It turns out that each cluster has unique trend of expression level and contains genes that tend to participate in the common processes.
+Cluster $1$ has decreasing expression level with time.
+From Table \ref{table:enrich}, we can see it concentrates genes that somatize the cell stress, such as oxidative heat-induce proteins.
+It mainly consists of observations that did not get profiled in \cite{tavazoie1999systematic}, as we can see from Table \ref{table:confusion} and Figure \ref{fig:5_clusters}.
+
+The mean gene expression level in Cluster $2$ decreases at the beginning and then increase.
+Our analysis found genes govern mitochondrial translation and mitochondrion organization are heavily concentrated in cluster $2$.
+Genes in \cite{tavazoie1999systematic}'s cluster $4$ and $8$ are heavily concentrated in this cluster, which also explains the coincidence between the enriched categories in our cluster $2$ and the functional groups in cluster $4$ and $8$ of    \cite{tavazoie1999systematic}.
+
+Our cluster $3$ is a periodic cluster where one can see two gene expression cycles from Figure \ref{fig:5_clusters}.
+Indeed, cell cycle genes that related to budding and cell polarity are commonly found in cluster $3$.
+In addition, one can see from Table \ref{table:enrich} that genes govern RNA processing and transcription are highly represented in this cluster.
+Comparing cluster $3$ to the clusters found in \cite{tavazoie1999systematic} from Table \ref{table:confusion} and Figure \ref{fig:5_clusters}, one can see that it mainly consists of genes from cluster $3$, $7$ and $14$ of \cite{tavazoie1999systematic}.
+Note that the cluster $7$ and $14$ are both periodic clusters highlighted in \cite{tavazoie1999systematic}.
+
+Cluster $4$ has increasing expression level with time.
+From Table \ref{table:enrich}, we can see genes related to cytoplasmic translation are highlighted in this cluster.
+Genes encode ribosomes are also overly presented in cluster $4$.
+This cluster corresponding to cluster $1$, the most notable functional grouping cluster of \cite{tavazoie1999systematic}.
+Indeed, one can see from Table \ref{table:confusion} that almost entire cluster $1$ of  \cite{tavazoie1999systematic} is contained in our cluster $4$.
+
+One can easily see from Figure \ref{fig:5_clusters} that our cluster $5$ is also a periodic cluster.
+Enrichment analysis shows that genes participated in the cell-cycle phase related processes such as DNA replication and DNA repair are highly concentrated in this cluster.
+Table \ref{table:confusion} shows that the entire cluster $2$ of \cite{tavazoie1999systematic} is contained in our cluster $5$, along with some genes from cluster $14$.
+Note that both cluster $2$ and $14$ are profiled periodic clusters in \cite{tavazoie1999systematic}.
+We can also see from Figure \ref{fig:5_clusters} that our cluster $5$ groups together genes with similar cycle phase in \cite{tavazoie1999systematic}'s cluster $2$ and $14$.
+
+In conclusion, our clusters group together genes that participate in the common biological processes.
+Our results are broadly align with the results in \cite{tavazoie1999systematic}, which clustered the data with $k=30$.
+In particular, the periodic cell cycle related genes are concentrated in cluster $5$ and cluster $3$, which consistent with the findings in \cite{tavazoie1999systematic} (cluster $2$, $7$ and $14$).
 Indeed, more than $70\%$ genes which function in mitotic cell cycle process are concentrated in these two clusters.
-Genes participated in the cell-cycle phase related processes such as DNA replication and DNA repair are highly concentrated in cluster $5$.
-Genes related to budding and cell polarity, as well as chromatin organization and transcription are commonly found in cluster $3$.
-These cyclic gene clusters are also reported in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
-
-Cluster $1$ concentrates genes that somatize the cell stress, such as oxidative heat-induce proteins. This cluster corresponding to the cluster $5$ and $11$ in \cite{dortet2008model}.
-Our analysis also found genes govern mitochondrial translation and mitochondrion organization are heavily concentrated in cluster $2$, while genes related to cytoplasmic translation are highlighted in cluster $4$.
-Genes encode ribosomes are also overly presented in cluster $4$.
-
-Our results are well align with the results in \cite{tavazoie1999systematic}, which clustered the data with $k=30$.
-We prepared a confusion matrix (Table \ref{table:confusion}) to compare our results with those in \cite{tavazoie1999systematic}.
-In specific, \cite{tavazoie1999systematic} profiled three periodic clusters: cluster $2$, cluster $7$ and cluster $14$, which corresponding to our cluster $3$ and $5$.
-Indeed, we can easily see from Table \ref{table:confusion} that their cluster $2$ is entirely contained in our cluster $5$, their cluster $7$ is contained in our cluster $3$ and the periodic cluster $14$ is divided into our cluster $3$ and $5$.
-So their periodic clusters coincide with ours.
-
-In addition, they also profiled several non-periodic clusters.
-Among those, their most notable functional grouping cluster $1$ is well contained inside our cluster $4$, their cluster $3$ is also in our cluster $3$ and their cluster $8$ is contained in our cluster $2$
-
-In conclusion, functional grouping of gene occurred in the $5$ clusters found by our proposed method.
-In particular, the periodic cell cycle related genes are concentrated in cluster $5$ and cluster $3$, which consistent with the findings in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
 The number of clusters found by our method is much smaller than the number of clusters used in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
 However, both studies found that most of their clusters have no characteristics, suggesting overestimate of the underlying diversity of biological expression classes in the data set \citep{tavazoie1999systematic}.
-Our result is more compact, while still able to identify the functional grouping of genes in clusters highlighted in previous studies.
+From Figure \ref{fig:5_clusters}, one can see clusters in \cite{tavazoie1999systematic} with similar behavior are merged in our cluster, which leads to smaller number of clusters.
+Our result is more parsimonious, while still able to identify the functional grouping of genes in clusters highlighted in previous studies.
 
@@
 		\centering
-	\includegraphics[width=5.5in, height=4in]{wei/5_clusters.pdf}
-	\caption{Yeast data set--mean expression profiles of the 5 clusters found by correlation-corrected Gabriel method}
+	\includegraphics[width=6.2in, height=6.2in]{wei/split_plots.pdf}
+	\caption{Yeast data set mean expression profiles. The $5$ clusters found by Gabriel method are on the top; Clusters profiled in \cite{tavazoie1999systematic} are on the right.}
 	\label{fig:5_clusters}
@@
 
-\begin{table}[H]
+\begin{table}[t]
 \begin{center}
@@
 \end{align*}
+
+\section{Confusion matrix} \label{app:confusion}
+\begin{sidewaystable}
+\captionsetup{justification=centering}
+\caption{\label{table:confusion-complete} Complete confusion matrix comparing $5$ clusters to $30$ clusters }
+\scalebox{0.67}{%
+\begin{tabular}{lcccccccccccccccccccccccccccccc|r}
+\hline
+   &  C$1$  & C$2$   &  C$3$  &  C$4$ & C$5$ &  C$6$ &  C$7$  & C$8$ & C$9$ & C$10$ &  C$11$& C$12$ & C$13$ & C$14$ & C$15$ & C$16$& C$17$& C$18$& C$19$& C$20$& C$21$& C$22$& C$23$& C$24$& C$25$& C$26$& C$27$& C$28$& C$29$& C$30$& Total \\ \hline
+ Cluster $1$  & $0$  &  $1$  & $0$  &  $0$  & $152$& $1$  & $1$  & $3$  & $142$& $12$ & $0$  & $3$  & $0$ & $0$ & $60$ & $7$& $9$& $3$& $62$& $11$& $1$& $33$& $2$& $1$& $2$& $5$& $9$& $0$& $27$& $3$& $550$\\
+ Cluster $2$  & $0$  &  $0$  & $0$  & $102$ & $0$  & $2$  & $10$ & $145$& $0$ & $27$  & $5$  & $68$ & $1$ & $1$ & $53$ & $0$& $0$& $16$& $2$& $0$& $13$& $15$& $18$& $38$& $57$& $3$& $0$& $7$& $2$& $5$& $590$\\
+ Cluster $3$  & $1$  &  $0$  & $91$ &  $2$  & $0$  & $12$  & $83$ & $0$ & $0$ & $24$  & $81$ & $5$  & $12$ & $29$& $2$ & $0$& $72$& $10$& $0$& $6$& $6$& $28$& $2$& $45$& $16$& $23$& $30$& $23$& $10$& $41$& $654$\\
+ Cluster $4$  & $161$&  $0$  & $11$ &  $66$ & $0$  & $85$  & $7$  & $0$ & $0$ & $20$  & $8$  & $3$  & $81$ & $6$ & $0$ & $9$& $1$& $67$& $1$& $2$& $44$& $9$& $25$& $1$& $1$& $19$& $5$& $0$& $1$& $1$& $634$\\
+ Cluster $5$  & $2$  & $185$ & $2$  &  $0$  & $0$  & $4$  & $0$  & $0$  & $5$ & $6$   & $0$  & $1$  & $5$  & $38$& $0$ & $83$& $1$& $5$& $8$& $65$& $6$& $0$& $22$& $0$& $0$& $0$& $20$& $38$& $11$& $10$& $517$ \\ \hline
+ Total   & $164$& $186$ & $104$ & $170$ & $152$ & $104$ & $101$ & $148$ & $147$&$89$ & $94$ & $80$  & $99$ & $74$& $115$ & $99$& $83$& $101$& $73$& $84$& $70$& $85$& $69$& $85$& $76$& $50$& $64$& $68$& $51$& $60$& $2945$    \\ \hline
+\end{tabular}
+}
+\end{sidewaystable}	
 \clearpage

### 2016-04-28 New simulation/wei/confusion_table.R

@@ -1,0 +1,159 @@
+Tava <- read.csv("Tava.csv", header= T)
+Tava2 <- Tava[,c(1,3)]
+#Tava2[,2] <- as.character(lapply(Tava2[,2],function(x)substr(x,1,7)))
+Tava2[,2] <- as.character(lapply(Tava2[,2],function(x)toupper(x)))
+names(Tava2)[2]<- c("Gene")
+
+Yeast <- read.table("yeast3000.txt")
+Data <- Yeast
+Data[,1] <- as.character(lapply(Data[,1],function(x)toupper(x)))
+
+set.seed(1)
+Kfit <- kmeans(Data[,2:16],5,nstart = 100)
+Data$Gabriel <- Kfit$cluster
+names(Data)[1] <- "Gene"
+
+##merge two data
+
+DATA <- cbind(Data,Tava2)
+
+##Get 5X30 confusion matrix
+Confusion <- matrix(NA,5,30)
+
+for(i in 1:5){
+	Subset <- DATA[DATA$Gabriel==i,];
+	for(j in 1:30){
+		Confusion[i,j] = sum(Subset$Cluster==j)
+	}
+}
+write.csv(Confusion,"Confusion_matrix.csv")
+
+###=============split plots==========
+DATA$Tava <- NA
+
+for(i in 1:nrow(DATA)){
+	if( DATA[i,"Cluster"] == 1 ){DATA[i,"Tava"] = "1"}
+	if( DATA[i,"Cluster"] == 2 ){DATA[i,"Tava"] = "2"}
+	if( DATA[i,"Cluster"] == 3 ){DATA[i,"Tava"] = "3"}
+	if( DATA[i,"Cluster"] == 4 ){DATA[i,"Tava"] = "4"}
+	if( DATA[i,"Cluster"] == 7 ){DATA[i,"Tava"] = "7"}
+	if( DATA[i,"Cluster"] == 8 ){DATA[i,"Tava"] = "8"}
+	if( DATA[i,"Cluster"] == 14 ){DATA[i,"Tava"] = "14"}
+	if( !(DATA[i,"Cluster"] %in% c(1:4,7,8,14)) ){DATA[i,"Tava"] = "Other"}
+}
+
+##==================for top plot================================
+##function SE
+SE <- function(x){sd(x)/sqrt(length(x))}
+
+for(i in 1:5){
+ 	Subset <- DATA[DATA$Gabriel==i,];
[…]

### 2016-04-29 New simulation/technometrics.tex
further revision of yeast data application

@@
 \subsection{Application on Yeast data set}
-The yeast Saccharomyces cerevisiae data set was collected by \cite{cho1998genome} to study the genome-wide characterization of mRNA transcript levels during the cell cycles of the budding yeast S.
+The yeast Saccharomyces cerevisiae data set was collected by \cite{cho1998genome} to study the characterization of the cell cycle related processes of the budding yeast S.
 cerevisiae.
-To obtain data,
-cdc28-13 cells were synchronized in late G1 phase by heat treatment, and the cell cycle was reinitiated by shifting cells to cooler temperature.
-Data was collected at $17$ time points at every $10$ minutes intervals covering almost $2$ complete cell cycles.
+To obtain data, cdc28 cells were synchronized by raising temperature to $37^\circ$C in late G1 phase, and then reinitiated by changing cells to cooler environment ($25^\circ$C).
+Data was collected at $17$ time points at every $10$ minutes interval covering almost $2$ complete cell cycles.
 Oligonucleotide microarrays was used to query $6,220$ gene expression profiles at these $17$ time points.
-This data has been the subject of several studies and is generally accepted as a reference \citep{dortet2008model}.
+Besides the original paper, several other studies have analyzed this data.
 
 \cite{tavazoie1999systematic} applied $k$-means clustering on the preprocessed data with $k=30$ and discovered several periodic clusters.
-The preprocessing procedure includes choosing the most variable $2,945$ geness among the total $6,220$ by metric of variation based on the normalized dispersion across time points (s.d/mean).
-Data at time points $90$ and $100$ minutes were removed because of their less efficient labelling of mRNA during the original chip hybridization.
+The preprocessing procedure includes choosing the most variable $2,945$ geness among the total $6,220$.
+The normalized dispersion across time points (s.d/mean) was used as the metric of variation for selection.
+Data at time points $90$ and $100$ minutes were removed because they were deemed unreliable.
 It results a $2,945$ by $15$ data matrix.
-The last step of the preprocessing procedure is to variance normalize each gene across the $15$ time points, which is done by subtracting the mean and then dividing by the standard deviation.
+The last step of the preprocessing procedure was to variance normalize each gene across the $15$ time points, which is done by subtracting the mean and then dividing by the standard deviation.
 The result is each gene has mean zero and standard deviation equal to one across the time points.
@@
 The algorithm found $5$ clusters.
+Note here we cluster the genes as observations into clusters and the $15$ time points are considered attributes of each gene.
 Since the dimensions (time points) are likely to be correlated, we also applied the correlation-corrected version of Gabriel method on the data which also result $5$ clusters.
 This numbers is much smaller compare to those of \cite{tavazoie1999systematic} and \cite{dortet2008model}.
-We will focus our analysis on this $5$ clusters found by the proposed method, whose result is summarized on the top panel of Figure \ref{fig:5_clusters}, where each cluster is presented by the mean expression level of all the genes within that cluster over $15$ time points.
+We will focus our analysis on the $5$ clusters found by the proposed method, whose result is summarized on the bottom panel of Figure \ref{fig:5_clusters}, where each cluster is presented by the mean expression level of all the genes within that cluster over $15$ time points.
 Similar as \cite{tavazoie1999systematic}, we found the members of each cluster to be significantly enriched for genes function in similar biological process.
-Each gene is mapped with the Gene Ontology (GO) Annotation in Saccharomyces Genome Database (SGD) database, which provides Gene Ontology (GO) annotation of the entire genome using GO slims, which are high-level subsets of GO terms that allow grouping of genes into broad categories, such as biological process, molecular function or cellular component.
+Each gene is mapped with the Gene Ontology (GO) Annotation in Saccharomyces Genome Database (SGD) database, which provides Gene Ontology (GO) annotation of the entire genome using GO slims.
+GO slims are subsets of GO terms that give broad overview of ontology content and allow grouping of genes into broad categories, such as biological process, molecular function or cellular component.
 We choose to group genes in terms of biological process, which is a recognized series of events or molecular functions.
@@
 The categories with the most significant grouping of gene inside each of the $5$ clusters are shown in Table \ref{table:enrich}.
+Only $p$-values less than $4.8\times10^{-4}$ ($0.05/103$) are reported here.
 
 To better compare our results with those of \cite{tavazoie1999systematic}, we also prepared confusion Table \ref{table:confusion} and Figure \ref{fig:5_clusters}.
-The number in cell $(i,j)$ of Table \ref{table:confusion} shows the number of observations that belong to both the $i$th cluster of ours and the $j$th cluster of \cite{tavazoie1999systematic}.
-The plots in Figure \ref{fig:5_clusters} have similar meaning -- the mini plot on first row and the $4$th column is the mean gene expression level of observations belong to both the first cluster of \cite{tavazoie1999systematic} and the $4$th cluster of ours.
+The number in cell $(i,j)$ of Table \ref{table:confusion} shows the number of observations that belong to both the $i$th cluster of \cite{tavazoie1999systematic}'s and the $j$th cluster of ours.
+The plots in Figure \ref{fig:5_clusters} have similar meaning -- the mini plot on first row and the $4$th column is the mean gene expression level of observations belong to both the first cluster of \cite{tavazoie1999systematic}'s and the $4$th cluster of ours.
 We only plot the mean expression level if the number of observations in that cell is greater than $20$.
@@
 		\centering
-	\includegraphics[width=6.2in, height=6.2in]{wei/split_plots.pdf}
-	\caption{Yeast data set mean expression profiles. The $5$ clusters found by Gabriel method are on the top; Clusters profiled in \cite{tavazoie1999systematic} are on the right.}
+	\includegraphics[width=5.7in, height=5.7in]{wei/split_plots.pdf}
+	\caption{Yeast data set mean expression profiles. The $5$ clusters found by Gabriel method are on the bottom; Clusters profiled in \cite{tavazoie1999systematic} are on the right.}
 	\label{fig:5_clusters}
@@
 
-\begin{table}[t]
-\begin{center}
-\captionsetup{justification=centering}
-\caption{\label{table:confusion} Confusion matrix comparing $5$ clusters to $30$ clusters }
-\begin{threeparttable}
-\begin{tabular}{lccccccc}
+\begin{table}[H]
+\caption{\label{table:confusion} \small{Confusion matrix comparing $5$ clusters to clusters profiled in \cite{tavazoie1999systematic}}}
+\centering
+\begin{tabular}{lcccccr}
   \hline
-   &  Cluster $1$  & Cluster $2$   &  Cluster $3$  &  Cluster $4$ &  Cluster $7$  & Cluster $8$  & Cluster $14$ \\ \hline 
- Cluster $1$  & $0$  &  $1$  & $0$  &  $0$  & $1$  & $3$  & $0$ \\
- Cluster $2$  & $0$  &  $0$  & $0$  & $102$ & $10$ & $145$& $1$ \\
- Cluster $3$  & $1$  &  $0$  & $91$ &  $2$  & $83$ & $0$  & $29$ \\
- Cluster $4$  & $161$&  $0$  & $11$ &  $66$ & $7$  & $0$  & $6$ \\
- Cluster $5$  & $2$  & $185$ & $2$  &  $0$  & $0$  & $0$  & $38$ \\ \hline      
+ & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
+  \hline
+Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
+  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
+  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
+  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
+  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
+  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
+  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
+  Other & 545 & 332 & 448 & 383 & 290 & 1998 \\ 
+  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ 
+   \hline
 \end{tabular}
- \begin{tablenotes}[para,flushleft]
-\footnotesize {Only the profiled clusters in \cite{tavazoie1999systematic} are listed here}.
-  \end{tablenotes}
-\end{threeparttable}
-\end{center}
 \end{table}
@@
 \section{Confusion matrix} \label{app:confusion}
-\begin{sidewaystable}
-\captionsetup{justification=centering}
+
+\begin{table}
+%\captionsetup{justification=centering}
 \caption{\label{table:confusion-complete} Complete confusion matrix comparing $5$ clusters to $30$ clusters }
-\scalebox{0.67}{%
-\begin{tabular}{lcccccccccccccccccccccccccccccc|r}
+\begin{tabular}{lccccc|c}
   \hline
-   &  C$1$  & C$2$   &  C$3$  &  C$4$ & C$5$ &  C$6$ &  C$7$  & C$8$ & C$9$ & C$10$ &  C$11$& C$12$ & C$13$ & C$14$ & C$15$ & C$16$& C$17$& C$18$& C$19$& C$20$& C$21$& C$22$& C$23$& C$24$& C$25$& C$26$& C$27$& C$28$& C$29$& C$30$& Total \\ \hline
- Cluster $1$  & $0$  &  $1$  & $0$  &  $0$  & $152$& $1$  & $1$  & $3$  & $142$& $12$ & $0$  & $3$  & $0$ & $0$ & $60$ & $7$& $9$& $3$& $62$& $11$& $1$& $33$& $2$& $1$& $2$& $5$& $9$& $0$& $27$& $3$& $550$\\
- Cluster $2$  & $0$  &  $0$  & $0$  & $102$ & $0$  & $2$  & $10$ & $145$& $0$ & $27$  & $5$  & $68$ & $1$ & $1$ & $53$ & $0$& $0$& $16$& $2$& $0$& $13$& $15$& $18$& $38$& $57$& $3$& $0$& $7$& $2$& $5$& $590$\\
- Cluster $3$  & $1$  &  $0$  & $91$ &  $2$  & $0$  & $12$  & $83$ & $0$ & $0$ & $24$  & $81$ & $5$  & $12$ & $29$& $2$ & $0$& $72$& $10$& $0$& $6$& $6$& $28$& $2$& $45$& $16$& $23$& $30$& $23$& $10$& $41$& $654$\\
- Cluster $4$  & $161$&  $0$  & $11$ &  $66$ & $0$  & $85$  & $7$  & $0$ & $0$ & $20$  & $8$  & $3$  & $81$ & $6$ & $0$ & $9$& $1$& $67$& $1$& $2$& $44$& $9$& $25$& $1$& $1$& $19$& $5$& $0$& $1$& $1$& $634$\\
- Cluster $5$  & $2$  & $185$ & $2$  &  $0$  & $0$  & $4$  & $0$  & $0$  & $5$ & $6$   & $0$  & $1$  & $5$  & $38$& $0$ & $83$& $1$& $5$& $8$& $65$& $6$& $0$& $22$& $0$& $0$& $0$& $20$& $38$& $11$& $10$& $517$ \\ \hline
- Total   & $164$& $186$ & $104$ & $170$ & $152$ & $104$ & $101$ & $148$ & $147$&$89$ & $94$ & $80$  & $99$ & $74$& $115$ & $99$& $83$& $101$& $73$& $84$& $70$& $85$& $69$& $85$& $76$& $50$& $64$& $68$& $51$& $60$& $2945$    \\ \hline
+ & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
+  \hline
+  Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
+  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
+  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
+  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
+  Cluster 5 & 152 & 0 & 0 & 0 & 0 & 152 \\ 
+  Cluster 6 & 1 & 2 & 12 & 85 & 4 & 104 \\ 
+  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
+  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
+  Cluster 9 & 142 & 0 & 0 & 0 & 5 & 147 \\ 
+  Cluster 10 & 12 & 27 & 24 & 20 & 6 & 89 \\ 
+  Cluster 11 & 0 & 5 & 81 & 8 & 0 & 94 \\ 
+  Cluster 12 & 3 & 68 & 5 & 3 & 1 & 80 \\ 
+  Cluster 13 & 0 & 1 & 12 & 81 & 5 & 99 \\ 
+  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
+  Cluster 15 & 60 & 53 & 2 & 0 & 0 & 115 \\ 
+  Cluster 16 & 7 & 0 & 0 & 9 & 83 & 99 \\ 
+  Cluster 17 & 9 & 0 & 72 & 1 & 1 & 83 \\ 
+  Cluster 18 & 3 & 16 & 10 & 67 & 5 & 101 \\ 
+  Cluster 19 & 62 & 2 & 0 & 1 & 8 & 73 \\ 
+  Cluster 20 & 11 & 0 & 6 & 2 & 65 & 84 \\ 
+  Cluster 21 & 1 & 13 & 6 & 44 & 6 & 70 \\ 
+  Cluster 22 & 33 & 15 & 28 & 9 & 0 & 85 \\ 
+  Cluster 23 & 2 & 18 & 2 & 25 & 22 & 69 \\ 
+  Cluster 24 & 1 & 38 & 45 & 1 & 0 & 85 \\ 
+  Cluster 25 & 2 & 57 & 16 & 1 & 0 & 76 \\ 
+  Cluster 26 & 5 & 3 & 23 & 19 & 0 & 50 \\ 
+  Cluster 27 & 9 & 0 & 30 & 5 & 20 & 64 \\ 
+  Cluster 28 & 0 & 7 & 23 & 0 & 38 & 68 \\ 
+  Cluster 29 & 27 & 2 & 10 & 1 & 11 & 51 \\ 
+  Cluster 30 & 3 & 5 & 41 & 1 & 10 & 60 \\ \hline
+  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ \hline
 \end{tabular}
-}
-\end{sidewaystable}	
+\end{table}
+
 \clearpage

### 2016-05-02 New simulation/technometrics.tex
more details on correlation correction procedure

@@
 A simply remedy for the high correlation is available if we assume common covariance structure among the $k$ clusters.
+The procedure consists of four steps as follows:
 \begin{enumerate}
@@
 \end{enumerate}
+Such correlation-correction idea is based on the assumption that each observation $x_i$ in $\dataX$ can be expressed as
+\[	x_i = \mu(G_i)+\varepsilon_i \]
+where $\mu(G_i)$ is the true cluster center for $x_i$ and $\varepsilon_i$ is noise with covariance matrix $\Sigma$.
+To reduce correlation between dimensions, we would like to transform $\Sigma$ into a diagonal matrix.
+A simple way to achieve this goal is multiplying $\dataX$ by $\Sigma^{-1/2}$, so the transformed $x_i$
+
+\[	\Sigma^{-1/2} x_i = \Sigma^{-1/2} \mu(G_i)+\Sigma^{-1/2} \varepsilon_i \]
+has identity covariance matrix $\mathbf{I}$.
+This is exactly what step $3$ does.
+However, before we can transform the data, we need to estimate the covariance matrix $\Sigma$.
+Step $1$ and $2$ estimate the covariance matrix by first applying Gabriel CV method on $\dataX$ to get preliminary estimate $\hat{k}$ and then estimate $\Sigma$ through
+
+\[	\hat{\Sigma} = \frac{1}{n-\hat{k}} \sum^n_{i=1} (x_i-\hat{\mu}_i)(x_i-\hat{\mu}_i)^T	\]
+where $\hat{\mu}_i$ is the center of the cluster which contains $x_i$ when applying $k$-means on $\dataX$ with parameter $\hat{k}$.
+
+In practice, $\hat{\Sigma}$ wouldn't be the efficient estimator of $\Sigma$ since step $1$ may yield $\hat{k} > k$.
+Nevertheless, above procedure works well with correlated noise as we shall see in the simulation section.
 

### 2016-06-14 cvclust.tex
revise through sec 4.1

@@
 with the property that items in the same group are more similar to each other
-than items in different groups \citep{hartigan1975clustering}.  These
+than items in different groups \citep{hartigan1975clustering}.  Such
 procedures are used in two main applicaitons: (a) exploratory analysis, where clusters
-reveal homogeneous sub-groups within a large sample \cite{TODO}; (b) data
+reveal homogeneous sub-groups within a large sample; (b) data
 reduction, where high-dimensional item attribute vectors get reduced discrete
@@
 within-cluster dispersion relative to what is expected under a reference
-distribution \citep{tibshirani2001estimating}. Similar methods include those
-developed by \citet{TODO}.
+distribution \citep{tibshirani2001estimating}.
+%Similar methods include those developed by \citet{TODO}.
 
 Other authors have proposed choosing $k$ by using information
-criteria.  \citet{sugar2003finding} proposed an approach that
+criteria.  For example, \citet{sugar2003finding} proposed an approach that
 minimizes the estimated ``distortion'', the average distance per dimension.
-\citet{fraley2002model}'s model-based method fits Gaussian mixture model
+Likewise, \citet{fraley2002model}'s model-based method fits Gaussian mixture model
 models to the data, then selects the number of mixture components, $k$, using
@@
 cross-validation, and it is adaptive to the characteristics of the data
-distribution. The essential idea is to devise a way to measure a form 
-of internal prediction error associated each choose of $k$, and then choose
-the $k$ with the smallest associated error.
-
-In this report, we extend the Wold and Gabriel methods to the clustering
-problem, specifically to choose an appropriate number of clusters for a
-dataset.  We prove that the Gabriel method is self-consistent, and we analyze
-some of its properties in the presence of noise.  We compare these methods to
-state-of-the-art algorithms, and show that both are competitive.
-
-\textit{TODO: Add outline}
-
-\section{Cross-validation for selecting the number of clusters}
+distribution. The essential idea is to devise a way to measure a form of
+internal prediction error associated each choose of $k$, and then choose the
+$k$ with the smallest associated error. We describe this method in detail in
+Section~\ref{sec:meth}.  In Section~\ref{sec:self-consistent}, we prove that
+our method is self-consistent. Then, in Section~\ref{sec:theory-gaussian}, we
+analyze the performance of our method in the presence of Gaussian noise. The
+theoretical analysis shows that the performance of our method degrades in the
+presence of correlated noise; to fix this, we propose a correction for
+correlated noise in Section~\ref{sec:corr-correct}. In
+Sections~\ref{sec:simulations} and~\ref{sec:empirical-validation}, we
+demonstrate that our method is competitive with other state-of-the-art
+procedures in both simulated and real data sets. Then, in
+Section~\ref{sec:application-yeast}, we apply our method to a Yeast cell cycle
+dataset. We conclude with a short discussion in Section~\ref{sec:discussion}.
+
+\section{Cross-validation for clustering}
 \label{sec:meth}
-Cross-validation is commonly used for model selection in supervised learning
-problems.  In these settings, the data comes in the form of $N$
+
+\subsection{Problem statement}
+
+Suppose that we are given a data matrix with $N$ rows and $P$ columns, and we
+are tasked with choosing an appropriate number $k$ of clusters to use for
+performing $k$-means clustering on the rows of the data matrix. Recall that
+the $k$-means procedure takes a set of observations $\{ x_1, \dotsc ,x_n \}$
+and finds a set of $k$ or cluster centers $A = \{ a_1, \dotsc, a_k \} $
+minimizing the within cluster dispersion
+\[
+  W(A) = \sum_{i=1}^{n} \min_{a \in A} \|x_i - a\|^2.
+\]
+This implicitly defines a cluster assignment rule
+\[
+  g(x) = \argmin_{g \in \{1, \dotsc, k\}} \|x - a_g\|^2,
+\]
+with ties broken arbitrarily.
+%We will assume that the $k$-means procedure
+%finds an optimal choice of , $A$, but we will not assume that this solution is
+%unique.
+
+We can consider the problem of choosing $k$, the number of clusters, to be a
+model selection problem. In other domains, especially supervised learning
+problems like regression and classification, cross-validation is popular for
+performing model selection.
+In these settings, the data comes in the form of $N$
 predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
@@
 \citet{gabriel2002biblot}, who proposed holding out a single column and a
-single row at each fold.  Owen and Perry proved that this procedure is
-self-consistent, in the sense that it performs the correct model selection in
-the absence of noise, and \citet{perry2009cross} provided more theoretical
-support.
+single row at each fold.
 
@@
 
-We now give the details of how to implement
-the Gabriel cross-validation to locate the optimal cluster number $k$. The
-Wold cross-validation algorithm is described in Appendix $A$.
-
-\subsection{Gabriel CV algorithm}
+In the sequel, we extend Gabriel cross-validation to the problem of selecting
+the number of clusters, $k$, automatically, and we provide theoretical and
+empirical support analogous to the consistency results proved by
+\citet{owen2009bi}.
+
+\subsection{Gabriel cross-validation}
 \label{sec:gabriel-cv-algorithm}
-We are given a data matrix with $N$ rows and $P$ columns.  In each fold of
-cross-validation, we permute the rows and columns of the data matrix and then
-partition the rows and columns as $N = n + m$ and $P = p + q$ for 
-non-negative integers $n$, $m$, $p$, and $q$.  We treat the first $p$
-columns as ``predictors'' and the last $q$ columns as ``responses'';
-similarly, we treat the first $n$ rows as ``training'' and the last $m$ rows
-as ``test''.  In block form, the permuted data matrix is
+
+Our version of Gabriel cross validation for clustering works by performing a
+sequence of ``folds'' over the data. We use these folds to estimate a version
+of prediction error (cross-validation error) for each possible value of $k$;
+we then choose the value $\hat k$ with the smallest cross-validation error.
+
+In each fold of our cross-validation procedure, we permute the rows and
+columns of the data matrix and then partition the rows and columns as $N = n +
+m$ and $P = p + q$ for positive integers $n$, $m$, $p$, and $q$.  We treat
+the first $p$ columns as ``predictors'' and the last $q$ columns as
+``responses''; similarly, we treat the first $n$ rows as ``training'' and the
+last $m$ rows as ``test''.  In block form, the permuted data matrix is
 \[
@@
 
-\section{Self-Consistency of Gabriel CV method}
-
-This section gives the self-consistency proof of the proposed Gabriel method.
-Specifically, we will show that under appropriate conditions, in the absence
-of noise, the Gabriel cross-validation procedure finds the optimal number of
-clusters.
-
-Because $k$-means algorithm is essential to the method, we review the
-procedure here.  Given a set of observations $\{ x_1, \dotsc ,x_n \}$, and a
-specified the number of clusters $k$, the goal of the $k$-means procedure is
-to find a set of $k$ or cluster centers $A = \{ a_1, \dotsc, a_k \} $
-minimizing the within cluster dispersion
-\[
-  W(A) = \sum_{i=1}^{n} \min_{a \in A} \|x_i - a\|^2.
-\]
-This implicitly defines a cluster assignment rule
-\[
-  g(x) = \argmin_{g \in \{1, \dotsc, k\}} \|x - a_g\|^2,
-\]
-with ties broken arbitrarily.  We will assume that the $k$-means procedure
-finds an optimal solution, $A$, but we will not assume that this solution is
-unique.
-
-It will suffice to analyze a single fold of the cross-validation procedure.
-As in in section~\ref{sec:gabriel-cv-algorithm} we assume that the $P$
-variables of the data set have been partitioned into $p$ predictor variables
-represented in vector~$X$ and $q$ response variables represented in
-vector~$Y$.  The $N$ observations have been divided into two sets: $n$ train
-observations and $m$ test observations.  The following theorem gives
-conditions for Gabriel CV to recover the true number of clusters in the
+TODO: describe theoretical support
+
+\section{Self-consistency}
+\label{sec:self-consistent}
+
+An important property of any estimation procedure is that in the absence of of
+noise, the procedure correctly estimates the truth. This property is called
+``self-consistency'' \citep{tarpey96}. We will now show that Gabriel
+cross-validation is self-consistent. That is, in the absence of noise, the
+Gabriel cross-validation procedure finds the optimal number of clusters.
+
+It will suffice to prove self-consistency for a single fold of the
+cross-validation procedure.  As in in section~\ref{sec:gabriel-cv-algorithm}
+we assume that the $P$ variables of the data set have been partitioned into
+$p$ predictor variables represented in vector~$X$ and $q$ response variables
+represented in vector~$Y$.  The $N$ observations have been divided into two
+sets: $n$ train observations and $m$ test observations.  The following theorem
+gives conditions for Gabriel CV to recover the true number of clusters in the
 absence of noise.
@@
 
-This theorem is implied by the following two lemmas.
-
-\begin{lemma}
+The result statement of Theorem~\ref{thm:self-consistency} follows from 
+Lemmas~\ref{lem:self-consistency1} and~\ref{lem:self-consistency2}, which we
+will now state and prove.
+
+\begin{lemma}\label{lem:self-consistency1}
 Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
@@
 
-\begin{lemma}
+\begin{lemma}\label{lem:self-consistency2}
 Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
@@
 \section{Analysis under Gaussian noise}
+\label{sec:theory-gaussian}
 
@@
 Theorem~\ref{thm:self-consistency} tells us that the Gabriel cross-validation
-method recovers the true number of clusters then the noise is negligible.
+method recovers the true number of clusters when the noise is negligible.
 While this result gives us some assurance that the procedure is well-behaved,
@@
 Gabriel cross-validation criterion chooses the correct answer $k = 1$ whenever
-$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.
+$\rho < 0.5$; the criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.
 
@@
 cross-validation criterion chooses the correct answer $k = 1$ whenever
-$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
+$|\rho| < 0.5$; the criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.}
 \label{fig:nullcorr-equal}
@@
 
-The reason why Gabriel CV method tends to select larger $k$ when correlation is high 
-between dimensions is that it resembles the naive method mentioned in section \ref{sec:intro} 
-under such circumstance. In fact, if $X$, $Y$ are perfectly correlated, Gabriel method is 
-equivalent to the naive method.
+Intuitively, when the correlation is high, the response feature, $Y$, looks
+similar to the predictor feature, $X$. Prediction error on $X$ always
+decreases with larger $k$. Thus, when the correlation is high, the prediction
+error for $Y$ will also decrease with larger $k$. This explains why
+cross-validation breaks down in the presense of strong correlation.
 
@@
 
-\textit{todo: add and revise old proof}
+\begin{proposition}
+Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
+of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^{p+q}$ is an
+independent draw from a mean-zero multivariate normal distribution with 
+covariance matrix $\Sigma_{XY}= \left( \begin{smallmatrix} \Sigma_{XX} & \Sigma_{XY} \\ 
+ \Sigma_{YX} & \Sigma_{YY} \end{smallmatrix}\right)$, with $\Sigma_{YY}$ has leading 
+eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. In this case, the data are drawn
+from a single cluster; the true number of clusters is~$1$.  If  $\frac{\sqrt{\lambda_1}}{2}
+ > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}}$,
+then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
+\end{proposition}
+
+\begin{proof}
+Let $X$ and $Y$ be jointly multivariate normal distributed with mean $\mathbf{0}$ and covariance matrix $\Sigma_{XY}$, i.e.
+\[	(X,Y) \sim \mathcal{N} \left( \mathbf{0}, \Sigma_{XY}\right)	\]
+where $\Sigma_{XY}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} \\  \Sigma_{YX} & \Sigma_{YY} \end{bmatrix}$.
+
+Let $\Sigma_{YY} = U \Lambda U^T$  be the eigendecomposition of $\Sigma_{YY}$, with leading eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$.
+Then the centroid of $k$-means applying on $(y_1,..,y_n)$ is on the first PC
+of $Y$,\[	E(u^T_1 Y|u^T_1 Y>0) = \bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1\] and 
+\[	E(u^T_1 Y|u^T_1 Y<0) = \bar{\mu}^Y_2 =-\sqrt{2 \lambda_1/\pi}u_1\]
+ where $u^T_1 Y \sim \mathcal{N}(0,\lambda_1)$.
+
+To compute $\bar{\mu}^X_1 = E(X|u^T_1 Y>0)$, we need to know the conditional distribution $X|u^T_1 Y$.
+Since $(X,Y)$ has multivariate normal distribution, $(X,u^T_1 Y)$ also has a multivariate normal distribution with mean $\mathbf{0}$ and covariance matrix
+$$\Sigma_{X,u^T_1 Y}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} u_1 \\  u^T_1 \Sigma_{YX} & \lambda_1 \end{bmatrix}$$
+The conditional distribution $X|u^T_1 Y$ is hence normal with mean
+ $$\mu_{X|u^T_1 Y} = \Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y $$
+Therefore, 
+\begin{align}
+\bar{\mu}^X_1 &= E(X \mid u^T_1 Y>0) \nonumber \\ \nonumber
+ 			  &= E\left(E[X \mid u^T_1 Y] \mid u^T_1 Y>0\right) \\ \nonumber
+ 			  &=  E\left(\Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y \mid u^T_1 Y>0\right)\\ \nonumber
+ 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 E(u^T_1 Y \mid u^T_1 Y>0) \\ \nonumber
+ 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 \sqrt{2 \lambda_1/\pi} \\ \nonumber
+ 			  &= \sqrt{2 / \lambda_1 \pi} \Sigma_{XY}u_1
+\end{align}
+Similar calculation yields $\bar{\mu}^X_2 = -\sqrt{2 / \lambda_1 \pi} \Sigma_{XY}u_1$.
+The decision rule to classify any observed value of $X$ to $\bar{\mu}^X_1$ is therefore
+\[	(\bar{\mu}^X_1)^T X >0	\hspace{0.2in}\text{or} \hspace{0.2in} u^T_1\Sigma_{YX}X>0\] 
+Since $u^T_1\Sigma_{YX}X$ is a linear combination of $X$, it also has normal distribution 
+\[	\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)	\]
+And $(Y,u^T_1\Sigma_{YX}X)$ also have multivariate normal distribution with mean $\mathbf{0}$ 
+and covariance matrix
+\[
+\begin{bmatrix}
+\Sigma_{YY} & \Sigma_{YX}\Sigma_{XY}u_1  \\
+u^T_1\Sigma_{YX}\Sigma_{XY} &  u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1
+\end{bmatrix}
+\]
+The conditional distribution of $Y|u^T_1\Sigma_{YX}X$ is also multivariate normal with mean 
+\[	
+\mu_{Y|u^T_1\Sigma_{YX}X } = \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1}u^T_1\Sigma_{YX}X	
+\]
+The $Y$ center for $u^T_1\Sigma_{YX}X>0$ is
+\begin{align}
+\hat{\mu}^Y_1 &= E(Y|u^T_1\Sigma_{YX}X>0) \nonumber \\ \nonumber
+ 		      & =  \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) \\ \nonumber
+\end{align}
+Note that $u^T_1\Sigma_{YX}X$ has normal distribution $\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)$, so
+\[
+E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) = \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}
+\]
+Therefore, we have the $Y$ center for $u^T_1\Sigma_{YX}X>0$ be
+\begin{align*}
+\hat{\mu}^Y_1 &= \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1} \hspace{0.1in} \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} \\
+&=\frac{\sqrt{2/\pi}}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} \Sigma_{YX}\Sigma_{XY}u_1
+\end{align*}
+
+Recall that $\bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1$, to judge if $CV(2) > CV(1)$, one only need to compare the distance between  $\hat{\mu}^Y_1$  and  $\bar{\mu}^Y_1$ with distance between  $\hat{\mu}^Y_1$ and grand mean $0$.
+By variance and bias decomposition of prediction MSE, when variance is the same, only bias influence the MSE.
+
+After some linear algebra manipulation, we get
+$||\hat{\mu}^Y_1 - \bar{\mu}^Y_1||^2 > ||\hat{\mu}^Y_1||^2$ or $CV(2) > CV(1)$ iff
+\[
+ \frac{\sqrt{\lambda_1}}{2} > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} 
+\]
+\end{proof}
 
@@
 \section{Adjusting for correlation}
-
-\textit{TODO: add more discussion here}
-
-When the correlation between dimensions are high, the proposed Gabriel CV
-method tends to overestimate the number $k$. A simply remedy for the high
-correlation is available if we assume common covariance structure among the
-$k$ clusters.
-
+\label{sec:corr-correct}
+
+When the correlation between dimensions are high, the proposed Gabriel CV method tends to overestimate the number $k$.
+A simply remedy for the high correlation is available if we assume common covariance structure among the $k$ clusters.
+The procedure consists of four steps as follows:
 \begin{enumerate}
@@
 \end{enumerate}
+Such correlation-correction idea is based on the assumption that each observation $x_i$ in $\dataX$ can be expressed as
+\[	x_i = \mu(G_i)+\varepsilon_i \]
+where $\mu(G_i)$ is the true cluster center for $x_i$ and $\varepsilon_i$ is noise with covariance matrix $\Sigma$.
+To reduce correlation between dimensions, we would like to transform $\Sigma$ into a diagonal matrix.
+A simple way to achieve this goal is multiplying $\dataX$ by $\Sigma^{-1/2}$, so the transformed $x_i$
+
+\[	\Sigma^{-1/2} x_i = \Sigma^{-1/2} \mu(G_i)+\Sigma^{-1/2} \varepsilon_i \]
+has identity covariance matrix $\mathbf{I}$.
+This is exactly what step $3$ does.
+However, before we can transform the data, we need to estimate the covariance matrix $\Sigma$.
+Step $1$ and $2$ estimate the covariance matrix by first applying Gabriel CV method on $\dataX$ to get preliminary estimate $\hat{k}$ and then estimate $\Sigma$ through
+
+\[	\hat{\Sigma} = \frac{1}{n-\hat{k}} \sum^n_{i=1} (x_i-\hat{\mu}_i)(x_i-\hat{\mu}_i)^T	\]
+where $\hat{\mu}_i$ is the center of the cluster which contains $x_i$ when applying $k$-means on $\dataX$ with parameter $\hat{k}$.
+
+In practice, $\hat{\Sigma}$ wouldn't be the efficient estimator of $\Sigma$ since step $1$ may yield $\hat{k} > k$.
+Nevertheless, above procedure works well with correlated noise as we shall see in the simulation section.
 
 \section{Performance in simulations}
+\label{sec:simulations}
 
@@
 \section{Empirical validation}
+\label{sec:empirical-validation}
 
@@
 
-\section{Application}
-
-\textit{TODO: add data applicaiton where number of clusters is unknown}
+\section{Application on Yeast data set}
+\label{sec:application-yeast}
+
+The yeast Saccharomyces cerevisiae data set was collected by \cite{cho1998genome} to study the characterization of the cell cycle related processes of the budding yeast S.
+cerevisiae.
+To obtain data, cdc28 cells were synchronized by raising temperature to $37^\circ$C in late G1 phase, and then reinitiated by changing cells to cooler environment ($25^\circ$C).
+Data was collected at $17$ time points at every $10$ minutes interval covering almost $2$ complete cell cycles.
+Oligonucleotide microarrays was used to query $6,220$ gene expression profiles at these $17$ time points.
+Besides the original paper, several other studies have analyzed this data.
+
+\cite{tavazoie1999systematic} applied $k$-means clustering on the preprocessed data with $k=30$ and discovered several periodic clusters.
+The preprocessing procedure includes choosing the most variable $2,945$ geness among the total $6,220$.
+The normalized dispersion across time points (s.d/mean) was used as the metric of variation for selection.
+Data at time points $90$ and $100$ minutes were removed because they were deemed unreliable.
+It results a $2,945$ by $15$ data matrix.
+The last step of the preprocessing procedure was to variance normalize each gene across the $15$ time points, which is done by subtracting the mean and then dividing by the standard deviation.
+The result is each gene has mean zero and standard deviation equal to one across the time points.
+
+\cite{dortet2008model} performed model-based clustering on the preprocessed data of \cite{tavazoie1999systematic} and  found $27$ clusters.
+They classified the $27$ clusters into $3$ major types: cyclic gene clusters, stress gene clusters and the clusters neither cyclically nor sensitive to the induced stress.
+
+We also applied our proposed Gabriel method on the preprocessed data of \cite{tavazoie1999systematic}.
+The algorithm found $5$ clusters.
+Note here we cluster the genes as observations into clusters and the $15$ time points are considered attributes of each gene.
+Since the dimensions (time points) are likely to be correlated, we also applied the correlation-corrected version of Gabriel method on the data which also result $5$ clusters.
+This numbers is much smaller compare to those of \cite{tavazoie1999systematic} and \cite{dortet2008model}.
+We will focus our analysis on the $5$ clusters found by the proposed method, whose result is summarized on the bottom panel of Figure \ref{fig:5_clusters}, where each cluster is presented by the mean expression level of all the genes within that cluster over $15$ time points.
+Similar as \cite{tavazoie1999systematic}, we found the members of each cluster to be significantly enriched for genes function in similar biological process.
+Each gene is mapped with the Gene Ontology (GO) Annotation in Saccharomyces Genome Database (SGD) database, which provides Gene Ontology (GO) annotation of the entire genome using GO slims.
+GO slims are subsets of GO terms that give broad overview of ontology content and allow grouping of genes into broad categories, such as biological process, molecular function or cellular component.
+We choose to group genes in terms of biological process, which is a recognized series of events or molecular functions.
+There are total $103$ categories for biological process.
+
+To identify biologically meaningful gene clusters from spurious ones, \cite{tavazoie1999systematic} introduced a method, enrichment analysis, to assess the statistical significance of the over-represented gene class in a cluster using hypergeometric distribution.
+Specifically, the $p$-value of observing $k$ genes from a particular category group within a cluster of size $m$ is given by:
+\[	P = 1-\sum^{k-1}_{i=1} \frac{{{2945-n}\choose{m-i}}{{n}\choose{i}} }{ {{2945} \choose{m}} }\] 
+where $n$ is the total genes in that group.
+It can be seen as testing the null hypothesis that genes are randomly  distributed among each cluster.
+The categories with the most significant grouping of gene inside each of the $5$ clusters are shown in Table \ref{table:enrich}.
+Only $p$-values less than $4.8\times10^{-4}$ ($0.05/103$) are reported here.
+
+To better compare our results with those of \cite{tavazoie1999systematic}, we also prepared confusion Table \ref{table:confusion} and Figure \ref{fig:5_clusters}.
+The number in cell $(i,j)$ of Table \ref{table:confusion} shows the number of observations that belong to both the $i$th cluster of \cite{tavazoie1999systematic}'s and the $j$th cluster of ours.
+The plots in Figure \ref{fig:5_clusters} have similar meaning -- the mini plot on first row and the $4$th column is the mean gene expression level of observations belong to both the first cluster of \cite{tavazoie1999systematic}'s and the $4$th cluster of ours.
+We only plot the mean expression level if the number of observations in that cell is greater than $20$.
+Although \cite{tavazoie1999systematic} found $30$ clusters, not all of them were discussed in \cite{tavazoie1999systematic}.
+We only compare our results with those clusters that were profiled in \cite{tavazoie1999systematic}.
+The complete confusion matrix is in Appendix \ref{app:confusion}.
+The original result of \cite{tavazoie1999systematic} is obtained from site \url{http://arep.med.harvard.edu/network_discovery/}.
+
+\begin{table}
+\begin{center}
+\captionsetup{justification=centering}
+\caption{\label{table:enrich} Enrichment of clusters for genes within biological process categories}
+\scalebox{0.8}{%
+\begin{tabular}{cclcl}
+\hline
+   Cluster & Cluster size($m$) & Biological process category(total genes) & genes within category($k$) &$p$-value \\ \hline
+ $1$ & $550$  &  response to oxidative stress ($55$)  &    $24$  & $1.5\times10^{-5}$ \\
+     &     &  response to chemical ($213$)    &    $64$     & $2.2\times10^{-5}$\\
+  $2$& $590$  &  mitochondrion organization ($159$) & $79$ &  $1.1\times10^{-16}$  \\  
+  	 &     &  mitochondrial translation ($51$) & $28$ 	&  $2.9\times10^{-8}$  \\  
+  	 &     &  generation of precursor metabolites and energy ($80$) &  $37$ & $7.3\times10^{-8}$   \\  
+ $3$ & $654$ &  transcription from RNA polymerase II promoter ($214$) & $75$ &  $5.5\times10^{-6}$    \\  
+  	 &     & mRNA processing ($67$) & $30$ 	& $2.7\times10^{-5}$    \\  
+  	 &     & mitotic cell cycle ($183$)  & $63$ 	& $6.2\times10^{-5}$  \\ 
+  $4$& $634$  & cytoplasmic translation ($134$)  &  $105$ &  $3.3\times10^{-47}$   \\  
+     &     &  ribosomal subunit biogenesis $(138)$ & $73$ 	&  $7.7\times10^{-17}$  \\  
+  	 &     &  rRNA processing ($131$) &  $61$	&  $5.7\times10^{-11}$  \\  
+  	 &     &  ribosome assembly ($36$) &  $21$	&  $1.5\times10^{-6}$  \\  
+$5$  &   $517$   &  chromosome segregation($106$) &  $53$      &  $6.0\times10^{-15}$ \\
+     &     &  cellular response to DNA damage stimulus ($172$)  &  $71$  & $3.6\times10^{-14}$     \\  
+  	 &     &  DNA repair ($147$)  &  $64$  & $3.7\times10^{-14}$     \\                 
+ 	 &     &  DNA replication ($78$)  & $42$   &  $1.8\times10^{-13}$   \\  
+  	 &     &  mitotic cell cycle ($183$)  &  $70$  &  $4.8\times10^{-12}$   \\
+
+\hline
+\end{tabular}
+}
+\end{center}
+\hspace{0.5in} \footnotesize {}
+\end{table}
+
+It turns out that each cluster has unique trend of expression level and contains genes that tend to participate in the common processes.
+Cluster $1$ has decreasing expression level with time.
+From Table \ref{table:enrich}, we can see it concentrates genes that somatize the cell stress, such as oxidative heat-induce proteins.
+It mainly consists of observations that did not get profiled in \cite{tavazoie1999systematic}, as we can see from Table \ref{table:confusion} and Figure \ref{fig:5_clusters}.
+
+The mean gene expression level in Cluster $2$ decreases at the beginning and then increase.
+Our analysis found genes govern mitochondrial translation and mitochondrion organization are heavily concentrated in cluster $2$.
+Genes in \cite{tavazoie1999systematic}'s cluster $4$ and $8$ are heavily concentrated in this cluster, which also explains the coincidence between the enriched categories in our cluster $2$ and the functional groups in cluster $4$ and $8$ of    \cite{tavazoie1999systematic}.
+
+Our cluster $3$ is a periodic cluster where one can see two gene expression cycles from Figure \ref{fig:5_clusters}.
+Indeed, cell cycle genes that related to budding and cell polarity are commonly found in cluster $3$.
+In addition, one can see from Table \ref{table:enrich} that genes govern RNA processing and transcription are highly represented in this cluster.
+Comparing cluster $3$ to the clusters found in \cite{tavazoie1999systematic} from Table \ref{table:confusion} and Figure \ref{fig:5_clusters}, one can see that it mainly consists of genes from cluster $3$, $7$ and $14$ of \cite{tavazoie1999systematic}.
+Note that the cluster $7$ and $14$ are both periodic clusters highlighted in \cite{tavazoie1999systematic}.
+
+Cluster $4$ has increasing expression level with time.
+From Table \ref{table:enrich}, we can see genes related to cytoplasmic translation are highlighted in this cluster.
+Genes encode ribosomes are also overly presented in cluster $4$.
+This cluster corresponding to cluster $1$, the most notable functional grouping cluster of \cite{tavazoie1999systematic}.
+Indeed, one can see from Table \ref{table:confusion} that almost entire cluster $1$ of  \cite{tavazoie1999systematic} is contained in our cluster $4$.
+
+One can easily see from Figure \ref{fig:5_clusters} that our cluster $5$ is also a periodic cluster.
+Enrichment analysis shows that genes participated in the cell-cycle phase related processes such as DNA replication and DNA repair are highly concentrated in this cluster.
+Table \ref{table:confusion} shows that the entire cluster $2$ of \cite{tavazoie1999systematic} is contained in our cluster $5$, along with some genes from cluster $14$.
+Note that both cluster $2$ and $14$ are profiled periodic clusters in \cite{tavazoie1999systematic}.
+We can also see from Figure \ref{fig:5_clusters} that our cluster $5$ groups together genes with similar cycle phase in \cite{tavazoie1999systematic}'s cluster $2$ and $14$.
+
+In conclusion, our clusters group together genes that participate in the common biological processes.
+Our results are broadly align with the results in \cite{tavazoie1999systematic}, which clustered the data with $k=30$.
+In particular, the periodic cell cycle related genes are concentrated in cluster $5$ and cluster $3$, which consistent with the findings in \cite{tavazoie1999systematic} (cluster $2$, $7$ and $14$).
+Indeed, more than $70\%$ genes which function in mitotic cell cycle process are concentrated in these two clusters.
+The number of clusters found by our method is much smaller than the number of clusters used in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
+However, both studies found that most of their clusters have no characteristics, suggesting overestimate of the underlying diversity of biological expression classes in the data set \citep{tavazoie1999systematic}.
+From Figure \ref{fig:5_clusters}, one can see clusters in \cite{tavazoie1999systematic} with similar behavior are merged in our cluster, which leads to smaller number of clusters.
+Our result is more parsimonious, while still able to identify the functional grouping of genes in clusters highlighted in previous studies.
+
+	\begin{figure}[H]
+		\centering
+	\includegraphics[width=5.7in, height=5.7in]{wei/split_plots.pdf}
+	\caption{Yeast data set mean expression profiles. The $5$ clusters found by Gabriel method are on the bottom; Clusters profiled in \cite{tavazoie1999systematic} are on the right.}
+	\label{fig:5_clusters}
+	\end{figure}
+
+\begin{table}[H]
+\caption{\label{table:confusion} \small{Confusion matrix comparing $5$ clusters to clusters profiled in \cite{tavazoie1999systematic}}}
+\centering
+\begin{tabular}{lcccccr}
+  \hline
+ & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
+  \hline
+Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
+  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
+  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
+  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
+  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
+  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
+  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
+  Other & 545 & 332 & 448 & 383 & 290 & 1998 \\ 
+  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ 
+   \hline
+\end{tabular}
+\end{table}
 
 \section{Discussion}
-\label{sec:conc}
+\label{sec:discussion}
 
@@
 
+\section{Confusion matrix} \label{app:confusion}
+
+\begin{table}
+%\captionsetup{justification=centering}
+\caption{\label{table:confusion-complete} Complete confusion matrix comparing $5$ clusters to $30$ clusters }
+\begin{tabular}{lccccc|c}
+  \hline
+ & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
+  \hline
+  Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
+  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
+  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
+  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
+  Cluster 5 & 152 & 0 & 0 & 0 & 0 & 152 \\ 
+  Cluster 6 & 1 & 2 & 12 & 85 & 4 & 104 \\ 
+  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
+  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
+  Cluster 9 & 142 & 0 & 0 & 0 & 5 & 147 \\ 
+  Cluster 10 & 12 & 27 & 24 & 20 & 6 & 89 \\ 
+  Cluster 11 & 0 & 5 & 81 & 8 & 0 & 94 \\ 
+  Cluster 12 & 3 & 68 & 5 & 3 & 1 & 80 \\ 
+  Cluster 13 & 0 & 1 & 12 & 81 & 5 & 99 \\ 
+  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
+  Cluster 15 & 60 & 53 & 2 & 0 & 0 & 115 \\ 
+  Cluster 16 & 7 & 0 & 0 & 9 & 83 & 99 \\ 
+  Cluster 17 & 9 & 0 & 72 & 1 & 1 & 83 \\ 
+  Cluster 18 & 3 & 16 & 10 & 67 & 5 & 101 \\ 
+  Cluster 19 & 62 & 2 & 0 & 1 & 8 & 73 \\ 
+  Cluster 20 & 11 & 0 & 6 & 2 & 65 & 84 \\ 
+  Cluster 21 & 1 & 13 & 6 & 44 & 6 & 70 \\ 
+  Cluster 22 & 33 & 15 & 28 & 9 & 0 & 85 \\ 
+  Cluster 23 & 2 & 18 & 2 & 25 & 22 & 69 \\ 
+  Cluster 24 & 1 & 38 & 45 & 1 & 0 & 85 \\ 
+  Cluster 25 & 2 & 57 & 16 & 1 & 0 & 76 \\ 
+  Cluster 26 & 5 & 3 & 23 & 19 & 0 & 50 \\ 
+  Cluster 27 & 9 & 0 & 30 & 5 & 20 & 64 \\ 
+  Cluster 28 & 0 & 7 & 23 & 0 & 38 & 68 \\ 
+  Cluster 29 & 27 & 2 & 10 & 1 & 11 & 51 \\ 
+  Cluster 30 & 3 & 5 & 41 & 1 & 10 & 60 \\ \hline
+  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ \hline
+\end{tabular}
+\end{table}
+
 \bibliographystyle{apalike}

### 2016-06-16 cvclust.tex
revise two-cluster theory

@@
 
+\newcommand{\Oh}{O}
 \newcommand{\OhP}{O_p}
@@
 
-\begin{theorem}\label{thm:self-consistency}
+\begin{proposition}\label{prop:self-consistency}
 
@@
 Gabriel CV correctly chooses $k = K$.
-\end{theorem}
-
-The result statement of Theorem~\ref{thm:self-consistency} follows from 
+\end{proposition}
+
+The result statement of Proposition~\ref{prop:self-consistency} follows from 
 Lemmas~\ref{lem:self-consistency1} and~\ref{lem:self-consistency2}, which we
@@
 \begin{lemma}\label{lem:self-consistency1}
-Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
+Suppose that the assumptions of Proposition~\ref{prop:self-consistency} are in
 force.  If $k < K$, then $\CV(k) > 0$.
@@
 \begin{lemma}\label{lem:self-consistency2}
-Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
+Suppose that the assumptions of Proposition~\ref{prop:self-consistency} are in
 force.  If $k \geq K$, then $\CV(k) = 0$.
@@
 
-Theorem~\ref{thm:self-consistency} tells us that the Gabriel cross-validation
+Proposition~\ref{prop:self-consistency} tells us that the Gabriel cross-validation
 method recovers the true number of clusters when the noise is negligible.
@@
 
-\begin{theorem}
+\begin{proposition}\label{prop:single-2d}
 Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
@@
 and $n$ increase.
-\end{theorem}
+\end{proposition}
 
@@
 error for $Y$ will also decrease with larger $k$. This explains why
-cross-validation breaks down in the presense of strong correlation.
-
-\subsection{Single Cluster in higher dimensions}
-
-\begin{proposition}
-Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
-of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^{p+q}$ is an
-independent draw from a mean-zero multivariate normal distribution with 
-covariance matrix $\Sigma_{XY}= \left( \begin{smallmatrix} \Sigma_{XX} & \Sigma_{XY} \\ 
- \Sigma_{YX} & \Sigma_{YY} \end{smallmatrix}\right)$, with $\Sigma_{YY}$ has leading 
-eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. In this case, the data are drawn
-from a single cluster; the true number of clusters is~$1$.  If  $\frac{\sqrt{\lambda_1}}{2}
- > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}}$,
-then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
-\end{proposition}
-
-\begin{proof}
-Let $X$ and $Y$ be jointly multivariate normal distributed with mean $\mathbf{0}$ and covariance matrix $\Sigma_{XY}$, i.e.
-\[	(X,Y) \sim \mathcal{N} \left( \mathbf{0}, \Sigma_{XY}\right)	\]
-where $\Sigma_{XY}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} \\  \Sigma_{YX} & \Sigma_{YY} \end{bmatrix}$.
-
-Let $\Sigma_{YY} = U \Lambda U^T$  be the eigendecomposition of $\Sigma_{YY}$, with leading eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$.
-Then the centroid of $k$-means applying on $(y_1,..,y_n)$ is on the first PC
-of $Y$,\[	E(u^T_1 Y|u^T_1 Y>0) = \bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1\] and 
-\[	E(u^T_1 Y|u^T_1 Y<0) = \bar{\mu}^Y_2 =-\sqrt{2 \lambda_1/\pi}u_1\]
- where $u^T_1 Y \sim \mathcal{N}(0,\lambda_1)$.
-
-To compute $\bar{\mu}^X_1 = E(X|u^T_1 Y>0)$, we need to know the conditional distribution $X|u^T_1 Y$.
-Since $(X,Y)$ has multivariate normal distribution, $(X,u^T_1 Y)$ also has a multivariate normal distribution with mean $\mathbf{0}$ and covariance matrix
-$$\Sigma_{X,u^T_1 Y}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} u_1 \\  u^T_1 \Sigma_{YX} & \lambda_1 \end{bmatrix}$$
-The conditional distribution $X|u^T_1 Y$ is hence normal with mean
- $$\mu_{X|u^T_1 Y} = \Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y $$
-Therefore, 
-\begin{align}
-\bar{\mu}^X_1 &= E(X \mid u^T_1 Y>0) \nonumber \\ \nonumber
- 			  &= E\left(E[X \mid u^T_1 Y] \mid u^T_1 Y>0\right) \\ \nonumber
- 			  &=  E\left(\Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y \mid u^T_1 Y>0\right)\\ \nonumber
- 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 E(u^T_1 Y \mid u^T_1 Y>0) \\ \nonumber
- 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 \sqrt{2 \lambda_1/\pi} \\ \nonumber
- 			  &= \sqrt{2 / \lambda_1 \pi} \Sigma_{XY}u_1
-\end{align}
-Similar calculation yields $\bar{\mu}^X_2 = -\sqrt{2 / \lambda_1 \pi} \Sigma_{XY}u_1$.
-The decision rule to classify any observed value of $X$ to $\bar{\mu}^X_1$ is therefore
-\[	(\bar{\mu}^X_1)^T X >0	\hspace{0.2in}\text{or} \hspace{0.2in} u^T_1\Sigma_{YX}X>0\] 
-Since $u^T_1\Sigma_{YX}X$ is a linear combination of $X$, it also has normal distribution 
-\[	\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)	\]
-And $(Y,u^T_1\Sigma_{YX}X)$ also have multivariate normal distribution with mean $\mathbf{0}$ 
-and covariance matrix
-\[
-\begin{bmatrix}
-\Sigma_{YY} & \Sigma_{YX}\Sigma_{XY}u_1  \\
-u^T_1\Sigma_{YX}\Sigma_{XY} &  u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1
-\end{bmatrix}
-\]
-The conditional distribution of $Y|u^T_1\Sigma_{YX}X$ is also multivariate normal with mean 
-\[	
-\mu_{Y|u^T_1\Sigma_{YX}X } = \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1}u^T_1\Sigma_{YX}X	
-\]
-The $Y$ center for $u^T_1\Sigma_{YX}X>0$ is
-\begin{align}
-\hat{\mu}^Y_1 &= E(Y|u^T_1\Sigma_{YX}X>0) \nonumber \\ \nonumber
- 		      & =  \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) \\ \nonumber
-\end{align}
-Note that $u^T_1\Sigma_{YX}X$ has normal distribution $\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)$, so
-\[
-E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) = \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}
-\]
-Therefore, we have the $Y$ center for $u^T_1\Sigma_{YX}X>0$ be
-\begin{align*}
-\hat{\mu}^Y_1 &= \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1} \hspace{0.1in} \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} \\
-&=\frac{\sqrt{2/\pi}}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} \Sigma_{YX}\Sigma_{XY}u_1
-\end{align*}
-
-Recall that $\bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1$, to judge if $CV(2) > CV(1)$, one only need to compare the distance between  $\hat{\mu}^Y_1$  and  $\bar{\mu}^Y_1$ with distance between  $\hat{\mu}^Y_1$ and grand mean $0$.
-By variance and bias decomposition of prediction MSE, when variance is the same, only bias influence the MSE.
-
-After some linear algebra manipulation, we get
-$||\hat{\mu}^Y_1 - \bar{\mu}^Y_1||^2 > ||\hat{\mu}^Y_1||^2$ or $CV(2) > CV(1)$ iff
-\[
- \frac{\sqrt{\lambda_1}}{2} > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} 
-\]
-\end{proof}
+cross-validation breaks down in the presence of strong correlation.
+
+In Appendix~\ref{sec:single-general-dim}, using similar techniques to those used
+to prove Proposition~\ref{prop:single-2d}, we derive an analogous result for correlated
+Gaussian noise in more than two dimensions. A similar phenomenon holds:
+the Gabriel cross-validation method fails when the first principal component
+of the $Y$ variables is strongly correlated with a linear combination of the
+$X$ variables.
 
@@
 We will now analyze a simple two-cluster setting, and derive conditions for
-Gabriel cross-validation to correctly prefer $k=2$ clusters to $k=1$.
-
-\begin{proposition}
+Gabriel cross-validation to correctly prefer $k=2$ clusters to $k=1$. In the
+statement of the result, $\Phi(\cdot)$ and $\varphi(\cdot)$ denote the
+standard normal cumulative distribution function and density, respectively.
+
+\begin{proposition}\label{prop:twoclust}
 Suppose that $\{(X_i,Y_i)\}_{i=1}^{n+m}$ is data from a single fold of Gabriel
@@
 from an equiprobable mixture of two multivariate normal distributions with
-identity covariance. Suppose that the first mixture component has mean $(\muX,
-\muY)$, and the second has mean $(-\muX, -\muY)$, where $\muX > 0$ and
-$\muY > 0$.  If $1 + 2\Phi(\mu^Y)+ \frac{2\varphi(\mu^Y)}{\mu^Y} < 4\Phi(\mu^X)$, then $\CV(2) < \CV(1)$ with
-probability tending to one as $m$ and $n$ increase.
+identity covariance. Suppose that the first mixture component has mean
+$\mu = (\muX, \muY)$ and the second has mean $-\mu = (-\muX, -\muY)$,
+where $\muX \geq 0$ and $\muY \geq 0$. 
+If 
+\(
+  2 \varphi(\muY) + \muY + 2 \muY \Phi(\muY) < 4 \muY \Phi(\muX),
+\)
+then $\CV(2) < \CV(1)$ with probability tending to one as $m$ and $n$ increase.
 \end{proposition}
@@
 \begin{proof}
-There are two clusters $G_1$ and $G_2$, where observations from $G_1$ are distributed as
-\[	\mathcal{N}\left( \begin{pmatrix} 
-    \mu^X \\
-    \mu^Y \\
-  \end{pmatrix}, \mathbf{I} \right)	\]
-and observations from $G_2$ are distributed as
-\[	\mathcal{N}\left( \begin{pmatrix} 
-    -\mu^X \\
-    -\mu^Y \\
-  \end{pmatrix}, \mathbf{I} \right)	\]
-where $\mu^X_1 > 0$ and $\mu^Y_1 > 0$. Let $G_i$ be the true cluster where observation $i$ is generated from, by assumption
-\[	P(G_i=G_1) = P(G_i=G_2) = 1/2	\]
-After applying $k$-means on $\{ Y_i\}_{i=1}^{n}$ with $k=2$, if $n$ is large enough, we have the estimated centroids $\bar{\mu}^Y_1$ and $\bar{\mu}^Y_2$ be close to $E(Y \mid Y>0)$ and $\E(Y \mid Y < 0)$,
-with errors will be of size $\OhP(n^{-1/2})$. Here
-\begin{align}
-E(Y \mid Y>0) &= E(Y_1 \mid Y_1 > 0) \cdot P(Y_1>0) + E(Y_2 \mid Y_2 > 0) \cdot P(Y_2>0) \nonumber \\ 
-   &= 2\varphi(\mu^Y)+2\mu^Y\Phi(\mu^Y)-\mu^Y  \\  \nonumber 
-\end{align}
-where $Y_1 \sim N(\mu^Y, 1)$ and $Y_2 \sim N(-\mu^Y, 1)$, and we used Lemma~\ref{lem:truncated-normal-moments}
-(Appendix~\ref{app:technical-lemmas}). Similarly, we have
-\begin{align}
-E(Y \mid Y<0) &= E(Y_1 \mid Y_1 < 0) \cdot P(Y_1<0) + E(Y_2 \mid Y_2 < 0) \cdot P(Y_2<0) \nonumber \\ 
-   &= -2\varphi(\mu^Y)-2\mu^Y\Phi(\mu^Y)+\mu^Y  \\  \nonumber 
-\end{align}
- where $\varphi()$ and $\Phi()$ are the standard normal probability and cumulative distribution function respectively.
-
-Same as in single cluster case, the classification rule learned from
-$\{(X_i, \hGY_i)\}_{i=1}^{n}$ variables will be determined according to
-whether $X > 0$, with the decision boundary be at $0 + \OhP(n^{-1/2})$.
-By symmetry, the CV error for points from $G_1$ is same as the points from $G_2$. Because $P(G=G_1) = P(G=G_2) =1/2$, the CV error can be calculated solely from $G_2$, that is
-
+There are two clusters: observations from cluster~$1$ are distributed as
+\(
+  \mathcal{N}(\mu, I)
+\)
+and observations from cluster~$2$ are distributed as
+\(
+\mathcal{N}(-\mu, I)
+\)
+where $\mu = (\muX, \muY)$.  Without loss of generality, $\muX \geq 0$ and
+$\muY \geq 0$.  Let $G_i$ be the true cluster of observation~$i$
+where, by assumption,
+\[
+  \Pr(G_i=1) = \Pr(G_i=2) = 1/2.
+\]
+After applying $k$-means to $\{ Y_i\}_{i=1}^{n}$ with $k=2$, if $n$ is large
+enough, then the estimated cluster means $\bar{\mu}^Y_1$ and $\bar{\mu}^Y_2$
+will be close to $\E(Y \mid Y>0)$ and $\E(Y \mid Y < 0)$, with errors of
+size $\OhP(n^{-1/2})$. To compute these quantities, let
+$(X_1, Y_1) \sim \mathcal{N}(\mu, I)$ and $(X_2, Y_2) \sim \mathcal{N}(-\mu, I)$
+be draws from the mixture components, and let $(X,Y)$ be defined such that
+$\Pr(X= X_1, Y = Y_1) = \Pr(X= X_2, Y = Y_2) = 1/2$. Then,
+\begin{align*}
+  \E(Y \mid Y>0)
+  &= \E(Y_1 \mid Y_1 > 0) \cdot \Pr(Y = Y_1 \mid Y >0)
+   + \E(Y_2 \mid Y_2 > 0) \cdot \Pr(Y = Y_2 \mid Y > 0)
+\\
+&=  \{ \muY + \varphi(\muY) / \Phi(\muY) \} \cdot \Phi(\muY)
++ [-\muY + \varphi(\muY) / \{ 1 - \Phi(\muY) \}] \cdot \{ 1 - \Phi(\muY) \}
+\\
+  &= 2\varphi(\muY)+ 2\muY\Phi(\muY)- \muY. 
+\end{align*}
+In the second line, we have used
+Lemma~\ref{lem:truncated-normal-moments} from
+Appendix~\ref{app:technical-lemmas} to compute the conditional expectations;
+$\varphi()$ and $\Phi()$ are the standard normal density and
+cumulative distribution function, respectively. By symmetry,
+\begin{align*}
+  \E(Y \mid Y<0) &= -\E(Y \mid Y > 0).
+\end{align*}
+
+The classification rule learned from the training data
+$\{(X_i, \hGY_i)\}_{i=1}^{n}$ will have its decision boundary at $0 + \OhP(n^{-1/2})$;
+that is, in the limit,
+observations will get classified as coming from cluster~1 when $X > 0$. 
+Set $a = \E(Y \mid Y > 0)$.
+Up to terms of order $\OhP(n^{-1/2})$,
+the cross-validation error from a single observation is distributed as
+\[
+  (Y - a)^2 1\{ X > 0 \} + (Y + a)^2 1\{X < 0\}.
+\]
+Using the fact that conditional on the mixture component, the $X$ and $Y$
+coordinates are independent, we can compute the expectation of the first
+summand as
+\begin{align*}
+  \E[(Y - a)^2 1\{ X > 0 \}]
+  &= (1/2) \E[(Y_1 - a)^2] \Pr(X_1 > 0)
+  + (1/2) \E[(Y_2 - a)^2] \Pr(X_2 > 0)
+\\
+&= (1/2) \Phi(\muX) \E(Y_1 - a)^2  + (1/2) \{ 1 - \Phi(\muX) \} \E(Y_2 - a)^2
+\\
+&=
+(1/2) [ 1  + \Phi(\muX) (\muY - a)^2 + \{ 1 - \Phi(\muX) \} (-\muY - a)^2 ].
+\end{align*}
+By a similar calculation, the expectation of the second summand is
+\[
+  \E[(Y + a)^2 1\{ X < 0 \}]
+  =
+  (1/2)[ 1 + \{ 1 - \Phi(\muX) \} (\muY + a)^2 + \Phi(\muX) (-\muY + a)^2 ].
+\]
+Adding the two terms, we get that the expected cross-validation error from a
+single observation is
+\[
+  1 + \Phi(\muX) (\muY - a)^2 + \{ 1 - \Phi(\muX) \} (\muY + a)^2
+    =
+  1 + (\muY)^2 + a \, \{  a + 2 \muY - 4 \muY \Phi(\muX) \}.
+\]
+Thus, the $k = 2$ cross-validation error on the test set is
 \begin{align*}
@@
       +
-      \| (Y_i - \bmuY_2) 1\{\hGX_i = 2\} \|^2, \hspace{0.2in}Y\sim N(-\mu^Y,1)
+      \| (Y_i - \bmuY_2) 1\{\hGX_i = 2\} \|^2
 \\
 &=
-    \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]
-      + \OhP(m^{-1/2}) + \OhP(n^{-1/2})\\
+1 + (\muY)^2 +  a \, \{ a + 2 \muY - 4 \muY \Phi(\muX) \}
++ \OhP(n^{-1/2}) + \OhP(m^{-1/2}).
 \end{align*}
-   With 
-   \begin{align}
- \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]  =& P(\hGX_i = 1)\cdot E[(Y - a)^2] + P(\hGX_i = 2)\cdot E[(Y + a)^2] \nonumber \\  \nonumber
- =& [1-\Phi(\mu^X)][var(Y)+(-\mu^Y - a)^2]+\\ \nonumber
-  & \Phi(\mu^X)[var(Y)+(-\mu^Y + a)^2]\\  \nonumber
-  =& [1-\Phi(\mu^X)][1+(\mu^Y + a)^2]+\Phi(\mu^X)[1+(\mu^Y -a )^2]   \\  \nonumber
-  =& 1+(\mu^Y + a)^2 -4a\Phi(\mu^X)\mu^Y  \nonumber
-\end{align}
-where $a$ is given by $(1)$.
-
-When $k=1$, the result is straight forward since the estimated centroid will approximately equal to
-$0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error
-will be
+
+When $k=1$, the $k$-means centroid is equal to the sample mean $\bar Y_n = (1/n)
+\sum_{i=1}^{n} Y_i$, approximately equal to 
+$\E(Y) = 0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error is
 \[
@@
 \]
-So if we have $1 + 2\Phi(\mu^Y)+ \frac{2\varphi(\mu^Y)}{\mu^Y} < 4\Phi(\mu^X)$, we have $\CV(2) < \CV(1)$
+Thus, if $a + 2 \muY - 4 \muY \Phi(\muX) < 0$, then $\CV(2) < \CV(1)$ with
+probability tending to one as $m$ and $n$ increase. Substituting the
+expression for $\E(Y \mid Y > 0)$ in place of $a$, the inequality holds
+precisely when
+\(
+  2 \varphi(\muY) + \muY + 2 \muY \Phi(\muY) < 4 \muY \Phi(\muX).
+\)
 \end{proof}
 
-We confirm this result with a simulation.  We perform $10$ replicates for each pair of $(\mu^X, \mu^Y)$, where both $\mu^X$ and $\mu^Y$ take value on grid of $[0,3]$ with step $0.1$.  In each
-replicate, we generate $20000$ observations from two multivariate normal distributions with identity covariance, where one has mean $(\muX, \muY)$ and the other one has mean $(-\muX, -\muY)$.
-We perform a single $2 \times 2$ fold of Gabriel cross-validation and report the
-times (out of $10$ replicates) when $k=2$ is selected by the algorithm in stead of $k=1$.
-Figure~\ref{fig:overlap-color_plot} shows the frequency $k=2$ is selected by the algorithm for each pair of $(\mu^X, \mu^Y)$.
-The dark spot means high number (close to $10$) is selected by the algorithm, which means algorithm very likely will pick $k=2$ over $k=1$ for the corresponding $(\mu^X, \mu^Y)$.
-While light spot means algorithm prefer $k=1$ for the corresponding value of $(\mu^X, \mu^Y)$.
-We can see the simulation result perfectly align with the theoretical curve (the black line), which separates the $k=2$ zone from the $k=1$ zone.
-It demonstrates that the Gabriel cross-validation works exactly as it suppose to under such setting.
-The position of dark spots shows that when the two clusters are reasonably apart (not overlapping too heavily) in both dimensions, the Gabriel cross-validation is asymptotically consistent.
-
-\begin{figure}[H]
+We confirm the result of Proposition~\ref{prop:twoclust} with a simulation.
+We perform $10$ replicates for each $(\mu^X, \mu^Y)$ pair, sweeping over
+a two-dimensional grid of values in the domain $[0,3] \times [0,3]$, with step
+size $0.1$ in each dimension.  In each
+replicate, we generate $N=20000$ observations from an equiprobably mixture
+of multivariate normal distributions with identity covariance, with one
+component having mean $(\muX, \muY)$ and
+the other component having mean $(-\muX, -\muY)$. We perform a single $2 \times 2$
+fold of Gabriel cross-validation and report the number of times
+(out of $10$ replicates) where $k=2$ is selected by the algorithm instead of $k=1$.
+Figure~\ref{fig:overlap-color_plot} shows the frequency with which $k=2$ is selected by
+the algorithm for each $(\mu^X, \mu^Y)$ pair. Darker (red) colors indicate
+higher numbers (close to $10$), situations where $k = 2$ is selected more
+often than $k = 1$. Ligher (blue) colors indicate that $k = 1$ is preferred.
+We can see the simulation result perfectly align with the
+theoretical curve (the black line), which separates the $k=2$ zone from the
+$k=1$ zone.
+
+\begin{figure}
 \centering
 \includegraphics[width=3in]{demo/overlap/color_plot.pdf}
-\caption{Number of times $k=2$ is selected out of $10$ replicates for each pair of $(\muX, \muY)$.
-The heat map shows the frequency $k=2$ is selected by the algorithm, with light means low number (of $k=2$) is selected and dark color means high number is selected.
-The black line is the theoretical curve, above which the algorithm suppose to pick $k=2$ and below which algorithm select $k=1$.
-}
+\caption{Number of times $k=2$ is selected out of $10$ replicates for each
+pair of $(\muX, \muY)$. The heat map shows the frequency $k=2$ is selected by
+the algorithm, with light (blue) means $k = 1$ is preferred to $k = 2$, and
+dark (red) indicates $k = 2$ is preferred to $k = 1$.
+The black line is the theoretical boundary determined from
+Proposition~\ref{prop:twoclust}.}
 \label{fig:overlap-color_plot}
@@
 converges to its limit $k_0$ given such limit exists.
+
+\section{Analysis of Gabriel method: Single cluster in more than two dimensions}
+\label{sec:single-general-dim}
+
+\begin{proposition}
+Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
+of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^{p+q}$ is an
+independent draw from a mean-zero multivariate normal distribution with 
+covariance matrix $\Sigma_{XY}= \left( \begin{smallmatrix} \Sigma_{XX} & \Sigma_{XY} \\ 
+ \Sigma_{YX} & \Sigma_{YY} \end{smallmatrix}\right)$, with $\Sigma_{YY}$ has leading 
+eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. In this case, the data are drawn
+from a single cluster; the true number of clusters is~$1$.  If  $\frac{\sqrt{\lambda_1}}{2}
+ > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}}$,
+then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
+\end{proposition}
+
+\begin{proof}
+Let $X$ and $Y$ be jointly multivariate normal distributed with mean $\mathbf{0}$ and covariance matrix $\Sigma_{XY}$, i.e.
+\[	(X,Y) \sim \mathcal{N} \left( \mathbf{0}, \Sigma_{XY}\right)	\]
+where $\Sigma_{XY}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} \\  \Sigma_{YX} & \Sigma_{YY} \end{bmatrix}$.
+
+Let $\Sigma_{YY} = U \Lambda U^T$  be the eigendecomposition of $\Sigma_{YY}$,
+with leading eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. Then
+the centroid of $k$-means applying on $(y_1,..,y_n)$ is on the first principal
+component
+of $Y$,\[	E(u^T_1 Y|u^T_1 Y>0) = \bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1\] and 
+\[	E(u^T_1 Y|u^T_1 Y<0) = \bar{\mu}^Y_2 =-\sqrt{2 \lambda_1/\pi}u_1\]
+ where $u^T_1 Y \sim \mathcal{N}(0,\lambda_1)$.
+
+To compute $\bar{\mu}^X_1 = E(X|u^T_1 Y>0)$, we need to know the conditional distribution $X|u^T_1 Y$.
+Since $(X,Y)$ has multivariate normal distribution, $(X,u^T_1 Y)$ also has a multivariate normal distribution with mean $\mathbf{0}$ and covariance matrix
+$$\Sigma_{X,u^T_1 Y}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} u_1 \\  u^T_1 \Sigma_{YX} & \lambda_1 \end{bmatrix}$$
+The conditional distribution $X|u^T_1 Y$ is hence normal with mean
+ $$\mu_{X|u^T_1 Y} = \Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y $$
+Therefore, 
+\begin{align}
+\bar{\mu}^X_1 &= E(X \mid u^T_1 Y>0) \nonumber \\ \nonumber
+ 			  &= E\left(E[X \mid u^T_1 Y] \mid u^T_1 Y>0\right) \\ \nonumber
+ 			  &=  E\left(\Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y \mid u^T_1 Y>0\right)\\ \nonumber
+ 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 E(u^T_1 Y \mid u^T_1 Y>0) \\ \nonumber
+ 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 \sqrt{2 \lambda_1/\pi} \\ \nonumber
+      &= \sqrt{2 / (\lambda_1 \pi)} \Sigma_{XY}u_1
+\end{align}
+Similar calculation yields $\bar{\mu}^X_2 = -\sqrt{2 / (\lambda_1 \pi)} \Sigma_{XY}u_1$.
+The decision rule to classify any observed value of $X$ to $\bar{\mu}^X_1$ is therefore
+\[	(\bar{\mu}^X_1)^T X >0	\hspace{0.2in}\text{or} \hspace{0.2in} u^T_1\Sigma_{YX}X>0\] 
+Since $u^T_1\Sigma_{YX}X$ is a linear combination of $X$, it also has normal distribution 
+\[	\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)	\]
+And $(Y,u^T_1\Sigma_{YX}X)$ also have multivariate normal distribution with mean $\mathbf{0}$ 
+and covariance matrix
+\[
+\begin{bmatrix}
+\Sigma_{YY} & \Sigma_{YX}\Sigma_{XY}u_1  \\
+u^T_1\Sigma_{YX}\Sigma_{XY} &  u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1
+\end{bmatrix}
+\]
+The conditional distribution of $Y|u^T_1\Sigma_{YX}X$ is also multivariate normal with mean 
+\[	
+\mu_{Y|u^T_1\Sigma_{YX}X } = \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1}u^T_1\Sigma_{YX}X	
+\]
+The $Y$ center for $u^T_1\Sigma_{YX}X>0$ is
+\begin{align}
+\hat{\mu}^Y_1 &= E(Y|u^T_1\Sigma_{YX}X>0) \nonumber \\ \nonumber
+ 		      & =  \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) \\ \nonumber
+\end{align}
+Note that $u^T_1\Sigma_{YX}X$ has normal distribution $\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)$, so
+\[
+E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) = \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}
+\]
+Therefore, we have the $Y$ center for $u^T_1\Sigma_{YX}X>0$ be
+\begin{align*}
+\hat{\mu}^Y_1 &= \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1} \hspace{0.1in} \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} \\
+&=\frac{\sqrt{2/\pi}}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} \Sigma_{YX}\Sigma_{XY}u_1
+\end{align*}
+
+Recall that $\bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1$, to judge if $CV(2) >
+CV(1)$, one only need to compare the distance between  $\hat{\mu}^Y_1$  and
+$\bar{\mu}^Y_1$ with distance between  $\hat{\mu}^Y_1$ and grand mean $0$. By
+the variance and bias decomposition of prediction MSE, when variance is the
+same, only bias influences the MSE.
+
+After some linear algebra manipulation, we get
+$\|\hat{\mu}^Y_1 - \bar{\mu}^Y_1\|^2 > \|\hat{\mu}^Y_1\|^2$ or $\CV(2) >
+\CV(1)$ if and only if
+\[
+ \frac{\sqrt{\lambda_1}}{2} > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} 
+\]
+\end{proof}
+
+\section{Technical Lemmas}
+\label{app:technical-lemmas}
+
+\begin{lemma}\label{lem:truncated-normal-moments}
+
+If $Z$ is a standard normal random variable, then
+\[
+  \E(Z \mid a < Z < b)
+    = - \frac{\varphi(b) - \varphi(a)}
+             {\Phi(b) - \Phi(a)}
+\]
+and
+\[
+  \E\{(Z - \delta)^2 \mid a < Z < b\}
+    = \delta^2 + 1
+    - \frac{  (b - 2 \delta) \varphi(b)
+            - (a - 2 \delta) \varphi(a)}
+           {\Phi(b) - \Phi(a)}
+\]
+for all constants $a$, $b$, and $\delta$, where $\varphi(z)$ and $\Phi(z)$ are
+the standard normal probability density and cumulative distribution functions.
+These expressions are valid for $a = -\infty$ or $b = \infty$ by taking
+limits.
+
+\end{lemma}
+\begin{proof}
+We will derive the expression for the second moment.  Integrate to get
+\begin{align*}
+  \E[ (Z - \delta)^2 1\{Z < b\}]
+    &= \int_{-\infty}^b (z - \delta)^2 \varphi(z) \, dz \\
+    &= (\delta^2 + 1) \Phi(b) - (b - 2 \delta) \varphi(b).
+\end{align*}
+Now,
+\[
+  \E\{(Z - \delta)^2 \mid a < Z < b\}
+    =
+    \frac{  \E[ (Z - \delta)^2 1\{Z < b\}]
+          - \E[ (Z - \delta)^2 1\{Z < a\}]}
+         { \Phi(b) - \Phi(a) }.
+\]
+\end{proof}
+
+Lemma~\ref{lem:truncated-normal-moments} has some important special cases:
+\begin{align*}
+  \E\{Z \mid Z > 0\} &= 2 \varphi(0) = \sqrt{2 / \pi}, \\
+  \E\{(Z - \delta)^2 \mid Z > 0 \}
+    &= \delta^2 + 1 - 4 \delta \varphi(0), \\
+  \E\{(Z - \delta)^2 \mid Z < 0 \}
+    &= \delta^2 + 1 + 4 \delta \varphi(0).
+\end{align*}
 

### 2016-06-16 cvclust.tex
revise through section 5

@@
 \DeclareMathOperator*{\var}{var}
+\newcommand{\T}{T}
 
@@
 %% even when the prediction is evaluated on an independent test set
-%% \citep{hastie2009elements}. Therefore, CV prefers larger $k$ if we do the CV
-%% naively.
+%% \citep{hastie2009elements}. Therefore, cross-validation prefers larger $k$ if
+%% we do the cross-validation naively.
 
@@
 the first $p$ columns as ``predictors'' and the last $q$ columns as
-``responses''; similarly, we treat the first $n$ rows as ``training'' and the
-last $m$ rows as ``test''.  In block form, the permuted data matrix is
+``responses''; similarly, we treat the first $n$ rows as ``train''
+observations and the last $m$ rows as ``test'' observations.  In block form,
+the permuted data matrix is
 \[
@@
 
-TODO: describe theoretical support
+In Section~\ref{sec:self-consistent}, we prove that this procedure is
+self-consistent, in the sense that it recover the correct value of $k$ in the
+absence of noise. Then, in Section~\ref{sec:theory-gaussian}, we analyze some
+of the properties of Gabriel cross-validation in the presence of Gaussian
+noise.
 
@@
 sets: $n$ train observations and $m$ test observations.  The following theorem
-gives conditions for Gabriel CV to recover the true number of clusters in the
-absence of noise.
+gives conditions for Gabriel cross-validation to recover the true number of
+clusters in the absence of noise.
 
@@
 Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
-cross-validation.  For any $k$, let $CV(k)$ be the cross-validation error for
+cross-validation.  For any $k$, let $\CV(k)$ be the cross-validation error for
 this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
@@
 \end{enumerate}
-Then $CV(k) < CV(K)$ for $k < K$, and $CV(k) = CV(K)$ for $k > K$, so that
-Gabriel CV correctly chooses $k = K$.
+Then $\CV(k) < \CV(K)$ for $k < K$, and $\CV(k) = \CV(K)$ for $k > K$, so that
+Gabriel cross-validation correctly chooses $k = K$.
 \end{proposition}
@@
 
-We confirm this result with a simulation.  We perform $10$ replicates.  In each
-replicate, we generate $20000$ observations from a mean-zero bivariate normal
-distribution with unit marginal variances and correlation $\rho$.  We perform
-a single $2 \times 2$ fold of Gabriel cross-validation and report the
-cross-validation mean squared error for the number of clusters $k$ ranging
-from $1$ to $5$.  Figure~\ref{fig:nullcorr-equal} shows the cross-validation
-errors for all $10$ replicates.  The simulation demonstrates that in the
-Gabriel cross-validation criterion chooses the correct answer $k = 1$ whenever
-$\rho < 0.5$; the criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.
+We confirm the result of Proposition~\ref{prop:single-2d} with a simulation.
+We perform $10$ replicates.  In each replicate, we generate $20000$
+observations from a mean-zero bivariate normal distribution with unit marginal
+variances and correlation $\rho$.  We perform a single $2 \times 2$ fold of
+Gabriel cross-validation and report the cross-validation mean squared error
+for the number of clusters $k$ ranging from $1$ to $5$.
+Figure~\ref{fig:nullcorr-equal} shows the cross-validation errors for all $10$
+replicates.  The simulation demonstrates that in the Gabriel cross-validation
+criterion chooses the correct answer $k = 1$ whenever $\rho < 0.5$; the
+criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.
 
@@
 
-When the correlation between dimensions are high, the proposed Gabriel CV method tends to overestimate the number $k$.
-A simply remedy for the high correlation is available if we assume common covariance structure among the $k$ clusters.
-The procedure consists of four steps as follows:
+Proposition~\ref{prop:single-2d} shows that when the correlation between
+dimensions is high, the Gabriel cross-validation method tends to overestimate
+the number of clusters,~$k$. To mitigate this effect, we propose a two-stage
+estimation procedure that attempts to transform the data to minimize the
+correlation between features. In the first stage, we get a preliminary
+estimate for the number of clusters,~$\hat k_0$, and we use this value to get
+an estimate of the noise covariance matrix. Then, in the second stage, we
+transform the data attempting to sphere the noise covariance, and re-estimate
+the number of clusters, getting a final estimate~$\hat k$.
+
+The details of the correlation correction procedure are as follow:
+
 \begin{enumerate}
-	\item Apply Gabriel CV method on the original data $\dataX$, get estimated number of cluster $\hat{k}$
-	\item Estimate the pooled covariance matrix $\hat{\Sigma}$ from the $\hat{k}$ clusters.
-	\item Let $\hat{\Sigma} = \Gamma\Lambda\Gamma'$ be the Eigendecomposition of $\hat{\Sigma}$, we rescale and rotate the original data $\dataX$ to get $ \widetilde{\dataX} = \dataX\Gamma\Lambda^{-1/2}Q$, where $Q$ is a random  orthonormal rotation matrix.
-	\item Apply Gabriel CV method again on the transformed data $ \widetilde{\dataX}$ to estimate $k$. 
+
+  \item Apply the Gabriel cross-validation method on the original data
+    $\dataX$ to get a preliminary estimate of the number of
+    clusters,~$\hat k_0$.
+
+  \item Apply $k$-means to the full data set with observations
+    $\dataX_1, \dataX_2, \dotsc, \dataX_N$ using $k = \hat k_0$ clusters.
+    For $i = 1, \dotsc, N$, let $\hat \mu_i$ denote the assigned cluster
+    mean for the $i$th observation.
+
+	\item Estimate the noise covariance matrix $\hat{\Sigma}$:
+    \[
+      \hat{\Sigma}
+        =
+        \frac{1}{N - \hat k_0}
+        \sum^N_{i=1}
+        (\dataX_i-\hat{\mu}_i)(\dataX_i-\hat{\mu}_i)^\T.
+    \]
+
+  \item \label{step:transform}
+    Compute the eigendecomposition $\hat{\Sigma} = \Gamma\Lambda\Gamma^\T$.
+    Choose a random (Haar distributed) $P \times P$ orthogonal matrix $Q$.
+    Rescale and rotate the original data matrix $\dataX$ to get a
+    transformed data matrix defined by
+    \[
+      \tilde{\dataX} = \dataX\Gamma\Lambda^{-1/2}Q.
+    \]
+
+  \item Apply Gabriel cross-validation method to transformed data matrix
+    $\tilde{\dataX}$ to get a final estimate for the number of
+    clusters,~$\hat k$.
+
 \end{enumerate}
-Such correlation-correction idea is based on the assumption that each observation $x_i$ in $\dataX$ can be expressed as
-\[	x_i = \mu(G_i)+\varepsilon_i \]
-where $\mu(G_i)$ is the true cluster center for $x_i$ and $\varepsilon_i$ is noise with covariance matrix $\Sigma$.
-To reduce correlation between dimensions, we would like to transform $\Sigma$ into a diagonal matrix.
-A simple way to achieve this goal is multiplying $\dataX$ by $\Sigma^{-1/2}$, so the transformed $x_i$
-
-\[	\Sigma^{-1/2} x_i = \Sigma^{-1/2} \mu(G_i)+\Sigma^{-1/2} \varepsilon_i \]
-has identity covariance matrix $\mathbf{I}$.
-This is exactly what step $3$ does.
-However, before we can transform the data, we need to estimate the covariance matrix $\Sigma$.
-Step $1$ and $2$ estimate the covariance matrix by first applying Gabriel CV method on $\dataX$ to get preliminary estimate $\hat{k}$ and then estimate $\Sigma$ through
-
-\[	\hat{\Sigma} = \frac{1}{n-\hat{k}} \sum^n_{i=1} (x_i-\hat{\mu}_i)(x_i-\hat{\mu}_i)^T	\]
-where $\hat{\mu}_i$ is the center of the cluster which contains $x_i$ when applying $k$-means on $\dataX$ with parameter $\hat{k}$.
-
-In practice, $\hat{\Sigma}$ wouldn't be the efficient estimator of $\Sigma$ since step $1$ may yield $\hat{k} > k$.
-Nevertheless, above procedure works well with correlated noise as we shall see in the simulation section.
+
+The noise covariance estimate assumes a shared covariance matrix for all
+$k$~clusters. Letting $G_i$ denote the cluster membership of the $i$th
+observation, and letting $\mu(g)$ denote the mean of cluster $g$ for $g = 1,
+\dotsc, k$, the model supposes that
+\[
+  \dataX_i = \mu(G_i)+\varepsilon_i,
+\]
+where~$\varepsilon_i$ has mean zero and covariance matrix $\Sigma$, independent
+of~$G_i$. If we knew $\Sigma$, then we could transform the observations as
+\begin{align*}
+  \Sigma^{-1/2} \dataX_i
+    &= \Sigma^{-1/2} \mu(G_i) + \Sigma^{-1/2} \varepsilon_i \\
+    &= \tilde \mu(G_i) + \tilde \varepsilon_i,
+\end{align*}
+where $\tilde \mu = \Sigma^{-1/2} \mu$ and
+$\tilde \varepsilon_i = \Sigma^{-1/2} \varepsilon_i$. The transformed data
+has the same number of clusters, but has noise covariance
+$\cov(\tilde \varepsilon_i) = I$. The matrix product $\Gamma \Lambda^{-1/2}$
+used in step~\ref{step:transform} is an estimate of $\Sigma^{-1/2}$.
+
+The transformation used in step~\ref{step:transform} uses a random orthogonal
+matrix~$Q$, which gets applied to the rows of $\dataX$ after 
+multiplying by the estimate of $\Sigma^{-1/2}$. We use this random orthogonal
+matrix for two reasons. First, it ensures that in expectation, each transformed cluster mean
+$Q \Lambda^{-1/2} \Gamma^\T \mu(g)$ for $g = 1, \dotsc, k$ is uniformly
+spread across all $P$ features. This ensures that the self-consistency
+conditions on the cluster centers enumerated in Proposition~\ref{prop:self-consistency}
+are likely to hold. The second reason for multiplying by $Q$ is to spread any
+remaining correlation in the noise evenly (in expectation) across all
+dimensions. The latter effect follows since if $Z$ is a random vector with
+covariance matrix $\Theta$, then $Q Z$ has covariance matrix $Q \Theta Q^\T$,
+which has expectation $E(Q \Theta Q^\T) = \tr(\Theta) I$.
+
+Our correlation correction is not backed by a rigorous theoretical
+justification. However, the simulations and empirical validation in
+Sections~\ref{sec:simulations} and~\ref{sec:empirical-validation}
+demonstrate the effectiveness of our ad-hoc adjustment procedure.
 
@@
 From the graph of setting $1$, we can see high correlation between dimensions cause problem for most 
-existing methods as well as the proposed Gabriel CV method and Wold method. Higher correlation clearly 
+existing methods as well as the proposed Gabriel cross-validation method and Wold method. Higher correlation clearly 
 degenerate most algorithms' ability to find the correct $k$ as data no longer spherical in $p$-dimension.
-And the recent proposed methods (Gabriel CV method, Gap statistic, Jump method) seem to be more sensitive to
+And the recent proposed methods (Gabriel cross-validation method, Gap statistic, Jump method) seem to be more sensitive to
 high correlation than those old methods such as CH-index and Hartigan statistics. The only two methods 
@@
 model-based BIC method. Note here the data are still generated from mixture of Gaussian. In contrast, our 
-proposed Gabriel CV method as well as its correlation-corrected version actually working better in higher 
+proposed Gabriel cross-validation method as well as its correlation-corrected version actually working better in higher 
 dimension.
@@
 Result of setting $4$ indicates most previously methods are sensitive to variance heterogeneity among 
-clusters, most notably the Gap statistics and model-based BIC method. The proposed Gabriel CV method
+clusters, most notably the Gap statistics and model-based BIC method. The
+proposed Gabriel cross-validation method
 and its correlation-corrected version consistently perform well in estimating $k$ and quite insensitive
@@
 
-In conclusion, the proposed Gabriel CV method compares well with current state-of-the-art methods and 
+In conclusion, the proposed Gabriel cross-validation method compares well with current state-of-the-art methods and 
 is very robust against variance heterogeneity, high dimension and heavy-tail data. One weakness of the 
-Gabriel CV method is that it's sensitive to the high correlation among dimensions. In such case, its 
+Gabriel cross-validation method is that it's sensitive to the high correlation among dimensions. In such case, its 
 correlation-corrected version will perform well under the assumption of common covariance structure.
@@
 separates the two parties very well with the lowest miss-classification error
-$10.43\%$ ($k=3$ has $14.78\%$). Our proposed Gabriel CV method, along with
+$10.43\%$ ($k=3$ has $14.78\%$). Our proposed Gabriel cross-validation method, along with
 several other algorithms correctly pick $k=2$ for this data set. Note that
@@
 of data through the uniqueness of each cluster center. We proved the self-consistency 
-for our proposed Gabriel CV method as well as its asymptotic property with Gaussian noise,
+for our proposed Gabriel cross-validation method as well as its asymptotic property with Gaussian noise,
 and showed the robustness of our method by simulation. Our method has very good performance
@@
 our proposed method can be safely applied. In case the high correlation does exist, 
-the correlation-corrected version of Gabriel CV method in section $4.3$ can be used if we can 
+the correlation-corrected version of Gabriel cross-validation method in section $4.3$ can be used if we can 
 assume common covariance structure. Other procedure such as leave out most columns for clustering 
@@
 concentric circles share the same cluster center \citep{hastie2009elements}. However, it's 
-possible that our proposed Gabriel CV method can be used on the transformed data set. In the
+possible that our proposed Gabriel cross-validation method can be used on the transformed data set. In the
 concentric circles case where spectral clustering is appropriate, our proposed method can be applied 
@@
 
-Recall that $\bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1$, to judge if $CV(2) >
-CV(1)$, one only need to compare the distance between  $\hat{\mu}^Y_1$  and
+Recall that $\bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1$, to judge if $\CV(2) >
+\CV(1)$, one only need to compare the distance between  $\hat{\mu}^Y_1$  and
 $\bar{\mu}^Y_1$ with distance between  $\hat{\mu}^Y_1$ and grand mean $0$. By
@@
 
-\section{Wold CV estimation} \label{app:foobar}
+\section{Wold cross-validation estimation} \label{app:foobar}
 \begin{itemize}
@@
 	\end{enumerate}
-	\item For each CV folder, repeat above procedure and obtain the $Error_k$ for each $k$
+	\item For each cross-validation fold, repeat above procedure and obtain the $Error_k$ for each $k$
 	\item Average $Error_k$ across all folders for each $k$, and then select the $k$ corresponding to the minimum average $Error_k$ 

### 2016-06-16 cvclust.tex
revise section 6

@@
 
-In this section, simulation is performed to evaluate the performance of our
-proposed methods in locating the ``correct'' number of clusters. We compare
-with a basket of existing methods including Gap statistics
+\subsection{Overview}
+
+In this section, we perform a set of simulations to evaluate the performance of our
+proposed method and the associated correlation correction described in
+Section~\ref{sec:corr-correct}.
+We compare our method
+with a basket of competing methods including the Gap statistic
 \citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
-\citep{fraley2002model}, CH-index \citep{calinski1974dendrite}, Hartigan
-statistics \citep{hartigan1975clustering}, Jump method
+\citep{fraley2002model}, the CH-index \citep{calinski1974dendrite}, Hartigan's
+statistic \citep{hartigan1975clustering}, the Jump method
 \citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
-Bootstrap stability \citep{fang2012selection} in following simulation
-settings. All methods are executed with their default parameter settings. 
-We select $p=q$ and $5-$fold cross-validation in row
-($m=\frac{1}{4}n$) as default parameter setting for our proposed Gabriel
-method. Note that set $p=q$ corresponding to $2-$fold cross-validation in
-column. Wold method (detailed in Appendix $A$) and correlation-corrected 
-Gabriel method (section $4.3$) are also included for comparison.
-
-Note that in all settings, cluster centers are randomly generated from 
-multivariate normal distribution $\mathcal{N}\left(\mathbf{0},\varsigma \mathbf{I}\right)$. 
-All clusters are well-separated, i.e. no overlapping. In fact, any simulated 
-clusters with minimum distance less than one unit was discarded, so there is
-clear definition of true number of clusters. The parameters $\varsigma$ is chosen 
-such that about half of the random realization were discarded. The idea is borrowed from
-\cite{tibshirani2001estimating}. The proportion of each method successfully picks the correct
-$k$ out of $100$ simulation trials is reported, along with its corresponding confident interval
-by Wilson's method \citep{wilson1927probable}.
-
-\begin{enumerate}
-
-  \item \textbf{Correlation between dimensions} -- Six clusters in $10$ dimensions. 
-  	Each cluster has $100$ or $50$
-    multivariate normal observations with common covariance matrix $\Sigma$ 
-    which has compound symmetric structure with $1$ in diagonal and $\rho$ 
-    off diagonal. $\rho$ takes value in $\{0,0.1,...,0.9\}$.
+and Bootstrap stability \citep{fang2012selection}.
+We use the default parameter settings for all competing methods. For Gabriel
+cross-validation, we perform $2$-fold cross-validation on the columns~($p=q$)
+and $5$-fold cross-validation on the rows~($m=n/4$).
+We also compare with Wold cross-validation, which we describe in
+Appendix~\ref{sec:wold-cv}.
+
+In all simulation settings, we randomly generate cluster centers by drawing
+from a multivariate normal distribution with covariance matrix $\tau I$,
+conditional on the cluster centers being well-separated (if the distance
+between any two cluster centers is less than $1$, then we re-draw a new set of
+cluster centers). We choose $\tau$ to make the probability the cluster
+centers being well-separated on the first draw to be equal to approximately
+50\%. Many of our simulation settings are chosen to mimic the settings used by
+\cite{tibshirani2001estimating}.
+
+For each setting, we perform $100$ replicates. We report the number of times
+that each method finds the correct number~$k$ of clusters. We also report 95\%
+confidence intervals for the proportions, using Wilson's
+method~\citep{wilson1927probable}.  The simulations demonstrate that overall,
+the proposed Gabriel cross-validation method and its correlation-corrected
+version compare well with the competing methods, and they are robust to
+variance heterogeneity, high dimensional data, and heavy-tail data.
+
+\subsection{Setting 1: Correlation between dimensions}
+
+We generate six clusters in $10$ dimensions.  Each cluster has $100$ or $50$
+multivariate normal observations with common covariance matrix $\Sigma$ which
+has compound symmetric structure with $1$ in diagonal and $\rho$ off diagonal.
+$\rho$ takes value in $\{0,0.1,...,0.9\}$.
 
@@
 
-  \item \textbf{Noise dimensions} --- Three clusters in $6$ dimensions. Each cluster has $1000$ or $500$
+We can see that high correlation between dimensions causes problem for most
+existing methods, including Gabriel cross-validation method without the
+correlation correction.  The only two methods that work well in the presence
+of high correlation are the Gaussian model-based BIC method
+\citep{fraley2002model} and the correlation-corrected Gabriel method.
+
+\subsection{Setting 2: Noise dimensions}
+
+We generate three clusters in $6$ dimensions. Each cluster has $1000$ or $500$
 mean zero multivariate normal observations with identity covariance matrix.
-    We add $P$ dimensions of noise to the data, which is randomly generated from uniform$[0,1]$. The noise
-    dimension $P$ takes values in $\{0,6,...,54\}$.
+We add $r$ dimensions of noise to the data, randomly generated from a uniform
+distribution on $[0,1]$. The noise dimension $r$ takes values in
+$\{0,6,...,54\}$.
 
@@
 
-  \item \textbf{High dimension} --- Eight clusters in $p$ dimensions, $p$ takes values 
-  	in $\{10,20,...,100\}$. Each cluster has $100$ or $50$ mean zero multivariate normal 
-  	observations with identity covariance matrix.
+Most methods are relatively insensitive to adding more noise dimensions; the
+one exception to this is the Jump method, which deteriorates significantly in
+the presence of extra noise dimensions.
+
+\subsection{Setting 3: High dimension}
+
+We generate eight clusters in $P$ dimensions, with $P$ taking values in
+$\{10,20,...,100\}$.  Each cluster has $100$ or $50$ mean zero multivariate
+normal observations with identity covariance matrix.
 
@@
 
-     \item \textbf{Variance heterogeneity} --- Three clusters in $20$ dimensions, each has
-     $60$ observations in it. Observations are generated from 
+Some methods, like Jump and Gap, are insensitive to higher dimensions while
+other methods deteriorate quickly with increasing dimension, most notably the
+Gaussian model-based BIC method.  Gabriel cross-validation and its
+correlation-corrected version tend to work better in higher dimensions.
+
+\subsection{Setting 4: Variance heterogeneity}
+
+We generate three clusters in $20$ dimensions. Each cluster has $60$
+observations.  Observations are generated from
 $\mathcal{N}\left(\mathbf{0},\sigma_1^2\mathbf{I}\right)$,
 $\mathcal{N}\left(\mathbf{0},\sigma_2^2\mathbf{I}\right)$ and
-       $\mathcal{N}\left(\mathbf{0},\sigma_3^2\mathbf{I}\right)$ where 
-       $\sigma_1^2 : \sigma_2^2: \sigma_3^2 = 1:\frac{1+R}{2}:R$. The maximum ratio $R$
-       takes values in $\{1,5,10,...,45\}$
+$\mathcal{N}\left(\mathbf{0},\sigma_3^2\mathbf{I}\right)$ where $\sigma_1^2 :
+\sigma_2^2: \sigma_3^2 = 1:\frac{1+R}{2}:R$. The maximum ratio $R$ takes
+values in $\{1,5,10,...,45\}$
 
@@
 
-	\item \textbf{Heavy tail data} --- Five clusters in $15$ dimensions, each has
-     $80$ observations in it. Observations have independent $t$ distribution in each dimension  
-     with degree of freedom $\nu$, which takes value in $\{11,10,...,2\}$
+This setting demonstrates that most existing methods are sensitive to variance
+heterogeneity, most notably the Gap method and the model-based BIC method. The
+proposed Gabriel cross-validation method and its correlation-corrected version
+consistently perform well in estimating $k$ and they are insensitive to
+variance heterogeneity.
+
+\subsection{Setting 5: Heavy tail data}
+
+We generate five clusters in $15$ dimensions. Each cluster has $80$
+observations.  Observations have independent $t$ distributions in each
+dimension, with degrees of freedom $\nu$ taking values in $\{11,10,...,2\}$
 
@@
 \includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting5/Facet.pdf}
-	\label{fig:setting4}
+\label{fig:setting5}
 \end{figure}
-\end{enumerate}
-
-From the graph of setting $1$, we can see high correlation between dimensions cause problem for most 
-existing methods as well as the proposed Gabriel cross-validation method and Wold method. Higher correlation clearly 
-degenerate most algorithms' ability to find the correct $k$ as data no longer spherical in $p$-dimension.
-And the recent proposed methods (Gabriel cross-validation method, Gap statistic, Jump method) seem to be more sensitive to
-high correlation than those old methods such as CH-index and Hartigan statistics. The only two methods 
-work well in high correlation are the Gaussian model-based BIC method \citep{fraley2002model} and the 
-correlation-corrected Gabriel method.
-
-Plot of setting $3$ shows some methods are insensitive to higher dimension such as Jump method and Gap 
-statistics, while others deteriorate quickly with increasing dimension, most notably the Gaussian 
-model-based BIC method. Note here the data are still generated from mixture of Gaussian. In contrast, our 
-proposed Gabriel cross-validation method as well as its correlation-corrected version actually working better in higher 
-dimension.
-
-Result of setting $4$ indicates most previously methods are sensitive to variance heterogeneity among 
-clusters, most notably the Gap statistics and model-based BIC method. The
-proposed Gabriel cross-validation method
-and its correlation-corrected version consistently perform well in estimating $k$ and quite insensitive
-to variance heterogeneity. The Wold method also performs well in this setting.
-
-The last non-Gaussian setting examines methods' performance in heavy-tail data. With degree of freedom 
-decreases, the tail becomes more flat and the Gaussian assumption becomes more inappropriate. It explains
-the sharp plunge of Gaussian model-based BIC method in the plot. For most methods, their performances are 
-relatively stable until the tail gets very heavy. In case where degree of freedom is $2$, Gap statistics 
-and Jump method dived considerably compare to the Gabriel method and Wold method.
-
-In conclusion, the proposed Gabriel cross-validation method compares well with current state-of-the-art methods and 
-is very robust against variance heterogeneity, high dimension and heavy-tail data. One weakness of the 
-Gabriel cross-validation method is that it's sensitive to the high correlation among dimensions. In such case, its 
-correlation-corrected version will perform well under the assumption of common covariance structure.
+
+This setting investigates performance in the presence of heavy-tailed data.
+When the degrees of freedom decreases, the tail becomes more flat and the
+Gaussian assumption becomes more appropriate.  For most methods, their
+performances are relatively stable until the tail gets very heavy. In the case
+where there are $2$ degrees of freedom, the Gap and Jump methods' performances
+deteriorate considerably relative to the Gabriel method.
 
@@
 
-\section{Wold cross-validation estimation} \label{app:foobar}
+\section{Wold cross-validation}
+\label{sec:wold-cv}
+
 \begin{itemize}

### 2016-06-16 cvclust.tex

@@
 \usepackage{amsthm}
+\usepackage{booktabs}
 \usepackage{enumerate}
@@
 
-\begin{figure}[H]
+\begin{figure}
 \centering
@@
 
-We also applied our proposed method to three real world data sets, two were
-obtained from the University of California Irvine machine learning repository.
-The first and third data sets are selected because there are clear number of
-clusters in those two data sets. The second data set is used as a benchmark
-data set since it was widely used in literature.
-
-The first one is congress voting data which consists of voting records of
-$98$th United States Congress, $2$nd session \citep{schlimmer1987concept}.
-This data set includes votes for each of the U.S. House of Representatives
-Congressmen on the $16$ key votes identified by the \textit{CQA}
-(Congressional Quarterly Almanac). For each vote, each Congressman either vote
-positively ``yea" (voted for/paired for/announced for),  negatively ``nay''
-(voted against/paired against/announced against) or position unknown ``?". We
-took out those records contain ``?" before comparing each algorithm. It
-results in $232$ remaining records, with $124$ democrat and $108$ republican.
-
-The second data set is the well-known Wisconsin breast cancer data set
-\citep{mangasarian1990pattern}. After excluding the records with missing data,
-this data set consists records of $683$ patients, each with measurements of
-nine attributes of their biopsy specimens. It is known that there exist at
-least two groups of patients: $444$ patients with benign specimens and $239$
-patients with malignant specimens.
-
-The third data set is gene expression data of $5$ types of brain tumours from
-\cite{de2008clustering}, which contains $42$ observations including $10$
-medulloblastomas, $10$ malignant gliomas, $10$ atypical teratoid/rhabdoid
-tumours (AT/RTs), $8$ primitive neuroectodermal tumours (PNETs) and $4$ normal
-cerebella. It was originally used by \cite{pomeroy2002prediction} to show that
-medulloblastomas are molecularly distinct from other types of brain tumours.
-
-Data was collected with Affymetrix microarrays, which estimates the number of
-RNA copies found in the cell sample (frozen specimens). Data was preprocessed
-as following: as a common procedure, all genes with expression level below
-$10$ are set to a minimum level threshold of $10$. The maximum threshold is
-set at $16,000$. Values below or above these thresholds are often unreliable.
-To remove uninformative genes, for each gene $j$ (attribute), the mean $m_j$
-among the samples was calculated. The $10\%$ largest and smallest values are
-discarded in order to remove extreme values. Based on the mean, following
-transformation is applied on every value $x_{ij}$ of sample $i$ and gene j \[
-y_{ij} = \log_2 (x_{ij}/m_j)	\]
-
-Genes with expression levels differing by at least $l$-flod in at least $c$
-samples from their mean expression level across the sample were selected. The
-parameter $l$ and $c$ were chosen so as to produce a filtered data set with at
-least $10\%$ of the original number of genes \citep{de2008clustering}. Such
-preprocessing procedure results in $1379$ genes selected. After filtering
-step, data is restored in the original scale (i.e. $x_{ij}$). So the data set
-is $42 \times 1379$, which serves as a high dimensional example.
-
-\begin{table}[H]
-\begin{center}
+To further validate our method, we applied it to three real world data sets
+with known clustering structure.
+
+The first data set is congressional voting data consisting of voting records
+of the second session of the $98$th United States Congress,
+\citep{schlimmer1987concept}.  This data set includes votes for legislators on
+the $P = 16$ key votes.  For each vote, each legislator either votes
+positively (``yea'') negatively (``nay''). We removed legislators with missing
+votes.  This results in $N = 232$ remaining records, with $124$ democrat and
+$108$ republican. There are $k = 2$ clusters of legislators, corresponding to
+political party.
+
+The second benchmark is the \citet{mangasarian1990pattern} Wisconsin breast
+cancer data set.  After excluding the records with missing data, this data set
+consists records of $N = 683$ patients, each with measurements of $P = 9$
+attributes of their biopsy specimens. It is known that there are $k = 2$
+groups of patients: $444$ patients with benign specimens and $239$ patients
+with malignant specimens. There is some disagreement as to what the ``true''
+value of $k$ should be for this data set; \citet{fujita2014non} have argued
+that the malign group is heterogeneous, and should be split into two smaller
+clusters, yielding $k = 3$.
+
+The third data set is gene expression data of $k = 5$ types of brain tumors
+from \citet{pomeroy2002prediction}, which contains $N = 42$ observations
+including $10$ medulloblastomas, $10$ malignant gliomas, $10$ atypical
+teratoid/rhabdoid tumors, $8$ primitive neuroectodermal tumours and $4$ normal
+cerebella. After preprocessing and feature selection, there are $P = 1379$
+variables, corresponding to log activation levels for 1379 genes.
+
+%% Data was collected with Affymetrix microarrays, which estimates the number of
+%% RNA copies found in the cell sample (frozen specimens). Data was preprocessed
+%% as following: as a common procedure, all genes with expression level below
+%% $10$ are set to a minimum level threshold of $10$. The maximum threshold is
+%% set at $16,000$. Values below or above these thresholds are often unreliable.
+%% To remove uninformative genes, for each gene $j$ (attribute), the mean $m_j$
+%% among the samples was calculated. The $10\%$ largest and smallest values are
+%% discarded in order to remove extreme values. Based on the mean, following
+%% transformation is applied on every value $x_{ij}$ of sample $i$ and gene j \[
+%% y_{ij} = \log_2 (x_{ij}/m_j)	\] 
+%% 
+%% Genes with expression levels differing by at least $l$-flod in at least $c$
+%% samples from their mean expression level across the sample were selected. The
+%% parameter $l$ and $c$ were chosen so as to produce a filtered data set with at
+%% least $10\%$ of the original number of genes \citep{de2008clustering}. Such
+%% preprocessing procedure results in $1379$ genes selected. After filtering
+%% step, data is restored in the original scale (i.e. $x_{ij}$). So the data set
+%% is $42 \times 1379$, which serves as a high dimensional example.
+
+\begin{table}
+\centering
 \captionsetup{justification=centering}
-\caption{\label{table2} Number of clusters selected by each algorithm}
-\begin{tabular}{lccccc}
-    \hline                  
- & Congress Voting && Breast Cancer && Brain Tumours \\ \hline                    
-CH-index & $2$ && $2$  && $2$    \\
-Hartigan & $3$ && $3$  &&  $4$   \\
-Jump & $10$ && $9$ && $1$  \\   
-Prediction strength & $2$ &&  $2$ && $1$    \\
-Bootstrap stability & $2$ &&  $2$ &&  $7$   \\ 
-Gap & $8$ && $10$ && $10$  \\   
-Gaussian-BIC & $2$ && $5$ && $2$  \\   
-Gabriel & $2$  && $3$  && $5$  \\    
-Gabriel-corr-correct & $2$  && $2$  && $5$  \\  
-Wold & $2$ && $3$ && $4$ \\  \hline
-
+\caption{\label{tab:benchmark} Number of clusters selected on benchmark datasets}
+\begin{tabular}{lccc}
+\toprule                  
+         & \multicolumn{3}{c}{Dataset} \\
+\cmidrule(l){2-4}
+Method   & Congress Voting & Breast Cancer & Brain Tumours \\
+\midrule                    
+Gabriel  & $2$             & $3$           & $5$  \\    
+Gabriel (corr.~correct) & $2$ & $2$           & $5$  \\  
+Wold     & $2$             & $3$           & $4$  \\
+Gap      & $8$             & $10$          & $10$ \\   
+BIC      & $2$         & $5$           & $2$  \\   
+CH       & $2$             & $2$           & $2$  \\
+Hartigan & $3$             & $3$           & $4$  \\
+Jump     & $10$            & $9$           & $1$  \\   
+PS       & $2$  & $2$           & $1$  \\
+Stab.    & $2$  & $2$           & $7$  \\ 
+\midrule 
+Ground Truth & $2$         & $2$ or $3$    & $5$  \\
+\bottomrule
 \end{tabular}
-\end{center}
-\hspace{0.5in} \footnotesize {All the algorithms executed with their default parameter settings with $k$ ranges from $1$ to $10$}
 \end{table}
 
-Since most congressmen vote based on their parties' interest, two parties
-(Democratic and Republican) represent two clusters in this data set. So the
-optimal number should be two. Close inspection shows $k$-means with $k=2$
-separates the two parties very well with the lowest miss-classification error
-$10.43\%$ ($k=3$ has $14.78\%$). Our proposed Gabriel cross-validation method, along with
-several other algorithms correctly pick $k=2$ for this data set. Note that
-CH-index and Bootstrap stability also return $k=2$, but $2$ is the lower bound
-those methods can select for $k$. So it's not clear they actually choose $k=2$
-or they hit the lower bound (they would pick $k=1$ if allowed).
-
-For the breast cancer data, it's known to have $2$ clusters based on whether
-it's benign specimens or malignant specimens. Although the proposed Gabriel
-method picks $k=3$, its correlation-corrected version correctly picks $k=2$.
-It shows the necessity to correct for the high correlation among data
-features. However, for this data $k=3$ is also an acceptable answer, as
-\cite{fujita2014non} noticed that the malign group is quite heterogeneous and
-can be further clustered into at least two subgroups.
-
-The brain tumours gene expression data is a typical high-dimension data with
-$p > n$. It is clear that the correct number of clusters should be $5$, the
-number of brain tumour types. Indeed, $k$-means results with $k=5$ show that
-the $4$ normal human cerebella are well separated from brain tumours as they
-form a single cluster.  Medulloblastomas, malignant gliomas and AT/RT tumours
-also form $3$ respective clusters, which means they can be separated from each
-other. Such results are similar to those found in
-\cite{pomeroy2002prediction}, whose analysis is based on the first $3$
-principle components of the original data.    From table \ref{table2} we can
-see only our proposed Gabriel method (as well as its correlation-corrected
-version) correctly selects $k=5$, underline how difficult it was to pick the
-right $k$ when dimension is high.
+We applied the Gabriel cross-validation method, the correlation-corrected
+version, and the competing methods described in Section~\ref{sec:simulations}
+to each of the three benchmark datasets.  In each dataset, we allowed the
+number of clusters, $k$, to range from $1$ to $10$.  Table~\ref{tab:benchmark}
+displays the results.  Both versions of the Gabriel method perform well on all
+three benchmark datasets. In fact, Gabriel cross-validation is the only method
+that correctly identifies the number of clusters in all three benchmark
+datasets.
+
+%% Since most congressmen vote based on their parties' interest, two parties
+%% (Democratic and Republican) represent two clusters in this data set. So the
+%% optimal number should be two. Close inspection shows $k$-means with $k=2$
+%% separates the two parties very well with the lowest miss-classification error
+%% $10.43\%$ ($k=3$ has $14.78\%$). Our proposed Gabriel cross-validation method, along with
+%% several other algorithms correctly pick $k=2$ for this data set. Note that
+%% CH-index and Bootstrap stability also return $k=2$, but $2$ is the lower bound
+%% those methods can select for $k$. So it's not clear they actually choose $k=2$
+%% or they hit the lower bound (they would pick $k=1$ if allowed). 
+%% 
+%% For the breast cancer data, it's known to have $2$ clusters based on whether
+%% it's benign specimens or malignant specimens. Although the proposed Gabriel
+%% method picks $k=3$, its correlation-corrected version correctly picks $k=2$.
+%% It shows the necessity to correct for the high correlation among data
+%% features. However, for this data $k=3$ is also an acceptable answer, as
+%% \cite{fujita2014non} noticed that the malign group is quite heterogeneous and
+%% can be further clustered into at least two subgroups.
+%% 
+%% 
+%% The brain tumours gene expression data is a typical high-dimension data with
+%% $p > n$. It is clear that the correct number of clusters should be $5$, the
+%% number of brain tumour types. Indeed, $k$-means results with $k=5$ show that
+%% the $4$ normal human cerebella are well separated from brain tumours as they
+%% form a single cluster.  Medulloblastomas, malignant gliomas and AT/RT tumours
+%% also form $3$ respective clusters, which means they can be separated from each
+%% other. Such results are similar to those found in
+%% \cite{pomeroy2002prediction}, whose analysis is based on the first $3$
+%% principle components of the original data.    From table \ref{table2} we can
+%% see only our proposed Gabriel method (as well as its correlation-corrected
+%% version) correctly selects $k=5$, underline how difficult it was to pick the
+%% right $k$ when dimension is high.
 

### 2016-06-17 cvclust.tex
revise through 8.4

@@
 
-\section{Application on Yeast data set}
+\section{Application to yeast cell cycle data}
 \label{sec:application-yeast}
 
-The yeast Saccharomyces cerevisiae data set was collected by \cite{cho1998genome} to study the characterization of the cell cycle related processes of the budding yeast S.
-cerevisiae.
-To obtain data, cdc28 cells were synchronized by raising temperature to $37^\circ$C in late G1 phase, and then reinitiated by changing cells to cooler environment ($25^\circ$C).
-Data was collected at $17$ time points at every $10$ minutes interval covering almost $2$ complete cell cycles.
-Oligonucleotide microarrays was used to query $6,220$ gene expression profiles at these $17$ time points.
-Besides the original paper, several other studies have analyzed this data.
-
-\cite{tavazoie1999systematic} applied $k$-means clustering on the preprocessed data with $k=30$ and discovered several periodic clusters.
-The preprocessing procedure includes choosing the most variable $2,945$ geness among the total $6,220$.
-The normalized dispersion across time points (s.d/mean) was used as the metric of variation for selection.
-Data at time points $90$ and $100$ minutes were removed because they were deemed unreliable.
-It results a $2,945$ by $15$ data matrix.
-The last step of the preprocessing procedure was to variance normalize each gene across the $15$ time points, which is done by subtracting the mean and then dividing by the standard deviation.
-The result is each gene has mean zero and standard deviation equal to one across the time points.
-
-\cite{dortet2008model} performed model-based clustering on the preprocessed data of \cite{tavazoie1999systematic} and  found $27$ clusters.
-They classified the $27$ clusters into $3$ major types: cyclic gene clusters, stress gene clusters and the clusters neither cyclically nor sensitive to the induced stress.
-
-We also applied our proposed Gabriel method on the preprocessed data of \cite{tavazoie1999systematic}.
-The algorithm found $5$ clusters.
-Note here we cluster the genes as observations into clusters and the $15$ time points are considered attributes of each gene.
-Since the dimensions (time points) are likely to be correlated, we also applied the correlation-corrected version of Gabriel method on the data which also result $5$ clusters.
-This numbers is much smaller compare to those of \cite{tavazoie1999systematic} and \cite{dortet2008model}.
-We will focus our analysis on the $5$ clusters found by the proposed method, whose result is summarized on the bottom panel of Figure \ref{fig:5_clusters}, where each cluster is presented by the mean expression level of all the genes within that cluster over $15$ time points.
-Similar as \cite{tavazoie1999systematic}, we found the members of each cluster to be significantly enriched for genes function in similar biological process.
-Each gene is mapped with the Gene Ontology (GO) Annotation in Saccharomyces Genome Database (SGD) database, which provides Gene Ontology (GO) annotation of the entire genome using GO slims.
-GO slims are subsets of GO terms that give broad overview of ontology content and allow grouping of genes into broad categories, such as biological process, molecular function or cellular component.
-We choose to group genes in terms of biological process, which is a recognized series of events or molecular functions.
-There are total $103$ categories for biological process.
-
-To identify biologically meaningful gene clusters from spurious ones, \cite{tavazoie1999systematic} introduced a method, enrichment analysis, to assess the statistical significance of the over-represented gene class in a cluster using hypergeometric distribution.
-Specifically, the $p$-value of observing $k$ genes from a particular category group within a cluster of size $m$ is given by:
-\[	P = 1-\sum^{k-1}_{i=1} \frac{{{2945-n}\choose{m-i}}{{n}\choose{i}} }{ {{2945} \choose{m}} }\] 
-where $n$ is the total genes in that group.
-It can be seen as testing the null hypothesis that genes are randomly  distributed among each cluster.
-The categories with the most significant grouping of gene inside each of the $5$ clusters are shown in Table \ref{table:enrich}.
-Only $p$-values less than $4.8\times10^{-4}$ ($0.05/103$) are reported here.
-
-To better compare our results with those of \cite{tavazoie1999systematic}, we also prepared confusion Table \ref{table:confusion} and Figure \ref{fig:5_clusters}.
-The number in cell $(i,j)$ of Table \ref{table:confusion} shows the number of observations that belong to both the $i$th cluster of \cite{tavazoie1999systematic}'s and the $j$th cluster of ours.
-The plots in Figure \ref{fig:5_clusters} have similar meaning -- the mini plot on first row and the $4$th column is the mean gene expression level of observations belong to both the first cluster of \cite{tavazoie1999systematic}'s and the $4$th cluster of ours.
-We only plot the mean expression level if the number of observations in that cell is greater than $20$.
-Although \cite{tavazoie1999systematic} found $30$ clusters, not all of them were discussed in \cite{tavazoie1999systematic}.
-We only compare our results with those clusters that were profiled in \cite{tavazoie1999systematic}.
-The complete confusion matrix is in Appendix \ref{app:confusion}.
-The original result of \cite{tavazoie1999systematic} is obtained from site \url{http://arep.med.harvard.edu/network_discovery/}.
+\subsection{Motivation}
+
+Now that we have established that Gabriel cross-validation can effectively
+estimate the number of clusters, we apply our method to a yeast cell cycle
+dataset. This dataset was collected by \cite{cho1998genome} to study the cell
+cycle of budding yeast Saccharomyces cerevisiae. Other authors, including
+\citet{tavazoie1999systematic} and \citet{dortet2008model} have used $k$-means
+and related methods to cluster the genes in the dataset, with $k$
+approximately equal to~30. In both of these analyses, the authors discard the
+majority of their clusters as uninterpretable or noise, focusing instead on a
+small number of clusters. In contrast to these previous analyses, Gabriel
+cross-validation finds a small number, $k = 5$ clusters, all of which are
+interpretable.
+
+\subsection{Data collection and preprocessing}
+
+To obtain the raw data, \citet{cho1998genome} first synchronized a collection
+of CDC28 yeast cells by raising their temperature to $37^\circ$C in the late
+G1 cell cycle phase, then they reinitiated the cell cycle by switching them to
+a cooler environment ($25^\circ$C). The authors collected data at $17$ time
+points spaced evenly at $10$-minute intervals, covering almost $2$ complete
+cell cycles. At each of the $17$ time points, they used
+oligonucleotide microarrays to measure $6220$ gene expression profiles.
+
+\citet{tavazoie1999systematic} preprocessed the raw data in an attempt to
+normalize the gene responses and remove noise. They reduced the original
+$6220$ gene expression profiles to just the $2945$ genes with the highest
+variances. Then, they removed the time points at $90$ and $100$ minutes,
+because they deemed the measurements at these time points to be unreliable.
+Finally, they centered and scaled the genes by subtracting the means and
+dividing by the standard deviations, as computed from the remaining $15$ time
+points.  After the preprocessing, the data matrix $\dataX$ has $N = 2945$
+genes and $P = 15$ time points.
+
+We obtained the preprocessed data and the \citet{tavazoie1999systematic}
+cluster analysis from \url{http://arep.med.harvard.edu/network_discovery/}.
+
+\subsection{Clustering}
+
+Following \citet{tavazoie1999systematic} and \citet{dortet2008model}, we
+treat the $N = 2945$ gene expression profiles as draws from a mixture
+distribution, and we perform $k$-means clustering to segment the genes
+according to their expression profiles across the $P = 15$ time points. 
+Both the original and the correlation-corrected version of Gabriel
+cross-validation find $k = 5$ clusters.
+
+The lower-left panel of Figure~\ref{fig:5_clusters} shows the average
+expression level for each cluster across the 15 time points, with error bars
+showing standard deviations. Cluster~$1$ has decreasing expression level with
+time. The mean gene expression level in Cluster~$2$ decreases at the beginning
+and then increases. Cluster~$3$ is a periodic cluster where one can see two
+periods corresponding to the two cell cycles. Cluster~$4$ has increasing
+expression level with time.  Cluster~$5$ is another periodic cluster.
+
+\subsection{Enrichment analysis}
+
+To further validate our clusters, we follow \citet{tavazoie1999systematic},
+performing an enrichment analysis to discover which functional gene groups are
+significantly over-represented in each cluster. In the Saccharomyces Genome
+Database, each gene is mapped to a set of Gene Ontology categories. We focus
+on the $103$ biological process categories.
+
+For each category and each cluster, we compute a $p$-value for the null
+hypothesis that genes from the category are distributed across all clusters
+without any bias towards the particular cluster in question. Under the null
+hypothesis, the number of genes from the category that end up in the cluster
+is distributed as a hypergeometric random variable. For each cluster, we
+compute $p$-values for all $103$ biological process categories, and we report
+those that are significantly enrigched in Table~\ref{table:enrich}.  Using a
+Bonferroni correction to control the family-wise error rate at level 5\%, we
+only report $p$-values that are less than $0.05 / 103 = 4.8\times10^{-4}$.
 
@@
 \captionsetup{justification=centering}
-\caption{\label{table:enrich} Enrichment of clusters for genes within biological process categories}
-\scalebox{0.8}{%
-\begin{tabular}{cclcl}
-\hline
-   Cluster & Cluster size($m$) & Biological process category(total genes) & genes within category($k$) &$p$-value \\ \hline
- $1$ & $550$  &  response to oxidative stress ($55$)  &    $24$  & $1.5\times10^{-5}$ \\
-     &     &  response to chemical ($213$)    &    $64$     & $2.2\times10^{-5}$\\
-  $2$& $590$  &  mitochondrion organization ($159$) & $79$ &  $1.1\times10^{-16}$  \\  
-  	 &     &  mitochondrial translation ($51$) & $28$ 	&  $2.9\times10^{-8}$  \\  
-  	 &     &  generation of precursor metabolites and energy ($80$) &  $37$ & $7.3\times10^{-8}$   \\  
- $3$ & $654$ &  transcription from RNA polymerase II promoter ($214$) & $75$ &  $5.5\times10^{-6}$    \\  
-  	 &     & mRNA processing ($67$) & $30$ 	& $2.7\times10^{-5}$    \\  
-  	 &     & mitotic cell cycle ($183$)  & $63$ 	& $6.2\times10^{-5}$  \\ 
-  $4$& $634$  & cytoplasmic translation ($134$)  &  $105$ &  $3.3\times10^{-47}$   \\  
-     &     &  ribosomal subunit biogenesis $(138)$ & $73$ 	&  $7.7\times10^{-17}$  \\  
-  	 &     &  rRNA processing ($131$) &  $61$	&  $5.7\times10^{-11}$  \\  
-  	 &     &  ribosome assembly ($36$) &  $21$	&  $1.5\times10^{-6}$  \\  
-$5$  &   $517$   &  chromosome segregation($106$) &  $53$      &  $6.0\times10^{-15}$ \\
-     &     &  cellular response to DNA damage stimulus ($172$)  &  $71$  & $3.6\times10^{-14}$     \\  
-  	 &     &  DNA repair ($147$)  &  $64$  & $3.7\times10^{-14}$     \\                 
- 	 &     &  DNA replication ($78$)  & $42$   &  $1.8\times10^{-13}$   \\  
-  	 &     &  mitotic cell cycle ($183$)  &  $70$  &  $4.8\times10^{-12}$   \\
-
-\hline
+\caption{\label{table:enrich} Biological process enrichment within gene clusters}
+\footnotesize
+\begin{tabular}{ccll}
+\toprule
+Cluster & Cluster Size & Process Category (In Cluster/Total Genes) & $p$-value \\
+\midrule
+$1$ & $550$ & response to oxidative stress ($24/55$)  & $1.5\times10^{-5}$  \\
+    &       & response to chemical ($64/213$)         & $2.2\times10^{-5}$  \\
+$2$ & $590$ & mitochondrion organization ($79/159$)   & $1.1\times10^{-16}$ \\
+    &       & mitochondrial translation ($28/51$)     & $2.9\times10^{-8}$  \\
+    &       & generation of precursor metabolites and energy ($37/80$) & $7.3\times10^{-8}$ \\
+$3$ & $654$ & transcription from RNA polymerase II promoter ($75/214$) & $5.5\times10^{-6}$ \\
+    &       & mRNA processing ($30/67$)               & $2.7\times10^{-5}$  \\
+    &       & mitotic cell cycle ($63/183$)           & $6.2\times10^{-5}$  \\
+$4$ & $634$ & cytoplasmic translation ($105/134$)     & $3.3\times10^{-47}$ \\
+    &       & ribosomal subunit biogenesis $(73/138)$ & $7.7\times10^{-17}$ \\
+    &       & rRNA processing ($61/131$)              & $5.7\times10^{-11}$ \\
+    &       & ribosome assembly ($21/36$)             & $1.5\times10^{-6}$  \\
+$5$ & $517$ & chromosome segregation($53/106$)        & $6.0\times10^{-15}$ \\
+    &       & cellular response to DNA damage stimulus ($71/172$)      & $3.6\times10^{-14}$ \\
+    &       & DNA repair ($64/147$)                   & $3.7\times10^{-14}$ \\
+    &       & DNA replication ($42/78$)               & $1.8\times10^{-13}$ \\
+    &       & mitotic cell cycle ($70/183$)           & $4.8\times10^{-12}$ \\
+\bottomrule
 \end{tabular}
-}
 \end{center}
@@
 
-It turns out that each cluster has unique trend of expression level and contains genes that tend to participate in the common processes.
-Cluster $1$ has decreasing expression level with time.
-From Table \ref{table:enrich}, we can see it concentrates genes that somatize the cell stress, such as oxidative heat-induce proteins.
-It mainly consists of observations that did not get profiled in \cite{tavazoie1999systematic}, as we can see from Table \ref{table:confusion} and Figure \ref{fig:5_clusters}.
-
-The mean gene expression level in Cluster $2$ decreases at the beginning and then increase.
-Our analysis found genes govern mitochondrial translation and mitochondrion organization are heavily concentrated in cluster $2$.
-Genes in \cite{tavazoie1999systematic}'s cluster $4$ and $8$ are heavily concentrated in this cluster, which also explains the coincidence between the enriched categories in our cluster $2$ and the functional groups in cluster $4$ and $8$ of    \cite{tavazoie1999systematic}.
-
-Our cluster $3$ is a periodic cluster where one can see two gene expression cycles from Figure \ref{fig:5_clusters}.
-Indeed, cell cycle genes that related to budding and cell polarity are commonly found in cluster $3$.
-In addition, one can see from Table \ref{table:enrich} that genes govern RNA processing and transcription are highly represented in this cluster.
-Comparing cluster $3$ to the clusters found in \cite{tavazoie1999systematic} from Table \ref{table:confusion} and Figure \ref{fig:5_clusters}, one can see that it mainly consists of genes from cluster $3$, $7$ and $14$ of \cite{tavazoie1999systematic}.
-Note that the cluster $7$ and $14$ are both periodic clusters highlighted in \cite{tavazoie1999systematic}.
-
-Cluster $4$ has increasing expression level with time.
-From Table \ref{table:enrich}, we can see genes related to cytoplasmic translation are highlighted in this cluster.
-Genes encode ribosomes are also overly presented in cluster $4$.
-This cluster corresponding to cluster $1$, the most notable functional grouping cluster of \cite{tavazoie1999systematic}.
-Indeed, one can see from Table \ref{table:confusion} that almost entire cluster $1$ of  \cite{tavazoie1999systematic} is contained in our cluster $4$.
-
-One can easily see from Figure \ref{fig:5_clusters} that our cluster $5$ is also a periodic cluster.
-Enrichment analysis shows that genes participated in the cell-cycle phase related processes such as DNA replication and DNA repair are highly concentrated in this cluster.
-Table \ref{table:confusion} shows that the entire cluster $2$ of \cite{tavazoie1999systematic} is contained in our cluster $5$, along with some genes from cluster $14$.
-Note that both cluster $2$ and $14$ are profiled periodic clusters in \cite{tavazoie1999systematic}.
-We can also see from Figure \ref{fig:5_clusters} that our cluster $5$ groups together genes with similar cycle phase in \cite{tavazoie1999systematic}'s cluster $2$ and $14$.
-
-In conclusion, our clusters group together genes that participate in the common biological processes.
-Our results are broadly align with the results in \cite{tavazoie1999systematic}, which clustered the data with $k=30$.
-In particular, the periodic cell cycle related genes are concentrated in cluster $5$ and cluster $3$, which consistent with the findings in \cite{tavazoie1999systematic} (cluster $2$, $7$ and $14$).
-Indeed, more than $70\%$ genes which function in mitotic cell cycle process are concentrated in these two clusters.
-The number of clusters found by our method is much smaller than the number of clusters used in \cite{tavazoie1999systematic} and \cite{dortet2008model}.
-However, both studies found that most of their clusters have no characteristics, suggesting overestimate of the underlying diversity of biological expression classes in the data set \citep{tavazoie1999systematic}.
-From Figure \ref{fig:5_clusters}, one can see clusters in \cite{tavazoie1999systematic} with similar behavior are merged in our cluster, which leads to smaller number of clusters.
-Our result is more parsimonious, while still able to identify the functional grouping of genes in clusters highlighted in previous studies.
+From Table \ref{table:enrich}, we can see that Cluster~1 is enriched with
+genes that somatize cell stress, such as oxidative heat-induce proteins.
+Cluster~2 contains genes that govern mitochondrial translation and
+mitochondrion organization.  Cluster~3, the first period cluster, contains
+cell cycle genes related to budding and cell polarity, along with genes that
+govern RNA processing and transcription.  Cluster~4 contains genes related to
+cytoplasmic translation and genes encoding ribosomes.  Cluster~5, the second
+periodic cluster, contains genes that participate cell-cycle processes, along
+with DNA replication and DNA repair.
+
+\subsection{Comparison with Tavazoie~clusters}
+
+To better compare our results with those of \cite{tavazoie1999systematic}, we
+also prepared confusion Table \ref{table:confusion} and Figure
+\ref{fig:5_clusters}. The number in cell $(i,j)$ of Table
+\ref{table:confusion} shows the number of observations that belong to both the
+$i$th cluster of \cite{tavazoie1999systematic}'s and the $j$th cluster of
+ours. The plots in Figure \ref{fig:5_clusters} have similar meaning -- the
+mini plot on first row and the $4$th column is the mean gene expression level
+of observations belong to both the first cluster of
+\cite{tavazoie1999systematic}'s and the $4$th cluster of ours. We only plot
+the mean expression level if the number of observations in that cell is
+greater than $20$. Although \cite{tavazoie1999systematic} found $30$ clusters,
+not all of them were discussed in \cite{tavazoie1999systematic}. We only
+compare our results with those clusters that were profiled in
+\cite{tavazoie1999systematic}. The complete confusion matrix is in Appendix
+\ref{app:confusion}.
+
+Cluster~1 mainly consists of observations that did not get profiled in
+\cite{tavazoie1999systematic}, as we can see from Table \ref{table:confusion}
+and Figure \ref{fig:5_clusters}.
+
+Genes in \cite{tavazoie1999systematic}'s cluster $4$ and $8$ are heavily
+concentrated in this our Cluster~2, which also explains the coincidence between the
+enriched categories in our cluster $2$ and the functional groups in cluster
+$4$ and $8$ of    \cite{tavazoie1999systematic}.
+
+Comparing cluster $3$ to the clusters
+found in \cite{tavazoie1999systematic} from Table \ref{table:confusion} and
+Figure \ref{fig:5_clusters}, one can see that it mainly consists of genes from
+cluster $3$, $7$ and $14$ of \cite{tavazoie1999systematic}. Note that the
+cluster $7$ and $14$ are both periodic clusters highlighted in
+\cite{tavazoie1999systematic}.
+
+Our Cluster~4 corresponds to Tavazoie's cluster $1$, the most notable functional grouping cluster of
+\cite{tavazoie1999systematic}. Indeed, one can see from Table
+\ref{table:confusion} that almost entire cluster $1$ of
+\cite{tavazoie1999systematic} is contained in our cluster $4$.
+
+Table \ref{table:confusion} shows
+that the entire cluster $2$ of \cite{tavazoie1999systematic} is contained in
+our cluster $5$, along with some genes from cluster $14$. Note that both
+cluster $2$ and $14$ are profiled periodic clusters in
+\cite{tavazoie1999systematic}. We can also see from Figure
+\ref{fig:5_clusters} that our cluster $5$ groups together genes with similar
+cycle phase in \cite{tavazoie1999systematic}'s cluster $2$ and $14$.
+
+In conclusion, our clusters group together genes that participate in the
+common biological processes. Our results are broadly align with the results in
+\cite{tavazoie1999systematic}, which clustered the data with $k=30$. In
+particular, the periodic cell cycle related genes are concentrated in cluster
+$5$ and cluster $3$, which consistent with the findings in
+\cite{tavazoie1999systematic} (cluster $2$, $7$ and $14$). Indeed, more than
+$70\%$ genes which function in mitotic cell cycle process are concentrated in
+these two clusters. The number of clusters found by our method is much smaller
+than the number of clusters used in \cite{tavazoie1999systematic} and
+\cite{dortet2008model}. However, both studies found that most of their
+clusters have no characteristics, suggesting overestimate of the underlying
+diversity of biological expression classes in the data set
+\citep{tavazoie1999systematic}. From Figure \ref{fig:5_clusters}, one can see
+clusters in \cite{tavazoie1999systematic} with similar behavior are merged in
+our cluster, which leads to smaller number of clusters. Our result is more
+parsimonious, while still able to identify the functional grouping of genes in
+clusters highlighted in previous studies.
 

### 2016-06-17 cvclust.tex
finish revising section 8

@@
 
+\begin{figure}
+		\centering
+	\includegraphics[width=5.7in, height=5.7in]{wei/split_plots.pdf}
+	\caption{Yeast data set mean expression profiles. The $5$ clusters found by
+  the Gabriel method are on the bottom; Clusters profiled in
+  \cite{tavazoie1999systematic} are on the right.}
+	\label{fig:5_clusters}
+\end{figure}
+
 Following \citet{tavazoie1999systematic} and \citet{dortet2008model}, we
@@
 on the $103$ biological process categories.
-
-For each category and each cluster, we compute a $p$-value for the null
-hypothesis that genes from the category are distributed across all clusters
-without any bias towards the particular cluster in question. Under the null
-hypothesis, the number of genes from the category that end up in the cluster
-is distributed as a hypergeometric random variable. For each cluster, we
-compute $p$-values for all $103$ biological process categories, and we report
-those that are significantly enrigched in Table~\ref{table:enrich}.  Using a
-Bonferroni correction to control the family-wise error rate at level 5\%, we
-only report $p$-values that are less than $0.05 / 103 = 4.8\times10^{-4}$.
 
@@
 $4$ & $634$ & cytoplasmic translation ($105/134$)     & $3.3\times10^{-47}$ \\
-    &       & ribosomal subunit biogenesis $(73/138)$ & $7.7\times10^{-17}$ \\
+    &       & ribosomal subunit biogenesis ($73/138$) & $7.7\times10^{-17}$ \\
     &       & rRNA processing ($61/131$)              & $5.7\times10^{-11}$ \\
     &       & ribosome assembly ($21/36$)             & $1.5\times10^{-6}$  \\
-$5$ & $517$ & chromosome segregation($53/106$)        & $6.0\times10^{-15}$ \\
+$5$ & $517$ & chromosome segregation ($53/106$)       & $6.0\times10^{-15}$ \\
     &       & cellular response to DNA damage stimulus ($71/172$)      & $3.6\times10^{-14}$ \\
@@
 
+For each category and each cluster, we compute a $p$-value for the null
+hypothesis that genes from the category are distributed across all clusters
+without any bias towards the particular cluster in question. Under the null
+hypothesis, the number of genes from the category that end up in the cluster
+is distributed as a hypergeometric random variable. For each cluster, we
+compute $p$-values for all $103$ biological process categories, and we report
+those that are significantly enrigched in Table~\ref{table:enrich}.  Using a
+Bonferroni correction to control the family-wise error rate at level 5\%, we
+only report $p$-values that are less than $0.05 / 103 = 4.8\times10^{-4}$.
+
 From Table \ref{table:enrich}, we can see that Cluster~1 is enriched with
@@
 Cluster~2 contains genes that govern mitochondrial translation and
-mitochondrion organization.  Cluster~3, the first period cluster, contains
-cell cycle genes related to budding and cell polarity, along with genes that
-govern RNA processing and transcription.  Cluster~4 contains genes related to
+mitochondrion organization. Cluster~3, the first period cluster, contains cell
+cycle genes related to budding and cell polarity, along with genes that govern
+RNA processing and transcription. Cluster~4 contains genes related to
 cytoplasmic translation and genes encoding ribosomes. Cluster~5, the second
@@
 
-To better compare our results with those of \cite{tavazoie1999systematic}, we
-also prepared confusion Table \ref{table:confusion} and Figure
-\ref{fig:5_clusters}. The number in cell $(i,j)$ of Table
-\ref{table:confusion} shows the number of observations that belong to both the
-$i$th cluster of \cite{tavazoie1999systematic}'s and the $j$th cluster of
-ours. The plots in Figure \ref{fig:5_clusters} have similar meaning -- the
-mini plot on first row and the $4$th column is the mean gene expression level
-of observations belong to both the first cluster of
-\cite{tavazoie1999systematic}'s and the $4$th cluster of ours. We only plot
-the mean expression level if the number of observations in that cell is
-greater than $20$. Although \cite{tavazoie1999systematic} found $30$ clusters,
-not all of them were discussed in \cite{tavazoie1999systematic}. We only
-compare our results with those clusters that were profiled in
-\cite{tavazoie1999systematic}. The complete confusion matrix is in Appendix
-\ref{app:confusion}.
-
-Cluster~1 mainly consists of observations that did not get profiled in
-\cite{tavazoie1999systematic}, as we can see from Table \ref{table:confusion}
-and Figure \ref{fig:5_clusters}.
-
-Genes in \cite{tavazoie1999systematic}'s cluster $4$ and $8$ are heavily
-concentrated in this our Cluster~2, which also explains the coincidence between the
-enriched categories in our cluster $2$ and the functional groups in cluster
-$4$ and $8$ of    \cite{tavazoie1999systematic}.
-
-Comparing cluster $3$ to the clusters
-found in \cite{tavazoie1999systematic} from Table \ref{table:confusion} and
-Figure \ref{fig:5_clusters}, one can see that it mainly consists of genes from
-cluster $3$, $7$ and $14$ of \cite{tavazoie1999systematic}. Note that the
-cluster $7$ and $14$ are both periodic clusters highlighted in
-\cite{tavazoie1999systematic}.
-
-Our Cluster~4 corresponds to Tavazoie's cluster $1$, the most notable functional grouping cluster of
-\cite{tavazoie1999systematic}. Indeed, one can see from Table
-\ref{table:confusion} that almost entire cluster $1$ of
-\cite{tavazoie1999systematic} is contained in our cluster $4$.
-
-Table \ref{table:confusion} shows
-that the entire cluster $2$ of \cite{tavazoie1999systematic} is contained in
-our cluster $5$, along with some genes from cluster $14$. Note that both
-cluster $2$ and $14$ are profiled periodic clusters in
-\cite{tavazoie1999systematic}. We can also see from Figure
-\ref{fig:5_clusters} that our cluster $5$ groups together genes with similar
-cycle phase in \cite{tavazoie1999systematic}'s cluster $2$ and $14$.
-
-In conclusion, our clusters group together genes that participate in the
-common biological processes. Our results are broadly align with the results in
-\cite{tavazoie1999systematic}, which clustered the data with $k=30$. In
-particular, the periodic cell cycle related genes are concentrated in cluster
-$5$ and cluster $3$, which consistent with the findings in
-\cite{tavazoie1999systematic} (cluster $2$, $7$ and $14$). Indeed, more than
-$70\%$ genes which function in mitotic cell cycle process are concentrated in
-these two clusters. The number of clusters found by our method is much smaller
-than the number of clusters used in \cite{tavazoie1999systematic} and
-\cite{dortet2008model}. However, both studies found that most of their
-clusters have no characteristics, suggesting overestimate of the underlying
-diversity of biological expression classes in the data set
-\citep{tavazoie1999systematic}. From Figure \ref{fig:5_clusters}, one can see
-clusters in \cite{tavazoie1999systematic} with similar behavior are merged in
-our cluster, which leads to smaller number of clusters. Our result is more
-parsimonious, while still able to identify the functional grouping of genes in
-clusters highlighted in previous studies.
-
-	\begin{figure}[H]
-		\centering
-	\includegraphics[width=5.7in, height=5.7in]{wei/split_plots.pdf}
-	\caption{Yeast data set mean expression profiles. The $5$ clusters found by Gabriel method are on the bottom; Clusters profiled in \cite{tavazoie1999systematic} are on the right.}
-	\label{fig:5_clusters}
-	\end{figure}
-
-\begin{table}[H]
-\caption{\label{table:confusion} \small{Confusion matrix comparing $5$ clusters to clusters profiled in \cite{tavazoie1999systematic}}}
+In the \citet{tavazoie1999systematic} analysis, those authors performed
+$k$-means clustering with $k = 30$; they found 23 of the clusters to be
+uninterpretable, and they found 7 clusters to be meaningful. To compare our
+clusters with the Tavazoie et al.\ clusters, we prepared a confusion matrix
+comparing our clusters with the 7 interpretable Tavazoie clusters in
+Table~\ref{table:confusion}.  (Appendix~\ref{app:confusion} contains a
+confusion matrix for all 30 Tavazoie clusters, including the non-interpretable
+ones.) Entry $(i,j)$ of the confusion matrix gives the number of genes in
+Tavazoie's Cluster~$i$ and our Cluster~$j$.
+
+\begin{table}
+\caption{\label{table:confusion} Confusion matrix comparing
+    the 5 clusters found by Gabriel cross-validation to the 7 interpretable
+    clusters found by the \cite{tavazoie1999systematic} analysis}
 \centering
 \begin{tabular}{lcccccr}
-  \hline
+  \toprule
  & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
-  \hline
+  \midrule
 Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
@@
   Other & 545 & 332 & 448 & 383 & 290 & 1998 \\ 
+  \midrule
   Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ 
-   \hline
+   \bottomrule
 \end{tabular}
 \end{table}
+
+Figure~\ref{fig:5_clusters} provides a more in-depth comparison with the
+Tavazoie clusters, using a graphical confusion matrix.  The plot in cell
+$(i,j)$ of the upper left part of this figure gives the mean expression level
+for genes in the intersection of Tavazoie's Cluster~$i$ and our Cluster~$j$;
+the plots in the margins give the mean expression levels for Tavazoie's
+clusters (top right) and our clusters (bottom left). In
+Figure~\ref{fig:5_clusters}, we only include a plot for cell $(i,j)$ if the
+number of genes in that cell is greater than $20$.
+
+Our Cluster~1 mainly consists of genes that Tavazoie et al.\ found to be in
+uninterpretable clusters. Our Cluster~2 contains high concentrations of
+Tavazoie's Clusters~4 and~8. Our first periodic cluster, Cluster~3, contains
+high concentrations of Tavazoie's Clulsters~3,~7, and~14; this is notable,
+because Tavazoie et al.\ highlighted their Clusters~7 and~14 as being
+periodic. Our Cluster~4 contains almost all of Tavazoie's Cluster~1, along
+with part of Tavazoie's Cluster~4. Finally, our second periodic cluster,
+Cluster~5, contains almost all of Tavazoie's Cluster~2, along with part of
+Tavazoie's Cluster 14; this, again, is notable, because Tavazoie et al.\
+highlited these clusters as being periodic.
+
+For the clusters that Tavazoie et al.\ were able to characterize, our analysis
+broadly agrees with the earlier clustering. The major difference between our
+analysis and that of \citet{tavazoie1999systematic} is that we are able to
+identify meaningful groups of genes with a much smaller value of $k$ ($k = 5$
+instead of $k = 30$), and we are able to interpret all of the clusters
+found by our analysis.
 

### 2016-06-17 cvclust.tex
revise section 9

@@
 comparing our clusters with the 7 interpretable Tavazoie clusters in
-Table~\ref{table:confusion}.  (Appendix~\ref{app:confusion} contains a
-confusion matrix for all 30 Tavazoie clusters, including the non-interpretable
-ones.) Entry $(i,j)$ of the confusion matrix gives the number of genes in
+Table~\ref{table:confusion}.
+%%(Appendix~\ref{app:confusion} contains a confusion matrix for all 30
+%% Tavazoie clusters, including the non-interpretable ones.)
+Entry $(i,j)$ of the confusion matrix gives the number of genes in
 Tavazoie's Cluster~$i$ and our Cluster~$j$.
@@
 
-In this paper, we proposed a novel approach to estimate the number of
-clusters. The intuition behind our proposed methods is to transfer the
-unsupervised learning problem into supervised learning problem via novel form
-of cross validation. Such approach is quite different from previous methods which 
-utilize the within/between cluster dispersion or stability criterion for selecting the 
-optimal $k$. Our method utilizes the connection between different dimensions (columns)
-of data through the uniqueness of each cluster center. We proved the self-consistency 
-for our proposed Gabriel cross-validation method as well as its asymptotic property with Gaussian noise,
-and showed the robustness of our method by simulation. Our method has very good performance
-in our limited simulation settings and real data application, and clearly the superior
-one when data has heterogeneous variance or is heavy-tailed.
-
-Besides no strong modeling assumption is required, our proposed method is robust against data set 
-with variance heterogeneity, unequal number of observations, non-Gaussian noise and high-dimension. 
-Such robustness is important in practice because for any given data, it's hard to tell whether
-its clusters have different number of observations, is the variance equal for each cluster,
-or what underlying noise distribution it has. The weakness of our proposed method is 
-that its theory assumes only week correlation between ``predictor'' columns and ``response'' 
-columns. In practice, many data sets don't exhibit high correlations between columns, where
-our proposed method can be safely applied. In case the high correlation does exist, 
-the correlation-corrected version of Gabriel cross-validation method in section $4.3$ can be used if we can 
-assume common covariance structure. Other procedure such as leave out most columns for clustering 
-may also be used to reduce its effect. However, the theory for such procedure has not
-been fully developed and it could be the future research topic.
-
-Another situation where our proposed method cannot be directly used is when the clusters 
-are non-convex. In such situation, $k$-means itself doesn't work well, for example two 
-concentric circles share the same cluster center \citep{hastie2009elements}. However, it's 
-possible that our proposed Gabriel cross-validation method can be used on the transformed data set. In the
-concentric circles case where spectral clustering is appropriate, our proposed method can be applied 
-on the eigenvector subspace of the graph Laplacian matrix inside the spectral clustering algorithm
-to find the optimal $k$. This can also be the future research topic.
+In this paper, we proposed a new approach to estimate the number of clusters
+to be used in $k$-means clustering. The intuition behind our proposed method
+is to transform the unsupervised learning problem into a supervised learning
+problem via a form of Gabriel cross validation.  We proved that our method is
+self-consistent, and we analyzed its behavior in some special cases of
+Gaussian mixture models.  Using simulations and real data examples, we
+demonstrated that our method has good performance, competitive with existing
+approaches. The simulations and empirical benchmarks demonstrate the
+advantages of our method. In the yeast cell cycle application, our method
+was able to identify meaningful gene groups with a small number of clusters.
+
+There are many other clustering algorithms that get used in practice besides
+$k$-means. We suspect that it should be possible to apply our method in the
+context of a spectral clustering procedure, after transforming by the
+eigenvectors of the Laplacian matrix. For other clustering schemes, including
+versions of hierarchical clustering, we are less certain about the viability
+of Gabriel cross-validation.  It is an open question as to whether Gabriel
+cross-validation can be extended to other clustering methods, and whether such
+extensions will perform well in practice.
+
+For $k$-means clustering, Gabriel cross-validation is competitive with
+other model selection methods, especially in the presence of high-dimensional,
+heterogeneous, or heavy-tailed data.
+
+%% Besides no strong modeling assumption is required, our proposed method is
+%% robust against data set with variance heterogeneity, unequal number of
+%% observations, non-Gaussian noise and high-dimension.  Such robustness is
+%% important in practice because for any given data, it's hard to tell whether
+%% its clusters have different number of observations, is the variance equal for
+%% each cluster, or what underlying noise distribution it has. The weakness of
+%% our proposed method is that its theory assumes only week correlation between
+%% ``predictor'' columns and ``response'' columns. In practice, many data sets
+%% don't exhibit high correlations between columns, where our proposed method can
+%% be safely applied. In case the high correlation does exist, the
+%% correlation-corrected version of Gabriel cross-validation method in section
+%% $4.3$ can be used if we can assume common covariance structure. Other
+%% procedure such as leave out most columns for clustering may also be used to
+%% reduce its effect. However, the theory for such procedure has not been fully
+%% developed and it could be the future research topic. 
+%% 
+%% 
+%% Another situation where our proposed method cannot be directly used is when
+%% the clusters are non-convex. In such situation, $k$-means itself doesn't work
+%% well, for example two concentric circles share the same cluster center
+%% \citep{hastie2009elements}. However, it's possible that our proposed Gabriel
+%% cross-validation method can be used on the transformed data set. In the
+%% concentric circles case where spectral clustering is appropriate, our proposed
+%% method can be applied on the eigenvector subspace of the graph Laplacian
+%% matrix inside the spectral clustering algorithm to find the optimal $k$. This
+%% can also be the future research topic.
 
@@
 
-\section{Clustering scree plot}
+\section{Clustering scree plot examples}
 
@@
 
-\section{Review of consistency results for clustering}
-
-Some proposed methods provide theoretical consistency result for choosing $k$.
-Assume data is a mixture of $G$ $p$-dimension clusters with equal prior,
-identically distributed with common covariance matrix and finite fourth
-moments in each dimension, \citet{sugar2003finding} shows the Jump method will
-pick the correct $k$ under the conditions that cluster centers are
-sufficiently separated and an appropriate transformation is used.
-Specifically, let $\Delta\sqrt{p}$ denotes the minimum Euclidean distance
-between cluster centers, \citet{sugar2003finding} shows that the jump will be
-maximized at $k=G$ given that $\Delta > 6$ and the existence of a positive $Y$
-such that
-\[
-  \left(\frac{p\Delta^2W}{9G}\right)^{-Y}
-    + \left(
-        W \left[\frac{2^{2H^*(X)}}{K^2_{max}2\pi e}\right]
-        - (\frac{\Delta}{6})^2(1-W)
-      \right)^{-Y} < 2
-\]
-and $\left(\frac{p\Delta^2W}{9G}\right)^{-Y} <1/2$, where $H^*(X)$ is the
-minimum entropy of the cluster membership over each dimension and $W =1-
-\frac{6^4V_{\mathbf{X}}}{(\Delta^2-36)^2}$ with $V_{\mathbf{X}} = \var\left(
-  1/p ||\mathbf{X}-\mu_j||^2_{\Gamma^{-1}} \mid \mathbf{X}
-\hspace{0.05in}\text{in jth cluster}\right)$
-
-\cite{tibshirani2005cluster} proved that the prediction strength method is
-consistent under the setting that the observations in $p$-dimension are
-uniformly distributed within one unit from their respective cluster centers,
-with minimum distance between the $G$ cluster centers is four unit. Under such
-well-separated setting, they show that 
-\[ ps(G) = 1+ o_p(1), \hspace{0.5in} \underset{k \geq G+1}{\text{sup}} ps(k)\leq \frac{2}{3}+ o_p(1) \]
-therefore $\hat{k}$ is consistent in estimating $G$.  When data does come from
-mixture of Gaussian distribution, the model-based method of
-\citet{fraley2002model} is consistent in estimating $k$ in low dimension.
-
-Let $\Psi$ denotes any given base-clustering algorithm (e.g $k$-means), and
-$\Psi(Z,k)$ the clustering obtained by applying $\Psi$ on data $Z$ with
-parameter $k$. Also let $k_0$ be the optimal number of cluster and
-$d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ be the measure of distance between clustering
-$\Psi(Z_1,k)$ and $\Psi(Z_2,k)$ where $Z_1$ and $Z_2$ are two independent
-samples from population. \citet{wang2010consistent} shows that their proposed
-method will consistently select $k = k_0$, given such $k_0$ exists and
-$d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ converges to zero at proper rate $r_k$.
-However, $k_0$ is defined as the most stable number for $\Psi$ among all the
-possible $k$, instead of the number of components in mixture model or
-well-separated compact regions commonly seen in literature. Therefore, their
-consistent result actually is convergent result, i.e. the selected $k$
-converges to its limit $k_0$ given such limit exists.
+%% \section{Review of consistency results for clustering}
+%% 
+%% Some proposed methods provide theoretical consistency result for choosing $k$.
+%% Assume data is a mixture of $G$ $p$-dimension clusters with equal prior,
+%% identically distributed with common covariance matrix and finite fourth
+%% moments in each dimension, \citet{sugar2003finding} shows the Jump method will
+%% pick the correct $k$ under the conditions that cluster centers are
+%% sufficiently separated and an appropriate transformation is used.
+%% Specifically, let $\Delta\sqrt{p}$ denotes the minimum Euclidean distance
+%% between cluster centers, \citet{sugar2003finding} shows that the jump will be
+%% maximized at $k=G$ given that $\Delta > 6$ and the existence of a positive $Y$
+%% such that
+%% \[
+%%   \left(\frac{p\Delta^2W}{9G}\right)^{-Y}
+%%     + \left(
+%%         W \left[\frac{2^{2H^*(X)}}{K^2_{max}2\pi e}\right]
+%%         - (\frac{\Delta}{6})^2(1-W)
+%%       \right)^{-Y} < 2
+%% \]
+%% and $\left(\frac{p\Delta^2W}{9G}\right)^{-Y} <1/2$, where $H^*(X)$ is the
+%% minimum entropy of the cluster membership over each dimension and $W =1-
+%% \frac{6^4V_{\mathbf{X}}}{(\Delta^2-36)^2}$ with $V_{\mathbf{X}} = \var\left(
+%%   1/p ||\mathbf{X}-\mu_j||^2_{\Gamma^{-1}} \mid \mathbf{X}
+%% \hspace{0.05in}\text{in jth cluster}\right)$
+%% 
+%% \cite{tibshirani2005cluster} proved that the prediction strength method is
+%% consistent under the setting that the observations in $p$-dimension are
+%% uniformly distributed within one unit from their respective cluster centers,
+%% with minimum distance between the $G$ cluster centers is four unit. Under such
+%% well-separated setting, they show that 
+%% \[ ps(G) = 1+ o_p(1), \hspace{0.5in} \underset{k \geq G+1}{\text{sup}} ps(k)\leq \frac{2}{3}+ o_p(1) \]
+%% therefore $\hat{k}$ is consistent in estimating $G$.  When data does come from
+%% mixture of Gaussian distribution, the model-based method of
+%% \citet{fraley2002model} is consistent in estimating $k$ in low dimension.
+%% 
+%% Let $\Psi$ denotes any given base-clustering algorithm (e.g $k$-means), and
+%% $\Psi(Z,k)$ the clustering obtained by applying $\Psi$ on data $Z$ with
+%% parameter $k$. Also let $k_0$ be the optimal number of cluster and
+%% $d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ be the measure of distance between clustering
+%% $\Psi(Z_1,k)$ and $\Psi(Z_2,k)$ where $Z_1$ and $Z_2$ are two independent
+%% samples from population. \citet{wang2010consistent} shows that their proposed
+%% method will consistently select $k = k_0$, given such $k_0$ exists and
+%% $d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ converges to zero at proper rate $r_k$.
+%% However, $k_0$ is defined as the most stable number for $\Psi$ among all the
+%% possible $k$, instead of the number of components in mixture model or
+%% well-separated compact regions commonly seen in literature. Therefore, their
+%% consistent result actually is convergent result, i.e. the selected $k$
+%% converges to its limit $k_0$ given such limit exists.
 
@@
 
-\section{Confusion matrix} \label{app:confusion}
-
-\begin{table}
-%\captionsetup{justification=centering}
-\caption{\label{table:confusion-complete} Complete confusion matrix comparing $5$ clusters to $30$ clusters }
-\begin{tabular}{lccccc|c}
-  \hline
- & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
-  \hline
-  Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
-  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
-  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
-  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
-  Cluster 5 & 152 & 0 & 0 & 0 & 0 & 152 \\ 
-  Cluster 6 & 1 & 2 & 12 & 85 & 4 & 104 \\ 
-  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
-  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
-  Cluster 9 & 142 & 0 & 0 & 0 & 5 & 147 \\ 
-  Cluster 10 & 12 & 27 & 24 & 20 & 6 & 89 \\ 
-  Cluster 11 & 0 & 5 & 81 & 8 & 0 & 94 \\ 
-  Cluster 12 & 3 & 68 & 5 & 3 & 1 & 80 \\ 
-  Cluster 13 & 0 & 1 & 12 & 81 & 5 & 99 \\ 
-  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
-  Cluster 15 & 60 & 53 & 2 & 0 & 0 & 115 \\ 
-  Cluster 16 & 7 & 0 & 0 & 9 & 83 & 99 \\ 
-  Cluster 17 & 9 & 0 & 72 & 1 & 1 & 83 \\ 
-  Cluster 18 & 3 & 16 & 10 & 67 & 5 & 101 \\ 
-  Cluster 19 & 62 & 2 & 0 & 1 & 8 & 73 \\ 
-  Cluster 20 & 11 & 0 & 6 & 2 & 65 & 84 \\ 
-  Cluster 21 & 1 & 13 & 6 & 44 & 6 & 70 \\ 
-  Cluster 22 & 33 & 15 & 28 & 9 & 0 & 85 \\ 
-  Cluster 23 & 2 & 18 & 2 & 25 & 22 & 69 \\ 
-  Cluster 24 & 1 & 38 & 45 & 1 & 0 & 85 \\ 
-  Cluster 25 & 2 & 57 & 16 & 1 & 0 & 76 \\ 
-  Cluster 26 & 5 & 3 & 23 & 19 & 0 & 50 \\ 
-  Cluster 27 & 9 & 0 & 30 & 5 & 20 & 64 \\ 
-  Cluster 28 & 0 & 7 & 23 & 0 & 38 & 68 \\ 
-  Cluster 29 & 27 & 2 & 10 & 1 & 11 & 51 \\ 
-  Cluster 30 & 3 & 5 & 41 & 1 & 10 & 60 \\ \hline
-  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ \hline
-\end{tabular}
-\end{table}
+%% \section{Confusion matrix} \label{app:confusion}
+%% 
+%% \begin{table}
+%%   \footnotesize
+%% %\captionsetup{justification=centering}
+%% \caption{\label{table:confusion-complete} Complete confusion matrix comparing
+%% $5$ clusters to $30$ clusters }
+%% \begin{tabular}{lccccc|c}
+%%   \toprule
+%%  & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
+%%   \midrule
+%%   Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
+%%   Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
+%%   Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
+%%   Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
+%%   Cluster 5 & 152 & 0 & 0 & 0 & 0 & 152 \\ 
+%%   Cluster 6 & 1 & 2 & 12 & 85 & 4 & 104 \\ 
+%%   Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
+%%   Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
+%%   Cluster 9 & 142 & 0 & 0 & 0 & 5 & 147 \\ 
+%%   Cluster 10 & 12 & 27 & 24 & 20 & 6 & 89 \\ 
+%%   Cluster 11 & 0 & 5 & 81 & 8 & 0 & 94 \\ 
+%%   Cluster 12 & 3 & 68 & 5 & 3 & 1 & 80 \\ 
+%%   Cluster 13 & 0 & 1 & 12 & 81 & 5 & 99 \\ 
+%%   Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
+%%   Cluster 15 & 60 & 53 & 2 & 0 & 0 & 115 \\ 
+%%   Cluster 16 & 7 & 0 & 0 & 9 & 83 & 99 \\ 
+%%   Cluster 17 & 9 & 0 & 72 & 1 & 1 & 83 \\ 
+%%   Cluster 18 & 3 & 16 & 10 & 67 & 5 & 101 \\ 
+%%   Cluster 19 & 62 & 2 & 0 & 1 & 8 & 73 \\ 
+%%   Cluster 20 & 11 & 0 & 6 & 2 & 65 & 84 \\ 
+%%   Cluster 21 & 1 & 13 & 6 & 44 & 6 & 70 \\ 
+%%   Cluster 22 & 33 & 15 & 28 & 9 & 0 & 85 \\ 
+%%   Cluster 23 & 2 & 18 & 2 & 25 & 22 & 69 \\ 
+%%   Cluster 24 & 1 & 38 & 45 & 1 & 0 & 85 \\ 
+%%   Cluster 25 & 2 & 57 & 16 & 1 & 0 & 76 \\ 
+%%   Cluster 26 & 5 & 3 & 23 & 19 & 0 & 50 \\ 
+%%   Cluster 27 & 9 & 0 & 30 & 5 & 20 & 64 \\ 
+%%   Cluster 28 & 0 & 7 & 23 & 0 & 38 & 68 \\ 
+%%   Cluster 29 & 27 & 2 & 10 & 1 & 11 & 51 \\ 
+%%   Cluster 30 & 3 & 5 & 41 & 1 & 10 & 60 \\ \midrule
+%%   Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ \bottomrule
+%% \end{tabular}
+%% \end{table}
 

### 2016-06-17 cvclust.tex
revise appendix D

@@
 
-\begin{itemize}
-	\item For each $k = 1,2,...,k_{max}$
+In Wold cross-validation, we perform ``speckled'' hold-outs in each fold, leaving out a
+random subset of the entries of the data matrix $\dataX \in \R^{N \times P}$. For each value
+of $k$ and each fold, we perform the following set of actions to get an
+estimate of cross-validation error, $\CV(k)$, which we average over all folds.
+
 \begin{enumerate}
-		\item Randomly draw some entries in $\mathbf{X}$ missing, keep those hold-out values in vector $V_{true}$
-		\item Impute the missing values with column mean or $0$, denote the imputed data as $\mathbf{X}_{new}$
-		\item Apply the iterative procedure below until converge or stopping criteria reached
-		\begin{itemize}
-			\item Apply $K$-mean on data set $\mathbf{X}_{new}$ with parameter $k$
-			\item Substitute each observation in $\mathbf{X}_{new}$ by its nearest center, get new data $\mathbf{X}^c_{new}$ ($\mathbf{X}_{new}$ keep the same)
-			\item Replace (impute) those imputed values in $\mathbf{X}_{new}$ with the corresponding entries in $\mathbf{X}^c_{new}$
-			\item Calculate the difference between the old and newly imputed values, check whether or not they coincide (converge) 
-		\end{itemize}
-		\item Obtain the last imputed entry values of converged $\mathbf{X}_{new}$, denote it by $V_{converge}$
-		\item Calculate the prediction error  $Error_k = ||V_{true} - V_{converge}||^2$ 
+  \item Randomly partition the set of indices $\{ 1, 2, \dotsc, N \} \times \{
+    1, 2, \dotsc, P \}$ into a train set $S_\text{train}$ and a test set
+    $S_\text{test}$.
+
+  \item Apply a $k$-means fitting procedure that can handle missing data to
+    the training data $\{ \dataX_{i,j} : (i,j) \in S_{\text{train}} \}$. This
+    gives a set of cluster means $\mu(1), \dotsc, \mu(k) \in \R^P$ and
+    cluster labels for the rows, $G_1, G_2, \dotsc, G_N$.
+
+  \item Compute the cross-validation error as
+    \[
+      \CV(k) = \sum_{(i,j) \in S_{\text{test}}} \{ \dataX_{i,j} - \mu_j(G_i) \}^2,
+    \]
+    where $\mu_j(G_i)$ denotes the $j$th component of $\mu(G_i)$.
 \end{enumerate}
-	\item For each cross-validation fold, repeat above procedure and obtain the $Error_k$ for each $k$
-	\item Average $Error_k$ across all folders for each $k$, and then select the $k$ corresponding to the minimum average $Error_k$ 
-\end{itemize}
 

### 2016-06-17 cvclust.tex
revise abstract and keywords

@@
 devised to choose the number of clusters automatically, but they often rely on
-strong modeling assumptions. We propose a data-driven approach to estimate the
-number of clusters based on a novel form of cross-validation. This differs
-from ordinary cross-validation, because clustering is fundamentally an
-unsupervised learning problem. Simulation and real data analysis results show
-that our proposed method outperforms existing methods, especially in
-high-dimensional settings with heavy-tailed data.
+strong modeling assumptions. This paper proposes a data-driven approach to
+estimate the number of clusters based on a novel form of cross-validation. The
+proposed method differs from ordinary cross-validation, because clustering is
+fundamentally an unsupervised learning problem. Simulation and real data
+analysis results show that the proposed method outperforms existing methods,
+especially in high-dimensional settings with heterogeneous or heavy-tailed
+noise. In a yeast cell cycle dataset, the proposed method finds a parsimonious
+clustering with interpretable gene groupings.
 \end{abstract}
@@
 \noindent%
-{\it Keywords:} clustering, unsupervised learning, model selection
+{\it Keywords:} clustering, cross-validation, $k$-means, model selection, unsupervised learning
 \vfill

### 2016-06-18 cvclust.tex
Correct a few typos

@@
 reveal homogeneous sub-groups within a large sample; (b) data
-reduction, where high-dimensional item attribute vectors get reduced discrete
+reduction, where high-dimensional item attribute vectors get reduced to discrete
 cluster labels \citep{jain1999data}.
@@
 ability'' to drive the model selection \citep{hastie2009elements}.  Most
-existing methods for choosing $k$ instead rely on on explicit or implicit
+existing methods for choosing $k$ instead rely on explicit or implicit
 assumptions about the data distribution, including it shape, scale, and
@@
 Several authors advocate choosing $k$ by performing a sequence of hypothesis
-tests will null and alternative hypotheses of the form $H_0 : k = k_0$ and
+tests with null and alternative hypotheses of the form $H_0 : k = k_0$ and
 $H_1: k > k_0$, starting with $k_0 = 1$ and proceeding sequentially with
@@
 When the degrees of freedom decreases, the tail becomes more flat and the
-Gaussian assumption becomes more appropriate.  For most methods, their
+Gaussian assumption becomes more inappropriate.  For most methods, their
 performances are relatively stable until the tail gets very heavy. In the case
@@
 the $P = 16$ key votes.  For each vote, each legislator either votes
-positively (``yea'') negatively (``nay''). We removed legislators with missing
+positively (``yea'') or negatively (``nay''). We removed legislators with missing
 votes.  This results in $N = 232$ remaining records, with $124$ democrat and

### 2016-06-20 cvclust.tex
fix a few typos

@@
   \title{\bf Estimating the number of clusters using cross-validation}
-  \author{Wei Fu and Patrick O. Perry \\
+  \author{Wei Fu and Patrick O.\ Perry \\
   Stern School of Business, New York University}
@@
 covariance matrix $\Theta$, then $Q Z$ has covariance matrix $Q \Theta Q^\T$,
-which has expectation $E(Q \Theta Q^\T) = \tr(\Theta) I$.
+which has expectation $\E(Q \Theta Q^\T) = \tr(\Theta) I$.
 
@@
 \cmidrule(l){2-4}
-Method   & Congress Voting & Breast Cancer & Brain Tumours \\
+Method   & Congress Voting & Breast Cancer & Brain Tumors \\
 \midrule                    
@@
 cytoplasmic translation and genes encoding ribosomes. Cluster~5, the second
-periodic cluster, contains genes that participate cell-cycle processes, along
+periodic cluster, contains genes that participate in cell-cycle processes, along
 with DNA replication and DNA repair.

### 2016-06-20 cvclust.tex
minor updates to jasa template

@@
 \usepackage{booktabs}
+\usepackage[singlelinecheck=false]{caption}
 \usepackage{enumerate}
@@
 \usepackage{float}
-%\usepackage{footnote}
-%\usepackage[bottom]{footmisc}
-\usepackage[singlelinecheck=false]{caption}
-
-\usepackage{graphicx,psfrag,epsf}
-\usepackage[space]{grffile}
+\usepackage{graphicx}
 \usepackage{natbib}
-\usepackage{url} % not crucial - just used below for the URL
+\usepackage{url}
 
@@
 % NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{1}
+\newcommand{\blind}{0}
 
@@
 
-\newtheorem{theorem}{Theorem}
 \newtheorem{lemma}{Lemma}
-\newtheorem{claim}{Claim}
 \newtheorem{proposition}{Proposition}
-\newtheorem{definition}{Definition}
-\newtheorem{corollary}{Corollary}
 
@@
 \newcommand{\toas}{\overset{\mathit{a.s.}}{\to}}
-
 \newcommand{\Oh}{O}
@@
 
-%\bibliographystyle{natbib}
-
 \def\spacingset#1{\renewcommand{\baselinestretch}%
@@
 
-\if1\blind
+\if0\blind
 {
   \title{\bf Estimating the number of clusters using cross-validation}
-  \author{Wei Fu and Patrick O.\ Perry \\
+  \author{Wei Fu and Patrick O.\ Perry \hspace{.2cm} \\
   Stern School of Business, New York University}
@@
 
-\if0\blind
+\if1\blind
 {
@@
 \noindent%
-{\it Keywords:} clustering, cross-validation, $k$-means, model selection, unsupervised learning
+{\it Keywords:} clustering, $k$-means, model selection, unsupervised learning
 \vfill
@@
 \bibliographystyle{apalike}
-
 \bibliography{references}
+
 \end{document}

### 2017-02-08 cvclust.tex
pdfminorversion=4, for mac
fix figure captions
add Wei's edits

--- 2016-06-18 cvclust.tex
@@
 steps~\ref{step:gabriel-cluster} and~\ref{step:gabriel-classify}.  In this
-report, we use $k$-means as the clustering algorithm.  For the classification
-step, we compute the mean value of $X$ for each class; we assign an
+report, we use $k$-means \citep{hartigan1979algorithm} as the clustering algorithm 
+and develop the theoretical properties of the proposed method based on $k$-means. 
+For the classification step, we compute the mean value of $X$ for each class; we assign an
 observation to class $g$ if that class has the closest mean (randomly breaking
@@
 It will suffice to prove self-consistency for a single fold of the
-cross-validation procedure.  As in in section~\ref{sec:gabriel-cv-algorithm}
+cross-validation procedure.  As in section~\ref{sec:gabriel-cv-algorithm}
 we assume that the $P$ variables of the data set have been partitioned into
@@
 
+Given $\mu(g) = \bigl(\muX(g),\muY(g)\bigr)$ are distinct for $g = 1, \dotsc, K$,
+~\ref{asn:self-consistency-distinct-mux} and ~\ref{asn:self-consistency-distinct-muy}
+are reasonable given not too small dimension $p$ and $q$.
 The result statement of Proposition~\ref{prop:self-consistency} follows from 
@@
 \end{proof}
-
+The inequality in Proposition~\ref{prop:twoclust} means that the proposed method
+would correctly pick $k = 2$ given that $(\muX, \muY)$ and 
+$(-\muX, -\muY)$ are not too close. 
 We confirm the result of Proposition~\ref{prop:twoclust} with a simulation.
@@
 We generate three clusters in $6$ dimensions. Each cluster has $1000$ or $500$
-mean zero multivariate normal observations with identity covariance matrix.
+multivariate normal observations with identity covariance matrix.
 We add $r$ dimensions of noise to the data, randomly generated from a uniform
@@
 We generate eight clusters in $P$ dimensions, with $P$ taking values in
-$\{10,20,...,100\}$.  Each cluster has $100$ or $50$ mean zero multivariate
+$\{10,20,...,100\}$.  Each cluster has $100$ or $50$ multivariate
 normal observations with identity covariance matrix.

### 2017-02-08 cvclust.tex
change line endings (DOS to UNIX)
add more clarification on the assumptions

@@
 It will suffice to prove self-consistency for a single fold of the
-cross-validation procedure.  As in section~\ref{sec:gabriel-cv-algorithm}
-we assume that the $P$ variables of the data set have been partitioned into
-$p$ predictor variables represented in vector~$X$ and $q$ response variables
+cross-validation procedure.  As in section~\ref{sec:gabriel-cv-algorithm} we
+assume that the $P$ variables of the data set have been partitioned into $p$
+predictor variables represented in vector~$X$ and $q$ response variables
 represented in vector~$Y$.  The $N$ observations have been divided into two
-sets: $n$ train observations and $m$ test observations.  The following theorem
+sets: $n$ train observations and $m$ test observations.  We state the
+assumptions for the self-consistency result in terms of a specific split; for
+the result to hold in general, with high probability, these assumptions would
+have to hold with high probability for a random split. The following theorem
 gives conditions for Gabriel cross-validation to recover the true number of
@@
 
-Given $\mu(g) = \bigl(\muX(g),\muY(g)\bigr)$ are distinct for $g = 1, \dotsc, K$,
-~\ref{asn:self-consistency-distinct-mux} and ~\ref{asn:self-consistency-distinct-muy}
-are reasonable given not too small dimension $p$ and $q$.
-The result statement of Proposition~\ref{prop:self-consistency} follows from 
+The proposition states that our method works well in the absence of noise,
+when each observation is equal to its cluster center.  The essential
+assumption here is assumption~\ref{asn:self-consistency-noiseless}, which states
+that there is no noise.  If we are willing to assume, say, that the cluster
+centers~$\mu(g) = \bigl(\muX(g),\muY(g)\bigr)$ for $g = 1, \dotsc, K$ were
+randomly drawn from a distribution with a density over $\R^{p+q}$, then
+assumptions~\ref{asn:self-consistency-distinct-mux}
+and~\ref{asn:self-consistency-distinct-muy} will hold with probability one for
+all splits of the data. Likewise, if the clusters are not too small (relative
+to $n$ and $m$), then assumptions~\ref{asn:self-consistency-train}
+and~\ref{asn:self-consistency-test} will likely hold for a random split of the
+data into test and train.
+
+Proposition~\ref{prop:self-consistency} follows from
 Lemmas~\ref{lem:self-consistency1} and~\ref{lem:self-consistency2}, which we
-will now state and prove.
+now state and prove.
 
@@
 
+Proposition~\ref{prop:single-2d} tells us that Gabriel cross-validation fails
+when there is strong correlation between the variables. To get around this, in
+practice we will transform the data to reduce correlation before performing
+cross-validation. We detail this approach in Section~\ref{sec:corr-correct}.
+
 \subsection{Two clusters in two dimensions}
@@
 We will now analyze a simple two-cluster setting, and derive conditions for
-Gabriel cross-validation to correctly prefer $k=2$ clusters to $k=1$. In the
-statement of the result, $\Phi(\cdot)$ and $\varphi(\cdot)$ denote the
-standard normal cumulative distribution function and density, respectively.
+Gabriel cross-validation to correctly prefer $k=2$ clusters to $k=1$.  The
+main assumption of the proposition is that the cluster centers are not too
+close. The precise definition of ``too close'' is stated in terms of
+$\Phi(\cdot)$ and $\varphi(\cdot)$, the standard normal cumulative
+distribution function and density, respectively.  The inequality is hard to
+interpret directly, but we show the boundary between ``too close'' and ``well
+separated'' in Fig.~\ref{fig:overlap-color_plot}, after the proof of the
+proposition.
 
@@
 where $\muX \geq 0$ and $\muY \geq 0$. 
-If 
+If the cluster centers are well separated, specifically such that
 \(
@@
 \end{proof}
-The inequality in Proposition~\ref{prop:twoclust} means that the proposed method
-would correctly pick $k = 2$ given that $(\muX, \muY)$ and 
-$(-\muX, -\muY)$ are not too close. 
 We confirm the result of Proposition~\ref{prop:twoclust} with a simulation.

### 2017-02-08 cvclust.tex
add acknowledgements

@@
 %% can also be the future research topic.
+
+\section*{Acknowledgements}
+
+We thank Rob Tibshirani for getting us started on this problem and
+for providing code for some initial simulations. We thank Art~Owen for
+providing us with a summary of the relevant theory on $k$-means clustering,
+and for giving us feedback on our theoretical results.  We also thank Cliff
+Hurvich, Josh Reed, and Jeff~Simonoff, for providing comments on an early
+draft of this manuscript and for suggesting further avenues of inquiry.
 

### 2018-02-27 wei-revision/simulation-results-table.R
add Wei's revision + notes

@@ -1,0 +1,53 @@
+setwd("C:/Users/fuwei/Desktop/cvclust/New simulation/demo/bench")
+
+nclusters <- list("setting1" = 6L,
+		     "setting2" = 3L,
+			 "setting3" = 8L,
+			 "setting4" = 3L,
+			 "setting5" = 5L)
+
+Scale_X <-  list("setting1" = seq(0.0,0.9,0.1),
+		     "setting2" = 6*seq(0,9,1),
+			 "setting3" = seq(10,100,10),
+			 "setting4" = c(1,seq(5,45,5)),
+			 "setting5" = seq(2,11,1))
+
+Label_X <- list("setting1" = "Rho",
+		    "setting2" = "Noise Dimension",
+			"setting3" = "Data Dimension",
+			"setting4" = "Max Variance Ratio",
+			"setting5" = "Degree of freedom")
+
+
+
+for (s in list.dirs(full.names=FALSE, recursive=FALSE)){
+	print(s)
+	print('------------------------------------------------')
+	rm(Result.path)
+
+	####----------------------Read from RDS and count the K-mean results-----------------
+	Counts <- function(x){
+		vector <- rep(NA,11)
+		for(j in 1:10){
+			vector[j] <- sum(x==j,na.rm=TRUE)
+		}
+		vector[11] <- 100-sum(vector[1:10]) ## vector[11] stores the NAs
+		return(vector)
+	}
+	###----------------------------------------------------------------------
+	Result.path <- file.path(s,"result.rds")
+	if(!file.exists(Result.path)){stop("file path doesn't exists")}
+
+	Fit_result <- readRDS(Result.path)
+
+	for(i in 1:length(Fit_result)){
+		count <- as.data.frame(apply(Fit_result[[i]], 2, Counts))
+		colnames(count) <- as.character(Scale_X[[s]])
+		rownames(count) <- c("1","2","3","4","5","6","7","8","9","10","NA")		
+		print(xtable(count,digits=0,caption = names(Fit_result)[i]),table.placement="H")
+	}
+	print('------------------------------------------------')
+}
+
+
+

### 2018-02-27 cvclust.tex
fix Ref 1 comments

@@
 than items in different groups \citep{hartigan1975clustering}.  Such
-procedures are used in two main applicaitons: (a) exploratory analysis, where clusters
+procedures are used in two main applications: (a) exploratory analysis, where clusters
 reveal homogeneous sub-groups within a large sample; (b) data
@@
 existing methods for choosing $k$ instead rely on explicit or implicit
-assumptions about the data distribution, including it shape, scale, and
+assumptions about the data distribution, including its shape, scale, and
 correlation structure.
@@
 Section~\ref{sec:meth}.  In Section~\ref{sec:self-consistent}, we prove that
-our method is self-consistent. Then, in Section~\ref{sec:theory-gaussian}, we
+our method is self-consistent in certain situations, meaning that in the
+absence of noise, it recovers the correct $k$. Then, in
+Section~\ref{sec:theory-gaussian}, we
 analyze the performance of our method in the presence of Gaussian noise. The
@@
 \subsection{Problem statement}
+\label{sec:problem-statement}
 
@@
 \[
-  W(A) = \sum_{i=1}^{n} \min_{a \in A} \|x_i - a\|^2.
+    W(A) = \sum_{i=1}^{n} \min_{g \in \{ 1, \dotsc, k \} } \|x_i - a_g\|^2.
 \]
@@
 \[
-  g(x) = \argmin_{g \in \{1, \dotsc, k\}} \|x - a_g\|^2,
+  G(x) = \argmin_{g \in \{1, \dotsc, k\}} \|x - a_g\|^2,
 \]
@@
 $N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
-``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
-``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
-matrix to get $\dataX$, a matrix with the $r$th test subset as its trailing
-rows.  We partition $\dataX$ as
+``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  We do this
+by randomly assigning a fold number $r_i$ in the range $\{1, \dotsc, K\}$
+for each row~$i$. For each ``fold'' $r$ in the range $1, \dotsc, K$, we
+permute the rows of the data matrix to get $\dataX$, a matrix with the
+$r$th test subset, the rows $i$ with $r_i = r$ as its trailing rows.  
+We partition $\dataX$ as
 \[
@@
 \]
-We use the training rows $[ \Xtrain\ \Ytrain ]$ to fit a regression model
+We use the training rows $[ \Xtrain\ \Ytrain ]$, containing roughly $N
+(K-1)/K$ of the observations to fit a regression model
 $\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
@@
     \textbf{Cluster:}
-    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding the
-    assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ and the
-    cluster means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
-    be the assigned cluster for row $i$.
+    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding
+    $k$ cluster centers, denoted $\bmuY_1, \dotsc, \bmuY_k$, identified
+    with the labels $1, \dotsc, k$. This also yields the
+    assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ defined analogously
+    to the assignment rule $G$ from Sec.~\ref{sec:problem-statement}, which
+    for any input $Y$, yields the label of the closest cluster center.
+    Set $\hGY_i = \hGY(Y_i)$ to be the assigned cluster label for row~$i$: the
+    label of the closest cluster mean to $Y_i$.
   \item \label{step:gabriel-classify}
@@
     and take $\hGY_1, \dotsc, \hGY_n$ to be corresponding class labels.  Use
-    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to train a classifier
-    $\hGX : \R^p \to \{ 1, \dotsc, k \}$.
+    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to fit a model that can
+    predicts a label from input vector $X$; denote the fitted
+    prediction function by
+    $\hGX : \R^p \to \{ 1, \dotsc, k \}$. In principle this prediction method
+    can be arbitrary but in the sequel we will use linear discriminant
+    analysis.
   \item \label{step:gabriel-predict}
@@
 In Section~\ref{sec:self-consistent}, we prove that this procedure is
-self-consistent, in the sense that it recover the correct value of $k$ in the
+self-consistent, in the sense that it recovers the correct value of $k$ in the
 absence of noise. Then, in Section~\ref{sec:theory-gaussian}, we analyze some
@@
 
-An important property of any estimation procedure is that in the absence of of
+An important property of any estimation procedure is that in the absence of
 noise, the procedure correctly estimates the truth. This property is called
@@
 this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
-We will assume that there are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with
+We will assume that there are $K$ ``true cluster centers''
+$\mu(1), \dotsc,\mu(K)$, with
 the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),

### 2018-03-01 cvclust.tex
handle AE1.

@@
 \citet{gabriel2002biblot}, who proposed holding out a single column and a
-single row at each fold.
+single row at each fold. Gabriel's cross validation is analogous to
+leave-one-out cross-validation, while Owen and Perry's generalization handles
+more general leave-outs.
 
@@
 
-In the sequel, we extend Gabriel cross-validation to the problem of selecting
-the number of clusters, $k$, automatically, and we provide theoretical and
-empirical support analogous to the consistency results proved by
-\citet{owen2009bi}.
+Gabriel's cross-validation the Owen and Perry generalization are designed
+specifically for choosing the rank of a low-rank matrix approximation like the
+singular value decomposition or the non-negative matrix factorization. While
+for some purposes it makes sense to view $k$-means clustering as a sort of low
+rank approximation to the data matrix, the discrete nature of clustering (each
+observation gets assigned to one of $k$ possible clusters) imposes additional
+structure on the problem that makes the prior procedures not directly
+applicable.
+
+In the sequel, we extend the version of Gabriel cross-validation described by
+Owen and Perry to the problem of selecting the number of clusters, $k$. To do
+so, we introduce a classification procedure into the cross-validation steps
+that takes the place of least squares regression in the matrix factorization
+problem formation. Our new procedure is able to select the number of clusters,
+$k$, automatically. For this procedure, we provide theoretical and empirical
+support analogous to the consistency results proved by \citet{owen2009bi}.
 

### 2018-03-01 cvclust.tex
AE point 2

@@
 
-In this section, we perform a set of simulations to evaluate the performance of our
-proposed method and the associated correlation correction described in
-Section~\ref{sec:corr-correct}.
-We compare our method
-with a basket of competing methods including the Gap statistic
+In this section, we perform a set of simulations to evaluate the performance
+of our proposed method and the associated correlation correction described in
+Section~\ref{sec:corr-correct}.  We compare our method with a basket of
+competing methods including the Gap statistic
 \citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
@@
 \citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
-and Bootstrap stability \citep{fang2012selection}.
-We use the default parameter settings for all competing methods. For Gabriel
-cross-validation, we perform $2$-fold cross-validation on the columns~($p=q$)
-and $5$-fold cross-validation on the rows~($m=n/4$).
-We also compare with Wold cross-validation, which we describe in
-Appendix~\ref{sec:wold-cv}.
+and Bootstrap stability \citep{fang2012selection}.  We use the default
+parameter settings for all competing methods.  For Gabriel cross-validation,
+we perform $2$-fold cross-validation on the columns~($p=q$) and $5$-fold
+cross-validation on the rows~($m=n/4$).  We also compare with Wold
+cross-validation, which we describe in Appendix~\ref{sec:wold-cv}.
+
+For the competing methods, some might take issue with the fact that we are
+using default parameter settings.  In many other benchmark comparisons, one
+would choose the tuning parameters using some form of cross-validation. In our
+case, this is not an option. We are in an \emph{unsupervised} setting, so
+there is no clear way to set the tuning parameters. We must rely on the
+defaults from the implementations of these methods.
 

### 2018-03-01 cvclust.tex
address third AE point

@@
   \item \label{step:transform}
-    Compute the eigendecomposition $\hat{\Sigma} = \Gamma\Lambda\Gamma^\T$.
+    Compute a decomposition $\hat{\Sigma} = \Gamma\Lambda\Gamma^\T$
+    where $\Gamma$ has orthonormal columns and
+    $\Lambda$ is diagonal and positive-definite.
     Choose a random (Haar distributed) $P \times P$ orthogonal matrix $Q$.
@@
 demonstrate the effectiveness of our ad-hoc adjustment procedure.
+
+In many applications, including our application in the sequel to a yeast cell
+cycle dataset, the number of observations $N$ is below the number of
+features~$P$. When this happens, the sample covariance matrix $\hat \Sigma$
+will not have full rank~$P$, and will be a poor estimate of the population
+covariance matrix. In these situations, though, it will still be the case that
+a decomposition $\hat \Sigma = \Gamma \Lambda \Gamma^\T$ exists with $\Gamma$
+having orthonormal columns and $\Lambda$ being diagonal and positive-definite.
+It will also be the case that $\Lambda^{-1/2}$ and hence $\tilde{\dataX}$
+exist. What fails in these situation is that, since $\hat \Sigma$ is no longer
+a good estimate of $\Sigma$, it will no longer be the case that our correction
+reliably transforms the noise covariance to $I$. Still, in these situations,
+applying the correlation correction may give better results than using the
+original untransformed data. In practice we inspect the resulting estimate
+$\hat k$ both with and without the correlation correction, and if the two
+answers agree, we can be more confident in our answer.
 

### 2018-03-02 cvclust.tex
start reviewer 2 comments

@@
 
+The proof of Proposition~\ref{prop:single-2d} relies on normality assumptions
+in two crucial ways: (i) assuming that $X$ and $Y$ have symmetric marginal
+distributions; and (ii) assuming that $E(Y|X) = \rho X$. We could relax the
+normality assumption by instead supposing that $X$ and $Y$ can be decomposed
+as $X = Z_1$ and $Y = \rho Z_1 + \sqrt{1-\rho^2}Z_2$ where $Z_1$ and $Z_2$ are
+independent random variables with mean zero, unit variance and symmetric
+distribution. Any bivariate distribution can be put in this form for
+uncorrelated $Z_1$ and $Z_2$, but only special distributions, including the
+bivariate normal, can be put in the form with independent $Z_1$ and $Z_2$.
+
 We confirm the result of Proposition~\ref{prop:single-2d} with a simulation.
@@
 \end{proof}
+
+In principle the proof of Proposition~\ref{prop:twoclust} generalizes to
+non-normal distributions easily, provided expressions for $\E(Y | Y > 0)$ and
+the other related quantities are available. Different distributions for
+$(X,Y)$ give rise to different cutoffs for when $k = 1$ cluster is preferred
+to $k = 2$. The normal distribution is only special here in that the required
+conditional expectations and the boundary where $\CV(1) = \CV(2)$ are
+computable in closed form.
+
 We confirm the result of Proposition~\ref{prop:twoclust} with a simulation.

### 2018-03-09 cvclust.tex
typos, update Rev 1 response

@@
 
-Gabriel's cross-validation the Owen and Perry generalization are designed
+Gabriel's cross-validation and the Owen and Perry generalization are designed
 specifically for choosing the rank of a low-rank matrix approximation like the
@@
 It will also be the case that $\Lambda^{-1/2}$ and hence $\tilde{\dataX}$
-exist. What fails in these situation is that, since $\hat \Sigma$ is no longer
+exist. What fails in these situations is that, since $\hat \Sigma$ is no longer
 a good estimate of $\Sigma$, it will no longer be the case that our correction

### 2018-03-09 cvclust.tex
add general-d proof to appendix

@@
 \end{proof}
+
+We have stated and prooved Proposition~\ref{prop:twoclust} for data in $d = 2$
+dimensions, but there is nothing essential about the particulars of the
+dimension; we generalize the result to arbitrary dimension in
+Appendex~\ref{app:twoclust-anyd}.
 
@@
 
-\section{Analysis of Gabriel method: Single cluster in more than two dimensions}
+\section{Analysis of single cluster in more than two dimensions}
 \label{sec:single-general-dim}
 
-\begin{proposition}
+\begin{proposition}\label{prop:single-general-dim}
 Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
@@
 \]
+\end{proof}
+
+\section{Analysis of two clusters in more than two dimensions}
+\label{app:twoclust-anyd}
+
+\begin{proposition}\label{prop:twoclust-anyd}
+Suppose that $\{(X_i,Y_i)\}_{i=1}^{n+m}$ is data from a single fold of Gabriel
+    cross-validation, where each $(X,Y)$ pair in with $X \in \mathbb{R}^P$ and $Y \in
+    \mathbb{R}^Q$ is an independent draw
+from an equiprobable mixture of two multivariate normal distributions with
+identity covariance. Suppose that the first mixture component has mean
+$\mu = (\muX, \muY)$ and the second has mean $-\mu = (-\muX, -\muY)$,
+where $\muX \geq 0$ and $\muY \geq 0$. 
+If the cluster centers are well separated, specifically such that
+\(
+  2 \varphi(\muY) + \muY + 2 \muY \Phi(\muY) < 4 \muY \Phi(\muX),
+\)
+then $\CV(2) < \CV(1)$ with probability tending to one as $m$ and $n$ increase.
+\end{proposition}
+
+\begin{proof}
+There are two clusters $G_1$ and $G_2$, where observations from $G_1$ are distributed as
+\[	\mathcal{N}\left( \begin{pmatrix} 
+    \mu^X \\
+    \mu^Y \\
+  \end{pmatrix}, \mathbf{I} \right)	\]
+and observations from $G_2$ are distributed as
+\[	\mathcal{N}\left( \begin{pmatrix} 
+    -\mu^X \\
+    -\mu^Y \\
+  \end{pmatrix}, \mathbf{I} \right)	\]
+here $\mu^X$ and $\mu^Y$ are all vectors. Let $G_i$ be the true cluster where observation $i$ is generated from, by assumption
+\[	P(G_i=G_1) = P(G_i=G_2) = 1/2	\]
+
+To simplify the notation, let 
+\[
+ \begin{pmatrix} 
+    X_1 \\
+    Y_1 \\
+  \end{pmatrix}  \sim
+  \mathcal{N}\left( \begin{pmatrix} 
+    \mu^X \\
+    \mu^Y \\
+  \end{pmatrix}, \mathbf{I} \right)
+\]  
+denote the observations from $G_1$ and 
+\[
+ \begin{pmatrix} 
+    X_2 \\
+    Y_2 \\
+  \end{pmatrix}  \sim
+  \mathcal{N}\left( \begin{pmatrix} 
+    -\mu^X \\
+    -\mu^Y \\
+  \end{pmatrix}, \mathbf{I} \right)
+\]  
+denote the observations from $G_2$
+
+Further, let's denote $\mu^X = \lambda_x e_x$ where $\lambda_x$ is a scalar denotes the distance of $\mu^X$ from origin and $e_x$ is the unit vector point at the same direction as $\mu^X$; $\mu^Y = \lambda_y e_y$ has the same interpretation.
+
+Apply $K$-mean on the $Y$-space, where the two clusters are $\mathcal{N}\left( \mu^Y, \mathbf{I}\right)$ and  $\mathcal{N}\left( -\mu^Y, \mathbf{I}\right)$, the $K$-mean centroids are $\bar{\mu}^Y_1$ and $\bar{\mu}^Y_2$ with $\bar{\mu}^Y_1 = -\bar{\mu}^Y_2$.
+Note that the boundary between the two clusters are $e^T_y Y > 0 $.
+So
+
+\begin{align}
+\bar{\mu}^Y_1 & = E(Y \mid Y>0) \\
+ &= E(Y_1 \mid e^T_y Y_1 > 0) \cdot P(e^T_y Y_1 >0) +E(Y_2 \mid e^T_y Y_2 > 0) \cdot P(e^T_y Y_2 >0)
+\end{align}
+
+Note that $e^T_y Y_1$ projects vector $Y_1$ on the direction of $e_y$.
+And because the $e_y$ is the same direct as $\mu^Y$, it goes through the center of the sphere $\mathcal{N}\left( \mu^Y, \mathbf{I}\right)$.
+Because the covariance matrix is $\mathbf{I}$, the sphere is symmetric around $e_y$.
+Therefore,
+\begin{align}
+E(Y_1 \mid e^T_y Y_1 = a ) & = a e_y
+\end{align}
+
+Also, $Y_1 \sim \mathcal{N}\left( \mu^Y, \mathbf{I}\right)$ so $e^T_y Y_1 \sim \mathcal{N}\left( \lambda_y, 1\right)$. We have
+
+\begin{align}
+E(Y_1 \mid e^T_y Y_1 > 0)  & = E \left[  E\left(  Y_1 \mid e^T_y Y_1  \right) \mid e^T_y Y_1 > 0  \right] \\
+ from \hspace{0.05in} (3) \hspace{0.05in} above	& =  E( e^T_y Y_1 e_y  \mid e^T_y Y_1 > 0  )  \\ 
+	& = e_y E( e^T_y Y_1 \mid e^T_y Y_1 > 0  ) 
+\end{align}
+
+Because $e^T_y Y_1 \sim \mathcal{N}\left( \lambda_y, 1\right)  = \lambda_y + Z$, where $Z$ is standard normal, by Lemma $3$ from Appendix C we have 
+\begin{align}
+E( e^T_y Y_1 \mid e^T_y Y_1 > 0  ) &= E(\lambda_y +  Z \mid Z > -  \lambda_y )  \\
+ &= \lambda_y + E(Z \mid Z > -  \lambda_y ) \\
+ &= \lambda_y + \frac{\varphi(\lambda_y)}{\Phi(\lambda_y)}
+\end{align}
+
+ where $\varphi()$ and $\Phi()$ are the standard normal probability and cumulative distribution function respectively. So, by $(6)$ we have
+
+\begin{align}
+E(Y_1 \mid e^T_y Y_1 > 0)  &= \left[ \lambda_y + \frac{\varphi(\lambda_y)}{\Phi(\lambda_y)}  \right] e_y
+\end{align}
+
+Similarly, we can have
+
+\begin{align}
+E(Y_2 \mid e^T_y Y_2 > 0)  &= - E(Y_1 \mid e^T_y Y_1 < 0) \\
+	&= - e_y E( e^T_y Y_1 \mid e^T_y Y_1 < 0  )  \\
+	&= \left[ \frac{\varphi(\lambda_y)}{1-\Phi(\lambda_y)} - \lambda_y  \right] e_y 
+\end{align}
+
+Because $e^T_y Y_1 \sim \mathcal{N}\left( \lambda_y, 1\right)  = \lambda_y + Z$, it's easy to get 
+\begin{align}
+P(e^T_y Y_1 > 0) &= P(Z > - \lambda_y) \\
+	&= \Phi(\lambda_y) 
+\end{align}
+
+By symmetry, we can get 
+\begin{align}
+P(e^T_y Y_2 > 0) &= 1- \Phi(\lambda_y)
+\end{align}
+
+Put everything together, we have
+
+\begin{align}
+\bar{\mu}^Y_1 & = E(Y_1 \mid e^T_y Y_1 > 0) \cdot P(e^T_y Y_1 >0) +E(Y_2 \mid e^T_y Y_2 > 0) \cdot P(e^T_y Y_2 >0) \\
+ &=  \left[ \lambda_y + \frac{\varphi(\lambda_y)}{\Phi(\lambda_y)}  \right]\cdot \Phi(\lambda_y) e_y + \left[ \frac{\varphi(\lambda_y)}{1-\Phi(\lambda_y)} - \lambda_y  \right]\cdot\left(1- \Phi(\lambda_y)\right) e_y  \\
+ & = [2\lambda_y \Phi(\lambda_y) + 2\varphi(\lambda_y) - \lambda_y]e_y 
+\end{align}
+
+After training the classifier, because of the identity covariance matrix, the classification boundary is $e^T_x X > 0$. So the $Y$ center for observation with $e^T_x X > 0$ is
+
+\begin{align}
+\hat{\mu}^Y_1 &= E(Y_1 \mid e^T_x X_1 > 0)\cdot P(e^T_x X_1 > 0) + E(Y_2 \mid e^T_x X_2 > 0)\cdot P(e^T_x X_2 > 0)\\ 
+\text{ X independent of Y}	&= E(Y_1) \cdot P(e^T_x X_1 > 0) + E(Y_2) \cdot P(e^T_x X_2 > 0)\\ 
+  & = \mu^Y \cdot P(e^T_x X_1 > 0) -  \mu^Y \cdot P(e^T_x X_2 > 0)\\
+  & = \mu^Y \left( P(e^T_x X_1 > 0) -  P(e^T_x X_2 > 0)  \right) \\
+   & = \mu^Y \left[ \Phi(\lambda_x) -  (1-\Phi(\lambda_x))  \right] \\
+   & = (2\Phi(\lambda_x) - 1)\mu^Y \\
+    & = (2\Phi(\lambda_x) - 1)\lambda_y e_y
+\end{align}
+
+Because of symmetry and $P(G_i=G_1) = P(G_i=G_2) = 1/2$, it's sufficient to show that for observations with  $e^T_x X > 0$, if $CV(2) < CV(1)$ then the Gabriel CV method correctly picks $k=2$ over $k=1$.
+
+Similar as in the proof of Proposition~\ref{prop:single-general-dim}, by the variance and bias decomposition of MSE, the variance is the same, so only the bias influences the result.
+Note the predicted center is grand $0$ for $CV(1)$, so to see if  $CV(2) < CV(1)$ one only need to see if $||\bar{\mu}^Y_1-\hat{\mu}^Y_1||^2 < ||\hat{\mu}^Y_1 - 0||^2$, which is true if
+\[ 2\Phi(\lambda_y) + 2 \frac{\varphi(\lambda_y)}{\lambda_y} < 4\Phi(\lambda_x)-1 \] 
+this result reduces to the original result of Proposition~\ref{prop:twoclust}
+if one sets $\lambda_x = \mu^X$ and $\lambda_y = \mu^Y$.
 \end{proof}

### 2018-03-09 cvclust.tex
add response about k

@@
 Our correlation correction is not backed by a rigorous theoretical
-justification. However, the simulations and empirical validation in
-Sections~\ref{sec:simulations} and~\ref{sec:empirical-validation}
-demonstrate the effectiveness of our ad-hoc adjustment procedure.
+justification. Further, since our initial estimate $\hat k_0$ tends to
+over-estimate the true number of clusters, our covariance estimate $\hat
+\Sigma$ will often be biased. Fortunately, we do not need to estimate $\Sigma$
+with high accuracy. For our correction to be useful, we only need $\hat
+\Sigma^{-1/2}$ to have lower within-cluster correlation than the raw data. In
+our simulations and in our empirical validation study in
+Sections~\ref{sec:simulations} and~\ref{sec:empirical-validation} we
+demonstrate that despite its theoretical shortcoming, our ad-hoc correlation
+adjustment often performs well in practice.
 

### 2018-03-09 cvclust.tex
more progress

@@
 \citet{gabriel2002biblot}, who proposed holding out a single column and a
-single row at each fold. Gabriel's cross validation is analogous to
+single row at each fold. Gabriel's cross-validation is analogous to
 leave-one-out cross-validation, while Owen and Perry's generalization handles
@@
 
-Our version of Gabriel cross validation for clustering works by performing a
+Our version of Gabriel cross-validation for clustering works by performing a
 sequence of ``folds'' over the data. We use these folds to estimate a version
@@
 
-We have stated and prooved Proposition~\ref{prop:twoclust} for data in $d = 2$
+We have stated and proved Proposition~\ref{prop:twoclust} for data in $d = 2$
 dimensions, but there is nothing essential about the particulars of the
@@
 justification. Further, since our initial estimate $\hat k_0$ tends to
-over-estimate the true number of clusters, our covariance estimate $\hat
-\Sigma$ will often be biased. Fortunately, we do not need to estimate $\Sigma$
-with high accuracy. For our correction to be useful, we only need $\hat
-\Sigma^{-1/2}$ to have lower within-cluster correlation than the raw data. In
-our simulations and in our empirical validation study in
-Sections~\ref{sec:simulations} and~\ref{sec:empirical-validation} we
-demonstrate that despite its theoretical shortcoming, our ad-hoc correlation
-adjustment often performs well in practice.
+over-estimate the true number of clusters in the presence of strong
+correlation, our covariance estimate $\hat \Sigma$ will often be biased.
+Fortunately, we do not need to estimate $\Sigma$ with high accuracy. For our
+correction to be useful, we only need $\tilde{\dataX}$ to have lower
+within-cluster correlation than the raw data~$\dataX$. In our simulations and
+in our empirical validation study in Sections~\ref{sec:simulations}
+and~\ref{sec:empirical-validation} we demonstrate that despite its theoretical
+shortcoming, our ad-hoc correlation adjustment often performs well in
+practice.
 
@@
 from a multivariate normal distribution with covariance matrix $\tau I$,
-conditional on the cluster centers being well-separated (if the distance
-between any two cluster centers is less than $1$, then we re-draw a new set of
-cluster centers). We choose $\tau$ to make the probability the cluster
+conditional on the cluster points being well-separated. By ``well-separated''
+we mean that the distance from any point to its own cluster center is less
+than the distance of that point to any other cluster center by a margin of
+at least $1.0$ units. In any replicate where the clusters are not
+well-separated, we discard the replicate and generate another set of points.
+We choose $\tau$ to make the probability the cluster
 centers being well-separated on the first draw to be equal to approximately
-50\%. Many of our simulation settings are chosen to mimic the settings used by
-\cite{tibshirani2001estimating}.
+50\%.  The ``well-separated'' condition and some of our simulation settings
+are chosen to mimic those used by~\cite{tibshirani2001estimating} in
+settings (c) and (d) of their Section~6.
 
@@
 The first data set is congressional voting data consisting of voting records
-of the second session of the $98$th United States Congress,
+of the second session of the $98$th United States Congress
 \citep{schlimmer1987concept}.  This data set includes votes for legislators on
@@
 is to transform the unsupervised learning problem into a supervised learning
-problem via a form of Gabriel cross validation.  We proved that our method is
+problem via a form of Gabriel cross-validation.  We proved that our method is
 self-consistent, and we analyzed its behavior in some special cases of

### 2018-03-09 cvclust.tex
address minor points

@@
 \includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting1/Facet.pdf}
+\caption{Performance as correlation increases.}
 \label{fig:setting1}
@@
 \includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting2/Facet.pdf}
+\caption{Performance as number of noise dimensions increases.}
 \label{fig:setting2}
@@
 \includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting3/Facet.pdf}
+\caption{Performance as number of dimensions increases.}
 \label{fig:setting3}
@@
 \includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting4/Facet.pdf}
+\caption{Performance as variance heterogeneity increases.}
 \label{fig:setting4}
@@
 \includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting5/Facet.pdf}
+\caption{Performance as tail width increases.}
 \label{fig:setting5}
@@
 independent draw from a mean-zero multivariate normal distribution with 
-covariance matrix $\Sigma_{XY}= \left( \begin{smallmatrix} \Sigma_{XX} & \Sigma_{XY} \\ 
+covariance matrix $\Sigma = \left( \begin{smallmatrix} \Sigma_{XX} & \Sigma_{XY} \\ 
  \Sigma_{YX} & \Sigma_{YY} \end{smallmatrix}\right)$, with $\Sigma_{YY}$ has leading 
@@
 \begin{proof}
-Let $X$ and $Y$ be jointly multivariate normal distributed with mean $\mathbf{0}$ and covariance matrix $\Sigma_{XY}$, i.e.
-\[	(X,Y) \sim \mathcal{N} \left( \mathbf{0}, \Sigma_{XY}\right)	\]
-where $\Sigma_{XY}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} \\  \Sigma_{YX} & \Sigma_{YY} \end{bmatrix}$.
+Let $X$ and $Y$ be jointly multivariate normal distributed with mean $\mathbf{0}$ and covariance matrix $\Sigma$, i.e.
+\[	(X,Y) \sim \mathcal{N} \left( \mathbf{0}, \Sigma\right)	\]
+where $\Sigma=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} \\  \Sigma_{YX} & \Sigma_{YY} \end{bmatrix}$.
 

### 2018-10-02 JCGS-R2/cvclust.tex
JCGS revision 2

--- 2018-03-09 cvclust.tex
@@
 \DeclareMathOperator*{\cov}{cov}
+\DeclareMathOperator*{\cor}{cor}
 \DeclareMathOperator*{\tr}{tr}
@@
 \centering
-\includegraphics[width=3in]{demo/nullcorr/equal.pdf}
+\includegraphics[width=3in]{main_code/demo/nullcorr/equal.pdf}
 \caption{Cross-validation error on $10$ replicates, with the number of
@@
 \centering
-\includegraphics[width=3in]{demo/overlap/color_plot.pdf}
+\includegraphics[width=3in]{main_code/demo/overlap/color_plot.pdf}
 \caption{Number of times $k=2$ is selected out of $10$ replicates for each
@@
 \citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
-and Bootstrap stability \citep{fang2012selection}.  We use the default
-parameter settings for all competing methods.  For Gabriel cross-validation,
+Slope method \citep{fujita2014non} and Bootstrap stability \citep{fang2012selection}. 
+We use the default parameter settings for all competing methods.  For Gabriel cross-validation,
 we perform $2$-fold cross-validation on the columns~($p=q$) and $5$-fold
@@
 \centering
-\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting1/Facet.pdf}
+\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting1/Facet.pdf}
 \caption{Performance as correlation increases.}
@@
 \centering
-\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting2/Facet.pdf}
+\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting2/Facet.pdf}
 \caption{Performance as number of noise dimensions increases.}
@@
 \centering
-\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting3/Facet.pdf}
+\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting3/Facet.pdf}
 \caption{Performance as number of dimensions increases.}
@@
 correlation-corrected version tend to work better in higher dimensions.
+Slope method also works better in higher dimensions.
 
@@
 \centering
-\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting4/Facet.pdf}
+\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting4/Facet.pdf}
 \caption{Performance as variance heterogeneity increases.}
@@
 This setting demonstrates that most existing methods are sensitive to variance
-heterogeneity, most notably the Gap method and the model-based BIC method. The
+heterogeneity, most notably the Gap method, the Slope method and the model-based BIC method. The
 proposed Gabriel cross-validation method and its correlation-corrected version
@@
 \centering
-\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting5/Facet.pdf}
+\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting5/Facet.pdf}
 \caption{Performance as tail width increases.}
@@
 performances are relatively stable until the tail gets very heavy. In the case
-where there are $2$ degrees of freedom, the Gap and Jump methods' performances
+where there are $2$ degrees of freedom, the Slope method, the Gap and Jump methods' performances
 deteriorate considerably relative to the Gabriel method.
@@
 Stab.    & $2$  & $2$           & $7$  \\ 
+Slope	 & $2$  & $4$           & $2$ \\
 \midrule 
@@
 		\centering
-	\includegraphics[width=5.7in, height=5.7in]{wei/split_plots.pdf}
+	\includegraphics[width=5.7in, height=5.7in]{main_code/data_examples/split_plots.pdf}
 	\caption{Yeast data set mean expression profiles. The $5$ clusters found by
@@
   \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{demo/elbow/correct-data.pdf}
+    \includegraphics[width=\linewidth]{main_code/demo/elbow/correct-data.pdf}
   \end{minipage}
@@
   \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{demo/elbow/correct-withinss.pdf}
+    \includegraphics[width=\linewidth]{main_code/demo/elbow/correct-withinss.pdf}
   \end{minipage}
@@
   \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{demo/elbow/incorrect-data.pdf}
+    \includegraphics[width=\linewidth]{main_code/demo/elbow/incorrect-data.pdf}
   \end{minipage}
@@
   \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{demo/elbow/incorrect-withinss.pdf}
+    \includegraphics[width=\linewidth]{main_code/demo/elbow/incorrect-withinss.pdf}
   \end{minipage}
@@
 
+The condition that ensures $\CV(1) < \CV(2)$ is tight but difficult to
+understand intuitively. We make understand this condition better by
+noting that it is equivalent to a condition on the correlation between the
+two linear combinations of $X$ and $Y$ variables:
+\begin{equation}\label{eqn:clust-anyd}
+    \cor(u_1^\T \Sigma_{YX} X, u_1^\T Y) < 1/2.
+\end{equation}
+A stronger condition that is easier to understand, then, is that the canonical
+correlation between $X$ and $Y$ is below $1/2$; when the latter is true the
+former follows and $\CV(1) < \CV(2)$. A necessary but not sufficient
+guarantee of the stronger condition is that all correlations between variables
+are below $1/2$.
+
+One might object to the condition~\eqref{eqn:clust-anyd} being stated in terms
+of a particular $(X, Y)$ split of the data.  In practice, one might want the
+condition to hold for all possible splits of the data. A sufficient guarantee
+that~\eqref{eqn:clust-anyd} holds for any $(X, Y)$ split of the data is that
+the data covariance matrix $\Sigma$ is diagonal. The ad-hoc adjustment we
+propose in Sec.~\ref{sec:corr-correct} attempts to transform the data to this
+form.
+
 \section{Analysis of two clusters in more than two dimensions}

### 2018-10-02 JCGS-R2/main_code/demo/bench/01_generate_data.R

@@ -1,0 +1,59 @@
+#!/usr/bin/Rscript --vanilla
+setwd("C:/Users/fuwei/Desktop/JCGS-R2/demo/bench") ## set the working directory correctly
+
+library("MASS")
+library("parallel")
+source("settings.R") # settings
+
+
+# Use the L'Ecuyer RNG so that we can have multiple streams of
+# random numbers
+rng_kind <- "L'Ecuyer-CMRG"
+rng_normal_kind <- "Inversion"
+
+
+nrep <- 100
+
+for (s in names(settings)) {
+    if (!dir.exists(s)) {
+        dir.create(s)
+    }
+	for(c in 1:10){
+		if (!dir.exists(file.path(s, paste0("C" ,c)))){
+			dir.create(file.path(s, paste0("C" ,c)))
+		}
+			
+		filename <- file.path(s, paste0("C" ,c), "replicates.rds")
+		if (!file.exists(filename)) {
+
+			cat("generating replicates for '", s, "' with Case ", c, "\n", sep="")
+
+			# generate the list of seeds
+			set.seed(settings[[s]]$seed, rng_kind, rng_normal_kind)
+			seeds <- vector("list", nrep)
+			seeds[[1L]] <- .Random.seed
+			for (r in seq_len(nrep - 1L))
+				seeds[[r+1L]] <- nextRNGStream(seeds[[r]])
+
+			# generate the replicates
+			replicates <- vector("list", nrep)
+
+			pb <- txtProgressBar(0, nrep, style=3)
+			for (r in seq_len(nrep)) {
+				# set the seed for this replicate
+				.Random.seed <<- seeds[[r]]
+				replicates[[r]] <- settings[[s]]$simulate(c)
+
+				# save the state of the RNG
+				replicates[[r]][["rng"]] <- list(seed = .Random.seed,
+												 kind = rng_kind,
+												 normal_kind = rng_normal_kind)
+
+				setTxtProgressBar(pb, r)
+			}
+			close(pb)
+
+			saveRDS(replicates, filename)
+		}
+	}
+}

### 2018-10-02 JCGS-R2/main_code/demo/bench/04_plot_results.R

@@ -1,0 +1,87 @@
+setwd("C:/Users/fuwei/Desktop/JCGS-R2/demo/bench") ## set the working directory correctly
+
+library("PropCIs")
+library(ggplot2)
+source("methods.R")
+
+nclusters <- list("setting1" = 6L,
+		     "setting2" = 3L,
+			 "setting3" = 8L,
+			 "setting4" = 3L,
+			 "setting5" = 5L)
+
+Scale_X <-  list("setting1" = seq(0.0,0.9,0.1),
+		     "setting2" = 6*seq(0,9,1),
+			 "setting3" = seq(10,100,10),
+			 "setting4" = c(1,seq(5,45,5)),
+			 "setting5" = seq(2,11,1))
+
+Label_X <- list("setting1" = "Rho",
+		    "setting2" = "Noise Dimension",
+			"setting3" = "Data Dimension",
+			"setting4" = "Max Variance Ratio",
+			"setting5" = "Degree of freedom")
+
+
+for (s in list.dirs(full.names=FALSE, recursive=FALSE)){
+	## remove the object so the if cond checkd out for next iter
+	rm(Result_data)
+	rm(Fit_result)
+
+	Result.path <- file.path(s,"result.rds")
+
+	if(!file.exists(Result.path)){
+		Fit.result <- list()
+
+		for (m in names(methods)) {
+	 		Matrix_result <- matrix(NA,100,10)
+			for(c in 1:10){
+				f.replicates <- file.path(s, paste0("C" ,c), "method",paste0(m ,".rds"))
+				if(file.exists(f.replicates)){
+					Matrix_result[,c]<- readRDS(f.replicates)
+				}
+			}
+			Fit.result[[m]] = Matrix_result
+		}
+		saveRDS(Fit.result, Result.path)
+	}
+	####----------------------Read from RDS and make plot-----------------
+
+	Fit_result <- readRDS(Result.path)
+	
+	for(i in 1:length(Fit_result)){
+		Proportion <- as.numeric(apply(Fit_result[[i]], 2, function(x) sum(x==nclusters[[s]],na.rm=T)))
+		X_axis <- Scale_X[[s]]
+		Methods <- rep(names(Fit_result)[i],10)
+
+		if(!exists("Result_data")){
+			Result_data <- as.data.frame(cbind(X_axis,Proportion,Methods))
+		}else{
+			Result_data <- rbind(Result_data,as.data.frame(cbind(X_axis,Proportion,Methods)))
+		}	
[…]

### 2018-10-02 JCGS-R2/main_code/demo/bench/05_create_tables.R

@@ -1,0 +1,53 @@
+setwd("C:/Users/fuwei/Desktop/JCGS-R2/main_code/demo/bench") ## set the working directory correctly
+
+library('xtable')
+
+nclusters <- list("setting1" = 6L,
+		     "setting2" = 3L,
+			 "setting3" = 8L,
+			 "setting4" = 3L,
+			 "setting5" = 5L)
+
+Scale_X <-  list("setting1" = seq(0.0,0.9,0.1),
+		     "setting2" = 6*seq(0,9,1),
+			 "setting3" = seq(10,100,10),
+			 "setting4" = c(1,seq(5,45,5)),
+			 "setting5" = seq(2,11,1))
+
+Label_X <- list("setting1" = "Rho",
+		    "setting2" = "Noise Dimension",
+			"setting3" = "Data Dimension",
+			"setting4" = "Max Variance Ratio",
+			"setting5" = "Degree of freedom")
+
+
+
+for (s in list.dirs(full.names=FALSE, recursive=FALSE)){
+	print(s)
+	print('------------------------------------------------')
+	rm(Result.path)
+
+	####----------------------Read from RDS and count the K-mean results-----------------
+	Counts <- function(x){
+		vector <- rep(NA,11)
+		for(j in 1:10){
+			vector[j] <- sum(x==j,na.rm=TRUE)
+		}
+		vector[11] <- 100-sum(vector[1:10]) ## vector[11] stores the NAs
+		return(vector)
+	}
+	###----------------------------------------------------------------------
+	Result.path <- file.path(s,"result.rds")
+	if(!file.exists(Result.path)){stop("file path doesn't exists")}
+
+	Fit_result <- readRDS(Result.path)
+
+	## create Latex table to be exported
+	for(i in 1:length(Fit_result)){
+		count <- as.data.frame(apply(Fit_result[[i]], 2, Counts))
+		colnames(count) <- as.character(Scale_X[[s]])
+		rownames(count) <- c("1","2","3","4","5","6","7","8","9","10","NA")		
+		print(xtable(count,digits=0,caption = names(Fit_result)[i]),table.placement="H")
+	}
+	print('------------------------------------------------')
+}

### 2018-10-02 JCGS-R2/main_code/demo/bench/methods.R

@@ -1,0 +1,72 @@
+# demo/bench/methods.R
+#
+# Depends:
+#   library("cstab")
+#   library("bcv")
+#   library("devtools")
+#   library("cluster")
+#   library("e1071")
+#   library("mclust")
+#   library("MASS")
+#   library("nnet")
+#   load_all("../../lib/fpc")
+#   load_all("../../lib/NbClust")
+#   source("../../lib/NbClust.R")
+#   source("../../code/classify.R")
+#   source("../../code/cluster.R")
+#   source("../../code/gabriel.R")
+#   source("../../code/jump.R")
+#   source("../../code/wold.R")
+#   
+
+methods <- list(
+    "gabriel-nearest-5x2" = function(x, maxcenters) {
+        cv <- cv.kmeans.gabriel(x, 5, 2, maxcenters,
+                                classify.method="nearest")
+        cv$centers
+    },
+	"gabriel-corr-correct2" = function(x, maxcenters) {
+       gabriel_cor_correct(x, maxcenters, type = 2)
+    },
+    "wold" = function(x, maxcenters) {
+        Wold_holdout(x, 5, max.k=maxcenters, Errortol=0.01)
+    },
+    "gap" = function(x, maxcenters) {
+        Gap <- cluster::clusGap(x, FUN = cluster_kmeans, K.max = maxcenters)
+        which.max(Gap[[1]][,3])
+    },
+    "BIC" = function(x, maxcenters) {
+        mcluster <- mclust::Mclust(x, G = 1:maxcenters,
+                                   control=mclust::emControl(itmax=c(10000L, 10000L)))
+        mcluster$G
+    },
+    "CH" = function(x, maxcenters) {
+        Ch <- NbClust::NbClust(x, min.nc = 2, max.nc = maxcenters,
+                               method = "kmeans", index = "ch")
+        Ch$Best.nc[[1]]
+    },
+    "Hartigan" = function(x, maxcenters) {
+        Hartigan <- NbClust::NbClust(x, min.nc = 2, max.nc = maxcenters,
+                                     method = "kmeans", index = "hartigan")
+        Hartigan$Best.nc[[1]]
+    },
+    "Jump" = function(x, maxcenters) {
+        Jump <- jump(x, maxcenters, plotjumps=FALSE, trace=FALSE)
+        Jump$maxjump
+    },
+    "PS" = function(x, maxcenters) {
+        PS <- fpc::prediction.strength(x, Gmin=2, Gmax=maxcenters,
[…]

### 2018-10-02 JCGS-R2/main_code/demo/nullcorr/01_run_simulation.R

@@ -1,0 +1,69 @@
+#!/usr/bin/Rscript --vanilla
+
+
+library("MASS")
+source("../../code/classify.R")
+source("../../code/cluster.R")
+
+
+set.seed(7627579) # from random.org
+
+ntrain <- 10000
+ntest <- ntrain
+n <- ntest + ntrain
+train <- seq_len(ntrain)
+test <- ntrain + seq_len(ntest)
+
+predictor <- 1
+response <- 2
+
+nrep <- 10
+
+sim <- NULL
+
+for (r in seq_len(nrep)) {
+    z <- matrix(rnorm(n * 2), n, 2)
+
+    cverr <- function(correlation, centers, prior)
+    {
+        r <- correlation
+        coef <- cbind(rbind(sqrt((1 + r) / 2),  sqrt((1 - r) / 2)),
+                      rbind(sqrt((1 + r) / 2), -sqrt((1 - r) / 2)))
+        x <- z %*% coef
+
+        if (centers == 1) {
+            fit <- colMeans(x[train,response,drop=FALSE])
+            err <- scale(x[test,response], center=fit, scale=FALSE)
+        } else {
+            cl <- cluster_kmeans(x[train,response], centers=centers)
+            cluster <- factor(cl$cluster, levels=seq_len(centers))
+            fit <- classify_lda(x[train,predictor,drop=FALSE], cluster, prior)
+            pred <- predict(fit, x[test,predictor,drop=FALSE])$class
+            err <- x[test,response] - cl$centers[pred,]
+        }
+
+        mse <- mean(err^2)
+        mse
+    }
+
+
+    correlation <- seq(-1, 1, length.out=21)
+    for (prior in c("equal", "proportions")) {
+        for (centers in c(1, 2, 3, 4, 5)) {
+            mse <- sapply(correlation, function(r) cverr(r, centers, prior))
+
+            sim <- rbind(sim,
+                         data.frame(replicate = r,
+                                    prior = prior,
+                                    correlation = correlation,
+                                    mse = mse,
+                                    centers = centers))
+        }
+    }
+}
+
+sim$replicate <- as.factor(sim$replicate)
+sim$centers <- as.factor(sim$centers)
+
+write.csv(sim, "nullcorr.csv", row.names=FALSE)
+

### 2019-01-10 JCGS-R2/response/response.tex
start response

\documentclass[11pt]{article}

\usepackage{amsmath}
\usepackage{framed}
\usepackage{fullpage}
\usepackage{hyperref} % must be last usepackage
\usepackage{color}
\usepackage[authoryear]{natbib}

\newcommand{\hilight}[1]{\colorbox{yellow}{#1}}
\newcommand{\todo}{\hilight{TODO}}

\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\let\PBS=\PreserveBackslash

\title{``Estimating the number of clusters using cross-validation''
Summary of changes to the revised manuscript}
\author{Wei Fu \and Patrick O. Perry}

\begin{document}

\maketitle
\setlength{\parindent}{0ex}
\setlength{\parskip}{1ex}
\vspace{2ex}

We thank the editors and reviewers for their time reading our submission, and
for providing many helpful suggestions for improvement.  We also thank you for
the opportunity to revise and resubmit our manuscript.  We have revised our
manuscript carefully taking into account your comments.

Detailed responses to reviewer comments follow below.

\section*{Response to Editor}

\begin{itemize}

\item \textit{Fix the figure captions.}

\textbf{Response:} Per your instruction, all figure captions now include
(1) what is the plot about, (2) specific details of plot, like what type of
display and how variables are mapped, (3) the most important thing that the
reader should learn.

\todo

\item \textit{Cite R packages.}

\textbf{Response:} We added citations for all packages used in our work: \todo

% You need to cite R and any packages used in your work, like ggplot2.

\item \textit{Reprodicibility.}

\textbf{Response:} \todo

% Your code is quite a mess. It is not clear to me how to reproduce the results
% you have reported. There is code in two different tar.gz files. Which is the
% code to use? In each there are several folders. This all needs to be
% organised so that it is easy for the reader to produce the results that you
% have reported in the paper.

\end{itemize}

\section*{Response to Associate Editor}

\begin{itemize}

\item \textit{Compare to the slope method}

\todo

\end{itemize}

\section*{Response to Reviewer 1}

(No further comments.)

\section*{Response to Reviewer 2}

\begin{itemize}

\item \textit{Bottom of page 4. The authors need to clarify the notation}

\textbf{Response:} \todo

%The fold numbers r and r_i are not well-defined. Please maintain the
%notational consistency. You defined only "r_i". What do you mean by: 'For each
%"fold" r'?

\item \textit{Page 6. Clarify that $n \gg K$, $m \gg L$.}

\textbf{Response:} \todo

%By my understanding, the authors must assume further that
%
%n  >> K
%m >> K	
%
%That is, they need to ensure that they are choosing n and m larger that K.
%This limitation should be clarified to the reader.

\item \textit{Section 2.2. Clarify.}

\textbf{Response:} \todo

%The authors need to clarify if and how the methods are applied
%to a non-regression setting.

\item \textit{Section 3. Note that slope is self consistent.}

\textbf{Response:} \todo

%Section 3. I am very confident that the slope statistics (Fujita et al) is
%also self-consistent under the absence of noise. This follows basically from
%the properties of finite differences.  On the other hand, the original
%Silhouette method is not self-consistent. The authors could explore this
%easily in this section in one paragraph.

\item \textit{Proposition 1 improvements.}

\textbf{Response:} \todo

We chose to keep the proof in the main text, but if there are space
constraints we will take your suggestion and move the proof to an appendix.

% 5. Proposition 1. 
% 
% 5.1 In order to check condition (iv), it seems the user must know something about K.
% 
% 5.2 Should it be CV(k)>CV(K)?
% 
% 5.3 My feeling is that once the reader fully understand the ingredients of the
% procedure, Proposition one renders trivial.  The authors could provide direct
% arguments for that. This seems to follow straightforward from the properties
% of estimation in a regression setting.

\item {\textit Proposition 2 improvements.}

\textbf{Response:} \todo

% Please, define the indicator function.

\item \textit{Page 8, line 14. Generalization.}

\textbf{Response:} \todo

We clarified what we meant by saying the proof ``generalizes to non-normal
distributions easily''. You rightly point out that it will not generalize to
all possible situations, but only to distributions with finite moments and
cases where both clusters have the same distribution.

\item \textit{Figure 2. Please define correctly the x,y axis}

\textbf{Response:} \todo

\item \textit{Page 20, first paragraph.}

\textbf{Response:} Your comment was that ``Proposition 2 only shows that if
$|\rho|<0.5$ the method works correctly.''Did you mean page 14 instead?  We
think you may have meant that our statement in the first paragraph of  page
14, ``Proposition~2 tells us that Gabriel cross-validation fails when there is
strong correlation between the variables'' is too strong.

\item \textit{Compare to standardized Euclidean norm $k$-means.}

\textbf{Response:} \todo

%10. Page 22, line 3. Maybe the standardized Euclidean norm in the k-means
%algorithm could produce good results.

\item \textit{Compare to slope}

\textbf{Response:} \todo

% 11. Page 23, line 35. Under normal distributions, the slope statistics
% proposed by Fujita et al. also works very good if the clusters are well
% separated. There are two R packages for computing the slope statistics: 
% 
% https://www.rdocumentation.org/packages/anocva/versions/0.1.1/topics/optimalSlope
% 
% and:
% 
% https://cran.r-project.org/web/packages/cstab/cstab.pdf
% 
% 
% 12. Page 25, line 31. The Slope statistic also works well under correlation.
% See the papers simulations (section 5).  Moreover, in Fujita et al., it is
% showed that some of the methods used here have bad performances under some
% scenarios which slope statistic was very good (high dimensional data, t and
% exponential distributions for each cluster, correlated items, Gaussian and
% uniform, heterogeneity).

\end{itemize}

\end{document}

### 2019-01-10 JCGS-R2/response/response.tex
reviewer comments

@@
 
-\item {\textit Proposition 2 improvements.}
+\item {\textit Proposition 2. Define the indicator function.}
 
-\textbf{Response:} \todo
+\textbf{Response:} Done.
 
-% Please, define the indicator function.
+\item \textit{Page 18, line 14. Generalization conditions.}
 
-\item \textit{Page 8, line 14. Generalization.}
+\textbf{Response:}
 
-\textbf{Response:} \todo
+We clarified what we meant by saying our proof generalized:
+\textit{
+In principle the proof of Proposition~3 generalizes to
+non-normal location families of distributions easily, provided that
+both cluster distributions are in the same family and expressions for
+$E(Y | Y > 0)$ and the other related quantities are available (and finite).
+}
 
-We clarified what we meant by saying the proof ``generalizes to non-normal
-distributions easily''. You rightly point out that it will not generalize to
-all possible situations, but only to distributions with finite moments and
-cases where both clusters have the same distribution.
+You rightly point out that it will not generalize to all possible situations,
+but only to distributions with finite moments and cases where both clusters
+have the same distribution.
 
@@
 
-\textbf{Response:} \todo
+\textbf{Response:} Done.
 
@@
 
-\item \textit{Compare to standardized Euclidean norm $k$-means.}
+\item \textit{Page 22, the standardized Euclidean norm $k$-means could produce
+good results.}
 
-\textbf{Response:} \todo
+\textbf{Response:} You may be correct, but we do not want to overwhelm the
+reader with two many variations of the same algorithm. The correlation-correction
+we propose is a slightly more general version of Euclidean standardization,
+and seems to perform adequately.
 
@@
 
-\textbf{Response:} \todo
+\textbf{Response:} As you suggested, we added comparisons to the slope method,
+using the \texttt{cstab} package. In our simulations, slope performs very well
+with noise dimensions and high-dimensional data; its performance degrades with
+correlated clusters, variance heterogeneity, and heavy-tailed data. Certainly
+there are some situations where slope is better than our method, but no method
+uniformly dominates the others. In our simulations, our correlation-corrected
+Gabriel CV method seems to be a good default choice.
 

### 2019-01-10 JCGS-R2/cvclust.tex
reviewer comments

@@
   \bmuX_j = \frac{\sum_{i=1}^{n} 1\{ \hGY_i = j \} X_i}
-                 {\sum_{i=1}^{n} 1\{ \hGY_i = j \}}.
-\]
+                 {\sum_{i=1}^{n} 1\{ \hGY_i = j \}},
+\]
+where $1\{ . \}$ denotes the indicator function.
 The classification rule $\hGX$ is defined by
@@
 In principle the proof of Proposition~\ref{prop:twoclust} generalizes to
-non-normal distributions easily, provided expressions for $\E(Y | Y > 0)$ and
-the other related quantities are available. Different distributions for
-$(X,Y)$ give rise to different cutoffs for when $k = 1$ cluster is preferred
-to $k = 2$. The normal distribution is only special here in that the required
-conditional expectations and the boundary where $\CV(1) = \CV(2)$ are
-computable in closed form.
+non-normal location families of distributions easily, provided that
+both cluster distributions are in the same family and expressions for
+$\E(Y | Y > 0)$ and the other related quantities are available (and finite).
+Different distributions for $(X,Y)$ give rise to different cutoffs for when $k
+= 1$ cluster is preferred to $k = 2$. The normal distribution is only special
+here in that the required conditional expectations and the boundary where
+$\CV(1) = \CV(2)$ are computable in closed form.
 

### 2019-01-10 JCGS-R2/response/response.tex
prop comments

@@
 
-\textbf{Response:} \todo
+\textbf{Response:} We fixed the typo you spotted ($CV(k) > CV(K)$), and we
+added a note that the proposition is purely for theoretical purposes:
+``The assumptions here are not checkable in practice, but the proposition
+suggest that Gabriel cross-validation might give a reasonable answer, at least
+when the noise level is low.''
 
-We chose to keep the proof in the main text, but if there are space
-constraints we will take your suggestion and move the proof to an appendix.
+You also suggested that we omit the proposition entirely. We chose to keep it,
+but if there are space constraints we will take your suggestion and move the
+proof to an appendix.
 

### 2019-01-11 JCGS-R2/response/response.tex
R2 self-consistency note

@@
 
-\item \textit{Section 3. Note that slope is self consistent.}
+\item \textit{Section 3. Note that other methods are self consistent.}
 
-\textbf{Response:} \todo
+\textbf{Response:} Per your suggestion, we added this paragraph:
+
+``Self-consistency by itself does not mean that a procedure performs well in
+practice. Self-consistency deals only with the no-noise situation.  Many other
+methods for selecting $K$ are likely self-consistent, but not all perform the
+same in practice. Self-consistency only suggests (but does not guarantee) that
+a procedure might perform reasonably when the noise level is low.''
+
+We did not specifically single out the slope statistic or Silhouette method
+specifically here because we have not taken the time to analyze those methods
+and we are unaware of proofs of their self-consistency or lack thereof.
 

### 2019-01-11 JCGS-R2/cvclust.tex
prop comments
R2 self-consistency note

@@
 
+Self-consistency by itself does not mean that a procedure performs well in
+practice. Self-consistency deals only with the no-noise situation.  Many other
+methods for selecting $K$ are likely self-consistent, but not all perform the
+same in practice. Self-consistency only suggests (but does not guarantee) that
+a procedure might perform reasonably when the noise level is low.
+
 It will suffice to prove self-consistency for a single fold of the
@@
 \end{enumerate}
-Then $\CV(k) < \CV(K)$ for $k < K$, and $\CV(k) = \CV(K)$ for $k > K$, so that
+Then $\CV(k) > \CV(K)$ for $k < K$, and $\CV(k) = \CV(K)$ for $k > K$, so that
 Gabriel cross-validation correctly chooses $k = K$.
@@
 The proposition states that our method works well in the absence of noise,
-when each observation is equal to its cluster center.  The essential
+when each observation is equal to its cluster center.  The assumptions here
+are not checkable in practice, but the proposition suggest that Gabriel
+cross-validation might give a reasonable answer, at least when the noise level
+is low.
+
+The essential
 assumption here is assumption~\ref{asn:self-consistency-noiseless}, which states

### 2019-01-11 JCGS-R2/response/response.tex
Require K <= N

@@
 
-\item \textit{Page 6. Clarify that $n \gg K$, $m \gg L$.}
+\item \textit{Page 6. Clarify that $N \geq K$, $M \geq K$.}
 
-\textbf{Response:} \todo
+\textbf{Response:}
+
+Added the sentence: ``For this cross-validation to be possible, we must have that
+$K \leq N$ and $L \leq M$.''
 
@@
 
-\item \textit{Section 2.2. Clarify.}
+\item \textit{Section 2.2. Clarify the context in which the method is applied.}
 
-\textbf{Response:} \todo
+\textbf{Response:}
 
-%The authors need to clarify if and how the methods are applied
-%to a non-regression setting.
+Changed the first sentence as follows: ``Our version of Gabriel
+cross-validation for clustering works is designed to find the best number of
+clusters to use for $N$~observations of a $P$-dimensional random variable with
+real-valued components.''
 

### 2019-01-11 JCGS-R2/cvclust.tex
Require K <= N

@@
 
-Our version of Gabriel cross-validation for clustering works by performing a
+Our version of Gabriel cross-validation for clustering works is designed to
+find the best number of clusters to use for $N$~observations of a
+$P$-dimensional random variable with real-valued components.
+The procedure works by performing a
 sequence of ``folds'' over the data. We use these folds to estimate a version
@@
 $(r,s)$ treats the $r$th row subset as ``test'', and the $s$th column subset
-as ``response''.  We typically take $K = 5$ and $L = 2$.  For the number of
+as ``response''. 
+For this cross-validation to be possible, we must have that
+$K \leq N$ and $L \leq M$.
+We typically take $K = 5$ and $L = 2$.  For the number of
 clusters, we select the value of $k$ that minimizes the average of $\CV(k)$

### 2019-01-11 JCGS-R2/cvclust.tex
finish R2 comments

@@
 by randomly assigning a fold number $r_i$ in the range $\{1, \dotsc, K\}$
-for each row~$i$. For each ``fold'' $r$ in the range $1, \dotsc, K$, we
-permute the rows of the data matrix to get $\dataX$, a matrix with the
+for each row~$i$. For each value $r$ in the range $1, \dotsc, K$, we define
+the $r$th fold as the matrix gotten by permuting the rows of the 
+original data matrix to get $\dataX$, a matrix with the
 $r$th test subset, the rows $i$ with $r_i = r$ as its trailing rows.  

### 2019-01-11 JCGS-R2/cvclust.tex
twocluster caption

--- 2019-01-11 JCGS-R2/cvclust.tex
@@
 by randomly assigning a fold number $r_i$ in the range $\{1, \dotsc, K\}$
-for each row~$i$. For each ``fold'' $r$ in the range $1, \dotsc, K$, we
-permute the rows of the data matrix to get $\dataX$, a matrix with the
+for each row~$i$. For each value $r$ in the range $1, \dotsc, K$, we define
+the $r$th ``fold'' as the matrix gotten by permuting the rows of the 
+original data matrix to get $\dataX$, a matrix with the
 $r$th test subset, the rows $i$ with $r_i = r$ as its trailing rows.  
@@
 \includegraphics[width=3in]{main_code/demo/nullcorr/equal.pdf}
-\caption{Cross-validation error on $10$ replicates, with the number of
-clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
-multivariate normal distribution with correlation $\rho$.  The Gabriel
-cross-validation criterion chooses the correct answer $k = 1$ whenever
+\caption{Gabriel cross-validation performance in a single-cluster simulation.
+    Gabriel cross-validation error $\CV(k)$ for $k$ ranging from 1 to~5,
+    averaged over $10$ replicates where
+    data is generated from two-dimensional
+multivariate normal distribution with correlation $\rho$.  The 
+    criterion chooses the correct answer $k = 1$ whenever
 $|\rho| < 0.5$; the criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.}
@@
 \includegraphics[width=3in]{main_code/demo/overlap/color_plot.pdf}
-\caption{Number of times $k=2$ is selected out of $10$ replicates for each
-pair of $(\muX, \muY)$. The heat map shows the frequency $k=2$ is selected by
-the algorithm, with light (blue) means $k = 1$ is preferred to $k = 2$, and
-dark (red) indicates $k = 2$ is preferred to $k = 1$.
+\caption{
+    Gabriel cross-validation performance in a two-cluster simulation.
+    For a grid of values for $\muX$ and $\muY$ ranging from $0$ to $3$,
+    we generated data according the manner described in
+    Prop.~\ref{prop:twoclust}, then computed Gabriel cross-validation
+    error $\CV(1)$ and $\CV(2)$. The heat map shows the proportion of times
+    that $\CV(2) < \CV(1)$ with
+    light (blue) indicating that $k = 1$ is preferred to $k = 2$, and
+dark (red) indicating that $k = 2$ is preferred to $k = 1$.
     The black line is the theoretical boundary determined from
-Proposition~\ref{prop:twoclust}.}
+Proposition~\ref{prop:twoclust}. The simulation confirms our theoretical
+    boundary.}
 \label{fig:overlap-color_plot}

### 2019-01-11 JCGS-R2/cvclust.tex
caption setting 1

@@
 \includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting1/Facet.pdf}
-\caption{Performance as correlation increases.}
+\caption{Setting 1 performance as correlation increases, faceted by estimation
+    method. For each estimate method, the subplot shows the proportion of
+    replicates in which estimated number of clusters agrees with the truth
+    versus the correlation between dimensions,~$\rho$. Gabriel
+    cross-validation performance degrades under moderate correlation, but
+    the correlation-corrected version performs well across all values of
+    $\rho$.}
 \label{fig:setting1}

### 2019-01-11 JCGS-R2/cvclust.tex
capture remaining simulations

@@
 \includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting2/Facet.pdf}
-\caption{Performance as number of noise dimensions increases.}
+\caption{Setting 2 performance as the number of noise dimensions increases,
+    faceted by estimation method. For each estimate method, the subplot shows the proportion of
+    replicates in which estimated number of clusters agrees with the truth
+    versus the number of noise dimensions,~$r$. Both variants of Gabriel
+    cross-validation perform reasonably under all settings of~$r$, though
+    not as well as some of the other methods.}
 \label{fig:setting2}
@@
 \includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting3/Facet.pdf}
+\caption{Setting 3 performance as the number of dimensions increases,
+    faceted by estimation method. For each estimate method, the subplot shows the proportion of
+    replicates in which estimated number of clusters agrees with the truth
+    versus the number of noise dimensions,~$r$. Both variants of Gabriel
+    cross-validation perform with perfect accuracy in the replicates
+    with $r \geq 30$ dimensions.}
+
 \caption{Performance as number of dimensions increases.}
@@
 \includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting4/Facet.pdf}
-\caption{Performance as variance heterogeneity increases.}
+\caption{Setting 4 performance as the variance heterogeneity increases,
+    faceted by estimation method. For each estimate method, the subplot shows the proportion of
+    replicates in which estimated number of clusters agrees with the truth
+    versus the variance heterogeneity parameter,  $R$. Both variants of Gabriel
+    cross-validation perform with near-perfect accuracy for all values of~$R$.}
 \label{fig:setting4}
@@
 \includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting5/Facet.pdf}
-\caption{Performance as tail width increases.}
+\caption{Setting 5 performance as the tail heaviness increases,
+    faceted by estimation method. For each estimate method, the subplot shows the proportion of
+    replicates in which estimated number of clusters agrees with the truth
+    versus the $t$~distributed noise degrees of freedom,~$\nu$, plotted
+    backwards so that $\nu$ decreases from $11$ to $2$ going from left to
+    right. Both variants of Gabriel cross-validation are robust to moderate
+    tail heaviness, with slower performance degradation than most of the other
+    estimation methods as the tail heaviness increases.}
 \label{fig:setting5}

### 2019-01-11 JCGS-R2/cvclust.tex
yeast caption

@@
 	\includegraphics[width=5.7in, height=5.7in]{main_code/data_examples/split_plots.pdf}
-	\caption{Yeast data set mean expression profiles. The $5$ clusters found by
-  the Gabriel method are on the bottom; Clusters profiled in
-  \cite{tavazoie1999systematic} are on the right.}
+	\caption{Comparison between  \cite{tavazoie1999systematic} clusters and
+    our proposed Gabriel cross-validation clusters for a Yeast gene expression
+    time series data set.  The plots on the vertical and horizontal
+    show the average expression time series for the \citeauthor{tavazoie1999systematic} 
+    clusters (1--4, 7, 8, 14, Other) and the Gabriel clusters (1--5),
+    respectively. Plot $(i,j)$ in the grid show the time series average
+    in the intersectios of  \citeauthor{tavazoie1999systematic} cluster~$i$
+    with Gabriel cluster~$j$, whenever there is overlap. The Gabriel clusters
+    capture the same general trends with a smaller number of clusters.
+    }
 	\label{fig:5_clusters}

### 2019-01-11 JCGS-R2/cvclust.tex
caption scree plot

@@
 \end{minipage}
-\caption{Left panels show the $(X,Y)$ data points; right panels
-  show the corresponding values of the within-cluster sum of squares $W_k$
-plotted against the number of clusters, $k$.}
+\caption{Scree plots for two data sets. Left panels show the sets of
+    two-dimensional data points, generated from four clusters, with plotting
+    symbol indicating the generated cluster. Right panels show the
+    corresponding values of the within-cluster sum of squares $W_k$
+    plotted against the number of clusters, $k$. The scree plot identifies
+    the correct number of clusters in the top row, but fails in the bottom
+    row.}
 \label{fig:elbow}

### 2019-01-11 JCGS-R2/response/response.tex
finish R2 comments
response to captions

@@
 
-\textbf{Response:} Per your instruction, all figure captions now include
-(1) what is the plot about, (2) specific details of plot, like what type of
-display and how variables are mapped, (3) the most important thing that the
-reader should learn.
-
-\todo
+\textbf{Response:} Thank you for the guidance here. Per your instruction, all
+figure captions now include (1)~what is the plot about, (2)~specific details
+of plot, and (3)~the most important thing that the reader should learn.
 
@@
 
-\item \textit{Bottom of page 4. The authors need to clarify the notation}
+\item \textit{Bottom of page 4. Clarify meaning of ``fold''.}
 
-\textbf{Response:} \todo
+\textbf{Response:} Added the following:
 
-%The fold numbers r and r_i are not well-defined. Please maintain the
-%notational consistency. You defined only "r_i". What do you mean by: 'For each
-%"fold" r'?
+``For each value $r$ in the range $1, \dotsc, K$, we define the $r$th `fold'
+as the matrix gotten by permuting the rows of the original data matrix to get
+$X$, a matrix with the $r$th test subset, the rows $i$ with $r_i = r$ as
+its trailing rows.''
 

### 2019-01-11 JCGS-R2/cvclust.tex
add R package citations

@@
 We confirm the result of Proposition~\ref{prop:single-2d} with a simulation.
+Here and throughout the text, our simulations use
+R~software environment~\citep{rproject}, with most figures generated
+with the support of the \texttt{ggplot2} library~\citep{ggplot2}.
 We perform $10$ replicates.  In each replicate, we generate $20000$
@@
 Slope method \citep{fujita2014non} and Bootstrap stability \citep{fang2012selection}. 
-We use the default parameter settings for all competing methods.  For Gabriel cross-validation,
+We use the default parameter settings for all competing methods, implented
+in the following R~packages:
+\texttt{cluster} \citep{cluster};
+\texttt{cstab} \citep{cstab};
+\texttt{e1071} \citep{e1071};
+\texttt{fpc} \citep{fpc};
+\texttt{MASS} \citep{MASS};
+\texttt{mclust} \citep{mclust};
+and
+\texttt{NbClust} \citep{NbClust}.
+For Gabriel cross-validation,
 we perform $2$-fold cross-validation on the columns~($p=q$) and $5$-fold

### 2019-01-11 JCGS-R2/response/response.tex
add R package citations
respond to AE

@@
 
-\textbf{Response:} We added citations for all packages used in our work: \todo
+\textbf{Response:}
+
+We added citations for the main packages used in our work, including for the R
+software environment itself. The packages cited include the following:
+\texttt{cluster},
+\texttt{cstab},
+\texttt{e1071}
+\texttt{fpc},
+\texttt{ggplot2},
+\texttt{MASS},
+\texttt{mclust},
+and
+\texttt{NbClust}.
 
@@
 
-\todo
+As requested, we added a comparison to the slope method. Slope performs better
+than our method in Setting 2 (and better than the non-correlation corrected
+version in Setting 1), but otherwise our method either performs comparably or
+performs better.
 

### 2019-06-29 cover/cover.tex
update with local files

\documentclass[12pt,letterpaper]{article}

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{slantsc}

\def\name{Patrick O. Perry}

\hypersetup{
  colorlinks = true,
  urlcolor = black,
  pdfauthor = {\name},
  pdfkeywords = {},
  pdftitle = {},
  pdfsubject = {},
  pdfpagemode = UseNone
}

\geometry{
  body={6.25in, 9.0in},
  left=1.25in,
  top=1.0in
}

\begin{document}
\pagenumbering{gobble}

\noindent
\begin{minipage}[t]{2in}
  \vspace{0pt}
  \includegraphics[width=1.42in]{logo.png}
\end{minipage}
\hfill
\begin{minipage}[t]{3in}
  \vspace{0pt}\raggedright
  {\footnotesize%
  {\scshape%
    Department~of~Information,~Operations,~\& \\
    Management Sciences \\
  }
  \vspace{\baselineskip}
  {\scshape%
    Henry Kaufman Management Center \\
    44 West 4\textsuperscript{th} Street, Suite 8-63 \\
    New York, NY 10012-1125 \\
    (212) 998-0194 \hspace{2ex} Fax: (212) 995-4227 \\
    Email:} pperry@stern.nyu.edu \\
  \vspace{\baselineskip}
  {\scshape {\bfseries Patrick O.\ Perry}} \\
  {\slshape%
    Assistant~Professor~of~Information,~Operations,~\& \\
    Management Sciences
  }
  }
\end{minipage}

\vspace{1\baselineskip}

\vspace{\stretch{1}}

\begin{flushright}
\today
\end{flushright}

\noindent
Dear Prof. Dianne Cook and JCGS Associate Editors,

\vspace{1\baselineskip}

Wei Fu and I are submitting revision of Manuscript ID JCGS-17-028. We have
corrected the mistakes and added discussion of the points raised by the AE and
the reviewers. The Associate Editor and Reviewer 1 mostly requested more
discussion of some of the difficult issues in the paper; we have added these
as suggested. Reviewer 2 requested that we generalize our Propositions by
removing normality and dimensionality assumptions; we have done this, either
with a new proof in the appendix or with discussion of the assumption after
the relevant proof. (We kept the assumptions in the main text as-is for the
sake of clarity of presentation.) Reviewer~2 also asked for the results in
table form rather than as figures. We have added these tables in a
supplementary appendix.

Not counting the supplementary appendix for the tables, the revision added
8 pages of material to the manuscript.

\vspace{1\baselineskip}

\noindent
Regards,

\vspace{4\baselineskip}
\noindent
Patrick Perry

\vspace{\stretch{1}}

\end{document}

### 2019-06-29 foobar/technometrics.tex
update with local files

--- 2016-01-28 New simulation/technometrics.tex
@@
 
-% operators
-\DeclareMathOperator*{\argmin}{arg\,min}
-\newcommand{\CV}{\operatorname{CV}}
-\newcommand{\PE}{\operatorname{PE}}
-\newcommand{\E}{\operatorname{E}}
-\DeclareMathOperator*{\cov}{cov}
-\DeclareMathOperator*{\tr}{tr}
-\DeclareMathOperator*{\var}{var}
-
-% convergence
-\newcommand{\toas}{\overset{\mathit{a.s.}}{\to}}
-
-\newcommand{\OhP}{O_p}
-
-% sets
-\newcommand{\R}{\mathbb{R}}
-\newcommand{\sA}{\mathcal{A}}
-\newcommand{\sbA}{\mathcal{\bar A}}
-\newcommand{\sC}{\mathcal{C}}
-\newcommand{\sE}{\mathcal{E}}
-\newcommand{\sN}{\mathcal{N}}
-\newcommand{\sX}{\mathcal{X}}
-
-% scalars
-\newcommand{\bpi}{\bar \pi}
-
-% vectors
-\newcommand{\muX}{\mu^{X}}
-\newcommand{\muY}{\mu^{Y}}
-\newcommand{\bmuX}{\bar \mu^{X}}
-\newcommand{\bmuY}{\bar \mu^{Y}}
-
-% matrices
-\newcommand{\dataX}{\mathfrak{X}}
-\newcommand{\SigmaY}{\Sigma^Y}
-\newcommand{\Xtrain}{X_{\text{train}}}
-\newcommand{\Ytrain}{Y_{\text{train}}}
-\newcommand{\Xtest}{X_{\text{test}}}
-\newcommand{\Ytest}{Y_{\text{test}}}
-
-% class labels
-\newcommand{\hGX}{\hat G^{X}}
-\newcommand{\hGY}{\hat G^{Y}}
 %----------------------We added packages----------------------------
@@
 \spacingset{1.45} % DON'T change the spacing!
-\section{Introduction}
-\label{sec:intro}
 
-As a main task of exploratory data analysis, clustering organizes unlabeled
-observations into groups such that observations in same group are more similar
-compare to those in different group. Clustering is an important topic in
-unsupervised learning because it can reveal the internal structure of data
-through grouping, segment the data through partitioning and summarize data for
-other purposes such as dimension reduction. It has being widely used in
-various fields such as psychology, biology, statistics and machine learning
-including pattern recognition, image segmentation etc \citep{jain1999data}.
-
-After being proposed more than $50$ years, $k$-means remains one of the most
-popular and widely used clustering algorithms \citep{jain2010data}. Like many
-other clustering methods, $k$-means requires an input parameter $k$, the
-number of clusters, to be specified by the user. Automatically and
-quantitatively deciding such parameter is important and yet unsolved problem
-\citep{fujita2014non}. Various methods have been proposed to tackle this
-difficulty. One ad hoc approach is to explore the relationship between $W_k$
-(within-cluster dispersion) and the number of cluster $k$ for a certain
-clustering method such as $k$-means. Since $W_k$ decreases as $k$ increases,
-one usually find the ``elbow" of curve obtain by plotting $W_k$ versus $k$ as
-the appropriate number of clusters. The example on the top row of
-Figure~\ref{fig:elbow} demonstrates such approach for data with $k=4$, where
-the ``elbow" point indeed reveals the true number of clusters. This is based
-on the idea that under partitioning data set has more impact than over
-partitioning data set in terms of $W_k$. However, locating the ``elbow" point
-is somewhat subjective and sometimes is not appropriate to select the optimal
-$k$. The second example on the bottom row of Figure \ref{fig:elbow} shows a
-situation where there is no clear choice of the ``elbow" point -- both $k=2$
-and $k=3$ can be viewed as the ``elbow" point. What's more, the true $k=4$ can
-never be selected as the optimal $k$ using such approach in this case since it
-can hardly be viewed as the ``elbow" of the curve.
-
-\begin{figure}
-\centering
-\begin{minipage}{\linewidth}
-  \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{demo/elbow/correct-data.pdf}
-  \end{minipage}
-  \hspace{0.05in}
-  \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{demo/elbow/correct-withinss.pdf}
-  \end{minipage}
-\end{minipage}
-\begin{minipage}{\linewidth}
-  \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{demo/elbow/incorrect-data.pdf}
-  \end{minipage}
-  \hspace{0.05in}
-  \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{demo/elbow/incorrect-withinss.pdf}
-  \end{minipage}
-\end{minipage}
-\caption{Left panels show the $(X,Y)$ data points; right panels
-  show the corresponding values of the within-cluster sum of squares $W_k$
-plotted against the number of clusters, $k$.}
-\label{fig:elbow}
-\end{figure}
-
-Recently, there are several new proposals to find the $k$ automatically. Gap
-statistics \citep{tibshirani2001estimating} estimates $k$ by comparing the
-change in within-cluster dispersion with that expected under an appropriate
-reference null distribution. Specifically, the graph of $\log(W_k)$ is
-compared with its expectation under an appropriate null reference distribution
-of the data. The value of $k$ associated with the largest gap between
-$\log(W_k)$ and the reference curve is selected as optimal $k$.
-\citet{sugar2003finding} proposed an approach which finds the number of
-clusters based on distortion, a quantity that measures the average distance,
-per dimension. It's backed by a rigorous theoretical justification based on
-information-theoretic ideas.   \citet{fraley2002model}'s Model-based method
-employs the EM algorithm to estimate the parameters in Gaussian mixture model,
-and select the best model ($k$) using BIC criterion. Stability-based criterion
-is also proposed to locate the best $k$ by some authors such as
-\citet{ben2001stability}, \citet{wang2010consistent} and
-\citet{fang2012selection}. \citet{chiang2010intelligent} provides a nice review
-of existing methods for finding the right $k$ in published literature.
-
-Most existing methods are either model based method requires strong modeling
-assumptions or otherwise lack of clear interpretation and theoretical justification.
-Although many view selecting the number of clusters as a model selection problem, very few
-approaches this problem from the prediction point of view. Select model with
-smallest prediction error via cross-validation is one of the simplest and most
-widely used model selection techniques in supervised learning. The lack of
-true class (label) in data set makes the adoption of cross-validation into
-unsupervised learning problem difficult. A naive algorithm may works as following:
-with the data split by rows into training set and test set, cluster the training 
-data with parameter $k$; then predict each observation in test data by the closest
-cluster center returned in last step. The prediction error is calculated by the 
-difference between the predicted cluster centers and observations' true value, and algorithm 
-picks the $k$ which minimizes such error. Such algorithm will always pick the maximum number 
-of $k$ allowed because more cluster centers means more tight fit to the data space, therefore
-smaller within-cluster dispersion even if the prediction is evaluated on an independent
-test set \citep{hastie2009elements}.
-
-One exception is \cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction
-strength. The strategy is to first cluster the test data and training data
-into $k$ clusters respectively. Then, for each pair of observations that are
-assigned to the same test cluster, algorithm determines whether they are also
-assigned to the same cluster based on the training centers. The intuition here
-is, if $k=k_0$, the true number of clusters, then the $k$ training set
-clusters will be similar to the $k$ test clusters, and hence will predict them
-well. However, a specifically defined prediction error measure is used in such
-procedure, which is quite different from the one commonly used in
-cross-validation procedure in supervised learning. \citet{wang2010consistent} also uses 
-cross-validation to select the optimal $k$. Instead of selecting $k$ which minimizes 
-prediction error, such method picks $k$ which minimizes the specifically defined 
-clustering instability. Note that these methods minimize measures that are specifically defined,
-which makes these methods hard to interpret or compare to other methods through analogy because
-the underlying measures are unique and not well understood.
-
-The prediction error measure in our proposed method is exactly the same as in supervised learning.
-Our method is a complete data-driven approach which doesn't rely on strong model assumption. Through novel form of partitioning data set, we effectively 
-transferred an unsupervised learning problem into a supervised leaning problem, which is one of
-the kind. By doing so, we are able to employing the cross-validation procedure in clustering 
-a similar way as in supervised learning problem, so that much of the intuition from supervised learning
-carries over. Hence, it's easy for reader to
-understand the intuition behind our proposed method. Simulation and real data
-application shows the superior performance of our proposed method compare with
-existing methods in high-dimension settings and heavy-tailed data. Since the
-embedded cross-validation procedure is well understood, it also makes our
-method potentially easily to be extended in future study.
-
-Some proposed methods provide theoretical consistency result for choosing $k$. 
-Assume data is a mixture of $G$ $p$-dimension clusters with equal prior, identically distributed with 
-common covariance matrix and finite fourth moments in each dimension, \citet{sugar2003finding} shows
-the Jump method will pick the correct $k$ under the conditions that cluster centers are sufficiently
-separated and an appropriate transformation is used. Specifically, let $\Delta\sqrt{p}$ denotes the 
-minimum Euclidean distance between cluster centers, \citet{sugar2003finding} shows that the jump will
-be maximized at $k=G$ given that $\Delta > 6$ and the existence of a positive $Y$ such that
-\[	\left(\frac{p\Delta^2W}{9G}\right)^{-Y}+\left(W\left[\frac{2^{2H^*(X)}}{K^2_{max}2\pi e}\right]-(\frac{\Delta}{6})^2(1-W) \right)^{-Y} < 2	\]
-and $\left(\frac{p\Delta^2W}{9G}\right)^{-Y} <1/2$, where $H^*(X)$ is the minimum entropy of the cluster 
-membership over each dimension and $W =1- \frac{6^4V_{\mathbf{X}}}{(\Delta^2-36)^2}$ with $V_{\mathbf{X}} = var\left( 1/p ||\mathbf{X}-\mu_j||^2_{\Gamma^{-1}} \mid \mathbf{X} \hspace{0.05in}\text{in jth cluster}\right)$
-
-\cite{tibshirani2005cluster} proved that the prediction strength method is consistent under the setting 
-that the observations in $p$-dimension are uniformly distributed within one unit from their respective cluster centers, with minimum distance between the $G$ cluster centers is four unit.
-Under such well-separated setting,
-they show that 
-\[ ps(G) = 1+ o_p(1), \hspace{0.5in} \underset{k \geq G+1}{\text{sup}} ps(k)\leq \frac{2}{3}+ o_p(1) \]
-therefore $\hat{k}$ is consistent in estimating $G$.  
-When data does come from mixture of Gaussian distribution, the model-based method of \citet{fraley2002model} is consistent in estimating $k$ in low dimension.
-
-Let $\Psi$ denotes any given base-clustering algorithm (e.g $k$-means), and $\Psi(Z,k)$ the clustering obtained by applying $\Psi$ on data $Z$ with parameter $k$.
-Also let $k_0$ be the optimal number of cluster and $d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ be the measure of distance between clustering $\Psi(Z_1,k)$ and $\Psi(Z_2,k)$ where $Z_1$ and $Z_2$ are two independent samples from population.
-\citet{wang2010consistent} shows that their proposed method will consistently select $k = k_0$, given such $k_0$ exists and $d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ converges to zero at proper rate $r_k$.
-However, $k_0$ is defined as the most stable number for $\Psi$ among all the possible $k$, instead of the number of components in mixture model or well-separated compact regions commonly seen in literature.
-Therefore, their consistent result actually is convergent result, i.e.
-the selected $k$ converges to its limit $k_0$ given such limit exists.
-
-\section{Cross-validation for selecting the number of clusters}
-\label{sec:meth}
-Cross-validation is commonly used for model selection in supervised learning
-problems.  In these settings, the data comes in the form of $N$
-predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
-\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
-$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
-``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
-``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
-matrix to get $\dataX$, a matrix with the $r$th test subset as its trailing
-rows.  We partition $\dataX$ as
-\[
-  \dataX =
-  \begin{bmatrix}
-    \Xtrain & \Ytrain \\
-    \Xtest  & \Ytest
-  \end{bmatrix}.
-\]
-We use the training rows $[ \Xtrain\ \Ytrain ]$ to fit a regression model
-$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
-test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
-or some variant thereof.  We choose the model with the smallest
-cross-validation error, averaged over all $K$ folds.
-
-In unsupervised learning problems like factor analysis and clustering, the
-features of the observations are not naturally partitioned into ``predictors''
-and ``responses'', so we cannot directly apply the cross-validation procedure
-described above.  For factor analysis, there are at least two versions of
-cross-validation.  \citet{wold78cross} proposed a ``speckled'' holdout, where
-in each fold we leave out a subset of the elements of the data matrix.  Wold's
-procedure works well empirically, but does not have any theoretical support,
-and it requires a factor analysis procedure that can handle missing data.
-\citet{owen2009bi} proposed a scheme called ``bi-cross-validation'' wherein
-each fold designates a subset of the data matrix columns to be response and a
-subset of the rows to be test data.  This generalized a procedure due to
-\citet{gabriel2002biblot}, who proposed holding out a single column and a
-single row at each fold.  Owen and Perry proved that this procedure is
-self-consistent, in the sense that it performs the correct model selection in
-the absence of noise, and \citet{perry2009cross} provided more theoretical
-support.
-
-In this report, we extend the Wold and Gabriel methods to the clustering
-problem, specifically to choose an appropriate number of clusters for a
-dataset.  We prove that the Gabriel method is self-consistent, and we analyze
-some of its properties in the presence of noise.  We compare these methods to
-state-of-the-art algorithms, and show that both are competitive.
-
-%% [POP] This paragraph doesn't fit into the section.  Maybe put it in the introduction?
-%% 
-%% Cross-Validation technique, one of the most commonly used and popular model
-%% selecting method in supervised learning, can not be used naively in
-%% unsupervised learning context. Since prediction error of new observation is
-%% calculated by its distance to the nearest cluster center, more cluster centers
-%% means much tighter fit to the feature space, and hence smaller distance
-%% (prediction error) of observation to its nearest cluster center. This holds
-%% even when the prediction is evaluated on an independent test set
-%% \citep{hastie2009elements}. Therefore, CV prefers larger $k$ if we do the CV
-%% naively.
-
-We now give the details of how to implement
-the Gabriel cross-validation to locate the optimal cluster number $k$. The
-Wold cross-validation algorithm is described in Appendix $A$.
-
-\subsection{Gabriel CV algorithm}
-\label{sec:gabriel-cv-algorithm}
-We are given a data matrix with $N$ rows and $P$ columns.  In each fold of
-cross-validation, we permute the rows and columns of the data matrix and then
-partition the rows and columns as $N = n + m$ and $P = p + q$ for 
-non-negative integers $n$, $m$, $p$, and $q$.  We treat the first $p$
-columns as ``predictors'' and the last $q$ columns as ``responses'';
-similarly, we treat the first $n$ rows as ``training'' and the last $m$ rows
-as ``test''.  In block form, the permuted data matrix is
-\[
-  \dataX
-  =
-  \begin{bmatrix}
-    \Xtrain & \Ytrain \\
-    \Xtest  & \Ytest
-  \end{bmatrix},
-\]
-where
-$\Xtrain \in \R^{n \times p}$,
-$\Ytrain \in \R^{n \times q}$,
-$\Xtest \in  \R^{m \times p}$,
-and
-$\Ytest \in  \R^{m \times q}$.
-
-Given such a partition of $\dataX$, we perform four steps for each value of
-$k$, the number of clusters:
-\begin{enumerate}
-  \item \label{step:gabriel-cluster}
-    \textbf{Cluster:}
-    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding the
-    assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ and the
-    cluster means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
-    be the assigned cluster for row $i$.
-  \item \label{step:gabriel-classify}
-    \textbf{Classify:}
-    Take $X_{1}, \dotsc, X_n$, the rows of $\Xtrain$ to be predictors,
-    and take $\hGY_1, \dotsc, \hGY_n$ to be corresponding class labels.  Use
-    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to train a classifier
-    $\hGX : \R^p \to \{ 1, \dotsc, k \}$.
-  \item \label{step:gabriel-predict}
-    \textbf{Predict:}
-    Apply the classifier to $X_{n+1}, \dotsc, X_{n+m}$, the rows of
-    $\Xtest$, yielding predicted classes $\hGX_i = \hGX(X_i)$ for
-    $i = n+1, \dotsc, n+m$.  For each value of $i$ in this range, compute
-    predicted response $\hat Y_i = \bmuY(\hGX_i)$, where
-    $\bmuY(g) = \bmuY_g$.
-  \item \label{step:gabriel-evaluate}
-    \textbf{Evaluate:}
-    Compute the cross-validation error
-    \[
-      \CV(k) = \frac{1}{m} \sum_{i=n+1}^{n+m} \|Y_i - \hat Y_i\|^2,
-    \]
-    where $Y_{n+1}, \dotsc, Y_{n+m}$ are the rows of $\Ytest$.
-\end{enumerate}
-\noindent
-In principle, we could use any clustering and classification methods in
-steps~\ref{step:gabriel-cluster} and~\ref{step:gabriel-classify}.  In this
-report, we use $k$-means as the clustering algorithm.  For the classification
-step, we compute the mean value of $X$ for each class; we assign an
-observation to class $g$ if that class has the closest mean (randomly breaking
-ties between classes).  The classification step is equivalent to linear
-discriminant analysis with equal class priors and identity noise covariance
-matrix.
-
-To choose the folds, we randomly partition the rows and columns into $K$ and
-$L$ subsets, respectively.  Each fold is indexed by a pair $(r,s)$ of
-integers, with $r \in \{1, \dotsc, K\}$ and $s \in \{1, \dotsc, L\}$.  Fold
-$(r,s)$ treats the $r$th row subset as ``test'', and the $s$th column subset
-as ``response''.  We typically take $K = 5$ and $L = 2$.  For the number of
-clusters, we select the value of $k$ that minimizes the average of $\CV(k)$
-over all $K \times L$ folds (choosing the smallest value of $k$ in the event
-of a tie).
 %------------------------------------------------------------------
-\section{Self-Consistency of Gabriel CV method}
-
-This section gives the self-consistency proof of the proposed Gabriel method.
-Specifically, we will show that under appropriate conditions, in the absence
-of noise, the Gabriel cross-validation procedure finds the optimal number of
-clusters.
-
-Because $k$-means algorithm is essential to the method, we review the
-procedure here.  Given a set of observations $\{ x_1, \dotsc ,x_n \}$, and a
-specified the number of clusters $k$, the goal of the $k$-means procedure is
-to find a set of $k$ or cluster centers $A = \{ a_1, \dotsc, a_k \} $
-minimizing the within cluster dispersion
-\[
-  W(A) = \sum_{i=1}^{n} \min_{a \in A} \|x_i - a\|^2.
-\]
-This implicitly defines a cluster assignment rule
-\[
-  g(x) = \argmin_{g \in \{1, \dotsc, k\}} \|x - a_g\|^2,
-\]
-with ties broken arbitrarily.  We will assume that the $k$-means procedure
-finds an optimal solution, $A$, but we will not assume that this solution is
-unique.
-
-It will suffice to analyze a single fold of the cross-validation procedure.
-As in in section~\ref{sec:gabriel-cv-algorithm} we assume that the $P$
-variables of the data set have been partitioned into $p$ predictor variables
-represented in vector~$X$ and $q$ response variables represented in
-vector~$Y$.  The $N$ observations have been divided into two sets: $n$ train
-observations and $m$ test observations.  The following theorem gives
-conditions for Gabriel CV to recover the true number of clusters in the
-absence of noise.
-
-\begin{theorem}\label{thm:self-consistency}
-
-Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
-cross-validation.  For any $k$, let $CV(k)$ be the cross-validation error for
-this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
-We will assume that there are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with
-the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
-\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
-\begin{enumerate}[label=(\roman*)]
-  \item \label{asn:self-consistency-noiseless}
-    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
-    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
-    $i = 1, \dotsc, n+m$.
-  \item \label{asn:self-consistency-distinct-mux}
-    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
-  \item \label{asn:self-consistency-distinct-muy}
-    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
-  \item \label{asn:self-consistency-train}
-    The training set contains at least one member of each cluster: for all $g$
-    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
-    $1, \dotsc, n$ such that $G_i = g$.
-  \item \label{asn:self-consistency-test}
-    The test set contains at least one member of each cluster: for all $g$ in
-    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
-    \dotsc, n+m$ such that $G_i = g$.
-\end{enumerate}
-Then $CV(k) < CV(K)$ for $k < K$, and $CV(k) = CV(K)$ for $k > K$, so that
-Gabriel CV correctly chooses $k = K$.
-\end{theorem}
-
-This theorem is implied by the following two lemmas.
-
-\begin{lemma}
-Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
-force.  If $k < K$, then $\CV(k) > 0$.
-\end{lemma}
-\begin{proof}
-By definition,
-\[
-  \CV(k)
-    =
-      \sum_{i=n+1}^{n+m}
-        \| Y_i - \bmuY (\hGX_i) \|^2,
-\]
-where $\bmuY(g)$ is the center of cluster $g$ returned from applying $k$-means
-to $Y_1, \dotsc, Y_n$.  Assumptions~\ref{asn:self-consistency-noiseless}
-and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
-set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all distinct values in
-$\{ \muY(1), \dotsc, \muY(K) \}$.
-Assumption~\ref{asn:self-consistency-distinct-muy} implies that there are
-exactly $K$ such distinct values.  However, there are only $k$ distinct values
-of $\bmuY(g)$.  Thus, at least one summand
-\(
-  \| Y_i - \bmuY(\hGX_i) \|^2
-\)
-is nonzero.  Therefore,
-\(
-  \CV(k) > 0.
-\)
-\end{proof}
-
-\begin{lemma}
-Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
-force.  If $k \geq K$, then $\CV(k) = 0$.
-\end{lemma}
-\begin{proof}
-From assumptions~\ref{asn:self-consistency-noiseless},
-\ref{asn:self-consistency-distinct-muy},
-and~\ref{asn:self-consistency-train}, we know the cluster centers
-gotten from applying $k$-means to $Y_1, \dotsc, Y_n$ must include
-$\muY(1), \dotsc, \muY(K)$.  Without loss of generality, suppose that
-$\bmuY(g) = \muY(g)$ for $g = 1, \dotsc, K$.  This implies that
-$\hGY_i = G_i$ for $i = 1, \dotsc, n$.  Thus, employing
-assumption~\ref{asn:self-consistency-noiseless} again, we get that
-$\bmuX(g) = \muX(g)$ for $g = 1, \dotsc, K$.
-
-Since assumption~\ref{asn:self-consistency-distinct-mux} ensures that
-$\muX(1), \dotsc, \muX(K)$ are all distinct, we must have that $\hGX_i = G_i$
-for all $i = 1, \dotsc, m+n$.  In particular, this implies that $\bmuY(\hGX_i)
-= Y_i$ for $i = 1, \dotsc, m+n$, so that $\CV(k) = 0$.
-\end{proof}
 %-------------------------------------------------------------
 \section{Analysis of Gabriel Cross-Validation with Gaussian Noise}
-
-\subsection{Single Cluster}
-Now we we analyze the asymptotic performance of Gabriel Cross-Validation, in
-the case of Gaussian noise. Our main result is that with single-cluster
-Gaussian data, if the predictor and response columns of $\dataX$ are weakly
-correlated or independent, then the method will correctly prefer $k = 1$ to
-$k = 2$ clusters. We first state the result in the case where $\dataX$ has two 
-columns, and later generalize this result to higher dimensions.
 
@@
 \end{proof}
-
-We confirm this result with a simulation.  We perform $10$ replicates.  In each
-replicate, we generate $20000$ observations from a mean-zero bivariate normal
-distribution with unit marginal variances and correlation $\rho$.  We perform
-a single $2 \times 2$ fold of Gabriel cross-validation and report the
-cross-validation mean squared error for the number of clusters $k$ ranging
-from $1$ to $5$.  Figure~\ref{fig:nullcorr-equal} shows the cross-validation
-errors for all $10$ replicates.  The simulation demonstrates that in the
-Gabriel cross-validation criterion chooses the correct answer $k = 1$ whenever
-$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=3in]{demo/nullcorr/equal.pdf}
-\caption{Cross-validation error on $10$ replicates, with the number of
-clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
-multivariate normal distribution with correlation $\rho$.  The Gabriel
-cross-validation criterion chooses the correct answer $k = 1$ whenever
-$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
-\label{fig:nullcorr-equal}
-\end{figure}
-
-The reason why Gabriel CV method tends to select larger $k$ when correlation is high 
-between dimensions is that it resembles the naive method mentioned in section \ref{sec:intro} 
-under such circumstance. In fact, if $X$, $Y$ are perfectly correlated, Gabriel method is 
-equivalent to the naive method.
 
@@
 
-\begin{theorem}
-Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
-of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
-independent draw from a mean-zero multivariate normal distribution with unit
-marginal variances and correlation $\rho$.  In this case, the data are drawn
-from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 0.5$,
-then $\CV(1) < \CV(k)$, $k > 1$ with probability tending to one as $m$ and $n$ increase.
-\end{theorem}
-
-\begin{proof}
-Given $(X_1, Y_1), \dotsc, (X_n, Y_n)$, let $A_k = \{a_1,a_2,...,a_k\}$ denotes the set of cluster centers from applying $k$-means to $\{ Y_i\}_{i=1}^{n}$ with parameter $k$.
-Because $\{ Y_i\}_{i=1}^{n}$ is symmetric, it's easy to see that the cluster centers are symmetric around $0$, i.e.
-if $a^* \in A_k$ then $-a^* \in A_k$.
-The $k$ clusters are symmetric in the same sense.
-Also, if $k$ is odd, then only one cluster contains both negative and positive $Y$ and it's symmetric with center $0$.
-The rest have pure positive or negative intervals; if $k$ is even, then all clusters have pure positive or negative intervals.
-
-Let $a_i$ denotes a cluster center with interval $[b_i, c_i]$.
-By above argument, either $0 \leq b_i < c_i$ ($b_i < c_i \leq 0$) or $b_i = -c_i$ which corresponding to $a_i=0$ with $k$ be odd.
-By symmetry, it's sufficient to only consider case $0 \leq b_i < c_i$ (with $a_i > 0 $) and $b_i = -c_i$ (with $a_i = 0 $) with $\rho \geq 0$.
-
-From joint normality of $X$ and $Y$, it follows that $X \mid Y$ is normal with mean $\rho Y$. Therefore, the mean value $\bar{X}_i$ of $X$ in cluster with $Y$ center $a_i$ is
-
-\begin{align*}
-	\bar{X}_i &= E[X \mid b_i \leq Y \leq c_i] \\
-			&= E[ E(X \mid Y) \mid b_i \leq Y \leq c_i] \\
-			&= E[\rho Y \mid b_i \leq Y \leq c_i] \\
-			&= \rho E[ Y \mid b_i \leq Y \leq c_i] \\
-			&= \rho a_i
-\end{align*}
-
-That is, the mean values of $X$ in the $k$ clusters is $\{\rho a_1,\rho a_2,...,\rho a_k\}$.
-Because we assign each observation to the closest $\bar{X}_i$, $i=1,2,...,k$, it's easy to see the corresponding interval on $X$ for $\bar{X}_i = \rho a_i$ is $[\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}]$, where $a_{i-1}$ and $a_{i+1}$ are the two adjacent centers to $a_i$ on $Y$.
-Note that $[\frac{ \left( a_{i-1}+a_i\right)}{2}, \frac{\left( a_{i}+a_{i+1}\right)}{2}] = [b_i, c_i]$ by the algorithm definition of $k$-means.
-The $Y$ center for such interval on $X$ is
-
-\begin{align*}
-	\hat{\mu}^Y_i &= E[Y \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
-	 &= E[ E(Y \mid X) \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
-	 &= E[ \rho X \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
-	 & =  \rho E[X \mid \frac{\rho \left( a_{i-1}+a_i\right)}{2} \leq X \leq \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] \\
-	  & =  \rho E[X \mid \rho b_i \leq X \leq \rho c_i] \\
-	  &\leq \rho a_i
-\end{align*}
-
-where the equality holds if $b_i = -c_i$ or $\rho = 1$ on the last step. This is because
-
-\begin{align*}
-	 E[X \mid \rho b_i \leq X \leq \rho c_i] &= E[Y \mid \rho b_i \leq Y \leq \rho c_i] \\
-	\mathtt{0 \leq b_i < c_i}, \hspace{0.05in} \mathtt{0 \leq \rho \leq 1} \hspace{0.1in} & \leq E[Y \mid b_i \leq Y \leq c_i] \\
-	&= a_i
-\end{align*}
-
-The first equation above is because the same marginal distribution of $X$ and $Y$.
-
-Note that Gabriel CV method predicts $Y$ to be $a_i$ for $X \in [\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}]$, and if $k=1$ the predicted $Y = 0$ for all $X$.
-Hence, to see if $CV(k) > CV(1)$ for  $X \in [\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}]$, one only need to see if $ a_i-\hat{\mu}^Y_i > \hat{\mu}^Y_i - 0$.
-Since $ \hat{\mu}^Y_i \leq \rho a_i$, it's clear that $CV(k) > CV(1)$ whenever $\rho < 0.5$.
-This is true to all segment $[\frac{\rho \left( a_{i-1}+a_i\right)}{2}, \frac{\rho \left( a_{i}+a_{i+1}\right)}{2}] = [\rho b_i, \rho c_i]$ except the one with $b_i = -c_i$, in which $CV(k) = CV(1)$.
-Since $k>1$ and at most one segment has $CV(k) = CV(1)$ while the rest have $CV(k) > CV(1)$ if $\rho < 0.5$, it is clear that the overall prediction error $CV(k) > CV(1)$.
-By symmetry, we can see $CV(k) > CV(1)$ if $|\rho| < 0.5$.
-The proof is complete.
-\end{proof}
-
-\subsection{Two Clusters}
-
-We will now analyze a simple two-cluster setting, and derive conditions for
-Gabriel cross-validation to correctly prefer $k=2$ clusters to $k=1$.
-
-\begin{proposition}
-Suppose that $\{(X_i,Y_i)\}_{i=1}^{n+m}$ is data from a single fold of Gabriel
-cross-validation, where each $(X,Y)$ pair in $\R^2$ is an independent draw
-from an equiprobable mixture of two multivariate normal distributions with
-identity covariance. Suppose that the first mixture component has mean $(\muX,
-\muY)$, and the second has mean $(-\muX, -\muY)$, where $\muX > 0$ and
-$\muY > 0$.  If $1 + 2\Phi(\mu^Y)+ \frac{2\varphi(\mu^Y)}{\mu^Y} < 4\Phi(\mu^X)$, then $\CV(2) < \CV(1)$ with
-probability tending to one as $m$ and $n$ increase.
-\end{proposition}
-
-\begin{proof}
-There are two clusters $G_1$ and $G_2$, where observations from $G_1$ are distributed as
-\[	\mathcal{N}\left( \begin{pmatrix} 
-    \mu^X \\
-    \mu^Y \\
-  \end{pmatrix}, \mathbf{I} \right)	\]
-and observations from $G_2$ are distributed as
-\[	\mathcal{N}\left( \begin{pmatrix} 
-    -\mu^X \\
-    -\mu^Y \\
-  \end{pmatrix}, \mathbf{I} \right)	\]
-where $\mu^X_1 > 0$ and $\mu^Y_1 > 0$. Let $G_i$ be the true cluster where observation $i$ is generated from, by assumption
-\[	P(G_i=G_1) = P(G_i=G_2) = 1/2	\]
-After applying $k$-means on $\{ Y_i\}_{i=1}^{n}$ with $k=2$, if $n$ is large enough, we have the estimated centroids $\bar{\mu}^Y_1$ and $\bar{\mu}^Y_2$ be close to $E(Y \mid Y>0)$ and $\E(Y \mid Y < 0)$,
-with errors will be of size $\OhP(n^{-1/2})$. Here
-\begin{align}
-E(Y \mid Y>0) &= E(Y_1 \mid Y_1 > 0) \cdot P(Y_1>0) + E(Y_2 \mid Y_2 > 0) \cdot P(Y_2>0) \nonumber \\ 
-   &= 2\varphi(\mu^Y)+2\mu^Y\Phi(\mu^Y)-\mu^Y  \\  \nonumber 
-\end{align}
-where $Y_1 \sim N(\mu^Y, 1)$ and $Y_2 \sim N(-\mu^Y, 1)$, and we used Lemma~\ref{lem:truncated-normal-moments}
-(Appendix~\ref{app:technical-lemmas}). Similarly, we have
-\begin{align}
-E(Y \mid Y<0) &= E(Y_1 \mid Y_1 < 0) \cdot P(Y_1<0) + E(Y_2 \mid Y_2 < 0) \cdot P(Y_2<0) \nonumber \\ 
-   &= -2\varphi(\mu^Y)-2\mu^Y\Phi(\mu^Y)+\mu^Y  \\  \nonumber 
-\end{align}
- where $\varphi()$ and $\Phi()$ are the standard normal probability and cumulative distribution function respectively.
-
-Same as in single cluster case, the classification rule learned from
-$\{(X_i, \hGY_i)\}_{i=1}^{n}$ variables will be determined according to
-whether $X > 0$, with the decision boundary be at $0 + \OhP(n^{-1/2})$.
-By symmetry, the CV error for points from $G_1$ is same as the points from $G_2$. Because $P(G=G_1) = P(G=G_2) =1/2$, the CV error can be calculated solely from $G_2$, that is
-
-\begin{align*}
-  \CV(2)
-  &=
-    \frac{1}{m}
-    \sum_{i=n+1}^{n+m}
-      \| (Y_i - \bmuY_1) 1\{\hGX_i = 1\} \|^2
-      +
-      \| (Y_i - \bmuY_2) 1\{\hGX_i = 2\} \|^2, \hspace{0.2in}Y\sim N(-\mu^Y,1)
-\\
-  &=
-    \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]
-      + \OhP(m^{-1/2}) + \OhP(n^{-1/2})\\
-   \end{align*}
-   With 
-   \begin{align}
- \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]  =& P(\hGX_i = 1)\cdot E[(Y - a)^2] + P(\hGX_i = 2)\cdot E[(Y + a)^2] \nonumber \\  \nonumber
- =& [1-\Phi(\mu^X)][var(Y)+(-\mu^Y - a)^2]+\\ \nonumber
-  & \Phi(\mu^X)[var(Y)+(-\mu^Y + a)^2]\\  \nonumber
-  =& [1-\Phi(\mu^X)][1+(\mu^Y + a)^2]+\Phi(\mu^X)[1+(\mu^Y -a )^2]   \\  \nonumber
-  =& 1+(\mu^Y + a)^2 -4a\Phi(\mu^X)\mu^Y  \nonumber
-\end{align}
-where $a$ is given by $(1)$.
-
-When $k=1$, the result is straight forward since the estimated centroid will approximately equal to
-$0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error
-will be
-\[
-  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
-         = 1 + (\mu^Y)^2 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
-\]
-So if we have $1 + 2\Phi(\mu^Y)+ \frac{2\varphi(\mu^Y)}{\mu^Y} < 4\Phi(\mu^X)$, we have $\CV(2) < \CV(1)$
-\end{proof}
-
-We confirm this result with a simulation.  We perform $10$ replicates for each pair of $(\mu^X, \mu^Y)$, where both $\mu^X$ and $\mu^Y$ take value on grid of $[0,3]$ with step $0.1$.  In each
-replicate, we generate $20000$ observations from two multivariate normal distributions with identity covariance, where one has mean $(\muX, \muY)$ and the other one has mean $(-\muX, -\muY)$.
-We perform a single $2 \times 2$ fold of Gabriel cross-validation and report the
-times (out of $10$ replicates) when $k=2$ is selected by the algorithm in stead of $k=1$.
-Figure~\ref{fig:overlap-color_plot} shows the frequency $k=2$ is selected by the algorithm for each pair of $(\mu^X, \mu^Y)$.
-The dark spot means high number (close to $10$) is selected by the algorithm, which means algorithm very likely will pick $k=2$ over $k=1$ for the corresponding $(\mu^X, \mu^Y)$.
-While light spot means algorithm prefer $k=1$ for the corresponding value of $(\mu^X, \mu^Y)$.
-We can see the simulation result perfectly align with the theoretical curve (the black line), which separates the $k=2$ zone from the $k=1$ zone.
-It demonstrates that the Gabriel cross-validation works exactly as it suppose to under such setting.
-The position of dark spots shows that when the two clusters are reasonably apart (not overlapping too heavily) in both dimensions, the Gabriel cross-validation is asymptotically consistent.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=3in]{demo/overlap/color_plot.pdf}
-\caption{Number of times $k=2$ is selected out of $10$ replicates for each pair of $(\muX, \muY)$.
-The heat map shows the frequency $k=2$ is selected by the algorithm, with light means low number (of $k=2$) is selected and dark color means high number is selected.
-The black line is the theoretical curve, above which the algorithm suppose to pick $k=2$ and below which algorithm select $k=1$.
-}
-\label{fig:overlap-color_plot}
-\end{figure}
-
-\subsection{Correlation adjustment for Gabriel method}
-When the correlation between dimensions are high, the proposed Gabriel CV method tends to overestimate the number $k$.
-A simply remedy for the high correlation is available if we assume common covariance structure among the $k$ clusters.
-\begin{enumerate}
-	\item Apply Gabriel CV method on the original data $\dataX$, get estimated number of cluster $\hat{k}$
-	\item Estimate the pooled covariance matrix $\hat{\Sigma}$ from the $\hat{k}$ clusters.
-	\item Let $\hat{\Sigma} = \Gamma\Lambda\Gamma'$ be the Eigendecomposition of $\hat{\Sigma}$, we rescale and rotate the original data $\dataX$ to get $ \widetilde{\dataX} = \dataX\Gamma\Lambda^{-1/2}Q$, where $Q$ is a random  orthonormal rotation matrix.
-	\item Apply Gabriel CV method again on the transformed data $ \widetilde{\dataX}$ to estimate $k$. 
-\end{enumerate}
-
-\section{Simulation}
-In this section, simulation is performed to evaluate the performance of our
-proposed methods in locating the ``correct" number of clusters. We compare
-with a basket of existing methods including Gap statistics
-\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
-\citep{fraley2002model}, CH-index \citep{calinski1974dendrite}, Hartigan
-statistics \citep{hartigan1975clustering}, Jump method
-\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
-Bootstrap stability \citep{fang2012selection} in following simulation
-settings. All methods are executed with their default parameter settings. 
-We select $p=q$ and $5-$fold cross-validation in row
-($m=\frac{1}{4}n$) as default parameter setting for our proposed Gabriel
-method. Note that set $p=q$ corresponding to $2-$fold cross-validation in
-column. Wold method (detailed in Appendix $A$) and correlation-corrected 
-Gabriel method (section $4.3$) are also included for comparison.
-
-Note that in all settings, cluster centers are randomly generated from 
-multivariate normal distribution $\mathcal{N}\left(\mathbf{0},\varsigma \mathbf{I}\right)$. 
-All clusters are well-separated, i.e. no overlapping. In fact, any simulated 
-clusters with minimum distance less than one unit was discarded, so there is
-clear definition of true number of clusters. The parameters $\varsigma$ is chosen 
-such that about half of the random realization were discarded. The idea is borrowed from
-\cite{tibshirani2001estimating}. The proportion of each method successfully picks the correct
-$k$ out of $100$ simulation trials is reported, along with its corresponding confident interval
-by Wilson's method \citep{wilson1927probable}.
-
-\begin{enumerate}
-
-  \item \textbf{Correlation between dimensions} -- Six clusters in $10$ dimensions. 
-  	Each cluster has $100$ or $50$
-    multivariate normal observations with common covariance matrix $\Sigma$ 
-    which has compound symmetric structure with $1$ in diagonal and $\rho$ 
-    off diagonal. $\rho$ takes value in $\{0,0.1,...,0.9\}$.
-
-	\begin{figure}[H]
-	\centering
-	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting1/Facet.pdf}
-	\label{fig:setting1}
-	\end{figure}
-
-  \item \textbf{Noise dimensions} --- Three clusters in $6$ dimensions. Each cluster has $1000$ or $500$
-    mean zero multivariate normal observations with identity covariance matrix. 
-    We add $P$ dimensions of noise to the data, which is randomly generated from uniform$[0,1]$. The noise
-    dimension $P$ takes values in $\{0,6,...,54\}$.
-
-	\begin{figure}[H]
-	\centering
-	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting2/Facet.pdf}
-	\label{fig:setting2}
-	\end{figure}
-
-  \item \textbf{High dimension} --- Eight clusters in $p$ dimensions, $p$ takes values 
-  	in $\{10,20,...,100\}$. Each cluster has $100$ or $50$ mean zero multivariate normal 
-  	observations with identity covariance matrix.
-
-	\begin{figure}[H]
-	\centering
-	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting3/Facet.pdf}
-	\label{fig:setting3}
-	\end{figure}
-
-     \item \textbf{Variance heterogeneity} --- Three clusters in $20$ dimensions, each has
-     $60$ observations in it. Observations are generated from 
-     $\mathcal{N}\left(\mathbf{0},\sigma_1^2\mathbf{I}\right)$,
-      $\mathcal{N}\left(\mathbf{0},\sigma_2^2\mathbf{I}\right)$ and
-       $\mathcal{N}\left(\mathbf{0},\sigma_3^2\mathbf{I}\right)$ where 
-       $\sigma_1^2 : \sigma_2^2: \sigma_3^2 = 1:\frac{1+R}{2}:R$. The maximum ratio $R$
-       takes values in $\{1,5,10,...,45\}$
-
-	\begin{figure}[H]
-	\centering
-	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting4/Facet.pdf}
-	\label{fig:setting4}
-	\end{figure}
-
-	\item \textbf{Heavy tail data} --- Five clusters in $15$ dimensions, each has
-     $80$ observations in it. Observations have independent $t$ distribution in each dimension  
-     with degree of freedom $\nu$, which takes value in $\{11,10,...,2\}$
-
-	\begin{figure}[H]
-	\centering
-	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting5/Facet.pdf}
-	\label{fig:setting4}
-	\end{figure}
-\end{enumerate}
-
-From the graph of setting $1$, we can see high correlation between dimensions cause problem for most 
-existing methods as well as the proposed Gabriel CV method and Wold method. Higher correlation clearly 
-degenerate most algorithms' ability to find the correct $k$ as data no longer spherical in $p$-dimension.
-And the recent proposed methods (Gabriel CV method, Gap statistic, Jump method) seem to be more sensitive to
-high correlation than those old methods such as CH-index and Hartigan statistics. The only two methods 
-work well in high correlation are the Gaussian model-based BIC method \citep{fraley2002model} and the 
-correlation-corrected Gabriel method.
-
-Plot of setting $3$ shows some methods are insensitive to higher dimension such as Jump method and Gap 
-statistics, while others deteriorate quickly with increasing dimension, most notably the Gaussian 
-model-based BIC method. Note here the data are still generated from mixture of Gaussian. In contrast, our 
-proposed Gabriel CV method as well as its correlation-corrected version actually working better in higher 
-dimension.
-
-Result of setting $4$ indicates most previously methods are sensitive to variance heterogeneity among 
-clusters, most notably the Gap statistics and model-based BIC method. The proposed Gabriel CV method
-and its correlation-corrected version consistently perform well in estimating $k$ and quite insensitive
-to variance heterogeneity. The Wold method also performs well in this setting.
-
-The last non-Gaussian setting examines methods' performance in heavy-tail data. With degree of freedom 
-decreases, the tail becomes more flat and the Gaussian assumption becomes more inappropriate. It explains
-the sharp plunge of Gaussian model-based BIC method in the plot. For most methods, their performances are 
-relatively stable until the tail gets very heavy. In case where degree of freedom is $2$, Gap statistics 
-and Jump method dived considerably compare to the Gabriel method and Wold method.
-
-In conclusion, the proposed Gabriel CV method compares well with current state-of-the-art methods and 
-is very robust against variance heterogeneity, high dimension and heavy-tail data. One weakness of the 
-Gabriel CV method is that it's sensitive to the high correlation among dimensions. In such case, its 
-correlation-corrected version will perform well under the assumption of common covariance structure.
-
-\section{Real data application}
-We also applied our proposed method to three real world data sets, two were obtained
-from the University of California Irvine machine learning repository. The
-first and third data sets are selected because there are clear number of
-clusters in those two data sets. The second data set is used as a benchmark
-data set since it was widely used in literature.
-
-The first one is congress voting data which consists of voting records of
-$98$th United States Congress, $2$nd session \citep{schlimmer1987concept}. 
-This data set includes votes for each of the U.S. House of Representatives Congressmen on the $16$ key votes
-identified by the \textit{CQA} (Congressional Quarterly Almanac). For each
-vote, each Congressman either vote positively ``yea" (voted for/paired
-for/announced for),  negatively ``nay'' (voted against/paired
-against/announced against) or position unknown ``?". We took out those records
-contain ``?" before comparing each algorithm. It results in $232$ remaining
-records, with $124$ democrat and $108$ republican.
-
-The second data set is the well-known Wisconsin breast cancer data set \citep{mangasarian1990pattern}.
-After excluding the records with missing data, this data set consists records of $683$ patients, each with measurements of nine attributes of their biopsy specimens.
-It is known that there exist at least two groups of patients: $444$ patients with benign specimens and $239$ patients with malignant specimens.
-
-The third data set is gene expression data of $5$ types of brain tumours from \cite{de2008clustering}, which contains $42$ observations including $10$ medulloblastomas, $10$ malignant gliomas, $10$ atypical teratoid/rhabdoid tumours (AT/RTs), $8$ primitive neuroectodermal tumours (PNETs) and $4$ normal cerebella.
-It was originally used by \cite{pomeroy2002prediction} to show that medulloblastomas are molecularly distinct from other types of brain tumours.
-There were $7129$ genes originally, but after feature selection, $1379$ genes left.
-So the data set is $42 \times 1379$, which serves as a high dimensional example.
-
-\begin{table}[H]
-\begin{center}
-\captionsetup{justification=centering}
-\caption{\label{table2} Number of clusters selected by each algorithm}
-\begin{tabular}{lccccc}
-    \hline                  
- & Congress Voting && Breast Cancer && Brain Tumours \\ \hline                    
-CH-index & $2$ && $2$  && $2$    \\
-Hartigan & $3$ && $3$  &&  $4$   \\
-Jump & $10$ && $9$ && $1$  \\   
-Prediction strength & $2$ &&  $2$ && $1$    \\
-Bootstrap stability & $2$ &&  $2$ &&  $7$   \\ 
-Gap & $8$ && $10$ && $10$  \\   
-Gaussian-Mix & $2$ && $5$ && $2$  \\   
-Gabriel & $2$  && $3$  && $5$  \\    
-Wold & $2$ && $3$ && $4$ \\  \hline
-
-\end{tabular}
-\end{center}
-\hspace{0.5in} \footnotesize {All the algorithms executed with their default parameter settings with $k$ ranges from $1$ to $10$}
-\end{table}
-
-Since most congressmen vote base on their parties' interest, $2$ parties
-(Democratic and Republican) represent two clusters in this data set. So the
-optimal number should be two. Close inspection shows $k$-means with
-$k=2$ separates the two parties very well with the lowest miss-classification
-error $10.43\%$ ($k=3$ has $14.78\%$). Note that CH-index and Bootstrap
-stability also return $k=2$, but $2$ is the lower bound those methods can
-select for $k$. So it's not clear they actually choose $k=2$ or they hit the
-lower bound (they would pick $k=1$ if allowed). For the breast cancer data,
-it's known to have at least $2$ cluster based on whether it's benign specimens
-or not. But it doesn't mean the optimal $k$ should be $2$.
-\cite{fujita2014non} noticed that the malign group is quite heterogeneous and
-can be further clustered into at least two subgroups. Hence, the result $k=3$
-given by our proposed Gabriel method make sense. For
-the brain tumours gene expression data set which is high-dimension, it is also clear that 
-the correct number of clusters should be $5$, each cluster corresponding to one type of 
-brain tumour. From the table we can see only our proposed
-Gabriel method correctly selects $k=5$, underline how
-difficult it was to pick the right $k$ when dimension is high.
-
-\section{Discussion}
-\label{sec:conc}
-In this paper, we proposed a novel approach to estimate the number of
-clusters. The intuition behind our proposed methods is to transfer the
-unsupervised learning problem into supervised learning problem via novel form
-of cross validation. Such approach is quite different from previous methods which 
-utilize the within/between cluster dispersion or stability criterion for selecting the 
-optimal $k$. Our method utilizes the connection between different dimensions (columns)
-of data through the uniqueness of each cluster center. We proved the self-consistency 
-for our proposed Gabriel CV method as well as its asymptotic property with Gaussian noise,
-and showed the robustness of our method by simulation. Our method has very good performance
-in our limited simulation settings and real data application, and clearly the superior
-one when data has heterogeneous variance or is heavy-tailed.
-
-Besides no strong modeling assumption is required, our proposed method is robust against data set 
-with variance heterogeneity, unequal number of observations, non-Gaussian noise and high-dimension. 
-Such robustness is important in practice because for any given data, it's hard to tell whether
-its clusters have different number of observations, is the variance equal for each cluster,
-or what underlying noise distribution it has. The weakness of our proposed method is 
-that its theory assumes only week correlation between ``predictor'' columns and ``response'' 
-columns. In practice, many data sets don't exhibit high correlations between columns, where
-our proposed method can be safely applied. In case the high correlation does exist, 
-the correlation-corrected version of Gabriel CV method in section $4.3$ can be used if we can 
-assume common covariance structure. Other procedure such as leave out most columns for clustering 
-may also be used to reduce its effect. However, the theory for such procedure has not
-been fully developed and it could be the future research topic.
-
-Another situation where our proposed method cannot be directly used is when the clusters 
-are non-convex. In such situation, $k$-means itself doesn't work well, for example two 
-concentric circles share the same cluster center \citep{hastie2009elements}. However, it's 
-possible that our proposed Gabriel CV method can be used on the transformed data set. In the
-concentric circles case where spectral clustering is appropriate, our proposed method can be applied 
-on the eigenvector subspace of the graph Laplacian matrix inside the spectral clustering algorithm
-to find the optimal $k$. This can also be the future research topic.
-
-\clearpage
-
 \section*{\textbf{APPENDIX}}
 \appendix
-\section{Wold CV estimation} \label{app:foobar}
-\begin{itemize}
-	\item For each $k = 1,2,...,k_{max}$
-	\begin{enumerate}
-		\item Randomly draw some entries in $\mathbf{X}$ missing, keep those hold-out values in vector $V_{true}$
-		\item Impute the missing values with column mean or $0$, denote the imputed data as $\mathbf{X}_{new}$
-		\item Apply the iterative procedure below until converge or stopping criteria reached
-		\begin{itemize}
-			\item Apply $K$-mean on data set $\mathbf{X}_{new}$ with parameter $k$
-			\item Substitute each observation in $\mathbf{X}_{new}$ by its nearest center, get new data $\mathbf{X}^c_{new}$ ($\mathbf{X}_{new}$ keep the same)
-			\item Replace (impute) those imputed values in $\mathbf{X}_{new}$ with the corresponding entries in $\mathbf{X}^c_{new}$
-			\item Calculate the difference between the old and newly imputed values, check whether or not they coincide (converge) 
-		\end{itemize}
-		\item Obtain the last imputed entry values of converged $\mathbf{X}_{new}$, denote it by $V_{converge}$
-		\item Calculate the prediction error  $Error_k = ||V_{true} - V_{converge}||^2$ 
-	\end{enumerate}
-	\item For each CV folder, repeat above procedure and obtain the $Error_k$ for each $k$
-	\item Average $Error_k$ across all folders for each $k$, and then select the $k$ corresponding to the minimum average $Error_k$ 
-\end{itemize}
 

### 2019-06-29 JCGS-R3/main_code/data_examples/Empirical_validation.R

@@ -1,0 +1,241 @@
+## change the working directory to correct location
+## You will need to change the directory path below
+ 
+setwd("C:/Users/fuwei/Desktop/JCGS-R2/data_examples")
+
+   library("cstab")
+   library("bcv")
+   library("devtools")
+   library("cluster")
+   library("e1071")
+   library("mclust")
+   library("MASS")
+   library("nnet")
+   load_all("../lib/fpc")
+   load_all("../lib/NbClust")
+   #source("../lib/NbClust.R")
+   source("../code/classify.R")
+   source("../code/cluster.R")
+   source("../code/gabriel.R")
+   source("../code/jump.R")
+   source("../code/wold.R")
+
+###============ Code to get Table 1 results=====================
+
+######------------Voting data-----------------------############
+Vote <- read.table("Voting.txt",sep=",")
+
+Id <- vector()
+
+for(i in 1:nrow(Vote)){
+      if(sum(Vote[i,] == "?")!=0){
+		Id <- c(Id,i)
+	}
+}
+
+vote <- Vote[-Id,-1]
+
+VV <- matrix(NA,232,16)
+
+for(i in 1:232){
+	for(j in 1:16){
+		if(vote[i,j]=="n"){
+			VV[i,j]=0
+		}else if(vote[i,j]=="y"){
+			VV[i,j]=1
+		}
+	}
+}
+
+set.seed(3)
+ID <- sample(1:232, 232)
+Vote <- VV[ID,]
+
+set.seed(1)
+cv.kmeans.gabriel(Vote, 5, 2, maxcenters=10, classify.method="nearest")$centers
+
+set.seed(1)		
+gabriel_cor_correct(Vote, maxcenters=10, type = 2)
+
+set.seed(1)		
+Wold_holdout(data = Vote, CV = 5, Errortol = 0.01,  max.k =10)
+
+set.seed(1)		
+Gap <- clusGap(Vote, FUN = kmeans, K.max =10)
+which(Gap[[1]][,3] == max(Gap[[1]][,3]))
+			
+set.seed(1)	
+mcluster <- Mclust(Vote, G = 1:10)
+ mcluster$G
+						
+set.seed(1)
[…]

### 2019-06-29 JCGS-R3/main_code/data_examples/confusion_table.R

@@ -1,0 +1,182 @@
+setwd("C:/Users/fuwei/Desktop/JCGS-R2/data_examples")
+
+Tava <- read.csv("Tava.csv", header= T)
+Tava2 <- Tava[,c(1,3)]
+#Tava2[,2] <- as.character(lapply(Tava2[,2],function(x)substr(x,1,7)))
+Tava2[,2] <- as.character(lapply(Tava2[,2],function(x)toupper(x)))
+names(Tava2)[2]<- c("Gene")
+
+Yeast <- read.table("yeast3000.txt")
+Data <- Yeast
+Data[,1] <- as.character(lapply(Data[,1],function(x)toupper(x)))
+
+set.seed(1)
+Kfit <- kmeans(Data[,2:16],5,nstart = 100)
+Data$Gabriel <- Kfit$cluster
+names(Data)[1] <- "Gene"
+
+##merge two data
+DATA <- cbind(Data,Tava2)
+
+##Get 30X5 confusion matrix
+Confusion <- matrix(NA,31,6)
+
+for(i in 1:30){
+	Subset <- DATA[DATA$Cluster==i,];
+	for(j in 1:5){
+		Confusion[i,j] = sum(Subset$Gabriel==j)
+	}
+	Confusion[i,6] = sum(Confusion[i,1:5])
+}
+Confusion[31,] <- colSums(Confusion[1:30,])
+Confusion <- as.data.frame(Confusion)
+colnames(Confusion) <- c("Cluster 1", "Cluster 2","Cluster 3","Cluster 4","Cluster 5","Total")
+
+for(i in 1:30){
+	rownames(Confusion)[i] = paste("Cluster ", i, sep="")
+}
+rownames(Confusion)[31] = "Total"
+
+##convert to latex table 
+library(xtable)
+options(xtable.floating = FALSE)
+xtable(Confusion,digits=0)
+
+###=============To get Table 3 in table ==========
+## reduce to 7X5
+reduced <- Confusion[c(1:4,7,8,14),]
+Other <- as.numeric(unname(Confusion[31,] - colSums(reduced)))
+Final <- rbind(reduced,Other,Confusion[31,])
+xtable(Final,digits=0)
+
+###=============split plots==========
+
+DATA$Tava <- NA
+
+for(i in 1:nrow(DATA)){
+	if( DATA[i,"Cluster"] == 1 ){DATA[i,"Tava"] = "1"}
+	if( DATA[i,"Cluster"] == 2 ){DATA[i,"Tava"] = "2"}
[…]

### 2019-06-29 JCGS-R3/main_code/demo/bench/02_fit_kmeans.R

@@ -1,0 +1,37 @@
+#!/usr/bin/Rscript --vanilla
+
+setwd("C:/Users/fuwei/Desktop/JCGS-R2/demo/bench") ## set the working directory correctly
+
+source("../../code/cluster.R") # cluster_kmeans
+library("parallel")
+
+kmax <- 10
+
+options(cl.cores=max(1, detectCores() - 1))
+cl <- makeCluster(getOption("cl.cores", 2))
+
+clusterExport(cl=cl, varlist=c("cluster_kmeans", "kmax"))
+
+for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
+	for(c in 1:10){
+		f.replicates <- file.path(s, paste0("C" ,c), "replicates.rds")
+		f.kmeans     <- file.path(s, paste0("C" ,c), "kmeans.rds")
+
+		if (file.exists(f.replicates) && !file.exists(f.kmeans)) {
+			cat("fitting k-means for '", s, "' with Case ", c, "\n", sep="")
+
+			replicates <- readRDS(f.replicates)
+			kmeans <- parLapply(cl, replicates, function(r) {
+					km <- vector("list", kmax)
+					for (k in seq_len(kmax)) {
+						km[[k]] <- cluster_kmeans(r$x, k)
+					}
+					km
+				})
+
+			saveRDS(kmeans, f.kmeans)
+		}
+	}
+}
+
+stopCluster(cl)

### 2019-06-29 JCGS-R3/main_code/demo/bench/03_select_nclusters.R

@@ -1,0 +1,83 @@
+#!/usr/bin/Rscript --vanilla
+setwd("C:/Users/fuwei/Desktop/JCGS-R2/demo/bench") ## set the working directory correctly
+
+library("cstab")
+library("bcv")
+library("cluster")
+library("devtools")
+library("e1071")
+library("mclust")
+library("MASS")
+library("nnet")
+library("parallel")
+load_all("../../lib/fpc", export_all=FALSE)
+load_all("../../lib/NbClust", export_all=FALSE)
+source("../../code/classify.R")
+source("../../code/cluster.R")
+source("../../code/gabriel.R")
+source("../../code/jump.R")
+source("../../code/wold.R")
+source("methods.R")
+
+
+options(cl.cores=max(1, detectCores() - 1))
+cl <- makeCluster(getOption("cl.cores", 2))
+
+kmax <- 10
+
+for (s in list.dirs(full.names=FALSE, recursive=FALSE)) {
+	for(c in 1:10){
+		f.replicates <- file.path(s, paste0("C" ,c), "replicates.rds")
+		
+		if (!file.exists(f.replicates))
+			next
+
+		replicates <- readRDS(f.replicates)
+		nrep <- length(replicates)
+
+		if (!dir.exists(file.path(s, paste0("C" ,c), "method")))
+			dir.create(file.path(s, paste0("C" ,c), "method"))
+
+		for (m in names(methods)) {
+			f.method <- file.path(s, paste0("C" ,c), "method", paste0(m, ".rds"))
+
+			if (!file.exists(f.method)) {
+				cat("applying method '", m, "' to '", s, "' with Case ", c, "\n", sep="")
+
+				method <- methods[[m]]
+				nclusters <- integer(nrep)
+
+				clusterExport(cl=cl, varlist=c("replicates", "method", "kmax",
+											   "classify_lda",
+											   "classify_nearest",
+											   "cluster_kmeans",
+											   "compute.jump",
+											   "cv.kmeans.gabriel",
+											   "gabriel_cor_correct", 
+											   "Uncorrelate",
+		 									   "Uncorrelate2",
+											   "Rotate",
+											   "Cov",
+											   "Impute",
+											   "jump",
+											   "kmeans.rndstart",
+											   "predict.classify_nearest",
+											   "mclustBIC",
+											   "WoldIter",
+											   "Wold_holdout"))
[…]

### 2019-06-29 JCGS-R3/main_code/demo/elbow/01_make_plots.R

@@ -1,0 +1,112 @@
+#!/usr/bin/Rscript --vanilla
+setwd("C:/Users/fuwei/Desktop/JCGS-R2/demo/bench") ## set the working directory correctly
+
+library("MASS") 	# mvrnorm
+library("RColorBrewer") # brewer.pal
+
+source("../../code/cluster.R") # cluster_kmeans
+
+
+
+##---------------------Top plot of figure 9-----------------------------------
+
+set.seed(3) # seed chosen to make well-separated clusters
+
+x1 <- mvrnorm(20, c(1.7, 3.0), 0.15 * diag(2))
+x2 <- mvrnorm(20, c(1.0, 1.0), 0.15 * diag(2))
+x3 <- mvrnorm(20, c(3.0, 1.5), 0.15 * diag(2))
+x4 <- mvrnorm(20, c(3.7, 3.7), 0.10 * diag(2))
+
+data <- rbind(data.frame(x=x1[,1], y=x1[,2], group="1"),
+	      data.frame(x=x2[,1], y=x2[,2], group="2"),
+	      data.frame(x=x3[,1], y=x3[,2], group="3"),
+	      data.frame(x=x4[,1], y=x4[,2], group="4"))
+
+
+
+mar <- c(3.25, 3.25, 1.75, 1.75)
+csi <- 0.2 # default value of par("csi")
+mai <- mar * csi
+w <- h <- 3
+mypar <- list(las=1, mai=mai, mgp=c(2.25, 0.75, 0), ps=11, tcl=-0.4)
+
+width <- w + sum(mai[c(1,3)])
+height <- h + sum(mai[c(2,4)])
+
+
+
+pdf("correct-data.pdf", width=width, height=height)
+palette(brewer.pal(6, "Set1"))
+par(mypar)
+
+plot(data$x, data$y, type="n", xlab="X", ylab="Y")
+points(data$x, data$y, pch=as.integer(data$group),
+       col=as.integer(data$group))
+axis(3, labels=FALSE)
+axis(4, labels=FALSE)
+dev.off()
+
+
+centers <- 1:10
+tot.withinss <- rep(NA, length(centers))
+for (i in seq_along(centers)) {
+	cl <- cluster_kmeans(cbind(data$x, data$y), centers[i])
+    tot.withinss[i] <- cl$tot.withinss
+}
+
+
+pdf("correct-withinss.pdf", width=width, height=height)
+par(mypar)
+
[…]

### 2019-06-29 JCGS-R3/main_code/demo/overlap/02_make_plot.R

@@ -1,0 +1,44 @@
+setwd("C:/Users/fuwei/Desktop/JCGS-R2/demo/overlap")
+
+Raw <- read.csv("Raw_result.csv")
+## "Group" column accidently deleted from above data file
+Raw$Group <- rep(1:(31*31),each=10)
+
+Number <- rep(NA,30*30)
+X_ax <- rep(NA,30*30)
+Y_ax <- rep(NA,30*30)
+
+for(i in unique(Raw$Group) ){
+	X_ax[i] <- unique(Raw[Raw$Group==i,]$U_x)
+	Y_ax[i] <- unique(Raw[Raw$Group==i,]$U_y)
+	Number[i] <-  sum(Raw[Raw$Group==i,]$k == 2)
+}
+
+library(plotrix)
+
+pdf("color_plot.pdf")
+
+plot(X_ax,Y_ax,col=color.scale(-Number,c(2,1,0),c(0,2,4),c(3,5,6)),pch=16,cex=2,xlab=expression(mu^X), ylab=expression(mu^Y))
+##-----------------------------------------------
+
+YtoX <- function(U_y){
+	Num <- 2*dnorm(U_y)+2*U_y*pnorm(U_y)-U_y
+	NM <- (Num/U_y +2)/4
+	if(NM<1){
+		return(qnorm(NM))
+	}else{
+		return(NA)
+	}
+}
+##-----------------------------------
+Y_vector <- seq(0.437,3,0.01)
+X_vector <- rep(NA,257)
+
+for(j in 1:257){
+	X_vector[j] <- YtoX(Y_vector[j])
+}
+
+
+lines(X_vector,Y_vector, lwd=3) 
+
+dev.off()

### 2019-06-29 jcgs/demo/nullcorr/02_plot_results.R

@@ -1,0 +1,49 @@
+#!/usr/bin/Rscript --vanilla
+
+
+library("RColorBrewer") # brewer.pal
+
+data <- read.csv("nullcorr.csv")
+data$replicate <- as.factor(data$replicate)
+data$centers <- as.factor(data$centers)
+
+mar <- c(3.25,3.25,0.75, 4.50 + 0.75)
+csi <- 0.2 # default value of par("csi")
+mai <- mar * csi
+w <- h <- 3
+
+for (prior in levels(data$prior)) {
+    filename <- paste0(prior, ".pdf")
+    pdf(filename, width=sum(w, mai[c(2,4)]), height=sum(h, mai[c(1,3)]))
+    palette(brewer.pal(6, "Set1"))
+
+    d <- data[data$prior == prior,]
+
+    par(las=1, mai=mai, mgp=c(2.25,0.75,0), ps=11, tcl=-0.4)
+    plot(d$correlation, d$mse, type="n",
+         xlab="Correlation", ylab="CV Error")
+    axis(3, label=FALSE)
+    axis(4, label=FALSE)
+
+    for (i in seq_len(nlevels(data$centers))) {
+        k <- levels(data$centers)[i]
+        dk <- data[data$centers == k & data$prior == prior,]
+        points(dk$correlation, dk$mse, pch=i, col=i)
+    }
+
+
+    centers <- levels(data$centers)
+    ncenters <- nlevels(data$centers)
+
+    usr <- par("usr")
+    cxy <- par("cxy")
+    legend.x <- usr[2] + 3 * cxy[1]
+    legend.y <- usr[3]
+
+    legend(legend.x, legend.y, xjust=0, yjust=0, xpd=TRUE, bty="n",
+           title="Clusters", title.adj=0,
+           legend=centers, pch=1:ncenters, col=1:ncenters)
+
+    dev.off()
+}
+

### 2019-06-29 mkarxiv.sh

@@ -1,0 +1,46 @@
+#!/usr/bin/env bash
+
+ARXIV=arXiv
+
+TITLE=mhlm
+
+FILES=( \
+    mhlm.bbl \
+    mhlm_supp.pdf \
+    macros.sty \
+    statsoc.cls \
+    figure/ml-10m-misclass.pdf \
+    figure/ml-10m-coef-pairs.png \
+    figure/sim-logistic.pdf \
+    figure/sim-linear.pdf \
+    figure/time-order.pdf
+)
+
+DISTDIR=${ARXIV}/${TITLE}
+
+
+rm -rf "${ARXIV}"
+
+mkdir -p "${DISTDIR}/figure"
+
+for FILE in "${FILES[@]}"
+do
+    cp "${FILE}" "${DISTDIR}/${FILE}"
+done
+
+awk '
+    $0 != "\\end{document}" {
+        print $0
+    }
+
+    END {
+        print "\\includepdf[pages={1-11}]{mhlm_supp}"
+        print "\\end{document}"
+    }
+' mhlm.tex > "${DISTDIR}/mhlm.tex"
+
+
+pushd ${ARXIV} > /dev/null
+rm -f ${TITLE}.tar.gz
+tar czvf "${TITLE}.tar.gz" "${TITLE}" > /dev/null
+popd > /dev/null

### 2019-06-29 JCGS-R3/appendix.tex
fix margin, split out appendix

--- 2019-01-11 JCGS-R2/cvclust.tex
@@
 \addtolength{\textwidth}{1in}%
-\addtolength{\textheight}{-.3in}%
+\addtolength{\textheight}{1.3in}%
 \addtolength{\topmargin}{-.8in}%
@@
 {
-  \title{\bf Estimating the number of clusters using cross-validation}
+  \title{\bf Supplement to ``Estimating the number of clusters using
+cross-validation''}
   \author{Wei Fu and Patrick O. Perry \\
@@
 \bigskip
-\begin{abstract}
-Many clustering methods, including $k$-means, require the user to specify the
-number of clusters as an input parameter. A variety of methods have been
-devised to choose the number of clusters automatically, but they often rely on
-strong modeling assumptions. This paper proposes a data-driven approach to
-estimate the number of clusters based on a novel form of cross-validation. The
-proposed method differs from ordinary cross-validation, because clustering is
-fundamentally an unsupervised learning problem. Simulation and real data
-analysis results show that the proposed method outperforms existing methods,
-especially in high-dimensional settings with heterogeneous or heavy-tailed
-noise. In a yeast cell cycle dataset, the proposed method finds a parsimonious
-clustering with interpretable gene groupings.
-\end{abstract}
-
-\noindent%
-{\it Keywords:} clustering, cross-validation, $k$-means, model selection, unsupervised learning
-\vfill
-
-\newpage
+
 \spacingset{1.45} % DON'T change the spacing!
-\section{Introduction}
-\label{sec:intro}
-
-A clustering procedure segments a collection of items into smaller groups,
-with the property that items in the same group are more similar to each other
-than items in different groups \citep{hartigan1975clustering}.  Such
-procedures are used in two main applications: (a) exploratory analysis, where clusters
-reveal homogeneous sub-groups within a large sample; (b) data
-reduction, where high-dimensional item attribute vectors get reduced to discrete
-cluster labels \citep{jain1999data}.
-
-With many clustering methods, including the popular $k$-means clustering
-procedure, the user must specify $k$, the number of clusters
-\citep{jain2010data}.  One popular ad-hoc device for selecting the number of
-clusters is to use an analogue of the principal components scree plot: plot
-the within-cluster dispersion $W_k$, as a function of the number of clusters
-$k$, looking for an ``elbow'' in the plot.  This approach is simple and often
-performs well, but it requires subjective judgment as to where the elbow is
-located, and as we demonstrate in Appendix~\ref{sec:elbow-fail}, the approach
-can easily fail.  In this report, we propose a new method to choose the number
-of clusters automatically.
-
-The problem of choosing $k$ has been well-studied, and dozens of methods have
-been proposed~\citep{chiang2010intelligent,fujita2014non}. The main difficulty
-in choosing $k$ is that clustering is fundamentally an ``unsupervised''
-learning problem, meaning that there is no obvious way to use ``prediction
-ability'' to drive the model selection \citep{hastie2009elements}.  Most
-existing methods for choosing $k$ instead rely on explicit or implicit
-assumptions about the data distribution, including its shape, scale, and
-correlation structure.
-
-Several authors advocate choosing $k$ by performing a sequence of hypothesis
-tests with null and alternative hypotheses of the form $H_0 : k = k_0$ and
-$H_1: k > k_0$, starting with $k_0 = 1$ and proceeding sequentially with
-higher values of $k_0$ until a test fails to reject $H_0$. The gap statistic
-method typifies this class of methods, with a test statistic that measures the
-within-cluster dispersion relative to what is expected under a reference
-distribution \citep{tibshirani2001estimating}.
-%Similar methods include those developed by \citet{TODO}.
-
-Other authors have proposed choosing $k$ by using information
-criteria.  For example, \citet{sugar2003finding} proposed an approach that
-minimizes the estimated ``distortion'', the average distance per dimension.
-Likewise, \citet{fraley2002model}'s model-based method fits Gaussian mixture model
-models to the data, then selects the number of mixture components, $k$, using
-the Bayesian Information Criterion (BIC).
-
-A third set of approaches is based on the idea of ``stability'', that clusters
-are meaningful if they manifest in multiple independent samples from the same
-population. \citet{ben2001stability}, \citet{tibshirani2005cluster},
-\citet{wang2010consistent} and \citet{fang2012selection} developed methods
-based on this idea.
-
-The procedure we propose in this report is based on a form of
-cross-validation, and it is adaptive to the characteristics of the data
-distribution. The essential idea is to devise a way to measure a form of
-internal prediction error associated each choose of $k$, and then choose the
-$k$ with the smallest associated error. We describe this method in detail in
-Section~\ref{sec:meth}.  In Section~\ref{sec:self-consistent}, we prove that
-our method is self-consistent in certain situations, meaning that in the
-absence of noise, it recovers the correct $k$. Then, in
-Section~\ref{sec:theory-gaussian}, we
-analyze the performance of our method in the presence of Gaussian noise. The
-theoretical analysis shows that the performance of our method degrades in the
-presence of correlated noise; to fix this, we propose a correction for
-correlated noise in Section~\ref{sec:corr-correct}. In
-Sections~\ref{sec:simulations} and~\ref{sec:empirical-validation}, we
-demonstrate that our method is competitive with other state-of-the-art
-procedures in both simulated and real data sets. Then, in
-Section~\ref{sec:application-yeast}, we apply our method to a Yeast cell cycle
-dataset. We conclude with a short discussion in Section~\ref{sec:discussion}.
-
-\section{Cross-validation for clustering}
-\label{sec:meth}
-
-\subsection{Problem statement}
-\label{sec:problem-statement}
-
-Suppose that we are given a data matrix with $N$ rows and $P$ columns, and we
-are tasked with choosing an appropriate number $k$ of clusters to use for
-performing $k$-means clustering on the rows of the data matrix. Recall that
-the $k$-means procedure takes a set of observations $\{ x_1, \dotsc ,x_n \}$
-and finds a set of $k$ or cluster centers $A = \{ a_1, \dotsc, a_k \} $
-minimizing the within cluster dispersion
-\[
-    W(A) = \sum_{i=1}^{n} \min_{g \in \{ 1, \dotsc, k \} } \|x_i - a_g\|^2.
-\]
-This implicitly defines a cluster assignment rule
-\[
-  G(x) = \argmin_{g \in \{1, \dotsc, k\}} \|x - a_g\|^2,
-\]
-with ties broken arbitrarily.
-%We will assume that the $k$-means procedure
-%finds an optimal choice of , $A$, but we will not assume that this solution is
-%unique.
-
-We can consider the problem of choosing $k$, the number of clusters, to be a
-model selection problem. In other domains, especially supervised learning
-problems like regression and classification, cross-validation is popular for
-performing model selection.
-In these settings, the data comes in the form of $N$
-predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
-\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
-$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
-``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  We do this
-by randomly assigning a fold number $r_i$ in the range $\{1, \dotsc, K\}$
-for each row~$i$. For each value $r$ in the range $1, \dotsc, K$, we define
-the $r$th ``fold'' as the matrix gotten by permuting the rows of the 
-original data matrix to get $\dataX$, a matrix with the
-$r$th test subset, the rows $i$ with $r_i = r$ as its trailing rows.  
-We partition $\dataX$ as
-\[
-  \dataX =
-  \begin{bmatrix}
-    \Xtrain & \Ytrain \\
-    \Xtest  & \Ytest
-  \end{bmatrix}.
-\]
-We use the training rows $[ \Xtrain\ \Ytrain ]$, containing roughly $N
-(K-1)/K$ of the observations to fit a regression model
-$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
-test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
-or some variant thereof.  We choose the model with the smallest
-cross-validation error, averaged over all $K$ folds.
-
-In unsupervised learning problems like factor analysis and clustering, the
-features of the observations are not naturally partitioned into ``predictors''
-and ``responses'', so we cannot directly apply the cross-validation procedure
-described above.  For factor analysis, there are at least two versions of
-cross-validation.  \citet{wold78cross} proposed a ``speckled'' holdout, where
-in each fold we leave out a subset of the elements of the data matrix.  Wold's
-procedure works well empirically, but does not have any theoretical support,
-and it requires a factor analysis procedure that can handle missing data.
-\citet{owen2009bi} proposed a scheme called ``bi-cross-validation'' wherein
-each fold designates a subset of the data matrix columns to be response and a
-subset of the rows to be test data.  This generalized a procedure due to
-\citet{gabriel2002biblot}, who proposed holding out a single column and a
-single row at each fold. Gabriel's cross-validation is analogous to
-leave-one-out cross-validation, while Owen and Perry's generalization handles
-more general leave-outs.
-
-%% [POP] This paragraph doesn't fit into the section.  Maybe put it in the introduction?
-%% 
-%% Cross-Validation technique, one of the most commonly used and popular model
-%% selecting method in supervised learning, can not be used naively in
-%% unsupervised learning context. Since prediction error of new observation is
-%% calculated by its distance to the nearest cluster center, more cluster centers
-%% means much tighter fit to the feature space, and hence smaller distance
-%% (prediction error) of observation to its nearest cluster center. This holds
-%% even when the prediction is evaluated on an independent test set
-%% \citep{hastie2009elements}. Therefore, cross-validation prefers larger $k$ if
-%% we do the cross-validation naively.
-
-Gabriel's cross-validation and the Owen and Perry generalization are designed
-specifically for choosing the rank of a low-rank matrix approximation like the
-singular value decomposition or the non-negative matrix factorization. While
-for some purposes it makes sense to view $k$-means clustering as a sort of low
-rank approximation to the data matrix, the discrete nature of clustering (each
-observation gets assigned to one of $k$ possible clusters) imposes additional
-structure on the problem that makes the prior procedures not directly
-applicable.
-
-In the sequel, we extend the version of Gabriel cross-validation described by
-Owen and Perry to the problem of selecting the number of clusters, $k$. To do
-so, we introduce a classification procedure into the cross-validation steps
-that takes the place of least squares regression in the matrix factorization
-problem formation. Our new procedure is able to select the number of clusters,
-$k$, automatically. For this procedure, we provide theoretical and empirical
-support analogous to the consistency results proved by \citet{owen2009bi}.
-
-\subsection{Gabriel cross-validation}
-\label{sec:gabriel-cv-algorithm}
-
-Our version of Gabriel cross-validation for clustering works is designed to
-find the best number of clusters to use for $N$~observations of a
-$P$-dimensional random variable with real-valued components.
-The procedure works by performing a
-sequence of ``folds'' over the data. We use these folds to estimate a version
-of prediction error (cross-validation error) for each possible value of $k$;
-we then choose the value $\hat k$ with the smallest cross-validation error.
-
-In each fold of our cross-validation procedure, we permute the rows and
-columns of the data matrix and then partition the rows and columns as $N = n +
-m$ and $P = p + q$ for positive integers $n$, $m$, $p$, and $q$.  We treat
-the first $p$ columns as ``predictors'' and the last $q$ columns as
-``responses''; similarly, we treat the first $n$ rows as ``train''
-observations and the last $m$ rows as ``test'' observations.  In block form,
-the permuted data matrix is
-\[
-  \dataX
-  =
-  \begin{bmatrix}
-    \Xtrain & \Ytrain \\
-    \Xtest  & \Ytest
-  \end{bmatrix},
-\]
-where
-$\Xtrain \in \R^{n \times p}$,
-$\Ytrain \in \R^{n \times q}$,
-$\Xtest \in  \R^{m \times p}$,
-and
-$\Ytest \in  \R^{m \times q}$.
-
-Given such a partition of $\dataX$, we perform four steps for each value of
-$k$, the number of clusters:
-\begin{enumerate}
-  \item \label{step:gabriel-cluster}
-    \textbf{Cluster:}
-    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding
-    $k$ cluster centers, denoted $\bmuY_1, \dotsc, \bmuY_k$, identified
-    with the labels $1, \dotsc, k$. This also yields the
-    assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ defined analogously
-    to the assignment rule $G$ from Sec.~\ref{sec:problem-statement}, which
-    for any input $Y$, yields the label of the closest cluster center.
-    Set $\hGY_i = \hGY(Y_i)$ to be the assigned cluster label for row~$i$: the
-    label of the closest cluster mean to $Y_i$.
-  \item \label{step:gabriel-classify}
-    \textbf{Classify:}
-    Take $X_{1}, \dotsc, X_n$, the rows of $\Xtrain$ to be predictors,
-    and take $\hGY_1, \dotsc, \hGY_n$ to be corresponding class labels.  Use
-    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to fit a model that can
-    predicts a label from input vector $X$; denote the fitted
-    prediction function by
-    $\hGX : \R^p \to \{ 1, \dotsc, k \}$. In principle this prediction method
-    can be arbitrary but in the sequel we will use linear discriminant
-    analysis.
-  \item \label{step:gabriel-predict}
-    \textbf{Predict:}
-    Apply the classifier to $X_{n+1}, \dotsc, X_{n+m}$, the rows of
-    $\Xtest$, yielding predicted classes $\hGX_i = \hGX(X_i)$ for
-    $i = n+1, \dotsc, n+m$.  For each value of $i$ in this range, compute
-    predicted response $\hat Y_i = \bmuY(\hGX_i)$, where
-    $\bmuY(g) = \bmuY_g$.
-  \item \label{step:gabriel-evaluate}
-    \textbf{Evaluate:}
-    Compute the cross-validation error
-    \[
-      \CV(k) = \frac{1}{m} \sum_{i=n+1}^{n+m} \|Y_i - \hat Y_i\|^2,
-    \]
-    where $Y_{n+1}, \dotsc, Y_{n+m}$ are the rows of $\Ytest$.
-\end{enumerate}
-\noindent
-In principle, we could use any clustering and classification methods in
-steps~\ref{step:gabriel-cluster} and~\ref{step:gabriel-classify}.  In this
-report, we use $k$-means \citep{hartigan1979algorithm} as the clustering algorithm 
-and develop the theoretical properties of the proposed method based on $k$-means. 
-For the classification step, we compute the mean value of $X$ for each class; we assign an
-observation to class $g$ if that class has the closest mean (randomly breaking
-ties between classes).  The classification step is equivalent to linear
-discriminant analysis with equal class priors and identity noise covariance
-matrix.
-
-To choose the folds, we randomly partition the rows and columns into $K$ and
-$L$ subsets, respectively.  Each fold is indexed by a pair $(r,s)$ of
-integers, with $r \in \{1, \dotsc, K\}$ and $s \in \{1, \dotsc, L\}$.  Fold
-$(r,s)$ treats the $r$th row subset as ``test'', and the $s$th column subset
-as ``response''. 
-For this cross-validation to be possible, we must have that
-$K \leq N$ and $L \leq M$.
-We typically take $K = 5$ and $L = 2$.  For the number of
-clusters, we select the value of $k$ that minimizes the average of $\CV(k)$
-over all $K \times L$ folds (choosing the smallest value of $k$ in the event
-of a tie).
-
-In Section~\ref{sec:self-consistent}, we prove that this procedure is
-self-consistent, in the sense that it recovers the correct value of $k$ in the
-absence of noise. Then, in Section~\ref{sec:theory-gaussian}, we analyze some
-of the properties of Gabriel cross-validation in the presence of Gaussian
-noise.
-
-\section{Self-consistency}
-\label{sec:self-consistent}
-
-An important property of any estimation procedure is that in the absence of
-noise, the procedure correctly estimates the truth. This property is called
-``self-consistency'' \citep{tarpey96}. We will now show that Gabriel
-cross-validation is self-consistent. That is, in the absence of noise, the
-Gabriel cross-validation procedure finds the optimal number of clusters.
-
-Self-consistency by itself does not mean that a procedure performs well in
-practice. Self-consistency deals only with the no-noise situation.  Many other
-methods for selecting $K$ are likely self-consistent, but not all perform the
-same in practice. Self-consistency only suggests (but does not guarantee) that
-a procedure might perform reasonably when the noise level is low.
-
-It will suffice to prove self-consistency for a single fold of the
-cross-validation procedure.  As in section~\ref{sec:gabriel-cv-algorithm} we
-assume that the $P$ variables of the data set have been partitioned into $p$
-predictor variables represented in vector~$X$ and $q$ response variables
-represented in vector~$Y$.  The $N$ observations have been divided into two
-sets: $n$ train observations and $m$ test observations.  We state the
-assumptions for the self-consistency result in terms of a specific split; for
-the result to hold in general, with high probability, these assumptions would
-have to hold with high probability for a random split. The following theorem
-gives conditions for Gabriel cross-validation to recover the true number of
-clusters in the absence of noise.
-
-\begin{proposition}\label{prop:self-consistency}
-
-Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
-cross-validation.  For any $k$, let $\CV(k)$ be the cross-validation error for
-this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
-We will assume that there are $K$ ``true cluster centers''
-$\mu(1), \dotsc,\mu(K)$, with
-the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
-\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
-\begin{enumerate}[label=(\roman*)]
-  \item \label{asn:self-consistency-noiseless}
-    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
-    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
-    $i = 1, \dotsc, n+m$.
-  \item \label{asn:self-consistency-distinct-mux}
-    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
-  \item \label{asn:self-consistency-distinct-muy}
-    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
-  \item \label{asn:self-consistency-train}
-    The training set contains at least one member of each cluster: for all $g$
-    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
-    $1, \dotsc, n$ such that $G_i = g$.
-  \item \label{asn:self-consistency-test}
-    The test set contains at least one member of each cluster: for all $g$ in
-    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
-    \dotsc, n+m$ such that $G_i = g$.
-\end{enumerate}
-Then $\CV(k) > \CV(K)$ for $k < K$, and $\CV(k) = \CV(K)$ for $k > K$, so that
-Gabriel cross-validation correctly chooses $k = K$.
-\end{proposition}
-
-The proposition states that our method works well in the absence of noise,
-when each observation is equal to its cluster center.  The assumptions here
-are not checkable in practice, but the proposition suggest that Gabriel
-cross-validation might give a reasonable answer, at least when the noise level
-is low.
-
-The essential
-assumption here is assumption~\ref{asn:self-consistency-noiseless}, which states
-that there is no noise.  If we are willing to assume, say, that the cluster
-centers~$\mu(g) = \bigl(\muX(g),\muY(g)\bigr)$ for $g = 1, \dotsc, K$ were
-randomly drawn from a distribution with a density over $\R^{p+q}$, then
-assumptions~\ref{asn:self-consistency-distinct-mux}
-and~\ref{asn:self-consistency-distinct-muy} will hold with probability one for
-all splits of the data. Likewise, if the clusters are not too small (relative
-to $n$ and $m$), then assumptions~\ref{asn:self-consistency-train}
-and~\ref{asn:self-consistency-test} will likely hold for a random split of the
-data into test and train.
-
-Proposition~\ref{prop:self-consistency} follows from
-Lemmas~\ref{lem:self-consistency1} and~\ref{lem:self-consistency2}, which we
-now state and prove.
-
-\begin{lemma}\label{lem:self-consistency1}
-Suppose that the assumptions of Proposition~\ref{prop:self-consistency} are in
-force.  If $k < K$, then $\CV(k) > 0$.
-\end{lemma}
-\begin{proof}
-By definition,
-\[
-  \CV(k)
-    =
-      \sum_{i=n+1}^{n+m}
-        \| Y_i - \bmuY (\hGX_i) \|^2,
-\]
-where $\bmuY(g)$ is the center of cluster $g$ returned from applying $k$-means
-to $Y_1, \dotsc, Y_n$.  Assumptions~\ref{asn:self-consistency-noiseless}
-and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
-set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all distinct values in
-$\{ \muY(1), \dotsc, \muY(K) \}$.
-Assumption~\ref{asn:self-consistency-distinct-muy} implies that there are
-exactly $K$ such distinct values.  However, there are only $k$ distinct values
-of $\bmuY(g)$.  Thus, at least one summand
-\(
-  \| Y_i - \bmuY(\hGX_i) \|^2
-\)
-is nonzero.  Therefore,
-\(
-  \CV(k) > 0.
-\)
-\end{proof}
-
-\begin{lemma}\label{lem:self-consistency2}
-Suppose that the assumptions of Proposition~\ref{prop:self-consistency} are in
-force.  If $k \geq K$, then $\CV(k) = 0$.
-\end{lemma}
-\begin{proof}
-From assumptions~\ref{asn:self-consistency-noiseless},
-\ref{asn:self-consistency-distinct-muy},
-and~\ref{asn:self-consistency-train}, we know the cluster centers
-gotten from applying $k$-means to $Y_1, \dotsc, Y_n$ must include
-$\muY(1), \dotsc, \muY(K)$.  Without loss of generality, suppose that
-$\bmuY(g) = \muY(g)$ for $g = 1, \dotsc, K$.  This implies that
-$\hGY_i = G_i$ for $i = 1, \dotsc, n$.  Thus, employing
-assumption~\ref{asn:self-consistency-noiseless} again, we get that
-$\bmuX(g) = \muX(g)$ for $g = 1, \dotsc, K$.
-
-Since assumption~\ref{asn:self-consistency-distinct-mux} ensures that
-$\muX(1), \dotsc, \muX(K)$ are all distinct, we must have that $\hGX_i = G_i$
-for all $i = 1, \dotsc, m+n$.  In particular, this implies that $\bmuY(\hGX_i)
-= Y_i$ for $i = 1, \dotsc, m+n$, so that $\CV(k) = 0$.
-\end{proof}
-
-\section{Analysis under Gaussian noise}
-\label{sec:theory-gaussian}
-
-\subsection{Single cluster in two dimensions}
-
-Proposition~\ref{prop:self-consistency} tells us that the Gabriel cross-validation
-method recovers the true number of clusters when the noise is negligible.
-While this result gives us some assurance that the procedure is well-behaved,
-we can bolster our confidence and gain insight into its workings by analyzing
-its behavior in the presence of noise.  We first study the case of a single
-cluster in two dimensions with correlated Gaussian noise.
-
-\begin{proposition}\label{prop:single-2d}
-Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
-of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
-independent draw from a mean-zero multivariate normal distribution with unit
-marginal variances and correlation $\rho$.  In this case, the data are drawn
-from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 1/2$,
-and $k > 1$, then $\CV(1) < \CV(k)$ with probability tending to one as $m$
-and $n$ increase.
-\end{proposition}
-
-\begin{proof}
-
-Throughout the proof we will assume that $\rho \geq 0$; a similar argument
-holds with minor modification when $\rho < 0$.
-
-Set $\hGY_1, \dotsc, \hGY_n$ to be the cluster labels gotten from applying
-$k$-means to $Y_1, \dotsc, Y_n$. Denote the cluster means by $\bmuY_1 \leq
-\bmuY_2 \leq \dotsb \leq \bmuY_k$.  Pollard's \citeyearpar{pollard1981strong}
-strong consistency theorem for $k$-means implies that for large $n$, the
-cluster centers are close to population clusters centers $a_1 < a_2 < \dotsb <
-a_k$. Specifically, $\bmuY_j = a_j + \OhP(n^{-1/2})$.  Since the distribution
-of $Y$ is symmetric, the population centers $a_1, a_2, \dotsc, a_k$ are
-symmetric about the origin.
-
-For $j$ in $1, \dotsc, k$, set
-\[
-  \bmuX_j = \frac{\sum_{i=1}^{n} 1\{ \hGY_i = j \} X_i}
-                 {\sum_{i=1}^{n} 1\{ \hGY_i = j \}},
-\]
-where $1\{ . \}$ denotes the indicator function.
-The classification rule $\hGX$ is defined by
-\(
-  \hGX(X) = \argmin_j \| \bmuX_j - X \|.
-\)
-Denote the boundaries between the population clusters as
-$b_j = (a_j + a_{j+1})/2$ for $j = 1, \dotsc, k-1$.
-Set $b_0 = -\infty$ and $b_k = +\infty$. 
-Then, $\bmuX_j$ is within $\OhP(n^{-1/2})$ of the following expectation:
-\begin{align*}
-  \E(X \mid b_j \leq Y \leq b_{j+1})
-    &= \E\{ E(X \mid Y) \mid b_j \leq Y \leq b_{j+1}\} \\
-    &= \E\{ \rho Y \mid b_j \leq Y \leq b_{j+1}\} \\
-    &= \rho \E\{ Y \mid b_j \leq Y \leq b_{j+1}\} \\
-    &= \rho a_j.
-\end{align*}
-That is, $\bmuX_j = \rho a_j + \OhP(n^{-1/2})$. For $j = 1, \dotsc, k-1$,
-the boundary between sample the classification based on $X$ to
-labels $j$ and $j+1$ is
-$(\bmuX_j + \bmuX_{j+1}) /2 = \rho b_j + \OhP(n^{-1/2})$.
-
-Set $\hGX_i = \hGX(X_i)$.
-The cross-validation error is
-\begin{align*}
-  \CV(k)
-  &= \frac{1}{m} \sum_{i=n+1}^{n+m} \|Y_i - \hat Y_i\|^2
-  \\
-  &= \frac{1}{m} \sum_{i=n+1}^{n+m}
-  \sum_{j=1}^{k} \|Y_i - \bmuY_j\|^2 1\{ \hGX_i = j \}
-  \\
-  &=
-  \sum_{j=1}^{k}
-    \pi_j
-    \E[ \|Y - a_j\|^2 \mid \rho b_j < X < \rho b_{j+1}]
-  + \OhP(n^{-1/2})
-  + \OhP(m^{-1/2}),
-\end{align*}
-where $\pi_j = \Pr(\rho b_j < X < \rho b_{j+1}).$
-
-For $j = 1, \dotsc, k$,
-set
-\(
-  \hmuY_j =
-  \E[ Y  \mid \rho b_j < X < \rho b_{j+1}].
-\)
-Note that
-\[
-  \CV(1)
-    = 
-  \sum_{j=1}^{k}
-    \pi_j
-    \E[ \|Y - 0\|^2 \mid \rho b_j < X < \rho b_{j+1}]
-  + \OhP(n^{-1/2})
-  + \OhP(m^{-1/2}).
-\]
-Thus, the difference in cross-validation errors is
-\begin{align*}
-\CV(k) - \CV(1)
-  &=
-  \sum_{j=1}^{k}
-    \pi_j
-    a_j
-    (a_j - 2 \hmuY_j)
-  + \OhP(n^{-1/2})
-  + \OhP(m^{-1/2}).
-\end{align*}
-For arbitrary $j$,
-\begin{align*}
-  \hmuY_j &=
-  \E[ E(Y\mid X)  \mid \rho b_j < X < \rho b_{j+1}]
-\\
-  &= \rho \E[ X  \mid \rho b_j < X < \rho b_{j+1}].
-\end{align*}
-Since $0 \leq \rho \leq 1$,
-in cases where $0 \leq b_j < b_{j+1}$, we have that
-$\hmuY_j \leq \rho a_j$; similarly, when $b_j < b_{j+1} \leq 0$, it follows
-that $\hmuY_j \geq \rho a_j$. In either of these two situations, if $\rho <
-1/2$, then
-\[
-  a_j (a_j - 2 \hmuY_j) > 0.
-\]
-The last situation to consider is when $b_j < 0 < b_{j+1}$, in which case
-$b_j = - b_{j+1}$; here, $\hmuY_j = a_j = 0$.
-Putting this all together, we have that as $n$ and $m$ tend to infinity,
-the probability that
-\(
-  \CV(k) > \CV(1)
-\)
-tends to one.
-\end{proof}
-
-The proof of Proposition~\ref{prop:single-2d} relies on normality assumptions
-in two crucial ways: (i) assuming that $X$ and $Y$ have symmetric marginal
-distributions; and (ii) assuming that $E(Y|X) = \rho X$. We could relax the
-normality assumption by instead supposing that $X$ and $Y$ can be decomposed
-as $X = Z_1$ and $Y = \rho Z_1 + \sqrt{1-\rho^2}Z_2$ where $Z_1$ and $Z_2$ are
-independent random variables with mean zero, unit variance and symmetric
-distribution. Any bivariate distribution can be put in this form for
-uncorrelated $Z_1$ and $Z_2$, but only special distributions, including the
-bivariate normal, can be put in the form with independent $Z_1$ and $Z_2$.
-
-We confirm the result of Proposition~\ref{prop:single-2d} with a simulation.
-We perform $10$ replicates.  In each replicate, we generate $20000$
-observations from a mean-zero bivariate normal distribution with unit marginal
-variances and correlation $\rho$.  We perform a single $2 \times 2$ fold of
-Gabriel cross-validation and report the cross-validation mean squared error
-for the number of clusters $k$ ranging from $1$ to $5$.
-Figure~\ref{fig:nullcorr-equal} shows the cross-validation errors for all $10$
-replicates.  The simulation demonstrates that in the Gabriel cross-validation
-criterion chooses the correct answer $k = 1$ whenever $\rho < 0.5$; the
-criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.
-
-\begin{figure}
-\centering
-\includegraphics[width=3in]{main_code/demo/nullcorr/equal.pdf}
-\caption{Gabriel cross-validation performance in a single-cluster simulation.
-    Gabriel cross-validation error $\CV(k)$ for $k$ ranging from 1 to~5,
-    averaged over $10$ replicates where
-    data is generated from two-dimensional
-multivariate normal distribution with correlation $\rho$.  The 
-    criterion chooses the correct answer $k = 1$ whenever
-$|\rho| < 0.5$; the criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.}
-\label{fig:nullcorr-equal}
-\end{figure}
-
-Intuitively, when the correlation is high, the response feature, $Y$, looks
-similar to the predictor feature, $X$. Prediction error on $X$ always
-decreases with larger $k$. Thus, when the correlation is high, the prediction
-error for $Y$ will also decrease with larger $k$. This explains why
-cross-validation breaks down in the presence of strong correlation.
-
-In Appendix~\ref{sec:single-general-dim}, using similar techniques to those used
-to prove Proposition~\ref{prop:single-2d}, we derive an analogous result for correlated
-Gaussian noise in more than two dimensions. A similar phenomenon holds:
-the Gabriel cross-validation method fails when the first principal component
-of the $Y$ variables is strongly correlated with a linear combination of the
-$X$ variables.
-
-Proposition~\ref{prop:single-2d} tells us that Gabriel cross-validation fails
-when there is strong correlation between the variables. To get around this, in
-practice we will transform the data to reduce correlation before performing
-cross-validation. We detail this approach in Section~\ref{sec:corr-correct}.
-
-\subsection{Two clusters in two dimensions}
-
-We will now analyze a simple two-cluster setting, and derive conditions for
-Gabriel cross-validation to correctly prefer $k=2$ clusters to $k=1$.  The
-main assumption of the proposition is that the cluster centers are not too
-close. The precise definition of ``too close'' is stated in terms of
-$\Phi(\cdot)$ and $\varphi(\cdot)$, the standard normal cumulative
-distribution function and density, respectively.  The inequality is hard to
-interpret directly, but we show the boundary between ``too close'' and ``well
-separated'' in Fig.~\ref{fig:overlap-color_plot}, after the proof of the
-proposition.
-
-\begin{proposition}\label{prop:twoclust}
-Suppose that $\{(X_i,Y_i)\}_{i=1}^{n+m}$ is data from a single fold of Gabriel
-cross-validation, where each $(X,Y)$ pair in $\R^2$ is an independent draw
-from an equiprobable mixture of two multivariate normal distributions with
-identity covariance. Suppose that the first mixture component has mean
-$\mu = (\muX, \muY)$ and the second has mean $-\mu = (-\muX, -\muY)$,
-where $\muX \geq 0$ and $\muY \geq 0$. 
-If the cluster centers are well separated, specifically such that
-\(
-  2 \varphi(\muY) + \muY + 2 \muY \Phi(\muY) < 4 \muY \Phi(\muX),
-\)
-then $\CV(2) < \CV(1)$ with probability tending to one as $m$ and $n$ increase.
-\end{proposition}
-
-\begin{proof}
-There are two clusters: observations from cluster~$1$ are distributed as
-\(
-  \mathcal{N}(\mu, I)
-\)
-and observations from cluster~$2$ are distributed as
-\(
-\mathcal{N}(-\mu, I)
-\)
-where $\mu = (\muX, \muY)$.  Without loss of generality, $\muX \geq 0$ and
-$\muY \geq 0$.  Let $G_i$ be the true cluster of observation~$i$
-where, by assumption,
-\[
-  \Pr(G_i=1) = \Pr(G_i=2) = 1/2.
-\]
-After applying $k$-means to $\{ Y_i\}_{i=1}^{n}$ with $k=2$, if $n$ is large
-enough, then the estimated cluster means $\bar{\mu}^Y_1$ and $\bar{\mu}^Y_2$
-will be close to $\E(Y \mid Y>0)$ and $\E(Y \mid Y < 0)$, with errors of
-size $\OhP(n^{-1/2})$. To compute these quantities, let
-$(X_1, Y_1) \sim \mathcal{N}(\mu, I)$ and $(X_2, Y_2) \sim \mathcal{N}(-\mu, I)$
-be draws from the mixture components, and let $(X,Y)$ be defined such that
-$\Pr(X= X_1, Y = Y_1) = \Pr(X= X_2, Y = Y_2) = 1/2$. Then,
-\begin{align*}
-  \E(Y \mid Y>0)
-  &= \E(Y_1 \mid Y_1 > 0) \cdot \Pr(Y = Y_1 \mid Y >0)
-   + \E(Y_2 \mid Y_2 > 0) \cdot \Pr(Y = Y_2 \mid Y > 0)
-\\
-&=  \{ \muY + \varphi(\muY) / \Phi(\muY) \} \cdot \Phi(\muY)
-+ [-\muY + \varphi(\muY) / \{ 1 - \Phi(\muY) \}] \cdot \{ 1 - \Phi(\muY) \}
-\\
-  &= 2\varphi(\muY)+ 2\muY\Phi(\muY)- \muY. 
-\end{align*}
-In the second line, we have used
-Lemma~\ref{lem:truncated-normal-moments} from
-Appendix~\ref{app:technical-lemmas} to compute the conditional expectations;
-$\varphi()$ and $\Phi()$ are the standard normal density and
-cumulative distribution function, respectively. By symmetry,
-\begin{align*}
-  \E(Y \mid Y<0) &= -\E(Y \mid Y > 0).
-\end{align*}
-
-The classification rule learned from the training data
-$\{(X_i, \hGY_i)\}_{i=1}^{n}$ will have its decision boundary at $0 + \OhP(n^{-1/2})$;
-that is, in the limit,
-observations will get classified as coming from cluster~1 when $X > 0$. 
-Set $a = \E(Y \mid Y > 0)$.
-Up to terms of order $\OhP(n^{-1/2})$,
-the cross-validation error from a single observation is distributed as
-\[
-  (Y - a)^2 1\{ X > 0 \} + (Y + a)^2 1\{X < 0\}.
-\]
-Using the fact that conditional on the mixture component, the $X$ and $Y$
-coordinates are independent, we can compute the expectation of the first
-summand as
-\begin{align*}
-  \E[(Y - a)^2 1\{ X > 0 \}]
-  &= (1/2) \E[(Y_1 - a)^2] \Pr(X_1 > 0)
-  + (1/2) \E[(Y_2 - a)^2] \Pr(X_2 > 0)
-\\
-&= (1/2) \Phi(\muX) \E(Y_1 - a)^2  + (1/2) \{ 1 - \Phi(\muX) \} \E(Y_2 - a)^2
-\\
-&=
-(1/2) [ 1  + \Phi(\muX) (\muY - a)^2 + \{ 1 - \Phi(\muX) \} (-\muY - a)^2 ].
-\end{align*}
-By a similar calculation, the expectation of the second summand is
-\[
-  \E[(Y + a)^2 1\{ X < 0 \}]
-  =
-  (1/2)[ 1 + \{ 1 - \Phi(\muX) \} (\muY + a)^2 + \Phi(\muX) (-\muY + a)^2 ].
-\]
-Adding the two terms, we get that the expected cross-validation error from a
-single observation is
-\[
-  1 + \Phi(\muX) (\muY - a)^2 + \{ 1 - \Phi(\muX) \} (\muY + a)^2
-    =
-  1 + (\muY)^2 + a \, \{  a + 2 \muY - 4 \muY \Phi(\muX) \}.
-\]
-Thus, the $k = 2$ cross-validation error on the test set is
-\begin{align*}
-  \CV(2)
-  &=
-    \frac{1}{m}
-    \sum_{i=n+1}^{n+m}
-      \| (Y_i - \bmuY_1) 1\{\hGX_i = 1\} \|^2
-      +
-      \| (Y_i - \bmuY_2) 1\{\hGX_i = 2\} \|^2
-\\
-&=
-1 + (\muY)^2 +  a \, \{ a + 2 \muY - 4 \muY \Phi(\muX) \}
-+ \OhP(n^{-1/2}) + \OhP(m^{-1/2}).
-\end{align*}
-
-When $k=1$, the $k$-means centroid is equal to the sample mean $\bar Y_n = (1/n)
-\sum_{i=1}^{n} Y_i$, approximately equal to 
-$\E(Y) = 0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error is
-\[
-  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
-         = 1 + (\mu^Y)^2 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
-\]
-Thus, if $a + 2 \muY - 4 \muY \Phi(\muX) < 0$, then $\CV(2) < \CV(1)$ with
-probability tending to one as $m$ and $n$ increase. Substituting the
-expression for $\E(Y \mid Y > 0)$ in place of $a$, the inequality holds
-precisely when
-\(
-  2 \varphi(\muY) + \muY + 2 \muY \Phi(\muY) < 4 \muY \Phi(\muX).
-\)
-\end{proof}
-
-We have stated and proved Proposition~\ref{prop:twoclust} for data in $d = 2$
-dimensions, but there is nothing essential about the particulars of the
-dimension; we generalize the result to arbitrary dimension in
-Appendex~\ref{app:twoclust-anyd}.
-
-In principle the proof of Proposition~\ref{prop:twoclust} generalizes to
-non-normal location families of distributions easily, provided that
-both cluster distributions are in the same family and expressions for
-$\E(Y | Y > 0)$ and the other related quantities are available (and finite).
-Different distributions for $(X,Y)$ give rise to different cutoffs for when $k
-= 1$ cluster is preferred to $k = 2$. The normal distribution is only special
-here in that the required conditional expectations and the boundary where
-$\CV(1) = \CV(2)$ are computable in closed form.
-
-We confirm the result of Proposition~\ref{prop:twoclust} with a simulation.
-We perform $10$ replicates for each $(\mu^X, \mu^Y)$ pair, sweeping over
-a two-dimensional grid of values in the domain $[0,3] \times [0,3]$, with step
-size $0.1$ in each dimension.  In each
-replicate, we generate $N=20000$ observations from an equiprobably mixture
-of multivariate normal distributions with identity covariance, with one
-component having mean $(\muX, \muY)$ and
-the other component having mean $(-\muX, -\muY)$. We perform a single $2 \times 2$
-fold of Gabriel cross-validation and report the number of times
-(out of $10$ replicates) where $k=2$ is selected by the algorithm instead of $k=1$.
-Figure~\ref{fig:overlap-color_plot} shows the frequency with which $k=2$ is selected by
-the algorithm for each $(\mu^X, \mu^Y)$ pair. Darker (red) colors indicate
-higher numbers (close to $10$), situations where $k = 2$ is selected more
-often than $k = 1$. Ligher (blue) colors indicate that $k = 1$ is preferred.
-We can see the simulation result perfectly align with the
-theoretical curve (the black line), which separates the $k=2$ zone from the
-$k=1$ zone.
-
-\begin{figure}
-\centering
-\includegraphics[width=3in]{main_code/demo/overlap/color_plot.pdf}
-\caption{
-    Gabriel cross-validation performance in a two-cluster simulation.
-    For a grid of values for $\muX$ and $\muY$ ranging from $0$ to $3$,
-    we generated data according the manner described in
-    Prop.~\ref{prop:twoclust}, then computed Gabriel cross-validation
-    error $\CV(1)$ and $\CV(2)$. The heat map shows the proportion of times
-    that $\CV(2) < \CV(1)$ with
-    light (blue) indicating that $k = 1$ is preferred to $k = 2$, and
-dark (red) indicating that $k = 2$ is preferred to $k = 1$.
-    The black line is the theoretical boundary determined from
-Proposition~\ref{prop:twoclust}. The simulation confirms our theoretical
-    boundary.}
-\label{fig:overlap-color_plot}
-\end{figure}
-
-\section{Adjusting for correlation}
-\label{sec:corr-correct}
-
-Proposition~\ref{prop:single-2d} shows that when the correlation between
-dimensions is high, the Gabriel cross-validation method tends to overestimate
-the number of clusters,~$k$. To mitigate this effect, we propose a two-stage
-estimation procedure that attempts to transform the data to minimize the
-correlation between features. In the first stage, we get a preliminary
-estimate for the number of clusters,~$\hat k_0$, and we use this value to get
-an estimate of the noise covariance matrix. Then, in the second stage, we
-transform the data attempting to sphere the noise covariance, and re-estimate
-the number of clusters, getting a final estimate~$\hat k$.
-
-The details of the correlation correction procedure are as follow:
-
-\begin{enumerate}
-
-  \item Apply the Gabriel cross-validation method on the original data
-    $\dataX$ to get a preliminary estimate of the number of
-    clusters,~$\hat k_0$.
-
-  \item Apply $k$-means to the full data set with observations
-    $\dataX_1, \dataX_2, \dotsc, \dataX_N$ using $k = \hat k_0$ clusters.
-    For $i = 1, \dotsc, N$, let $\hat \mu_i$ denote the assigned cluster
-    mean for the $i$th observation.
-
-	\item Estimate the noise covariance matrix $\hat{\Sigma}$:
-    \[
-      \hat{\Sigma}
-        =
-        \frac{1}{N - \hat k_0}
-        \sum^N_{i=1}
-        (\dataX_i-\hat{\mu}_i)(\dataX_i-\hat{\mu}_i)^\T.
-    \]
-
-  \item \label{step:transform}
-    Compute a decomposition $\hat{\Sigma} = \Gamma\Lambda\Gamma^\T$
-    where $\Gamma$ has orthonormal columns and
-    $\Lambda$ is diagonal and positive-definite.
-    Choose a random (Haar distributed) $P \times P$ orthogonal matrix $Q$.
-    Rescale and rotate the original data matrix $\dataX$ to get a
-    transformed data matrix defined by
-    \[
-      \tilde{\dataX} = \dataX\Gamma\Lambda^{-1/2}Q.
-    \]
-
-  \item Apply Gabriel cross-validation method to transformed data matrix
-    $\tilde{\dataX}$ to get a final estimate for the number of
-    clusters,~$\hat k$.
-
-\end{enumerate}
-
-The noise covariance estimate assumes a shared covariance matrix for all
-$k$~clusters. Letting $G_i$ denote the cluster membership of the $i$th
-observation, and letting $\mu(g)$ denote the mean of cluster $g$ for $g = 1,
-\dotsc, k$, the model supposes that
-\[
-  \dataX_i = \mu(G_i)+\varepsilon_i,
-\]
-where~$\varepsilon_i$ has mean zero and covariance matrix $\Sigma$, independent
-of~$G_i$. If we knew $\Sigma$, then we could transform the observations as
-\begin{align*}
-  \Sigma^{-1/2} \dataX_i
-    &= \Sigma^{-1/2} \mu(G_i) + \Sigma^{-1/2} \varepsilon_i \\
-    &= \tilde \mu(G_i) + \tilde \varepsilon_i,
-\end{align*}
-where $\tilde \mu = \Sigma^{-1/2} \mu$ and
-$\tilde \varepsilon_i = \Sigma^{-1/2} \varepsilon_i$. The transformed data
-has the same number of clusters, but has noise covariance
-$\cov(\tilde \varepsilon_i) = I$. The matrix product $\Gamma \Lambda^{-1/2}$
-used in step~\ref{step:transform} is an estimate of $\Sigma^{-1/2}$.
-
-The transformation used in step~\ref{step:transform} uses a random orthogonal
-matrix~$Q$, which gets applied to the rows of $\dataX$ after 
-multiplying by the estimate of $\Sigma^{-1/2}$. We use this random orthogonal
-matrix for two reasons. First, it ensures that in expectation, each transformed cluster mean
-$Q \Lambda^{-1/2} \Gamma^\T \mu(g)$ for $g = 1, \dotsc, k$ is uniformly
-spread across all $P$ features. This ensures that the self-consistency
-conditions on the cluster centers enumerated in Proposition~\ref{prop:self-consistency}
-are likely to hold. The second reason for multiplying by $Q$ is to spread any
-remaining correlation in the noise evenly (in expectation) across all
-dimensions. The latter effect follows since if $Z$ is a random vector with
-covariance matrix $\Theta$, then $Q Z$ has covariance matrix $Q \Theta Q^\T$,
-which has expectation $E(Q \Theta Q^\T) = \tr(\Theta) I$.
-
-Our correlation correction is not backed by a rigorous theoretical
-justification. Further, since our initial estimate $\hat k_0$ tends to
-over-estimate the true number of clusters in the presence of strong
-correlation, our covariance estimate $\hat \Sigma$ will often be biased.
-Fortunately, we do not need to estimate $\Sigma$ with high accuracy. For our
-correction to be useful, we only need $\tilde{\dataX}$ to have lower
-within-cluster correlation than the raw data~$\dataX$. In our simulations and
-in our empirical validation study in Sections~\ref{sec:simulations}
-and~\ref{sec:empirical-validation} we demonstrate that despite its theoretical
-shortcoming, our ad-hoc correlation adjustment often performs well in
-practice.
-
-In many applications, including our application in the sequel to a yeast cell
-cycle dataset, the number of observations $N$ is below the number of
-features~$P$. When this happens, the sample covariance matrix $\hat \Sigma$
-will not have full rank~$P$, and will be a poor estimate of the population
-covariance matrix. In these situations, though, it will still be the case that
-a decomposition $\hat \Sigma = \Gamma \Lambda \Gamma^\T$ exists with $\Gamma$
-having orthonormal columns and $\Lambda$ being diagonal and positive-definite.
-It will also be the case that $\Lambda^{-1/2}$ and hence $\tilde{\dataX}$
-exist. What fails in these situations is that, since $\hat \Sigma$ is no longer
-a good estimate of $\Sigma$, it will no longer be the case that our correction
-reliably transforms the noise covariance to $I$. Still, in these situations,
-applying the correlation correction may give better results than using the
-original untransformed data. In practice we inspect the resulting estimate
-$\hat k$ both with and without the correlation correction, and if the two
-answers agree, we can be more confident in our answer.
-
-\section{Performance in simulations}
-\label{sec:simulations}
-
-\subsection{Overview}
-
-In this section, we perform a set of simulations to evaluate the performance
-of our proposed method and the associated correlation correction described in
-Section~\ref{sec:corr-correct}.  We compare our method with a basket of
-competing methods including the Gap statistic
-\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
-\citep{fraley2002model}, the CH-index \citep{calinski1974dendrite}, Hartigan's
-statistic \citep{hartigan1975clustering}, the Jump method
-\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
-Slope method \citep{fujita2014non} and Bootstrap stability \citep{fang2012selection}. 
-We use the default parameter settings for all competing methods.  For Gabriel cross-validation,
-we perform $2$-fold cross-validation on the columns~($p=q$) and $5$-fold
-cross-validation on the rows~($m=n/4$).  We also compare with Wold
-cross-validation, which we describe in Appendix~\ref{sec:wold-cv}.
-
-For the competing methods, some might take issue with the fact that we are
-using default parameter settings.  In many other benchmark comparisons, one
-would choose the tuning parameters using some form of cross-validation. In our
-case, this is not an option. We are in an \emph{unsupervised} setting, so
-there is no clear way to set the tuning parameters. We must rely on the
-defaults from the implementations of these methods.
-
-In all simulation settings, we randomly generate cluster centers by drawing
-from a multivariate normal distribution with covariance matrix $\tau I$,
-conditional on the cluster points being well-separated. By ``well-separated''
-we mean that the distance from any point to its own cluster center is less
-than the distance of that point to any other cluster center by a margin of
-at least $1.0$ units. In any replicate where the clusters are not
-well-separated, we discard the replicate and generate another set of points.
-We choose $\tau$ to make the probability the cluster
-centers being well-separated on the first draw to be equal to approximately
-50\%.  The ``well-separated'' condition and some of our simulation settings
-are chosen to mimic those used by~\cite{tibshirani2001estimating} in
-settings (c) and (d) of their Section~6.
-
-For each setting, we perform $100$ replicates. We report the number of times
-that each method finds the correct number~$k$ of clusters. We also report 95\%
-confidence intervals for the proportions, using Wilson's
-method~\citep{wilson1927probable}.  The simulations demonstrate that overall,
-the proposed Gabriel cross-validation method and its correlation-corrected
-version compare well with the competing methods, and they are robust to
-variance heterogeneity, high dimensional data, and heavy-tail data.
-
-\subsection{Setting 1: Correlation between dimensions}
-
-We generate six clusters in $10$ dimensions.  Each cluster has $100$ or $50$
-multivariate normal observations with common covariance matrix $\Sigma$ which
-has compound symmetric structure with $1$ in diagonal and $\rho$ off diagonal.
-$\rho$ takes value in $\{0,0.1,...,0.9\}$.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting1/Facet.pdf}
-\caption{Setting 1 performance as correlation increases, faceted by estimation
-    method. For each estimate method, the subplot shows the proportion of
-    replicates in which estimated number of clusters agrees with the truth
-    versus the correlation between dimensions,~$\rho$. Gabriel
-    cross-validation performance degrades under moderate correlation, but
-    the correlation-corrected version performs well across all values of
-    $\rho$.}
-\label{fig:setting1}
-\end{figure}
-
-We can see that high correlation between dimensions causes problem for most
-existing methods, including Gabriel cross-validation method without the
-correlation correction.  The only two methods that work well in the presence
-of high correlation are the Gaussian model-based BIC method
-\citep{fraley2002model} and the correlation-corrected Gabriel method.
-
-\subsection{Setting 2: Noise dimensions}
-
-We generate three clusters in $6$ dimensions. Each cluster has $1000$ or $500$
-multivariate normal observations with identity covariance matrix.
-We add $r$ dimensions of noise to the data, randomly generated from a uniform
-distribution on $[0,1]$. The noise dimension $r$ takes values in
-$\{0,6,...,54\}$.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting2/Facet.pdf}
-\caption{Setting 2 performance as the number of noise dimensions increases,
-    faceted by estimation method. For each estimate method, the subplot shows the proportion of
-    replicates in which estimated number of clusters agrees with the truth
-    versus the number of noise dimensions,~$r$. Both variants of Gabriel
-    cross-validation perform reasonably under all settings of~$r$, though
-    not as well as some of the other methods.}
-\label{fig:setting2}
-\end{figure}
-
-Most methods are relatively insensitive to adding more noise dimensions; the
-one exception to this is the Jump method, which deteriorates significantly in
-the presence of extra noise dimensions.
-
-\subsection{Setting 3: High dimension}
-
-We generate eight clusters in $P$ dimensions, with $P$ taking values in
-$\{10,20,...,100\}$.  Each cluster has $100$ or $50$ multivariate
-normal observations with identity covariance matrix.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting3/Facet.pdf}
-\caption{Setting 3 performance as the number of dimensions increases,
-    faceted by estimation method. For each estimate method, the subplot shows the proportion of
-    replicates in which estimated number of clusters agrees with the truth
-    versus the number of noise dimensions,~$r$. Both variants of Gabriel
-    cross-validation perform with perfect accuracy in the replicates
-    with $r \geq 30$ dimensions.}
-
-\caption{Performance as number of dimensions increases.}
-\label{fig:setting3}
-\end{figure}
-
-Some methods, like Jump and Gap, are insensitive to higher dimensions while
-other methods deteriorate quickly with increasing dimension, most notably the
-Gaussian model-based BIC method.  Gabriel cross-validation and its
-correlation-corrected version tend to work better in higher dimensions.
-Slope method also works better in higher dimensions.
-
-\subsection{Setting 4: Variance heterogeneity}
-
-We generate three clusters in $20$ dimensions. Each cluster has $60$
-observations.  Observations are generated from
-$\mathcal{N}\left(\mathbf{0},\sigma_1^2\mathbf{I}\right)$,
-$\mathcal{N}\left(\mathbf{0},\sigma_2^2\mathbf{I}\right)$ and
-$\mathcal{N}\left(\mathbf{0},\sigma_3^2\mathbf{I}\right)$ where $\sigma_1^2 :
-\sigma_2^2: \sigma_3^2 = 1:\frac{1+R}{2}:R$. The maximum ratio $R$ takes
-values in $\{1,5,10,...,45\}$
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting4/Facet.pdf}
-\caption{Setting 4 performance as the variance heterogeneity increases,
-    faceted by estimation method. For each estimate method, the subplot shows the proportion of
-    replicates in which estimated number of clusters agrees with the truth
-    versus the variance heterogeneity parameter,  $R$. Both variants of Gabriel
-    cross-validation perform with near-perfect accuracy for all values of~$R$.}
-\label{fig:setting4}
-\end{figure}
-
-This setting demonstrates that most existing methods are sensitive to variance
-heterogeneity, most notably the Gap method, the Slope method and the model-based BIC method. The
-proposed Gabriel cross-validation method and its correlation-corrected version
-consistently perform well in estimating $k$ and they are insensitive to
-variance heterogeneity.
-
-\subsection{Setting 5: Heavy tail data}
-
-We generate five clusters in $15$ dimensions. Each cluster has $80$
-observations.  Observations have independent $t$ distributions in each
-dimension, with degrees of freedom $\nu$ taking values in $\{11,10,...,2\}$
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting5/Facet.pdf}
-\caption{Setting 5 performance as the tail heaviness increases,
-    faceted by estimation method. For each estimate method, the subplot shows the proportion of
-    replicates in which estimated number of clusters agrees with the truth
-    versus the $t$~distributed noise degrees of freedom,~$\nu$, plotted
-    backwards so that $\nu$ decreases from $11$ to $2$ going from left to
-    right. Both variants of Gabriel cross-validation are robust to moderate
-    tail heaviness, with slower performance degradation than most of the other
-    estimation methods as the tail heaviness increases.}
-\label{fig:setting5}
-\end{figure}
-
-This setting investigates performance in the presence of heavy-tailed data.
-When the degrees of freedom decreases, the tail becomes more flat and the
-Gaussian assumption becomes more inappropriate.  For most methods, their
-performances are relatively stable until the tail gets very heavy. In the case
-where there are $2$ degrees of freedom, the Slope method, the Gap and Jump methods' performances
-deteriorate considerably relative to the Gabriel method.
-
-\section{Empirical validation}
-\label{sec:empirical-validation}
-
-To further validate our method, we applied it to three real world data sets
-with known clustering structure.
-
-The first data set is congressional voting data consisting of voting records
-of the second session of the $98$th United States Congress
-\citep{schlimmer1987concept}.  This data set includes votes for legislators on
-the $P = 16$ key votes.  For each vote, each legislator either votes
-positively (``yea'') or negatively (``nay''). We removed legislators with missing
-votes.  This results in $N = 232$ remaining records, with $124$ democrat and
-$108$ republican. There are $k = 2$ clusters of legislators, corresponding to
-political party.
-
-The second benchmark is the \citet{mangasarian1990pattern} Wisconsin breast
-cancer data set.  After excluding the records with missing data, this data set
-consists records of $N = 683$ patients, each with measurements of $P = 9$
-attributes of their biopsy specimens. It is known that there are $k = 2$
-groups of patients: $444$ patients with benign specimens and $239$ patients
-with malignant specimens. There is some disagreement as to what the ``true''
-value of $k$ should be for this data set; \citet{fujita2014non} have argued
-that the malign group is heterogeneous, and should be split into two smaller
-clusters, yielding $k = 3$.
-
-The third data set is gene expression data of $k = 5$ types of brain tumors
-from \citet{pomeroy2002prediction}, which contains $N = 42$ observations
-including $10$ medulloblastomas, $10$ malignant gliomas, $10$ atypical
-teratoid/rhabdoid tumors, $8$ primitive neuroectodermal tumours and $4$ normal
-cerebella. After preprocessing and feature selection, there are $P = 1379$
-variables, corresponding to log activation levels for 1379 genes.
-
-%% Data was collected with Affymetrix microarrays, which estimates the number of
-%% RNA copies found in the cell sample (frozen specimens). Data was preprocessed
-%% as following: as a common procedure, all genes with expression level below
-%% $10$ are set to a minimum level threshold of $10$. The maximum threshold is
-%% set at $16,000$. Values below or above these thresholds are often unreliable.
-%% To remove uninformative genes, for each gene $j$ (attribute), the mean $m_j$
-%% among the samples was calculated. The $10\%$ largest and smallest values are
-%% discarded in order to remove extreme values. Based on the mean, following
-%% transformation is applied on every value $x_{ij}$ of sample $i$ and gene j \[
-%% y_{ij} = \log_2 (x_{ij}/m_j)	\] 
-%% 
-%% Genes with expression levels differing by at least $l$-flod in at least $c$
-%% samples from their mean expression level across the sample were selected. The
-%% parameter $l$ and $c$ were chosen so as to produce a filtered data set with at
-%% least $10\%$ of the original number of genes \citep{de2008clustering}. Such
-%% preprocessing procedure results in $1379$ genes selected. After filtering
-%% step, data is restored in the original scale (i.e. $x_{ij}$). So the data set
-%% is $42 \times 1379$, which serves as a high dimensional example.
-
-\begin{table}
-\centering
-\captionsetup{justification=centering}
-\caption{\label{tab:benchmark} Number of clusters selected on benchmark datasets}
-\begin{tabular}{lccc}
-\toprule                  
-         & \multicolumn{3}{c}{Dataset} \\
-\cmidrule(l){2-4}
-Method   & Congress Voting & Breast Cancer & Brain Tumours \\
-\midrule                    
-Gabriel  & $2$             & $3$           & $5$  \\    
-Gabriel (corr.~correct) & $2$ & $2$           & $5$  \\  
-Wold     & $2$             & $3$           & $4$  \\
-Gap      & $8$             & $10$          & $10$ \\   
-BIC      & $2$         & $5$           & $2$  \\   
-CH       & $2$             & $2$           & $2$  \\
-Hartigan & $3$             & $3$           & $4$  \\
-Jump     & $10$            & $9$           & $1$  \\   
-PS       & $2$  & $2$           & $1$  \\
-Stab.    & $2$  & $2$           & $7$  \\ 
-Slope	 & $2$  & $4$           & $2$ \\
-\midrule 
-Ground Truth & $2$         & $2$ or $3$    & $5$  \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-We applied the Gabriel cross-validation method, the correlation-corrected
-version, and the competing methods described in Section~\ref{sec:simulations}
-to each of the three benchmark datasets.  In each dataset, we allowed the
-number of clusters, $k$, to range from $1$ to $10$.  Table~\ref{tab:benchmark}
-displays the results.  Both versions of the Gabriel method perform well on all
-three benchmark datasets. In fact, Gabriel cross-validation is the only method
-that correctly identifies the number of clusters in all three benchmark
-datasets.
-
-%% Since most congressmen vote based on their parties' interest, two parties
-%% (Democratic and Republican) represent two clusters in this data set. So the
-%% optimal number should be two. Close inspection shows $k$-means with $k=2$
-%% separates the two parties very well with the lowest miss-classification error
-%% $10.43\%$ ($k=3$ has $14.78\%$). Our proposed Gabriel cross-validation method, along with
-%% several other algorithms correctly pick $k=2$ for this data set. Note that
-%% CH-index and Bootstrap stability also return $k=2$, but $2$ is the lower bound
-%% those methods can select for $k$. So it's not clear they actually choose $k=2$
-%% or they hit the lower bound (they would pick $k=1$ if allowed). 
-%% 
-%% For the breast cancer data, it's known to have $2$ clusters based on whether
-%% it's benign specimens or malignant specimens. Although the proposed Gabriel
-%% method picks $k=3$, its correlation-corrected version correctly picks $k=2$.
-%% It shows the necessity to correct for the high correlation among data
-%% features. However, for this data $k=3$ is also an acceptable answer, as
-%% \cite{fujita2014non} noticed that the malign group is quite heterogeneous and
-%% can be further clustered into at least two subgroups.
-%% 
-%% 
-%% The brain tumours gene expression data is a typical high-dimension data with
-%% $p > n$. It is clear that the correct number of clusters should be $5$, the
-%% number of brain tumour types. Indeed, $k$-means results with $k=5$ show that
-%% the $4$ normal human cerebella are well separated from brain tumours as they
-%% form a single cluster.  Medulloblastomas, malignant gliomas and AT/RT tumours
-%% also form $3$ respective clusters, which means they can be separated from each
-%% other. Such results are similar to those found in
-%% \cite{pomeroy2002prediction}, whose analysis is based on the first $3$
-%% principle components of the original data.    From table \ref{table2} we can
-%% see only our proposed Gabriel method (as well as its correlation-corrected
-%% version) correctly selects $k=5$, underline how difficult it was to pick the
-%% right $k$ when dimension is high.
-
-\section{Application to yeast cell cycle data}
-\label{sec:application-yeast}
-
-\subsection{Motivation}
-
-Now that we have established that Gabriel cross-validation can effectively
-estimate the number of clusters, we apply our method to a yeast cell cycle
-dataset. This dataset was collected by \cite{cho1998genome} to study the cell
-cycle of budding yeast Saccharomyces cerevisiae. Other authors, including
-\citet{tavazoie1999systematic} and \citet{dortet2008model} have used $k$-means
-and related methods to cluster the genes in the dataset, with $k$
-approximately equal to~30. In both of these analyses, the authors discard the
-majority of their clusters as uninterpretable or noise, focusing instead on a
-small number of clusters. In contrast to these previous analyses, Gabriel
-cross-validation finds a small number, $k = 5$ clusters, all of which are
-interpretable.
-
-\subsection{Data collection and preprocessing}
-
-To obtain the raw data, \citet{cho1998genome} first synchronized a collection
-of CDC28 yeast cells by raising their temperature to $37^\circ$C in the late
-G1 cell cycle phase, then they reinitiated the cell cycle by switching them to
-a cooler environment ($25^\circ$C). The authors collected data at $17$ time
-points spaced evenly at $10$-minute intervals, covering almost $2$ complete
-cell cycles. At each of the $17$ time points, they used
-oligonucleotide microarrays to measure $6220$ gene expression profiles.
-
-\citet{tavazoie1999systematic} preprocessed the raw data in an attempt to
-normalize the gene responses and remove noise. They reduced the original
-$6220$ gene expression profiles to just the $2945$ genes with the highest
-variances. Then, they removed the time points at $90$ and $100$ minutes,
-because they deemed the measurements at these time points to be unreliable.
-Finally, they centered and scaled the genes by subtracting the means and
-dividing by the standard deviations, as computed from the remaining $15$ time
-points.  After the preprocessing, the data matrix $\dataX$ has $N = 2945$
-genes and $P = 15$ time points.
-
-We obtained the preprocessed data and the \citet{tavazoie1999systematic}
-cluster analysis from \url{http://arep.med.harvard.edu/network_discovery/}.
-
-\subsection{Clustering}
-
-\begin{figure}
-		\centering
-	\includegraphics[width=5.7in, height=5.7in]{main_code/data_examples/split_plots.pdf}
-	\caption{Comparison between  \cite{tavazoie1999systematic} clusters and
-    our proposed Gabriel cross-validation clusters for a Yeast gene expression
-    time series data set.  The plots on the vertical and horizontal
-    show the average expression time series for the \citeauthor{tavazoie1999systematic} 
-    clusters (1--4, 7, 8, 14, Other) and the Gabriel clusters (1--5),
-    respectively. Plot $(i,j)$ in the grid show the time series average
-    in the intersectios of  \citeauthor{tavazoie1999systematic} cluster~$i$
-    with Gabriel cluster~$j$, whenever there is overlap. The Gabriel clusters
-    capture the same general trends with a smaller number of clusters.
-    }
-	\label{fig:5_clusters}
-\end{figure}
-
-Following \citet{tavazoie1999systematic} and \citet{dortet2008model}, we
-treat the $N = 2945$ gene expression profiles as draws from a mixture
-distribution, and we perform $k$-means clustering to segment the genes
-according to their expression profiles across the $P = 15$ time points. 
-Both the original and the correlation-corrected version of Gabriel
-cross-validation find $k = 5$ clusters.
-
-The lower-left panel of Figure~\ref{fig:5_clusters} shows the average
-expression level for each cluster across the 15 time points, with error bars
-showing standard deviations. Cluster~$1$ has decreasing expression level with
-time. The mean gene expression level in Cluster~$2$ decreases at the beginning
-and then increases. Cluster~$3$ is a periodic cluster where one can see two
-periods corresponding to the two cell cycles. Cluster~$4$ has increasing
-expression level with time.  Cluster~$5$ is another periodic cluster.
-
-\subsection{Enrichment analysis}
-
-To further validate our clusters, we follow \citet{tavazoie1999systematic},
-performing an enrichment analysis to discover which functional gene groups are
-significantly over-represented in each cluster. In the Saccharomyces Genome
-Database, each gene is mapped to a set of Gene Ontology categories. We focus
-on the $103$ biological process categories.
-
-\begin{table}
-\begin{center}
-\captionsetup{justification=centering}
-\caption{\label{table:enrich} Biological process enrichment within gene clusters}
-\footnotesize
-\begin{tabular}{ccll}
-\toprule
-Cluster & Cluster Size & Process Category (In Cluster/Total Genes) & $p$-value \\
-\midrule
-$1$ & $550$ & response to oxidative stress ($24/55$)  & $1.5\times10^{-5}$  \\
-    &       & response to chemical ($64/213$)         & $2.2\times10^{-5}$  \\
-$2$ & $590$ & mitochondrion organization ($79/159$)   & $1.1\times10^{-16}$ \\
-    &       & mitochondrial translation ($28/51$)     & $2.9\times10^{-8}$  \\
-    &       & generation of precursor metabolites and energy ($37/80$) & $7.3\times10^{-8}$ \\
-$3$ & $654$ & transcription from RNA polymerase II promoter ($75/214$) & $5.5\times10^{-6}$ \\
-    &       & mRNA processing ($30/67$)               & $2.7\times10^{-5}$  \\
-    &       & mitotic cell cycle ($63/183$)           & $6.2\times10^{-5}$  \\
-$4$ & $634$ & cytoplasmic translation ($105/134$)     & $3.3\times10^{-47}$ \\
-    &       & ribosomal subunit biogenesis ($73/138$) & $7.7\times10^{-17}$ \\
-    &       & rRNA processing ($61/131$)              & $5.7\times10^{-11}$ \\
-    &       & ribosome assembly ($21/36$)             & $1.5\times10^{-6}$  \\
-$5$ & $517$ & chromosome segregation ($53/106$)       & $6.0\times10^{-15}$ \\
-    &       & cellular response to DNA damage stimulus ($71/172$)      & $3.6\times10^{-14}$ \\
-    &       & DNA repair ($64/147$)                   & $3.7\times10^{-14}$ \\
-    &       & DNA replication ($42/78$)               & $1.8\times10^{-13}$ \\
-    &       & mitotic cell cycle ($70/183$)           & $4.8\times10^{-12}$ \\
-\bottomrule
-\end{tabular}
-\end{center}
-\hspace{0.5in} \footnotesize {}
-\end{table}
-
-For each category and each cluster, we compute a $p$-value for the null
-hypothesis that genes from the category are distributed across all clusters
-without any bias towards the particular cluster in question. Under the null
-hypothesis, the number of genes from the category that end up in the cluster
-is distributed as a hypergeometric random variable. For each cluster, we
-compute $p$-values for all $103$ biological process categories, and we report
-those that are significantly enrigched in Table~\ref{table:enrich}.  Using a
-Bonferroni correction to control the family-wise error rate at level 5\%, we
-only report $p$-values that are less than $0.05 / 103 = 4.8\times10^{-4}$.
-
-From Table \ref{table:enrich}, we can see that Cluster~1 is enriched with
-genes that somatize cell stress, such as oxidative heat-induce proteins.
-Cluster~2 contains genes that govern mitochondrial translation and
-mitochondrion organization. Cluster~3, the first period cluster, contains cell
-cycle genes related to budding and cell polarity, along with genes that govern
-RNA processing and transcription. Cluster~4 contains genes related to
-cytoplasmic translation and genes encoding ribosomes. Cluster~5, the second
-periodic cluster, contains genes that participate cell-cycle processes, along
-with DNA replication and DNA repair.
-
-\subsection{Comparison with Tavazoie~clusters}
-
-In the \citet{tavazoie1999systematic} analysis, those authors performed
-$k$-means clustering with $k = 30$; they found 23 of the clusters to be
-uninterpretable, and they found 7 clusters to be meaningful. To compare our
-clusters with the Tavazoie et al.\ clusters, we prepared a confusion matrix
-comparing our clusters with the 7 interpretable Tavazoie clusters in
-Table~\ref{table:confusion}.
-%%(Appendix~\ref{app:confusion} contains a confusion matrix for all 30
-%% Tavazoie clusters, including the non-interpretable ones.)
-Entry $(i,j)$ of the confusion matrix gives the number of genes in
-Tavazoie's Cluster~$i$ and our Cluster~$j$.
-
-\begin{table}
-\caption{\label{table:confusion} Confusion matrix comparing
-    the 5 clusters found by Gabriel cross-validation to the 7 interpretable
-    clusters found by the \cite{tavazoie1999systematic} analysis}
-\centering
-\begin{tabular}{lcccccr}
-  \toprule
- & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
-  \midrule
-Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
-  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
-  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
-  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
-  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
-  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
-  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
-  Other & 545 & 332 & 448 & 383 & 290 & 1998 \\ 
-  \midrule
-  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ 
-   \bottomrule
-\end{tabular}
-\end{table}
-
-Figure~\ref{fig:5_clusters} provides a more in-depth comparison with the
-Tavazoie clusters, using a graphical confusion matrix.  The plot in cell
-$(i,j)$ of the upper left part of this figure gives the mean expression level
-for genes in the intersection of Tavazoie's Cluster~$i$ and our Cluster~$j$;
-the plots in the margins give the mean expression levels for Tavazoie's
-clusters (top right) and our clusters (bottom left). In
-Figure~\ref{fig:5_clusters}, we only include a plot for cell $(i,j)$ if the
-number of genes in that cell is greater than $20$.
-
-Our Cluster~1 mainly consists of genes that Tavazoie et al.\ found to be in
-uninterpretable clusters. Our Cluster~2 contains high concentrations of
-Tavazoie's Clusters~4 and~8. Our first periodic cluster, Cluster~3, contains
-high concentrations of Tavazoie's Clulsters~3,~7, and~14; this is notable,
-because Tavazoie et al.\ highlighted their Clusters~7 and~14 as being
-periodic. Our Cluster~4 contains almost all of Tavazoie's Cluster~1, along
-with part of Tavazoie's Cluster~4. Finally, our second periodic cluster,
-Cluster~5, contains almost all of Tavazoie's Cluster~2, along with part of
-Tavazoie's Cluster 14; this, again, is notable, because Tavazoie et al.\
-highlited these clusters as being periodic.
-
-For the clusters that Tavazoie et al.\ were able to characterize, our analysis
-broadly agrees with the earlier clustering. The major difference between our
-analysis and that of \citet{tavazoie1999systematic} is that we are able to
-identify meaningful groups of genes with a much smaller value of $k$ ($k = 5$
-instead of $k = 30$), and we are able to interpret all of the clusters
-found by our analysis.
-
-\section{Discussion}
-\label{sec:discussion}
-
-In this paper, we proposed a new approach to estimate the number of clusters
-to be used in $k$-means clustering. The intuition behind our proposed method
-is to transform the unsupervised learning problem into a supervised learning
-problem via a form of Gabriel cross-validation.  We proved that our method is
-self-consistent, and we analyzed its behavior in some special cases of
-Gaussian mixture models.  Using simulations and real data examples, we
-demonstrated that our method has good performance, competitive with existing
-approaches. The simulations and empirical benchmarks demonstrate the
-advantages of our method. In the yeast cell cycle application, our method
-was able to identify meaningful gene groups with a small number of clusters.
-
-There are many other clustering algorithms that get used in practice besides
-$k$-means. We suspect that it should be possible to apply our method in the
-context of a spectral clustering procedure, after transforming by the
-eigenvectors of the Laplacian matrix. For other clustering schemes, including
-versions of hierarchical clustering, we are less certain about the viability
-of Gabriel cross-validation.  It is an open question as to whether Gabriel
-cross-validation can be extended to other clustering methods, and whether such
-extensions will perform well in practice.
-
-For $k$-means clustering, Gabriel cross-validation is competitive with
-other model selection methods, especially in the presence of high-dimensional,
-heterogeneous, or heavy-tailed data.
-
-%% Besides no strong modeling assumption is required, our proposed method is
-%% robust against data set with variance heterogeneity, unequal number of
-%% observations, non-Gaussian noise and high-dimension.  Such robustness is
-%% important in practice because for any given data, it's hard to tell whether
-%% its clusters have different number of observations, is the variance equal for
-%% each cluster, or what underlying noise distribution it has. The weakness of
-%% our proposed method is that its theory assumes only week correlation between
-%% ``predictor'' columns and ``response'' columns. In practice, many data sets
-%% don't exhibit high correlations between columns, where our proposed method can
-%% be safely applied. In case the high correlation does exist, the
-%% correlation-corrected version of Gabriel cross-validation method in section
-%% $4.3$ can be used if we can assume common covariance structure. Other
-%% procedure such as leave out most columns for clustering may also be used to
-%% reduce its effect. However, the theory for such procedure has not been fully
-%% developed and it could be the future research topic. 
-%% 
-%% 
-%% Another situation where our proposed method cannot be directly used is when
-%% the clusters are non-convex. In such situation, $k$-means itself doesn't work
-%% well, for example two concentric circles share the same cluster center
-%% \citep{hastie2009elements}. However, it's possible that our proposed Gabriel
-%% cross-validation method can be used on the transformed data set. In the
-%% concentric circles case where spectral clustering is appropriate, our proposed
-%% method can be applied on the eigenvector subspace of the graph Laplacian
-%% matrix inside the spectral clustering algorithm to find the optimal $k$. This
-%% can also be the future research topic.
-
-\section*{Acknowledgements}
-
-We thank Rob Tibshirani for getting us started on this problem and
-for providing code for some initial simulations. We thank Art~Owen for
-providing us with a summary of the relevant theory on $k$-means clustering,
-and for giving us feedback on our theoretical results.  We also thank Cliff
-Hurvich, Josh Reed, and Jeff~Simonoff, for providing comments on an early
-draft of this manuscript and for suggesting further avenues of inquiry.
-
-\appendix
 
@@
 
-\bibliographystyle{apalike}
-
-\bibliography{references}
 \end{document}

### 2019-06-29 JCGS-R3/cvclust.tex
fix margin, split out appendix

--- 2019-01-11 JCGS-R2/cvclust.tex
@@
 In each fold of our cross-validation procedure, we permute the rows and
-columns of the data matrix and then partition the rows and columns as $N = n +
-m$ and $P = p + q$ for positive integers $n$, $m$, $p$, and $q$.  We treat
+columns of the data matrix and then partition the rows and columns as $N = n + m$ and $P = p + q$ for positive integers $n$, $m$, $p$, and $q$.  We treat
 the first $p$ columns as ``predictors'' and the last $q$ columns as
@@
 
-\appendix
-
-\section{Clustering scree plot examples}
-
-\label{sec:elbow-fail}
-
-The top row of Figure~\ref{fig:elbow}
-displays an example where the elbow in $W_k$ corresponds to the true number $k
-= 4$ of mixture components in the data-generating mechanism. The elbow
-approach is simple and often performs well, but it requires subjective
-judgment as to where the elbow is located, and, as the bottom row of
-Figure~\ref{fig:elbow} demonstrates, the approach can easily fail.
-
-\begin{figure}
-\centering
-\begin{minipage}{\linewidth}
-  \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{main_code/demo/elbow/correct-data.pdf}
-  \end{minipage}
-  \hspace{0.05in}
-  \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{main_code/demo/elbow/correct-withinss.pdf}
-  \end{minipage}
-\end{minipage}
-\begin{minipage}{\linewidth}
-  \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{main_code/demo/elbow/incorrect-data.pdf}
-  \end{minipage}
-  \hspace{0.05in}
-  \begin{minipage}{0.45\linewidth}
-    \includegraphics[width=\linewidth]{main_code/demo/elbow/incorrect-withinss.pdf}
-  \end{minipage}
-\end{minipage}
-\caption{Scree plots for two data sets. Left panels show the sets of
-    two-dimensional data points, generated from four clusters, with plotting
-    symbol indicating the generated cluster. Right panels show the
-    corresponding values of the within-cluster sum of squares $W_k$
-    plotted against the number of clusters, $k$. The scree plot identifies
-    the correct number of clusters in the top row, but fails in the bottom
-    row.}
-\label{fig:elbow}
-\end{figure}
-
-%% \section{Review of consistency results for clustering}
-%% 
-%% Some proposed methods provide theoretical consistency result for choosing $k$.
-%% Assume data is a mixture of $G$ $p$-dimension clusters with equal prior,
-%% identically distributed with common covariance matrix and finite fourth
-%% moments in each dimension, \citet{sugar2003finding} shows the Jump method will
-%% pick the correct $k$ under the conditions that cluster centers are
-%% sufficiently separated and an appropriate transformation is used.
-%% Specifically, let $\Delta\sqrt{p}$ denotes the minimum Euclidean distance
-%% between cluster centers, \citet{sugar2003finding} shows that the jump will be
-%% maximized at $k=G$ given that $\Delta > 6$ and the existence of a positive $Y$
-%% such that
-%% \[
-%%   \left(\frac{p\Delta^2W}{9G}\right)^{-Y}
-%%     + \left(
-%%         W \left[\frac{2^{2H^*(X)}}{K^2_{max}2\pi e}\right]
-%%         - (\frac{\Delta}{6})^2(1-W)
-%%       \right)^{-Y} < 2
-%% \]
-%% and $\left(\frac{p\Delta^2W}{9G}\right)^{-Y} <1/2$, where $H^*(X)$ is the
-%% minimum entropy of the cluster membership over each dimension and $W =1-
-%% \frac{6^4V_{\mathbf{X}}}{(\Delta^2-36)^2}$ with $V_{\mathbf{X}} = \var\left(
-%%   1/p ||\mathbf{X}-\mu_j||^2_{\Gamma^{-1}} \mid \mathbf{X}
-%% \hspace{0.05in}\text{in jth cluster}\right)$
-%% 
-%% \cite{tibshirani2005cluster} proved that the prediction strength method is
-%% consistent under the setting that the observations in $p$-dimension are
-%% uniformly distributed within one unit from their respective cluster centers,
-%% with minimum distance between the $G$ cluster centers is four unit. Under such
-%% well-separated setting, they show that 
-%% \[ ps(G) = 1+ o_p(1), \hspace{0.5in} \underset{k \geq G+1}{\text{sup}} ps(k)\leq \frac{2}{3}+ o_p(1) \]
-%% therefore $\hat{k}$ is consistent in estimating $G$.  When data does come from
-%% mixture of Gaussian distribution, the model-based method of
-%% \citet{fraley2002model} is consistent in estimating $k$ in low dimension.
-%% 
-%% Let $\Psi$ denotes any given base-clustering algorithm (e.g $k$-means), and
-%% $\Psi(Z,k)$ the clustering obtained by applying $\Psi$ on data $Z$ with
-%% parameter $k$. Also let $k_0$ be the optimal number of cluster and
-%% $d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ be the measure of distance between clustering
-%% $\Psi(Z_1,k)$ and $\Psi(Z_2,k)$ where $Z_1$ and $Z_2$ are two independent
-%% samples from population. \citet{wang2010consistent} shows that their proposed
-%% method will consistently select $k = k_0$, given such $k_0$ exists and
-%% $d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ converges to zero at proper rate $r_k$.
-%% However, $k_0$ is defined as the most stable number for $\Psi$ among all the
-%% possible $k$, instead of the number of components in mixture model or
-%% well-separated compact regions commonly seen in literature. Therefore, their
-%% consistent result actually is convergent result, i.e. the selected $k$
-%% converges to its limit $k_0$ given such limit exists.
-
-\section{Analysis of single cluster in more than two dimensions}
-\label{sec:single-general-dim}
-
-\begin{proposition}\label{prop:single-general-dim}
-Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
-of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^{p+q}$ is an
-independent draw from a mean-zero multivariate normal distribution with 
-covariance matrix $\Sigma = \left( \begin{smallmatrix} \Sigma_{XX} & \Sigma_{XY} \\ 
- \Sigma_{YX} & \Sigma_{YY} \end{smallmatrix}\right)$, with $\Sigma_{YY}$ has leading 
-eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. In this case, the data are drawn
-from a single cluster; the true number of clusters is~$1$.  If  $\frac{\sqrt{\lambda_1}}{2}
- > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}}$,
-then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
-\end{proposition}
-
-\begin{proof}
-Let $X$ and $Y$ be jointly multivariate normal distributed with mean $\mathbf{0}$ and covariance matrix $\Sigma$, i.e.
-\[	(X,Y) \sim \mathcal{N} \left( \mathbf{0}, \Sigma\right)	\]
-where $\Sigma=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} \\  \Sigma_{YX} & \Sigma_{YY} \end{bmatrix}$.
-
-Let $\Sigma_{YY} = U \Lambda U^T$  be the eigendecomposition of $\Sigma_{YY}$,
-with leading eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. Then
-the centroid of $k$-means applying on $(y_1,..,y_n)$ is on the first principal
-component
-of $Y$,\[	E(u^T_1 Y|u^T_1 Y>0) = \bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1\] and 
-\[	E(u^T_1 Y|u^T_1 Y<0) = \bar{\mu}^Y_2 =-\sqrt{2 \lambda_1/\pi}u_1\]
- where $u^T_1 Y \sim \mathcal{N}(0,\lambda_1)$.
-
-To compute $\bar{\mu}^X_1 = E(X|u^T_1 Y>0)$, we need to know the conditional distribution $X|u^T_1 Y$.
-Since $(X,Y)$ has multivariate normal distribution, $(X,u^T_1 Y)$ also has a multivariate normal distribution with mean $\mathbf{0}$ and covariance matrix
-$$\Sigma_{X,u^T_1 Y}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} u_1 \\  u^T_1 \Sigma_{YX} & \lambda_1 \end{bmatrix}$$
-The conditional distribution $X|u^T_1 Y$ is hence normal with mean
- $$\mu_{X|u^T_1 Y} = \Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y $$
-Therefore, 
-\begin{align}
-\bar{\mu}^X_1 &= E(X \mid u^T_1 Y>0) \nonumber \\ \nonumber
- 			  &= E\left(E[X \mid u^T_1 Y] \mid u^T_1 Y>0\right) \\ \nonumber
- 			  &=  E\left(\Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y \mid u^T_1 Y>0\right)\\ \nonumber
- 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 E(u^T_1 Y \mid u^T_1 Y>0) \\ \nonumber
- 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 \sqrt{2 \lambda_1/\pi} \\ \nonumber
-      &= \sqrt{2 / (\lambda_1 \pi)} \Sigma_{XY}u_1
-\end{align}
-Similar calculation yields $\bar{\mu}^X_2 = -\sqrt{2 / (\lambda_1 \pi)} \Sigma_{XY}u_1$.
-The decision rule to classify any observed value of $X$ to $\bar{\mu}^X_1$ is therefore
-\[	(\bar{\mu}^X_1)^T X >0	\hspace{0.2in}\text{or} \hspace{0.2in} u^T_1\Sigma_{YX}X>0\] 
-Since $u^T_1\Sigma_{YX}X$ is a linear combination of $X$, it also has normal distribution 
-\[	\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)	\]
-And $(Y,u^T_1\Sigma_{YX}X)$ also have multivariate normal distribution with mean $\mathbf{0}$ 
-and covariance matrix
-\[
-\begin{bmatrix}
-\Sigma_{YY} & \Sigma_{YX}\Sigma_{XY}u_1  \\
-u^T_1\Sigma_{YX}\Sigma_{XY} &  u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1
-\end{bmatrix}
-\]
-The conditional distribution of $Y|u^T_1\Sigma_{YX}X$ is also multivariate normal with mean 
-\[	
-\mu_{Y|u^T_1\Sigma_{YX}X } = \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1}u^T_1\Sigma_{YX}X	
-\]
-The $Y$ center for $u^T_1\Sigma_{YX}X>0$ is
-\begin{align}
-\hat{\mu}^Y_1 &= E(Y|u^T_1\Sigma_{YX}X>0) \nonumber \\ \nonumber
- 		      & =  \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) \\ \nonumber
-\end{align}
-Note that $u^T_1\Sigma_{YX}X$ has normal distribution $\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)$, so
-\[
-E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) = \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}
-\]
-Therefore, we have the $Y$ center for $u^T_1\Sigma_{YX}X>0$ be
-\begin{align*}
-\hat{\mu}^Y_1 &= \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1} \hspace{0.1in} \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} \\
-&=\frac{\sqrt{2/\pi}}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} \Sigma_{YX}\Sigma_{XY}u_1
-\end{align*}
-
-Recall that $\bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1$, to judge if $\CV(2) >
-\CV(1)$, one only need to compare the distance between  $\hat{\mu}^Y_1$  and
-$\bar{\mu}^Y_1$ with distance between  $\hat{\mu}^Y_1$ and grand mean $0$. By
-the variance and bias decomposition of prediction MSE, when variance is the
-same, only bias influences the MSE.
-
-After some linear algebra manipulation, we get
-$\|\hat{\mu}^Y_1 - \bar{\mu}^Y_1\|^2 > \|\hat{\mu}^Y_1\|^2$ or $\CV(2) >
-\CV(1)$ if and only if
-\[
- \frac{\sqrt{\lambda_1}}{2} > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} 
-\]
-\end{proof}
-
-The condition that ensures $\CV(1) < \CV(2)$ is tight but difficult to
-understand intuitively. We make understand this condition better by
-noting that it is equivalent to a condition on the correlation between the
-two linear combinations of $X$ and $Y$ variables:
-\begin{equation}\label{eqn:clust-anyd}
-    \cor(u_1^\T \Sigma_{YX} X, u_1^\T Y) < 1/2.
-\end{equation}
-A stronger condition that is easier to understand, then, is that the canonical
-correlation between $X$ and $Y$ is below $1/2$; when the latter is true the
-former follows and $\CV(1) < \CV(2)$. A necessary but not sufficient
-guarantee of the stronger condition is that all correlations between variables
-are below $1/2$.
-
-One might object to the condition~\eqref{eqn:clust-anyd} being stated in terms
-of a particular $(X, Y)$ split of the data.  In practice, one might want the
-condition to hold for all possible splits of the data. A sufficient guarantee
-that~\eqref{eqn:clust-anyd} holds for any $(X, Y)$ split of the data is that
-the data covariance matrix $\Sigma$ is diagonal. The ad-hoc adjustment we
-propose in Sec.~\ref{sec:corr-correct} attempts to transform the data to this
-form.
-
-\section{Analysis of two clusters in more than two dimensions}
-\label{app:twoclust-anyd}
-
-\begin{proposition}\label{prop:twoclust-anyd}
-Suppose that $\{(X_i,Y_i)\}_{i=1}^{n+m}$ is data from a single fold of Gabriel
-    cross-validation, where each $(X,Y)$ pair in with $X \in \mathbb{R}^P$ and $Y \in
-    \mathbb{R}^Q$ is an independent draw
-from an equiprobable mixture of two multivariate normal distributions with
-identity covariance. Suppose that the first mixture component has mean
-$\mu = (\muX, \muY)$ and the second has mean $-\mu = (-\muX, -\muY)$,
-where $\muX \geq 0$ and $\muY \geq 0$. 
-If the cluster centers are well separated, specifically such that
-\(
-  2 \varphi(\muY) + \muY + 2 \muY \Phi(\muY) < 4 \muY \Phi(\muX),
-\)
-then $\CV(2) < \CV(1)$ with probability tending to one as $m$ and $n$ increase.
-\end{proposition}
-
-\begin{proof}
-There are two clusters $G_1$ and $G_2$, where observations from $G_1$ are distributed as
-\[	\mathcal{N}\left( \begin{pmatrix} 
-    \mu^X \\
-    \mu^Y \\
-  \end{pmatrix}, \mathbf{I} \right)	\]
-and observations from $G_2$ are distributed as
-\[	\mathcal{N}\left( \begin{pmatrix} 
-    -\mu^X \\
-    -\mu^Y \\
-  \end{pmatrix}, \mathbf{I} \right)	\]
-here $\mu^X$ and $\mu^Y$ are all vectors. Let $G_i$ be the true cluster where observation $i$ is generated from, by assumption
-\[	P(G_i=G_1) = P(G_i=G_2) = 1/2	\]
-
-To simplify the notation, let 
-\[
- \begin{pmatrix} 
-    X_1 \\
-    Y_1 \\
-  \end{pmatrix}  \sim
-  \mathcal{N}\left( \begin{pmatrix} 
-    \mu^X \\
-    \mu^Y \\
-  \end{pmatrix}, \mathbf{I} \right)
-\]  
-denote the observations from $G_1$ and 
-\[
- \begin{pmatrix} 
-    X_2 \\
-    Y_2 \\
-  \end{pmatrix}  \sim
-  \mathcal{N}\left( \begin{pmatrix} 
-    -\mu^X \\
-    -\mu^Y \\
-  \end{pmatrix}, \mathbf{I} \right)
-\]  
-denote the observations from $G_2$
-
-Further, let's denote $\mu^X = \lambda_x e_x$ where $\lambda_x$ is a scalar denotes the distance of $\mu^X$ from origin and $e_x$ is the unit vector point at the same direction as $\mu^X$; $\mu^Y = \lambda_y e_y$ has the same interpretation.
-
-Apply $K$-mean on the $Y$-space, where the two clusters are $\mathcal{N}\left( \mu^Y, \mathbf{I}\right)$ and  $\mathcal{N}\left( -\mu^Y, \mathbf{I}\right)$, the $K$-mean centroids are $\bar{\mu}^Y_1$ and $\bar{\mu}^Y_2$ with $\bar{\mu}^Y_1 = -\bar{\mu}^Y_2$.
-Note that the boundary between the two clusters are $e^T_y Y > 0 $.
-So
-
-\begin{align}
-\bar{\mu}^Y_1 & = E(Y \mid Y>0) \\
- &= E(Y_1 \mid e^T_y Y_1 > 0) \cdot P(e^T_y Y_1 >0) +E(Y_2 \mid e^T_y Y_2 > 0) \cdot P(e^T_y Y_2 >0)
-\end{align}
-
-Note that $e^T_y Y_1$ projects vector $Y_1$ on the direction of $e_y$.
-And because the $e_y$ is the same direct as $\mu^Y$, it goes through the center of the sphere $\mathcal{N}\left( \mu^Y, \mathbf{I}\right)$.
-Because the covariance matrix is $\mathbf{I}$, the sphere is symmetric around $e_y$.
-Therefore,
-\begin{align}
-E(Y_1 \mid e^T_y Y_1 = a ) & = a e_y
-\end{align}
-
-Also, $Y_1 \sim \mathcal{N}\left( \mu^Y, \mathbf{I}\right)$ so $e^T_y Y_1 \sim \mathcal{N}\left( \lambda_y, 1\right)$. We have
-
-\begin{align}
-E(Y_1 \mid e^T_y Y_1 > 0)  & = E \left[  E\left(  Y_1 \mid e^T_y Y_1  \right) \mid e^T_y Y_1 > 0  \right] \\
- from \hspace{0.05in} (3) \hspace{0.05in} above	& =  E( e^T_y Y_1 e_y  \mid e^T_y Y_1 > 0  )  \\ 
-	& = e_y E( e^T_y Y_1 \mid e^T_y Y_1 > 0  ) 
-\end{align}
-
-Because $e^T_y Y_1 \sim \mathcal{N}\left( \lambda_y, 1\right)  = \lambda_y + Z$, where $Z$ is standard normal, by Lemma $3$ from Appendix C we have 
-\begin{align}
-E( e^T_y Y_1 \mid e^T_y Y_1 > 0  ) &= E(\lambda_y +  Z \mid Z > -  \lambda_y )  \\
- &= \lambda_y + E(Z \mid Z > -  \lambda_y ) \\
- &= \lambda_y + \frac{\varphi(\lambda_y)}{\Phi(\lambda_y)}
-\end{align}
-
- where $\varphi()$ and $\Phi()$ are the standard normal probability and cumulative distribution function respectively. So, by $(6)$ we have
-
-\begin{align}
-E(Y_1 \mid e^T_y Y_1 > 0)  &= \left[ \lambda_y + \frac{\varphi(\lambda_y)}{\Phi(\lambda_y)}  \right] e_y
-\end{align}
-
-Similarly, we can have
-
-\begin{align}
-E(Y_2 \mid e^T_y Y_2 > 0)  &= - E(Y_1 \mid e^T_y Y_1 < 0) \\
-	&= - e_y E( e^T_y Y_1 \mid e^T_y Y_1 < 0  )  \\
-	&= \left[ \frac{\varphi(\lambda_y)}{1-\Phi(\lambda_y)} - \lambda_y  \right] e_y 
-\end{align}
-
-Because $e^T_y Y_1 \sim \mathcal{N}\left( \lambda_y, 1\right)  = \lambda_y + Z$, it's easy to get 
-\begin{align}
-P(e^T_y Y_1 > 0) &= P(Z > - \lambda_y) \\
-	&= \Phi(\lambda_y) 
-\end{align}
-
-By symmetry, we can get 
-\begin{align}
-P(e^T_y Y_2 > 0) &= 1- \Phi(\lambda_y)
-\end{align}
-
-Put everything together, we have
-
-\begin{align}
-\bar{\mu}^Y_1 & = E(Y_1 \mid e^T_y Y_1 > 0) \cdot P(e^T_y Y_1 >0) +E(Y_2 \mid e^T_y Y_2 > 0) \cdot P(e^T_y Y_2 >0) \\
- &=  \left[ \lambda_y + \frac{\varphi(\lambda_y)}{\Phi(\lambda_y)}  \right]\cdot \Phi(\lambda_y) e_y + \left[ \frac{\varphi(\lambda_y)}{1-\Phi(\lambda_y)} - \lambda_y  \right]\cdot\left(1- \Phi(\lambda_y)\right) e_y  \\
- & = [2\lambda_y \Phi(\lambda_y) + 2\varphi(\lambda_y) - \lambda_y]e_y 
-\end{align}
-
-After training the classifier, because of the identity covariance matrix, the classification boundary is $e^T_x X > 0$. So the $Y$ center for observation with $e^T_x X > 0$ is
-
-\begin{align}
-\hat{\mu}^Y_1 &= E(Y_1 \mid e^T_x X_1 > 0)\cdot P(e^T_x X_1 > 0) + E(Y_2 \mid e^T_x X_2 > 0)\cdot P(e^T_x X_2 > 0)\\ 
-\text{ X independent of Y}	&= E(Y_1) \cdot P(e^T_x X_1 > 0) + E(Y_2) \cdot P(e^T_x X_2 > 0)\\ 
-  & = \mu^Y \cdot P(e^T_x X_1 > 0) -  \mu^Y \cdot P(e^T_x X_2 > 0)\\
-  & = \mu^Y \left( P(e^T_x X_1 > 0) -  P(e^T_x X_2 > 0)  \right) \\
-   & = \mu^Y \left[ \Phi(\lambda_x) -  (1-\Phi(\lambda_x))  \right] \\
-   & = (2\Phi(\lambda_x) - 1)\mu^Y \\
-    & = (2\Phi(\lambda_x) - 1)\lambda_y e_y
-\end{align}
-
-Because of symmetry and $P(G_i=G_1) = P(G_i=G_2) = 1/2$, it's sufficient to show that for observations with  $e^T_x X > 0$, if $CV(2) < CV(1)$ then the Gabriel CV method correctly picks $k=2$ over $k=1$.
-
-Similar as in the proof of Proposition~\ref{prop:single-general-dim}, by the variance and bias decomposition of MSE, the variance is the same, so only the bias influences the result.
-Note the predicted center is grand $0$ for $CV(1)$, so to see if  $CV(2) < CV(1)$ one only need to see if $||\bar{\mu}^Y_1-\hat{\mu}^Y_1||^2 < ||\hat{\mu}^Y_1 - 0||^2$, which is true if
-\[ 2\Phi(\lambda_y) + 2 \frac{\varphi(\lambda_y)}{\lambda_y} < 4\Phi(\lambda_x)-1 \] 
-this result reduces to the original result of Proposition~\ref{prop:twoclust}
-if one sets $\lambda_x = \mu^X$ and $\lambda_y = \mu^Y$.
-\end{proof}
-
-\section{Technical Lemmas}
-\label{app:technical-lemmas}
-
-\begin{lemma}\label{lem:truncated-normal-moments}
-
-If $Z$ is a standard normal random variable, then
-\[
-  \E(Z \mid a < Z < b)
-    = - \frac{\varphi(b) - \varphi(a)}
-             {\Phi(b) - \Phi(a)}
-\]
-and
-\[
-  \E\{(Z - \delta)^2 \mid a < Z < b\}
-    = \delta^2 + 1
-    - \frac{  (b - 2 \delta) \varphi(b)
-            - (a - 2 \delta) \varphi(a)}
-           {\Phi(b) - \Phi(a)}
-\]
-for all constants $a$, $b$, and $\delta$, where $\varphi(z)$ and $\Phi(z)$ are
-the standard normal probability density and cumulative distribution functions.
-These expressions are valid for $a = -\infty$ or $b = \infty$ by taking
-limits.
-
-\end{lemma}
-\begin{proof}
-We will derive the expression for the second moment.  Integrate to get
-\begin{align*}
-  \E[ (Z - \delta)^2 1\{Z < b\}]
-    &= \int_{-\infty}^b (z - \delta)^2 \varphi(z) \, dz \\
-    &= (\delta^2 + 1) \Phi(b) - (b - 2 \delta) \varphi(b).
-\end{align*}
-Now,
-\[
-  \E\{(Z - \delta)^2 \mid a < Z < b\}
-    =
-    \frac{  \E[ (Z - \delta)^2 1\{Z < b\}]
-          - \E[ (Z - \delta)^2 1\{Z < a\}]}
-         { \Phi(b) - \Phi(a) }.
-\]
-\end{proof}
-
-Lemma~\ref{lem:truncated-normal-moments} has some important special cases:
-\begin{align*}
-  \E\{Z \mid Z > 0\} &= 2 \varphi(0) = \sqrt{2 / \pi}, \\
-  \E\{(Z - \delta)^2 \mid Z > 0 \}
-    &= \delta^2 + 1 - 4 \delta \varphi(0), \\
-  \E\{(Z - \delta)^2 \mid Z < 0 \}
-    &= \delta^2 + 1 + 4 \delta \varphi(0).
-\end{align*}
-
-\section{Wold cross-validation}
-\label{sec:wold-cv}
-
-In Wold cross-validation, we perform ``speckled'' hold-outs in each fold, leaving out a
-random subset of the entries of the data matrix $\dataX \in \R^{N \times P}$. For each value
-of $k$ and each fold, we perform the following set of actions to get an
-estimate of cross-validation error, $\CV(k)$, which we average over all folds.
-
-\begin{enumerate}
-  \item Randomly partition the set of indices $\{ 1, 2, \dotsc, N \} \times \{
-    1, 2, \dotsc, P \}$ into a train set $S_\text{train}$ and a test set
-    $S_\text{test}$.
-
-  \item Apply a $k$-means fitting procedure that can handle missing data to
-    the training data $\{ \dataX_{i,j} : (i,j) \in S_{\text{train}} \}$. This
-    gives a set of cluster means $\mu(1), \dotsc, \mu(k) \in \R^P$ and
-    cluster labels for the rows, $G_1, G_2, \dotsc, G_N$.
-
-  \item Compute the cross-validation error as
-    \[
-      \CV(k) = \sum_{(i,j) \in S_{\text{test}}} \{ \dataX_{i,j} - \mu_j(G_i) \}^2,
-    \]
-    where $\mu_j(G_i)$ denotes the $j$th component of $\mu(G_i)$.
-\end{enumerate}
-
-%% \section{Confusion matrix} \label{app:confusion}
-%% 
-%% \begin{table}
-%%   \footnotesize
-%% %\captionsetup{justification=centering}
-%% \caption{\label{table:confusion-complete} Complete confusion matrix comparing
-%% $5$ clusters to $30$ clusters }
-%% \begin{tabular}{lccccc|c}
-%%   \toprule
-%%  & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
-%%   \midrule
-%%   Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
-%%   Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
-%%   Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
-%%   Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
-%%   Cluster 5 & 152 & 0 & 0 & 0 & 0 & 152 \\ 
-%%   Cluster 6 & 1 & 2 & 12 & 85 & 4 & 104 \\ 
-%%   Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
-%%   Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
-%%   Cluster 9 & 142 & 0 & 0 & 0 & 5 & 147 \\ 
-%%   Cluster 10 & 12 & 27 & 24 & 20 & 6 & 89 \\ 
-%%   Cluster 11 & 0 & 5 & 81 & 8 & 0 & 94 \\ 
-%%   Cluster 12 & 3 & 68 & 5 & 3 & 1 & 80 \\ 
-%%   Cluster 13 & 0 & 1 & 12 & 81 & 5 & 99 \\ 
-%%   Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
-%%   Cluster 15 & 60 & 53 & 2 & 0 & 0 & 115 \\ 
-%%   Cluster 16 & 7 & 0 & 0 & 9 & 83 & 99 \\ 
-%%   Cluster 17 & 9 & 0 & 72 & 1 & 1 & 83 \\ 
-%%   Cluster 18 & 3 & 16 & 10 & 67 & 5 & 101 \\ 
-%%   Cluster 19 & 62 & 2 & 0 & 1 & 8 & 73 \\ 
-%%   Cluster 20 & 11 & 0 & 6 & 2 & 65 & 84 \\ 
-%%   Cluster 21 & 1 & 13 & 6 & 44 & 6 & 70 \\ 
-%%   Cluster 22 & 33 & 15 & 28 & 9 & 0 & 85 \\ 
-%%   Cluster 23 & 2 & 18 & 2 & 25 & 22 & 69 \\ 
-%%   Cluster 24 & 1 & 38 & 45 & 1 & 0 & 85 \\ 
-%%   Cluster 25 & 2 & 57 & 16 & 1 & 0 & 76 \\ 
-%%   Cluster 26 & 5 & 3 & 23 & 19 & 0 & 50 \\ 
-%%   Cluster 27 & 9 & 0 & 30 & 5 & 20 & 64 \\ 
-%%   Cluster 28 & 0 & 7 & 23 & 0 & 38 & 68 \\ 
-%%   Cluster 29 & 27 & 2 & 10 & 1 & 11 & 51 \\ 
-%%   Cluster 30 & 3 & 5 & 41 & 1 & 10 & 60 \\ \midrule
-%%   Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ \bottomrule
-%% \end{tabular}
-%% \end{table}
-
 \bibliographystyle{apalike}

### 2019-06-29 JCGS-R3/appendix.tex
move self-consistency to appendix

@@
 \end{figure}
+
+\section{Self-consistency}
+\label{sec:self-consistent}
+
+An important property of any estimation procedure is that in the absence of
+noise, the procedure correctly estimates the truth. This property is called
+``self-consistency'' \citep{tarpey96}. We will now show that Gabriel
+cross-validation is self-consistent. That is, in the absence of noise, the
+Gabriel cross-validation procedure finds the optimal number of clusters.
+
+Self-consistency by itself does not mean that a procedure performs well in
+practice. Self-consistency deals only with the no-noise situation.  Many other
+methods for selecting $K$ are likely self-consistent, but not all perform the
+same in practice. Self-consistency only suggests (but does not guarantee) that
+a procedure might perform reasonably when the noise level is low.
+
+It will suffice to prove self-consistency for a single fold of the
+cross-validation procedure.  As in section~\ref{sec:gabriel-cv-algorithm} we
+assume that the $P$ variables of the data set have been partitioned into $p$
+predictor variables represented in vector~$X$ and $q$ response variables
+represented in vector~$Y$.  The $N$ observations have been divided into two
+sets: $n$ train observations and $m$ test observations.  We state the
+assumptions for the self-consistency result in terms of a specific split; for
+the result to hold in general, with high probability, these assumptions would
+have to hold with high probability for a random split. The following theorem
+gives conditions for Gabriel cross-validation to recover the true number of
+clusters in the absence of noise.
+
+\begin{proposition}\label{prop:self-consistency}
+
+Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
+cross-validation.  For any $k$, let $\CV(k)$ be the cross-validation error for
+this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
+We will assume that there are $K$ ``true cluster centers''
+$\mu(1), \dotsc,\mu(K)$, with
+the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
+\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
+\begin{enumerate}[label=(\roman*)]
+  \item \label{asn:self-consistency-noiseless}
+    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
+    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
+    $i = 1, \dotsc, n+m$.
+  \item \label{asn:self-consistency-distinct-mux}
+    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
+  \item \label{asn:self-consistency-distinct-muy}
+    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
+  \item \label{asn:self-consistency-train}
+    The training set contains at least one member of each cluster: for all $g$
+    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
+    $1, \dotsc, n$ such that $G_i = g$.
+  \item \label{asn:self-consistency-test}
+    The test set contains at least one member of each cluster: for all $g$ in
+    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
+    \dotsc, n+m$ such that $G_i = g$.
+\end{enumerate}
+Then $\CV(k) > \CV(K)$ for $k < K$, and $\CV(k) = \CV(K)$ for $k > K$, so that
+Gabriel cross-validation correctly chooses $k = K$.
+\end{proposition}
+
+The proposition states that our method works well in the absence of noise,
+when each observation is equal to its cluster center.  The assumptions here
+are not checkable in practice, but the proposition suggest that Gabriel
+cross-validation might give a reasonable answer, at least when the noise level
+is low.
+
+The essential
+assumption here is assumption~\ref{asn:self-consistency-noiseless}, which states
+that there is no noise.  If we are willing to assume, say, that the cluster
+centers~$\mu(g) = \bigl(\muX(g),\muY(g)\bigr)$ for $g = 1, \dotsc, K$ were
+randomly drawn from a distribution with a density over $\R^{p+q}$, then
+assumptions~\ref{asn:self-consistency-distinct-mux}
+and~\ref{asn:self-consistency-distinct-muy} will hold with probability one for
+all splits of the data. Likewise, if the clusters are not too small (relative
+to $n$ and $m$), then assumptions~\ref{asn:self-consistency-train}
+and~\ref{asn:self-consistency-test} will likely hold for a random split of the
+data into test and train.
+
+Proposition~\ref{prop:self-consistency} follows from
+Lemmas~\ref{lem:self-consistency1} and~\ref{lem:self-consistency2}, which we
+now state and prove.
+
+\begin{lemma}\label{lem:self-consistency1}
+Suppose that the assumptions of Proposition~\ref{prop:self-consistency} are in
+force.  If $k < K$, then $\CV(k) > 0$.
+\end{lemma}
+\begin{proof}
+By definition,
+\[
+  \CV(k)
+    =
+      \sum_{i=n+1}^{n+m}
+        \| Y_i - \bmuY (\hGX_i) \|^2,
+\]
+where $\bmuY(g)$ is the center of cluster $g$ returned from applying $k$-means
+to $Y_1, \dotsc, Y_n$.  Assumptions~\ref{asn:self-consistency-noiseless}
+and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
+set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all distinct values in
+$\{ \muY(1), \dotsc, \muY(K) \}$.
+Assumption~\ref{asn:self-consistency-distinct-muy} implies that there are
+exactly $K$ such distinct values.  However, there are only $k$ distinct values
+of $\bmuY(g)$.  Thus, at least one summand
+\(
+  \| Y_i - \bmuY(\hGX_i) \|^2
+\)
+is nonzero.  Therefore,
+\(
+  \CV(k) > 0.
+\)
+\end{proof}
+
+\begin{lemma}\label{lem:self-consistency2}
+Suppose that the assumptions of Proposition~\ref{prop:self-consistency} are in
+force.  If $k \geq K$, then $\CV(k) = 0$.
+\end{lemma}
+\begin{proof}
+From assumptions~\ref{asn:self-consistency-noiseless},
+\ref{asn:self-consistency-distinct-muy},
+and~\ref{asn:self-consistency-train}, we know the cluster centers
+gotten from applying $k$-means to $Y_1, \dotsc, Y_n$ must include
+$\muY(1), \dotsc, \muY(K)$.  Without loss of generality, suppose that
+$\bmuY(g) = \muY(g)$ for $g = 1, \dotsc, K$.  This implies that
+$\hGY_i = G_i$ for $i = 1, \dotsc, n$.  Thus, employing
+assumption~\ref{asn:self-consistency-noiseless} again, we get that
+$\bmuX(g) = \muX(g)$ for $g = 1, \dotsc, K$.
+
+Since assumption~\ref{asn:self-consistency-distinct-mux} ensures that
+$\muX(1), \dotsc, \muX(K)$ are all distinct, we must have that $\hGX_i = G_i$
+for all $i = 1, \dotsc, m+n$.  In particular, this implies that $\bmuY(\hGX_i)
+= Y_i$ for $i = 1, \dotsc, m+n$, so that $\CV(k) = 0$.
+\end{proof}
 

### 2019-06-29 JCGS-R3/cvclust.tex
move self-consistency to appendix

@@
 
+%% TODO: appendix
+
 In Section~\ref{sec:self-consistent}, we prove that this procedure is
@@
 noise.
-
-\section{Self-consistency}
-\label{sec:self-consistent}
-
-An important property of any estimation procedure is that in the absence of
-noise, the procedure correctly estimates the truth. This property is called
-``self-consistency'' \citep{tarpey96}. We will now show that Gabriel
-cross-validation is self-consistent. That is, in the absence of noise, the
-Gabriel cross-validation procedure finds the optimal number of clusters.
-
-Self-consistency by itself does not mean that a procedure performs well in
-practice. Self-consistency deals only with the no-noise situation.  Many other
-methods for selecting $K$ are likely self-consistent, but not all perform the
-same in practice. Self-consistency only suggests (but does not guarantee) that
-a procedure might perform reasonably when the noise level is low.
-
-It will suffice to prove self-consistency for a single fold of the
-cross-validation procedure.  As in section~\ref{sec:gabriel-cv-algorithm} we
-assume that the $P$ variables of the data set have been partitioned into $p$
-predictor variables represented in vector~$X$ and $q$ response variables
-represented in vector~$Y$.  The $N$ observations have been divided into two
-sets: $n$ train observations and $m$ test observations.  We state the
-assumptions for the self-consistency result in terms of a specific split; for
-the result to hold in general, with high probability, these assumptions would
-have to hold with high probability for a random split. The following theorem
-gives conditions for Gabriel cross-validation to recover the true number of
-clusters in the absence of noise.
-
-\begin{proposition}\label{prop:self-consistency}
-
-Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
-cross-validation.  For any $k$, let $\CV(k)$ be the cross-validation error for
-this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
-We will assume that there are $K$ ``true cluster centers''
-$\mu(1), \dotsc,\mu(K)$, with
-the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
-\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
-\begin{enumerate}[label=(\roman*)]
-  \item \label{asn:self-consistency-noiseless}
-    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
-    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
-    $i = 1, \dotsc, n+m$.
-  \item \label{asn:self-consistency-distinct-mux}
-    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
-  \item \label{asn:self-consistency-distinct-muy}
-    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
-  \item \label{asn:self-consistency-train}
-    The training set contains at least one member of each cluster: for all $g$
-    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
-    $1, \dotsc, n$ such that $G_i = g$.
-  \item \label{asn:self-consistency-test}
-    The test set contains at least one member of each cluster: for all $g$ in
-    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
-    \dotsc, n+m$ such that $G_i = g$.
-\end{enumerate}
-Then $\CV(k) > \CV(K)$ for $k < K$, and $\CV(k) = \CV(K)$ for $k > K$, so that
-Gabriel cross-validation correctly chooses $k = K$.
-\end{proposition}
-
-The proposition states that our method works well in the absence of noise,
-when each observation is equal to its cluster center.  The assumptions here
-are not checkable in practice, but the proposition suggest that Gabriel
-cross-validation might give a reasonable answer, at least when the noise level
-is low.
-
-The essential
-assumption here is assumption~\ref{asn:self-consistency-noiseless}, which states
-that there is no noise.  If we are willing to assume, say, that the cluster
-centers~$\mu(g) = \bigl(\muX(g),\muY(g)\bigr)$ for $g = 1, \dotsc, K$ were
-randomly drawn from a distribution with a density over $\R^{p+q}$, then
-assumptions~\ref{asn:self-consistency-distinct-mux}
-and~\ref{asn:self-consistency-distinct-muy} will hold with probability one for
-all splits of the data. Likewise, if the clusters are not too small (relative
-to $n$ and $m$), then assumptions~\ref{asn:self-consistency-train}
-and~\ref{asn:self-consistency-test} will likely hold for a random split of the
-data into test and train.
-
-Proposition~\ref{prop:self-consistency} follows from
-Lemmas~\ref{lem:self-consistency1} and~\ref{lem:self-consistency2}, which we
-now state and prove.
-
-\begin{lemma}\label{lem:self-consistency1}
-Suppose that the assumptions of Proposition~\ref{prop:self-consistency} are in
-force.  If $k < K$, then $\CV(k) > 0$.
-\end{lemma}
-\begin{proof}
-By definition,
-\[
-  \CV(k)
-    =
-      \sum_{i=n+1}^{n+m}
-        \| Y_i - \bmuY (\hGX_i) \|^2,
-\]
-where $\bmuY(g)$ is the center of cluster $g$ returned from applying $k$-means
-to $Y_1, \dotsc, Y_n$.  Assumptions~\ref{asn:self-consistency-noiseless}
-and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
-set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all distinct values in
-$\{ \muY(1), \dotsc, \muY(K) \}$.
-Assumption~\ref{asn:self-consistency-distinct-muy} implies that there are
-exactly $K$ such distinct values.  However, there are only $k$ distinct values
-of $\bmuY(g)$.  Thus, at least one summand
-\(
-  \| Y_i - \bmuY(\hGX_i) \|^2
-\)
-is nonzero.  Therefore,
-\(
-  \CV(k) > 0.
-\)
-\end{proof}
-
-\begin{lemma}\label{lem:self-consistency2}
-Suppose that the assumptions of Proposition~\ref{prop:self-consistency} are in
-force.  If $k \geq K$, then $\CV(k) = 0$.
-\end{lemma}
-\begin{proof}
-From assumptions~\ref{asn:self-consistency-noiseless},
-\ref{asn:self-consistency-distinct-muy},
-and~\ref{asn:self-consistency-train}, we know the cluster centers
-gotten from applying $k$-means to $Y_1, \dotsc, Y_n$ must include
-$\muY(1), \dotsc, \muY(K)$.  Without loss of generality, suppose that
-$\bmuY(g) = \muY(g)$ for $g = 1, \dotsc, K$.  This implies that
-$\hGY_i = G_i$ for $i = 1, \dotsc, n$.  Thus, employing
-assumption~\ref{asn:self-consistency-noiseless} again, we get that
-$\bmuX(g) = \muX(g)$ for $g = 1, \dotsc, K$.
-
-Since assumption~\ref{asn:self-consistency-distinct-mux} ensures that
-$\muX(1), \dotsc, \muX(K)$ are all distinct, we must have that $\hGX_i = G_i$
-for all $i = 1, \dotsc, m+n$.  In particular, this implies that $\bmuY(\hGX_i)
-= Y_i$ for $i = 1, \dotsc, m+n$, so that $\CV(k) = 0$.
-\end{proof}
 

### 2019-06-29 JCGS-R3/appendix.tex
move cluster material to appendix

@@
 
+\section{Enrichment analysis}
+
+To further validate our clusters, we follow \citet{tavazoie1999systematic},
+performing an enrichment analysis to discover which functional gene groups are
+significantly over-represented in each cluster. In the Saccharomyces Genome
+Database, each gene is mapped to a set of Gene Ontology categories. We focus
+on the $103$ biological process categories.
+
+\begin{table}
+\begin{center}
+\captionsetup{justification=centering}
+\caption{\label{table:enrich} Biological process enrichment within gene clusters}
+\footnotesize
+\begin{tabular}{ccll}
+\toprule
+Cluster & Cluster Size & Process Category (In Cluster/Total Genes) & $p$-value \\
+\midrule
+$1$ & $550$ & response to oxidative stress ($24/55$)  & $1.5\times10^{-5}$  \\
+    &       & response to chemical ($64/213$)         & $2.2\times10^{-5}$  \\
+$2$ & $590$ & mitochondrion organization ($79/159$)   & $1.1\times10^{-16}$ \\
+    &       & mitochondrial translation ($28/51$)     & $2.9\times10^{-8}$  \\
+    &       & generation of precursor metabolites and energy ($37/80$) & $7.3\times10^{-8}$ \\
+$3$ & $654$ & transcription from RNA polymerase II promoter ($75/214$) & $5.5\times10^{-6}$ \\
+    &       & mRNA processing ($30/67$)               & $2.7\times10^{-5}$  \\
+    &       & mitotic cell cycle ($63/183$)           & $6.2\times10^{-5}$  \\
+$4$ & $634$ & cytoplasmic translation ($105/134$)     & $3.3\times10^{-47}$ \\
+    &       & ribosomal subunit biogenesis ($73/138$) & $7.7\times10^{-17}$ \\
+    &       & rRNA processing ($61/131$)              & $5.7\times10^{-11}$ \\
+    &       & ribosome assembly ($21/36$)             & $1.5\times10^{-6}$  \\
+$5$ & $517$ & chromosome segregation ($53/106$)       & $6.0\times10^{-15}$ \\
+    &       & cellular response to DNA damage stimulus ($71/172$)      & $3.6\times10^{-14}$ \\
+    &       & DNA repair ($64/147$)                   & $3.7\times10^{-14}$ \\
+    &       & DNA replication ($42/78$)               & $1.8\times10^{-13}$ \\
+    &       & mitotic cell cycle ($70/183$)           & $4.8\times10^{-12}$ \\
+\bottomrule
+\end{tabular}
+\end{center}
+\hspace{0.5in} \footnotesize {}
+\end{table}
+
+For each category and each cluster, we compute a $p$-value for the null
+hypothesis that genes from the category are distributed across all clusters
+without any bias towards the particular cluster in question. Under the null
+hypothesis, the number of genes from the category that end up in the cluster
+is distributed as a hypergeometric random variable. For each cluster, we
+compute $p$-values for all $103$ biological process categories, and we report
+those that are significantly enrigched in Table~\ref{table:enrich}.  Using a
+Bonferroni correction to control the family-wise error rate at level 5\%, we
+only report $p$-values that are less than $0.05 / 103 = 4.8\times10^{-4}$.
+
+From Table \ref{table:enrich}, we can see that Cluster~1 is enriched with
+genes that somatize cell stress, such as oxidative heat-induce proteins.
+Cluster~2 contains genes that govern mitochondrial translation and
+mitochondrion organization. Cluster~3, the first period cluster, contains cell
+cycle genes related to budding and cell polarity, along with genes that govern
+RNA processing and transcription. Cluster~4 contains genes related to
+cytoplasmic translation and genes encoding ribosomes. Cluster~5, the second
+periodic cluster, contains genes that participate cell-cycle processes, along
+with DNA replication and DNA repair.
+
+\section{Comparison with Tavazoie~clusters}
+
+In the \citet{tavazoie1999systematic} analysis, those authors performed
+$k$-means clustering with $k = 30$; they found 23 of the clusters to be
+uninterpretable, and they found 7 clusters to be meaningful. To compare our
+clusters with the Tavazoie et al.\ clusters, we prepared a confusion matrix
+comparing our clusters with the 7 interpretable Tavazoie clusters in
+Table~\ref{table:confusion}.
+%%(Appendix~\ref{app:confusion} contains a confusion matrix for all 30
+%% Tavazoie clusters, including the non-interpretable ones.)
+Entry $(i,j)$ of the confusion matrix gives the number of genes in
+Tavazoie's Cluster~$i$ and our Cluster~$j$.
+
+\begin{table}
+\caption{\label{table:confusion} Confusion matrix comparing
+    the 5 clusters found by Gabriel cross-validation to the 7 interpretable
+    clusters found by the \cite{tavazoie1999systematic} analysis}
+\centering
+\begin{tabular}{lcccccr}
+  \toprule
+ & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
+  \midrule
+Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
+  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
+  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
+  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
+  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
+  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
+  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
+  Other & 545 & 332 & 448 & 383 & 290 & 1998 \\ 
+  \midrule
+  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ 
+   \bottomrule
+\end{tabular}
+\end{table}
+
+Figure~\ref{fig:5_clusters} provides a more in-depth comparison with the
+Tavazoie clusters, using a graphical confusion matrix.  The plot in cell
+$(i,j)$ of the upper left part of this figure gives the mean expression level
+for genes in the intersection of Tavazoie's Cluster~$i$ and our Cluster~$j$;
+the plots in the margins give the mean expression levels for Tavazoie's
+clusters (top right) and our clusters (bottom left). In
+Figure~\ref{fig:5_clusters}, we only include a plot for cell $(i,j)$ if the
+number of genes in that cell is greater than $20$.
+
+Our Cluster~1 mainly consists of genes that Tavazoie et al.\ found to be in
+uninterpretable clusters. Our Cluster~2 contains high concentrations of
+Tavazoie's Clusters~4 and~8. Our first periodic cluster, Cluster~3, contains
+high concentrations of Tavazoie's Clulsters~3,~7, and~14; this is notable,
+because Tavazoie et al.\ highlighted their Clusters~7 and~14 as being
+periodic. Our Cluster~4 contains almost all of Tavazoie's Cluster~1, along
+with part of Tavazoie's Cluster~4. Finally, our second periodic cluster,
+Cluster~5, contains almost all of Tavazoie's Cluster~2, along with part of
+Tavazoie's Cluster 14; this, again, is notable, because Tavazoie et al.\
+highlited these clusters as being periodic.
+
+For the clusters that Tavazoie et al.\ were able to characterize, our analysis
+broadly agrees with the earlier clustering. The major difference between our
+analysis and that of \citet{tavazoie1999systematic} is that we are able to
+identify meaningful groups of genes with a much smaller value of $k$ ($k = 5$
+instead of $k = 30$), and we are able to interpret all of the clusters
+found by our analysis.
+
 \end{document}

### 2019-06-29 JCGS-R3/cvclust.tex
move cluster material to appendix

@@
 
-\subsection{Enrichment analysis}
-
-To further validate our clusters, we follow \citet{tavazoie1999systematic},
-performing an enrichment analysis to discover which functional gene groups are
-significantly over-represented in each cluster. In the Saccharomyces Genome
-Database, each gene is mapped to a set of Gene Ontology categories. We focus
-on the $103$ biological process categories.
-
-\begin{table}
-\begin{center}
-\captionsetup{justification=centering}
-\caption{\label{table:enrich} Biological process enrichment within gene clusters}
-\footnotesize
-\begin{tabular}{ccll}
-\toprule
-Cluster & Cluster Size & Process Category (In Cluster/Total Genes) & $p$-value \\
-\midrule
-$1$ & $550$ & response to oxidative stress ($24/55$)  & $1.5\times10^{-5}$  \\
-    &       & response to chemical ($64/213$)         & $2.2\times10^{-5}$  \\
-$2$ & $590$ & mitochondrion organization ($79/159$)   & $1.1\times10^{-16}$ \\
-    &       & mitochondrial translation ($28/51$)     & $2.9\times10^{-8}$  \\
-    &       & generation of precursor metabolites and energy ($37/80$) & $7.3\times10^{-8}$ \\
-$3$ & $654$ & transcription from RNA polymerase II promoter ($75/214$) & $5.5\times10^{-6}$ \\
-    &       & mRNA processing ($30/67$)               & $2.7\times10^{-5}$  \\
-    &       & mitotic cell cycle ($63/183$)           & $6.2\times10^{-5}$  \\
-$4$ & $634$ & cytoplasmic translation ($105/134$)     & $3.3\times10^{-47}$ \\
-    &       & ribosomal subunit biogenesis ($73/138$) & $7.7\times10^{-17}$ \\
-    &       & rRNA processing ($61/131$)              & $5.7\times10^{-11}$ \\
-    &       & ribosome assembly ($21/36$)             & $1.5\times10^{-6}$  \\
-$5$ & $517$ & chromosome segregation ($53/106$)       & $6.0\times10^{-15}$ \\
-    &       & cellular response to DNA damage stimulus ($71/172$)      & $3.6\times10^{-14}$ \\
-    &       & DNA repair ($64/147$)                   & $3.7\times10^{-14}$ \\
-    &       & DNA replication ($42/78$)               & $1.8\times10^{-13}$ \\
-    &       & mitotic cell cycle ($70/183$)           & $4.8\times10^{-12}$ \\
-\bottomrule
-\end{tabular}
-\end{center}
-\hspace{0.5in} \footnotesize {}
-\end{table}
-
-For each category and each cluster, we compute a $p$-value for the null
-hypothesis that genes from the category are distributed across all clusters
-without any bias towards the particular cluster in question. Under the null
-hypothesis, the number of genes from the category that end up in the cluster
-is distributed as a hypergeometric random variable. For each cluster, we
-compute $p$-values for all $103$ biological process categories, and we report
-those that are significantly enrigched in Table~\ref{table:enrich}.  Using a
-Bonferroni correction to control the family-wise error rate at level 5\%, we
-only report $p$-values that are less than $0.05 / 103 = 4.8\times10^{-4}$.
-
-From Table \ref{table:enrich}, we can see that Cluster~1 is enriched with
-genes that somatize cell stress, such as oxidative heat-induce proteins.
-Cluster~2 contains genes that govern mitochondrial translation and
-mitochondrion organization. Cluster~3, the first period cluster, contains cell
-cycle genes related to budding and cell polarity, along with genes that govern
-RNA processing and transcription. Cluster~4 contains genes related to
-cytoplasmic translation and genes encoding ribosomes. Cluster~5, the second
-periodic cluster, contains genes that participate cell-cycle processes, along
-with DNA replication and DNA repair.
-
-\subsection{Comparison with Tavazoie~clusters}
-
-In the \citet{tavazoie1999systematic} analysis, those authors performed
-$k$-means clustering with $k = 30$; they found 23 of the clusters to be
-uninterpretable, and they found 7 clusters to be meaningful. To compare our
-clusters with the Tavazoie et al.\ clusters, we prepared a confusion matrix
-comparing our clusters with the 7 interpretable Tavazoie clusters in
-Table~\ref{table:confusion}.
-%%(Appendix~\ref{app:confusion} contains a confusion matrix for all 30
-%% Tavazoie clusters, including the non-interpretable ones.)
-Entry $(i,j)$ of the confusion matrix gives the number of genes in
-Tavazoie's Cluster~$i$ and our Cluster~$j$.
-
-\begin{table}
-\caption{\label{table:confusion} Confusion matrix comparing
-    the 5 clusters found by Gabriel cross-validation to the 7 interpretable
-    clusters found by the \cite{tavazoie1999systematic} analysis}
-\centering
-\begin{tabular}{lcccccr}
-  \toprule
- & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
-  \midrule
-Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
-  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
-  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
-  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
-  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
-  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
-  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
-  Other & 545 & 332 & 448 & 383 & 290 & 1998 \\ 
-  \midrule
-  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ 
-   \bottomrule
-\end{tabular}
-\end{table}
-
-Figure~\ref{fig:5_clusters} provides a more in-depth comparison with the
-Tavazoie clusters, using a graphical confusion matrix.  The plot in cell
-$(i,j)$ of the upper left part of this figure gives the mean expression level
-for genes in the intersection of Tavazoie's Cluster~$i$ and our Cluster~$j$;
-the plots in the margins give the mean expression levels for Tavazoie's
-clusters (top right) and our clusters (bottom left). In
-Figure~\ref{fig:5_clusters}, we only include a plot for cell $(i,j)$ if the
-number of genes in that cell is greater than $20$.
-
-Our Cluster~1 mainly consists of genes that Tavazoie et al.\ found to be in
-uninterpretable clusters. Our Cluster~2 contains high concentrations of
-Tavazoie's Clusters~4 and~8. Our first periodic cluster, Cluster~3, contains
-high concentrations of Tavazoie's Clulsters~3,~7, and~14; this is notable,
-because Tavazoie et al.\ highlighted their Clusters~7 and~14 as being
-periodic. Our Cluster~4 contains almost all of Tavazoie's Cluster~1, along
-with part of Tavazoie's Cluster~4. Finally, our second periodic cluster,
-Cluster~5, contains almost all of Tavazoie's Cluster~2, along with part of
-Tavazoie's Cluster 14; this, again, is notable, because Tavazoie et al.\
-highlited these clusters as being periodic.
-
-For the clusters that Tavazoie et al.\ were able to characterize, our analysis
-broadly agrees with the earlier clustering. The major difference between our
-analysis and that of \citet{tavazoie1999systematic} is that we are able to
-identify meaningful groups of genes with a much smaller value of $k$ ($k = 5$
-instead of $k = 30$), and we are able to interpret all of the clusters
-found by our analysis.
+%% TODO: enrichment analysis
+
+%% TODO: mention comparison with Tavazoie clusters
 

### 2019-06-29 JCGS-R3/cvclust.tex
rearrange figures

@@
 
+\begin{figure}
+\centering
+\includegraphics[width=3in]{main_code/demo/overlap/color_plot.pdf}
+\caption{
+    Gabriel cross-validation performance in a two-cluster simulation.
+    For a grid of values for $\muX$ and $\muY$ ranging from $0$ to $3$,
+    we generated data according the manner described in
+    Prop.~\ref{prop:twoclust}, then computed Gabriel cross-validation
+    error $\CV(1)$ and $\CV(2)$. The heat map shows the proportion of times
+    that $\CV(2) < \CV(1)$ with
+    light (blue) indicating that $k = 1$ is preferred to $k = 2$, and
+dark (red) indicating that $k = 2$ is preferred to $k = 1$.
+    The black line is the theoretical boundary determined from
+Proposition~\ref{prop:twoclust}. The simulation confirms our theoretical
+    boundary.}
+\label{fig:overlap-color_plot}
+\end{figure}
+
 In principle the proof of Proposition~\ref{prop:twoclust} generalizes to
@@
 $k=1$ zone.
-
-\begin{figure}
-\centering
-\includegraphics[width=3in]{main_code/demo/overlap/color_plot.pdf}
-\caption{
-    Gabriel cross-validation performance in a two-cluster simulation.
-    For a grid of values for $\muX$ and $\muY$ ranging from $0$ to $3$,
-    we generated data according the manner described in
-    Prop.~\ref{prop:twoclust}, then computed Gabriel cross-validation
-    error $\CV(1)$ and $\CV(2)$. The heat map shows the proportion of times
-    that $\CV(2) < \CV(1)$ with
-    light (blue) indicating that $k = 1$ is preferred to $k = 2$, and
-dark (red) indicating that $k = 2$ is preferred to $k = 1$.
-    The black line is the theoretical boundary determined from
-Proposition~\ref{prop:twoclust}. The simulation confirms our theoretical
-    boundary.}
-\label{fig:overlap-color_plot}
-\end{figure}
 
@@
 \centering
-\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting1/Facet.pdf}
+\includegraphics[width=5.5in, height=2.75in]{main_code/demo/bench/setting1/Facet.pdf}
 \caption{Setting 1 performance as correlation increases, faceted by estimation
@@
 \centering
-\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting2/Facet.pdf}
+\includegraphics[width=5.5in, height=2.75in]{main_code/demo/bench/setting2/Facet.pdf}
 \caption{Setting 2 performance as the number of noise dimensions increases,
@@
 \centering
-\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting3/Facet.pdf}
+\includegraphics[width=5.5in, height=2.75in]{main_code/demo/bench/setting3/Facet.pdf}
 \caption{Setting 3 performance as the number of dimensions increases,
@@
 \centering
-\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting4/Facet.pdf}
+\includegraphics[width=5.5in, height=2.75in]{main_code/demo/bench/setting4/Facet.pdf}
 \caption{Setting 4 performance as the variance heterogeneity increases,
@@
 \centering
-\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting5/Facet.pdf}
+\includegraphics[width=5.5in, height=2.75in]{main_code/demo/bench/setting5/Facet.pdf}
 \caption{Setting 5 performance as the tail heaviness increases,
@@
 \caption{\label{tab:benchmark} Number of clusters selected on benchmark datasets}
+    {\footnotesize
 \begin{tabular}{lccc}
@@
 \end{tabular}
+    }
 \end{table}
@@
 
-\subsection{Data collection and preprocessing}
-
-To obtain the raw data, \citet{cho1998genome} first synchronized a collection
-of CDC28 yeast cells by raising their temperature to $37^\circ$C in the late
-G1 cell cycle phase, then they reinitiated the cell cycle by switching them to
-a cooler environment ($25^\circ$C). The authors collected data at $17$ time
-points spaced evenly at $10$-minute intervals, covering almost $2$ complete
-cell cycles. At each of the $17$ time points, they used
-oligonucleotide microarrays to measure $6220$ gene expression profiles.
-
-\citet{tavazoie1999systematic} preprocessed the raw data in an attempt to
-normalize the gene responses and remove noise. They reduced the original
-$6220$ gene expression profiles to just the $2945$ genes with the highest
-variances. Then, they removed the time points at $90$ and $100$ minutes,
-because they deemed the measurements at these time points to be unreliable.
-Finally, they centered and scaled the genes by subtracting the means and
-dividing by the standard deviations, as computed from the remaining $15$ time
-points.  After the preprocessing, the data matrix $\dataX$ has $N = 2945$
-genes and $P = 15$ time points.
-
-We obtained the preprocessed data and the \citet{tavazoie1999systematic}
-cluster analysis from \url{http://arep.med.harvard.edu/network_discovery/}.
-
-\subsection{Clustering}
-
 \begin{figure}
@@
 \end{figure}
+
+\subsection{Data collection and preprocessing}
+
+To obtain the raw data, \citet{cho1998genome} first synchronized a collection
+of CDC28 yeast cells by raising their temperature to $37^\circ$C in the late
+G1 cell cycle phase, then they reinitiated the cell cycle by switching them to
+a cooler environment ($25^\circ$C). The authors collected data at $17$ time
+points spaced evenly at $10$-minute intervals, covering almost $2$ complete
+cell cycles. At each of the $17$ time points, they used
+oligonucleotide microarrays to measure $6220$ gene expression profiles.
+
+\citet{tavazoie1999systematic} preprocessed the raw data in an attempt to
+normalize the gene responses and remove noise. They reduced the original
+$6220$ gene expression profiles to just the $2945$ genes with the highest
+variances. Then, they removed the time points at $90$ and $100$ minutes,
+because they deemed the measurements at these time points to be unreliable.
+Finally, they centered and scaled the genes by subtracting the means and
+dividing by the standard deviations, as computed from the remaining $15$ time
+points.  After the preprocessing, the data matrix $\dataX$ has $N = 2945$
+genes and $P = 15$ time points.
+
+We obtained the preprocessed data and the \citet{tavazoie1999systematic}
+cluster analysis from \url{http://arep.med.harvard.edu/network_discovery/}.
+
+\subsection{Clustering}
 

### 2019-06-29 JCGS-R3/cvclust.tex
fix self-consistency, wold refs

@@
 performs well, but it requires subjective judgment as to where the elbow is
-located, and as we demonstrate in Appendix~\ref{sec:elbow-fail}, the approach
-can easily fail.  In this report, we propose a new method to choose the number
-of clusters automatically.
+located, and can easily fail (we demonstrate such a failure in Section~1 of
+the supplementary appendix).  In this report, we propose a new method to
+choose the number of clusters automatically.
 
@@
 $k$ with the smallest associated error. We describe this method in detail in
-Section~\ref{sec:meth}.  In Section~\ref{sec:self-consistent}, we prove that
-our method is self-consistent in certain situations, meaning that in the
-absence of noise, it recovers the correct $k$. Then, in
+Section~\ref{sec:meth}.  Then, in
 Section~\ref{sec:theory-gaussian}, we
@@
 
-%% TODO: appendix
-
-In Section~\ref{sec:self-consistent}, we prove that this procedure is
-self-consistent, in the sense that it recovers the correct value of $k$ in the
-absence of noise. Then, in Section~\ref{sec:theory-gaussian}, we analyze some
-of the properties of Gabriel cross-validation in the presence of Gaussian
-noise.
-
 \section{Analysis under Gaussian noise}
@@
 
+We turn now to theoretical properties of Gabriel cross-validation. In
+Section~2 of the supplementary appendix, show that the procedure is
+self-consistent, in the sense that it recovers the correct value of $k$ in the
+absence of noise.  This gives us some assurance that the procedure is
+well-behaved, but we can bolster our confidence and gain insight into its
+workings by analyzing its behavior in the presence of noise. To that end, we
+devote the present section to analysis of the procedure in the presence of
+Gaussian noise.
+
 \subsection{Single cluster in two dimensions}
 
-Proposition~\ref{prop:self-consistency} tells us that the Gabriel cross-validation
-method recovers the true number of clusters when the noise is negligible.
-While this result gives us some assurance that the procedure is well-behaved,
-we can bolster our confidence and gain insight into its workings by analyzing
-its behavior in the presence of noise.  We first study the case of a single
+We first study the case of a single
 cluster in two dimensions with correlated Gaussian noise.
@@
 
-In Appendix~\ref{sec:single-general-dim}, using similar techniques to those used
+In Section~3 of the supplementary appendix,
+%\ref{sec:single-general-dim}
+using similar techniques to those used
 to prove Proposition~\ref{prop:single-2d}, we derive an analogous result for correlated
@@
 In the second line, we have used
-Lemma~\ref{lem:truncated-normal-moments} from
-Appendix~\ref{app:technical-lemmas} to compute the conditional expectations;
+Lemma~3 from Section~5 of the supplementary appendix
+    to compute the conditional expectations;
 $\varphi()$ and $\Phi()$ are the standard normal density and
@@
 dimension; we generalize the result to arbitrary dimension in
-Appendex~\ref{app:twoclust-anyd}.
+Section~4 of the supplementary appendix.
+%Appendex~\ref{app:twoclust-anyd}.
 
@@
 $Q \Lambda^{-1/2} \Gamma^\T \mu(g)$ for $g = 1, \dotsc, k$ is uniformly
-spread across all $P$ features. This ensures that the self-consistency
-conditions on the cluster centers enumerated in Proposition~\ref{prop:self-consistency}
-are likely to hold. The second reason for multiplying by $Q$ is to spread any
+spread across all $P$ features. This ensures that the cluster centers
+are well identified.
+The second reason for multiplying by $Q$ is to spread any
 remaining correlation in the noise evenly (in expectation) across all
@@
 cross-validation on the rows~($m=n/4$).  We also compare with Wold
-cross-validation, which we describe in Appendix~\ref{sec:wold-cv}.
+cross-validation, which we describe in section~6 of the supplementary
+appendix. %Appendix~\ref{sec:wold-cv}.
 

### 2019-06-30 JCGS-R3/cvclust.tex
spacing

@@
   \Sigma^{-1/2} \dataX_i
-    &= \Sigma^{-1/2} \mu(G_i) + \Sigma^{-1/2} \varepsilon_i \\
-    &= \tilde \mu(G_i) + \tilde \varepsilon_i,
+    &= \Sigma^{-1/2} \mu(G_i) + \Sigma^{-1/2} \varepsilon_i 
+    = \tilde \mu(G_i) + \tilde \varepsilon_i,
 \end{align*}
@@
 
-%% TODO: enrichment analysis
-
-%% TODO: mention comparison with Tavazoie clusters
+To validate our clusters, in Section~7 of the supplementary appendix we follow
+\citet{tavazoie1999systematic} by performing an enrichment analysis to
+discover which functional gene groups are significantly over-represented in
+each cluster.  We find that Cluster~1 is enriched with genes that somatize
+cell stress, such as oxidative heat-induce proteins.  Cluster~2 contains genes
+that govern mitochondrial translation and mitochondrion organization.
+Cluster~3, the first period cluster, contains cell cycle genes related to
+budding and cell polarity, along with genes that govern RNA processing and
+transcription. Cluster~4 contains genes related to cytoplasmic translation and
+genes encoding ribosomes. Cluster~5, the second periodic cluster, contains
+genes that participate cell-cycle processes, along with DNA replication and
+DNA repair.
+
+For the clusters that Tavazoie et al.\ were able to characterize, our analysis
+broadly agrees with the earlier clustering (Section~8 of the supplementary
+appendix contains a more detailed comparison). The major difference between our
+analysis and that of \citet{tavazoie1999systematic} is that we are able to
+identify meaningful groups of genes with a much smaller value of $k$ ($k = 5$
+instead of $k = 30$), and we are able to interpret all of the clusters
+found by our analysis.
 

### 2019-06-30 JCGS-R3/cvclust.tex
figure placing

@@
 \caption{Setting 1 performance as correlation increases, faceted by estimation
-    method. For each estimate method, the subplot shows the proportion of
+    method. For each method, the subplot shows the proportion of
     replicates in which estimated number of clusters agrees with the truth
@@
 \caption{Setting 2 performance as the number of noise dimensions increases,
-    faceted by estimation method. For each estimate method, the subplot shows the proportion of
+    faceted by estimation method. For each method, the subplot shows the proportion of
     replicates in which estimated number of clusters agrees with the truth
@@
 \caption{Setting 3 performance as the number of dimensions increases,
-    faceted by estimation method. For each estimate method, the subplot shows the proportion of
+    faceted by estimation method. For each method, the subplot shows the proportion of
     replicates in which estimated number of clusters agrees with the truth
     versus the number of noise dimensions,~$r$. Both variants of Gabriel
-    cross-validation perform with perfect accuracy in the replicates
+    cross-validation have perfect accuracy in the replicates
     with $r \geq 30$ dimensions.}
 
-\caption{Performance as number of dimensions increases.}
 \label{fig:setting3}
@@
 \caption{Setting 4 performance as the variance heterogeneity increases,
-    faceted by estimation method. For each estimate method, the subplot shows the proportion of
+    faceted by estimation method. For each method, the subplot shows the proportion of
     replicates in which estimated number of clusters agrees with the truth
     versus the variance heterogeneity parameter,  $R$. Both variants of Gabriel
-    cross-validation perform with near-perfect accuracy for all values of~$R$.}
+    cross-validation have near-perfect accuracy for all values of~$R$.}
 \label{fig:setting4}
@@
 \caption{Setting 5 performance as the tail heaviness increases,
-    faceted by estimation method. For each estimate method, the subplot shows the proportion of
+    faceted by estimation method. For each method, the subplot shows the proportion of
     replicates in which estimated number of clusters agrees with the truth
@@
     tail heaviness, with slower performance degradation than most of the other
-    estimation methods as the tail heaviness increases.}
+    methods as the tail heaviness increases.}
 \label{fig:setting5}

### 2019-06-30 JCGS-R3/appendix.tex
fix appendix bib refs
fix remaining refs

@@
 It will suffice to prove self-consistency for a single fold of the
-cross-validation procedure.  As in section~\ref{sec:gabriel-cv-algorithm} we
+cross-validation procedure.  As in section~2.2 of the main article, we
+%\ref{sec:gabriel-cv-algorithm} 
 assume that the $P$ variables of the data set have been partitioned into $p$
@@
 cross-validation.  For any $k$, let $\CV(k)$ be the cross-validation error for
-this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
+this fold, computed as described in Section~2.2 of the main article.
+%\ref{sec:gabriel-cv-algorithm}.
 We will assume that there are $K$ ``true cluster centers''
@@
 the data covariance matrix $\Sigma$ is diagonal. The ad-hoc adjustment we
-propose in Sec.~\ref{sec:corr-correct} attempts to transform the data to this
-form.
+propose in Section~4 of the main article%.~\ref{sec:corr-correct} 
+attempts to transform the data to this form.
 
@@
 \[ 2\Phi(\lambda_y) + 2 \frac{\varphi(\lambda_y)}{\lambda_y} < 4\Phi(\lambda_x)-1 \] 
-this result reduces to the original result of Proposition~\ref{prop:twoclust}
+this result reduces to the original result of Proposition~2
+of the main article %\ref{prop:twoclust}
 if one sets $\lambda_x = \mu^X$ and $\lambda_y = \mu^Y$.
@@
 
-Figure~\ref{fig:5_clusters} provides a more in-depth comparison with the
+Figure~8 of the main article %\ref{fig:5_clusters}
+provides a more in-depth comparison with the
 Tavazoie clusters, using a graphical confusion matrix.  The plot in cell
@@
 clusters (top right) and our clusters (bottom left). In
-Figure~\ref{fig:5_clusters}, we only include a plot for cell $(i,j)$ if the
+Figure~8 (main article), %\ref{fig:5_clusters},
+we only include a plot for cell $(i,j)$ if the
 number of genes in that cell is greater than $20$.
@@
 
+\bibliographystyle{apalike}
+\bibliography{references}
+
 \end{document}

### 2019-06-30 JCGS-R3/cvclust.tex
fix remaining refs

@@
 For the clusters that Tavazoie et al.\ were able to characterize, our analysis
-broadly agrees with the earlier clustering (Section~8 of the supplementary
-appendix contains a more detailed comparison). The major difference between our
-analysis and that of \citet{tavazoie1999systematic} is that we are able to
-identify meaningful groups of genes with a much smaller value of $k$ ($k = 5$
-instead of $k = 30$), and we are able to interpret all of the clusters
-found by our analysis.
+broadly agrees with the earlier clustering. We can see this in
+Figure~\ref{fig:5_clusters}, which gives us a graphical confusion matrix
+comparing the two analyses.  The plot in cell $(i,j)$ of the upper left part
+of this figure gives the mean expression level for genes in the intersection
+of Tavazoie's Cluster~$i$ and our Cluster~$j$; the plots in the margins give
+the mean expression levels for Tavazoie's clusters (top right) and our
+clusters (bottom left).  We only include a plot for cell $(i,j)$ if the number
+of genes in that cell is greater than $20$.  Section~8 of the supplementary
+appendix contains a more detailed comparison.
+The major difference between our analysis and that of
+\citet{tavazoie1999systematic} is that we are able to identify meaningful
+groups of genes with a much smaller value of $k$ ($k = 5$ instead of $k =
+30$), and we are able to interpret all of the clusters found by our analysis.
 
